### 2020-08-20 R/IPM practice.Rmd
initialize IPM practice Rmd

---
title: "Heliconia IPM Practice"
author: "Eric R. Scott"
date: "2020-08-20"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(skimr)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Practice constructing an IPM using (a subset of) the [Heliconia demographic data](https://github.com/embruna/HeliconiaDataPaper/tree/master/data_clean).

# Load Data

```{r data, echo=TRUE}
demog <- read_csv(here("analysis", "data", "raw_data", "Ha_survey_with_Zombies.csv"), col_names = TRUE,
               cols(plot = col_character(),
                    bdffp_reserve_no = col_character(),
                    shts = col_integer(),
                    year = col_integer(),
                    infl = col_integer()))

unique(demog$ranch)
unique(demog$code.notes)
unique(demog$code2)
unique(demog$size)

skim(demog)
```

## Data Dictionary

`demog`: Demography data for *Heliconia acuminatata* experiment at BDFFP

- `plot` (character):
- `size` (character): fragment size
- `ranch` (character): "Dimona", "Esteio-Colosso", or "PortoAlegre".  The three ranches that make up BDFFP.
- `bdffp_reserve_no` (character): Official BDFFP reserve designation
- `tag_number` (numeric): plant tag# within plot (not unique)
- `row` (character): spatial location in plot
- `column` (numeric): spatial location in plot
- `year` (numeric): year of survey
- `ht` (numeric): plant height.  Plants not straightened before measurement.
- `shts` (numeric): number of shoots
- `infl` (numeric): number of infloresences
- `code.notes` (character): 
    - "plant missing" = plant was not found in that survey.  May not be dead, may be found in subsequent surveys.
    - "dead" = plant lost all above-ground tissue.  May resprout in a subsequent year.
    - "ULY" = 
    - "sdlg" = seedling, defined as a year 1 plant.
    - "tag missing" = ID tag labeling the plant was missing.  Plants were given new tags and researchers did their best to match up new tag and old tag. **(that column isn't in this version though)**
- `code2` (character): "initial.tag.yr" = marks the first year plant appeard in a survey.  Not necessarily a seedling.

# Analysis

### 2020-08-25 R/IPM practice.Rmd
built a practice IPM for one plot.

@@
 library(skimr)
+library(performance)
+library(lme4)
+library(bbmle)
+library(broom)
+library(broom.mixed)
+library(popbio)
 
@@
 conflict_prefer("filter", "dplyr")
+conflict_prefer("slice", "dplyr")
 ```
@@
 *Last compiled: `r Sys.Date()`*
+
+# TODO:
+
+- I feel like there is maybe a more sophisticated method of dealing with "zombie" plants.  Take another look at Elizabeth's lectures or whats-her-name's paper on terrestrial orchids.
+- Not sure how to incorporate overdispersion (significant random effect of `tag_number`)
 
@@
                     year = col_integer(),
-                    infl = col_integer()))
+                    infl = col_integer(),
+                    tag_number = col_character()))
 
@@
 - `shts` (numeric): number of shoots
-- `infl` (numeric): number of infloresences
+- `infl` (numeric): number of inflorescences
 - `code.notes` (character): 
@@
     - "tag missing" = ID tag labeling the plant was missing.  Plants were given new tags and researchers did their best to match up new tag and old tag. **(that column isn't in this version though)**
-- `code2` (character): "initial.tag.yr" = marks the first year plant appeard in a survey.  Not necessarily a seedling.
+- `code2` (character): "initial.tag.yr" = marks the first year plant appeared in a survey.  Not necessarily a seedling.
 
-# Analysis
+# Data Wrangling
+
+Take subset for messing around
+
+```{r}
+km41 <- demog %>% filter(bdffp_reserve_no == "1501") %>% select(-bdffp_reserve_no)
+km41 %>% count(plot)
+```
+
+Narrow down to just one plot for now
+
+```{r}
+km41 <- km41 %>% filter(plot == "5750")
+km41 %>% count(tag_number) %>% count(n)
+
+km41 %>% count(tag_number) %>% ggplot() + geom_histogram(aes(n))
+```
+Plants have up to 12 years of data
+
+## Survival and reproduction
+
+I need to have a column for fate.  Made difficult by "zombie" plants.
+
+```{r}
+km41
+```
+When is a plant really dead?  When it's the last non-`NA` value for that particular tag number before the max date.
+
+```{r}
+km41 %>% group_by(tag_number) %>% summarize(max(year))
+max(km41$year)
+```
+
+```{r}
+km41_2 <- 
+  km41 %>%
+  group_by(tag_number) %>% 
+  # Routine for figuring out last year of data collected
+  mutate(x = cumsum(ht > 0 & !is.na(ht))) %>%
+  # mutate(x = cumsum(lag(ht) > 0 & !is.na(lag(ht)))) %>%
+  mutate(n = 1:n()) %>% 
+  mutate(fate = if_else(
+    n == which.max(x==max(x)),
+    "died",
+    NA_character_
+  )) %>% 
+  # remove rows after plant is dead
+  filter(row_number() <= which(fate == "died")[1]) %>% 
+  # remove "died" if last year of data (2009)
+  ungroup() %>% 
+  mutate(fate = if_else(year == max(year), NA_character_, fate)) %>% 
+  select(-n, -x)
+# km41_2 %>% filter(fate == "died") %>% count(year)
+```
+```{r}
+km41_2 %>% count(fate)
+```
+359 plants died
+
+```{r}
+km41_2 <- km41_2 %>% mutate(surv = ifelse(is.na(fate), 1, 0))
+```
+
+```{r}
+km41_2 <- km41_2 %>% mutate(flowered = infl > 0)
+```
+
+## Just flowering individuals
+
+```{r}
+km41_flowered <- km41 %>% filter(infl > 0)
+```
+
+## Lagged height for growth
+
+```{r}
+km41_2 <-
+  km41_2 %>% 
+  group_by(tag_number) %>% 
+  mutate(ht_prev = lag(ht))
+```
+
+# IPM
+
+Gotta look this up, but assuming that reproduction happens before survival
+
+## Kernel
+
+In words:
+
+The number of individuals in a size class in year t+1 is a function of
+
+$$
+K(z^\prime,z) = p_b(z)b(z)p_rc_0(z^\prime) + s(z)G(z, z^\prime)
+$$
+Reporduction:
+
+- $p_b(z)$: probability of flowering as a function of size
+- $b(z)$: number of inflorescences as a function of size. 
+- $p_r$ mean # of seedlings produced per inflorescence.  This varies by year, not sure how to incorporate environmental stochasticity.
+- $c_0(z^\prime)$ size distribution of new recruits
+
+Growth & survival:
+
+- $s(z)$ is survival probability as a function of size
+- $G(z, z^\prime)$ is growth as a function of size.
+
+## Reproduction
+
+In this particular dataset the only information on reproduction is number of inflorescences per plant and total number of seedlings in a year.
+
+### Probability of flowering
+
+#### Model
+
+Start with most complex: quadratic response, environmental stochasticity, and random variation among plants.
+
+```{r}
+meo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+```
+Check random effects
+
+```{r}
+mo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|tag_number), family = binomial, data = km41_2)
+me_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|year), family = binomial, data = km41_2)
+m_pflwr2  <-   glm(flowered ~ ht + I(ht^2), family = binomial, data = km41_2)
+AICtab(meo_pflwr2, me_pflwr2, mo_pflwr2, m_pflwr2)
+```
+Keep both random effects.
+
+Check quadratic
+
+```{r}
+meo_pflwr <- glmer(flowered ~ ht + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+AICtab(meo_pflwr, meo_pflwr2)
+```
+
+Keep quadratic
+
+```{r}
+augment(meo_pflwr2, type.predict = "response") %>% 
+  # filter(tag_number %in% c(1, 101)) %>%
+  ggplot(aes(x = ht, y = .fitted)) +
+  geom_line(alpha = 0.2, aes(group = year)) +
+  stat_summary_bin(geom = "point", fun = mean, aes(y = as.numeric(flowered)), bins = 80, alpha = 0.5)
+```
+
+Some years had super high survival, others very low.  Variation among plants.
+
+#### Function
+
+```{r}
+fixef(meo_pflwr2)
+```
+
+$$
+p(flower) = inv.logit(-7.423 + 0.181ht - 0.0008ht^2)
+$$
+
+Not sure how to incorporate random effects
+
+```{r}
+p_bz <- function(z) {
+  coefs <- fixef(meo_pflwr2)
+  linear.p <- coefs[1] + coefs[2]*z + coefs[3]*z^2
+  p <- 1/(1+exp(-linear.p))
+  return(as.numeric(p))
+}
+# predict(mm_pflwr2, newdata = data.frame(ht = 20, tag_number = "new"), allow.new.levels = TRUE) %>% plogis() ==
+# p_bz(20)
+```
+
+### Number of infloresences
+
+#### Model
+
+```{r}
+meo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = poisson, data = km41_flowered)
+```
+```{r}
+me_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|year), family = poisson, data = km41_flowered)
+mo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|tag_number), family = poisson, data = km41_flowered)
+m_flwrs2 <- glm(infl ~ ht + I(ht^2), family = poisson, data = km41_flowered)
+AICtab(meo_flwrs2, me_flwrs2, mo_flwrs2, m_flwrs2)
+```
+
+No random effects.
+
+Check quadratic
+
+```{r}
+m_flwrs <- glm(infl ~ ht, family = poisson, data = km41_flowered)
+AICtab(m_flwrs, m_flwrs2)
+```
+
+Go with simpler linear glm
+
+```{r}
+augment(m_flwrs, type.predict = "response") %>% 
+  ggplot(aes(x = ht, y = .fitted)) +
+  geom_line() +
+  geom_jitter(aes(y = infl), alpha = 0.5, width = 0, height = 0.1)
+```
+
+#### Function
+
+```{r}
+b_z <- function(z) {
+  coefs <- coef(m_flwrs)
+  infl <- coefs[1] + coefs[2]*z
+  return(as.numeric(infl))
+}
+```
+
+### Probability of recruitment
+
+#### Model
+
+Estimated as number of recruits / number of inflorescences in year prev. year
+
+```{r}
+km41_rec_sum <- 
+  km41 %>%
+  group_by(year) %>% 
+  summarize(recruits = sum(code.notes == "sdlg (1)", na.rm = TRUE),
+            seeds_est = sum(lag(infl), na.rm = TRUE)) %>% 
+  filter(year > 1998)
+km41_rec_sum
+```
+
+```{r}
+m <- glm(recruits ~ 1, offset = log(seeds_est), family = poisson, data = km41_rec_sum)
+coef(m)
+```
+
+#### Function
+
+```{r}
+p_r <- as.numeric(exp(coef(m)))
+p_r
+```
+
+So like 1.64 seedlings per inflorescence
+
+### Size distribution of new recruits
+
+#### Model
+
+recruitment size
+
+```{r}
+km41_rec <- km41 %>% filter(code.notes == "sdlg (1)")
+m_rec <- lm(ht ~ 1, data = km41_rec)
+```
+
+```{r}
+coef(m_rec)
+summary(m_rec)$sigma
+```
+
+#### Function
+
+```{r}
+c_0z1 <- function(z1) {
+  mu <- coef(m_rec)[1]
+  sig <- summary(m_rec)$sigma
+  p.deRecr <- dnorm(z1, mean = mu, sd = sig)
+  return(p.deRecr)
+}
+# c_0z1(20)
+```
+
+## Growth
+
+### Model
+
+Try modeling growth as height_t+1 ~ height_t.  Needs a random effect of tag_number, I think?
+```{r}
+meo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|tag_number) + (1|year), data = km41_2)
+```
+Test random effects
+
+```{r}
+me_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|year), data = km41_2)
+mo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|tag_number), data = km41_2)
+m_grow2 <- lm(ht ~ ht_prev + I(ht_prev^2), data = km41_2)
+AICtab(meo_grow2, me_grow2, mo_grow2, m_grow2)
+```
+
+Environmental stochastsicity, but not differences among plants
+
+check quadratic
+
+```{r}
+me_grow <- lmer(ht ~ ht_prev + (1|year), data = km41_2)
+AICtab(me_grow, me_grow2)
+```
+
+Quadratic is better.
+
+```{r}
+augment(me_grow2) %>% 
+ggplot(aes(x = ht_prev, y = ht)) + 
+  geom_point(alpha = 0.2) +
+  geom_line(aes(y = .fitted, group = year), color = "blue", alpha = 0.2)
+```
+### Function
+
+```{r}
+fixef(me_grow2)
+```
+
+$$
+ht_{t+1} = 2.143 + 1.056ht - 0.002ht^2
+$$
+
+```{r}
+summary(me_grow2)$sigma
+summary(me_grow2)$varcor
+```
+
+Not sure how to take random effect into account
+
+```{r}
+G_z1z <- function(z1, z) {
+  coefs <- fixef(me_grow2)
+  sig <- summary(me_grow2)$sigma #or shoudl this incorporate mixed effects somehow?
+
+  mu <- coefs[1] + coefs[2]*z + coefs[3]*z^2
+  p.den.grow <- dnorm(z1, mean = mu, sd = sig) #pdf of new size for current size
+  return(p.den.grow)
+}
+```
+
+## Survival
+
+### Model
+```{r}
+meo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+```
+```{r}
+me_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|year), family = binomial, data = km41_2)
+mo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|tag_number), family = binomial, data = km41_2)
+m_surv2 <- glm(surv ~ ht + I(ht^2), family = binomial, data = km41_2)
+AICtab(meo_surv2, me_surv2, mo_surv2, m_surv2)
+```
+
+Only environmental stochasticity.
+
+Check quadratic
+
+```{r}
+me_surv <- glmer(surv ~ ht + (1|year), family = binomial, data = km41_2)
+
+AICtab(me_surv, me_surv2)
+```
+
+Quadratic is better
+
+```{r}
+augment(me_surv2, type.predict = "response") %>% 
+  ggplot(aes(x = ht, y = .fitted)) +
+  geom_line(aes(group = year), color = "blue", alpha = 0.5) +
+  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), bins = 80, alpha = 0.5)
+```
+
+Survival is better in some years than others, especially for very small plants
+
+### Function
+
+```{r}
+fixef(me_surv2)
+```
+
+$$
+p(survive) = inv.logit(0.920 + 0.142ht - 0.0009ht^2)
+$$
+
+```{r}
+s_z <- function(z) {
+  coefs <- fixef(me_surv2)
+  linear.p <- coefs[1] + coefs[2]*z + coefs[3]*z^2
+  p <- 1/(1+exp(-linear.p))
+  return(as.numeric(p))
+}
+```
+
+# Fecundity kernel
+
+p_b(z)b(z)p_rc_0(z^\prime)
+
+```{r}
+F_z1z <- function(z1, z) {
+  return(p_bz(z) * b_z(z) * p_r * c_0z1(z1))
+}
+```
+
+# Survival kernel
+
+s(z)G(z, z^\prime)
+
+```{r}
+P_z1z <- function(z1, z) {
+  return(s_z(z) * G_z1z(z, z1))
+}
+```
+
+# Run IPM
+
+## Decide meshpoints
+
+```{r}
+range(km41$ht[km41$ht>0], na.rm = TRUE)
+```
+
+So I'll go from 0.4 to 200??
+
+```{r}
+hist(km41$ht[km41$ht>0])
+```
+## Projection matrix
+
+```{r}
+mk_K <- function(m, L, U) {
+
+	# mesh points 
+	h <- (U - L)/m
+	meshpts <- L + ((1:m) - 1/2) * h
+	P <- h * (outer(meshpts, meshpts, P_z1z))
+	F <- h * (outer(meshpts, meshpts, F_z1z))
+	K <- P + F
+	return(list(K = K, meshpts = meshpts, P = P, F = F))
+}
+```
+
+```{r}
+out <- mk_K(100, 0.4, 200)
+```
+
+# Results
+
+```{r}
+lambda(out$K)
+plot(out$meshpts, stable.stage(out$K), type = "l")
+```
+
+```{r}
+hist(km41$ht)
+```
+Kinda matches?
+
+```{r}
+sens <- sensitivity(out$K)
+
+rownames(sens) <- out$meshpts
+colnames(sens) <- out$meshpts
+matplot2(sens)
+```
+
+Can't remember how to interpret this

### 2020-08-25 R/IPM practice.Rmd
added notes

@@
 In this particular dataset the only information on reproduction is number of inflorescences per plant and total number of seedlings in a year.
+However, Bruna et al.
+2004 has some additional data on reproduction:
+
+- each inflorescence has 20--25 flowers
+- proportion of flowers developing into fruits = 0.30±0.24 (range = 0--0.88)
+- Number of seeds per plant = 4.74±6.10 (range = 0--27)
+- Number of seeds per flower = 0.27±0.33 (range = 0--1)
+
+But the last step, seedlings per seed, isn't there, so I'm not sure if this info is relevant for this IPM, especially if none of the above vital rates is expected to differ by plant size.
 

### 2020-09-01 R/Workflow for yearly lambda.Rmd
Notes on how to programatically get lambdas for every transition year with a simple matrix model.

---
title: "R Notebook"
output: html_notebook
---

Use `popbio` to make a projection matrix

```{r}
library(tidyverse)
library(popbio)
```

One year at a time.  Good. That's what I want

```{r}
trans01 <- subset(merge(test.census, test.census, by = "plant", sort =FALSE),
                    year.x==2001 & year.y==2002 )
test.census %>% filter(year == 2002)
```

Add individual fertilities using "anonymous reproduction"  based on the proportional reproductive outputs of flowering plants and the total number of seedlings at the end of the projection interval

```{r}
# trans01$seedferts <- trans01$fruits.x / sum(trans01$fruits.x) * 5 #5 = number of seedlings in 2002.
trans01 <-
  trans01 %>% 
  mutate(seedferts = fruits.x / sum(fruits.x) * 5)

trans01
```
```{r}
stages <- c("seedling", "vegetative", "reproductive")
```

One way to specify columns:

```{r}
projection.matrix(trans01, stage.x, stage.y, seedferts, stages)
```

BEST to use column default (fertility column (seedling) now matches stage class name) AND order stages in dataframe

```{r}
# names(trans01)[c(3, 6, 8)] <- c("stage", "fate", "seedling")
# trans01$stage <- ordered(trans01$stage, stages)
trans01 <-
  rename(trans01, stage = stage.x, fate = stage.y, seedling = seedferts) %>% 
  mutate(stage = ordered(stage, c("seedling", "vegetative", "reproductive")))

trans01
```
Names in `stage` and `fate` have to match.  `stage` is ordinal. `seedling` column matches first stage to indicate fertility rates.

```{r}
projection.matrix(trans01)
```

# Try for every year

```{r}
census_wide <-
  test.census %>% 
  pivot_wider(names_from = year, values_from = c(stage, fruits))
```

```{r}
census_wide %>% 
  #need fertility columns for each year
  # mutate(fert_2001 = fruits_2001 / sum(fruits_2001, na.rm = TRUE) * sum(str_detect(stage_2002, "seedling"), na.rm = TRUE))
  mutate(across(starts_with("fruits"), ~ . / sum(., na.rm = TRUE), .names = "{col}.tot"))
```
```{r}
#complete data frame with all plant IDs.  Necessary for lags to work right.
full.census <-
  left_join(
    test.census %>% 
      expand(plant = 1:18, year) %>% 
      arrange(year),
    test.census
  )

full.census <-
  full.census %>%
  # generate transitions
  group_by(plant) %>% 
  mutate(transition = paste0(year, "-", lead(year))) %>% 
  mutate(fate = lead(stage)) %>% 
  # calculate anonymous reproduction
  group_by(year) %>% 
  mutate(num_seedling = sum(str_detect(fate, "seedling"), na.rm = TRUE)) %>% 
  mutate(seedling = fruits / sum(fruits, na.rm = TRUE) * num_seedling) %>% 
  # remove plants that started off a transition dead
  filter(stage != "dead", !is.na(fate)) %>% 
  mutate(stage = ordered(stage, c("seedling", "vegetative", "reproductive")))
```

```{r}
testyr <- full.census %>% filter(year == 2001)
projection.matrix(as.data.frame(testyr), stage = stage, fate = fate, fertility = seedling)
projection.matrix(trans01, stage = stage, fate = fate, fertility = seedling)
projection.matrix(full.census %>% as.data.frame())
```

Get lambda for every transition

```{r}
full.census %>% 
  as.data.frame() %>%  
  group_by(transition) %>% 
  nest() %>% 
  mutate(K = map(data,
                 ~projection.matrix(as.data.frame(.x))
  )) %>% #projection.matrix() doesn't work with tibbles for some reason
  mutate(lambda = map_dbl(K, ~ lambda(.x)))
```

### 2020-09-03 notes/Climate vital rates modeling.Rmd
moved notebooks to notes directory

---
title: "Drought effects on vital rates"
author: "Eric R. Scott"
date: "Date Initiated"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(conflicted)
library(lubridate)
library(equatiomatic)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Goal

Here's what needs to happen.  One script needs to calculate demographic rates for each transition (year) and plot combination and use those to get a lambda for each transition.  Rows would be year:plot combinations.

A second script needs to put SPEI values in a wide format with rows = years and cols = months starting at whatever month the census was done in (I think February).  So the colnames would be like:

feb, jan, dec, nov, oct, sep, aug, jul, jun, may, apr, mar

A third script would combine these datasets and implement the DLNLM (gasparini) / FLM (teller) GAM approach.

# Purpose

Begin to play around with modeling effects of drought on vital rates

# Load Data

```{r data, echo=TRUE}
precip <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
demog <- read_csv(here("analysis", "data", "raw_data", "Ha_survey_with_Zombies.csv"), col_names = TRUE,
               cols(plot = col_character(),
                    bdffp_reserve_no = col_character(),
                    shts = col_integer(),
                    year = col_integer(),
                    infl = col_integer(),
                    tag_number = col_character(),
                    HA_ID_Number = col_character())) %>% 
  clean_names()
min(demog$year)
max(demog$year)
```

# Wrangling

Need to match up demography with climate in some way.  As a preliminary pass, I'll calculate average SPI/SPEI and total precip for the growing season (MONTHS?) for each year and match the rain gauge sites to the demographic plots.

```{r}
unique(demog$ranch)
unique(demog$bdffp_reserve_no)
demog %>% filter(bdffp_reserve_no == "1202")
unique(precip$site)
```

km41 = 1501
km37 = ??
florestal = 1301
cabo frio = 3402
Porto Alegre = 3114
Colosso = 1104, 1202
Dimona = 2206, 2108, 2107

colosso_clust = 1301, 3402, 1104, 1202

```{r}
demog <- 
  demog %>% 
  mutate(site = case_when(
    bdffp_reserve_no == "1501" ~ "km_clust",
    bdffp_reserve_no == "3114" ~ "porto_alegre",
    bdffp_reserve_no %in% c("1301", "3402", "1104", "1202") ~ "colosso_clust",
    bdffp_reserve_no %in% c("2206", "2108", "2107") ~ "dimona"
  ))
```

Aggregate precip data by growing season so just one value per year.

Flowering season =  jan - april
Fruits ripening = may - june
Dry season = june - dec

```{r}
ggplot(precip, aes(x = month(date, label = TRUE), y = precip_tot)) + 
  geom_violin(draw_quantiles = .5)
```
```{r}
flwr <-
  precip %>% 
  mutate(month = month(date),
         year = year(date)) %>% 
  filter(month %in% c(1:4)) %>% 
  group_by(year, site) %>% 
  summarize(spi_flwr = mean(spi),
            spei_flwr = mean(spei),
            precip_flwr = sum(precip_tot))

fruit <-
  precip %>% 
  mutate(month = month(date),
         year = year(date)) %>% 
  filter(month %in% c(5:6)) %>% 
  group_by(year, site) %>% 
  summarize(spi_fruit = mean(spi),
            spei_fruit = mean(spei),
            precip_fruit = sum(precip_tot))

dry <-
  precip %>% 
  mutate(month = month(date),
         year = year(date)) %>% 
  filter(month %in% c(7:12)) %>% 
  group_by(year, site) %>% 
  summarize(spi_dry = mean(spi, na.rm = TRUE),
            spei_dry = mean(spei, na.rm = TRUE),
            precip_dry = sum(precip_tot))
wet <-
  precip %>% 
  mutate(month = month(date),
         year = year(date)) %>% 
  filter(month %in% c(1:5)) %>% 
  group_by(year, site) %>% 
  summarize(spi_wet = mean(spi, na.rm = TRUE),
            spei_wet = mean(spei, na.rm = TRUE),
            precip_wet = sum(precip_tot))

precip <- full_join(flwr, fruit,by = c("year", "site")) %>% full_join(dry, by = c("year", "site")) %>% full_join(wet, by = c("year", "site"))
```
join to demography data

```{r}
full <- left_join(demog, precip, by = c("year", "site"))
```

# Analysis

Ok, let's start with something simple like growth.  Census happen in late Jan or Feb, so change in height would be predicted by previous year's rainy season.

```{r}
full <- full %>% 
  group_by(ha_id_number) %>% 
  arrange(ha_id_number, year) %>% 
  mutate(ht_prev = lag(ht))
```

```{r}
m <- lm(ht ~ ht_prev*spi_wet, data = full)
summary(m)
anova(m)
extract_eq(m, use_coefs = TRUE)
```
$$
\operatorname{ht} = 5.77 + 0.87(\operatorname{ht\_prev}) - 0.14(\operatorname{spi\_wet}) + 0.01(\operatorname{ht\_prev} \times \operatorname{spi\_wet}) + \epsilon
$$

- higher SPI in the wet season (wetter) = less growth.
- positive interaction term means larger plants (more/less?) impacted by spi in the wet season

### 2020-09-03 notes/IPM practice.Rmd
moved notebooks to notes directory

--- 2020-08-25 R/IPM practice.Rmd
@@
 library(popbio)
+library(janitor)
 
@@
                     infl = col_integer(),
-                    tag_number = col_character()))
+                    tag_number = col_character(),
+                    HA_ID_Number = col_character())) %>% 
+  clean_names()
 
 unique(demog$ranch)
-unique(demog$code.notes)
+unique(demog$code_notes)
 unique(demog$code2)
-unique(demog$size)
+unique(demog$habitat)
 
@@
 - `plot` (character):
-- `size` (character): fragment size
-- `ranch` (character): "Dimona", "Esteio-Colosso", or "PortoAlegre".  The three ranches that make up BDFFP.
+- `habitat` (character): fragment size
+- `ranch` (character): "DIM" = Dimona, "EST" = Esteio-Colosso, or "PAL" = PortoAlegre.  The three ranches that make up BDFFP.
 - `bdffp_reserve_no` (character): Official BDFFP reserve designation
+- `ha_id_number` (numeric): A unique ID number for each plant after resolving duplicate tags and such.
 - `tag_number` (numeric): plant tag# within plot (not unique)
@@
     - "sdlg" = seedling, defined as a year 1 plant.
-    - "tag missing" = ID tag labeling the plant was missing.  Plants were given new tags and researchers did their best to match up new tag and old tag. **(that column isn't in this version though)**
+    - "tag missing" = ID tag labeling the plant was missing.  Plants were given new tags and researchers did their best to match up new tag and old tag.
 - `code2` (character): "initial.tag.yr" = marks the first year plant appeared in a survey.  Not necessarily a seedling.
@@
 ```{r}
-km41 %>% group_by(tag_number) %>% summarize(max(year))
+km41 %>% group_by(ha_id_number) %>% summarize(max(year))
 max(km41$year)
@@
   km41 %>%
-  group_by(tag_number) %>% 
+  group_by(ha_id_number) %>% 
   # Routine for figuring out last year of data collected
@@
 ```
-359 plants died
+362 plants died
 
@@
   km41_2 %>% 
-  group_by(tag_number) %>% 
+  group_by(ha_id_number) %>% 
   mutate(ht_prev = lag(ht))
@@
 - Number of seeds per plant = 4.74±6.10 (range = 0--27)
+- Number of sees per fruit =  1, 2, or 3 with equal probability.
 - Number of seeds per flower = 0.27±0.33 (range = 0--1)
@@
 ```{r}
-meo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+meo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|ha_id_number) + (1|year), family = binomial, data = km41_2)
 ```
@@
 ```{r}
-mo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|tag_number), family = binomial, data = km41_2)
+mo_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|ha_id_number), family = binomial, data = km41_2)
 me_pflwr2 <- glmer(flowered ~ ht + I(ht^2) + (1|year), family = binomial, data = km41_2)
@@
 ```{r}
-meo_pflwr <- glmer(flowered ~ ht + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+meo_pflwr <- glmer(flowered ~ ht + (1|ha_id_number) + (1|year), family = binomial, data = km41_2)
 AICtab(meo_pflwr, meo_pflwr2)
@@
 ```{r}
-meo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = poisson, data = km41_flowered)
+meo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|ha_id_number) + (1|year), family = poisson, data = km41_flowered)
 ```
@@
 me_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|year), family = poisson, data = km41_flowered)
-mo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|tag_number), family = poisson, data = km41_flowered)
+mo_flwrs2 <- glmer(infl ~ ht + I(ht^2) + (1|ha_id_number), family = poisson, data = km41_flowered)
 m_flwrs2 <- glm(infl ~ ht + I(ht^2), family = poisson, data = km41_flowered)
@@
   group_by(year) %>% 
-  summarize(recruits = sum(code.notes == "sdlg (1)", na.rm = TRUE),
+  summarize(recruits = sum(code_notes == "sdlg (1)", na.rm = TRUE),
             seeds_est = sum(lag(infl), na.rm = TRUE)) %>% 
@@
 ```{r}
-km41_rec <- km41 %>% filter(code.notes == "sdlg (1)")
+km41_rec <- km41 %>% filter(code_notes == "sdlg (1)")
 m_rec <- lm(ht ~ 1, data = km41_rec)
@@
 ```{r}
-meo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|tag_number) + (1|year), data = km41_2)
+meo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|ha_id_number) + (1|year), data = km41_2)
 ```
@@
 me_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|year), data = km41_2)
-mo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|tag_number), data = km41_2)
+mo_grow2 <- lmer(ht ~ ht_prev + I(ht_prev^2) + (1|ha_id_number), data = km41_2)
 m_grow2 <- lm(ht ~ ht_prev + I(ht_prev^2), data = km41_2)
@@
 ```{r}
-meo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|tag_number) + (1|year), family = binomial, data = km41_2)
+meo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|ha_id_number) + (1|year), family = binomial, data = km41_2)
 ```
@@
 me_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|year), family = binomial, data = km41_2)
-mo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|tag_number), family = binomial, data = km41_2)
+mo_surv2 <- glmer(surv ~ ht + I(ht^2) + (1|ha_id_number), family = binomial, data = km41_2)
 m_surv2 <- glm(surv ~ ht + I(ht^2), family = binomial, data = km41_2)
@@
 mk_K <- function(m, L, U) {
-
 	# mesh points 
@@
 colnames(sens) <- out$meshpts
-matplot2(sens)
+matplot2(sens, xlab = "Height")
 ```
@@
 Can't remember how to interpret this
+
+```{r}
+elas <- elasticity(out$K)
+rownames(elas) <- colnames(elas) <- out$meshpts
+matplot2(elas, xlab = "Height")
+```

### 2020-09-03 notes/IPMpack test.Rmd
moved notebooks to notes directory

---
title: "R Notebook"
output: html_notebook
---

```{r}
library(IPMpack)
```

Needs particular data.frame format with columns:

- `size`
- `sizeNext`
-  `surv` (1 or 0)
- `fec1`, `fec2`, ...: e.g.
  - `fec1`: probability of reproduction
  - `fec2`: number of reproductive structures (flowers)
  - `fec3`: number of propagules (seeds per flower)
- `stage` Use either this or size.  If size is not `NA`, then this must be `"continuous"`. If this column is missing, it assumes "continuous" for all entries.  Can mix continuous and discrete with this column.
- `stageNext`
- `number` number of individuals corresponding to each row
- `covariate` discrete covariate
- `covariateNext` 
- any other covariates named whatever you want (e.g. SPEI)
- `offspringNext` either "sexual" or "clonal"

```{r}
dff <- generateData()
head(dff)
```

```{r}
plot(dff$size, dff$sizeNext, xlab = "Size at t", ylab = "Size at t+1")
```

Make objects from data frame
uhhnhh what is "size2"?  It's not in the data frame!  Ughhh. I hate this shit.

```{r}
# gr1 <- makeGrowthObj(dataf = dff, Formula = sizeNext~size+size2)

gr1 <- makeGrowthObj(dataf = dff, Formula = sizeNext ~ size + I(size^2))
sv1 <- makeSurvObj(dff, surv ~ size + I(size^2))
```

```{r}
picGrow(dff, gr1)
picSurv(dff, sv1, ncuts = 20)
```

Make $P$ matrix (survival and growth)

```{r}
Pmatrix <- makeIPMPmatrix(nBigMatrix = 50,
                          minSize = -5, maxSize = 35,
                          growObj = gr1, survObj = sv1,
                          correction = "constant")
Pmatrix
# plot(Pmatrix) #nope
image(Pmatrix, xlab = "size at t", ylab = "size at t+1")
persp(Pmatrix, xlab = "size at t", ylab = "size at t+1", zlab = "transition probability")
```

How life expectency relates to size:

```{r}
LE <- meanLifeExpect(Pmatrix)
plot(Pmatrix@meshpoints, LE, type = "l", xlab = "size at t", ylab = "life expectency")
```

Fertilty
```{r}
fv1 <- makeFecObj(dff, Formula = fec ~ size, Family = "gaussian", Transform = "log")
```

### 2020-09-03 notes/Fumana commented IPM code.R

@@ -1,0 +1,351 @@
+##
+### Commented R code for IPM analyses done in Dahlgren et al., submitted to Ecology May 2015
+##
+#
+# Analyses were run in R v. 3.0.0, and using packages gamm4 v. 0.1-6, lme4 v. 0.999999-2 and mgcv v. 1.7-22.
+
+###
+## load gamm package 
+library(gamm4)
+
+###vital rate gamms:
+##k=3 puts a cap on the maximum wiggliness of the (default) "thin-plate splines" (lowest possible number of 'knots')
+##sum.precX are summer precipitation in years t and t+1 (for simplicitly/readability, the temperature variables are not included here)
+##prop.dens.t is plot density per year (as proportional to maximum observed density)
+##random factors are included to account for that climatic variables vary only among years, and density across plots
+##individual id was not included as a grouping variable to reduce model complexity
+##all variables are contained in a data frame object named 'dat'
+
+frutmod <-
+  gamm4(
+    frut1 ~ s(log.sizet, k = 3) + s(sum.prec.t1, k = 3) + s(prop.dens.t, k = 3),
+    random = ~ (1 | yrt) + (1 | Plot),
+    family = poisson,
+    data = dat
+  )
+
+survmod <-
+  gamm4(
+    survt.t1 ~ s(log.sizet, k = 3) + s(sum.prec.t, k = 3) + s(sum.prec.t1, k = 3) + s(prop.dens.t, k = 3),
+    random = ~ (1 | yrt) + (1 | Plot),
+    family = binomial,
+    data = dat
+  )
+
+sizemod <-
+  gamm4(
+    log.sizet1 ~ s(log.sizet, k = 3) + s(sum.prec.t, k = 3) + s(sum.prec.t1, k = 3) + s(prop.dens.t, k = 3),
+    random = ~ (1 | yrt) + (1 | Plot),
+    family = gaussian,
+    data = dat
+  )
+
+flowmod <-
+  gamm4(
+    flowt1 ~ s(log.sizet, k = 3) + s(sum.prec.t, k = 3) + s (sum.prec.t1, k = 3) + s(prop.dens.t, k = 3),
+    random = ~ (1 | yrt) + (1 | Plot),
+    family = binomial,
+    data = dat
+  )
+
+
+
+## IPM ----
+
+########Vital rate functions-------------------------------------------------------
+## the generic predict() function is used to get vital rate components for the kernel
+
[…]

### 2020-09-10 R/competition wrangling.Rmd
started working on wrangling data for modeling density dependence.

---
title: "Competition"
author: "Eric R. Scott"
date: "2020-09-09"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(glue)
library(car)
source(here("R", "utils.R"))

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Competition kernel. Teller et al.

I want to use the `row` and `colunn` to figure out how many neighbors each plant has. I should exclude plants on edges, because I don't know about their neighbors outside of the plot.

# Load Data

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
```

```{r}
ha
```
Make numeric row and column numbers for easier data handling later.

```{r}
# as.factor(ha$row) #levels are alphabetical order by default, which is perfect.  Except for row "L", not sure what that's about.
ha <- 
  ha %>% 
  mutate(row_number = as.integer(as.factor(row)), .after = row,
         column = as.integer(column)) %>% 
  filter(row != "L") # just get rid of the few plants in a row "L".--only affects one plot
```

Start with one plot

```{r}
ha_test <- ha %>% filter(plot == first(plot))
```

To calculate neigbors at a certain distance I first calculate the number of plants in each grid cell, as a matrix using the `xtabs()` function.  I do this for each year separately, and eventually will apply this to every year x plot combination.

```{r}
ha_g <- 
  ha_test %>% 
  add_column(plant = 1) %>% 
  group_by(year)

neighbors <-
  ha_g %>% 
  group_split() %>% 
  map(~xtabs(plant ~ row + column, .x)) %>% 
  set_names(group_keys(ha_g)$year)
head(neighbors, 3)
```

Then, for every grid cell, I create a masking matrix using a custom function, `mat_rect()`, in utils.R.

```{r}
I1 <- mat_rect(D = 0, nrow = 5, ncol = 10)
```

Then for each year in `neighbors`, I multiply the xtabs matrix by the masking matrix and take the sum to get the number of plants in that distance around each focal plant.

```{r}
n_plants_1 <-
  map_df(neighbors,
  function(x) map_dbl(I1, ~ sum(x * .x)) %>% 
    enframe(value = "n_plants_1") %>% 
    separate(name, into = c("row", "column"), sep = ",", convert = TRUE),
  .id = "year"
) %>% 
  mutate(year = as.integer(year)) #for joining later
n_plants_1
```
Then, I can join this with the original data frame

```{r}
ha_test <- left_join(ha_test, n_plants_1, by = c("row_number" = "row", "column", "year"))
```

# Model effects of density
## growth

```{r}
m <- glm(ht ~ ht_prev * n_plants_1, family = gaussian(link = "identity"), data = ha_test)
summary(m)
car::Anova(m)
```
```{r}
ha %>% count(plot, year, row, column)
```

### 2020-09-10 R/utils.R

@@ -12,3 +12,115 @@
 `%^%` <- function(x, k) {
   matrixcalc::matrix.power(x, k)
 }
+
+
+#' Create masking matrices with rectangles
+#' 
+#' Creates a list of matrices with all 0's except for a rectangle of 1's around
+#' a focal "cell". Used for figuring out how many neighboring plants each plant
+#' has.
+#'
+#' @param D distance from focal cell for rectangle
+#' @param nrow number of rows of matrix
+#' @param ncol number of columns of matrix
+#' @author Lyndsay Troyer
+#' @return a list of matrices
+#' @export
+#' 
+#' @import glue
+#' 
+#' @examples
+#' I <- mat_rect(D = 1, nrow = 5, ncol = 5)
+#' y <- matrix(round(runif(25, 1, 10), 0), nrow = 5, ncol = 5)
+#' map_dbl(I, ~ sum(y * .x)) %>%
+#'   enframe(value = "n_plants_2") %>%
+#'   separate(name, into = c("row", "column"), sep = ",", convert = TRUE)
+mat_rect <- function(D, nrow, ncol) {
+  mlist <- vector("list", 1)
+  for(r in 1:nrow) {
+    for(c in 1:ncol) {
+      m <- matrix(0, ncol = ncol, nrow = nrow)
+      r1 <- r + D
+      r1 <- r1[r1 <= nrow]
+      c1 <- c + (-D:D)
+      c1 <- c1[c1 > 0 & c1 <= ncol]
+      m[r1, c1] <- 1
+      
+      r2 <- r - D
+      r2 <- r2[r2 > 0]
+      c2 <- c - (-D:D)
+      c2 <- c2[c2 > 0 & c2 <= ncol]
+      m[r2, c2] <- 1
+      
+      r3 <- r - (-D:D)
+      r3 <- r3[r3 > 0 & r3 <= nrow]
+      c3 <- c - D
+      c3 <- c3[c3 > 0]
+      m[r3, c3] <- 1
+      
+      r4 <- r + (-D:D)
+      r4 <- r4[r4 > 0 & r4 <= nrow]
+      c4 <- c + D
+      c4 <- c4[c4 <= ncol]
+      m[r4, c4] <- 1
+      
+      m <- list(m)
+      names(m) <- glue::glue("{r}, {c}")
+      mlist <- c(mlist, m)
+    }
+  }
+  return(mlist[-1])
+}
+
+
[…]

### 2020-09-10 R/competition wrangling.Rmd
functioning, but untested function to calculate density of neighboring plants.

@@
 library(car)
+library(skimr)
+
 source(here("R", "utils.R"))
@@
 
-```{r}
-ha
-```
 Make numeric row and column numbers for easier data handling later.
@@
 
-Start with one plot
+```{r}
+skim(ha)
+```
+
+# Calculating density at different distances
+
+## Start with one plot
 
@@
 
-To calculate neigbors at a certain distance I first calculate the number of plants in each grid cell, as a matrix using the `xtabs()` function.
+To calculate neighbors at a certain distance I first calculate the number of plants in each grid cell, as a matrix using the `xtabs()` function.
 I do this for each year separately, and eventually will apply this to every year x plot combination.
@@
 
-# Model effects of density
-## growth
+## Scaling up
+
+One difficulty is that the size of plots varies:
 
 ```{r}
-m <- glm(ht ~ ht_prev * n_plants_1, family = gaussian(link = "identity"), data = ha_test)
-summary(m)
-car::Anova(m)
+ha %>% 
+  group_by(plot) %>% 
+  summarize(nrows = max(row_number), ncols = max(column))
 ```
+This means there has to be a different list of masking matrices for each plot, in addition to a different density matrix for each plot:year.
+
+Also, occasionally the density matrix `neighbors` isn't consistent from year to year.  E.g. a column has no plants in it in one year and has a plant in it the next year.
+
+So, It might be easiest to split by plot, do this stuff on each plot, then re-join?
+
 ```{r}
-ha %>% count(plot, year, row, column)
+ha_plots <- ha %>% group_by(plot) %>% group_split()
 ```
+
+```{r}
+#function to calculate neighbor density for a single plot (every year).
+df_plot <- ha_plots[[4]]
+calc_dens_neighbor <-
+  function(df_plot, D) {
+  ha_g <- 
+    df_plot %>% 
+    add_column(plant = 1) %>% 
+    group_by(year)
+
+  neighbors <-
+    ha_g %>% 
+    group_split() %>% 
+    map(~xtabs(plant ~ row + column, .x)) %>% 
+    set_names(group_keys(ha_g)$year)
+
+  I <- mat_rect(
+    D = D,
+    nrow = max(df_plot$row_number),
+    ncol = max(df_plot$column)
+  )
+
+  n_plants <-
+    map_df(neighbors,
+           function(x) map_dbl(I, ~ sum(x * .x)) %>%
+             enframe(value = glue::glue("n_plants_{D}")) %>%
+             separate(name, into = c("row", "column"), sep = ",", convert = TRUE),
+           .id = "year"
+    ) %>%
+    mutate(year = as.integer(year))
+
+  out <- left_join(df_plot, n_plants, by = c("row_number" = "row", "column", "year"))
+  return(out)
+}
+```
+
+For one plot:
+
+```{r}
+calc_dens_neighbor(ha_plots[[1]], D = 3)
+```
+
+But, doesn't work when a plot has empty rows or columns in some years but not others.
+
+```{r}
+calc_dens_neighbor(ha_plots[[4]], D = 3)
+```
+
+Alternative with list columns
+
+```{r}
+calc_dens_neighbor <-
+  function(df_plot, D) {
+
+    calc_dens <- function(neighbors, I, D) {
+      out <- map_dbl(I, ~ sum(neighbors * .x)) %>%
+        enframe(value = glue::glue("n_plants_{D}")) %>%
+        separate(name, into = c("row_number", "column"), sep = ",", convert = TRUE)
+      return(out)
+    }
+
+    df_nested <-
+      df_plot %>% 
+      add_column(plant = 1) %>% 
+      group_nest(year) %>% 
+      mutate(neighbors = map(data, ~xtabs(plant ~ row + column, .x))) %>% 
+      mutate(I = map(neighbors, ~{
+        mat_rect(
+          D = D,
+          nrow = nrow(.x),
+          ncol = ncol(.x)
+        )
+      }))
+
+    out_nested <- 
+      df_nested %>% 
+      mutate(n_plants = map2(neighbors, I, ~calc_dens(.x, .y, D)))
+
+    out_joined <- out_nested %>%
+      mutate(d = map2(data, n_plants,
+                      ~left_join(.x, .y, by = c("row_number", "column"))))
+
+    out <- out_joined$d %>% bind_rows()
+    return(out)
+  }
+```
+
+```{r}
+ha_plots %>% map_df(~calc_dens_neighbor(.x, 2))
+```

### 2020-09-14 R/competition wrangling.Rmd
finish wrangling density data

--- 2020-09-10 R/competition wrangling.Rmd
@@
 
-Start with one plot
+```{r}
+skim(ha)
+```
+
+# Calculating density at different distances
+First, a small example as proof of concept:
+```{r}
+ha_test <-
+  ha %>%
+  filter(plot == first(plot)) %>%
+  filter(year == first(year)) %>%
+  filter(row_number <=3, column <=3) %>% 
+  add_column(plant = 1) #just a dummy indicator column for use later
+ha_test
+```
+
+I'll make use of `expand_grid()` to create every combination of row and column (cell), and `dist()` to calculate the distance between every cell and every other cell.
 
 ```{r}
-ha_test <- ha %>% filter(plot == first(plot))
+#all possible combination of row and column ID
+grid_coords <- expand_grid(row = 1:3, col = 1:3) %>% as.matrix()
+
+#calculate distrance from every coord to every other coord
+d <- dist(grid_coords*10, #because each grid cell is 10m x 10m
+          diag = FALSE,
+          upper = TRUE) %>% 
+  as.matrix()
+
+rownames(d) <- colnames(d) <- paste(grid_coords[ , 1], grid_coords[ , 2], sep = ",")
+d
 ```
 
-To calculate neigbors at a certain distance I first calculate the number of plants in each grid cell, as a matrix using the `xtabs()` function.
-I do this for each year separately, and eventually will apply this to every year x plot combination.
+Now I can turn this into a tidy/long format. This shows the distance from (`row_from`, `column_from`) to (`row_to`, `column_to`).
 
 ```{r}
-ha_g <- 
-  ha_test %>% 
-  add_column(plant = 1) %>% 
-  group_by(year)
-
-neighbors <-
-  ha_g %>% 
-  group_split() %>% 
-  map(~xtabs(plant ~ row + column, .x)) %>% 
-  set_names(group_keys(ha_g)$year)
-head(neighbors, 3)
+dist_df <-
+  as.data.frame(d) %>% 
+  rownames_to_column(var = "from") %>% 
+  pivot_longer(-from, names_to = "to", values_to = "dist") %>% 
+  separate(from, into = c("row_from", "column_from")) %>% 
+  separate(to, into = c("row_to", "column_to")) %>% 
+  mutate(across(where(is.character), as.integer))
+head(dist_df)
 ```
 
-Then, for every grid cell, I create a masking matrix using a custom function, `mat_rect()`, in utils.R.
+Then we can summarize the real data to figure out how many plants are in each grid cell.  Here, I'm just using a small subset that corresponds to a single 3x3 plot
 
 ```{r}
-I1 <- mat_rect(D = 0, nrow = 5, ncol = 10)
+plant_summary <-
+  ha_test %>% 
+  group_by(row_number, column) %>% 
+  summarize(n_plants = n()) %>% 
+  rename(row = row_number)
+plant_summary
+```
+Now I can join these two data frames.
+
+```{r}
+dist_full <- full_join(dist_df, plant_summary, by = c("row_to" = "row", "column_to" = "column"))
+head(dist_full)
+```
+The way to read this is for a plant in [1,1] the cell [1,1] is 0m away and has 3 plants in it.
+For a plant in [1,1], cell [1,2] is 10m away and there is 1 plant in it.
+For a plant in [1,1], cell [2,3] is 22.36m away and there are 9 plants in it, etc.
+
+Here's another way to visualize this using `xtabs()`:
+
+```{r}
+xtabs(plant ~ row_number + column, ha_test)
 ```
 
-Then for each year in `neighbors`, I multiply the xtabs matrix by the masking matrix and take the sum to get the number of plants in that distance around each focal plant.
+So now I can filter this and summarize to calculate neighbor density at arbitrary distances for every grid cell.
 
 ```{r}
-n_plants_1 <-
-  map_df(neighbors,
-  function(x) map_dbl(I1, ~ sum(x * .x)) %>% 
-    enframe(value = "n_plants_1") %>% 
-    separate(name, into = c("row", "column"), sep = ",", convert = TRUE),
-  .id = "year"
-) %>% 
-  mutate(year = as.integer(year)) #for joining later
-n_plants_1
+#density between 5m and 15m away
+dist_5_15<-
+  dist_full %>% 
+  filter(dist >=5, dist <15) %>% 
+  group_by(row_from, column_from) %>% 
+  summarize(n_neighbors_5_15 = sum(n_plants, na.rm = TRUE),
+            n_cells = n(),
+            density_5_15 = n_neighbors_5_15/n_cells)
 ```
-Then, I can join this with the original data frame
+So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2
+
+Then I can join that back to the original data set
 
 ```{r}
-ha_test <- left_join(ha_test, n_plants_1, by = c("row_number" = "row", "column", "year"))
+full_test <- full_join(ha_test, dist_5_15, by = c("row_number" = "row_from", "column" = "column_from"))
+head(full_test)
 ```
 
-# Model effects of density
-## growth
+## Scaling up
+
+I'm going to scale up using a nested dataframe and the magic of list-columns!
+The advantage of this over splitting the one dataframe into a dataframe for every plot:year combo and using a for-loop or `map()` is that it keeps the plot and year structure intact and I don't have to figure out how to re-join the data at the end.
+
+First, nest the dataframe by plot and year.
 
 ```{r}
-m <- glm(ht ~ ht_prev * n_plants_1, family = gaussian(link = "identity"), data = ha_test)
-summary(m)
-car::Anova(m)
+ha_nest <-
+  ha %>% 
+  group_by(plot, year) %>% 
+  group_nest()
 ```
+
+Then, do the setup by calculating the `grid_coords` and `d` for every plot:year combo.
+
 ```{r}
-ha %>% count(plot, year, row, column)
+ha_nest_dist <- 
+  ha_nest %>% 
+  #every combination of row and column:
+  mutate(grid_coords = map(
+    data, 
+    ~expand_grid(row = min(.x$row_number):max(.x$row_number),
+                 col = min(.x$column):max(.x$column)) %>% as.matrix()
+    )) %>% 
+  #matrix describing euclidean distance from every point to every other point:
+  mutate(d = map(
+    grid_coords,
+    ~{
+      d <- dist(.x * 10, #because each grid cell is 10m x 10m
+                diag = FALSE,
+                upper = TRUE) %>% 
+        as.matrix()
+       colnames(d) <- rownames(d) <- paste(.x[,1], .x[,2], sep = ",")
+      d
+    }
+    )) %>% 
+  #tidy data frame version of d
+  mutate(dist_df = map(d, ~{
+    as.data.frame(.x) %>% 
+      rownames_to_column(var = "from") %>% 
+      pivot_longer(-from, names_to = "to", values_to = "dist") %>% 
+      separate(from, into = c("row_from", "column_from")) %>% 
+      separate(to, into = c("row_to", "column_to")) %>% 
+      mutate(across(where(is.character), as.integer))
+  }
+  ))
 ```
+
+Now, create that summary tibble counting the number of plants in each grid cell for every plot:year combo, join, and unnest.
+
+```{r}
+ha_dist <-
+  ha_nest_dist %>% 
+  #summarize each year:plot to count number of plants in each grid cell
+  mutate(n_plants_sum = map(data, ~{
+    .x %>% 
+      add_column(plant = 1) %>% 
+      group_by(row_number, column) %>% 
+      summarize(n_plants = n()) %>% 
+      rename(row = row_number)
+  }
+  )) %>% 
+  #join with tidied distance matrix to get dataframe for distance from every row
+  #and column to every other row and column
+  mutate(dist_full = map2(dist_df, n_plants_sum, ~{
+    full_join(.x, .y, by = c("row_to" = "row", "column_to" = "column"))
+  } 
+  )) %>% 
+  select(plot, year, dist_full) %>% 
+  unnest(dist_full)
+
+ha_dist
+```
+
+Now, with `ha_dist`, I can filter by distance, calculate a density, and the join to the original data.
+
+```{r}
+dens_0_10 <-
+  ha_dist %>% 
+  filter(dist>0 & dist <=10) %>% 
+  group_by(plot, year, row_from, column_from) %>% 
+  summarize(density_0_10 = sum(n_plants, na.rm = TRUE)/n())
+left_join(ha, dens_0_10, by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
+  arrange(plot, year, row_number, column)
+```
+
+I'll do this for different distances programmatically.
+
+```{r}
+width <- 10 #meters
+max <- 100 #meters
+L <- seq(0, max-width, by = width)
+U <- seq(0 + width, max, by = width)
+
+out <- 
+  map2(L, U, ~{
+  ha_dist %>%
+    filter(dist >= .x & dist < .y) %>% 
+    group_by(plot, year, row_from, column_from) %>%
+    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_plants, na.rm = TRUE)/n())
+})
+
+x <- out[[1]]
+for (i in 2:length(out)) {
+  x <- full_join(x, out[[i]], by = c("plot", "year",  "row_from", "column_from"))
+}  
+x
+```
+
+Join to original data
+
+```{r}
+ha_new <- 
+  left_join(ha, x,
+          by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
+  arrange(plot, year, row_number, column)
+ha_new
+```
+
+```{r}
+library(naniar)
+vis_miss(ha_new %>% select(starts_with("density")))
+```
+The furthest distance I can estimate density for all plants is 60m.
+Past that, in order to estimate effects of density, I would have to drop plants.
+Even for closer plants though, the estimates may be biased because they do not include plants outside of the plot.
+That is, neighbor density for plants near the edges is based on a smaller sample than for plants near the center.

### 2020-09-14 R/density dependence.Rmd
started exploring density dependence

---
title: "Density dependence"
author: "Eric R. Scott"
date: "2020-09-13"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(lme4)
library(bbmle)
library(car)
library(broom)
library(broom.mixed)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Start by looking for density dependence on growth, survival, and reproduction in continuous forests using the number of plants in the same grid-cell as a metric for density.

# Load Data

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```
Filter to just continuous forest
```{r}
ha_cf <- 
  ha %>% 
  filter(habitat == "CF") %>% 
  mutate(log_ht = log(ht),
         log_ht_prev = log(ht_prev))
```

I don't know what a height of 0 means...

```{r}
ha_cf <- ha_cf %>% filter(ht > 0, ht_prev > 0)
```

# Growth

```{r}
m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
summary(m)
m2 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year), data = ha_cf)
m3 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|plot), data = ha_cf)
m4 <- lm(log_ht ~ log_ht_prev + density_0_10, data = ha_cf)
AICtab(m, m2, m3, m4)
```

keep both random effects, possibly remove random effect of plot.

```{r}
m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
Anova(m)
```

Density matters?

```{r}
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()
```

residuals look leptokurtic.

```{r}
augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```

Try scaled T distribution

```{r}
library(mgcv)
library(gamm4)
m <- gamm4(log_ht ~ log_ht_prev + density_0_10, random = ~ (1|year) + (1|plot), family = scat, data = ha_cf)
m <- gam(log_ht ~ log_ht_prev + density_0_10, family = scat, data = ha_cf)
coef(m)
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()

augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```

Maybe something like a gamma distribution on un-transformed values?  Or use shoots..

```{r}
m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + (1|year) + (1|plot), data = ha_cf)
Anova(m)
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()

augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```
Ok, yeah.  That's a good reason to use # of shoots maybe?

```{r}
Anova(m)
```

Density matters!

```{r}
m <- lmer(log(shts) ~ log(shts_prev) * density_0_10 + (1|year) + (1|plot), data = ha_cf)
fixef(m)
Anova(m)
```

Interaction doesn't! (bless)

```{r}
m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + density_10_20 + density_20_30 + (1|year) + (1|plot), data = ha_cf)
fixef(m)
Anova(m)
```

Ah, density further away matters as well.  Collinearity, probably, so don't trust p-values!

Ok, gotta do GAMs I guess!

### 2020-09-15 R/competition wrangling.Rmd
add derived data for density calculations
changed density to be #shoots per cell

@@
 First, a small example as proof of concept:
+
 ```{r}
@@
   group_by(row_number, column) %>% 
-  summarize(n_plants = n()) %>% 
+  summarize(n_plants = n(),
+            n_shts = sum(shts)) %>% 
   rename(row = row_number)
@@
 #density between 5m and 15m away
-dist_5_15<-
+dist_5_15 <-
   dist_full %>% 
@@
             n_cells = n(),
-            density_5_15 = n_neighbors_5_15/n_cells)
+            density_5_15 = n_neighbors_5_15/n_cells,
+            shtsper_5_15 = sum(n_shts, na.rm = TRUE)/n())
 ```
-So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2
+So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2 or 3.66 shoots / 100m^2
 
@@
 
-Now, create that summary tibble counting the number of plants in each grid cell for every plot:year combo, join, and unnest.
+Now, create that summary tibble counting the number of shoots in each grid cell for every plot:year combo, join, and unnest.
 
@@
   #summarize each year:plot to count number of plants in each grid cell
-  mutate(n_plants_sum = map(data, ~{
+  mutate(n_shts_sum = map(data, ~{
     .x %>% 
-      add_column(plant = 1) %>% 
+      # add_column(plant = 1) %>% 
       group_by(row_number, column) %>% 
-      summarize(n_plants = n()) %>% 
+      summarize(n_shts = sum(shts)) %>% 
       rename(row = row_number)
@@
   #and column to every other row and column
-  mutate(dist_full = map2(dist_df, n_plants_sum, ~{
+  mutate(dist_full = map2(dist_df, n_shts_sum, ~{
     full_join(.x, .y, by = c("row_to" = "row", "column_to" = "column"))
@@
   group_by(plot, year, row_from, column_from) %>% 
-  summarize(density_0_10 = sum(n_plants, na.rm = TRUE)/n())
+  summarize(density_0_10 = sum(n_shts, na.rm = TRUE)/n())
 left_join(ha, dens_0_10, by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
@@
     group_by(plot, year, row_from, column_from) %>%
-    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_plants, na.rm = TRUE)/n())
+    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_shts, na.rm = TRUE)/n())
 })
@@
 That is, neighbor density for plants near the edges is based on a smaller sample than for plants near the center.
+
+# Write to cleaned data
+
+```{r}
+write_rds(ha_new, here("analysis", "data", "derived_data", "ha_density.rds"))
+```

### 2020-09-15 R/density dependence.Rmd
changed density to be #shoots per cell

@@
 library(broom.mixed)
+library(patchwork)
 
@@
 ```
+
 Filter to just continuous forest
+
 ```{r}
@@
 I don't know what a height of 0 means...
+Height of 0 probably means it was knocked over, or very small, or dead.
+
+Scale year
 
 ```{r}
-ha_cf <- ha_cf %>% filter(ht > 0, ht_prev > 0)
+ha_cf <- 
+  ha_cf %>% 
+  ungroup() %>% 
+  mutate(year_post = year - min(year))
 ```
@@
 
+For this I'm going to use shoot number. Very closely related to height, but better behaved distribution and not sensitive to plants being knocked over.
+
 ```{r}
-m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
-summary(m)
-m2 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year), data = ha_cf)
-m3 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|plot), data = ha_cf)
-m4 <- lm(log_ht ~ log_ht_prev + density_0_10, data = ha_cf)
-AICtab(m, m2, m3, m4)
+mpois <- glm(shts ~ log(shts_prev) * density_0_10,
+             family = poisson(link = "log"),
+             data = ha_cf)
+lm <- lm(shts ~ log(shts_prev) * density_0_10,
+         data = ha_cf)
+mnorm <- glm(shts ~ log(shts_prev) * density_0_10,
+             family = gaussian(link = "log"),
+             data = ha_cf)
+mpois2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
+                family = poisson(link = "log"),
+                data = ha_cf)
+
+mnorm2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
+                family = gaussian(link = "log"),
+                data = ha_cf)
+
+AICtab(mpois, mnorm, lm, mpois2, mnorm2)
 ```
 
-keep both random effects, possibly remove random effect of plot.
+Lognormal distribution fits best, even after attempting to account of overdispersion with random effects.  I guess despite being discrete, this isn't really "count" data *per se*.
+
+Model doesn't converge though.
 
 ```{r}
-m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
-Anova(m)
+mnorm3 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post),
+               family = gaussian(link = "log"),
+               data = ha_cf)
+
+AICtab(mnorm2, mnorm3)
 ```
 
-Density matters?
+Can remove plot ID.
 
 ```{r}
-augment(m) %>% 
-  ggplot(aes(.resid)) +
-  geom_histogram()
+mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post),
+               family = gaussian(link = "log"),
+               data = ha_cf)
+AICtab(mnorm3, mnorm4)
 ```
 
-residuals look leptokurtic.
+Keep plant ID as a random effect.
 
 ```{r}
-augment(m) %>% 
-  ggplot(aes(sample = .resid)) +
-  geom_qq() + geom_qq_line()
+mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number),
+               family = gaussian(link = "log"),
+               data = ha_cf)
+AICtab(mnorm3, mnorm4, mnorm5)
 ```
 
-Try scaled T distribution
+Can't remove year as random effect.
+
+Inspect residuals:
 
 ```{r}
-library(mgcv)
-library(gamm4)
-m <- gamm4(log_ht ~ log_ht_prev + density_0_10, random = ~ (1|year) + (1|plot), family = scat, data = ha_cf)
-m <- gam(log_ht ~ log_ht_prev + density_0_10, family = scat, data = ha_cf)
-coef(m)
-augment(m) %>% 
+h <- augment(mnorm2) %>% 
   ggplot(aes(.resid)) +
@@
 
-augment(m) %>% 
+qq <- augment(mnorm2) %>% 
   ggplot(aes(sample = .resid)) +
   geom_qq() + geom_qq_line()
+
+resid<- augment(mnorm2) %>% 
+  ggplot(aes(x = .fitted, y = .resid)) + geom_point() 
+h/(qq|resid)
 ```
+I don't love this, but it is a better fit than a poisson distribution.
 
-Maybe something like a gamma distribution on un-transformed values?  Or use shoots..
+# TODO:
 
-```{r}
-m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + (1|year) + (1|plot), data = ha_cf)
-Anova(m)
-augment(m) %>% 
-  ggplot(aes(.resid)) +
-  geom_histogram()
-
-augment(m) %>% 
-  ggplot(aes(sample = .resid)) +
-  geom_qq() + geom_qq_line()
-```
-Ok, yeah.  That's a good reason to use # of shoots maybe?
-
-```{r}
-Anova(m)
-```
-
-Density matters!
-
-```{r}
-m <- lmer(log(shts) ~ log(shts_prev) * density_0_10 + (1|year) + (1|plot), data = ha_cf)
-fixef(m)
-Anova(m)
-```
-
-Interaction doesn't! (bless)
-
-```{r}
-m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + density_10_20 + density_20_30 + (1|year) + (1|plot), data = ha_cf)
-fixef(m)
-Anova(m)
-```
-
-Ah, density further away matters as well.  Collinearity, probably, so don't trust p-values!
-
-Ok, gotta do GAMs I guess!
+- Density should be total shoot number
+- Should actually be density in previous year
+- ~~Shoots should be poisson distribution, not log-transformed?~~
+- incorporate 1-ha fragments and use habitat type as a predictor.
+- Implement FLM approach to density dependence.

### 2020-09-15 R/density dependence.Rmd
change density to lagged density.

@@
 ```
+Actually interested in density in the previous year
+
+```{r}
+ha <-
+  ha %>% 
+  group_by(ha_id_number) %>% 
+  arrange(ha_id_number, year) %>% 
+  mutate(across(starts_with("density"), lag)) %>% 
+  ungroup() %>% 
+  #scale year
+  mutate(year_post = year - min(year))
+```
 
@@
 Height of 0 probably means it was knocked over, or very small, or dead.
-
-Scale year
-
-```{r}
-ha_cf <- 
-  ha_cf %>% 
-  ungroup() %>% 
-  mutate(year_post = year - min(year))
-```
 
@@
 
-Can remove plot ID.
+Can't remove plot ID.
 
 ```{r}
-mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post),
+mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post) + (1|plot),
                family = gaussian(link = "log"),
                data = ha_cf)
-AICtab(mnorm3, mnorm4)
+AICtab(mnorm2, mnorm4)
 ```
 
-Keep plant ID as a random effect.
+Can't remove plant ID
 
 ```{r}
-mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number),
+mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|plot),
                family = gaussian(link = "log"),
                data = ha_cf)
-AICtab(mnorm3, mnorm4, mnorm5)
+AICtab(mnorm2, mnorm3, mnorm4, mnorm5)
 ```
@@
 ```{r}
-h <- augment(mnorm2) %>% 
-  ggplot(aes(.resid)) +
+ha_aug <- augment(mnorm2)
+h <-
+  ggplot(ha_aug, aes(.resid)) +
   geom_histogram()
 
-qq <- augment(mnorm2) %>% 
-  ggplot(aes(sample = .resid)) +
+qq <-
+  ggplot(ha_aug, aes(sample = .resid)) +
   geom_qq() + geom_qq_line()
 
-resid<- augment(mnorm2) %>% 
-  ggplot(aes(x = .fitted, y = .resid)) + geom_point() 
+resid<- 
+  ggplot(ha_aug, aes(x = .fitted, y = .resid)) + geom_point() 
 h/(qq|resid)
@@
 
-- Density should be total shoot number
-- Should actually be density in previous year
+- ~~Density should be total shoot number~~
+- ~~Should actually be density in previous year~~
 - ~~Shoots should be poisson distribution, not log-transformed?~~

### 2020-09-15 R/density dependence.Rmd
added survival analysis

@@
 library(patchwork)
+library(performance)
 
@@
   arrange(ha_id_number, year) %>% 
+  # mutate(across(starts_with("density"), lag, .names = "prev_{col}")) %>%
   mutate(across(starts_with("density"), lag)) %>% 
   ungroup() %>% 
+  # filter(!is.na(prev_density_0_10)) %>% 
   #scale year
@@
   filter(habitat == "CF") %>% 
-  mutate(log_ht = log(ht),
-         log_ht_prev = log(ht_prev))
+  mutate(log_shts = log(shts),
+         log_shts_prev = log(shts_prev))
+
 ```
@@
 
+## Distribution
+
+Number of shoots is discrete, but not *really* count data.  I'll try a poisson distribution and a lognormal distribution.
+
 ```{r}
-mpois <- glm(shts ~ log(shts_prev) * density_0_10,
+mpois <- glm(shts ~ log_shts_prev + density_0_10 + density_10_20,
              family = poisson(link = "log"),
              data = ha_cf)
-lm <- lm(shts ~ log(shts_prev) * density_0_10,
+lm <- lm(shts ~ log_shts_prev + density_0_10 + density_10_20,
          data = ha_cf)
-mnorm <- glm(shts ~ log(shts_prev) * density_0_10,
+mnorm <- glm(shts ~ log_shts_prev + density_0_10 + density_10_20,
              family = gaussian(link = "log"),
              data = ha_cf)
-mpois2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
+mpois2 <- glmer(shts ~ log_shts_prev + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post) + (1|plot),
                 family = poisson(link = "log"),
@@
 
-mnorm2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
+check_overdispersion(mpois2) #actually underdispersed
+
+mnorm2 <- glmer(shts ~ log_shts_prev + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post) + (1|plot),
                 family = gaussian(link = "log"),
@@
 AICtab(mpois, mnorm, lm, mpois2, mnorm2)
+compare_performance(mpois, mnorm, mpois2, mnorm2, rank = TRUE)
 ```
 
-Lognormal distribution fits best, even after attempting to account of overdispersion with random effects.  I guess despite being discrete, this isn't really "count" data *per se*.
-
-Model doesn't converge though.
+Lognormal distribution fits best according to AIC, although poisson has lower RMSE.
+I guess despite being discrete, this isn't really "count" data *per se*.
+The lognormal model doesn't problems with singularity that the poisson model has.
+Both have problems with heteroscedasticity.
 
 ```{r}
-mnorm3 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post),
+check_convergence(mnorm2)
+check_singularity(mpois2)
+check_singularity(mnorm2)
+x<-check_heteroscedasticity(mnorm2)
+plot(x)
+```
+
+## Random effects
+
+```{r}
+mnorm3 <- glmer(shts ~ log_shts_prev * density_0_10 + (1|ha_id_number) + (1|year_post),
                family = gaussian(link = "log"),
@@
 ```{r}
-mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post) + (1|plot),
+mnorm4 <- glmer(shts ~ log_shts_prev * density_0_10 + (1|year_post) + (1|plot),
                family = gaussian(link = "log"),
@@
 ```{r}
-mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|plot),
+mnorm5 <- glmer(shts ~ log_shts_prev * density_0_10 + (1|ha_id_number) + (1|plot),
                family = gaussian(link = "log"),
@@
 
-Can't remove year as random effect.
-
-Inspect residuals:
+## Does density matter beyond immediate surroundings?
 
 ```{r}
-ha_aug <- augment(mnorm2)
-h <-
-  ggplot(ha_aug, aes(.resid)) +
-  geom_histogram()
+Anova(mnorm2)
+```
 
-qq <-
-  ggplot(ha_aug, aes(sample = .resid)) +
-  geom_qq() + geom_qq_line()
+Apparently not! Weird! It is highly significant when I was accidentally modeling density in the same year. This is not due to removing rows from the first year either.  This is somewhat concerning!
 
-resid<- 
-  ggplot(ha_aug, aes(x = .fitted, y = .resid)) + geom_point() 
-h/(qq|resid)
+# Survival
+
+```{r}
+m_surv <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot) + (1|ha_id_number) + (1|year), 
+                family = binomial, data = ha_cf)
 ```
-I don't love this, but it is a better fit than a poisson distribution.
+
+## Random effects
+
+```{r}
+m_surv2 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot) + (1|year), 
+                family = binomial, data = ha_cf)
+AICtab(m_surv, m_surv2)
+```
+Can remove plant ID
+```{r}
+m_surv3 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|year), 
+                family = binomial, data = ha_cf)
+AICtab(m_surv2, m_surv3)
+```
+Can't remove plot
+
+```{r}
+m_surv4 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot), 
+                family = binomial, data = ha_cf)
+AICtab(m_surv2, m_surv4)
+```
+Can't remove year.
+
+## Density dependence?
+
+```{r}
+Anova(m_surv2)
+```
+
+Yes, but only immediate grid cell
 
@@
 
+- try negative binomial distribution for shoot number
 - ~~Density should be total shoot number~~

### 2020-09-15 R/density dependence.Rmd
added flowering probability

@@
 
+Should this be a function of `log_shts` or `log_shts_prev`?  Think about how you've coded survival.
+
 ```{r}
-m_surv <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot) + (1|ha_id_number) + (1|year), 
+m_surv <- glmer(surv ~ log_shts + density_0_10 + density_10_20 + (1|plot) + (1|ha_id_number) + (1|year), 
                 family = binomial, data = ha_cf)
@@
 ```{r}
-m_surv2 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot) + (1|year), 
+m_surv2 <- glmer(surv ~ log_shts + density_0_10 + density_10_20 + (1|plot) + (1|year), 
                 family = binomial, data = ha_cf)
@@
 ```{r}
-m_surv3 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|year), 
+m_surv3 <- glmer(surv ~ log_shts + density_0_10 + density_10_20 + (1|year), 
                 family = binomial, data = ha_cf)
@@
 ```{r}
-m_surv4 <- glmer(surv ~ log_shts_prev + density_0_10 + density_10_20 + (1|plot), 
+m_surv4 <- glmer(surv ~ log_shts + density_0_10 + density_10_20 + (1|plot), 
                 family = binomial, data = ha_cf)
@@
 Anova(m_surv2)
+fixef(m_surv2) %>% plogis()
 ```
 
-Yes, but only immediate grid cell
+Marginal, and only immediate grid cell.  Density has positive affect on survival.  So, probably just good patches.
+
+# Flowering probability
+
+```{r}
+ggplot(ha_cf, aes(infl)) + geom_histogram()
+```
+Make a binary column
+
+```{r}
+ha_cf <- 
+  ha_cf %>% 
+  mutate(flwr = as.integer(infl>0), .before = infl)
+
+#the way suvival is currently coded, a plant can flower and then not survive in the same year, so no need to filter out dead plants.
+ha_cf %>% filter(surv == 0)
+```
+
+```{r}
+m_flwr <- glmer(flwr ~ log_shts_prev + density_0_10 + density_10_20 + (1|ha_id_number) + (1|plot) + (1|year), family = binomial, data = ha_cf)
+```
+
+## Random effects
+
+```{r}
+m_flwr2 <- glmer(flwr ~ log_shts_prev + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year), family = binomial, data = ha_cf)
+AICtab(m_flwr, m_flwr2)
+```
+can get rid of plot ID
+
+```{r}
+m_flwr3 <- glmer(flwr ~ log_shts_prev + density_0_10 + density_10_20 + (1|year), family = binomial, data = ha_cf)
+AICtab(m_flwr, m_flwr2, m_flwr3)
+```
+keep plant ID
+
+```{r}
+m_flwr4 <- glmer(flwr ~ log_shts_prev + density_0_10 + density_10_20 + (1|ha_id_number), family = binomial, data = ha_cf)
+AICtab(m_flwr, m_flwr2, m_flwr3, m_flwr4)
+```
+
+keep year
+
+## Density dependence?
+
+```{r}
+Anova(m_flwr2)
+fixef(m_flwr2) %>% plogis()
+```
+
+Ok, there's an effect, but oddly only past 20m.  Again, positive density dependence.
 
@@
 
+- Check number of inflorescencesa given flowered?
+- check that survival is coded right so that it is a function of `ht_prev`
 - try negative binomial distribution for shoot number

### 2020-09-16 R/density dependence.Rmd
checked convergence issue and cleaned up code.  Added 1-ha fragments and habitat as a predictor

--- 2020-09-15 R/density dependence.Rmd
@@
 ```
-Actually interested in density in the previous year
 
@@
   arrange(ha_id_number, year) %>% 
-  mutate(across(starts_with("density"), lag)) %>% 
   ungroup() %>% 
@@
   filter(habitat == "CF") %>% 
-  mutate(log_ht = log(ht),
-         log_ht_prev = log(ht_prev))
+  mutate(log_shts = log(shts),
+         log_shts_next = log(shts_next))
+```
+
+Or filter to just 1-ha and cf
+
+```{r}
+ha_sub <-
+  ha %>% 
+  filter(habitat %in% c("CF", "1-ha")) %>% 
+  mutate(log_shts = log(shts),
+         log_shts_next = log(shts_next),
+         habitat = as.factor(habitat))
 ```
@@
 
+# General model structure
+
+For each vital rate I'll include the number of shoots and habitat as fixed effects and plant ID and year as random effects.
+I'm not including plot since there are few plots in each habitat type, and preliminary modeling shows the variance for plot is the lowest of those three random effects.
+
+```{r}
+ha_sub %>% 
+  group_by(habitat) %>% 
+  summarize(length(unique(plot)))
+```
+
 # Growth
@@
 
+## Distribution
+
+Number of shoots is discrete, but not *really* count data.  I'll try a poisson distribution and a lognormal distribution.
+
 ```{r}
-mpois <- glm(shts ~ log(shts_prev) * density_0_10,
+mpois <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
                 family = poisson(link = "log"),
-             data = ha_cf)
-lm <- lm(shts ~ log(shts_prev) * density_0_10,
-         data = ha_cf)
-mnorm <- glm(shts ~ log(shts_prev) * density_0_10,
+                data = ha_sub)
+
+check_overdispersion(mpois2) #actually underdispersed
+
+mnorm <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
                 family = gaussian(link = "log"),
-             data = ha_cf)
-mpois2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
-                family = poisson(link = "log"),
-                data = ha_cf)
-
-mnorm2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
-                family = gaussian(link = "log"),
-                data = ha_cf)
-
-AICtab(mpois, mnorm, lm, mpois2, mnorm2)
+                data = ha_sub)
 ```
 
-Lognormal distribution fits best, even after attempting to account of overdispersion with random effects.  I guess despite being discrete, this isn't really "count" data *per se*.
+But how bad is this convergence failure?
 
-Model doesn't converge though.
+Is the max|grad| calculation correct?
 
 ```{r}
-mnorm3 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post),
-               family = gaussian(link = "log"),
-               data = ha_cf)
+derivs1 <- mnorm@optinfo$derivs
+sc_grad1 <- with(derivs1, solve(Hessian, gradient))
+max(abs(sc_grad1))
+```
+No, it's actually a lot smaller than `tol`, so nothing to worry about
 
-AICtab(mnorm2, mnorm3)
+```{r}
+AICtab(mpois, mnorm)
+compare_performance(mpois, mnorm, rank = TRUE)
 ```
 
-Can't remove plot ID.
+Lognormal distribution fits best according to AIC, although poisson has lower RMSE.
+I guess despite being discrete, this isn't really "count" data *per se*.
+The lognormal model doesn't have problems with singularity that the poisson model has.
+Both have problems with heteroscedasticity.
+I'll go with log-normal.
 
 ```{r}
-mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post) + (1|plot),
+m_growth <- mnorm
+```
+
+```{r}
+x<-check_heteroscedasticity(m_growth)
+plot(x)
+```
+
+## Random effects
+
+```{r}
+m_growth2 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year_post),
                family = gaussian(link = "log"),
-               data = ha_cf)
-AICtab(mnorm2, mnorm4)
+               data = ha_sub)
+AICtab(m_growth, m_growth2)
 ```
@@
 ```{r}
-mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|plot),
+m_growth3 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
                family = gaussian(link = "log"),
-               data = ha_cf)
-AICtab(mnorm2, mnorm3, mnorm4, mnorm5)
+               data = ha_sub)
+AICtab(m_growth, m_growth2, m_growth3)
 ```
 
-Can't remove year as random effect.
-
-Inspect residuals:
+## Does density matter beyond immediate surroundings?
 
 ```{r}
-ha_aug <- augment(mnorm2)
-h <-
-  ggplot(ha_aug, aes(.resid)) +
-  geom_histogram()
+Anova(m_growth)
+```
 
-qq <-
-  ggplot(ha_aug, aes(sample = .resid)) +
-  geom_qq() + geom_qq_line()
+Odd, density further out is significant, but density close is not.
 
-resid<- 
-  ggplot(ha_aug, aes(x = .fitted, y = .resid)) + geom_point() 
-h/(qq|resid)
+```{r}
+fixef(m_growth) %>% exp()
 ```
-I don't love this, but it is a better fit than a poisson distribution.
+
+Both weakly positive density dependence
+
+# Survival
+
+Pre-reproductive census, so plants live/die in same year as `shts` and `ht`
+
+```{r}
+m_surv <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year), 
+                family = binomial, data = ha_sub)
+```
+
+## Random effects
+
+```{r}
+m_surv2 <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year), 
+                family = binomial, data = ha_sub)
+AICtab(m_surv, m_surv2)
+```
+
+Close to equivalent, so remove plant ID.  This makes sense, as plants only die once!
+
+```{r}
+m_surv3 <- glm(surv ~ log_shts + habitat + density_0_10 + density_10_20, 
+                family = binomial, data = ha_sub)
+AICtab(m_surv2, m_surv3)
+```
+Can't remove year.
+
+## Density dependence?
+
+```{r}
+Anova(m_surv2)
+fixef(m_surv2) %>% plogis()
+```
+
+Marginal, and only immediate grid cell.  Density has positive affect on survival.  So, probably just good patches.
+
+# Flowering probability
+
+```{r}
+ggplot(ha_sub, aes(infl)) + geom_histogram()
+```
+Make a binary column
+
+```{r}
+ha_sub <- 
+  ha_sub %>% 
+  mutate(flwr = as.integer(infl>0), .before = infl)
+
+#the way suvival is currently coded, a plant can flower and then not survive in the same year, so no need to filter out dead plants.
+ha_sub %>% filter(surv == 0 & flwr == 1)
+```
+
+```{r}
+m_flwr <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year),
+                family = binomial, data = ha_sub)
+```
+
+## Random effects
+
+```{r}
+m_flwr2 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year),
+                 family = binomial, data = ha_sub)
+AICtab(m_flwr, m_flwr2)
+```
+Can't remove plant ID.
+
+```{r}
+m_flwr3 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
+                 family = binomial, data = ha_sub)
+AICtab(m_flwr, m_flwr2, m_flwr3)
+```
+
+Can't remove year
+
+## Density dependence?
+
+```{r}
+Anova(m_flwr)
+fixef(m_flwr) %>% plogis()
+```
+
+Ok, there's an effect, but oddly only past 20m.  Again, positive density dependence.
 
@@
 
+- Check number of inflorescencesa given flowered?
+- ~~check that survival is coded right so that it is a function of `ht_prev`~~
+- try negative binomial distribution for shoot number
 - ~~Density should be total shoot number~~
@@
 - ~~Shoots should be poisson distribution, not log-transformed?~~
-- incorporate 1-ha fragments and use habitat type as a predictor.
+- ~~incorporate 1-ha fragments and use habitat type as a predictor.~~
 - Implement FLM approach to density dependence.

### 2020-09-16 R/competition wrangling.Rmd
corrected density at closest distance by subtracting shoots of focal plant.

@@
 I want to use the `row` and `colunn` to figure out how many neighbors each plant has. I should exclude plants on edges, because I don't know about their neighbors outside of the plot.
+# TODO:
+
+- density at distance 0 shouldn't include the focal plant
 
@@
 ```{r}
-#density between 5m and 15m away
-dist_5_15 <-
+#density between 0m and 10m away
+dist_0_10 <-
   dist_full %>% 
-  filter(dist >=5, dist <15) %>% 
+  filter(dist >=0, dist <10) %>% 
   group_by(row_from, column_from) %>% 
-  summarize(n_neighbors_5_15 = sum(n_plants, na.rm = TRUE),
+  summarize(n_neighbors_0_10 = sum(n_plants, na.rm = TRUE),
             n_cells = n(),
-            density_5_15 = n_neighbors_5_15/n_cells,
-            shtsper_5_15 = sum(n_shts, na.rm = TRUE)/n())
+            density_0_10 = n_neighbors_0_10/n_cells,
+            shtsper_0_10 = sum(n_shts, na.rm = TRUE)/n())
 ```
@@
 ```{r}
-full_test <- full_join(ha_test, dist_5_15, by = c("row_number" = "row_from", "column" = "column_from"))
+full_test <- full_join(ha_test, dist_0_10, by = c("row_number" = "row_from", "column" = "column_from"))
 head(full_test)
+```
+For the nearest distances (0-10m), I need to subtract the number of shoots in the focal plant.
+
+```{r}
+full_test %>% 
+  # mutate(shtsper_0_10 = shtsper_0_10 - shts) %>% 
+  arrange(row, column)
+
+full_test %>% 
+  group_by(row, column) %>% 
+  summarize(n = n(), n_shts = sum(shts))
 ```
@@
   ha_dist %>% 
-  filter(dist>0 & dist <=10) %>% 
+  filter(dist>=0 & dist <10) %>% 
   group_by(plot, year, row_from, column_from) %>% 
@@
 ```
+Finally, I need to correct the closest density by subtracting shoots of the focal plant.
+
+```{r}
+ha_new <- ha_new %>% mutate(density_0_10 = density_0_10 - shts)
+```
 

### 2020-09-17 R/density dependence.Rmd
figuring out crossbasis smooth

@@
 conflict_prefer("filter", "dplyr")
+conflict_prefer("select", "dplyr")
 ```
@@
 
-check_overdispersion(mpois2) #actually underdispersed
+check_overdispersion(mpois) #actually underdispersed
 
@@
 ```{r}
-fixef(m_growth) %>% exp()
+fixef(m_growth)
 ```
@@
 Anova(m_surv2)
-fixef(m_surv2) %>% plogis()
+fixef(m_surv2)
 ```
 
-Marginal, and only immediate grid cell.  Density has positive affect on survival.  So, probably just good patches.
+Marginal effects on survival.
 
@@
 Anova(m_flwr)
-fixef(m_flwr) %>% plogis()
+fixef(m_flwr)
 ```
@@
 
+Probability of flowering increases with size and is lower in continuous forest
+
+# Kernel approach
+
+So it seems like there are effect of density at least out to 20m.
+I'd like to take the approach in Teller et al.
+2016, but the code is so convoluted that it is not adaptable.
+I want to start by trying to adapt the `dlnm` approach instead.
+
+## Set up the crossbasis
+
+```{r}
+Q <-
+  ha_sub %>%
+  select(density_0_10:density_50_60) %>% 
+  as.matrix()
+
+L <-
+  matrix(seq(5, 55, length.out = 6),
+         nrow = nrow(Q), 
+         ncol = 6,
+         byrow = TRUE)
+```
+
+## GAM
+
+```{r}
+library(dlnm)
+library(mgcv)
+
+# Q_log <- log(Q+1)
+g_growth <-
+  gam(shts_next ~ s(log_shts) +
+        habitat +
+        s(Q, L,
+          bs = "cb",
+          k = c(5, 4),
+          xt = list(bs = "cr")),
+      family = gaussian(link = "log"),
+      data = ha_sub,
+      method = "REML")
+```
+
+## Predict
+
+```{r}
+x <- crosspred("Q", model = g_growth, cumul = TRUE)
+summary(x)
+```
+
+## Plot crossbasis
+
+### Contour plot
+
+```{r}
+plot.crosspred(x,
+               ptype = "contour",
+               xlab = "#shoots/100m^2",
+               ylab = "distance (m)",
+               cumul = FALSE)
+```
+So this shows negative density dependence that seems to be strongest at about 40-50 m away from the focal plant.
+
+### Slices
+
+We can also plot slices through the contour plot at different distances or different shoot densities
+
+```{r}
+plot.crosspred(x, ptype = "slice", var = 100, xlab = "distance (m)", main = "Effect of 100 shoots/100m^2")
+plot.crosspred(x, ptype = "slice", var = 10, xlab = "distance (m)", main = "Effect of 10 shoots/100m^2")
+```
+
+### Cumulative
+
+I think this shows the cumulative effect of density summed/integrated over all distances.
+So, for example, a plant with about 30 shoots/100m^2 surrounding it out to the maximum distance (60m) would have the largest number of shoots, on average.
+
+```{r}
+plot(x, ptype = "overall", xlab = "density (shoots/100m^2)", cumul = TRUE)
+```
+
+There is no built-in function to get cumulative effects of density over all distances.
+
+## Other shapes
+
+It is odd that the strength of competition seems to be increasing with distance.
+I'd like to be able to force the distance dimension to be monotonically decreasing as was done in Teller et al (I think?).
+Teller does this *after* fitting a spline using `mk_dist_wts()`, which is a function that takes the spline coefficients, and predicted values from the spline, and transforms it into a non-negative, monotonic decreasing function.
+
+You can override one of the dimensions of the cross basis with any custom function.
+
+```{r}
+#use arglag to override lag dimension of crossbasis, use argvar to override plant density dimension.
+lin <- dlnm:::lin
+xt <- list(bs = "cr", arglag = list(fun = "lin"))
+
+g_growth <- 
+  gam(shts_next ~ s(log_shts) +
+        habitat +
+        s(Q, L,
+          bs = "cb",
+          k = c(5, 4),
+          fx = c(F, T), #turns off penalization for the lag dimension
+          xt = xt),
+      family = gaussian(link = "log"),
+      data = ha_sub,
+      gamma = 1.5)
+
+gam.check(g_growth)
+```
+
+```{r}
+x <- crosspred("Q", model = g_growth)
+plot.crosspred(x, ptype = "contour", xlab = "#shoots/100m^2", ylab = "distance (m)")
+
+plot.crosspred(x, ptype = "slice", var = 100)
+plot.crosspred(x, ptype = "slice", var = 10)
+```
+Now the kernel is the same shape at every value of density.  However, this relationship doesn't make any sense.  Fewer knots?  More penalty?  Custom function?
+
+```{r}
+anova(g_growth)
+```
+
 # TODO:
 
-- Check number of inflorescencesa given flowered?
+- ~~is density at 0 distance including the focal plant?  It shouldn't.~~
+- Check number of inflorescences given flowered?
+- ~~log-transform density (nah, not right now.  DOesn't help with convergence)~~
 - ~~check that survival is coded right so that it is a function of `ht_prev`~~

### 2020-09-21 R/demography_wrangling.R
data with x and y coords for one year

@@ -56,7 +56,7 @@ demog <-
 demog <- 
   demog %>%
   select(ranch, bdffp_reserve_no, plot, habitat, #site level
-         row, column,ha_id_number, tag_number, #plot level
+         row, column, x_09, y_09, ha_id_number, tag_number, #plot level
          year,
          ht, ht_next, shts, shts_next, infl, surv, code_notes, code2) #plant level
 

### 2020-09-21 R/competition wrangling.Rmd
calculate number of plants within rings of different radii

@@
   filter(year == first(year)) %>%
-  filter(row_number <=3, column <=3) %>% 
+  filter(row_number <= 3, column <= 3) %>% 
   add_column(plant = 1) #just a dummy indicator column for use later
@@
   dist_full %>% 
-  filter(dist >=0, dist <10) %>% 
+  filter(dist >= 0, dist < 10) %>% 
   group_by(row_from, column_from) %>% 
@@
   ha_dist %>% 
-  filter(dist>=0 & dist <10) %>% 
+  filter(dist>= 0 & dist < 10) %>% 
   group_by(plot, year, row_from, column_from) %>% 
@@
 max <- 100 #meters
-L <- seq(0, max-width, by = width)
+L <- seq(0, max - width, by = width)
 U <- seq(0 + width, max, by = width)
@@
 ```
+
+# Using exact x, y coords
+```{r}
+library(furrr)
+plan(multisession)
+```
+
+For a subset of plants in 2009, exact x,y coords in the plot were taken.  Use these to estimate density.
+
+```{r}
+ha_xy <-
+  ha_new %>% 
+  filter(!is.na(x_09) & !is.na(y_09))
+ha_xy
+```
+It looks like the x, y coords are *within* each grid cell (e.g. A1).  Need to offset them to get relative coords for entire plot.  I'll make A1 = 0,0
+
+```{r}
+ha_xy <-
+  ha_xy %>% 
+  mutate(x_adj = x_09 + ((column - 1) * 10),
+         y_adj = y_09 + ((row_number - 1) * 10),
+         .after = y_09)
+```
+
+```{r}
+ha_xy_s <-
+  ha_xy %>% 
+  group_by(ha_id_number) %>% 
+  summarize(x = mean(x_adj),
+            y = mean(y_adj))
+#down to 3,571--still too big to do dist(), I think?
+
+test <- ha_xy_s %>% select(x, y) %>% as.matrix()
+rownames(test) <- paste(test[,1], test[,2], sep = ",")
+head(test)
+dist(test[1:5,])
+
+```
+Might not be the best approach
+
+```{r}
+eu_dist <- function(p1, p2) {
+  sqrt(sum((p1 - p2)^2))
+}
+x = c(0,0)
+y = c(3,3)
+eu_dist(x, y)
+```
+
+For every ha_id_number, find only plants within 0-10 m, sum the number of shoots.
+
+```{r}
+ha_xy %>% 
+  group_by(ha_id_number) %>% 
+  rowwise() %>% 
+  filter(eu_dist(
+    c(first(cur_data()$x_adj), first(cur_data()$y_adj)),
+    c(x_adj, y_adj)) < 10)
+```
+
+Nope
+
+Start with single plot
+
+```{r}
+y1p1 <- ha_xy %>% 
+  filter(year == first(year),
+         plot == first(plot))
+y1p1
+```
+
+```{r}
+add_density_cols <- function(df, L, U, by, verbose = FALSE) {
+
+  eu_dist <- function(p1, p2) {
+    sqrt(sum((p1 - p2)^2))
+  }
+
+  d <- seq(L, U, by = by)
+
+  df_out <- 
+    future_map_dfr(1:nrow(df), function(r) { #loop through rows/plants
+      #get coordinates for that plant
+      focal_plant <- t(df[r, c("x_adj", "y_adj")])
+
+      #calculate distance to every other plant
+      focal_df <-
+        df %>%
+        rowwise() %>%
+        mutate(dist = eu_dist(focal_plant, c(x_adj, y_adj)))
+
+      #get total number of shoots at every ring
+      future_map2_dfc(lag(d)[-1], d[-1],
+               ~ transmute(focal_df,
+                           !!glue("shts_{.x}_{.y}") := shts * (dist > .x & dist <= .y))) %>%
+        ungroup() %>% 
+        summarize(across(everything(), ~sum(., na.rm = TRUE)))
+    })
+
+  df_out <- bind_cols(df, df_out) #join to original data
+  return(df_out)
+}
+# tic()
+# add_density_cols(y1p1, L = 0, U = 10, by = 5)
+# toc()
+```
+
+If I want this to be in units of shoots/m^2, then I need to calculate the area of the donut:
+
+$$
+A = \pi(U^2 - L^2)
+$$
+But also, like, truncate that donut if it passes outside of the plot.  I don't know how to do that.
+
+```{r}
+tic()
+ha_xy2 <-
+  ha_xy %>% 
+  group_by(plot, year) %>% 
+  group_split() %>% 
+  head() %>% #test with 6 df to see how long it will take.
+  future_map(~{
+      add_density_cols(.x, L = 0, U = 10, by = 5) 
+    })
+toc()
+
+```
+
+Uggggh only works for first plot???

### 2020-09-23 R/competition wrangling.Rmd
updated todo
Removed density calculations based on grid cells and replaced with calculations based on exact x,y coordinates.

--- 2020-09-15 R/competition wrangling.Rmd
@@
 library(skimr)
+library(furrr)
+library(janitor)
 
@@
 
-Competition kernel. Teller et al.
+# TODO:
 
-I want to use the `row` and `colunn` to figure out how many neighbors each plant has. I should exclude plants on edges, because I don't know about their neighbors outside of the plot.
+- filter out focal plants where donut goes out of plot. (filter(distance_to_edge > outer_circle_radius))
+- Using all years is not estimating density accurately.
+Focal plants are not moving, but in other years there are neighbors without known coordinates.
+Solution: only use the three transitions where we know the coords of every plant?
+Or assume births and deaths don't change density significantly?
 
@@
 ```{r data, echo=TRUE}
+# demographic data
 ha <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
+# data from previous study: Uriarte M, Bruna EM, Rubim P, et al (2010) Effects
+# of forest fragmentation on the seedling recruitment of a tropical herb:
+# assessing seed vs. safe-site limitation. Ecology 91:1317–1328.
+# https://doi.org/10.1890/09-0785.1
+
+coords <-
+  read.table(here("analysis", "data", "raw_data", "adults0607_light.txt")) %>%
+  as_tibble() %>%
+  clean_names()
 ```
 
-Make numeric row and column numbers for easier data handling later.
+`ha` contains x and y coords for each plant from 2009 *within each grid cell*.
+`coords` contains x and y coordinates for each plant in 2006 and 2007 relative to the entire plot.
+I need to join these data sources and offset the coordinates in the `ha` dataset correctly.
+I'm going to remove `year` from the `coords` dataset and assume that plants aren't moving.
+
+# Join data
+
+Getting things to match up:
 
 ```{r}
-# as.factor(ha$row) #levels are alphabetical order by default, which is perfect.  Except for row "L", not sure what that's about.
-ha <- 
-  ha %>% 
-  mutate(row_number = as.integer(as.factor(row)), .after = row,
-         column = as.integer(column)) %>% 
-  filter(row != "L") # just get rid of the few plants in a row "L".--only affects one plot
+coords2 <- 
+  coords %>% 
+  select(plot = site, tag_number, row, column, plot_x, plot_y, x, y) %>% 
+  mutate(plot = case_when(
+    plot == "Cabo Frio" ~ "CaboFrio-CF",
+    plot == "Dimona CF" ~ "Dimona-CF",
+    plot == "Florestal" ~ "Florestal-CF",
+    plot == "PA-CF" ~ "PortoAlegre-CF",
+    TRUE ~ plot
+  )) 
+```
+
+Summarize so there is only one set of coordinates per plant. Averaging as I'm assuming any difference in coordinates between '06 and '07 is not because the plant moved.
+
+```{r}
+coords3 <-
+  coords2 %>%
+  group_by(plot, tag_number) %>% 
+  # rowwise() %>% 
+  summarize(across(where(is.character), first),
+            across(where(is.numeric), mean)) %>% 
+  rename(x_07 = x,
+         y_07 = y)
 ```
@@
 ```{r}
-skim(ha)
+ha_coords <- left_join(ha, coords3, by = c("plot", "row", "column", "tag_number"))
+```
+# Tidy
+
+Now, use the plot_x and plot_y variables to correctly offset the x and y coords from 2009.
+
+```{r}
+ha_coords2 <- 
+  ha_coords %>% 
+  mutate(x_adj = x_09 + ((plot_x - 1) * 10),
+         y_adj = y_09 + ((plot_y - 1) * 10),
+         .after = y_09)
 ```
 
-# Calculating density at different distances
-First, a small example as proof of concept:
+Now, I'll just average across the three years and assume that any differences are measurement error and not because plants moved.
 
 ```{r}
-ha_test <-
-  ha %>%
-  filter(plot == first(plot)) %>%
-  filter(year == first(year)) %>%
-  filter(row_number <=3, column <=3) %>% 
-  add_column(plant = 1) #just a dummy indicator column for use later
-ha_test
+ha_coords3 <-
+  ha_coords2 %>% 
+  rowwise() %>% 
+  mutate(x = mean(c(x_adj, x_07), na.rm = TRUE),
+         y = mean(c(y_adj, y_07), na.rm = TRUE)) %>% 
+  #replace NaN's generated when mean of only NAs is taken
+  mutate(across(c(x, y), ~ifelse(is.nan(.), NA, .))) %>% 
+  #remove intermediate coordinates
+  select(ranch, bdffp_reserve_no, plot, habitat,
+         ha_id_number, tag_number, row, column,
+         x, y, year, ht, ht_next, shts, shts_next,
+         infl, surv, code_notes, code2)
 ```
 
-I'll make use of `expand_grid()` to create every combination of row and column (cell), and `dist()` to calculate the distance between every cell and every other cell.
+# Filter
+
+To model the effect of density, I'm only going to use the subset of plants that I have coordinates for.
+
+**NOTE:** There is a **problem** with this.  By using all the years and filtering the data to just include plants with coordinates from 2006, 2007, and 2009, I am not estimating density accurately.
+```{r}
+ha_xy <-
+  ha_coords3 %>% 
+  filter(!is.na(x), !is.na(y)) %>% 
+  ungroup()
+```
+
+# Calculate density
+
+Now the challenge is to calculate the density of surrounding shoots for every focal plant at different distances.
 
 ```{r}
-#all possible combination of row and column ID
-grid_coords <- expand_grid(row = 1:3, col = 1:3) %>% as.matrix()
-
-#calculate distrance from every coord to every other coord
-d <- dist(grid_coords*10, #because each grid cell is 10m x 10m
-          diag = FALSE,
-          upper = TRUE) %>% 
-  as.matrix()
-
-rownames(d) <- colnames(d) <- paste(grid_coords[ , 1], grid_coords[ , 2], sep = ",")
-d
+# function to calculate euclidean distance between two points
+eu_dist <- function(p1, p2) {
+  sqrt(sum((p1 - p2)^2))
+}
+# x = c(0,0)
+# y = c(3,3)
+# eu_dist(x, y)
 ```
 
-Now I can turn this into a tidy/long format. This shows the distance from (`row_from`, `column_from`) to (`row_to`, `column_to`).
+## Proof of concept
+
+Start with single plot
 
 ```{r}
-dist_df <-
-  as.data.frame(d) %>% 
-  rownames_to_column(var = "from") %>% 
-  pivot_longer(-from, names_to = "to", values_to = "dist") %>% 
-  separate(from, into = c("row_from", "column_from")) %>% 
-  separate(to, into = c("row_to", "column_to")) %>% 
-  mutate(across(where(is.character), as.integer))
-head(dist_df)
-```
-
-Then we can summarize the real data to figure out how many plants are in each grid cell.  Here, I'm just using a small subset that corresponds to a single 3x3 plot
-
-```{r}
-plant_summary <-
-  ha_test %>% 
-  group_by(row_number, column) %>% 
-  summarize(n_plants = n(),
-            n_shts = sum(shts)) %>% 
-  rename(row = row_number)
-plant_summary
-```
-Now I can join these two data frames.
-
-```{r}
-dist_full <- full_join(dist_df, plant_summary, by = c("row_to" = "row", "column_to" = "column"))
-head(dist_full)
-```
-The way to read this is for a plant in [1,1] the cell [1,1] is 0m away and has 3 plants in it.
-For a plant in [1,1], cell [1,2] is 10m away and there is 1 plant in it.
-For a plant in [1,1], cell [2,3] is 22.36m away and there are 9 plants in it, etc.
-
-Here's another way to visualize this using `xtabs()`:
-
-```{r}
-xtabs(plant ~ row_number + column, ha_test)
-```
-
-So now I can filter this and summarize to calculate neighbor density at arbitrary distances for every grid cell.
-
-```{r}
-#density between 5m and 15m away
-dist_5_15 <-
-  dist_full %>% 
-  filter(dist >=5, dist <15) %>% 
-  group_by(row_from, column_from) %>% 
-  summarize(n_neighbors_5_15 = sum(n_plants, na.rm = TRUE),
-            n_cells = n(),
-            density_5_15 = n_neighbors_5_15/n_cells,
-            shtsper_5_15 = sum(n_shts, na.rm = TRUE)/n())
-```
-So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2 or 3.66 shoots / 100m^2
-
-Then I can join that back to the original data set
-
-```{r}
-full_test <- full_join(ha_test, dist_5_15, by = c("row_number" = "row_from", "column" = "column_from"))
-head(full_test)
-```
-
-## Scaling up
-
-I'm going to scale up using a nested dataframe and the magic of list-columns!
-The advantage of this over splitting the one dataframe into a dataframe for every plot:year combo and using a for-loop or `map()` is that it keeps the plot and year structure intact and I don't have to figure out how to re-join the data at the end.
-
-First, nest the dataframe by plot and year.
-
-```{r}
-ha_nest <-
-  ha %>% 
-  group_by(plot, year) %>% 
-  group_nest()
-```
-
-Then, do the setup by calculating the `grid_coords` and `d` for every plot:year combo.
-
-```{r}
-ha_nest_dist <- 
-  ha_nest %>% 
-  #every combination of row and column:
-  mutate(grid_coords = map(
-    data, 
-    ~expand_grid(row = min(.x$row_number):max(.x$row_number),
-                 col = min(.x$column):max(.x$column)) %>% as.matrix()
-    )) %>% 
-  #matrix describing euclidean distance from every point to every other point:
-  mutate(d = map(
-    grid_coords,
-    ~{
-      d <- dist(.x * 10, #because each grid cell is 10m x 10m
-                diag = FALSE,
-                upper = TRUE) %>% 
-        as.matrix()
-       colnames(d) <- rownames(d) <- paste(.x[,1], .x[,2], sep = ",")
-      d
-    }
-    )) %>% 
-  #tidy data frame version of d
-  mutate(dist_df = map(d, ~{
-    as.data.frame(.x) %>% 
-      rownames_to_column(var = "from") %>% 
-      pivot_longer(-from, names_to = "to", values_to = "dist") %>% 
-      separate(from, into = c("row_from", "column_from")) %>% 
-      separate(to, into = c("row_to", "column_to")) %>% 
-      mutate(across(where(is.character), as.integer))
-  }
-  ))
-```
-
-Now, create that summary tibble counting the number of shoots in each grid cell for every plot:year combo, join, and unnest.
-
-```{r}
-ha_dist <-
-  ha_nest_dist %>% 
-  #summarize each year:plot to count number of plants in each grid cell
-  mutate(n_shts_sum = map(data, ~{
-    .x %>% 
-      # add_column(plant = 1) %>% 
-      group_by(row_number, column) %>% 
-      summarize(n_shts = sum(shts)) %>% 
-      rename(row = row_number)
-  }
-  )) %>% 
-  #join with tidied distance matrix to get dataframe for distance from every row
-  #and column to every other row and column
-  mutate(dist_full = map2(dist_df, n_shts_sum, ~{
-    full_join(.x, .y, by = c("row_to" = "row", "column_to" = "column"))
-  } 
-  )) %>% 
-  select(plot, year, dist_full) %>% 
-  unnest(dist_full)
-
-ha_dist
-```
-
-Now, with `ha_dist`, I can filter by distance, calculate a density, and the join to the original data.
-
-```{r}
-dens_0_10 <-
-  ha_dist %>% 
-  filter(dist>0 & dist <=10) %>% 
-  group_by(plot, year, row_from, column_from) %>% 
-  summarize(density_0_10 = sum(n_shts, na.rm = TRUE)/n())
-left_join(ha, dens_0_10, by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
-  arrange(plot, year, row_number, column)
-```
-
-I'll do this for different distances programmatically.
-
-```{r}
-width <- 10 #meters
-max <- 100 #meters
-L <- seq(0, max-width, by = width)
-U <- seq(0 + width, max, by = width)
-
-out <- 
-  map2(L, U, ~{
-  ha_dist %>%
-    filter(dist >= .x & dist < .y) %>% 
-    group_by(plot, year, row_from, column_from) %>%
-    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_shts, na.rm = TRUE)/n())
-})
-
-x <- out[[1]]
-for (i in 2:length(out)) {
-  x <- full_join(x, out[[i]], by = c("plot", "year",  "row_from", "column_from"))
-}  
-x
-```
-
-Join to original data
-
-```{r}
-ha_new <- 
-  left_join(ha, x,
-          by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
-  arrange(plot, year, row_number, column)
-ha_new
+y1p1 <- ha_xy %>% 
+  filter(year == first(year),
+         plot == first(plot))
+y1p1
 ```
@@
 ```{r}
-library(naniar)
-vis_miss(ha_new %>% select(starts_with("density")))
+add_density_cols <- function(df, L, U, by, verbose = FALSE) {
+
+  eu_dist <- function(p1, p2) {
+    sqrt(sum((p1 - p2)^2))
+  }
+
+  d <- seq(L, U, by = by)
+
+  df_out <- 
+    future_map_dfr(1:nrow(df), function(r) { #loop through rows/plants
+      #get coordinates for that plant
+      focal_plant <- t(df[r, c("x", "y")])
+
+      #calculate distance to every other plant
+      focal_df <-
+        df %>%
+        rowwise() %>%
+        mutate(dist = eu_dist(focal_plant, c(x, y)))
+
+      #get total number of shoots at every ring
+      future_map2_dfc(lag(d)[-1], d[-1],
+               ~ transmute(focal_df,
+                           !!glue("shts_{.x}_{.y}") := shts * (dist > .x & dist <= .y))) %>%
+        ungroup() %>% 
+        summarize(across(everything(), ~sum(., na.rm = TRUE)))
+    })
+
+  df_out <- bind_cols(df, df_out) #join to original data
+  return(df_out)
+}
+
+add_density_cols(y1p1, L = 0, U = 10, by = 5)
+
 ```
-The furthest distance I can estimate density for all plants is 60m.
-Past that, in order to estimate effects of density, I would have to drop plants.
-Even for closer plants though, the estimates may be biased because they do not include plants outside of the plot.
-That is, neighbor density for plants near the edges is based on a smaller sample than for plants near the center.
 
-# Write to cleaned data
+If I want this to be in units of shoots/m^2, then I need to calculate the area of the donut:
+
+$$
+A = \pi(U^2 - L^2)
+$$
+But also, like, truncate that donut if it passes outside of the plot.  I don't know how to do that.
 
 ```{r}
-write_rds(ha_new, here("analysis", "data", "derived_data", "ha_density.rds"))
+plan(multiprocess)
 ```
+
+## Apply to entire dataset
+
+```{r}
+ha_xy2 <-
+  ha_xy %>% 
+  group_by(plot, year) %>% 
+  group_split() %>% 
+  # head() %>% #test with 6 df to see how long it will take.
+  future_map(~{
+      add_density_cols(.x, L = 0, U = 10, by = 5) 
+    }) %>% 
+  bind_rows()
+beepr::beep(4)
+```
+```{r}
+ha_xy2
+```
+
+Yay! It worked! And it didn't even run that long!
+
+## Remove plants on edge
+
+I should remove plants on the edge because the entire "donut" isn't included for them.
+I can do this by finding the distance to the nearest edge, then checking if that distance is greater than the outer radius for the donuts, and if so, setting that density value to NA.
+
+To do this, I think I need to know how big the plots are.  Should have kept the plot_x and plot_y variables, maybe.
+
+```{r}
+
+```

### 2020-09-23 R/competition wrangling.Rmd
Set density for plants on edges to NA

@@
 I should remove plants on the edge because the entire "donut" isn't included for them.
-I can do this by finding the distance to the nearest edge, then checking if that distance is greater than the outer radius for the donuts, and if so, setting that density value to NA.
+I can do this by finding the distance to the nearest edge, then checking if that distance is greater than the outer radius for the donuts, and if so, setting that density value to `NA`.
 
-To do this, I think I need to know how big the plots are.  Should have kept the plot_x and plot_y variables, maybe.
+Plots are 10 columns by 5 rows so the x should range from 0--100 and the y should range from 0--50.
 
 ```{r}
+# example plant at 94, 45
+# min(c(94, 100 - 94, 45, 50 - 45))
 
+ha_density <-
+  ha_xy2 %>% 
+  rowwise() %>% 
+  mutate(
+    dist_to_edge = min(x, 100 - x, y, 50 - y),
+    shts_0_5 = ifelse(
+      dist_to_edge < 5,
+      NA,
+      shts_0_5
+    ),
+    shts_5_10 = ifelse(
+      dist_to_edge < 10,
+      NA,
+      shts_5_10
+    ))
+
+ha_density
 ```
+# Write to file
+
+```{r}
+write_rds(ha_density, here("analysis", "data", "derived_data", "ha_density.rds"))
+```

### 2020-09-23 R/competition wrangling.Rmd
added reminder that focal plant is excluded

@@
                            !!glue("shts_{.x}_{.y}") := shts * (dist > .x & dist <= .y))) %>% 
+        #excludes focal plant with dist == 0
         ungroup() %>% 
@@
 beepr::beep(4)
-```
-```{r}
 ha_xy2

### 2020-09-23 R/density dependence.Rmd
switched analysis to use coordinate based competition.

@@
 
-I don't know what a height of 0 means...
-Height of 0 probably means it was knocked over, or very small, or dead.
+Technically, the density data are only for 2006, 2007, and 2009, so I should really only use those three transition years.  This might still not be 100% correct.
+
+```{r}
+ha_sub_sub <-
+  ha_sub %>% 
+  filter(year %in% c(2006, 2007, 2009))
+```
 
@@
 ```{r}
-mpois <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
+mpois <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                 family = poisson(link = "log"),
@@
 
-mnorm <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year_post),
+mnorm <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                 family = gaussian(link = "log"),
@@
 ```{r}
-x<-check_heteroscedasticity(m_growth)
+x <- check_heteroscedasticity(m_growth)
 plot(x)
@@
 ```{r}
-m_growth2 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year_post),
+m_growth2 <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year_post),
                family = gaussian(link = "log"),
@@
 ```{r}
-m_growth3 <- glmer(shts_next ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
+m_growth3 <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number),
                family = gaussian(link = "log"),
@@
 Anova(m_growth)
-```
-
-Odd, density further out is significant, but density close is not.
-
-```{r}
 fixef(m_growth)
@@
 
-Both weakly positive density dependence
+Positive density dependence at 0--5 meters
+
+Same for 3-year subset?
+
+```{r}
+mgrowth_sub <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
+                family = gaussian(link = "log"),
+                data = ha_sub_sub)
+Anova(mgrowth_sub)
+```
+
+No longer significant.
 
@@
 ```{r}
-m_surv <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year), 
+m_surv <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year), 
                 family = binomial, data = ha_sub)
@@
 ```{r}
-m_surv2 <- glmer(surv ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year), 
+m_surv2 <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year), 
                 family = binomial, data = ha_sub)
@@
 ```{r}
-m_surv3 <- glm(surv ~ log_shts + habitat + density_0_10 + density_10_20, 
+m_surv3 <- glm(surv ~ log_shts + habitat + shts_0_5 + shts_5_10, 
                 family = binomial, data = ha_sub)
@@
 
+Does this hold for the 3-year subset?
+
+```{r}
+m_surv2_sub <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year), 
+                family = binomial, data = ha_sub_sub)
+Anova(m_surv2_sub)
+```
+
+No longer significant
+
 # Flowering probability
@@
   ha_sub %>% 
-  mutate(flwr = as.integer(infl>0), .before = infl)
+  mutate(flwr = as.integer(infl > 0), .before = infl)
 
@@
 ```{r}
-m_flwr <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number) + (1|year),
+m_flwr <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year),
                 family = binomial, data = ha_sub)
@@
 ```{r}
-m_flwr2 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|year),
+m_flwr2 <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year),
                  family = binomial, data = ha_sub)
@@
 ```{r}
-m_flwr3 <- glmer(flwr ~ log_shts + habitat + density_0_10 + density_10_20 + (1|ha_id_number),
+m_flwr3 <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number),
                  family = binomial, data = ha_sub)
@@
 
-Ok, there's an effect, but oddly only past 20m.  Again, positive density dependence.
+Ok, there's an effect, but oddly only at the further distance.  Again, positive density dependence.
 
@@
 
+No flowering data for 2006, 2007, and 2009.  These models won't run if the data are subset to only include these years.
+
 # Kernel approach
+
+**Not working yet**
 
@@
 
-```{r}
+```{r eval=FALSE}
 Q <-
@@
 
-```{r}
+```{r eval=FALSE}
 library(dlnm)
@@
 
+```{r eval=FALSE}
+g_surv <-
+  gam(surv ~ s(log_shts) +
+        habitat +
+        s(Q, L,
+          bs = "cb",
+          k = c(5, 4),
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = ha_sub,
+      method = "REML")
+```
+
 ## Predict
 
-```{r}
+```{r eval=FALSE}
+# x <- crosspred("Q", model = g_surv, cumul = TRUE)
 x <- crosspred("Q", model = g_growth, cumul = TRUE)
@@
 
-```{r}
+```{r eval=FALSE}
 plot.crosspred(x,
@@
 ```
-So this shows negative density dependence that seems to be strongest at about 40-50 m away from the focal plant.
+So this shows negative density dependence on growth that seems to be strongest at about 40-50 m away from the focal plant.
 
@@
 
-```{r}
+```{r eval=FALSE}
 plot.crosspred(x, ptype = "slice", var = 100, xlab = "distance (m)", main = "Effect of 100 shoots/100m^2")
@@
 
-```{r}
+```{r eval=FALSE}
 plot(x, ptype = "overall", xlab = "density (shoots/100m^2)", cumul = TRUE)
@@
 
-```{r}
+```{r eval=FALSE}
 #use arglag to override lag dimension of crossbasis, use argvar to override plant density dimension.
@@
 
-```{r}
+```{r eval=FALSE}
 x <- crosspred("Q", model = g_growth)
@@
 
-```{r}
+```{r eval=FALSE}
 anova(g_growth)

### 2020-09-23 R/competition wrangling.Rmd
filtered out plants near edges

--- 2020-09-23 R/competition wrangling.Rmd
@@
 ```{r}
-add_density_cols <- function(df, L, U, by, verbose = FALSE) {
+# df <- y1p1
+# L = 0
+# U = 10
+# by = 0.5
+add_density_cols <- function(df, L, U, by, na_edges = TRUE) {
 
@@
       focal_plant <- t(df[r, c("x", "y")])
+      # focal_plant <- c(0,0)
+
+      if (na_edges) {
+        #don't do calculations for rings that go outside the plot
+        edge_dist <- min(c(focal_plant, c(100, 50) - focal_plant))
+        d2 <- d[which(d < edge_dist)]
+      } else {
+        d2 <- d
+      }
 
@@
         rowwise() %>%
-        mutate(dist = eu_dist(focal_plant, c(x, y)))
+        mutate(dist = eu_dist(focal_plant, c(x, y))) %>% 
+        ungroup()
+
+      # If plant is near edge, return NAs
 
       #get total number of shoots at every ring
-      future_map2_dfc(lag(d)[-1], d[-1],
-               ~ transmute(focal_df,
-                           !!glue("shts_{.x}_{.y}") := shts * (dist > .x & dist <= .y))) %>%
-        ungroup() %>% 
-        summarize(across(everything(), ~sum(., na.rm = TRUE)))
+      if (length(d2) > 1) {
+        future_map2_dfc(lag(d2)[-1], d2[-1],
+                        ~ transmute(
+                          focal_df,
+                          !!glue("shts_{.x}_{.y}") :=
+                            shts * (dist > .x & dist <= .y)
+                        )
+        ) %>% 
+          #excludes focal plant with dist == 0
+          summarize(across(everything(), ~sum(., na.rm = TRUE)), .groups = "drop")
+      } else {
+        # if all the rings go outside the plot, then return an empty row
+        tibble_row()
+      }
     })
@@
   future_map(~{
-      add_density_cols(.x, L = 0, U = 10, by = 5) 
+      add_density_cols(.x, L = 0, U = 10, by = 0.5) 
     }) %>% 
@@
 ```
+
+Aggregate some of the distances for simpler modeling.
+
 ```{r}
-ha_xy2
+ha_xy3 <-
+  ha_xy2 %>% 
+  rowwise() %>%
+  mutate(shts_0_5 = sum(c_across(shts_0_0.5:shts_4.5_5)),
+         shts_5_10 = sum(c_across(shts_5_5.5:shts_9.5_10)))
+
+ha_density <- ha_xy3
 ```
@@
 
-## Remove plants on edge
-
-I should remove plants on the edge because the entire "donut" isn't included for them.
-I can do this by finding the distance to the nearest edge, then checking if that distance is greater than the outer radius for the donuts, and if so, setting that density value to NA.
-
-To do this, I think I need to know how big the plots are.  Should have kept the plot_x and plot_y variables, maybe.
+# Write to file
 
 ```{r}
-
+write_rds(ha_density, here("analysis", "data", "derived_data", "ha_density.rds"))
 ```

### 2020-09-24 R/competition wrangling.Rmd
divide by area of donuts

@@
 
-- filter out focal plants where donut goes out of plot. (filter(distance_to_edge > outer_circle_radius))
+- ~~filter out focal plants where donut goes out of plot. (filter(distance_to_edge > outer_circle_radius))~~
+- Do I need to divide by area of donut?
 - Using all years is not estimating density accurately.
@@
 # by = 0.5
-add_density_cols <- function(df, L, U, by, na_edges = TRUE) {
+add_density_cols <- function(df, L, U, by, na_edges = TRUE, per_area = TRUE) {
 
@@
           summarize(across(everything(), ~sum(., na.rm = TRUE)), .groups = "drop")
+
       } else {
@@
     })
+
+  if (per_area) {
+    # Divide by areas of donuts to get shoots/m^2
+    d3 <- d[1:(ncol(df_out) + 1)] #in case largest radii are outside of plot for every plant
+    A <- pi * (d3[-1]^2 - lag(d3)[-1]^2) #area of donut
+    df_out <- map2_dfc(df_out, A, ~.x/.y) #divide each column by each area
+  }
 
@@
 ```
-
-If I want this to be in units of shoots/m^2, then I need to calculate the area of the donut:
-
-$$
-A = \pi(U^2 - L^2)
-$$
-But also, like, truncate that donut if it passes outside of the plot.  I don't know how to do that.
 

### 2020-09-24 R/density dependence.Rmd
messing with gams

@@
 
-But how bad is this convergence failure?
-
-Is the max|grad| calculation correct?
-
-```{r}
-derivs1 <- mnorm@optinfo$derivs
-sc_grad1 <- with(derivs1, solve(Hessian, gradient))
-max(abs(sc_grad1))
-```
-No, it's actually a lot smaller than `tol`, so nothing to worry about
-
 ```{r}
@@
 
+Plant ID:
+
 ```{r}
-m_growth2 <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year_post),
-               family = gaussian(link = "log"),
-               data = ha_sub)
+m_growth2 <- update(m_growth, .~. - (1|ha_id_number))
 AICtab(m_growth, m_growth2)
@@
 
+Year:
+
 ```{r}
-m_growth3 <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number),
-               family = gaussian(link = "log"),
-               data = ha_sub)
+m_growth3 <- update(m_growth, .~. - (1|year_post))
 AICtab(m_growth, m_growth2, m_growth3)
 ```
+Can't remove year
 
@@
 ```{r}
-mgrowth_sub <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
-                family = gaussian(link = "log"),
-                data = ha_sub_sub)
+mgrowth_sub <- update(m_growth, data = ha_sub_sub)
 Anova(mgrowth_sub)
@@
 
+Plant ID:
+
 ```{r}
-m_surv2 <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year), 
-                family = binomial, data = ha_sub)
+m_surv2 <- update(m_surv, .~. - (1|ha_id_number))
 AICtab(m_surv, m_surv2)
@@
 Close to equivalent, so remove plant ID.  This makes sense, as plants only die once!
+
+Year:
 
@@
 
-Marginal effects on survival.
+Positive density dependence for survival.
 
@@
 ```{r}
-m_surv2_sub <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year), 
-                family = binomial, data = ha_sub_sub)
+m_surv2_sub <- update(m_surv2, data = ha_sub_sub)
 Anova(m_surv2_sub)
+fixef(m_surv2_sub)
 ```
 
-No longer significant
+Yes, nearly identical coefs.
 
@@
 ```{r}
-m_flwr <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year),
+m_flwr <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                 family = binomial, data = ha_sub)
@@
 
+Plant ID:
+
 ```{r}
-m_flwr2 <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|year),
-                 family = binomial, data = ha_sub)
+m_flwr2 <- update(m_flwr, .~. - (1|ha_id_number))
 AICtab(m_flwr, m_flwr2)
@@
 
+Year:
+
 ```{r}
-m_flwr3 <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number),
-                 family = binomial, data = ha_sub)
+m_flwr3 <- update(m_flwr, .~. - (1|year_post))
 AICtab(m_flwr, m_flwr2, m_flwr3)
@@
   ha_sub %>%
-  select(density_0_10:density_50_60) %>%
+  select(shts_0_0.5:shts_4.5_5) %>%
   as.matrix()
@@
 L <-
-  matrix(seq(5, 55, length.out = 6),
+  matrix(seq(0.25, 4.75, by = 0.5), #use midpoint of r1 and r2
          nrow = nrow(Q),
-         ncol = 6,
+         ncol = ncol(Q),
          byrow = TRUE)
@@
           bs = "cb",
-          k = c(5, 4),
+          # k = c(10, 10),
           xt = list(bs = "cr")),
@@
       method = "REML")
+gam.check(g_growth)
+anova(g_growth)
 ```
+
+Marginally significant effect of conspecific density on growth
 
@@
 g_surv <-
-  gam(surv ~ s(log_shts) +
+  gam(surv ~ s(log_shts, k = 12) +
         habitat +
@@
       method = "REML")
+gam.check(g_surv)
+anova(g_surv)
 ```
+No significant effect of density
 
@@
 ```{r eval=FALSE}
-# x <- crosspred("Q", model = g_surv, cumul = TRUE)
-x <- crosspred("Q", model = g_growth, cumul = TRUE)
-summary(x)
+hist(Q)
+pred_growth <- crosspred("Q", model = g_growth, cumul = TRUE, cen = mean(Q, na.rm = T), from = 0, to = 8)
+summary(pred_growth)
 ```
@@
 
-```{r eval=FALSE}
-plot.crosspred(x,
+```{r}
+plot.crosspred(pred_growth,
                ptype = "contour",
-               xlab = "#shoots/100m^2",
+               xlab = "#shoots/m^2",
                ylab = "distance (m)",
-               cumul = FALSE)
+               cumul = FALSE,
+               main = "Relative to mean competition")
 ```
-So this shows negative density dependence on growth that seems to be strongest at about 40-50 m away from the focal plant.
+Non-monotonic density dependence that decays steeply beyond 1m
 
@@
 ```{r eval=FALSE}
-plot.crosspred(x, ptype = "slice", var = 100, xlab = "distance (m)", main = "Effect of 100 shoots/100m^2")
-plot.crosspred(x, ptype = "slice", var = 10, xlab = "distance (m)", main = "Effect of 10 shoots/100m^2")
+plot.crosspred(pred_growth, ptype = "slice", var = 0.2, xlab = "distance (m)", main = "0.2 shoots/m^2", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", var = 1, xlab = "distance (m)", main = "1 shoots/m^2", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", var = 2, xlab = "distance (m)", main = "2 shoots/m^2", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", var = 8, xlab = "distance (m)", main = "8 shoots/m^2", ylim = c(0.9,1.1))
 ```
 
-### Cumulative
-
-I think this shows the cumulative effect of density summed/integrated over all distances.
-So, for example, a plant with about 30 shoots/100m^2 surrounding it out to the maximum distance (60m) would have the largest number of shoots, on average.
-
-```{r eval=FALSE}
-plot(x, ptype = "overall", xlab = "density (shoots/100m^2)", cumul = TRUE)
+```{r}
+plot.crosspred(pred_growth, ptype = "slice", lag = 0.25, xlab = "density (shoots/m^2)", main = "0.25m", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", lag = 1.25, xlab = "density (shoots/m^2)", main = "1.25m", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", lag = 3.25, xlab = "density (shoots/m^2)", main = "3.25m", ylim = c(0.9,1.1))
+plot.crosspred(pred_growth, ptype = "slice", lag = 4.25, xlab = "density (shoots/m^2)", main = "4.25m", ylim = c(0.9,1.1))
 ```
-
-There is no built-in function to get cumulative effects of density over all distances.
 

### 2020-09-24 R/density dependence.Rmd
incorporated random effects into GAMs.

@@
 ```{r}
-m_surv <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year), 
+m_surv <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post), 
                 family = binomial, data = ha_sub)
@@
           # k = c(10, 10),
-          xt = list(bs = "cr")),
+          xt = list(bs = "cr")) +
+        s(year_post, bs = "re"),
       family = gaussian(link = "log"),
@@
 g_surv <-
-  gam(surv ~ s(log_shts, k = 12) +
+  gam(surv ~ s(log_shts) +
         habitat +
@@
           # k = c(10, 10),
-          xt = list(bs = "cr")),
+          xt = list(bs = "cr")) +
+        s(year_post, bs = "re"),
       family = binomial,
@@
 ```
-No significant effect of density
+Marginally significant effect of density on survival
 
@@
 
+Growth:
 ```{r eval=FALSE}
@@
 pred_growth <- crosspred("Q", model = g_growth, cumul = TRUE, cen = mean(Q, na.rm = T), from = 0, to = 8)
-summary(pred_growth)
+pred_surv <- crosspred("Q", model = g_surv, cumul = TRUE, cen = mean(Q, na.rm = T), from = 0, to = 8)
+# 
+# summary(pred_growth)
 ```
 
-## Plot crossbasis
-
-### Contour plot
+## Plot Growth
 
@@
 ```
+
+## Plot Survival
+
+```{r}
+plot.crosspred(pred_surv,
+               ptype = "contour",
+               xlab = "#shoots/m^2",
+               ylab = "distance (m)",
+               cumul = FALSE,
+               main = "Relative to mean competition")
+```
+Postive density dependence that drops off steeply after 1m
 

### 2020-09-30 R/competition wrangling.Rmd
Calculates density separately for 3 years with data.  Closes #4

@@
 
-# Purpose
+# TODO:
+ - move main function to .R file and document
 
-# TODO:
+# Summary
 
-- ~~filter out focal plants where donut goes out of plot. (filter(distance_to_edge > outer_circle_radius))~~
-- Do I need to divide by area of donut?
-- Using all years is not estimating density accurately.
-Focal plants are not moving, but in other years there are neighbors without known coordinates.
-Solution: only use the three transitions where we know the coords of every plant?
-Or assume births and deaths don't change density significantly?
+In order to test for density dependence, we used data from three surveys where exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded.
+THe 2009 survey was to assess the effects of safe-site limitation on recruitment (Bruna et al.
+2010) and the 2006 and 2007 surveys were for ____ (unpublished data).
+For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al.
+(date).
+Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e.
+a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m.
+If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value.
+Conspecific density was estimated as the number of *H.
+acuminata* shoots per m^2^.
+We used the three transition years for which we had density data to model the effects of conspecific density on survival, growth, and reproduction of *H.
+acuminata* (in a separate .Rmd file).
 
@@
 
-coords <-
+coords_0607 <-
   read.table(here("analysis", "data", "raw_data", "adults0607_light.txt")) %>%
@@
 I need to join these data sources and offset the coordinates in the `ha` dataset correctly.
-I'm going to remove `year` from the `coords` dataset and assume that plants aren't moving.
 
@@
 ```{r}
-coords2 <- 
-  coords %>% 
-  select(plot = site, tag_number, row, column, plot_x, plot_y, x, y) %>% 
+coords_0607 <- 
+  coords_0607 %>% 
+  select(year = yr, plot = site, tag_number, row, column, plot_x, plot_y, x, y) %>% 
   mutate(plot = case_when(
@@
   )) 
-```
-
-Summarize so there is only one set of coordinates per plant. Averaging as I'm assuming any difference in coordinates between '06 and '07 is not because the plant moved.
-
-```{r}
-coords3 <-
-  coords2 %>%
-  group_by(plot, tag_number) %>% 
-  # rowwise() %>% 
-  summarize(across(where(is.character), first),
-            across(where(is.numeric), mean)) %>% 
-  rename(x_07 = x,
-         y_07 = y)
+# filter(coords_0607, is.na(x))
 ```
@@
 ```{r}
-ha_coords <- left_join(ha, coords3, by = c("plot", "row", "column", "tag_number"))
+coords_09 <-
+  ha %>% 
+  filter(year == 2009, !is.na(x_09), !is.na(y_09))
+# filter(coords_09, is.na(x_09))
 ```
-# Tidy
 
+Adjust x and y coords to be for entire plot and not just within row/col cell
+
+```{r}
+coord_adj <-
+  bind_rows(
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 5:1),
+    tibble(column = 1:10,
+           plot_x = 1:10)
+  ) %>% add_column(plot = "2107"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 10:1)
+  ) %>% add_column(plot = "2108"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 1:10)
+  ) %>% add_column(plot = "Dimona-CF"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 5:1),
+    tibble(column = 1:10,
+           plot_x = 1:10)
+  ) %>% add_column(plot = "5753"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 10:1)
+  ) %>% add_column(plot = "PortoAlegre-CF"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 1:10)
+  ) %>% add_column(plot = "CaboFrio-CF"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 10:1)
+  ) %>% add_column(plot = "5751"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 10:1)
+  ) %>% add_column(plot = "Florestal-CF"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:10],
+           plot_y = 10:1),
+    tibble(column = 6:10,
+           plot_x = 5:1)
+  ) %>% add_column(plot = "5750"),
+
+  expand_grid(
+    tibble(row = LETTERS[1:5],
+           plot_y = 1:5),
+    tibble(column = 1:10,
+           plot_x = 1:10)
+  ) %>% add_column(plot = "5756")
+)
+```
+
+Get the `plot_x` and `plot_y` variables into the 2009 data to adjust the coordinates correctly.
+
+```{r}
+coords_09.2 <- left_join(coords_09, coord_adj, by = c("plot", "row", "column"))
+```
+
+There are still plants with row/column combinations that I wouldn't expect:
+
+```{r}
+coords_09.2 %>% filter(is.na(plot_x)) %>% 
+  select(plot, row, column, ha_id_number)
+```
 Now, use the plot_x and plot_y variables to correctly offset the x and y coords from 2009.
@@
 ```{r}
-ha_coords2 <- 
-  ha_coords %>% 
+coords_09.3 <- 
+  coords_09.2 %>% 
   mutate(x_adj = x_09 + ((plot_x - 1) * 10),
@@
 
-Now, I'll just average across the three years and assume that any differences are measurement error and not because plants moved.
+# Bind rows to combine data
 
 ```{r}
-ha_coords3 <-
-  ha_coords2 %>% 
-  rowwise() %>% 
-  mutate(x = mean(c(x_adj, x_07), na.rm = TRUE),
-         y = mean(c(y_adj, y_07), na.rm = TRUE)) %>% 
-  #replace NaN's generated when mean of only NAs is taken
-  mutate(across(c(x, y), ~ifelse(is.nan(.), NA, .))) %>% 
-  #remove intermediate coordinates
-  select(ranch, bdffp_reserve_no, plot, habitat,
-         ha_id_number, tag_number, row, column,
-         x, y, year, ht, ht_next, shts, shts_next,
-         infl, surv, code_notes, code2)
+ha_coords <-
+  bind_rows(
+  coords_09.3 %>% 
+  select(year, plot, tag_number, row, column, x = x_adj, y = y_adj),
+  coords_0607 %>% select(-plot_x, -plot_y)
+)
+ha_coords
 ```
 
-# Filter
+# Join to demography data
 
@@
 
-**NOTE:** There is a **problem** with this.  By using all the years and filtering the data to just include plants with coordinates from 2006, 2007, and 2009, I am not estimating density accurately.
 ```{r}
-ha_xy <-
-  ha_coords3 %>% 
-  filter(!is.na(x), !is.na(y)) %>% 
-  ungroup()
+ha_xy <- right_join(ha %>% select(-x_09, -y_09), ha_coords, by = c("plot", "row", "column", "tag_number", "year"))
 ```
@@
 
-add_density_cols(y1p1, L = 0, U = 10, by = 5)
+add_density_cols(y1p1, L = 0, U = 5, by = 0.5)
 
@@
 ```{r}
-ha_xy2 <-
+ha_density <-
   ha_xy %>% 
@@
   future_map(~{
-      add_density_cols(.x, L = 0, U = 10, by = 0.5) 
+      add_density_cols(.x, L = 0, U = 5, by = 0.5) 
     }) %>% 
@@
 ```
-
-Aggregate some of the distances for simpler modeling.
-
-```{r}
-ha_xy3 <-
-  ha_xy2 %>% 
-  rowwise() %>%
-  mutate(shts_0_5 = sum(c_across(shts_0_0.5:shts_4.5_5)),
-         shts_5_10 = sum(c_across(shts_5_5.5:shts_9.5_10)))
-
-ha_density <- ha_xy3
-```
-
-Yay! It worked! And it didn't even run that long!
 

### 2020-09-30 R/utils.R
generalized function for marginal effects of cross-basis smooths.

@@ -138,53 +138,71 @@ mat_diamond <- function(D, nrow, ncol) {
 }
 
 # WIP: function to get data for plotting results from GAM with crossbasis smooth.
-# 
-# library(rlang)
-# pred_cb <- function(Q, L, model) {
-#   Q_name <- enquo(Q)
-#   L_name <- enquo(L)
-#   df <- model$model
-#   
-#   testvals <- seq(min(Q), max(Q), length.out = 200)
-#   Q_new <- matrix(mean(Q), nrow = length(testvals), ncol = lag)
-#   L_new <- matrix(1:ncol(L), nrow = nrow(Q_new), ncol = ncol(L), byrow = TRUE)
-#   
-#   #keep all other numeric terms constant and make new levels for factors
-#   #TODO: Figure out how to predict marginal effect with fixed effects.
-#   newdata <-
-#     df %>% 
-#     summarize(
-#       across(c(-!!L_name, -!!Q_name) & where(is.numeric), mean),
-#       across(where(is.factor), ~factor(".newdata"))
-#       # across(where(is.factor), ~factor("1-ha"))
-#     ) 
-#   newdata <- uncount(newdata, 200) %>% add_column(!!L_name := L_new)
-#   
-#   resp <- array(dim = c(length(testvals), ncol(Q_new)))
-#   rownames(resp) <- testvals
-#   
-#   #loop through columns of matrix, replace with testvals, predict fitted.
-#   for(i in 1:ncol(Q_new)) {
-#     P1_i <- Q_new
-#     P1_i[, i] <- testvals
-#     resp[, i] <-
-#       suppressWarnings( #new levels of random effects are on purpose
-#         predict(
-#           model,
-#           newdata = newdata %>% add_column(!!Q_name := P1_i),
-#           type = "response"
-#         )
-#       )
-#   }
-#   out <-
-#     resp %>%
-#     as_tibble(rownames = "x") %>%
-#     pivot_longer(
-#       cols = starts_with("V"),
-#       names_to = "lag",
-#       names_prefix = "V",
-#       values_to = "fitted"
-#     ) %>%
-#     mutate(lag = as.double(lag), x = as.double(x)) 
-#   return(out)
-# }
+library(rlang)
+pred_cb <- 
+  function(Q, L, model) {
+  # Q_name <- quo(Q)
+  # L_name <- quo(L)
+  Q_name <- enquo(Q)
+  L_name <- enquo(L)
+  df <- model$model
+
[…]

### 2020-09-30 R/density dependence.Rmd
simplified density dependence analysis to only use 3 years of data.  Corrected calculation of density, re-ran models.

--- 2020-09-24 R/density dependence.Rmd
@@
 
+# TODO
+
+- Get confidence intervals for cross-basis predictions.  Plot slices with confidence bands.
+
 # Purpose
@@
 ```
+This is the main *Heliconia* dataset, but with densities of conspecifics (shoots/m^2^) at different rings with inner and outer radii in meters.
+So, for example, the `shts_8_8.5` column is the number of *Heliconia* shoots/m^2^ in a ring 8 to 8.5 m around the focal plant.
+
+The `x` and `y` coordinates are averages of data from 2006, 2007, and 2009, but joined to the full range of the dataset--assuming that plants don't move.
+
+## Tidying and subsetting
 
@@
   ha %>%
-  group_by(ha_id_number) %>% 
+  ungroup() %>% 
   arrange(ha_id_number, year) %>% 
-  ungroup() %>% 
-  #scale year
-  mutate(year_post = year - min(year))
+  #scale year, and convert to factor since I'll be using as random effect
+  mutate(year_post = as_factor(year - min(year)))
 ```
 
-Filter to just continuous forest
-
-```{r}
-ha_cf <- 
-  ha %>% 
-  filter(habitat == "CF") %>% 
-  mutate(log_shts = log(shts),
-         log_shts_next = log(shts_next))
-```
-
-Or filter to just 1-ha and cf
+Use just 1-ha and cf
 
@@
          log_shts_next = log(shts_next),
-         habitat = as.factor(habitat))
-```
-
-Technically, the density data are only for 2006, 2007, and 2009, so I should really only use those three transition years.  This might still not be 100% correct.
-
-```{r}
-ha_sub_sub <-
-  ha_sub %>% 
-  filter(year %in% c(2006, 2007, 2009))
+         habitat = as.factor(habitat),
+         ha_id_number = as.factor(ha_id_number))
 ```
@@
 For each vital rate I'll include the number of shoots and habitat as fixed effects and plant ID and year as random effects.
-I'm not including plot since there are few plots in each habitat type, and preliminary modeling shows the variance for plot is the lowest of those three random effects.
+For the mixed effects models I'll use density at just two rings (0--5m and 5--10m) as fixed effects.
+There are probably too few plots (4 1-ha and 6CF) to use plot as a random effect, but I don't know how to code it as a fixed effect right now.
+
+```
+vital_rate ~ log(shts) + habitat + shts_0_2.5 + shts_2.5_5 + (1|year) + (1|ha_id_number) + (1|habitat:plot)
+```
 
@@
 ```{r}
-mpois <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
+mpois <-
+  glmer(
+    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
     family = poisson(link = "log"),
-                data = ha_sub)
+    data = ha_sub
+  )
 
@@
 
-mnorm <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
+mnorm <-
+  glmer(
+    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
     family = gaussian(link = "log"),
-                data = ha_sub)
+    data = ha_sub
+  )
 ```
@@
 Both have problems with heteroscedasticity.
+A log-transformation of the response doesn't fix the heteroscedatsticyt either.
 I'll go with log-normal.
@@
 
+Plot:
+
+```{r}
+m_growth2 <- update(m_growth, .~. - (1|habitat:plot))
+AICtab(m_growth, m_growth2)
+```
+
 Plant ID:
@@
 ```{r}
-m_growth2 <- update(m_growth, .~. - (1|ha_id_number))
-AICtab(m_growth, m_growth2)
+m_growth3 <- update(m_growth2, .~. - (1|ha_id_number))
+AICtab(m_growth2, m_growth3)
 ```
@@
 ```{r}
-m_growth3 <- update(m_growth, .~. - (1|year_post))
-AICtab(m_growth, m_growth2, m_growth3)
+m_growth4 <- update(m_growth2, .~. - (1|year_post))
+AICtab(m_growth, m_growth2, m_growth3, m_growth4)
 ```
-Can't remove year
+Can remove year
 
@@
 ```{r}
-Anova(m_growth)
-fixef(m_growth)
+Anova(m_growth4)
+fixef(m_growth4)
 ```
 
-Positive density dependence at 0--5 meters
-
-Same for 3-year subset?
-
-```{r}
-mgrowth_sub <- update(m_growth, data = ha_sub_sub)
-Anova(mgrowth_sub)
-```
-
-No longer significant.
+No density dependence
 
@@
 ```{r}
-m_surv <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year), 
-                family = binomial, data = ha_sub)
+m_surv <-
+  glmer(
+    surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) + (1|year_post) + (1|habitat:plot), 
+    family = binomial, data = ha_sub
+  )
 ```
@@
 
+Plot:
+
+```{r}
+m_surv2 <- update(m_surv, .~. - (1|habitat:plot))
+AICtab(m_surv, m_surv2)
+```
+
 Plant ID:
@@
 ```{r}
-m_surv2 <- update(m_surv, .~. - (1|ha_id_number))
-AICtab(m_surv, m_surv2)
+m_surv3 <- update(m_surv2, .~. - (1|ha_id_number))
+AICtab(m_surv2, m_surv3)
 ```
@@
 ```{r}
-m_surv3 <- glm(surv ~ log_shts + habitat + shts_0_5 + shts_5_10, 
+m_surv4 <- glm(surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5, 
                 family = binomial, data = ha_sub)
-AICtab(m_surv2, m_surv3)
+AICtab(m_surv4, m_surv3)
 ```
-Can't remove year.
+glm is best
 
@@
 ```{r}
-Anova(m_surv2)
-fixef(m_surv2)
+Anova(m_surv4)
+coef(m_surv4)
 ```
@@
 
-Does this hold for the 3-year subset?
-
-```{r}
-m_surv2_sub <- update(m_surv2, data = ha_sub_sub)
-Anova(m_surv2_sub)
-fixef(m_surv2_sub)
-```
-
-Yes, nearly identical coefs.
-
 # Flowering probability
 
-```{r}
-ggplot(ha_sub, aes(infl)) + geom_histogram()
-```
 Make a binary column
@@
 
-#the way suvival is currently coded, a plant can flower and then not survive in the same year, so no need to filter out dead plants.
-ha_sub %>% filter(surv == 0 & flwr == 1)
+unique(ha_sub$flwr)
 ```
 
-```{r}
-m_flwr <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
-                family = binomial, data = ha_sub)
+Can't do this model because no variation in flowering.
+
+```{r eval=FALSE}
+m_flwr <-
+  glmer(
+    flwr ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
+    family = binomial, data = ha_sub
+  )
 ```
-
-## Random effects
-
-Plant ID:
-
-```{r}
-m_flwr2 <- update(m_flwr, .~. - (1|ha_id_number))
-AICtab(m_flwr, m_flwr2)
-```
-Can't remove plant ID.
-
-Year:
-
-```{r}
-m_flwr3 <- update(m_flwr, .~. - (1|year_post))
-AICtab(m_flwr, m_flwr2, m_flwr3)
-```
-
-Can't remove year
-
-## Density dependence?
-
-```{r}
-Anova(m_flwr)
-fixef(m_flwr)
-```
-
-Ok, there's an effect, but oddly only at the further distance.  Again, positive density dependence.
-
-Probability of flowering increases with size and is lower in continuous forest
-
-No flowering data for 2006, 2007, and 2009.  These models won't run if the data are subset to only include these years.
 
@@
 
-**Not working yet**
-
-So it seems like there are effect of density at least out to 20m.
-I'd like to take the approach in Teller et al.
-2016, but the code is so convoluted that it is not adaptable.
-I want to start by trying to adapt the `dlnm` approach instead.
+I'd like to take the approach in Teller et al. 2016, but the code is so convoluted that it is not adaptable. Instead, I'll start by trying to adapt the `dlnm` approach.
 
@@
 
-```{r eval=FALSE}
+```{r}
 Q <-
@@
 
-## GAM
-
-```{r eval=FALSE}
+```{r}
 library(dlnm)
 library(mgcv)
+library(visibly) # better diagnostic visulazations.  one of several packages for this.
+source(here("R", "utils.R"))
+```
 
-# Q_log <- log(Q+1)
+## Growth
+
+```{r}
 g_growth <-
-  gam(shts_next ~ s(log_shts) +
+  gam(shts_next ~ 
+        s(log_shts) + #Essentially linear, could replace with parametric
         habitat +
@@
           bs = "cb",
-          # k = c(10, 10),
-          xt = list(bs = "cr")),
+          xt = list(bs = "cr")), #+
+        # s(ha_id_number, bs = "re"), #this takes waaay too long to converge
       family = gaussian(link = "log"),
@@
       method = "REML")
-gam.check(g_growth)
+```
+
+```{r}
+plot_gam_check(g_growth)
+```
+
+```{r}
 anova(g_growth)
@@
 
-Marginally significant effect of conspecific density on growth
+Effect of size on size(t+1):
 
-```{r eval=FALSE}
+```{r}
+plot(g_growth, pages = 1)
+```
+
+Effects of density:
+
+```{r}
+pred_growth <- pred_cb(Q, L, g_growth)
+ggplot(pred_growth, aes(x = x, y = lag, z = fitted, fill = fitted)) +
+  geom_raster(interpolate = TRUE) +
+  geom_contour(color = "grey50", size = 0.4) + 
+  scale_y_continuous("distance (m)", expand = c(0,0)) +
+  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
+  scale_fill_viridis_c("#shoots (t+1)", option = "A")
+```
+
+## Survival
+
+```{r}
 g_surv <-
-  gam(surv ~ s(log_shts, k = 12) +
+  gam(surv ~
+        s(log_shts) +
         habitat +
@@
       method = "REML")
-gam.check(g_surv)
-anova(g_surv)
-```
-No significant effect of density
-
-## Predict
-
-```{r eval=FALSE}
-hist(Q)
-pred_growth <- crosspred("Q", model = g_growth, cumul = TRUE, cen = mean(Q, na.rm = T), from = 0, to = 8)
-summary(pred_growth)
-```
-
-## Plot crossbasis
-
-### Contour plot
-
-```{r}
-plot.crosspred(pred_growth,
-               ptype = "contour",
-               xlab = "#shoots/m^2",
-               ylab = "distance (m)",
-               cumul = FALSE,
-               main = "Relative to mean competition")
-```
-Non-monotonic density dependence that decays steeply beyond 1m
-
-### Slices
-
-We can also plot slices through the contour plot at different distances or different shoot densities
-
-```{r eval=FALSE}
-plot.crosspred(pred_growth, ptype = "slice", var = 0.2, xlab = "distance (m)", main = "0.2 shoots/m^2", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", var = 1, xlab = "distance (m)", main = "1 shoots/m^2", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", var = 2, xlab = "distance (m)", main = "2 shoots/m^2", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", var = 8, xlab = "distance (m)", main = "8 shoots/m^2", ylim = c(0.9,1.1))
 ```
@@
 ```{r}
-plot.crosspred(pred_growth, ptype = "slice", lag = 0.25, xlab = "density (shoots/m^2)", main = "0.25m", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", lag = 1.25, xlab = "density (shoots/m^2)", main = "1.25m", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", lag = 3.25, xlab = "density (shoots/m^2)", main = "3.25m", ylim = c(0.9,1.1))
-plot.crosspred(pred_growth, ptype = "slice", lag = 4.25, xlab = "density (shoots/m^2)", main = "4.25m", ylim = c(0.9,1.1))
+# plot_gam_check(g_surv)
+anova(g_surv)
 ```
 
-## Other shapes
+Effect of size on survival:
 
-It is odd that the strength of competition seems to be increasing with distance.
-I'd like to be able to force the distance dimension to be monotonically decreasing as was done in Teller et al (I think?).
-Teller does this *after* fitting a spline using `mk_dist_wts()`, which is a function that takes the spline coefficients, and predicted values from the spline, and transforms it into a non-negative, monotonic decreasing function.
-
-You can override one of the dimensions of the cross basis with any custom function.
-
-```{r eval=FALSE}
-#use arglag to override lag dimension of crossbasis, use argvar to override plant density dimension.
-lin <- dlnm:::lin
-xt <- list(bs = "cr", arglag = list(fun = "lin"))
-
-g_growth <- 
-  gam(shts_next ~ s(log_shts) +
-        habitat +
-        s(Q, L,
-          bs = "cb",
-          k = c(5, 4),
-          fx = c(F, T), #turns off penalization for the lag dimension
-          xt = xt),
-      family = gaussian(link = "log"),
-      data = ha_sub,
-      gamma = 1.5)
-
-gam.check(g_growth)
+```{r}
+plot(g_surv, pages = 1)
 ```
 
-```{r eval=FALSE}
-x <- crosspred("Q", model = g_growth)
-plot.crosspred(x, ptype = "contour", xlab = "#shoots/100m^2", ylab = "distance (m)")
-
-plot.crosspred(x, ptype = "slice", var = 100)
-plot.crosspred(x, ptype = "slice", var = 10)
+```{r}
+pred_surv <- pred_cb(Q, L, g_surv)
+ggplot(pred_surv, aes(x = x, y = lag, z = fitted, fill = fitted)) +
+  geom_raster(interpolate = TRUE) +
+  geom_contour(color = "grey50", size = 0.4) + 
+  scale_y_continuous("distance (m)", expand = c(0,0)) +
+  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
+  scale_fill_viridis_c("P(survival)", option = "A")
 ```
-Now the kernel is the same shape at every value of density.  However, this relationship doesn't make any sense.  Fewer knots?  More penalty?  Custom function?
-
-```{r eval=FALSE}
-anova(g_growth)
-```
-
-# TODO:
-
-- ~~is density at 0 distance including the focal plant?  It shouldn't.~~
-- Check number of inflorescences given flowered?
-- ~~log-transform density (nah, not right now.  DOesn't help with convergence)~~
-- ~~check that survival is coded right so that it is a function of `ht_prev`~~
-- try negative binomial distribution for shoot number
-- ~~Density should be total shoot number~~
-- ~~Should actually be density in previous year~~
-- ~~Shoots should be poisson distribution, not log-transformed?~~
-- ~~incorporate 1-ha fragments and use habitat type as a predictor.~~
-- Implement FLM approach to density dependence.

### 2020-10-05 R/competition wrangling.Rmd
added some combined annuli because it was easier than figuring out how to do a weighted average after the fact.
added se to marginal effects function.  Moved euclidean distance functiuon.

@@
 
-```{r}
-# function to calculate euclidean distance between two points
-eu_dist <- function(p1, p2) {
-  sqrt(sum((p1 - p2)^2))
-}
-# x = c(0,0)
-# y = c(3,3)
-# eu_dist(x, y)
-```
-
 ## Proof of concept
@@
   }) %>% 
+  future_map(~{
+    add_density_cols(.x, L = 0, U = 5, by = 2.5)
+  }) %>% 
   bind_rows()

### 2020-10-05 R/utils.R

@@ -139,10 +139,29 @@ mat_diamond <- function(D, nrow, ncol) {
 
 # WIP: function to get data for plotting results from GAM with crossbasis smooth.
 library(rlang)
+#TODO: change name of function to something that indicates marginal effects.
+#' Title
+#'
+#' @param Q 
+#' @param L 
+#' @param model 
+#'
+#' @return
+#' @export
+#' @import rlang
+#' @import purrr
+#' @import dplyr
+#' @import mgcv
+#'
+#' @examples
 pred_cb <- 
   function(Q, L, model) {
   # Q_name <- quo(Q)
   # L_name <- quo(L)
+    
+  if (!inherits(model, "gam")) {
+    abort("This is only for GAMs made with the `mgcv` package including cross-basis smooths from the `dlnm` package.")
+  } 
   Q_name <- enquo(Q)
   L_name <- enquo(L)
   df <- model$model
@@ -158,11 +177,11 @@ pred_cb <-
   # Set parametric factors to reference level.
   terms_raneff <-
     model$smooth %>% 
-    map_if(~inherits(.x, "random.effect"),
-           ~pluck(.x, "term"),
-           .else = function(x) return(NULL)) %>%
-    compact() %>% 
-    as_vector()
+    purrr::map_if(~inherits(.x, "random.effect"),
+                  ~pluck(.x, "term"),
+                  .else = function(x) return(NULL)) %>%
+    purrr::compact() %>% 
+    purrr::as_vector()
   
   terms_fac <- names(model$xlevels)
   
@@ -170,7 +189,7 @@ pred_cb <-
   #breaks if there is a fixed-effect factor put in as a character vector.
   newdata <-
     df %>%
-    summarize(
+    dplyr::summarize(
       across(c(-!!L_name, -!!Q_name) & where(is.numeric), mean),
       across(all_of(terms_raneff) & where(is.factor), ~factor(".newdata")),
       across(all_of(terms_fac) & where(is.factor), ~factor(levels(.x)[1], levels = levels(.x)))
@@ -180,22 +199,24 @@ pred_cb <-
   resp <- array(dim = c(length(testvals), ncol(Q_new)))
   rownames(resp) <- testvals
   colnames(resp) <- lvals
-
+  se <- resp
   #loop through columns of matrix representing different lags/distances, replace
   #with testvals, predict response.
   for (i in 1:ncol(Q_new)) {
     P1_i <- Q_new
[…]

### 2020-10-05 R/utils.R
documented functions better

@@ -152,6 +152,7 @@ library(rlang)
 #' @import purrr
 #' @import dplyr
 #' @import mgcv
+#' @import tidyr
 #'
 #' @examples
 pred_cb <- 
@@ -162,8 +163,8 @@ pred_cb <-
   if (!inherits(model, "gam")) {
     abort("This is only for GAMs made with the `mgcv` package including cross-basis smooths from the `dlnm` package.")
   } 
-  Q_name <- enquo(Q)
-  L_name <- enquo(L)
+  Q_name <- rlang::enquo(Q)
+  L_name <- rlang::enquo(L)
   df <- model$model
 
   testvals <- seq(min(Q, na.rm = TRUE), max(Q, na.rm = TRUE), length.out = 200)
@@ -218,23 +219,23 @@ pred_cb <-
   }
   fitted <-
     resp %>%
-    as_tibble(rownames = "x", .name_repair = "unique") %>%
-    pivot_longer(
+    dplyr::as_tibble(rownames = "x", .name_repair = "unique") %>%
+    tidyr::pivot_longer(
       cols = -x,
       names_to = "lag",
       values_to = "fitted"
     ) %>%
-    mutate(lag = as.double(lag), x = as.double(x))
+    dplyr::mutate(lag = as.double(lag), x = as.double(x))
   
   se.fitted <-
     se %>%
-    as_tibble(rownames = "x", .name_repair = "unique") %>%
-    pivot_longer(
+    dplyr::as_tibble(rownames = "x", .name_repair = "unique") %>%
+    tidyr::pivot_longer(
       cols = -x,
       names_to = "lag",
       values_to = "se.fit"
     ) %>%
-    mutate(lag = as.double(lag), x = as.double(x))
+    dplyr::mutate(lag = as.double(lag), x = as.double(x))
   
   out <- full_join(fitted, se.fitted, by = c("x", "lag"))
   

### 2020-10-05 R/density dependence.Rmd
prototype of excluding predicted values too far from data

@@
   #scale year, and convert to factor since I'll be using as random effect
-  mutate(year_post = as_factor(year - min(year)))
+  mutate(year_post = as_factor(year - min(year)),
+         log_shts = log(shts),
+         log_shts_next = log(shts_next),
+         habitat = as.factor(habitat),
+         ha_id_number = as.factor(ha_id_number),
+         plot = as.factor(plot))
 ```
 
-Use just 1-ha and cf
+Emilio suggested starting with a couple of small subsets of data to make sure I'm getting things right.
+
+1. Just continuous forest
+2. Just a single plot (e.g. 5750)
+3. Just plants with the median # of shoots (e.g. 3)
 
 ```{r}
-ha_sub <-
-  ha %>% 
-  filter(habitat %in% c("CF", "1-ha")) %>% 
-  mutate(log_shts = log(shts),
-         log_shts_next = log(shts_next),
-         habitat = as.factor(habitat),
-         ha_id_number = as.factor(ha_id_number))
+# ha_sub <-
+#   ha %>%
+#   filter(habitat %in% c("CF", "1-ha"))
+
+# ha_sub <-
+#   ha %>%
+#   filter(habitat == "CF")
+
+# ha_sub <-
+#   ha %>%
+#   filter(plot == 5750)
+
+# ha_sub <-
+#   ha %>%
+#   filter(shts == median(shts, na.rm = TRUE))
+
+ha_sub <- ha
 ```
+
+```{r}
+ha_sub %>% 
+  group_by(year, plot) %>% 
+  summarize(n_distinct(row), n_distinct(column))
+```
+There's something funky going on in 2009 with row and column positions. I opened an issue and this will get fixed eventually.
 
@@
   glmer(
-    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
-      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
+    shts_next ~
+      log_shts + 
+      habitat +
+      shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) +
+      (1|plot) +
+      (1|year_post),
     family = poisson(link = "log"),
@@
   glmer(
-    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
-      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
+    shts_next ~ 
+      log_shts +
+      habitat +
+      shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) +
+      (1|plot) +
+      (1|year_post),
     family = gaussian(link = "log"),
@@
 ```{r}
-m_growth2 <- update(m_growth, .~. - (1|habitat:plot))
+m_growth2 <- update(m_growth, .~. - (1|plot))
 AICtab(m_growth, m_growth2)
@@
 
-No density dependence
+- Just 5750: marginaly significant negative density dependence at 2.5--5m
+- Only CF: no significant effect of density dependence
+- Only 3-shoot plants: no significant effect of density dependence
+- Full dataset: No density dependence
 
@@
   glmer(
-    surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
-      (1|ha_id_number) + (1|year_post) + (1|habitat:plot), 
+    surv ~
+      log_shts +
+      habitat +
+      shts_0_2.5 + shts_2.5_5 +
+      (1|ha_id_number) +
+      (1|plot) +
+      (1|year_post), 
     family = binomial, data = ha_sub
@@
 ```{r}
-m_surv2 <- update(m_surv, .~. - (1|habitat:plot))
+m_surv2 <- update(m_surv, .~. - (1|plot))
 AICtab(m_surv, m_surv2)
@@
 ```{r}
-m_surv4 <- glm(surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5, 
+m_surv4 <- glm(surv ~ 
+                 log_shts +
+                 habitat +
+                 shts_0_2.5 + shts_2.5_5, 
                 family = binomial, data = ha_sub)
@@
 
-Positive density dependence for survival.
+- Just 5750: no density dependence
+- Only CF: no density dependence
+- Only 3-shoot plants: Marginally significant density dependence.  Negative at 0--2.5m, positive at 2.5--5m
+- Full dataset: no density dependence
 
@@
   gam(shts_next ~ 
-        s(log_shts) + #Essentially linear, could replace with parametric
+        # s(ha_id_number, bs = "re") + #this takes waaay too long to converge
+        s(plot, bs = "re") +
+        log_shts + #when modeled as a spline, edf is essentially a line
         habitat +
@@
           bs = "cb",
-          xt = list(bs = "cr")), #+
-        # s(ha_id_number, bs = "re"), #this takes waaay too long to converge
+          xt = list(bs = "cr")),
       family = gaussian(link = "log"),
@@
 pred_growth <- pred_cb(Q, L, g_growth)
-ggplot(pred_growth, aes(x = x, y = lag, z = fitted, fill = fitted)) +
+```
+
+Which predicted values are too far from data to sensibly plot?  (this is modeled after `mgcv::exclude.too.far`, which does not work for these models)
+
+```{r}
+dat <-
+  ha_sub %>%
+  select(shts_0.5_1:shts_4.5_5) %>% 
+  pivot_longer(everything(),
+               names_prefix = "shts_",
+               names_sep = "_",
+               names_to = c("inner", "outer"),
+               names_transform = list(inner = as.numeric, outer = as.numeric),
+               values_to = "x") %>% 
+  mutate(lag = (inner + outer) / 2)
+
+which.too.far <- function(grid, data, x, y, dist = 0.1) { 
+  x <- enquo(x)
+  y <- enquo(y)
+  # x <- quo(x)
+  # y <- quo(lag)
+
+  grid1 <-
+    grid %>% 
+    mutate(min_g_x = min(!!x, na.rm = TRUE),
+           min_g_y = min(!!y, na.rm = TRUE),
+           g_x = !!x - min_g_x,
+           g_y = !!y - min_g_y) %>% 
+    mutate(max_g_x = max(g_x, na.rm = TRUE),
+           max_g_y = max(g_y, na.rm = TRUE),
+           g_x = g_x / max_g_x,
+           g_y = g_y / max_g_y)
+
+  data1 <-
+    data %>% 
+    mutate(d_x = (!!x - first(grid1$min_g_x)) / first(grid1$max_g_x),
+           d_y = (!!y - first(grid1$min_g_y)) / first(grid1$max_g_y))
+
+  if (dist < 0)
+    stop("supplied dist negative")
+
+  #where dat is a 2-column matrix of x and y coords of the true data used to build the model
+  min_dist <- function(g_x, g_y, dat) {
+    dat[,1] <- dat[,1] - g_x
+    dat[,2] <- dat[,2] - g_y
+    min(
+      sqrt(
+        (dat[,1]^2 + dat[,2]^2)
+      ), na.rm = TRUE
+    )
+  }
+  o <- 
+    grid1 %>%
+    rowwise() %>%
+    mutate(min_dist = min_dist(g_x, g_y, cbind(data1$d_x, data1$d_y)),
+           too.far = min_dist > dist) %>% 
+    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y)
+  return(o)
+}
+
+pred_growth <- which.too.far(pred_growth, dat, x = x, y = lag)
+```
+
+```{r}
+ggplot(pred_growth %>% filter(min_dist < 0.1),
+       aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
   geom_raster(interpolate = TRUE) +
@@
 
+Plot a "slice" through 5m
+
+```{r}
+ggplot(pred_growth %>% filter(lag == 0.75),
+       aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
+  geom_line() + 
+  geom_ribbon(alpha = 0.3) +
+  geom_point(data = ha_sub, aes(x = shts_0.5_1, y = shts_next), inherit.aes = FALSE) +
+  coord_cartesian(ylim = c(0,20))
+```
+
+- TODO: Try with Florestal (really huge plants there)
+- 5750 only: essentially flat surface, except for positive density dependence at 5m.  Seems spurious, would want to see CI.
+- CF only: similar.  Essentially flat except for impossibly large positive density dependence at 5 m (expected value of 200+ shoots at highest conspecific density)
+- 3-shoot plants only: No significant effect of density.
+- Full dataset: Significant, but same pattern.  Large effects only at 5m.  Probably driven by a few extreme values of high-density plots.
+
 ## Survival
@@
   gam(surv ~
+        s(plot, bs = "re") +
         s(log_shts) +
@@
           bs = "cb",
-          # k = c(10, 10),
           xt = list(bs = "cr")),
@@
 pred_surv <- pred_cb(Q, L, g_surv)
-ggplot(pred_surv, aes(x = x, y = lag, z = fitted, fill = fitted)) +
+ggplot(pred_surv, aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
   geom_raster(interpolate = TRUE) +
@@
 ```
+
+- 5750 only: No significant density dependence
+- CF only: Also not significant.
+- 3-shoot plants only: no significant density dependence
+- Full dataset: no significant effect of density dependence.

### 2020-10-07 R/density dependence.Rmd
finalizing and writing about density dependence

@@
 - Get confidence intervals for cross-basis predictions.  Plot slices with confidence bands.
+- Try with just Florestal (really huge plants there)
 
@@
           bs = "cb",
+          k = c(2,2),
           xt = list(bs = "cr")),
@@
 ```{r}
-ggplot(pred_growth %>% filter(min_dist < 0.1),
-       aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
+pred_growth %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
   geom_raster(interpolate = TRUE) +
@@
 ```{r}
-ggplot(pred_growth %>% filter(lag == 0.75),
-       aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
+pred_growth %>%
+  filter(min_dist < 0.1) %>% 
+  filter(lag == 0.75) %>% 
+  ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
   geom_line() + 
   geom_ribbon(alpha = 0.3) +
-  geom_point(data = ha_sub, aes(x = shts_0.5_1, y = shts_next), inherit.aes = FALSE) +
+  geom_point(data = ha_sub, aes(x = shts_0_0.5, y = shts_next), inherit.aes = FALSE) +
   coord_cartesian(ylim = c(0,20))
@@
 
-- TODO: Try with Florestal (really huge plants there)
 - 5750 only: essentially flat surface, except for positive density dependence at 5m.  Seems spurious, would want to see CI.

### 2020-10-07 analysis/paper/appendix_density_dependence.Rmd
finalizing and writing about density dependence

---
title: "Density dependence"
author: "Eric R. Scott"
date: '2020-09-13'
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(inclue = FALSE)
library(tidyverse)
library(here)
library(conflicted)
library(lme4)
library(bbmle)
library(car)
library(broom)
library(broom.mixed)
# library(patchwork)
# library(performance)
library(dlnm)
library(mgcv)
# library(visibly) # better diagnostic visulazations.  one of several packages for this.
source(here("R", "utils.R"))

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# TODO:

-   Remove unecessary dependencies

-   Move functions to .R file

-   Rearrange analysis with GAM first

# Methods

## Data collection

The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years <!--# methods for collecting these data? -->. In 2006 and 2007, surveys were conducted to \_\_\_\_\_\_\_\_\_ . In 2009 a similar survey was conducted to assess the effects of safe-site limitation on recruitment [@uriarte2010]. For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al. [-@teller2016]. Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e. a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m. If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value. Conspecific density was estimated as the number of H. acuminata shoots per m2 in each annulus. Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in H. acuminata [@bruna2002; @brunaEffectsForestFragmentation2002].

We used the three transition years for which we had density data to test for effects of conspecific density on growth and survival of *H. acuminata*. We were unable to test for the effects of density dependence on probability of flowering, as surveys of reproduction were not conducted in these years <!--# or perhaps no plants flowered? -->.

```{r data}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
# ha
```

```{r tidy}
ha <-
  ha %>%
  ungroup() %>% 
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)),
         log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number),
         plot = as.factor(plot))
```

## Statistical Modeling

We modeled density dependence using a penalized distributed lag non-linear model (DLNM) approach with the `dlnm` package in R [@gasparriniPenalizedFrameworkDistributed2017; @gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This approach is similar to that of [-@teller2016], in that it makes no assumptions about the shape of the competition kernel. Instead, it fits a two-dimensional spline allowing for a non-linear response to the intensity of conspecific density as well as the distance of conspecifics. We implemented this in a generalized additive model (GAM) framework taking advantage of the penalization features of the `mgcv` package ( <!--# CITATION  -->).

```{r}
Q <-
  ha %>%
  select(shts_0_0.5:shts_4.5_5) %>%
  as.matrix()

L <-
  matrix(seq(0.25, 4.75, by = 0.5), #use midpoint of r1 and r2
         nrow = nrow(Q),
         ncol = ncol(Q),
         byrow = TRUE)
```

```{r}
g_growth <-
  gam(shts_next ~ 
        # s(ha_id_number, bs = "re") + #this takes waaay too long to converge
        s(plot, bs = "re") +
        s(log_shts, bs = "cs") + #when modeled as a spline, edf is essentially a line
        # s(log_shts, bs = "cs", by = habitat) + #kind of like an interaction.
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(2,2),
          xt = list(bs = "cr")),
      family = gaussian(link = "log"),
      data = ha,
      method = "REML")
```

Growth was modeled as:

$$
shoots_{t+1} \sim s(log(shoots_t)) + habitat + sldkfjad;lsfk
$$

Where \_\_\_\_\_ is a cross-basis smooth of number of conspecific shoots per square meter $Q$ at distances, $D$, equal to the midpoint of the inner and outer radius of the 5 annuli for which density was calculated. Each dimension was modeled with \_\_\_\_ knots. With a penalized spline, the number of knots is not very important as <!--# finish thought and cite (maybe Colin Edwards?) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat. We did not include a random effect of plant ID, as the model did not converge. Although the response, shoot number in year t+1, is a discrete variable, a log-normal distribution was used as it was a better fit than a poisson distribution ($\Delta$AIC = 527.2).

```{r}
g_growth_pois <- update(g_growth, family = poisson(link = "log"))
AICtab(g_growth, g_growth_pois)
```

Survival was modeled using the same predictors as for growth, but with a binomial family error distribution.

```{r}
g_surv <-
  gam(surv ~
        s(plot, bs = "re") +
        s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          xt = list(bs = "cr")),
      family = binomial,
      data = ha,
      method = "REML")
```

Because the DLNM approach does not allow for modeling an interaction between conspecific density and habitat type, we used the outputs of these cross-basis smooths to fit generalized linear mixed-effects models with interaction terms. By inspecting the marginal effects plots from the cross-basis smooths, we identified a range of distances over which conspecific density had the greatest effect and used density over that distance as a predictor in the GLMMs. <!--# revise and also talk more about what interactions and random effects included -->

# Results

### Growth

```{r}
# g_growth2 <- update(g_growth, ~-s(plot, bs = "re"))
# AICtab(g_growth, g_growth2)
# g_growth3 <- update(g_growth, ~-s(log_shts) + log_shts)
# AICtab(g_growth, g_growth2, g_growth3)
# Best model:
anova(g_growth)
```

```{r growth-plot, echo=FALSE, include=TRUE}
#TODO: show only s(log_shts), shift reference to intercept for model.
plot(g_growth, pages = 1)
```

Number of shoots in the current year had a significant effect on size in year t + 1 (F = 353.8, edf = 5.2, P \< 0.0001). The relationship between $log(shoots_{t+1})$ and $log(shoots_t)$ was essentially linear for all but the largest size classes \@ref(fig:growth-plot). There were no significant differences between habitat types (F = 0.012, df = 1, P = 0.912). The cross-basis smooth describing the effects of conspecific density on growth was significant (F = 2.78, edf = 18.4, P \< 0.001).

```{r}
pred_growth <- pred_cb(Q, L, g_growth)
```

```{r}
#TODO:
# - marginal_cb() function should add the min_dist column.
# - plot_marginal_cb() function should do ggplot stuff and include a too.far argument
# - All this moved into a .R file that is source()d
dat <-
  ha %>%
  select(shts_0.5_1:shts_4.5_5) %>% 
  pivot_longer(everything(),
               names_prefix = "shts_",
               names_sep = "_",
               names_to = c("inner", "outer"),
               names_transform = list(inner = as.numeric, outer = as.numeric),
               values_to = "x") %>% 
  mutate(lag = (inner + outer) / 2)

which.too.far <- function(grid, data, x, y, dist = 0.1) { 
  x <- enquo(x)
  y <- enquo(y)
  # x <- quo(x)
  # y <- quo(lag)

  grid1 <-
    grid %>% 
    mutate(min_g_x = min(!!x, na.rm = TRUE),
           min_g_y = min(!!y, na.rm = TRUE),
           g_x = !!x - min_g_x,
           g_y = !!y - min_g_y) %>% 
    mutate(max_g_x = max(g_x, na.rm = TRUE),
           max_g_y = max(g_y, na.rm = TRUE),
           g_x = g_x / max_g_x,
           g_y = g_y / max_g_y)

  data1 <-
    data %>% 
    mutate(d_x = (!!x - first(grid1$min_g_x)) / first(grid1$max_g_x),
           d_y = (!!y - first(grid1$min_g_y)) / first(grid1$max_g_y))

  if (dist < 0)
    stop("supplied dist negative")

  #where dat is a 2-column matrix of x and y coords of the true data used to build the model
  min_dist <- function(g_x, g_y, dat) {
    dat[,1] <- dat[,1] - g_x
    dat[,2] <- dat[,2] - g_y
    min(
      sqrt(
        (dat[,1]^2 + dat[,2]^2)
      ), na.rm = TRUE
    )
  }
  o <- 
    grid1 %>%
    rowwise() %>%
    mutate(min_dist = min_dist(g_x, g_y, cbind(data1$d_x, data1$d_y)),
           too.far = min_dist > dist) %>% 
    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y)
  return(o)
}

pred_growth <- which.too.far(pred_growth, dat, x = x, y = lag)
```

```{r growth-cb, echo=FALSE, include=TRUE}
pred_growth %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  # geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A") +
  theme_bw()
```

Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average. The effects of conspecific density seemed strongest around 1.5 -- 2 meters from the focal plant \@ref(fig:growth-cb).

```{r growth-slice, echo=FALSE, include=TRUE}
pred_growth %>%
  filter(min_dist < 0.1) %>% 
  filter(lag == 1.75) %>% 
  ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  geom_line() + 
  geom_ribbon(alpha = 0.3) +
  geom_jitter(data = ha,
             aes(x = shts_1.5_2, y = shts_next),
             alpha = 0.1,
             inherit.aes = FALSE) +
  labs(title = "annulus 1.5–2m", 
       x = "conspecific density (shoots/m^2)",
       y = "#shoots (t+1)")
```

In that annulus, conspecific density showed no effect (i.e. a flat line) below around 4 shoots/m^2^ \@ref(fig:growth-slice). Only a few focal plants had a conspecific density greater than 4 shoots/m^2^, so the support for positive density dependence in this spline is driven strongly by only a few points.

### Survival

```{r}
# plot_gam_check(g_surv)
anova(g_surv)
```

Plant size had a significant, nearly quadratic, effect on survival (F = 126.59, edf = 2.1, P \< 0.001) \@ref(fig:surv-plot). However, there were no significant effects of habitat ($\chi^2= 0.26$, df = 1, P = 0.613) or conspecific density ($\chi^2 = 126.59$, edf = 2, P = 0.074) on survival.

```{r surv-plot, include=TRUE, echo=FALSE}
#TODO:
# - plot only size vs survival
# - back-transform shit
# - shift intercept
plot(g_surv, pages = 1)
```

TODO:

-   re-fit growth with GLMM using density out to 2m, include interactions between habitat and growth, habitat and conspecific density. Include random effects of plot, plant ID, and year.

-   If non-significant, conclude that effects seen in GAM are largely driven by only a few plants that happened to have large neighbors (could even remove a single plant and see how it changes)

-   Hopefully can say there is no evidence for density dependence.

### 2020-10-07 analysis/paper/paper.Rmd
finalizing and writing about density dependence

---
title: "Title Goes Here"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF]
      correspondence: true
  - Paulo Rubim:
      email: rubimpaulo@gmail.com
      institute: [UFSC]
      correspondence: false
  - Maria Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      intitute: [UF]
      correspondence: false

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UFSC: Departamento do Botânica, Universidade Federal de São Carlos, CP 676, CEP 13565-905, São Carlos, SP, Brazil
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  density dependence; vital rates; *Heliconia acuminata*
highlights: |
  These are the highlights
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

# Background

# Methods

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible understory monocot native to the nonflooded rainforests of central Amazonia. *Heliconia* are rhizomatious, but clonal reproduction via underground runners is rare in *H. acuminata*, unlike other congeners [@bruna2004]. Vegetative shoots grow from the rhizome as well as one or more flowering shoots when plants reproduce. Each flowering shoot has a single inflorescence of 20--25 flowers [@bruna2002]. Flowers are pollinated nearly exclusively by two species of hummingbirds at this site <!--# species? -->. Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011].

## Site

-   All field data were collected at the Biological Dynamics of Forest Fragments Project (BDFFP)---a long term experimental site located about 80km north of Manaus (2º30' S, 60ºW) [@laurance2011; @laurance2002; @bierregaard1992]. The experimental plots are located in BDFFP reserves in nonflooded tropical lowland rainforest with nutrient-poor, well-draining, xanthic ferrosols.

-   Something about seasonality

-   BDFFP is comprised of replicated forest fragments of 1, 10, and 100ha as well as continuous forest plots that were established between 1980 and 1984. Fragments were maintained by clearing secondary growth around fragments occasionally <!--# Best reference for this? -->. Demographic plots ($50 \times 100$m) were established in <!--# year --> to track the growth, reproduction, and survival of *Heliconia acuminata*. Six of these plots were established in continuous forest, four were in 1-ha fragments, and three were in 10-ha fragments. The plots in continuous forest were established at haphazardly chosen locations 500--4000m from any mature forest borders, plots in 1-ha fragments were established on a randomly selected half of the fragment, and plots in 10-ha fragments were established in the center.

## Data collection

The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years <!--# methods for collecting these data? -->. In 2006 and 2007, surveys were conducted to \_\_\_\_\_\_\_\_\_ . In 2009 a similar survey was conducted to assess the effects of safe-site limitation on recruitment [@uriarte2010]. For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al. [-@teller2016]. Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e. a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m. If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value. Conspecific density was estimated as the number of *H. acuminata* shoots per m^2^ in each annulus. Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in *H. acuminata* [@bruna2002; @brunaEffectsForestFragmentation2002].

We used the three transition years for which we had density data to model the effects of conspecific density on survival, growth, and reproduction of *H. acuminata*. The collection of demographic data is described elsewhere [DATA PAPER], but briefly...

# Results

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2020-10-07 analysis/paper/appendix_density_dependence.Rmd
extracted marginal effects for crossbasis smooths function to .R file.  Removed unused functions and dependencies

@@
 library(conflicted)
-library(lme4)
 library(bbmle)
-library(car)
-library(broom)
-library(broom.mixed)
-# library(patchwork)
-# library(performance)
 library(dlnm)
 library(mgcv)
-# library(visibly) # better diagnostic visulazations.  one of several packages for this.
-source(here("R", "utils.R"))
+
+# library(lme4)
+
+# library(car)
+# library(broom)
+# library(broom.mixed)
+
+source(here("R", "crossbasis.R"))
 
@@
 
--   Remove unecessary dependencies
+-   Remove unnecessary dependencies
 
--   Move functions to .R file
-
--   Rearrange analysis with GAM first
+-   Rearrange analysis with GAM first (i.e. finish adding `glmer()` code)
 
@@
 ```{r}
-pred_growth <- pred_cb(Q, L, g_growth)
-```
-
-```{r}
-#TODO:
-# - marginal_cb() function should add the min_dist column.
-# - plot_marginal_cb() function should do ggplot stuff and include a too.far argument
-# - All this moved into a .R file that is source()d
-dat <-
-  ha %>%
-  select(shts_0.5_1:shts_4.5_5) %>% 
-  pivot_longer(everything(),
-               names_prefix = "shts_",
-               names_sep = "_",
-               names_to = c("inner", "outer"),
-               names_transform = list(inner = as.numeric, outer = as.numeric),
-               values_to = "x") %>% 
-  mutate(lag = (inner + outer) / 2)
-
-which.too.far <- function(grid, data, x, y, dist = 0.1) { 
-  x <- enquo(x)
-  y <- enquo(y)
-  # x <- quo(x)
-  # y <- quo(lag)
-
-  grid1 <-
-    grid %>% 
-    mutate(min_g_x = min(!!x, na.rm = TRUE),
-           min_g_y = min(!!y, na.rm = TRUE),
-           g_x = !!x - min_g_x,
-           g_y = !!y - min_g_y) %>% 
-    mutate(max_g_x = max(g_x, na.rm = TRUE),
-           max_g_y = max(g_y, na.rm = TRUE),
-           g_x = g_x / max_g_x,
-           g_y = g_y / max_g_y)
-
-  data1 <-
-    data %>% 
-    mutate(d_x = (!!x - first(grid1$min_g_x)) / first(grid1$max_g_x),
-           d_y = (!!y - first(grid1$min_g_y)) / first(grid1$max_g_y))
-
-  if (dist < 0)
-    stop("supplied dist negative")
-
-  #where dat is a 2-column matrix of x and y coords of the true data used to build the model
-  min_dist <- function(g_x, g_y, dat) {
-    dat[,1] <- dat[,1] - g_x
-    dat[,2] <- dat[,2] - g_y
-    min(
-      sqrt(
-        (dat[,1]^2 + dat[,2]^2)
-      ), na.rm = TRUE
-    )
-  }
-  o <- 
-    grid1 %>%
-    rowwise() %>%
-    mutate(min_dist = min_dist(g_x, g_y, cbind(data1$d_x, data1$d_y)),
-           too.far = min_dist > dist) %>% 
-    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y)
-  return(o)
-}
-
-pred_growth <- which.too.far(pred_growth, dat, x = x, y = lag)
+pred_growth <- cb_margeff(Q, L, g_growth)
 ```

### 2020-10-07 R/crossbasis.R

@@ -1,0 +1,181 @@
+#' Calculate marginal effects of a crossbasis smooth
+#' 
+#' Calculate marginal effects of a crossbasis smooth created with the `cb` basis
+#' from the `dlnm` package in a model created with `gam()`. Everything is kept
+#' average (or at reference value for factors) and response values are predicted
+#' using the range of values of Q at each lag defined by L.
+#'
+#' @param Q The matrix of predictor values used to generate a crossbasis smooth
+#' @param L The matrix of lags used to generate a crossbasis smooth
+#' @param model a gam with a crossbasis smooth
+#' @param meshpts The number of meshpoints for values of Q to use to generate
+#'   fitted values
+#' @param calc_dist logical; Calculate distance between predicted values and
+#'   actual data points?  This is inspired by the `too.far` argument of
+#'   `plot.gam()`.  If `TRUE` (default), it adds the column `min_dist` which is
+#'   useful for filtering data before plotting because "smooths tend to go wild
+#'   away from data".
+#'
+#' @return a tibble suitable for plotting marginal effects as a heatmap or
+#'   contour plot.  `x` is the meshpoint values across the range of the
+#'   predictor, `Q`; `lag` is the values of `L`; `fitted` and `se.fit` are the
+#'   results of `predict.gam()`; min_dist is the euclidean distance on the unit
+#'   square from the fitted values to the actual data used to fit the model.
+#' @export
+#' @import rlang
+#' @import purrr
+#' @import dplyr
+#' @import mgcv
+#' @import tidyr
+#'
+#' @examples
+cb_margeff <- 
+  function(Q, L, model, meshpts = 200, calc_dist = TRUE) {
+    # Q_name <- quo(Q)
+    # L_name <- quo(L)
+    
+    if (!inherits(model, "gam")) {
+      abort("This is only for GAMs made with the `mgcv` package including cross-basis smooths from the `dlnm` package.")
+    } 
+    
+    Q_name <- rlang::enquo(Q)
+    L_name <- rlang::enquo(L)
+    df <- model$model
+    
+    testvals <- seq(min(Q, na.rm = TRUE), max(Q, na.rm = TRUE), length.out = meshpts)
+    Q_new <- matrix(mean(Q, na.rm = TRUE), nrow = length(testvals), ncol = ncol(L))
[…]

### 2020-10-07 R/utils.R

@@ -14,239 +14,19 @@
 }
 
 
-#' Create masking matrices with rectangles
-#' 
-#' Creates a list of matrices with all 0's except for a rectangle of 1's around
-#' a focal "cell". Used for figuring out how many neighboring plants each plant
-#' has.
+#' Calculate euclidean distance between two points
 #'
-#' @param D distance from focal cell for rectangle
-#' @param nrow number of rows of matrix
-#' @param ncol number of columns of matrix
-#' @author Lyndsay Troyer
-#' @return a list of matrices
+#' @param p1 numeric vector of length 2; x and y coordinates of point 1
+#' @param p2 numeric vector of length 2; x and y coordinates of point 2
+#'
+#' @return Euclidean distance between the points
 #' @export
-#' 
-#' @import glue
-#' 
-#' @examples
-#' I <- mat_rect(D = 1, nrow = 5, ncol = 5)
-#' y <- matrix(round(runif(25, 1, 10), 0), nrow = 5, ncol = 5)
-#' map_dbl(I, ~ sum(y * .x)) %>%
-#'   enframe(value = "n_plants_2") %>%
-#'   separate(name, into = c("row", "column"), sep = ",", convert = TRUE)
-mat_rect <- function(D, nrow, ncol) {
-  mlist <- vector("list", 1)
-  for(r in 1:nrow) {
-    for(c in 1:ncol) {
-      m <- matrix(0, ncol = ncol, nrow = nrow)
-      r1 <- r + D
-      r1 <- r1[r1 <= nrow]
-      c1 <- c + (-D:D)
-      c1 <- c1[c1 > 0 & c1 <= ncol]
-      m[r1, c1] <- 1
-      
-      r2 <- r - D
-      r2 <- r2[r2 > 0]
-      c2 <- c - (-D:D)
-      c2 <- c2[c2 > 0 & c2 <= ncol]
-      m[r2, c2] <- 1
-      
-      r3 <- r - (-D:D)
-      r3 <- r3[r3 > 0 & r3 <= nrow]
-      c3 <- c - D
-      c3 <- c3[c3 > 0]
-      m[r3, c3] <- 1
-      
-      r4 <- r + (-D:D)
-      r4 <- r4[r4 > 0 & r4 <= nrow]
-      c4 <- c + D
-      c4 <- c4[c4 <= ncol]
-      m[r4, c4] <- 1
-      
-      m <- list(m)
[…]

### 2020-10-07 analysis/paper/appendix_density_dependence.Rmd
fixed up plots of 1d smooths.

@@
 library(mgcv)
+# library(visibly)
 
@@
 # TODO:
-
--   Remove unnecessary dependencies
 
@@
 ```{r growth-plot, echo=FALSE, include=TRUE}
-#TODO: show only s(log_shts), shift reference to intercept for model.
-plot(g_growth, pages = 1)
+Q_new <- 
+  g_growth$model %>% 
+  pull(Q) %>%
+  as_tibble() %>% 
+  summarize(across(everything(), mean)) %>%
+  as.matrix()
+L_new <- L[1,, drop = FALSE]
+
+newdata <- tibble(log_shts = seq(0, 2.565, length.out = 100), habitat = "1-ha", plot = "new", Q = Q_new, L = L_new)
+
+fitted <- augment(g_growth, newdata = newdata, se_fit = TRUE, conf.level = 0.95)
+ggplot(fitted, aes(x = exp(log_shts), y = exp(.fitted))) +
+  geom_jitter(data = ha %>%
+                filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
+              aes(x = shts, y = shts_next),
+              alpha = 0.2) +
+  geom_line(color = "blue") +
+  geom_ribbon(aes(ymin = exp(.fitted - .se.fit*1.96), ymax = exp(.fitted + .se.fit*1.96)), alpha = 0.4) +
+  labs(x = "#shoots (t)", y = "#shoots (t + 1)") +
+  theme_bw()
 ```
@@
   ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
-  geom_line() + 
-  geom_ribbon(alpha = 0.3) +
-  geom_jitter(data = ha,
+  geom_jitter(data = ha %>%
+                filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
               aes(x = shts_1.5_2, y = shts_next),
-             alpha = 0.1,
+              alpha = 0.2,
               inherit.aes = FALSE) +
+  geom_line(color = "blue") + 
+  geom_ribbon(alpha = 0.4) +
+
   labs(title = "annulus 1.5–2m", 
        x = "conspecific density (shoots/m^2)",
-       y = "#shoots (t+1)")
+       y = "#shoots (t+1)") +
+  theme_bw()
 ```
@@
 ```{r surv-plot, include=TRUE, echo=FALSE}
-#TODO:
-# - plot only size vs survival
-# - back-transform shit
-# - shift intercept
-plot(g_surv, pages = 1)
+Q_new <- 
+  g_surv$model %>% 
+  pull(Q) %>%
+  as_tibble() %>% 
+  summarize(across(everything(), mean)) %>%
+  as.matrix()
+L_new <- L[1,, drop = FALSE]
+
+newdata <- tibble(log_shts = seq(0, 2.565, length.out = 100), habitat = "1-ha", plot = "new", Q = Q_new, L = L_new)
+
+fitted <- augment(g_surv, newdata = newdata, se_fit = TRUE)
+
+#for rug
+rugdat <- 
+  ha %>%
+  filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))) %>% 
+  mutate(survived = ifelse(surv == 1, shts, NA),
+         died = ifelse(surv == 0, shts, NA))
+
+ggplot(fitted, aes(x = exp(log_shts), y = plogis(.fitted))) +
+  geom_line(color = "blue") +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit*1.96), ymax = plogis(.fitted + .se.fit*1.96)), alpha = 0.4) +
+  # geom_rug(data = rugdat,
+  #          aes(x = survived),
+  #          inherit.aes = FALSE,
+  #          sides = "t",
+  #          alpha = 0.2) +
+  # geom_rug(data = rugdat,
+  #          aes(x = died),
+  #          inherit.aes = FALSE,
+  #          sides = "b",
+  #          alpha = 0.2) +
+  stat_summary_bin(geom = "point",
+                   fun = mean,
+                   aes(y = surv),
+                   bins = 1000,
+                   data = ha %>%
+                     filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.)))) +
+  labs(x = "#shoots (t)", y = "survival probability") +
+  scale_x_continuous(n.breaks = 7) +
+  theme_bw()
 ```

### 2020-10-15 R/competition wrangling.Rmd
wrapped up analysis of density dependence

@@
   future_map(~{
-    add_density_cols(.x, L = 0, U = 5, by = 2.5)
+    add_density_cols(.x, L = 0, U = 3, by = 3)
   }) %>% 

### 2020-10-15 analysis/paper/appendix_density_dependence.Rmd
wrapped up analysis of density dependence

@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(inclue = FALSE)
+```{r setup, include=FALSE, echo=FALSE}
+knitr::opts_chunk$set(
+  include = FALSE,
+  echo = FALSE,
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  comment = "#>",
+  fig.path = "../figures/",
+  dpi = 300)
 library(tidyverse)
@@
 library(mgcv)
-# library(visibly)
-
-# library(lme4)
-
-# library(car)
-# library(broom)
-# library(broom.mixed)
+library(broom)
+library(broom.mixed)
+library(lme4)
+library(car)
 
@@
 
-# TODO:
+# TODO
 
--   Rearrange analysis with GAM first (i.e. finish adding `glmer()` code)
+-   figure captions
+
+-   revise
+
+-   equations for models
+
+-   Explain why heatmap is missing a chunk (search literature for other 2d gam papers that use `too.far` argument?)
 
@@
 
-The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years <!--# methods for collecting these data?
--->.
-In 2006 and 2007, surveys were conducted to \_\_\_\_\_\_\_\_\_ .
-In 2009 a similar survey was conducted to assess the effects of safe-site limitation on recruitment [@uriarte2010].
-For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al.
+The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years.
+In 2006 and 2007, locations of plants were recorded to the nearest meter as part of a study to assess the effects of safe-site limitation on recruitment [@uriarte2010].
+In 2009, plants were mapped to ....
+<!--# rubim study -->For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al.
 [-@teller2016].
@@
 Instead, it fits a two-dimensional spline allowing for a non-linear response to the intensity of conspecific density as well as the distance of conspecifics.
-We implemented this in a generalized additive model (GAM) framework taking advantage of the penalization features of the `mgcv` package ( <!--# CITATION  -->).
+We implemented this in a generalized additive model (GAM) framework taking advantage of the penalization features of the `mgcv` package [@woodGeneralizedAdditiveModels2017].
 
@@
   gam(shts_next ~ 
-        # s(ha_id_number, bs = "re") + #this takes waaay too long to converge
+        # s(ha_id_number, bs = "re") + #this doesn't seem to converge
         s(plot, bs = "re") +
@@
           bs = "cb",
-          # k = c(2,2),
+          # k = c(3,3),
           xt = list(bs = "cr")),
@@
       method = "REML")
+g_growth$smooth
 ```
 
-Growth was modeled as:
+Growth was modeled as: (still not sure how to do this. Summation is from Gasparini et al, but also written as simply as $s(\textbf{q}_{i,t})$
 
 $$
-shoots_{t+1} \sim s(log(shoots_t)) + habitat + sldkfjad;lsfk
+size_{t+1} = \sum_{d=d_0}^D f \cdot w(x_{i,d}, d)
 $$
 
-Where \_\_\_\_\_ is a cross-basis smooth of number of conspecific shoots per square meter $Q$ at distances, $D$, equal to the midpoint of the inner and outer radius of the 5 annuli for which density was calculated.
-Each dimension was modeled with \_\_\_\_ knots.
-With a penalized spline, the number of knots is not very important as <!--# finish thought and cite (maybe Colin Edwards?) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat.
+Where $f \cdot w(x, d)$ is a two dimensional crossbasis smooth composed of two marginal functions: $f(x)$, the effect of conspecific density, $x$ , on growth, and the additional distance response function $w(d)$ that models the effect of density at different annuli, $d$, equal to the midpoint of the inner and outer radius of the 5 annuli for which density was calculated.
+Each dimension was modeled with 10 knots.
+With a penalized spline, the number of knots is not very important as ...
+<!--# finish thought and cite (maybe Colin Edwards?) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat.
 We did not include a random effect of plant ID, as the model did not converge.
@@
 
-### Growth
+## Growth
 
@@
 
-```{r growth-plot, echo=FALSE, include=TRUE}
+```{r growth-plot, include=TRUE, fig.cap="this is a figure"}
 Q_new <- 
@@
 Number of shoots in the current year had a significant effect on size in year t + 1 (F = 353.8, edf = 5.2, P \< 0.0001).
-The relationship between $log(shoots_{t+1})$ and $log(shoots_t)$ was essentially linear for all but the largest size classes \@ref(fig:growth-plot).
+The relationship between $log(shoots_{t+1})$ and $log(shoots_t)$ was essentially linear for all but the largest size classes (Figure \@ref(fig:growth-plot)).
 There were no significant differences between habitat types (F = 0.012, df = 1, P = 0.912).
@@
 
-```{r growth-cb, echo=FALSE, include=TRUE}
+```{r growth-cb, include=TRUE, fig.cap="this is a figure"}
 pred_growth %>%
@@
 Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average.
-The effects of conspecific density seemed strongest around 1.5 -- 2 meters from the focal plant \@ref(fig:growth-cb).
+The effects of conspecific density seemed strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).
 
-```{r growth-slice, echo=FALSE, include=TRUE}
+```{r growth-slice, include=TRUE, fig.cap="this is a figure"}
 pred_growth %>%
@@
 In that annulus, conspecific density showed no effect (i.e.
-a flat line) below around 4 shoots/m^2^ \@ref(fig:growth-slice).
-Only a few focal plants had a conspecific density greater than 4 shoots/m^2^, so the support for positive density dependence in this spline is driven strongly by only a few points.
+a flat line) below around 4 shoots m^-2^ (Figure \@ref(fig:growth-slice)).
+Only a few focal plants had a conspecific density greater than 4 shoots m^-2^, so the support for positive density dependence in this spline is driven strongly by only a few points.
+Additionally, it is important to note that p-values from splines in generalized additive model are approximate and likely too low [@woodGeneralizedAdditiveModels2017].
 
-### Survival
+## Survival
 
@@
 
-Plant size had a significant, nearly quadratic, effect on survival (F = 126.59, edf = 2.1, P \< 0.001) \@ref(fig:surv-plot).
-However, there were no significant effects of habitat ($\chi^2= 0.26$, df = 1, P = 0.613) or conspecific density ($\chi^2 = 126.59$, edf = 2, P = 0.074) on survival.
+Plant size had a significant, nearly quadratic, effect on survival (F = 126.59, edf = 2.1, P \< 0.001) (Figure \@ref(fig:surv-plot)).
+However, there were no significant effects of habitat ($\chi^2$ = 0.26, df = 1, P = 0.613) or conspecific density ($\chi^2$ = 126.59, edf = 2, P = 0.074) on survival.
 
-```{r surv-plot, include=TRUE, echo=FALSE}
+```{r surv-plot, include=TRUE, fig.cap="this is a figure"}
 Q_new <- 
@@
   geom_ribbon(aes(ymin = plogis(.fitted - .se.fit*1.96), ymax = plogis(.fitted + .se.fit*1.96)), alpha = 0.4) +
-  # geom_rug(data = rugdat,
-  #          aes(x = survived),
-  #          inherit.aes = FALSE,
-  #          sides = "t",
-  #          alpha = 0.2) +
-  # geom_rug(data = rugdat,
-  #          aes(x = died),
-  #          inherit.aes = FALSE,
-  #          sides = "b",
-  #          alpha = 0.2) +
+  geom_rug(data = rugdat,
+           aes(x = survived),
+           inherit.aes = FALSE,
+           sides = "t",
+           alpha = 0.2) +
+  geom_rug(data = rugdat,
+           aes(x = died),
+           inherit.aes = FALSE,
+           sides = "b",
+           alpha = 0.2) +
   stat_summary_bin(geom = "point",
@@
 
-TODO:
+## Mixed Models Approach
 
--   re-fit growth with GLMM using density out to 2m, include interactions between habitat and growth, habitat and conspecific density. Include random effects of plot, plant ID, and year.
+Because the DLNM approach does not allow for interaction terms and did not converge with the inclusion of certain random effects, we repeated the analysis using a more standard generalized linear mixed effects model approach.
+Because the effects of conspecific density dropped off steeply beyond 3m from the focal plant (Figure \@ref(fig:growth-cb)), we re-calculated conspecific shoot density within a radius of 3m and used this as a predictor.
+We used AIC comparisons to test for significance of random effects, then interactions, then used a marginal hypothesis test to test for significance of main effects of log(shoot number), habitat, and conspecific density.
+Marginal hypothesis tests were conducted with the `Anova` function from the `car` package [@foxCompanionAppliedRegression2019].
 
--   If non-significant, conclude that effects seen in GAM are largely driven by only a few plants that happened to have large neighbors (could even remove a single plant and see how it changes)
+```{r eval=FALSE, include=FALSE}
+pred_growth %>%
+  filter(min_dist < 0.1) %>% 
+  ungroup() %>% 
+  filter(fitted == max(fitted))
 
--   Hopefully can say there is no evidence for density dependence.
+pred_growth %>%
+  filter(min_dist < 0.1) %>% 
+  filter(near(x, 5.01, tol = 0.01)) %>% 
+  ggplot(aes(x = lag, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
+  # geom_jitter(data = ha %>%
+  #               filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
+  #             aes(x = shts_1.5_2, y = shts_next),
+  #             alpha = 0.2,
+  #             inherit.aes = FALSE) +
+  # geom_smooth(color = "blue", se = FALSE) + 
+  geom_line(color = "blue") +
+  geom_ribbon(alpha = 0.4) +
+
+  labs(title = "Conspecific density of 5 shoots/m^2", 
+       x = "Midpoint distance from focal plant (m)",
+       y = "#shoots (t+1)") +
+  theme_bw()
+```
+
+```{r}
+mnorm <-
+  glmer(
+    shts_next ~ 
+      log_shts *
+      habitat *
+      shts_0_3 +
+      (1|ha_id_number) +
+      (1|plot) +
+      (1|year_post),
+    family = gaussian(link = "log"),
+    data = ha
+  )
+# Random effects
+mnorm2 <- update(mnorm, . ~ . - (1|ha_id_number))
+AICtab(mnorm, mnorm2) #keep ID number
+mnorm3 <- update(mnorm, .~. - (1|plot)) 
+AICtab(mnorm, mnorm2, mnorm3) #remove plot
+mnorm4 <- update(mnorm3, .~.-(1|year_post)) #remove year
+AICtab(mnorm, mnorm2, mnorm3, mnorm4)
+
+# Interactions
+mnorm5 <- update(mnorm4, .~. - log_shts:habitat:shts_0_3)
+AICtab(mnorm4, mnorm5) #good to remove 3-way interaction
+mnorm6 <- update(mnorm5, .~. -log_shts:habitat)
+mnorm7 <- update(mnorm5, .~. -log_shts:shts_0_3)
+mnorm8 <- update(mnorm5, .~. -habitat:shts_0_3)
+mnorm9 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:shts_0_3)
+mnorm10 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:habitat)
+mnorm11 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:habitat -log_shts:shts_0_3)
+AICtab(mnorm5, mnorm6, mnorm7, mnorm8, mnorm9, mnorm10, mnorm11) #good to remove all interactions
+
+Anova(mnorm11)
+fixef(mnorm11)
+```
+
+**No longer significant**
+
+In the GLMM, only plant ID number was retained as a random effect.
+Plant size (log(shoot number)) was a significant predictor of plant size in the next year ($\chi^2$ = 1422.4, df = 1, P \< 0.001).
+Conspecific density within 3m of the focal plant did not have a significant effect of plant size in the next year ($\chi^2$ = 1.26, df = 1, P = 0.2623).
+
+```{r glmer-plot, fig.cap="this"}
+
+ha_sub <- ha %>% filter(!is.na(shts_0_3), !is.na(log_shts))
+newdata <-
+  tibble(log_shts = mean(ha_sub$log_shts),
+         habitat = "1-ha", 
+         ha_id_number = "new",
+         shts_0_3 = seq(min(ha_sub$shts_0_3, na.rm = TRUE),
+                          max(ha_sub$shts_0_3, na.rm = TRUE),
+                          length.out = 50))
+
+augment(mnorm11, newdata = newdata, re.form = NA) %>% 
+  ggplot(aes(x = shts_0_3, y = exp(.fitted))) +
+  geom_jitter(data = ha_sub, aes(x = shts_0_3, y = shts_next), alpha = 0.2) +
+  geom_line(color = "blue") +
+  # geom_ribbon() + #don't have CI yet.  Need to do something like bootMer to get them with bootstrapping.
+  labs(x = "shoot density within 3m (shoots/m^2)", y = "shoots (t+1)") +
+  theme_bw()
+```
+
+# Discussion
+
+We found no meaningful statistically significant effect of conspecific density on the growth or survival of *H.
+accuminata.* In the GAM approach, the statistically significant effect we saw was driven by a few (\< 20) plants with with a high density of conspecific shoots.
+The GAM model also did not take into account variation in growth among individual plants, which was shown to be important in the GLMER model.
+In the GLMER, the effects of density are not significant.
+Because of this, we feel it is safe to exclude density dependence from further demographic models.
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::

### 2020-10-15 R/crossbasis.R

@@ -175,7 +175,8 @@ add_min_dist <- function(df, Q_name, L, pred) {
     grid %>% 
     rowwise() %>% 
     mutate(min_dist = min_dist(g_x, g_y, cbind(d$d_x, d$d_y))) %>% 
-    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y)
+    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y) %>% 
+    ungroup()
   
   return(out)
 }

### 2020-10-21 analysis/paper/appendix_density_dependence.Rmd
updated to use scaled-t family distribution

@@
 ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
-# ha
+ha$year %>% range()
 ```
@@
         s(plot, bs = "re") +
-        s(log_shts, bs = "cs") + #when modeled as a spline, edf is essentially a line
-        # s(log_shts, bs = "cs", by = habitat) + #kind of like an interaction.
-        habitat +
+        # s(log_shts, bs = "cr", by = habitat) + #prelim modeling shows this is a linear relationship (edf ~ 1)
+        habitat * log_shts +
         s(Q, L,
@@
           # k = c(3,3),
-          xt = list(bs = "cr")),
+          xt = list(bs = c("cr"))),
       family = gaussian(link = "log"),
@@
       method = "REML")
-g_growth$smooth
 ```
@@
 We did not include a random effect of plant ID, as the model did not converge.
-Although the response, shoot number in year t+1, is a discrete variable, a log-normal distribution was used as it was a better fit than a poisson distribution ($\Delta$AIC = 527.2).
+Although the response, shoot number in year t+1, is a discrete variable, a scaled t distribution with a log-link was used as it was a better fit than a poisson distribution ($\Delta$AIC = 810.5) and a log-normal distribution ($\Delta$AIC = 294).
 
@@
 g_growth_pois <- update(g_growth, family = poisson(link = "log"))
-AICtab(g_growth, g_growth_pois)
+g_growth_scat <- update(g_growth, family = scat(link = "log")) #because residuals are leptokurtic with gaussian
+AICtab(g_growth, g_growth_pois, g_growth_scat)
 ```
@@
         s(plot, bs = "re") +
-        s(log_shts) +
+        s(log_shts, bs = "cr", by = habitat) +
         habitat +
@@
       method = "REML")
+# plot(g_surv)
 ```
@@
 # Best model:
-anova(g_growth)
+anova(g_growth_scat)
+# plot(g_growth_scat)
 ```

### 2020-10-22 R/utils.R
added nearest() function to be used for filtering to plot slices of crossbasis results.

@@ -30,3 +30,18 @@ eu_dist <- function(p1, p2) {
   sqrt(sum((p1 - p2)^2))
 }
 
+#' Which element(s) of a vector is (are) nearest to some value?
+#' 
+#' For use in dplyr::filter() to get all the rows with the value closest to some target value.
+#'
+#' @param x a numeric vector
+#' @param val a numeric scalar
+#' @seealso dplyr::near()
+#'
+#' @return a logical vector
+#' @export
+#'
+#' @examples
+nearest <- function(x, val) {
+  abs(x - val) == min(abs(x - val))
+}

### 2020-10-22 R/crossbasis.R
documenting change to meshpts arg

@@ -8,14 +8,14 @@
 #' @param Q The matrix of predictor values used to generate a crossbasis smooth
 #' @param L The matrix of lags used to generate a crossbasis smooth
 #' @param model a gam with a crossbasis smooth
-#' @param meshpts The number of meshpoints for values of Q to use to generate
-#'   fitted values
+#' @param meshpts vector of length 2; The number of meshpoints for values of Q
+#'   and L, respectively, to use to generate fitted values
 #' @param calc_dist logical; Calculate distance between predicted values and
 #'   actual data points?  This is inspired by the `too.far` argument of
 #'   `plot.gam()`.  If `TRUE` (default), it adds the column `min_dist` which is
 #'   useful for filtering data before plotting because "smooths tend to go wild
-#'   away from data".
-#'
+#'   away from data". This step takes a long time, so if you don't need it,
+#'   consider setting to `FALSE`
 #' @return a tibble suitable for plotting marginal effects as a heatmap or
 #'   contour plot.  `x` is the meshpoint values across the range of the
 #'   predictor, `Q`; `lag` is the values of `L`; `fitted` and `se.fit` are the

### 2020-10-22 analysis/paper/appendix_density_dependence.Rmd
Exclude saplings, include interaction.

@@
 source(here("R", "crossbasis.R"))
+source(here("R", "utils.R"))
 
@@
 
+-   Remove saplings and update stats
+
 # Methods
@@
 If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value.
-Conspecific density was estimated as the number of H.
+Conspecific density was estimated as the size of H.
 acuminata shoots per m2 in each annulus.
-Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in H.
+<!--# Add analyses using height and shoots + height -->Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in H.
 acuminata [@bruna2002; @brunaEffectsForestFragmentation2002].
@@
 ha$year %>% range()
+ha$code_notes %>% unique()
 ```
@@
   ha %>%
-  ungroup() %>% 
+  filter(code_notes != "sdlg (1)" | is.na(code_notes)) %>% #remove seedlings
   arrange(ha_id_number, year) %>% 
@@
          log_shts_next = log(shts_next),
+         # log_ht = log(ht),
+         # log_ht_next = log(ht_next),
          habitat = as.factor(habitat),
@@
 ```{r}
-g_growth <-
+g_growth_scat <-
   gam(shts_next ~ 
@@
         # s(log_shts, bs = "cr", by = habitat) + #prelim modeling shows this is a linear relationship (edf ~ 1)
-        habitat * log_shts +
+        log_shts *
+        habitat +
         s(Q, L,
@@
           xt = list(bs = c("cr"))),
-      family = gaussian(link = "log"),
+      family = scat(link = "log"),
       data = ha,
@@
 With a penalized spline, the number of knots is not very important as ...
-<!--# finish thought and cite (maybe Colin Edwards?) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat.
+<!--# finish thought and cite (maybe Noam Ross) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat.
 We did not include a random effect of plant ID, as the model did not converge.
@@
 
-```{r}
+```{r eval=FALSE, include=FALSE}
 g_growth_pois <- update(g_growth, family = poisson(link = "log"))
-g_growth_scat <- update(g_growth, family = scat(link = "log")) #because residuals are leptokurtic with gaussian
+g_growth_norm <- update(g_growth, family = gaussian(link = "log")) #residuals are leptokurtic with gaussian
 AICtab(g_growth, g_growth_pois, g_growth_scat)
@@
 
-Number of shoots in the current year had a significant effect on size in year t + 1 (F = 353.8, edf = 5.2, P \< 0.0001).
-The relationship between $log(shoots_{t+1})$ and $log(shoots_t)$ was essentially linear for all but the largest size classes (Figure \@ref(fig:growth-plot)).
-There were no significant differences between habitat types (F = 0.012, df = 1, P = 0.912).
-The cross-basis smooth describing the effects of conspecific density on growth was significant (F = 2.78, edf = 18.4, P \< 0.001).
+Number of shoots in the current year had a significant effect on size in year t + 1 ($\chi^2$ = 796.8, df = 1, P \< 0.0001) (Figure \@ref(fig:growth-plot)).
+There was no significant effect of habitat or the interaction between habitat and size on size in year t + 1.
+The cross-basis smooth describing the effects of conspecific density on growth was significant ($\chi^2$ = 23.2, edf = 6.4, P = 0.004).
 
@@
 Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average.
-The effects of conspecific density seemed strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).
+The effects of conspecific density were strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).
 
@@
   filter(min_dist < 0.1) %>%
-  filter(lag == 1.75) %>% 
+  filter(nearest(lag, 2)) %>% 
   ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
@@
 
-  labs(title = "annulus 1.5–2m", 
+  labs(title = "Near 2m from focal plant", 
        x = "conspecific density (shoots/m^2)",
@@
 
-Plant size had a significant, nearly quadratic, effect on survival (F = 126.59, edf = 2.1, P \< 0.001) (Figure \@ref(fig:surv-plot)).
-However, there were no significant effects of habitat ($\chi^2$ = 0.26, df = 1, P = 0.613) or conspecific density ($\chi^2$ = 126.59, edf = 2, P = 0.074) on survival.
+Plant size had a significant, nearly quadratic (on a logit scale), effect on survival in continuous forest ($\chi^2$ = 112.2, edf = 2.1, P \< 0.001), but a linear (on a logit scale) effect in 1 ha fragments ($\chi^2$ = 15.63, edf = 1, P \< 0.001) (Figure \@ref(fig:surv-plot)).
+However, there were no significant effects of habitat ($\chi^2$ = 0.002, df = 1, P = 0.964) or conspecific density ($\chi^2$ = 5.326, edf = 2, P = 0.0699) on survival.
 
@@
 
-newdata <- tibble(log_shts = seq(0, 2.565, length.out = 100), habitat = "1-ha", plot = "new", Q = Q_new, L = L_new)
+#TODO: make this have both 1-ha and CF
+
+newdata <-
+  tibble(
+    log_shts = rep(seq(0, 2.565, length.out = 100), 2),
+    habitat = rep(c("1-ha", "CF"), each = 100),
+    plot = "new",
+    Q = Q_new,
+    L = L_new
+  )
 
@@
 
-ggplot(fitted, aes(x = exp(log_shts), y = plogis(.fitted))) +
-  geom_line(color = "blue") +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit*1.96), ymax = plogis(.fitted + .se.fit*1.96)), alpha = 0.4) +
+ggplot(fitted, aes(x = exp(log_shts), y = plogis(.fitted), color = habitat)) +
+  geom_line() +
+  geom_ribbon(aes(
+    ymin = plogis(.fitted - .se.fit * 1.96),
+    ymax = plogis(.fitted + .se.fit * 1.96),
+    color = NULL,
+    fill = habitat
+  ), alpha = 0.4)+
   geom_rug(data = rugdat,
-           aes(x = survived),
+           aes(x = survived, color = habitat),
            inherit.aes = FALSE,
@@
   geom_rug(data = rugdat,
-           aes(x = died),
+           aes(x = died, color = habitat),
            inherit.aes = FALSE,
@@
   filter(min_dist < 0.1) %>% 
-  ungroup() %>% 
-  filter(fitted == max(fitted))
-
-pred_growth %>%
-  filter(min_dist < 0.1) %>% 
-  filter(near(x, 5.01, tol = 0.01)) %>% 
+  filter(nearest(x, 5)) %>% 
   ggplot(aes(x = lag, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
@@
       (1|year_post),
-    family = gaussian(link = "log"),
+    family = gaussian(link = "log"), #can't use scat() in glmer()
     data = ha

### 2020-10-27 R/crossbasis.R
allow passing (partial) newdata to cb_margeff

@@ -8,6 +8,9 @@
 #' @param Q The matrix of predictor values used to generate a crossbasis smooth
 #' @param L The matrix of lags used to generate a crossbasis smooth
 #' @param model a gam with a crossbasis smooth
+#' @param ref_data reference data passed to the `newdata` argument of `predict`
+#'   (optional). This one-row data frame should contain values for all model
+#'   terms **except** `Q` and `L`.
 #' @param meshpts vector of length 2; The number of meshpoints for values of Q
 #'   and L, respectively, to use to generate fitted values
 #' @param calc_dist logical; Calculate distance between predicted values and
@@ -30,7 +33,7 @@
 #'
 #' @examples
 cb_margeff <- 
-  function(Q, L, model, meshpts = c(50, 50), calc_dist = TRUE) {
+  function(Q, L, model, ref_data = NULL, meshpts = c(50, 50), calc_dist = TRUE) {
     # Q_name <- quo(Q)
     # L_name <- quo(L)
     
@@ -61,17 +64,19 @@ cb_margeff <-
     
     terms_fac <- names(model$xlevels)
     
-    #TODO newdata columns must be the same class as the model data.  I think this
-    #breaks if there is a fixed-effect factor put in as a character vector.
-    newdata <-
-      df %>%
-      dplyr::summarize(
-        across(c(-!!L_name, -!!Q_name) & where(is.numeric), mean),
-        across(all_of(terms_raneff) & where(is.factor), ~factor(".newdata")),
-        across(all_of(terms_fac) & where(is.factor), ~factor(levels(.x)[1], levels = levels(.x)))
-      )
-    newdata <- uncount(newdata, meshpts[1]) %>% add_column(!!L_name := L_new)
     
+    if (is.null(ref_data)) {
+      #TODO newdata columns must be the same class as the model data.  I think this
+      #breaks if there is a fixed-effect factor put in as a character vector.
+      ref_data <-
+        df %>%
+        dplyr::summarize(
+          across(c(-!!L_name, -!!Q_name) & where(is.numeric), mean),
+          across(all_of(terms_raneff) & where(is.factor), ~factor(".newdata")),
+          across(all_of(terms_fac) & where(is.factor), ~factor(levels(.x)[1], levels = levels(.x)))
+        )
+    }
+    newdata <- uncount(ref_data, meshpts[1]) %>% add_column(!!L_name := L_new)
     resp <- array(dim = c(length(testvals), ncol(Q_new)))
[…]

### 2020-10-27 R/vital rate models.Rmd
modeling vital rates with DLNM

---
title: "FLM for growth as function of SPEI"
author: "Eric R. Scott"
date: "Date Initiated"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(mgcv)
library(dlnm)
library(bbmle)
library(lubridate)
library(glue)
library(patchwork)

library(conflicted)
conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("slice", "dplyr")

source(here("R", "crossbasis.R"))
source(here("R", "utils.R"))
```

*Last compiled: `r Sys.Date()`*

# TODO:

- ~~Take out saplings~~
- Separate models for CF and fragment
- Separate plots (models) for 2 shoots, 5 shoots
- Make presentation that describes DLNM for Maria
- Add random effects

- Start building IPM for subset of Heliconia data (in separate file)

# Purpose

Take a crack at modeling height as a function of previous height and lagged, non-linear, SPEI.

# Load Data

```{r data, echo=TRUE}
spei <- read_csv(here("analysis", "data", "derived_data", "spei_lags.csv"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
```

## Data Dictionary

`spei`: lagged SPEI for each site cluster

- `site` (character): site cluster, one of "colosso_clust", "dimona", "km_clust", or "porto_alegre"
- `census_year` (numeric): the census year to join with.  
- `spei_1` (numeric): SPEI value for January of the census year
- `spei_2` (numeric): SPEI value for December of the previous year
- `spei_3` (numeric): SPEI value for November of the previous year
- `spei_4`-- `spei_24`: ... and so on out to January of 2 years prior.

`demog`: demographic surveys of *Heliconia acuminata*

- `plot` (character): either `bdffp_reserve_no` or continuous forest indicated by "-CF"
- `habitat` (character): either 1-ha, 10-ha fragments or continuous forest "CF"
- `ranch` (character): DIM = Dimona, EST = Estaciao, PAL = Porto Alegre
- `bdffp_reserve_no` (character): Official bdffp reserve number, when there is one.
- `ha_id_number` (character): Unique plant ID
- `tag_number` (character): Plant ID within plot:year.  May have changed over lifetime of plant.
- `row` (character): Location in plot
- `column` (numeric): Location in plot
- `year` (integer): Survey year
- `ht` (numeric): plant height
- `shts` (integer): number of shoots
- `infl` (integer): number of inflorescences
- `code_notes` (character): Important ones: "sdlg (1)" = seedling, "dead (2)" = dead
- `code2` (character): "initial.yr.tag" = first year that plant is in the study.  May not necessarily be a seedling.
- `surv` (integer): did the plant survive to the next year? 1 = survived, 0 = died, NA = last year of data, fate unknown.

# Join data sets

`demog$year` = `spei$census_year`

Getting site to match up will take more work.

```{r}
unique(demog$plot)
unique(demog$bdffp_reserve_no)
unique(spei$site)

demog %>% filter(code_notes ==  "dead (2)", year == 2009)
```

km41 = 1501
km37 = ??
florestal = 1301, Florestal-CF
cabo frio = 3402, CaboFrio-CF
Porto Alegre = 3114, PortoAlegre-CF
Colosso = 1104, 1202
Dimona = 2206, 2108, 2107, Dimona-CF

colosso_clust = 1301, 3402, 1104, 1202, CaboFrio-CF, Florestal-CF

```{r}
demog <- 
  demog %>% 
  mutate(rain_site = case_when(
    bdffp_reserve_no %in% c("1301", "3402", "1104", "1202")      ~ "colosso_clust",
    plot %in% c("CaboFrio-CF", "Florestal-CF")                   ~ "colosso_clust",
    bdffp_reserve_no %in% c("2206", "2108", "2107")              ~ "dimona",
    plot == "Dimona-CF"                                          ~ "dimona",
    bdffp_reserve_no == "3114"                                   ~ "porto_alegre",
    plot == "PortoAlegre-CF"                                     ~ "porto_alegre",
    bdffp_reserve_no == "1501"                                   ~ "km_clust",
    TRUE ~ NA_character_
  ))
```

```{r}
demog_full <-
  left_join(demog, spei, by = c("rain_site" = "site", "year" = "census_year")) %>% 
  mutate(log_shts = log(shts), .after = shts)
```

`gam` wants factors!

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
```

# Splitting up data

To simplify initial modeling attempts, I'll split the data into seedlings vs. post-seedlings and by habitat.  I'll build the Q and L matrixes describing exposure and lag for each subset.

```{r}
demog_post <- 
  demog_full %>% 
  filter(code_notes != "sdlg (1)" | is.na(code_notes))

demog_post_cf <-
  demog_post %>% 
  filter(habitat == "CF")

demog_post_1ha <-
  demog_post %>% 
  filter(habitat == "1-ha")

demog_sdlg <- 
  demog_full %>% 
  filter(code_notes == "sdlg (1)")

demog_sdlg_cf <- 
  demog_sdlg %>% 
  filter(habitat == "CF")

demog_sdlg_1ha <- 
  demog_sdlg %>% 
  filter(habitat == "1-ha")
```

```{r}
Q_post_cf <-
  demog_post_cf %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_post_1ha <-
  demog_post_1ha %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_sdlg_cf <-
  demog_sdlg_cf %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

Q_sdlg_1ha <-
  demog_sdlg_1ha %>% 
  select(starts_with("spei_")) %>% 
  as.matrix()

L <-
  matrix(1:24,
         nrow = nrow(demog_full),
         ncol = 24,
         byrow = TRUE)

colnames(L) <- paste0("L_", 1:24)

L_post_cf <- L[1:nrow(Q_post_cf), ]
L_post_1ha <- L[1:nrow(Q_post_1ha), ]
L_sdlg_cf <- L[1:nrow(Q_sdlg_cf), ]
L_sdlg_1ha <- L[1:nrow(Q_sdlg_1ha), ]
```

## Growth

Start with no random effects

Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.

### Fit GAMs

Fit the GAM with crossbasis smooth for lagged drought

```{r}
g_post_cf <-
  gam(shts_next ~ 
        s(log_shts) + #does this need to be log # shoots?
        # s(shts) +
        s(Q_post_cf, L_post_cf,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
gam.check(g_post_cf) #looks good
# summary(g_post_cf)
anova(g_post_cf)
```

```{r}
g_post_1ha <-
  gam(shts_next ~ 
        s(log_shts) +
        s(Q_post_1ha, L_post_1ha,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
gam.check(g_post_1ha)
anova(g_post_1ha)
```

### Get plot data

This is pretty slow. It could probably be faster if I understood matrix math better.

```{r}
# use mean size plant across both habitats so plots are comparable
ref_data <-
  demog_post %>%
  dplyr::summarize(log_shts = log(mean(shts, na.rm = TRUE))) #log of mean shoots, not mean of log_shts
# exp(1.032814) # 2.8 shoots
```

```{r}
pred_g_post_cf <- cb_margeff(Q_post_cf, L_post_cf, g_post_cf, ref_data = ref_data)
pred_g_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, g_post_1ha, ref_data = ref_data)
```

Lag labels

```{r}
# spei_1 is january
labs <- c("Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
```

Shared color bar limits so scales can be combined.

```{r}
shoot_lim <- c(min(pred_g_post_1ha$fitted, pred_g_post_cf$fitted),
               max(pred_g_post_1ha$fitted, pred_g_post_cf$fitted))
```

### Plots

**Raster plots:**

```{r}
p_growth_cf <-
  pred_g_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
# p_growth_cf
```

```{r}
p_growth_1ha <-
  pred_g_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
# p_growth_1ha
```

**Slice through time at severe drought**

```{r}
p_growth_cf_slice <-
  pred_g_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "Continuous Forest", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_growth_cf_slice
```

```{r}
p_growth_1ha_slice <-
  pred_g_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "shoots (t+1)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_growth_1ha_slice
```

**Slice through individual months**

I find this less useful, but just shows that relationships are slightly quadratic at most complicated.  This would be more useful if there was a way to show which lines had "slopes" significantly different from 0 (i.e. don't totally overlap a flat line).

```{r}
plotdf <-
  pred_g_post_cf %>% 
  mutate(lag = round(lag)) %>% 
  group_by(x, lag) %>% 
  summarize(across(everything(),  mean))

foo <- function(x) {
  d <- 1:24
  names(d)<-labs
  x <- as.numeric(x)
  # s <- glue("lag {x} ({names(d[x])})")
  s <- names(d[x])
  return(s)
}

p_growth_cf_months <-
  plotdf %>% 
  ggplot(aes(x = x, y = exp(fitted))) +
  geom_ribbon(aes(
    ymin = exp(fitted - se.fit * 1.96),
    ymax = exp(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("SPEI3") +
  scale_y_continuous("shoots (t+1)") +
  facet_wrap(~lag, labeller = as_labeller(foo)) +
  theme_bw()
```

## Survival

### Fit GAMs

```{r}
s_post_cf <-
  gam(surv ~ 
        s(log_shts) +
        s(Q_post_cf, L_post_cf,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_cf)
```

```{r}
s_post_1ha <-
  gam(surv ~ 
        s(log_shts) +
        s(Q_post_1ha, L_post_1ha,
          bs = "cb",
          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(s_post_1ha)
```

### Get plot data

This is pretty slow. It could probably be faster if I understood matrix math better.

```{r}
#big plants
# ref_data <- tibble(log_shts = 1.6)

pred_s_post_cf <- cb_margeff(Q_post_cf, L_post_cf, s_post_cf, ref_data = ref_data)
pred_s_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, s_post_1ha, ref_data = ref_data)
```

Shared color bar limits so scales can be combined.

```{r}
surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
# plogis(surv_lim)
```

### Plots

**Raster plots:**

```{r}
p_surv_cf <-
  pred_s_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_surv_cf
```

```{r}
p_surv_1ha <-
  pred_s_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_surv_1ha
```

**Slice through time at severe drought**

```{r}
p_surv_cf_slice <-
  pred_s_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_surv_1ha_slice <-
  pred_s_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 1:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

# Results

## Growth

```{r include=FALSE}
g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])

g_df <-
  bind_rows(
    "CF" = g_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = g_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

I'm not 100% sure this is correct since it's not done with `predict()`

```{r}
g_df %>% 
  ggplot(aes(x = exp(x), y = exp(fit + coef(g_post_cf)[1]))) +
  geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
              aes(x = shts, y = shts_next), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = exp(fit + coef(g_post_cf)[1] - se),
                  ymax = exp(fit + coef(g_post_cf)[1] + se)),
              alpha = 0.2,
              fill = "blue") +
  labs(x = "size (#shoots)", y = "size (t + 1)") +
  facet_wrap(~habitat) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```
Plants don't grow much

```{r}
# anova(g_post_cf)
# anova(g_post_1ha)
```

There's a significant effect of lagged drought on number of shoots in the next year.

```{r}
(p_growth_cf | p_growth_1ha) +
  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
  plot_layout(guides = "collect")
```

It looks like drought in October is bad for growth in both CF and fragments, but drought in June and July is much worse for plants in continuous forest.

Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`Q`).  Might not be entirely comparable?

Here's a slice through SPEI = -1.5 (severe drought)

```{r}
(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
  p_growth_1ha_slice +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
```

Again, you can see that drought immediately preceding the census (Jan of that year) and in the previous October are bad for growth, but dry Decembers and Septembers are good for growth.

## Survival

```{r include=FALSE}
s_cf <- plot.gam(s_post_cf, shift = coef(s_post_cf)[1])
s_1ha <- plot.gam(s_post_1ha, shift = coef(s_post_1ha)[1])

s_df <-
  bind_rows(
    "CF" = s_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = s_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

I'm not 100% sure this is correct since it's not done with `predict()`

```{r}
s_df %>% 
  ggplot(aes(x = exp(x), y = plogis(fit + coef(s_post_cf)[1]))) +
  geom_line() +
  geom_ribbon(aes(ymin = plogis(fit + coef(s_post_cf)[1] - se),
                  ymax = plogis(fit + coef(s_post_cf)[1] + se)),
              alpha = 0.2) +
  labs(x = "size (#shoots)", y = "P(survival)") +
  facet_wrap(~habitat) +
  stat_summary_bin(data = demog_post %>% filter(habitat %in% c("1-ha", "CF")),
                   geom = "point",
                   fun = mean,
                   bins = 7,
                   aes(y = surv, x = shts)) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```

```{r}
# anova(s_post_cf)
# anova(s_post_1ha)
```

Drought also has a significant effect on survival in both habitat types.

```{r}
(p_surv_cf|p_surv_1ha) + 
  plot_annotation(title = "Effect of drought on survival") +
  plot_layout(guides = "collect")
```

I'm doubting things now because this makes it seem like the survival of an averaged sized plant (~2.3 shoots) is higher in fragments than in continuous forest.  I should really use an average sized plant in both habitats combined to get marginal effects.  Either way, the surface for 1-ha fragments is basically flat except for a slight *positive* effect of recent drought on survival.  In continuous forest, drought in the previous February and March has a negative impact on survival.  Wet conditions in the previous April–October have a positive effect on survival.

```{r}
demog_post %>% 
  ggplot(aes(surv)) +
  geom_bar(aes(y = stat(prop))) +
  facet_wrap(~habitat)

demog_post %>% 
  group_by(habitat) %>% 
  summarize(survived = sum(surv == 1, na.rm = TRUE),
            dead = sum(surv == 0, na.rm = TRUE),
            mean_shts = mean(shts, na.rm = TRUE)) %>% 
  mutate(p.surv = survived / (survived + dead))
```

Here's a slice through SPEI = -1.5 (severe drought)

```{r}
(p_surv_cf_slice / p_surv_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
```

### 2020-10-27 analysis/paper/appendix_density_dependence.Rmd
minor formatting changes

@@
 We modeled density dependence using a penalized distributed lag non-linear model (DLNM) approach with the `dlnm` package in R [@gasparriniPenalizedFrameworkDistributed2017; @gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-This approach is similar to that of [-@teller2016], in that it makes no assumptions about the shape of the competition kernel.
+This approach is similar to that of Teller et al.
+[-@teller2016], in that it makes no assumptions about the shape of the competition kernel.
 Instead, it fits a two-dimensional spline allowing for a non-linear response to the intensity of conspecific density as well as the distance of conspecifics.
@@
 Q_new <- 
-  g_growth$model %>% 
+  g_growth_scat$model %>% 
   pull(Q) %>%
@@
 
-fitted <- augment(g_growth, newdata = newdata, se_fit = TRUE, conf.level = 0.95)
+fitted <-
+  augment(
+    g_growth_scat,
+    newdata = newdata,
+    se_fit = TRUE,
+    conf.level = 0.95
+  )
 ggplot(fitted, aes(x = exp(log_shts), y = exp(.fitted))) +
@@
 ```{r}
-pred_growth <- cb_margeff(Q, L, g_growth)
+pred_growth <- cb_margeff(Q, L, g_growth_scat)
 ```
@@
 
-Plant size had a significant, nearly quadratic (on a logit scale), effect on survival in continuous forest ($\chi^2$ = 112.2, edf = 2.1, P \< 0.001), but a linear (on a logit scale) effect in 1 ha fragments ($\chi^2$ = 15.63, edf = 1, P \< 0.001) (Figure \@ref(fig:surv-plot)).
+Plant size had a significant, nearly quadratic (on a logit scale), effect on survival in continuous forest ($\chi^2$ = 112.2, edf = 2.1, P \< 0.001), but a linear (on a logit scale) effect in 1 ha fragments ($\chi^2$ = 15.63, edf = 1, P \< 0.001), but their 95% confidence intervals overlap (Figure \@ref(fig:surv-plot)).
 However, there were no significant effects of habitat ($\chi^2$ = 0.002, df = 1, P = 0.964) or conspecific density ($\chi^2$ = 5.326, edf = 2, P = 0.0699) on survival.

### 2020-10-28 R/competition wrangling.Rmd
revised row and column locations and re-ran density dependence analysis.

@@
   future_map(~{
-    add_density_cols(.x, L = 0, U = 3, by = 3)
+    add_density_cols(.x, L = 0, U = 1, by = 1) #first meter seems most important
   }) %>% 

### 2020-10-28 R/vital rate models.Rmd
revised row and column locations and re-ran density dependence analysis.

@@
                    fun = mean,
-                   bins = 7,
+                   bins = 20,
                    aes(y = surv, x = shts)) +
@@
   plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
+
+demog_full %>% ggplot(aes(shts)) + geom_histogram(bins = 20)
+hist(demog_full$shts)
+```
+```{r}
+demog_full %>% 
+  filter(shts >= 15)
+demog_full %>% 
+  filter(ha_id_number == 5310) %>% View()
 ```

### 2020-10-28 analysis/paper/appendix_density_dependence.Rmd
revised row and column locations and re-ran density dependence analysis.

@@
 
--   Remove saplings and update stats
+-   Update stats in writing
 
@@
 ```{r eval=FALSE, include=FALSE}
-g_growth_pois <- update(g_growth, family = poisson(link = "log"))
-g_growth_norm <- update(g_growth, family = gaussian(link = "log")) #residuals are leptokurtic with gaussian
-AICtab(g_growth, g_growth_pois, g_growth_scat)
+g_growth_pois <- update(g_growth_scat, family = poisson(link = "log"))
+g_growth_norm <- update(g_growth_scat, family = gaussian(link = "log")) #residuals are leptokurtic with gaussian
+AICtab(g_growth_norm, g_growth_pois, g_growth_scat)
 ```
@@
 Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average.
-The effects of conspecific density were strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).
+The effects of conspecific density were strongest around 0--1 meters from the focal plant (Figure \@ref(fig:growth-cb)).
 
@@
   filter(min_dist < 0.1) %>%
-  filter(nearest(lag, 2)) %>% 
+  filter(nearest(lag, 0.5)) %>% 
   ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
@@
                 filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
-              aes(x = shts_1.5_2, y = shts_next),
+              aes(x = shts_0_0.5, y = shts_next),
               alpha = 0.2,
@@
 
-  labs(title = "Near 2m from focal plant", 
+  labs(title = "Near 0.5m from focal plant", 
        x = "conspecific density (shoots/m^2)",
@@
 Plant size had a significant, nearly quadratic (on a logit scale), effect on survival in continuous forest ($\chi^2$ = 112.2, edf = 2.1, P \< 0.001), but a linear (on a logit scale) effect in 1 ha fragments ($\chi^2$ = 15.63, edf = 1, P \< 0.001), but their 95% confidence intervals overlap (Figure \@ref(fig:surv-plot)).
-However, there were no significant effects of habitat ($\chi^2$ = 0.002, df = 1, P = 0.964) or conspecific density ($\chi^2$ = 5.326, edf = 2, P = 0.0699) on survival.
+However, there were no significant effects of habitat ($\chi^2$ = 0.002, df = 1, P = 0.964) or conspecific density ($\chi^2$ = 5.36, edf = 2, P = 0.0688) on survival.
 
@@
 
-```{r eval=FALSE, include=FALSE}
-pred_growth %>%
-  filter(min_dist < 0.1) %>% 
-  filter(nearest(x, 5)) %>% 
-  ggplot(aes(x = lag, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
-  # geom_jitter(data = ha %>%
-  #               filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
-  #             aes(x = shts_1.5_2, y = shts_next),
-  #             alpha = 0.2,
-  #             inherit.aes = FALSE) +
-  # geom_smooth(color = "blue", se = FALSE) + 
-  geom_line(color = "blue") +
-  geom_ribbon(alpha = 0.4) +
-
-  labs(title = "Conspecific density of 5 shoots/m^2", 
-       x = "Midpoint distance from focal plant (m)",
-       y = "#shoots (t+1)") +
-  theme_bw()
-```
-
 ```{r}
@@
       habitat *
-      shts_0_3 +
+      shts_0_1 +
       (1|ha_id_number) +
@@
 # Interactions
-mnorm5 <- update(mnorm4, .~. - log_shts:habitat:shts_0_3)
+mnorm5 <- update(mnorm4, .~. - log_shts:habitat:shts_0_1)
 AICtab(mnorm4, mnorm5) #good to remove 3-way interaction
 mnorm6 <- update(mnorm5, .~. -log_shts:habitat)
-mnorm7 <- update(mnorm5, .~. -log_shts:shts_0_3)
-mnorm8 <- update(mnorm5, .~. -habitat:shts_0_3)
-mnorm9 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:shts_0_3)
-mnorm10 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:habitat)
-mnorm11 <- update(mnorm5, .~. -habitat:shts_0_3 -log_shts:habitat -log_shts:shts_0_3)
-AICtab(mnorm5, mnorm6, mnorm7, mnorm8, mnorm9, mnorm10, mnorm11) #good to remove all interactions
+mnorm7 <- update(mnorm5, .~. -log_shts:shts_0_1)
+mnorm8 <- update(mnorm5, .~. -habitat:shts_0_1)
+mnorm9 <- update(mnorm5, .~. -habitat:shts_0_1 -log_shts:shts_0_1)
+mnorm10 <- update(mnorm5, .~. -habitat:shts_0_1 -log_shts:habitat)
+mnorm11 <- update(mnorm5, .~. -habitat:shts_0_1 -log_shts:habitat -log_shts:shts_0_1)
+AICtab(mnorm5, mnorm6, mnorm7, mnorm8, mnorm9, mnorm10, mnorm11) #good to remove all interactions except size by density
 
-Anova(mnorm11)
-fixef(mnorm11)
+Anova(mnorm10)
+fixef(mnorm10)
 ```
@@
 **No longer significant**
+
+marginally significant size x conspecific density interaction (p = 0.046). No significant main effect of density.
 
@@
 
-ha_sub <- ha %>% filter(!is.na(shts_0_3), !is.na(log_shts))
+ha_sub <- ha %>% filter(!is.na(shts_0_1), !is.na(log_shts))
 newdata <-
@@
          ha_id_number = "new",
-         shts_0_3 = seq(min(ha_sub$shts_0_3, na.rm = TRUE),
-                          max(ha_sub$shts_0_3, na.rm = TRUE),
+         shts_0_1 = seq(min(ha_sub$shts_0_1, na.rm = TRUE),
+                          max(ha_sub$shts_0_1, na.rm = TRUE),
                           length.out = 50))
@@
 augment(mnorm11, newdata = newdata, re.form = NA) %>% 
-  ggplot(aes(x = shts_0_3, y = exp(.fitted))) +
-  geom_jitter(data = ha_sub, aes(x = shts_0_3, y = shts_next), alpha = 0.2) +
+  ggplot(aes(x = shts_0_1, y = exp(.fitted))) +
+  geom_jitter(data = ha_sub, aes(x = shts_0_1, y = shts_next), alpha = 0.2) +
   geom_line(color = "blue") +

### 2020-10-30 R/vital rate models.Rmd
updating text and cleaning up code.

--- 2020-10-27 R/vital rate models.Rmd
@@
 
-- ~~Take out saplings~~
-- Separate models for CF and fragment
-- Separate plots (models) for 2 shoots, 5 shoots
-- Make presentation that describes DLNM for Maria
-- Add random effects
+- ~~Remove seedlings from dataset, for now~~
+- ~~Separate models for CF and fragment, for now~~
+- Add model for probability of reproduction and other vital rates
+- Model seedlings separately?
+- Add random effects, interactions, other complexities
+- Explore FGAM as alternative to DLNM (seems like they are doing the same thing, but maybe FGAM allows modeling of interactions?)
 
@@
 
-Take a crack at modeling height as a function of previous height and lagged, non-linear, SPEI.
+Take a crack at modeling vital rates as a function of previous height and lagged, non-linear, SPEI.
 
@@
 ```
-Plants don't grow much
+Plants don't grow much.
 
 ```{r}
-# anova(g_post_cf)
-# anova(g_post_1ha)
+anova(g_post_cf)
+anova(g_post_1ha)
 ```
@@
 ```{r}
-# anova(s_post_cf)
-# anova(s_post_1ha)
+anova(s_post_cf)
+anova(s_post_1ha)
 ```
@@
 
-I'm doubting things now because this makes it seem like the survival of an averaged sized plant (~2.3 shoots) is higher in fragments than in continuous forest.
-I should really use an average sized plant in both habitats combined to get marginal effects.
-Either way, the surface for 1-ha fragments is basically flat except for a slight *positive* effect of recent drought on survival.
+The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.
 In continuous forest, drought in the previous February and March has a negative impact on survival.
-Wet conditions in the previous April–October have a positive effect on survival.
-
-```{r}
-demog_post %>% 
-  ggplot(aes(surv)) +
-  geom_bar(aes(y = stat(prop))) +
-  facet_wrap(~habitat)
-
-demog_post %>% 
-  group_by(habitat) %>% 
-  summarize(survived = sum(surv == 1, na.rm = TRUE),
-            dead = sum(surv == 0, na.rm = TRUE),
-            mean_shts = mean(shts, na.rm = TRUE)) %>% 
-  mutate(p.surv = survived / (survived + dead))
-```
+Wet conditions or very dry conditions in the previous August–October have a positive effect on survival.
 

### 2020-11-09 README.Rmd
Trying to make this no longer behave like an R package.  Might break things.

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# HeliconiaDemography

<!-- badges: start -->
[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio)
[![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/run-tests/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
<!-- badges: end -->

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

The simplest way to explore the text, code and data is to click on
[binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio)
to open an instance of RStudio in your browser, which will have the
compendium files ready to work with. Binder uses rocker-project.org
Docker images to ensure a consistent and reproducible computational
environment. These Docker images can also be used locally.

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### 2020-11-09 README.Rmd
Revert "Trying to make this no longer behave like an R package.  Might break things."
This reverts commit 32734dcbc63e2e1005f6aac130f5329c997f1c55.

@@
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio)
-[![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/run-tests/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
+[![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/R-CMD-check/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
 <!-- badges: end -->

### 2020-11-09 tests/testthat.R
got tests working, wrote tests for utils.

@@ -1,3 +1,4 @@
 library(testthat)
-
-test_dir("tests/testthat", reporter = c("check", "progress"))
+library(HeliconiaDemography)
+# test_dir("tests/testthat", reporter = c("check", "progress"))
+test_check("HeliconiaDemography")

### 2020-11-09 tests/testthat/test-utils.R

@@ -1,5 +1,5 @@
-library(here)
-source(here("R", "utils.R"))
+# library(here)
+# source(here("R", "utils.R"))
 test_that("matrix iteration works", {
   A <- matrix(1:9, nrow = 3)
   expect_equal(A %*% A, A %^% 2)
@@ -7,3 +7,14 @@ test_that("matrix iteration works", {
 })
 
 
+test_that("eu_dist() works", {
+  x = c(0,0)
+  y = c(3,3)
+  expect_equal(eu_dist(x, y), sqrt(3^2 + 3^2))
+})
+
+test_that("nearest() helper works", {
+  x <- seq(1, 10, 0.3)
+  expect_equal(which(nearest(x, 5)), 14)
+  expect_equal(x[nearest(x, 5)], 4.9)
+})

### 2020-11-09 analysis/supplementary-materials/appendix_density_dependence.Rmd
reorganized files to make R CMD check happier

--- 2020-10-28 analysis/paper/appendix_density_dependence.Rmd
@@
   comment = "#>",
-  fig.path = "../figures/",
+  fig.path = "",
   dpi = 300)
@@
 
-source(here("R", "crossbasis.R"))
-source(here("R", "utils.R"))
+library(HeliconiaDemography)
 

### 2020-11-09 R/crossbasis.R
added examples and documented with roxygen2

@@ -32,6 +32,21 @@
 #' @import tidyr
 #'
 #' @examples
+#' \dontrun{
+#' library(dlnm)
+#' library(mgcv)
+#' library(tsModel)
+#' data("chicagoNMMAPS")
+#' Q <- Lag(chicagoNMMAPS$temp, 0:25) #temperature data, lagged
+#' L <- matrix(0:25,nrow(Q),ncol(Q),byrow=TRUE) #matrix of 0-25
+#' # Fit DLNM model
+#' gam1 <- gam(death ~ s(Q, L, bs="cb", k=10) + s(pm10) + dow,
+#'             family=quasipoisson(), 
+#'             data = chicagoNMMAPS,
+#'             method='REML')
+#' # Calculate marginal effect of lagged temperature, all else being held average.
+#' cb_margeff(Q, L, gam1)
+#' }
 cb_margeff <- 
   function(Q, L, model, ref_data = NULL, meshpts = c(50, 50), calc_dist = TRUE) {
     # Q_name <- quo(Q)
@@ -118,7 +133,7 @@ cb_margeff <-
       ) %>%
       dplyr::mutate(lag = as.double(lag), x = as.double(x))
     
-    pred <- full_join(fitted, se.fitted, by = c("x", "lag"))
+    pred <- dplyr::full_join(fitted, se.fitted, by = c("x", "lag"))
     if (isTRUE(calc_dist)) {
       out <- add_min_dist(df, Q_name, L, pred)
     } else {

### 2020-11-09 R/utils.R

@@ -1,14 +1,19 @@
 #' Matrix iteration operator
-#'
+#' 
+#' A wrapper around [matrixcalc::matrix.power]. A%^%k multiplies matrix A by itself k times.
+#' 
+#' @rdname matrix_power_operator
 #' @param x a matrix
 #' @param k a power
 #' 
 #' @importFrom matrixcalc matrix.power
-#' @return
+#' @return a matrix
 #' @export
 #'
 #' @examples
 #' A <- matrix(1:9, nrow = 3)
+#' B <- A %^% 3
+#' B == A %*% A %*% A
 `%^%` <- function(x, k) {
   matrixcalc::matrix.power(x, k)
 }
@@ -42,6 +47,8 @@ eu_dist <- function(p1, p2) {
 #' @export
 #'
 #' @examples
+#' x <- seq(1, 10, 0.3)
+#' nearest(x, 5)
 nearest <- function(x, val) {
   abs(x - val) == min(abs(x - val))
 }

### 2020-11-09 README.Rmd
edits to README

--- 2020-11-09 README.Rmd
@@
 <!-- badges: start -->
+
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) 
 [![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/run-tests/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
+
 <!-- badges: end -->
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Authors, (YYYY). *`r Title`*. Name of journal/book <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Authors, (`r format(Sys.Date(), "%Y")`). *Compendium of R code and data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
+The **R** directory contains functions used in the analysis.
+
 The **analysis** directory contains:
 
-  - [:file\_folder: paper](/analysis/paper): R Markdown source document
-    for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+-   [:file\_folder: wrangling](/analysis/wrangling): .R and .Rmd files with code related to data wrangling (i.e. converting raw data into cleaned data)
+-   [:file\_folder: paper](/analysis/paper): R Markdown source document for manuscript.
+Includes code to reproduce the figures and tables generated by the analysis.
+It also has a rendered version, `paper.docx`, suitable for reading (the code is replaced by figures and tables in this file)
 -   [:file\_folder: data](/analysis/data): Data used in the analysis.
-  - [:file\_folder: figures](/analysis/figures): Plots and other
-    illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
+-   [:file\_folder: figures](/analysis/figures): Plots and other illustrations
+-   [:file\_folder: supplementary-materials](/analysis/supplementary-materials): Supplementary materials including notes and other documents prepared and collected during the analysis.
 
-## How to run in your broswer or download and run locally
+## How to run in your browser or download and run locally
 
-This research compendium has been developed using the statistical programming
-language R. To work with the compendium, you will need
-installed on your computer the [R software](https://cloud.r-project.org/)
-itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
+This research compendium has been developed using the statistical programming language R.
+To work with the compendium, you will need installed on your computer the [R software](https://cloud.r-project.org/) itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
 
-The simplest way to explore the text, code and data is to click on
-[binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio)
-to open an instance of RStudio in your browser, which will have the
-compendium files ready to work with. Binder uses rocker-project.org
-Docker images to ensure a consistent and reproducible computational
-environment. These Docker images can also be used locally.
+The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
+Binder uses rocker-project.org Docker images to ensure a consistent and reproducible computational environment.
+These Docker images can also be used locally.
 
-You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+You can download the compendium as a zip from from this URL: [master.zip](/archive/master.zip). After unzipping:
+
 -   open the `.Rproj` file in RStudio
-- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
-[DESCRIPTION](/DESCRIPTION) file).
+-   run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
 -   finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### 2020-11-10 R/crossbasis.R
fixed startup warnings

@@ -25,7 +25,7 @@
 #'   results of `predict.gam()`; min_dist is the euclidean distance on the unit
 #'   square from the fitted values to the actual data used to fit the model.
 #' @export
-#' @import rlang
+#' @importFrom rlang enquo
 #' @import purrr
 #' @import dplyr
 #' @import mgcv
@@ -152,6 +152,8 @@ cb_margeff <-
 #' @param Q_name quosure; the name for the Q matrix
 #' @param L matrix; the actual L matrix
 #' @param pred data frame; the predicted values
+#' @import purrr
+#' 
 #'
 #' @return a tibble
 add_min_dist <- function(df, Q_name, L, pred) {
@@ -159,7 +161,7 @@ add_min_dist <- function(df, Q_name, L, pred) {
     df %>%
     pull(!!Q_name) %>% 
     as_tibble() %>% 
-    set_names(L[1, ]) %>% 
+    purrr::set_names(L[1, ]) %>% 
     pivot_longer(everything(),
                  names_to = "lag",
                  names_transform = list(lag = as.numeric),

### 2020-11-10 R/utils.R
added plotting function for SPEI rectangles

@@ -52,3 +52,28 @@ eu_dist <- function(p1, p2) {
 nearest <- function(x, val) {
   abs(x - val) == min(abs(x - val))
 }
+
+
+#' Add SPI or SPEI indicators to a ggplot
+#'
+#' @param p a ggplot object
+#' @param ... other arguments passed to [annotate()], e.g. `alpha`
+#'
+#' @return a ggplot object
+#' @export
+#'
+#' @examples
+annotate_spei <- function(p, ...) {
+  spei_rects <- list(
+    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -4.2, ymax = -2, fill = "#e31a1c", ...),
+    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -2, ymax = -1.5, fill = "#fd8d3c", ...),
+    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -1.5, ymax = -1, fill = "#fecc5c", ...),
+    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -1, ymax = 0, fill = "#ffffb2", ...) 
+  )
+  #add rectangles
+  P2 <- p + spei_rects
+  #now, move the last 4 layers to the bottom
+  nlayers <- length(P2$layers)
+  P2$layers <- c(P2$layers[(nlayers-3):nlayers], P2$layers[1:(nlayers-4)])
+  return(P2)
+}

### 2020-11-10 analysis/paper/vital rate models.Rmd
reorganized files to make R CMD check happier
added Xavier gridded data SPEI.

--- 2020-10-30 R/vital rate models.Rmd
@@
 library(patchwork)
+library(sf)
 
@@
 
-source(here("R", "crossbasis.R"))
-source(here("R", "utils.R"))
+# library(HeliconiaDemography) #load functions in this compendium
+load_all()
 ```
@@
 
+- 
 - ~~Remove seedlings from dataset, for now~~
@@
 - Add random effects, interactions, other complexities
-- Explore FGAM as alternative to DLNM (seems like they are doing the same thing, but maybe FGAM allows modeling of interactions?)
-
-- Start building IPM for subset of Heliconia data (in separate file)
+- Match SPEI data to demographic plots programtically by lat lon data
 
@@
 spei <- read_csv(here("analysis", "data", "derived_data", "spei_lags.csv"))
+xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
 demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
+plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
+```
+# Join data sets
+
+## Gridded data
+Get only rows for february (the month of the census), and year for joining
+
+```{r}
+xa2 <- 
+  xa %>%
+  filter(month(yearmonth) == 2) %>% 
+  mutate(year = year(yearmonth), .after =yearmonth) %>% 
+  separate(latlon, into = c("lat", "lon"), sep = "_")
+
 ```
 
-## Data Dictionary
+Now, for the demography data, get year and plot coords
 
-`spei`: lagged SPEI for each site cluster
+```{r}
+# demog$plot %>% unique() %>% sort()
+# plot_coords$plot %>% unique() %>% sort()
+demog <-
+  right_join(plot_coords %>% select(plot, lon, lat),
+             demog,
+             by = "plot") %>% 
+  select(-x_09, -y_09) #remove to avoid confusion since i'm not using here.
+```
 
-- `site` (character): site cluster, one of "colosso_clust", "dimona", "km_clust", or "porto_alegre"
-- `census_year` (numeric): the census year to join with.  
-- `spei_1` (numeric): SPEI value for January of the census year
-- `spei_2` (numeric): SPEI value for December of the previous year
-- `spei_3` (numeric): SPEI value for November of the previous year
-- `spei_4`-- `spei_24`: ... and so on out to January of 2 years prior.
+Now join by year and whether plot is in grid cell or not.
+Probably need to construct `sf` object for grid cells (knowing that the lat lon are the centers and the resolution is 0.25º) and then use some `sf_` function to check if a point is inside of a grid cell.
+See: https://stackoverflow.com/questions/59766153/left-join-based-on-closest-lat-lon-in-r
 
-`demog`: demographic surveys of *Heliconia acuminata*
-
-- `plot` (character): either `bdffp_reserve_no` or continuous forest indicated by "-CF"
-- `habitat` (character): either 1-ha, 10-ha fragments or continuous forest "CF"
-- `ranch` (character): DIM = Dimona, EST = Estaciao, PAL = Porto Alegre
-- `bdffp_reserve_no` (character): Official bdffp reserve number, when there is one.
-- `ha_id_number` (character): Unique plant ID
-- `tag_number` (character): Plant ID within plot:year.  May have changed over lifetime of plant.
-- `row` (character): Location in plot
-- `column` (numeric): Location in plot
-- `year` (integer): Survey year
-- `ht` (numeric): plant height
-- `shts` (integer): number of shoots
-- `infl` (integer): number of inflorescences
-- `code_notes` (character): Important ones: "sdlg (1)" = seedling, "dead (2)" = dead
-- `code2` (character): "initial.yr.tag" = first year that plant is in the study.  May not necessarily be a seedling.
-- `surv` (integer): did the plant survive to the next year? 1 = survived, 0 = died, NA = last year of data, fate unknown.
-
-# Join data sets
+## Imputed data:
 

### 2020-11-10 analysis/wrangling/weather_wrangling.R

@@ -19,6 +19,9 @@ df_spei <-
          yearmonth = yearmonth(date)) %>% 
   select(date, year, month, yearmonth, site, spei)
 
+# An alternative way of doing lags that might be useful in the future.
+# df_spei %>% 
+#   mutate(Q = tsModel::Lag(spei, 1:24))
 
 df_wide <- 
   df_spei %>% 
@@ -39,3 +42,32 @@ spei_lag
 #spei_1 is Jan, spei_2, is Dec of the previous year, spei_3 is Nov of the previous year, etc.
 
 write_csv(spei_lag, here("analysis", "data", "derived_data", "spei_lags.csv"))
+
+# Repeat with gridded data from Xavier et al.
+xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
+library(SPEI)
+library(tsModel)
+xa_spei <-
+  xa %>% 
+  unite(latlon, lat, lon) %>% 
+  mutate(yearmonth = yearmonth(date)) %>% 
+  group_by(latlon, yearmonth) %>% 
+  summarize(precip = sum(precip), eto = sum(eto)) %>% 
+  mutate(cb = precip - eto) %>% 
+  as_tsibble(index = yearmonth, key = latlon) %>% 
+  mutate(spei = as.numeric(spei(cb, scale = 3)$fitted))
+
+
+xa_spei_lags <-
+  xa_spei %>%
+  mutate(Q_spei = Lag(spei, 1:24)) 
+
+# p <-
+#   ggplot(xa_spei_lags, aes(x = yearmonth, y = spei, color = latlon)) +
+#   geom_line() +
+#   facet_wrap(~year(yearmonth), scales = "free_x")
+
+# annotate_spei(p)  
+
+write_rds(xa_spei_lags, here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
+read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))

### 2020-11-10 README.Rmd
switched GH action to a more standard one for packages.

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) 
-[![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/run-tests/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
-
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) [![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/R-CMD-check/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
 <!-- badges: end -->

### 2020-11-12 analysis/paper/vital rate models.Rmd
updated to work with new SPEI history matrix-dataframes

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
 library(tidyverse)
@@
 library(sf)
+library(tsibble)
 
@@
 # library(HeliconiaDemography) #load functions in this compendium
-load_all()
+# devtools::load_all()
 ```
@@
 ```{r data, echo=TRUE}
-spei <- read_csv(here("analysis", "data", "derived_data", "spei_lags.csv"))
+spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
 xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
@@
   separate(latlon, into = c("lat", "lon"), sep = "_")
-
+xa2
 ```
@@
 unique(spei$site)
-
-demog %>% filter(code_notes ==  "dead (2)", year == 2009)
 ```
@@
 
+Get SPEI history only for census months (assuming all censuses done in February)
+
+```{r}
+spei_census <- 
+  spei %>% 
+  filter(month(yearmonth) == 2) %>% 
+  mutate(year = year(yearmonth), .before = site) %>% 
+  select(-yearmonth)
+```
+
+Join
+
 ```{r}
 demog_full <-
-  left_join(demog, spei, by = c("rain_site" = "site", "year" = "census_year")) %>% 
+  left_join(demog, spei_census, by = c("rain_site" = "site", "year")) %>% 
   mutate(log_shts = log(shts), .after = shts)
+
+# demog_full %>% select(spei, spei_history)
 ```
@@
 
-```{r}
-Q_post_cf <-
-  demog_post_cf %>% 
-  select(starts_with("spei_")) %>% 
-  as.matrix()
-
-Q_post_1ha <-
-  demog_post_1ha %>% 
-  select(starts_with("spei_")) %>% 
-  as.matrix()
-
-Q_sdlg_cf <-
-  demog_sdlg_cf %>% 
-  select(starts_with("spei_")) %>% 
-  as.matrix()
-
-Q_sdlg_1ha <-
-  demog_sdlg_1ha %>% 
-  select(starts_with("spei_")) %>% 
-  as.matrix()
-
-L <-
-  matrix(1:24,
-         nrow = nrow(demog_full),
-         ncol = 24,
-         byrow = TRUE)
-
-colnames(L) <- paste0("L_", 1:24)
-
-L_post_cf <- L[1:nrow(Q_post_cf), ]
-L_post_1ha <- L[1:nrow(Q_post_1ha), ]
-L_sdlg_cf <- L[1:nrow(Q_sdlg_cf), ]
-L_sdlg_1ha <- L[1:nrow(Q_sdlg_1ha), ]
-```
-
 ## Growth
@@
   gam(shts_next ~ 
-        s(log_shts) + #does this need to be log # shoots?
+        s(log_shts) + #does this *need* to be log # shoots?
         # s(shts) +
-        s(Q_post_cf, L_post_cf,
+        s(spei_history, L,
           bs = "cb",
-          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
           xt = list(bs = "cr")),
@@
         s(log_shts) +
-        s(Q_post_1ha, L_post_1ha,
+        s(spei_history, L,
           bs = "cb",
-          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
           xt = list(bs = "cr")),
@@
 ```{r}
-pred_g_post_cf <- cb_margeff(Q_post_cf, L_post_cf, g_post_cf, ref_data = ref_data)
-pred_g_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, g_post_1ha, ref_data = ref_data)
+pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data)
+pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
 ```
@@
 ```{r}
-# spei_1 is january
-labs <- c("Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
+# first column is February, the month of the census
+labs <- c("Feb", "Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
 ```
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 1:24, labels = labs, expand = c(0,0)) +
+  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
@@
   theme_bw()
-# p_growth_cf
+p_growth_cf
 ```
@@
   theme_bw()
-# p_growth_1ha
+p_growth_1ha
 ```
@@
   geom_line() +
-  scale_x_continuous("", breaks = 1:24, labels = labs) +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
   labs(title = "Continuous Forest", y = "shoots (t+1)") +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p_growth_cf_slice
+p_growth_cf_slice
 ```
@@
   geom_line() +
-  scale_x_continuous("", breaks = 1:24, labels = labs) +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
   labs(title = "1-ha Fragments", y = "shoots (t+1)") +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p_growth_1ha_slice
+p_growth_1ha_slice
 ```
@@
 foo <- function(x) {
-  d <- 1:24
+  d <- 1:25
   names(d)<-labs
@@
         s(log_shts) +
-        s(Q_post_cf, L_post_cf,
+        s(spei_history, L,
           bs = "cb",
-          k = c(3, 24),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
           xt = list(bs = "cr")),
@@
 ```{r}
-#big plants
-# ref_data <- tibble(log_shts = 1.6)
-
-pred_s_post_cf <- cb_margeff(Q_post_cf, L_post_cf, s_post_cf, ref_data = ref_data)
-pred_s_post_1ha <- cb_margeff(Q_post_1ha, L_post_1ha, s_post_1ha, ref_data = ref_data)
+pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
+pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
 ```
@@
   geom_line() +
-  scale_x_continuous("", breaks = 1:24, labels = labs) +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
   labs(title = "Continuous Forest", y = "P(survival)") +
@@
   geom_line() +
-  scale_x_continuous("", breaks = 1:24, labels = labs) +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
   labs(title = "1-ha Fragments", y = "P(survival)") +
@@
 ```{r include=FALSE}
+#TODO: switch to predict() so y axis is correct
 g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
@@
 
-It looks like drought in October is bad for growth in both CF and fragments, but drought in June and July is much worse for plants in continuous forest.
+It looks like drought in October is bad for growth in fragments, but drought in June and July is much worse for plants in continuous forest.
 
-Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`Q`).  Might not be entirely comparable?
+Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`spei_history`).
+Might not be entirely comparable?
+Different penalties to the smooths make these hard to compare as well.
 

### 2020-11-12 analysis/paper/paper.Rmd
updated citation style and worked on outline

@@
 ---
-title: "Title Goes Here"
+title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
 author:
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/ecology.csl" # Insert path for the bib-style
 abstract: |
@@
 keywords: |
-  density dependence; vital rates; *Heliconia acuminata*
+  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
 highlights: |
@@
 
+## Habitat fragmentation
+
+Habitat fragmentation is bad, pervasive, and getting worse
+
+-   stats about global land use change <!--# start this broad or start with the Amazon? -->
+
+-   stats about amazon deforestation [@malhi2008]
+
+-   stats about how much land is near an edge (refs in Schwartz et al and betts 2019)
+
+-   Why fragmentation bad
+
+    -   microclimate changes [@ewersFragmentationImpairsMicroclimate2013]
+
+    -   metapopulation and connectivity stuff (necessary?)
+
+-   Plants in the tropics are less well adapted to fragmentation because lack of historical distrubance and less variable climate [@betts2019]
+
+## Climate change and drought
+
+-   The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
+
+-   The El Niño--Southern Oscillation (ENSO) reduces wet-season rainfall in the Amazon <!--# is this correct?
+Or only in the south?
+-->.
+ENSO-induced droughts are predicted to increase in frequency and severity.
+[@caiIncreasingFrequencyExtreme2014].
+
+-   Also sea surface temperature related drought.
+Warming of north Atlantic results longer and more intense dry season in southern and eastern Amazonia [@phillipsDroughtSensitivityAmazon2009; @malhi2008]
+
+-   drought effects on plant vital rates
+
+-   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]
+
+-   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]
+
+-   2010 drought [@lewis2010AmazonDrought2011]
+
+-   [@zengCausesImpacts20052008]
+
+## Interactions
+
+-   However, our understanding of how climate change and habitat fragmentation might interact to impact organisms is limited.
+
+-   [@schwartz2019] (mention how tree size classes affected differently, but we don't know if this is also true for understory herbaceous plants.
+
+-   [@malhi2008]
+
+-   The 1997--1998 ENSO drought caused greater tree mortality and increased leaf litter fall in experimentally fragmented forests compared to intact forest in the Amazon [@laurancePositiveFeedbacksForest2001].
+
+-   Furthermore, the interactive effects of fragmentation and drought may be difficult to detect in long-lived species (e.g.
+Heliconia) [bruna paper that shows growth is reduced after 10 years slightly, but impacts reproduction]
+
+## This experiment
+
+-   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments since fragments less buffered from changes in humidity and more sensitive to desiccation.
+Under this hypothesis we would expect drought to reduce growth, survival, reproduction and recruitment more in fragmented landscapes compared to intact forest.
+
+-   It's also possible that plants in fragments have already been filtered such that the survivors are already adapted to more variable and desiccating conditions associated with forest edges [@schwartz2019].
+This may result in drought having a less noticeable impact on plants in fragments, although we would expect fragmentation to still have an impact on survival, growth, and reproduction.
+
+-   We set out to answer the following questions: 1) Does drought have different impacts on plants in forest fragments compared to intact forest?
+2) Do the magnitudes and directions of these effects differ among vital rates?
+3) Do the effects of drought, fragmentation, and their interaction vary with plant size [@tredennickSizebyenvironmentInteractionsNeglected2018] <!--# read this referrebce --> ?
+
 # Background
@@
 -->.
-Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011].
+Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011; @brunaHeliconiaAcuminataSeed2014].
 
@@
 
-## Data collection
-
-The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years <!--# methods for collecting these data?
--->.
-In 2006 and 2007, surveys were conducted to \_\_\_\_\_\_\_\_\_ .
-In 2009 a similar survey was conducted to assess the effects of safe-site limitation on recruitment [@uriarte2010].
-For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al.
-[-@teller2016].
-Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e.
-a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m.
-If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value.
-Conspecific density was estimated as the number of *H.
-acuminata* shoots per m^2^ in each annulus.
-Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in *H.
-acuminata* [@bruna2002; @brunaEffectsForestFragmentation2002].
-
-We used the three transition years for which we had density data to model the effects of conspecific density on survival, growth, and reproduction of *H.
-acuminata*.
-The collection of demographic data is described elsewhere [DATA PAPER], but briefly...
-
 # Results
-
-<!-- Here's some example analysis code: -->
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
-
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
-```
-
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
-
-```{r demo-inline-code}
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
 
@@
 
-# Acknowledgements
-
-<!-- The following line inserts a page break  -->
-
-\newpage
+# Acknowledgments
 

### 2020-11-12 analysis/supplementary-materials/appendix_density_dependence.Rmd
updated citation style and worked on outline

@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/ecology.csl" # Insert path for the bib-style
 ---
@@
 -   Update stats in writing
+
+-   Cite [@bruna2004] (reproductive success is independent of local floral density. Not sure if scales are comparable though, need to read paper first)
 

### 2020-11-12 analysis/paper/paper.Rmd
refined outline

@@
 
-## Habitat fragmentation
+1.  Little is known about how organisms respond to multiple anthropogenic stresses simultaneously <!--# (see Lakeman-Fraser and Ewers 2014)  -->
+
+2.  The Amazon Basin, an important place, has been fragmented, and although deforestation has slowed in recent years, it is continuing.
+
+    1.  fragmentation is bad because...
+
+3.  Simultaneously, there is a drying trend in the Amazon and also increasing intensity and frequency of droughts caused by ENSO (wet-season) and sea surface warming (dry-season).
+
+    1.  Drought is bad because...
+
+4.  The effects of drought and fragmentation could be additive, or interactive.
+
+    1.  The effects of drought could be more severe in fragments because of reduced microclimate buffering.
+
+    2.
+Alternatively, plants in fragmented habitats could be more resilient to drought because the individuals surviving in fragmented forests are already resilient in the face of a relatively dry and variable climate, even in non-drought years.
+
+    3.
+Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+Smaller/larger plants might be more susceptible to drought.
+So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+<!--# this paragraph might be better in conclusion.
+Depends on if it makes sense without introducing the study system.
+BUT, I feel like there should be some justification for looking for interaction with plant size (if that's something we can even do).
+-->
+
+5.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
+
+6.
+Specifically, we want to know 1) Does drought have different impacts on plants in forest fragments compared to intact forest?
+2) Do the magnitudes and directions of these effects differ among vital rates?
+3) Do the effects of drought, fragmentation, and their interaction vary with plant size
+
+# Notes for Introduction (first attempt at outline):
 
@@
 
--   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments since fragments less buffered from changes in humidity and more sensitive to desiccation.
+-   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments because of reduced microclimate buffering.
 Under this hypothesis we would expect drought to reduce growth, survival, reproduction and recruitment more in fragmented landscapes compared to intact forest.

### 2020-11-12 analysis/paper/paper.Rmd
remove Rubim as author, knit to Word

@@
       correspondence: true
-  - Paulo Rubim:
-      email: rubimpaulo@gmail.com
-      institute: [UFSC]
-      correspondence: false
   - Maria Uriarte:
@@
   - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
-  - UFSC: Departamento do Botânica, Universidade Federal de São Carlos, CP 676, CEP 13565-905, São Carlos, SP, Brazil
   - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
@@
 
+If TRMM data is used, they suggest: "The TMPA data were provided by the\
+NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
+PPS, which develop and compute the TMPA as a contribution to TRMM."
+
 # References

### 2020-11-12 analysis/supplementary-materials/appendix_SPEI.Rmd
Created appendix for SPEI data

---
title: "Appendix 1: Quantifying Drought"
author: "Eric R. Scott"
date: '2020-09-13'
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  include = FALSE,
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "",
  dpi = 300)
library(tidyverse)
library(here)
library(conflicted)
library(SPEI)
library(tsibble)
library(lubridate)

library(HeliconiaDemography)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Outline

-   A review of global precip datasets (Sun et al. 2018) recommends ground-based products over satellite products for Brazil.

-   The global SPEI database is calculated from ground-based product (CRU) and a uses method for estimating evapotranspiration that takes more than just temperature into account.

-   There's also a ground-based product specifically for Brazil (Xavier et al).

-   We have rain gauge data that we can get SPEI from if we use imputation (informed by EMBRAPA station data)

```{r read-data}
# Precip Sources
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
trmm <-

# Already calculated SPEI  
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei)

bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))

```

# Methods

## Data Source Descriptions

### BDFFP rain gauge data

Daily rain gauge data collected from 1987--2010 at BDFFP camps. The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010. There we also a large number of observations that were accumulated rainfall over two or more days. We removed multi-day accumulations as it was impossible to know how rainfall was distributed over the prior days. Missing values were then imputed after averaging rainfall data by ranch. This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps. Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed. Multiple imputation produced \_\_\_ <!--# 10? --> imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using potential evapotranspiration from Xavier et al. [-@xavierDailyGriddedMeteorological2016] to calculate climatic water balance <!--# why not use mean temperature from weather stations? -->. Then, the <!--# number --> resulting SPEI datasets were averaged to produce a single dataset.

### Ground-based gridded products

-   Daily gridded meteorological variables in Brazil [@xavierDailyGriddedMeteorological2016]

    -   Resolution 0.25º x 0.25º

    -   Includes precipitation and potential evapotranspiration

    -   PET calculated by ... method

    -   Source: <!--# number --> ground-based weather stations across Brazil

    -   Date range: 1980--2015

-   Daily precipitation grids for South America [@liebmannDailyPrecipitationGrids2005] <!--# Not sure this one even makes sense to use, as it doesn't include ETO or temperature data, so I'd have to get that from Xavier or another source. -->

    -   Resolution 1º x 1º

    -   Includes only precipitation

    -   PET calculated by \_\_\_ method using \_\_\_ data

    -   Source: <!--# number --> ground-based weather stations across South America

    -   Date range: 1987--2012

-   Global SPEI [@santiago2017]

    -   Resolution: 0.5º x 0.5º

    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the FAO-56 Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.

    -   Source: CRU TS version 4.03, a gridded product based on land stations.

    -   Date range: 1901--2018

### Satelite and reanalysis products

-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->

    -   Resolution: 0.25º x 0.25º

    -   Includes:

    -   PET calculated by\_\_\_ method using\_\_\_ data

    -   Date range: 1997--2015

-   PERSIANN-CDR

    -   only rainfall, but recommended by [@zhaoEvaluatingDroughtMonitoringUtility2019] for Amazonia.

    -   0.25º resolution

    -   1983--present

```{r xa-spei}
#aggregate monthly, then calculate climate water balance.
xa_mon <-
  xa %>% 
  unite(latlon, lat, lon, sep = ", ") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, latlon) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = latlon)

xa_spei <-
  xa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(latlon) %>%
  group_nest() %>% 
  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(xa_spei, data)) %>% 
  select(-cb)
```

```{r sa-spei}
# sa # I'm not sure this one is worth adding to the comparison since there is no temperature or ETO data associated with it.

```

```{r trmm-spei}
library(ncdf4)
nc_open()
```

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

### 2020-11-13 analysis/paper/paper.Rmd
added notes from conversation with Emilio

@@
 
-1.  Little is known about how organisms respond to multiple anthropogenic stresses simultaneously <!--# (see Lakeman-Fraser and Ewers 2014)  -->
+1.
+Little is known about how organisms respond to multiple anthropogenic stresses simultaneously <!--# (see Lakeman-Fraser and Ewers 2014)  Might not be a good introduction if we can't do much with interactions.
+--> ...
+how multiple sources of biotic and abiotic stress at different scales affect demography.
+...
+increasingly antrhropogenic...
 
-2.  The Amazon Basin, an important place, has been fragmented, and although deforestation has slowed in recent years, it is continuing.
+    -   "multiple environmental factors" instead of "anthropogenic stresses"?
 
-    1.  fragmentation is bad because...
+    -   Cite Pedro F. Quintana-Ascencio in intro---he will likely be a reviewer.
+
+    -   Cite Jennifer Williams
+
+    -   Emphasize demography.
+We know about interactions on physiology and other stuff, but we don't know as much about how demography of plants is shaped by how populations respond to multiple stresses, increasingly anthropogenic.
+
+    -   Especially difficult when lagged effects are expected.
+Part of the challenge in doing this modeling is the statistics are new <!--# maybe goes in discussion --> .
+"Move beyond the statistical shackles of interaction terms".LOL
+
+2.  The Amazon Basin (or tropical lowland forests), an important place, has been fragmented, and although deforestation has slowed in recent years, it is continuing.
+
+    1.  fragmentation is bad because loss of microclimate buffering, edge effects.
 
@@
 
-4.  The effects of drought and fragmentation could be additive, or interactive.
+4.  The effects of fragmentation and drought could be additive, or interactive. Could be both positive or negative depending on the vital rate.
 
@@
 
-    3.
+5.
+In addition to variation among vital rates, responses to drought and fragmentation cuold depend on life history stage or plant size.
+Large established adults could repsond differently than seedlings.
+This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
+
+    1.
 Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
@@
 So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-<!--# this paragraph might be better in conclusion.
-Depends on if it makes sense without introducing the study system.
-BUT, I feel like there should be some justification for looking for interaction with plant size (if that's something we can even do).
--->
 
-5.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
+6.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
 
-6.
+    1.
 Specifically, we want to know 1) Does drought have different impacts on plants in forest fragments compared to intact forest?
 2) Do the magnitudes and directions of these effects differ among vital rates?
-3) Do the effects of drought, fragmentation, and their interaction vary with plant size
+3) Do the effects of drought, fragmentation, and their interaction vary with plant size <!--# are there size dependent effects of drought and fragmentation?
+4) are the effects of drought and fragmentation indpenedent, additive, or synergistic -->
 
@@
 
+metapopulation dynamics stuff goes here.
+
 # Conclusion

### 2020-11-17 R/utils.R
fixed bug where annote_spei was extending y limits.

@@ -64,8 +64,9 @@ nearest <- function(x, val) {
 #'
 #' @examples
 annotate_spei <- function(p, ...) {
+  ymin <- min(ggplot_build(p)$layout$panel_scales_y[[1]]$range$range[1], -2.5)
   spei_rects <- list(
-    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -4.2, ymax = -2, fill = "#e31a1c", ...),
+    annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = ymin, ymax = -2, fill = "#e31a1c", ...),
     annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -2, ymax = -1.5, fill = "#fd8d3c", ...),
     annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -1.5, ymax = -1, fill = "#fecc5c", ...),
     annotate(geom = "rect", xmin = -Inf, xmax = Inf, ymin = -1, ymax = 0, fill = "#ffffb2", ...) 

### 2020-11-17 analysis/paper/vital rate models.Rmd
figures for lab meeting talk

@@
 library(tsibble)
+library(broom)
 
@@
 # devtools::load_all()
+```
+
+```{r}
+#define plotting themes
+
+theme_talk <- theme_bw() + theme(text = element_text(size = 18))
 ```
@@
 ```
+
+```{r}
+length(unique(demog$ha_id_number))
+range(demog$year)
+2009-1998
+```
+
 # Join data sets
@@
 
+**TODO:**
 Now join by year and whether plot is in grid cell or not.
@@
   demog_full %>% 
-  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
+  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
+  mutate(flowered = as.numeric(infl > 0))
 ```
@@
 
+## Flowering Probability
+
+### Fit GAMs
+
+```{r}
+f_post_cf <-
+  gam(flowered ~ 
+        s(log_shts) +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_cf,
+      gamma = 1.2, #increases smoothness when above 1
+      method = "REML")
+anova(f_post_cf)
+```
+
+```{r}
+f_post_1ha <-
+  gam(flowered ~ 
+        s(log_shts) +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_1ha,
+      gamma = 1.2, #increases smoothness when above 1
+      method = "REML")
+anova(f_post_1ha)
+```
+
+### Get plot data
+
+This is pretty slow. It could probably be faster if I understood matrix math better.
+
+```{r}
+pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
+pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
+```
+
+Shared color bar limits so scales can be combined.
+
+```{r}
+flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
+                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
+# plogis(flowered_lim)
+```
+
+### Plots
+
+**Raster plots:**
+
+```{r}
+p_flowered_cf <-
+  pred_f_post_cf %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_flowered_cf
+```
+
+```{r}
+p_flowered_1ha <-
+  pred_f_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
+  labs(title = "1-ha Fragments") +
+  theme_bw()
+p_flowered_1ha
+```
+
+**Slice through time at severe drought**
+
+```{r}
+p_flowered_cf_slice <-
+  pred_f_post_cf %>% 
+  filter(nearest(x, -1.50)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  labs(title = "Continuous Forest", y = "P(flowering)") +
+  theme_linedraw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# p_surv_cf_slice
+```
+
+```{r}
+p_flowered_1ha_slice <-
+  pred_f_post_1ha %>% 
+  filter(nearest(x, -1.5)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  labs(title = "1-ha Fragments", y = "P(flowering)") +
+  theme_linedraw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# p_surv_1ha_slice
+```
+
 # Results
@@
 ```{r include=FALSE}
-#TODO: switch to predict() so y axis is correct
-g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
-g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])
+Q <- demog_post$spei_history
+L <- demog_post$L
+
+newdata_cf <- 
+  tibble(
+    log_shts = log(
+      seq(min(demog_post_cf$shts, na.rm = TRUE),
+          max(demog_post_cf$shts, na.rm = TRUE),
+          length.out = 50)
+    ),
+    spei_history = matrix(mean(Q), ncol = ncol(Q), nrow = 50),
+    L = L[1:50, ]
+  )
+
+newdata_1ha <- 
+  tibble(
+    log_shts = log(
+      seq(min(demog_post_1ha$shts, na.rm = TRUE),
+          max(demog_post_1ha$shts, na.rm = TRUE),
+          length.out = 50)
+    ),
+    spei_history = matrix(mean(Q), ncol = ncol(Q), nrow = 50),
+    L = L[1:50, ]
+  )
+
+g_cf <- augment(g_post_cf, newdata = newdata_cf, type.predict = "link", se_fit = TRUE)
+g_1ha <- augment(g_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)
 
@@
   bind_rows(
-    "CF" = g_cf[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
-    "1-ha" = g_1ha[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
+    "CF" = select(g_cf, log_shts, .fitted, .se.fit),
+    "1-ha" = select(g_1ha, log_shts, .fitted, .se.fit),
     .id = "habitat"
@@
 
-I'm not 100% sure this is correct since it's not done with `predict()`
-
 ```{r}
-g_df %>% 
-  ggplot(aes(x = exp(x), y = exp(fit + coef(g_post_cf)[1]))) +
+p_growth_shoots <-
+  ggplot(g_df, aes(x = exp(log_shts), y = exp(.fitted))) +
   geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
@@
   geom_line(color = "blue") +
-  geom_ribbon(aes(ymin = exp(fit + coef(g_post_cf)[1] - se),
-                  ymax = exp(fit + coef(g_post_cf)[1] + se)),
+  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96),
+                  ymax = exp(.fitted + .se.fit * 1.96)),
               alpha = 0.2,
@@
   facet_wrap(~habitat) +
-  theme_bw() +
-  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
+  theme_talk
+p_growth_shoots
+ggsave(here("growth_shoots.png"), p_growth_shoots, width = 10, height = 5)
 ```
@@
 ```{r}
+p_growth_spei <-
   (p_growth_cf | p_growth_1ha) +
   plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
-  plot_layout(guides = "collect")
+  plot_layout(guides = "collect") & theme_talk
+p_growth_spei
+ggsave(here("growth_spei.png"), p_growth_spei, height = 7, width = 14)
 ```
@@
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
+ggsave(here("growth slice.png"))
 ```
@@
   )
+
+s_cf <-  augment(s_post_cf, newdata = newdata_cf, type.predict = "link", se_fit = TRUE)
+s_1ha <- augment(s_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)
+
+s_df <-
+  bind_rows(
+    "CF" = select(s_cf, log_shts, .fitted, .se.fit),
+    "1-ha" = select(s_1ha, log_shts, .fitted, .se.fit),
+    .id = "habitat"
+  )
+
 ```
@@
 ```{r}
-s_df %>% 
-  ggplot(aes(x = exp(x), y = plogis(fit + coef(s_post_cf)[1]))) +
+p_surv_shoots <-
+  ggplot(s_df, aes(x = exp(log_shts), y = plogis(.fitted))) +
   geom_line() +
-  geom_ribbon(aes(ymin = plogis(fit + coef(s_post_cf)[1] - se),
-                  ymax = plogis(fit + coef(s_post_cf)[1] + se)),
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96),
+                  ymax = plogis(.fitted + .se.fit * 1.96)),
               alpha = 0.2) +
@@
                    aes(y = surv, x = shts)) +
-  theme_bw() +
-  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
+  theme_talk
+ggsave(here("surv_shoots.png"), p_surv_shoots, width = 10, height = 5)
 ```
@@
 ```{r}
+p_surv_spei <- 
   (p_surv_cf|p_surv_1ha) + 
   plot_annotation(title = "Effect of drought on survival") +
-  plot_layout(guides = "collect")
+  plot_layout(guides = "collect") & theme_talk
+ggsave(here("surv_spei.png"), p_surv_spei, width = 14, height = 7)
 ```
@@
 ```{r}
-(p_surv_cf_slice / p_surv_1ha_slice) +
+((p_surv_cf_slice + theme(axis.text.x = element_blank())) / p_surv_1ha_slice) +
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
+ggsave(here("survival slice.png"))
 ```
+
+## Flowering
+
+```{r include=FALSE}
+f_cf <-  augment(f_post_cf,  newdata = newdata_cf,  type.predict = "link", se_fit = TRUE)
+f_1ha <- augment(f_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)
+f_df <-
+  bind_rows(
+    "CF" = select(f_cf, log_shts, .fitted, .se.fit),
+    "1-ha" = select(f_1ha, log_shts, .fitted, .se.fit),
+    .id = "habitat"
+  )
+```
+
+```{r}
+p_flowered_shoots <-
+  ggplot(f_df, aes(x = exp(log_shts), y = plogis(.fitted))) +
+  facet_wrap(~habitat) +
+  geom_line() +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96),
+                  ymax = plogis(.fitted + .se.fit * 1.96)),
+              alpha = 0.2) +
+  labs(x = "size (#shoots)", y = "P(flowering)") +
+  stat_summary_bin(data = bind_rows(demog_post_1ha, demog_post_cf),
+                   geom = "point",
+                   fun = ~mean(., na.rm = TRUE),
+                   bins = 20,
+                   aes(y = flowered, x = shts)) +
+  theme_talk
+p_flowered_shoots
+ggsave(here("flowering_shoots.png"), p_flowered_shoots, width = 10, height = 5)
+```
+So something weird is going on with the continuous forest data.  All I know is that if I do everything the same, but take the crossbasis smooth out of the GAM, then it looks right:
+
+```{r}
+flowergam <-
+  gam(flowered ~ s(log_shts),
+      family = binomial,
+      data = demog_post_cf,
+      gamma = 1.2, #increases smoothness when above 1
+      method = "REML")
+
+augment(flowergam, newdata = newdata_cf) %>% 
+  ggplot(aes(x = exp(log_shts), y = plogis(.fitted))) +
+  geom_line() +
+  stat_summary_bin(geom = "point",
+                   data = demog_post_cf %>% filter(!is.na(flowered)),
+                   aes(x = exp(log_shts), y = flowered),
+                   fun = mean,
+                   bins = 20) 
+```
+
+```{r}
+p_flowered_spei <- 
+  (p_flowered_cf|p_flowered_1ha) + 
+  plot_annotation(title = "Effect of drought on flowering") +
+  plot_layout(guides = "collect") & theme_talk
+ggsave(here("flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
+```
+
+The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.
+In continuous forest, drought in the previous February and March has a negative impact on survival.
+Wet conditions or very dry conditions in the previous August–October have a positive effect on survival.
+
+Here's a slice through SPEI = -1.5 (severe drought)
+
+```{r}
+((p_flowered_cf_slice + theme(axis.text.x = element_blank())) / p_flowered_1ha_slice) +
+  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
+ggsave(here("flowering slice.png"))
+```

### 2020-11-17 analysis/supplementary-materials/appendix_SPEI.Rmd
figures for lab meeting talk

@@
 ```{r trmm-spei}
-library(ncdf4)
-nc_open()
+# library(ncdf4)
+# nc_open()
+```
+
+# For Cronians talk
+Just gonna make a simple-ish plot (lol) with major droughts labeled
+
+```{r}
+xa_df <-
+  xa_spei %>% 
+  group_by(yearmonth) %>% 
+  #Summarize across grid cells
+  summarize(`Xavier et al.` = mean(xa_spei))
+
+comb <- left_join(xa_df, gspei %>% 
+  mutate(yearmonth = yearmonth(date), date = NULL)) %>% 
+  rename(`Global SPEI` = g_spei) %>% 
+  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
+  filter(yearmonth > yearmonth(ymd("1990-01-01")))
+```
+
+```{r}
+p <-
+  ggplot(comb %>% filter(dataset == "Xavier et al."), aes(x = yearmonth, y = spei, color = dataset)) +
+  geom_line() +
+  scale_color_manual(values = c("#3b1255", "darkgreen")) +
+  labs(y = "SPEI")
+p <- annotate_spei(p)
+
+p <-
+  p +
+  scale_x_yearmonth(date_breaks = "3 years",
+                    date_labels = "%Y") +
+  theme_bw() +
+  theme(axis.title.x = element_blank(), 
+        panel.grid = element_blank(),
+        legend.position = "bottom")
+p
+ggsave(here("spei_xavier.png"), p, width = 10, height = 4, units = "in")
+```
+
+Try labeling droughts
+
+```{r}
+library(gghighlight)
+p + gghighlight(year(yearmonth) ==2010, use_direct_label = FALSE)
 ```

### 2020-11-18 analysis/paper/paper.Rmd
cleaned up and fleshed out outline of introduction

@@
       - --lua-filter=../templates/pagebreak.lua
+linkcolor: blue
 bibliography: references.bib
@@
 1.
-Little is known about how organisms respond to multiple anthropogenic stresses simultaneously <!--# (see Lakeman-Fraser and Ewers 2014)  Might not be a good introduction if we can't do much with interactions.
---> ...
-how multiple sources of biotic and abiotic stress at different scales affect demography.
-...
-increasingly antrhropogenic...
-
-    -   "multiple environmental factors" instead of "anthropogenic stresses"?
+Organisms experience multiple sources of natural and anthropogenic stress simultaneously.
+However, the impacts of multiple sources of stress on population dynamics are not well understood [@tyeDemographicMenageTrois2016].
 
@@
 
-2.  The Amazon Basin (or tropical lowland forests), an important place, has been fragmented, and although deforestation has slowed in recent years, it is continuing.
+2.
+As a result of intense deforestation, tropical lowland forests are becoming an increasingly fragmented landscape.
+This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.
 
-    1.  fragmentation is bad because loss of microclimate buffering, edge effects.
+    -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?
 
-3.  Simultaneously, there is a drying trend in the Amazon and also increasing intensity and frequency of droughts caused by ENSO (wet-season) and sea surface warming (dry-season).
+    -   This can result in ...
 
-    1.  Drought is bad because...
-
-4.  The effects of fragmentation and drought could be additive, or interactive. Could be both positive or negative depending on the vital rate.
-
-    1.  The effects of drought could be more severe in fragments because of reduced microclimate buffering.
-
-    2.
-Alternatively, plants in fragmented habitats could be more resilient to drought because the individuals surviving in fragmented forests are already resilient in the face of a relatively dry and variable climate, even in non-drought years.
-
-5.
-In addition to variation among vital rates, responses to drought and fragmentation cuold depend on life history stage or plant size.
-Large established adults could repsond differently than seedlings.
-This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
-
-    1.
-Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
-Smaller/larger plants might be more susceptible to drought.
-So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-
-6.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
-
-    1.
-Specifically, we want to know 1) Does drought have different impacts on plants in forest fragments compared to intact forest?
-2) Do the magnitudes and directions of these effects differ among vital rates?
-3) Do the effects of drought, fragmentation, and their interaction vary with plant size <!--# are there size dependent effects of drought and fragmentation?
-4) are the effects of drought and fragmentation indpenedent, additive, or synergistic -->
-
-# Notes for Introduction (first attempt at outline):
-
-Habitat fragmentation is bad, pervasive, and getting worse
-
--   stats about global land use change <!--# start this broad or start with the Amazon? -->
-
--   stats about amazon deforestation [@malhi2008]
-
--   stats about how much land is near an edge (refs in Schwartz et al and betts 2019)
-
--   Why fragmentation bad
-
-    -   microclimate changes [@ewersFragmentationImpairsMicroclimate2013]
-
-    -   metapopulation and connectivity stuff (necessary?)
-
--   Plants in the tropics are less well adapted to fragmentation because lack of historical distrubance and less variable climate [@betts2019]
-
-## Climate change and drought
-
--   The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
-
--   The El Niño--Southern Oscillation (ENSO) reduces wet-season rainfall in the Amazon <!--# is this correct?
-Or only in the south?
--->.
-ENSO-induced droughts are predicted to increase in frequency and severity.
-[@caiIncreasingFrequencyExtreme2014].
-
--   Also sea surface temperature related drought.
-Warming of north Atlantic results longer and more intense dry season in southern and eastern Amazonia [@phillipsDroughtSensitivityAmazon2009; @malhi2008]
-
--   drought effects on plant vital rates
+3.
+Simultaneously, these fragmented habitats are experiencing increased drought stress as a result of climate change.
+The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
+Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
 
@@
 
--   [@zengCausesImpacts20052008]
+    -   Drought is bad because...
 
-## Interactions
+    -   Combined environmental stresses, such as drought and fragmentation, have the potential to exacerbate or dampen effects on population dynamics [@tyeDemographicMenageTrois2016].
 
--   However, our understanding of how climate change and habitat fragmentation might interact to impact organisms is limited.
+4.
+Plant populations may be more susceptible to drought in fragmented forests.
+Fragmentation results in changes in forest microclimate that could exacerbate the effects of drought on plant populations.
+Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
 
--   [@schwartz2019] (mention how tree size classes affected differently, but we don't know if this is also true for understory herbaceous plants.
+    -   Trees near forest edges experience increased mortality after droughts compared to forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
 
--   [@malhi2008]
+    -    Increased mortality after droughts disproportionately affects large trees.
 
--   The 1997--1998 ENSO drought caused greater tree mortality and increased leaf litter fall in experimentally fragmented forests compared to intact forest in the Amazon [@laurancePositiveFeedbacksForest2001].
+5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.
 
--   Furthermore, the interactive effects of fragmentation and drought may be difficult to detect in long-lived species (e.g.
-Heliconia) [bruna paper that shows growth is reduced after 10 years slightly, but impacts reproduction]
+    -   [@betts2019].
 
-## This experiment
+6.
+<!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.
+Need to do some more reading about this though.
+-->In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
+Large established adults could respond differently than seedlings.
+This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
 
--   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments because of reduced microclimate buffering.
-Under this hypothesis we would expect drought to reduce growth, survival, reproduction and recruitment more in fragmented landscapes compared to intact forest.
+    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+Smaller/larger plants might be more susceptible to drought.
+So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
 
--   It's also possible that plants in fragments have already been filtered such that the survivors are already adapted to more variable and desiccating conditions associated with forest edges [@schwartz2019].
-This may result in drought having a less noticeable impact on plants in fragments, although we would expect fragmentation to still have an impact on survival, growth, and reproduction.
-
--   We set out to answer the following questions: 1) Does drought have different impacts on plants in forest fragments compared to intact forest?
-2) Do the magnitudes and directions of these effects differ among vital rates?
-3) Do the effects of drought, fragmentation, and their interaction vary with plant size [@tredennickSizebyenvironmentInteractionsNeglected2018] <!--# read this referrebce --> ?
-
-# Background
+7.
+We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
+Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
+2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
+3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?
 
 # Methods
+
+<!--# Emilio, feel free to edit -->
 
@@
 ## Site
+
+<!--# Emilio: please edit -->
 

### 2020-11-19 analysis/supplementary-materials/appendix_SPEI.Rmd
work on plot to compare data sources. Added EMBRAPA station data

@@
 sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
-trmm <-
+# trmm <-
 
@@
              delim = ";") %>% 
-  rename(date = "days since 1900-1-1", g_spei = spei)
+  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
+  mutate(yearmonth = yearmonth(date), date = NULL)
 
@@
 
+### Meterological stations
+
+-   EMBRAPA stations
+
+    -   Source:
+
+    -   Date Range:
+
+    -   PET calcualted by ... method
+
 ### Ground-based gridded products
@@
     -   Date range: 1980--2015
-
--   Daily precipitation grids for South America [@liebmannDailyPrecipitationGrids2005] <!--# Not sure this one even makes sense to use, as it doesn't include ETO or temperature data, so I'd have to get that from Xavier or another source.
--->
-
-    -   Resolution 1º x 1º
-
-    -   Includes only precipitation
-
-    -   PET calculated by \_\_\_ method using \_\_\_ data
-
-    -   Source: <!--# number --> ground-based weather stations across South America
-
-    -   Date range: 1987--2012
 
@@
 
-```{r sa-spei}
-# sa # I'm not sure this one is worth adding to the comparison since there is no temperature or ETO data associated with it.
+```{r embrapa-spei}
+rpde <-
+  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv")) %>% 
+  select(date = Data,
+         precip = Precipitacao,
+         eto = Evatotranspiracao,
+  )
+manaus <-
+  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Manaus_1980-01-01_2016-12-31.csv"))%>% 
+  select(date = Data,
+         precip = Precipitacao,
+         eto = Evatotranspiracao,
+  )
 
+embrapa_mon <-
+  bind_rows("rpde" = rpde, "manaus" = manaus, .id = "station") %>% 
+  mutate(yearmonth = yearmonth(date), .after = date) %>% 
+  group_by(yearmonth, station) %>% 
+  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
+  as_tsibble(index = yearmonth, key = station)
+
+embrapa_spei <-
+  embrapa_mon %>% 
+  mutate(cb = precip - eto) %>% 
+  group_by(station) %>% 
+  group_nest() %>% 
+  mutate(st_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
+  unnest(c(st_spei, data)) %>% 
+  select(-cb)
+
+embrapa_wide <-
+  embrapa_spei %>% 
+  select(yearmonth, station, st_spei) %>% 
+  pivot_wider(names_from= station, values_from = st_spei, names_prefix = "spei_") %>% 
+  rowwise() %>% 
+  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
+  ungroup()
 ```
@@
 
-# For Cronians talk
-Just gonna make a simple-ish plot (lol) with major droughts labeled
+# Plot
+
+Create a plot comparing data sources with known drought years labeled.
+
+## Combine into single tidy data frame
 
 ```{r}
+# take average across site
 xa_df <-
@@
   #Summarize across grid cells
-  summarize(`Xavier et al.` = mean(xa_spei))
+  summarize(xa_spei = mean(xa_spei))
 
-comb <- left_join(xa_df, gspei %>% 
-  mutate(yearmonth = yearmonth(date), date = NULL)) %>% 
-  rename(`Global SPEI` = g_spei) %>% 
+comb <-
+  full_join(xa_df, gspei) %>% 
+  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
+  rename(`Global SPEI` = g_spei,
+         `Xavier et al.` = xa_spei,
+         `EMBRAPA` = st_spei) %>% 
   pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
@@
 ```
+
+Annotate worst droughts.  
+Expected:
+
+- 2004 & 2006: weak ENSO
+- 2002 & 2009: moderate ENSO
+- 1997: severe ENSO
+- 2005 & 2010: severe SST
+
+```{r}
+comb <- 
+  comb %>% 
+  mutate(year = year(yearmonth)) %>% 
+  group_by(year) %>% 
+  mutate(min_mon = spei == min(spei)) %>% 
+  mutate(drought_label = case_when(
+    year %in% c(2004, 2006) & min_mon ~ "weak ENSO",
+    year %in% c(2002, 2009) & min_mon ~ "moderate ENSO",
+    year == 1997 & min_mon ~ "severe ENSO",
+    year %in% c(2005, 2010) & min_mon ~ "severe SST"
+    ))
+```
+
+```{r}
+labs <- comb %>% 
+  filter(!is.na(drought_label)) %>% 
+  group_by(drought_label) %>% 
+  mutate(xend = mean(yearmonth))
+
+p <-
+  ggplot(comb, aes(x = yearmonth, y = spei)) +
+  geom_line(aes(linetype = dataset), size = 0.5) +
+  # geom_label_repel(data = comb %>% filter(!is.na(drought_label)),
+  #                  aes(label = drought_label, y = -2.5))
+  geom_segment(data = labs,
+               aes(xend = yearmonth, yend = spei -1)) +
+  geom_label(data = labs,
+             aes(x = yearmonth, y = spei-1, label = year, fill = drought_label)) +
+  scale_x_yearmonth(date_breaks = "2 years") +
+  scale_color_brewer(type = "qual", palette = 2) +
+  coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
+  theme_bw()
+annotate_spei(p)
+```
+
+For Cronian's talk:
 
@@
 
-Try labeling droughts
-
-```{r}
-library(gghighlight)
-p + gghighlight(year(yearmonth) ==2010, use_direct_label = FALSE)
-```
-
 # References

### 2020-11-20 analysis/paper/paper.Rmd
EB added study site, study species, and demog surveys to methods + acknlowedgements info.
Need to replace current citations with bib format.

@@
 # Methods
+## Study site
 
-<!--# Emilio, feel free to edit -->
+The Biological Dynamics of Forest Fragments Project (BDFFP) is located ~70 km north of Manaus, Brazil (2º30' S, 60ºW).
+In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig.
+--).
+These fragments were isolated from 1980-1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
+The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al.
+2001).
+The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30-37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al.
+1992), and an understory dominated by stemless palms (Scariot 1999).
+The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content.
+Mean annual temperature in the region is 26º C (range=19–39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).
 
-*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible understory monocot native to the nonflooded rainforests of central Amazonia.
-*Heliconia* are rhizomatious, but clonal reproduction via underground runners is rare in *H.
-acuminata*, unlike other congeners [@bruna2004].
-Vegetative shoots grow from the rhizome as well as one or more flowering shoots when plants reproduce.
-Each flowering shoot has a single inflorescence of 20--25 flowers [@bruna2002].
-Flowers are pollinated nearly exclusively by two species of hummingbirds at this site <!--# species?
--->.
-Pollinated ovaries develop into fleshy fruits containing a maximum of 3 seeds which are dispersed exclusively by birds [@uriarte2011; @brunaHeliconiaAcuminataSeed2014].
+## Focal species 
+*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon (Kress 1990), and it is the most abundant understory herb at the BDFFP (Ribeiro et al.
+2010).
+While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others -- such as *H.
+acuminata* -- are found at lower densities in shaded forest understories.
+Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory (Berry and Kress 1991).
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* (Bruna et al.
+2004); reproducctive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@bruna2002].
+Fruits mature April-May and are consumed by a thrush and several species of manakin (Uriarte et al.
+2011).
+Dispersed seeds (1-3 seeds per fruit,	$\bar{x}$=2) germinate ~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments (Bruna 1999, 2002).
+*Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.
 
-## Site
-
-<!--# Emilio: please edit -->
-
--   All field data were collected at the Biological Dynamics of Forest Fragments Project (BDFFP)---a long term experimental site located about 80km north of Manaus (2º30' S, 60ºW) [@laurance2011; @laurance2002; @bierregaard1992].
-The experimental plots are located in BDFFP reserves in nonflooded tropical lowland rainforest with nutrient-poor, well-draining, xanthic ferrosols.
-
--   Something about seasonality
-
--   BDFFP is comprised of replicated forest fragments of 1, 10, and 100ha as well as continuous forest plots that were established between 1980 and 1984.
-Fragments were maintained by clearing secondary growth around fragments occasionally <!--# Best reference for this?
--->.
-Demographic plots ($50 \times 100$m) were established in <!--# year --> to track the growth, reproduction, and survival of *Heliconia acuminata*.
-Six of these plots were established in continuous forest, four were in 1-ha fragments, and three were in 10-ha fragments.
-The plots in continuous forest were established at haphazardly chosen locations 500--4000m from any mature forest borders, plots in 1-ha fragments were established on a randomly selected half of the fragment, and plots in 10-ha fragments were established in the center.
+## Demographic data collection
+In 1997-1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H.
+acuminata*.
+The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m-41 km.
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table).
+Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf.
+Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship (Bruna and Kress 2002, Bruna et al.
+2002).
+In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings.
+There is a seven-fold range in H.
+acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments (Bruna and Kress 2002).
 
@@
 # Acknowledgments
+
+We thank ----, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
+We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
+Financial support was provided by the U.S.
+National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
+This is publication No.
+---- in the BDFFP Technical Series.
 

### 2020-12-03 analysis/paper/paper.Rmd
I'm about to re-arrange the introduction

--- 2020-11-18 analysis/paper/paper.Rmd
@@
 
+    -   "Over 70000 km of new forest edges are being created annually in the Brazilian Amazon by human activities (Broadbent et al., 2008)"
+
     -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?
@@
 
-6.
-<!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.
-Need to do some more reading about this though.
--->In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
+6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->
+
+    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
 Large established adults could respond differently than seedlings.

### 2020-12-03 analysis/paper/paper.Rmd
worked on intro

@@
 # Introduction
+
+Habitat fragmentation and drought are major threats to biodiversity, and it has been hypothesized that they could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+Forest fragmentation is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008].
+
+Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest.
+Drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001] <!--# not sure if this belongs here, for flow into next paragraph -->.
+
+To date, however, this hypothesis remains untested for three primary reasons.
+First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants.
+<!--# example of juvenile but not reproductive study with drought -->.
+<!--# example of reproductive but not juvenile study with drought -->.
+Additionally, while most studies of tropical plant demography focus on trees, little is known about the understory plants that comprise up to \_\_\_\_% of biomass in tropical forests and are ecologically important <!--# citations -->.
+
+Second, there is a growing literature on how tropical plants respond to severe droughts, but few studies have compared the responses of plants in continuous forest with those in fragments.
+Laurance et al.
+[-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
+However, in a temperate system...
+<!--# oak paper --> [@forner2020].
+
+Finally, the long-term data needed to test population-level hypotheses about climate change-fragmentation synergies are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
+However, without multi-decadal sampling one cannot capture enough droughts to quantify their ecological impact, nor the variation in vital rates needed to parameterize demographic models of long-lived species in changing climates [@morrisQuantitativeConservationBiology2002; @teller2016].
+Long-term data are also needed because while some demographic effects of fragmentation or drought on can be detected immediately, others may take years to manifest.
 

### 2020-12-03 analysis/paper/vital rate models.Rmd
minor edits and notes

@@
 
-# library(HeliconiaDemography) #load functions in this compendium
+library(HeliconiaDemography) #load functions in this compendium
 # devtools::load_all()
@@
         s(log_shts) + #does this *need* to be log # shoots?
-        # s(shts) +
+        # s(year, bs = "re") + #random effect of year
+        s(plot, bs = "re") + #random effect of plot
         s(spei_history, L,
@@
         s(log_shts) +
+        s(plot, bs = "re") +
         s(spei_history, L,
@@
   demog_post %>%
-  dplyr::summarize(log_shts = log(mean(shts, na.rm = TRUE))) #log of mean shoots, not mean of log_shts
+  dplyr::summarize(log_shts = log(mean(shts, na.rm = TRUE)),  #log of mean shoots, not mean of log_shts
+                   plot = "newlevel")
 # exp(1.032814) # 2.8 shoots
@@
 ```
+**TODO** Check concurvity between log_shts and climate history.  That's the only thing that makes sense to me.
+```{r}
+concurvity(f_post_cf, full = FALSE)
+```
 

### 2020-12-03 analysis/supplementary-materials/appendix_SPEI.Rmd
minor edits and notes

@@
 library(lubridate)
+library(glue)
 
@@
 
-    -   Source:
+    -   Estacao Rio Preto da Eva: -2.65º, -59.65º
 
-    -   Date Range:
+    -   Estacao Manaus: -2.65º, -60.35º (note, this is not in Manaus city, but NW of it.)
 
-    -   PET calcualted by ... method
+    -   Date Range: 1980-present
+
+    -   Unclear how evapotranspiration is calculated from the [information provided](https://www.embrapa.br/en/busca-de-solucoes-tecnologicas/-/produto-servico/6044/infoclima---informacoes-climaticas).
 
@@
     -   Resolution 0.25º x 0.25º
-
     -   Includes precipitation and potential evapotranspiration
-
-    -   PET calculated by ... method
-
-    -   Source: <!--# number --> ground-based weather stations across Brazil
-
+    -   PET calculated by Pennman-Monteith method
+    -   Source: 3,625 ground-based weather stations across Brazil
     -   Date range: 1980--2015
@@
 
-    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the FAO-56 Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
+    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
 
-    -   Source: CRU TS version 4.03, a gridded product based on land stations.
+    -   Source: Precipitation from CRU TS version 4.03, a gridded product based on land stations.
 
@@
     -   Resolution: 0.25º x 0.25º
-
     -   Includes:
-
     -   PET calculated by\_\_\_ method using\_\_\_ data
-
     -   Date range: 1997--2015
@@
 
-# Plot
-
-Create a plot comparing data sources with known drought years labeled.
-
-## Combine into single tidy data frame
+# Combine into single tidy data frame
 
@@
   full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
-  rename(`Global SPEI` = g_spei,
-         `Xavier et al.` = xa_spei,
-         `EMBRAPA` = st_spei) %>% 
   pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
-  filter(yearmonth > yearmonth(ymd("1990-01-01")))
+  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
+  mutate(year = year(yearmonth), .before = yearmonth)
 ```
 
-Annotate worst droughts.  
+# Table of Expected Droughts
+Annotate worst droughts.\
 Expected:
@@
 ```{r}
-comb <- 
+spei_table <-
   comb %>% 
-  mutate(year = year(yearmonth)) %>% 
+  group_by(year, dataset) %>% 
+  mutate(min = min(spei)) %>% 
+  filter(spei == min) %>% 
+  mutate(month = month(yearmonth, label = TRUE)) %>% 
+  # summarize(min = min, month = month) %>%
+  filter(year %in% c(2002, 2004, 2006, 2009, 1997, 2005, 2010)) %>% 
+  select(year, dataset, spei, month) %>% 
+  mutate(dataset = str_remove(dataset, "_spei")) %>% 
+  pivot_wider(names_from = dataset, values_from = c(spei, month)) %>% 
+  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), .after = "year") %>% 
+  mutate(`Global SPEI` = glue("{round(spei_g, 2)} ({month_g})"),
+         `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
+         `EMBRAPA` = glue("{round(spei_st, 2)} ({month_st})")) %>% 
+  ungroup()
+```
+
+```{r}
+library(gt)
+gt(spei_table) %>% 
+  tab_style(
+    style = list(
+      cell_fill(color = "#ffffb2")
+    ),
+    locations = list(
+      cells_body(
+        columns = vars(`Global SPEI`),
+        rows = spei_g < 0 & spei_g >= -1
+      ),
+      cells_body(
+        columns = vars(EMBRAPA),
+        rows = spei_st < 0 & spei_st >= -1
+      ),
+      cells_body(
+        columns = vars(`Xavier et al.`),
+        rows = spei_xa < 0 & spei_xa >= -1
+      )
+    )
+  ) %>% 
+  tab_style(
+    style = list(
+      cell_fill(color = "#fecc5c")
+    ),
+    locations = list(
+      cells_body(
+        columns = vars(`Global SPEI`),
+        rows = spei_g < -1 & spei_g >= -1.5
+      ),
+      cells_body(
+        columns = vars(EMBRAPA),
+        rows = spei_st < -1 & spei_st >= -1.5
+      ),
+      cells_body(
+        columns = vars(`Xavier et al.`),
+        rows = spei_xa < -1 & spei_xa >= -1.5
+      )
+    )
+  ) %>% 
+  tab_style(
+    style = list(
+      cell_fill(color = "#fd8d3c")
+    ),
+    locations = list(
+      cells_body(
+        columns = vars(`Global SPEI`),
+        rows = spei_g < -1.5 & spei_g >= -2
+      ),
+      cells_body(
+        columns = vars(EMBRAPA),
+        rows = spei_st < -1.5 & spei_st >= -2
+      ),
+      cells_body(
+        columns = vars(`Xavier et al.`),
+        rows = spei_xa < -1.5 & spei_xa >= -2
+      )
+    )
+  ) %>% 
+  tab_style(
+    style = list(
+      cell_fill(color = "#e31a1c")
+    ),
+    locations = list(
+      cells_body(
+        columns = vars(`Global SPEI`),
+        rows = spei_g < -2
+      ),
+      cells_body(
+        columns = vars(EMBRAPA),
+        rows = spei_st < -2
+      ),
+      cells_body(
+        columns = vars(`Xavier et al.`),
+        rows = spei_xa < -2
+      )
+    )
+  ) %>% 
+  cols_hide(starts_with(c("spei_", "month_"))) %>% 
+  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`))
+
+```
+
+# Plot
+
+Create a plot comparing data sources with known drought years labeled.
+
+```{r}
+spei_plot <- 
+  comb %>% 
   group_by(year) %>% 
@@
     year %in% c(2005, 2010) & min_mon ~ "severe SST"
+    )) %>% 
+  mutate(dataset = case_when(
+    dataset == "g_spei"  ~ "Global SPEI",   
+    dataset == "xa_spei" ~ "Xavier et al.",
+    dataset == "st_spei" ~ "EMBRAPA"
   ))
@@
 ```{r}
-labs <- comb %>% 
+labs <- spei_plot %>% 
   filter(!is.na(drought_label)) %>% 
@@
 p <-
-  ggplot(comb, aes(x = yearmonth, y = spei)) +
+  ggplot(spei_plot, aes(x = yearmonth, y = spei)) +
   geom_line(aes(linetype = dataset), size = 0.5) +
-  # geom_label_repel(data = comb %>% filter(!is.na(drought_label)),
-  #                  aes(label = drought_label, y = -2.5))
   geom_segment(data = labs,
@@
 ```
+
+-   Are there still mismatches with 12-month SPEI?
 

### 2020-12-03 R/crossbasis.R

@@ -25,7 +25,7 @@
 #'   results of `predict.gam()`; min_dist is the euclidean distance on the unit
 #'   square from the fitted values to the actual data used to fit the model.
 #' @export
-#' @importFrom rlang enquo
+#' @importFrom rlang enquo abort
 #' @import purrr
 #' @import dplyr
 #' @import mgcv
@@ -197,6 +197,7 @@ add_min_dist <- function(df, Q_name, L_name, pred) {
     )
   }
   
+  #I think this is the super slow step.  Would be great if it didn't have to be rowwise().
   out <-
     grid %>% 
     rowwise() %>% 

### 2020-12-03 analysis/paper/paper.Rmd

--- 2020-11-20 analysis/paper/paper.Rmd
@@
 
+Habitat fragmentation and drought are major threats to biodiversity, and it has been hypothesized that they could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+Forest fragmentation is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008].
+
+Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest.
+Drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001] <!--# not sure if this belongs here, for flow into next paragraph -->.
+
+To date, however, this hypothesis remains untested for three primary reasons.
+First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants.
+<!--# example of juvenile but not reproductive study with drought -->.
+<!--# example of reproductive but not juvenile study with drought -->.
+Additionally, while most studies of tropical plant demography focus on trees, little is known about the understory plants that comprise up to \_\_\_\_% of biomass in tropical forests and are ecologically important <!--# citations -->.
+
+Second, there is a growing literature on how tropical plants respond to severe droughts, but few studies have compared the responses of plants in continuous forest with those in fragments.
+Laurance et al.
+[-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
+However, in a temperate system...
+<!--# oak paper --> [@forner2020].
+
+Finally, the long-term data needed to test population-level hypotheses about climate change-fragmentation synergies are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
+However, without multi-decadal sampling one cannot capture enough droughts to quantify their ecological impact, nor the variation in vital rates needed to parameterize demographic models of long-lived species in changing climates [@morrisQuantitativeConservationBiology2002; @teller2016].
+Long-term data are also needed because while some demographic effects of fragmentation or drought on can be detected immediately, others may take years to manifest.
+
 1.
@@
 This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.
+
+    -   "Over 70000 km of new forest edges are being created annually in the Brazilian Amazon by human activities (Broadbent et al., 2008)"
 

### 2020-12-03 analysis/paper/paper.Rmd
edited citations

@@
 # Methods
+
 ## Study site
 
-The Biological Dynamics of Forest Fragments Project (BDFFP) is located ~70 km north of Manaus, Brazil (2º30' S, 60ºW).
+The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
 In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig.
@@
 The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content.
-Mean annual temperature in the region is 26º C (range=19–39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).
 
 ## Focal species
-*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon (Kress 1990), and it is the most abundant understory herb at the BDFFP (Ribeiro et al.
-2010).
+
+*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
 While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others -- such as *H.
 acuminata* -- are found at lower densities in shaded forest understories.
-Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory (Berry and Kress 1991).
+Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a].
 In our sites *H.
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* (Bruna et al.
-2004); reproducctive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@bruna2002].
-Fruits mature April-May and are consumed by a thrush and several species of manakin (Uriarte et al.
-2011).
-Dispersed seeds (1-3 seeds per fruit,	$\bar{x}$=2) germinate ~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments (Bruna 1999, 2002).
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+Dispersed seeds (1-3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
 *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.
@@
 ## Demographic data collection
+
 In 1997-1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H.
@@
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf.
-Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship (Bruna and Kress 2002, Bruna et al.
-2002).
+Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
 In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings.
 There is a seven-fold range in H.
-acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments (Bruna and Kress 2002).
+acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 

### 2020-12-04 analysis/paper/paper.Rmd
re-organized to original outline

@@
 
-Habitat fragmentation and drought are major threats to biodiversity, and it has been hypothesized that they could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
-Forest fragmentation is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008].
+1.
+Organisms experience multiple sources of natural and anthropogenic stress simultaneously.
+However, the impacts of multiple sources of stress on population dynamics are not well understood.
+Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics.
+<!--# examples of interactions on demography -->For example, drought and habitat fragmentation ...[@tyeDemographicMenageTrois2016].
 
-Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+    -   Cite Pedro F. Quintana-Ascencio in intro---he will likely be a reviewer.
+
+    -   Cite Jennifer Williams
+
+    -   Emphasize demography.
+We know about interactions on physiology and other stuff, but we don't know as much about how demography of plants is shaped by how populations respond to multiple stresses, increasingly anthropogenic.
+
+2.
+Habitat fragmentation is one such source of environmental stress.
+Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest farther than 1 km from an edge has increased to 75% compared to 90% historically [@haddadHabitatFragmentationIts2015].
+Fragmentation can impact plant population dynamics through ...
+<!--# more about consequences of fragmentation pull from slides from Emilio-->
+
+3.
+Simultaneously, fragmented tropical forests are experiencing increased drought stress as a result of climate change.
+The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
+Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
+
+    -   Drought is bad because...
+
+    -   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]
+
+    -   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]
+
+    -   2010 drought [@lewis2010AmazonDrought2011]
+
+4.
+It has been hypothesized that drought and fragmentation could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+Fragmentation results in changes in forest microclimate near edges including increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
 This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest.
-Drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001] <!--# not sure if this belongs here, for flow into next paragraph -->.
+For example, drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
 
-To date, however, this hypothesis remains untested for three primary reasons.
+5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.
+
+    -   [@betts2019].
+
+6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->
+
+    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
+Large established adults could respond differently than seedlings.
+This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
+
+    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+Smaller/larger plants might be more susceptible to drought.
+So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+
+<!--# transition better -->To date, however, this hypothesis remains untested for three primary reasons.
 First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants.
@@
 
-1.
-Organisms experience multiple sources of natural and anthropogenic stress simultaneously.
-However, the impacts of multiple sources of stress on population dynamics are not well understood [@tyeDemographicMenageTrois2016].
+<!--# more about the heliconia dataset here -->
 
-    -   Cite Pedro F. Quintana-Ascencio in intro---he will likely be a reviewer.
-
-    -   Cite Jennifer Williams
-
-    -   Emphasize demography.
-We know about interactions on physiology and other stuff, but we don't know as much about how demography of plants is shaped by how populations respond to multiple stresses, increasingly anthropogenic.
-
-    -   Especially difficult when lagged effects are expected.
-Part of the challenge in doing this modeling is the statistics are new <!--# maybe goes in discussion --> .
-"Move beyond the statistical shackles of interaction terms".LOL
-
-2.
-As a result of intense deforestation, tropical lowland forests are becoming an increasingly fragmented landscape.
-This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.
-
-    -   "Over 70000 km of new forest edges are being created annually in the Brazilian Amazon by human activities (Broadbent et al., 2008)"
-
-    -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?
-
-    -   This can result in ...
-
-3.
-Simultaneously, these fragmented habitats are experiencing increased drought stress as a result of climate change.
-The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
-Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
-
-    -   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]
-
-    -   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]
-
-    -   2010 drought [@lewis2010AmazonDrought2011]
-
-    -   Drought is bad because...
-
-    -   Combined environmental stresses, such as drought and fragmentation, have the potential to exacerbate or dampen effects on population dynamics [@tyeDemographicMenageTrois2016].
-
-4.
-Plant populations may be more susceptible to drought in fragmented forests.
-Fragmentation results in changes in forest microclimate that could exacerbate the effects of drought on plant populations.
-Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
-
-    -   Trees near forest edges experience increased mortality after droughts compared to forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
-
-    -   Increased mortality after droughts disproportionately affects large trees.
-
-5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.
-
-    -   [@betts2019].
-
-6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->
-
-    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
-Large established adults could respond differently than seedlings.
-This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
-
-    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
-Smaller/larger plants might be more susceptible to drought.
-So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-
-7.
 We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.

### 2020-12-07 analysis/paper/paper.Rmd
worked on methods a bit

@@
 Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics.
-<!--# examples of interactions on demography -->For example, drought and habitat fragmentation ...[@tyeDemographicMenageTrois2016].
+<!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].
 
@@
 
-<!--# transition better -->To date, however, this hypothesis remains untested for three primary reasons.
+<!--# transition better -->
+
+To date, however, this hypothesis remains untested for three primary reasons.
 First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants.
@@
 In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig.
---).
-These fragments were isolated from 1980-1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
+-- <!--# fig ref -->).
+These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
 The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al.
-2001).
-The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30-37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al.
-1992), and an understory dominated by stemless palms (Scariot 1999).
+2001) <!--# replace ref -->.
+The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al.
+1992) <!--# replace ref -->, and an understory dominated by stemless palms (Scariot 1999).
 The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content.
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others -- such as *H.
-acuminata* -- are found at lower densities in shaded forest understories.
+While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H.
+acuminata*---are found at lower densities in shaded forest understories.
 Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a].
@@
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X <!--# from data?
+--> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
-Dispersed seeds (1-3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
+Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
 *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.
@@
 
-In 1997-1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H.
+In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H.
 acuminata*.
-The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m-41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table).
+The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf.
@@
 acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+
+## Weather data
+
+Monthly precipitation and potential evapotranspiration records ( <!--# date range -->) were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
+We calculated 3-month SPEI using the `SPEI` package in \`r version\$version.string\` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+<!--# something about why SPEI-3 is appropriate.
+-->  Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->.
+We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
+
+## Statistical Modeling
 

### 2020-12-07 analysis/supplementary-materials/appendix_SPEI.Rmd
added TRMM to spei analysis

@@
 *Last compiled: `r Sys.Date()`*
+# TODO
+
+- Write
+- Fix colors and labels and such on fig S1
+- Make a figure S2 that compares SPEI across BDFFP from east to west (3 panels)
+- Maybe add imputed data to this (or maybe not)
 
@@
     -   Resolution: 0.25º x 0.25º
-    -   Includes:
-    -   PET calculated by\_\_\_ method using\_\_\_ data
-    -   Date range: 1997--2015
+    -   Includes: Monthly precip
+    -   Doesn't include PET, so I guess I'll take that from the ground stations??
+    -   Date range: 1998--2019
 
@@
     -   1983--present
+
+# Xavier et al.
 
@@
 ```
+# EMBRAPA stations
 
@@
 
+# TRMM
+
 ```{r trmm-spei}
-# library(ncdf4)
-# nc_open()
+trmm <- read_csv(here("analysis", "data", "raw_data", "trmm.csv"))
+trmm2 <-
+  trmm %>% 
+  unite(latlon, lat, long, sep = ", ") %>% 
+  mutate(yearmonth = yearmonth(date), .after = date)
+# fortunately xa_mon has same latlon, so can join eto
+
+unique(xa_mon$latlon) %in% unique(trmm2$latlon)
+
+trmm_ts <- 
+  left_join(trmm2, xa_mon %>% select(yearmonth, latlon, eto), by = c("yearmonth", "latlon")) %>% 
+  #I guess I only need grid cells that are also in the Xavier et al. dataset?
+  filter(!is.na(eto)) %>% 
+  as_tsibble(index = yearmonth, key = latlon)
+
+trmm_spei <- 
+  trmm_ts %>% 
+  mutate(cb = precip - eto) %>% 
+  group_by(latlon) %>% 
+  group_nest() %>% 
+  mutate(trmm_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
+  unnest(c(trmm_spei, data)) %>% 
+  select(-cb)
 ```
@@
 
+trmm_df <-
+  trmm_spei %>% 
+  group_by(yearmonth) %>% 
+  summarize(trmm_spei = mean(trmm_spei))
+
 comb <-
   full_join(xa_df, gspei) %>% 
+  full_join(trmm_df) %>% 
   full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
@@
          `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
-         `EMBRAPA` = glue("{round(spei_st, 2)} ({month_st})")) %>% 
+         `EMBRAPA` = glue("{round(spei_st, 2)} ({month_st})"),
+         `TRMM` = glue("{round(spei_trmm, 2)} ({month_trmm})")) %>% 
   ungroup()
@@
         rows = spei_xa < 0 & spei_xa >= -1
+      ),
+      cells_body(
+        columns = vars(TRMM),
+        rows = spei_trmm < 0 & spei_trmm >= -1
       )
@@
         rows = spei_xa < -1 & spei_xa >= -1.5
+      ),
+      cells_body(
+        columns = vars(TRMM),
+        rows = spei_trmm < -1 & spei_trmm >= -1.5
       )
@@
         rows = spei_xa < -1.5 & spei_xa >= -2
+      ),
+      cells_body(
+        columns = vars(TRMM),
+        rows = spei_trmm < -1.5 & spei_trmm >= -2
       )
@@
         rows = spei_xa < -2
+      ),
+      cells_body(
+        columns = vars(TRMM),
+        rows = spei_trmm < -2
       )
@@
   cols_hide(starts_with(c("spei_", "month_"))) %>% 
-  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`))
-
+  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM))
 ```
@@
   group_by(year) %>% 
-  mutate(min_mon = spei == min(spei)) %>% 
+  mutate(min_mon = spei == min(spei, na.rm = TRUE)) %>% 
   mutate(drought_label = case_when(
@@
     dataset == "xa_spei" ~ "Xavier et al.",
-    dataset == "st_spei" ~ "EMBRAPA"
+    dataset == "st_spei" ~ "EMBRAPA",
+    dataset == "trmm_spei" ~ "TRMM"
   ))
@@
 
-```{r}
-labs <- spei_plot %>% 
+```{r fig.height=5, fig.width=10}
+labs <-
+  spei_plot %>%
   filter(!is.na(drought_label)) %>% 
   group_by(drought_label) %>% 
-  mutate(xend = mean(yearmonth))
+  mutate(xend = mean(yearmonth, na.rm = TRUE))
 
@@
   ggplot(spei_plot, aes(x = yearmonth, y = spei)) +
-  geom_line(aes(linetype = dataset), size = 0.5) +
+  geom_line(aes(color = dataset), size = 0.5) +
   geom_segment(data = labs,
@@
 annotate_spei(p)
-```
-
--   Are there still mismatches with 12-month SPEI?
-
-For Cronian's talk:
-
-```{r}
-p <-
-  ggplot(comb %>% filter(dataset == "Xavier et al."), aes(x = yearmonth, y = spei, color = dataset)) +
-  geom_line() +
-  scale_color_manual(values = c("#3b1255", "darkgreen")) +
-  labs(y = "SPEI")
-p <- annotate_spei(p)
-
-p <-
-  p +
-  scale_x_yearmonth(date_breaks = "3 years",
-                    date_labels = "%Y") +
-  theme_bw() +
-  theme(axis.title.x = element_blank(), 
-        panel.grid = element_blank(),
-        legend.position = "bottom")
-p
-ggsave(here("spei_xavier.png"), p, width = 10, height = 4, units = "in")
+ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
 ```

### 2021-01-04 analysis/supplementary-materials/appendix_SPEI.Rmd
add imputed data to table

@@
 ```
+
 # EMBRAPA stations
@@
 
+# Imputed data (averaged across for BDFFP)
+
+```{r}
+bdffp_spei <- 
+  bdffp %>% group_by(date) %>% 
+  summarize(bdffp_spei = mean(spei)) %>% 
+  mutate(yearmonth = yearmonth(date), date = NULL)
+```
+
 # Combine into single tidy data frame
@@
   full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
+  full_join(bdffp_spei) %>% 
   pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
@@
   group_by(year, dataset) %>% 
-  mutate(min = min(spei)) %>% 
+  mutate(min = min(spei, na.rm = TRUE)) %>% 
   filter(spei == min) %>% 
@@
          `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
-         `EMBRAPA` = glue("{round(spei_st, 2)} ({month_st})"),
-         `TRMM` = glue("{round(spei_trmm, 2)} ({month_trmm})")) %>% 
+         EMBRAPA = glue("{round(spei_st, 2)} ({month_st})"),
+         TRMM = glue("{round(spei_trmm, 2)} ({month_trmm})"),
+         BDFFP = glue("{round(spei_bdffp, 2)} ({month_bdffp})")) %>% 
   ungroup()
@@
 library(gt)
+table <- 
   gt(spei_table) %>% 
@@
         rows = spei_trmm < 0 & spei_trmm >= -1
+      ),
+      cells_body(
+        columns = vars(BDFFP),
+        rows = spei_bdffp < 0 & spei_bdffp >= -1
       )
@@
         rows = spei_trmm < -1 & spei_trmm >= -1.5
+      ),
+      cells_body(
+        columns = vars(BDFFP),
+        rows = spei_bdffp < -1 & spei_bdffp >= -1.5
       )
@@
         rows = spei_trmm < -1.5 & spei_trmm >= -2
+      ),
+      cells_body(
+        columns = vars(BDFFP),
+        rows = spei_bdffp < -1.5 & spei_bdffp >= -2
       )
@@
         rows = spei_trmm < -2
+      ),
+      cells_body(
+        columns = vars(BDFFP),
+        rows = spei_bdffp < -2
       )
@@
   cols_hide(starts_with(c("spei_", "month_"))) %>% 
-  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM))
+  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM, BDFFP))
+table
+gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
 ```
@@
     dataset == "st_spei" ~ "EMBRAPA",
-    dataset == "trmm_spei" ~ "TRMM"
+    dataset == "trmm_spei" ~ "TRMM",
+    dataset == "bdffp_spei" ~ "BDFFP imputed"
   ))
@@
   scale_x_yearmonth(date_breaks = "2 years") +
-  scale_color_brewer(type = "qual", palette = 2) +
+  scale_color_brewer(type = "qual", palette = "Set1") +
   coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
   theme_bw()
-annotate_spei(p)
+annotate_spei(p, alpha = 0.7)
 ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))

### 2021-01-05 analysis/paper/paper.Rmd
added TRMM to spei analysis
Switched models to use SPEI data calculated from Xavier et al. gridded precip.

@@
 
-Monthly precipitation and potential evapotranspiration records ( <!--# date range -->) were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+Monthly precipitation and potential evapotranspiration records were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
 SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
@@
 <!--# something about why SPEI-3 is appropriate.
--->  Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->.
-We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
+--> Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+We compared the SPEI values calculated using this data source to <!--# others --> including <!--# TRMM, near-by weather stations, global SPEI --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
 

### 2021-01-05 analysis/paper/vital rate models.Rmd
Switched models to use SPEI data calculated from Xavier et al. gridded precip.

--- 2020-11-12 analysis/paper/vital rate models.Rmd
@@
 
-# library(HeliconiaDemography) #load functions in this compendium
+library(HeliconiaDemography) #load functions in this compendium
 # devtools::load_all()
+```
+
+```{r}
+#define plotting themes
+theme_talk <- theme_bw() + theme(text = element_text(size = 18))
 ```
@@
 - Add random effects, interactions, other complexities
-- Match SPEI data to demographic plots programtically by lat lon data
+- Try integrative measure of size.  Some paper you read did some formula with height and leaf number. Plantago?
 
@@
 ```{r data, echo=TRUE}
-spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
 xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
@@
 ```
+
+```{r}
+length(unique(demog$ha_id_number))
+range(demog$year)
+2009-1998
+```
+
 # Join data sets
 
-## Gridded data
+## Prep precip data
 Get only rows for february (the month of the census), and year for joining
@@
 
-Now, for the demography data, get year and plot coords
+Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2
 
 ```{r}
-# demog$plot %>% unique() %>% sort()
-# plot_coords$plot %>% unique() %>% sort()
-demog <-
-  right_join(plot_coords %>% select(plot, lon, lat),
-             demog,
-             by = "plot") %>% 
-  select(-x_09, -y_09) #remove to avoid confusion since i'm not using here.
+unique(xa2$lat)
+mean(c(-2.25, -2.5))
+```
+So, from -2.25 to -2.5
+
+```{r}
+unique(xa2$lon)
+mean(c(-59.5, -59.75))
+mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
+mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
 ```
 
-Now join by year and whether plot is in grid cell or not.
-Probably need to construct `sf` object for grid cells (knowing that the lat lon are the centers and the resolution is 0.25º) and then use some `sf_` function to check if a point is inside of a grid cell.
-See: https://stackoverflow.com/questions/59766153/left-join-based-on-closest-lat-lon-in-r
-
-## Imputed data:
-
-`demog$year` = `spei$census_year`
-
-Getting site to match up will take more work.
+```{r}
+plot_coords
+```
 
 ```{r}
-unique(demog$plot)
-unique(demog$bdffp_reserve_no)
-unique(spei$site)
+demog2 <-
+  demog %>% 
+  mutate(lon = case_when(
+    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
+    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
+    TRUE ~ NA_character_
+  ), .before = plot)
+demog_full <- 
+  full_join(demog2, xa2, by = c("lon", "year")) %>% 
+  select(-lon, -lat, -yearmonth)
 ```
 
-km41 = 1501
-km37 = ??
-florestal = 1301, Florestal-CF
-cabo frio = 3402, CaboFrio-CF
-Porto Alegre = 3114, PortoAlegre-CF
-Colosso = 1104, 1202
-Dimona = 2206, 2108, 2107, Dimona-CF
-
-colosso_clust = 1301, 3402, 1104, 1202, CaboFrio-CF, Florestal-CF
-
-```{r}
-demog <- 
-  demog %>% 
-  mutate(rain_site = case_when(
-    bdffp_reserve_no %in% c("1301", "3402", "1104", "1202")      ~ "colosso_clust",
-    plot %in% c("CaboFrio-CF", "Florestal-CF")                   ~ "colosso_clust",
-    bdffp_reserve_no %in% c("2206", "2108", "2107")              ~ "dimona",
-    plot == "Dimona-CF"                                          ~ "dimona",
-    bdffp_reserve_no == "3114"                                   ~ "porto_alegre",
-    plot == "PortoAlegre-CF"                                     ~ "porto_alegre",
-    bdffp_reserve_no == "1501"                                   ~ "km_clust",
-    TRUE ~ NA_character_
-  ))
-```
-
-Get SPEI history only for census months (assuming all censuses done in February)
-
-```{r}
-spei_census <- 
-  spei %>% 
-  filter(month(yearmonth) == 2) %>% 
-  mutate(year = year(yearmonth), .before = site) %>% 
-  select(-yearmonth)
-```
-
-Join
+## Tidy
 
@@
 demog_full <-
-  left_join(demog, spei_census, by = c("rain_site" = "site", "year")) %>% 
-  mutate(log_shts = log(shts), .after = shts)
-
-# demog_full %>% select(spei, spei_history)
+  demog_full %>% 
+  mutate(log_shts = log(shts), .after = shts) %>% 
+  mutate(log_shts_next = log(shts_next), .after = shts_next)
 ```
@@
   demog_full %>% 
-  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
+  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
+  mutate(flowered = as.numeric(infl > 0), .after = infl)
+# View(demog_full)
 ```
@@
 
-To simplify initial modeling attempts, I'll split the data into seedlings vs. post-seedlings and by habitat.  I'll build the Q and L matrixes describing exposure and lag for each subset.
+Because I can't model interactions with DLNMs, I can either include factors like habitat as predictors (different intercepts, same functional form, same data) or fit separate models (potentially different functional forms, but not using all the data).
 
@@
         s(log_shts) + #does this *need* to be log # shoots?
-        # s(shts) +
+        # s(year, bs = "re") + #random effect of year
+        s(plot, bs = "re") + #random effect of plot
         s(spei_history, L,
           bs = "cb",
-          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 25),  #unimodal-ish response to drought, but one knot per month lag.
           xt = list(bs = "cr")),
@@
 
+Inspect the GAM
+
 ```{r}
-gam.check(g_post_cf) #looks good
-# summary(g_post_cf)
-anova(g_post_cf)
+gam.check(g_post_cf) 
 ```
+
+Suggests possibly more knots for s(log_shts), but it looks like a straight line to me
+
+```{r}
+plot(g_post_cf, pages=1)
+```
+
+```{r}
+summary(g_post_cf)
+```
+
+Remember, R-sq is probably so high because size this year is a very good predictor of size next year.
 
@@
         s(log_shts) +
+        s(plot, bs = "re") +
         s(spei_history, L,
           bs = "cb",
-          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 25),  #unimodal response to drought, but one knot per month lag.
           xt = list(bs = "cr")),
@@
 gam.check(g_post_1ha)
-anova(g_post_1ha)
+```
+
+Looks great
+
+```{r}
+summary(g_post_1ha)
 ```
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p_surv_cf_slice
+p_surv_cf_slice
 ```
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+p_surv_1ha_slice
+```
+
+## Flowering Probability
+
+### Fit GAMs
+```{r}
+x <- demog_post_cf$flowered
+sum(x, na.rm = TRUE) / length(x)
+```
+
+```{r}
+f_post_cf <-
+  gam(flowered ~ 
+        s(log_shts) +
+        s(plot, bs = "re") +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_cf,
+      gamma = 1.2, #increases smoothness when above 1
+      method = "REML")
+```
+
+```{r}
+# gam.check(f_post_cf)
+```
+
+```{r}
+summary(f_post_cf)
+```
+
+```{r}
+f_post_1ha <-
+  gam(flowered ~ 
+        s(log_shts) +
+        s(plot, bs = "re") +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_1ha,
+      gamma = 1.2, #increases smoothness when above 1
+      method = "REML")
+```
+
+```{r}
+summary(f_post_1ha)
+```
+
+### Get plot data
+
+This is pretty slow. It could probably be faster if I understood matrix math better.
+
+```{r}
+pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
+pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
+```
+
+Shared color bar limits so scales can be combined.
+
+```{r}
+flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
+                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
+plogis(flowered_lim)
+```
+
+### Plots
+
+```{r}
+p_flowered_cf <-
+  pred_f_post_cf %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_flowered_cf
+```
+Something is wrong.  Flowering probability should not be 1.
+
+```{r}
+p_flowered_1ha <-
+  pred_f_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
+  labs(title = "1-ha Fragments") +
+  theme_bw()
+p_flowered_1ha
+```
+
+**Slice through time at severe drought**
+
+```{r}
+p_flowered_cf_slice <-
+  pred_f_post_cf %>% 
+  filter(nearest(x, -1.50)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  labs(title = "Continuous Forest", y = "P(flowering)") +
+  theme_linedraw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# p_surv_cf_slice
+```
+
+```{r}
+p_flowered_1ha_slice <-
+  pred_f_post_1ha %>% 
+  filter(nearest(x, -1.5)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  labs(title = "1-ha Fragments", y = "P(flowering)") +
+  theme_linedraw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 # p_surv_1ha_slice
@@
 
-# Results
-
+# Plots
 ## Growth
-
-```{r include=FALSE}
-#TODO: switch to predict() so y axis is correct
-g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
-g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])
-
-g_df <-
-  bind_rows(
-    "CF" = g_cf[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
-    "1-ha" = g_1ha[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
-    .id = "habitat"
-  )
+```{r}
+p_growth_spei <-
+  (p_growth_cf | p_growth_1ha) +
+  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
+  plot_layout(guides = "collect") & theme_talk
+p_growth_spei
+ggsave(here("growth_spei_xa.png"), p_growth_spei, height = 7, width = 14)
 ```
 
-I'm not 100% sure this is correct since it's not done with `predict()`
+It looks like drought in September/October is bad for growth.
 
-```{r}
-g_df %>% 
-  ggplot(aes(x = exp(x), y = exp(fit + coef(g_post_cf)[1]))) +
-  geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
-              aes(x = shts, y = shts_next), alpha = 0.1) +
-  geom_line(color = "blue") +
-  geom_ribbon(aes(ymin = exp(fit + coef(g_post_cf)[1] - se),
-                  ymax = exp(fit + coef(g_post_cf)[1] + se)),
-              alpha = 0.2,
-              fill = "blue") +
-  labs(x = "size (#shoots)", y = "size (t + 1)") +
-  facet_wrap(~habitat) +
-  theme_bw() +
-  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
-```
-Plants don't grow much.
-
-```{r}
-anova(g_post_cf)
-anova(g_post_1ha)
-```
-
-There's a significant effect of lagged drought on number of shoots in the next year.
-
-```{r}
-(p_growth_cf | p_growth_1ha) +
-  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
-  plot_layout(guides = "collect")
-```
-
-It looks like drought in October is bad for growth in fragments, but drought in June and July is much worse for plants in continuous forest.
-
-Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`spei_history`).
-Might not be entirely comparable?
-Different penalties to the smooths make these hard to compare as well.
+Reminder that these are still fit with two different models and might not be entirely comparable.  Different penalties to the smooths make these hard to compare as well.
 
@@
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
+ggsave(here("growth slice_xa.png"))
 ```
@@
 
-```{r include=FALSE}
-s_cf <- plot.gam(s_post_cf, shift = coef(s_post_cf)[1])
-s_1ha <- plot.gam(s_post_1ha, shift = coef(s_post_1ha)[1])
-
-s_df <-
-  bind_rows(
-    "CF" = s_cf[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
-    "1-ha" = s_1ha[[1]][c("x", "fit", "se")] %>% 
-      bind_cols(),
-    .id = "habitat"
-  )
-```
-
-I'm not 100% sure this is correct since it's not done with `predict()`
-
-```{r}
-s_df %>% 
-  ggplot(aes(x = exp(x), y = plogis(fit + coef(s_post_cf)[1]))) +
-  geom_line() +
-  geom_ribbon(aes(ymin = plogis(fit + coef(s_post_cf)[1] - se),
-                  ymax = plogis(fit + coef(s_post_cf)[1] + se)),
-              alpha = 0.2) +
-  labs(x = "size (#shoots)", y = "P(survival)") +
-  facet_wrap(~habitat) +
-  stat_summary_bin(data = demog_post %>% filter(habitat %in% c("1-ha", "CF")),
-                   geom = "point",
-                   fun = mean,
-                   bins = 20,
-                   aes(y = surv, x = shts)) +
-  theme_bw() +
-  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
-```
-
-```{r}
-anova(s_post_cf)
-anova(s_post_1ha)
-```
-
 Drought also has a significant effect on survival in both habitat types.
@@
 ```{r}
+p_surv_spei <- 
   (p_surv_cf|p_surv_1ha) + 
   plot_annotation(title = "Effect of drought on survival") +
-  plot_layout(guides = "collect")
+  plot_layout(guides = "collect") & theme_talk
+p_surv_spei
+ggsave(here("surv_spei_xa.png"), p_surv_spei, width = 14, height = 7)
 ```
@@
 The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.
-In continuous forest, drought in the previous February and March has a negative impact on survival.
-Wet conditions or very dry conditions in the previous August–October have a positive effect on survival.
+In continuous forest, wet conditions or very dry conditions in the previous August–October have a positive effect on survival.
 
@@
 ```{r}
-(p_surv_cf_slice / p_surv_1ha_slice) +
+((p_surv_cf_slice + theme(axis.text.x = element_blank())) / p_surv_1ha_slice) +
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
+ggsave(here("survival slice_xa.png"))
 ```
+Pretty flat.
+
+## Flowering
+
+```{r}
+p_flowered_spei <- 
+  (p_flowered_cf|p_flowered_1ha) + 
+  plot_annotation(title = "Effect of drought on flowering") +
+  plot_layout(guides = "collect") & theme_talk
+p_flowered_spei
+ggsave(here("flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
+```
+
+The surface for continuous forest is perfectly flat (did I do something wrong with the color scale?) and for 1-ha fragments there is a strange (anomalous?) negative effect of high SPEI (wet conditions) in the summer 2 years ago.
+
+Here's a slice through SPEI = -1.5 (severe drought)
+
+```{r}
+((p_flowered_cf_slice + theme(axis.text.x = element_blank())) / p_flowered_1ha_slice) +
+  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
+ggsave(here("flowering slice.png"))
+```

### 2021-01-05 analysis/paper/paper.Rmd
added description of DLNM method

--- 2020-12-07 analysis/paper/paper.Rmd
@@
 
+We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
+DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, called a crossbasis function.
+This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
+Crossbasis functions were created using the *dlnm* package in R with a maximum lag of \_\_\_\_ <!--# how far back might drought impact heliconia?
+-->months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
+This particular maximum lag time was determined by ....
+<!--# preliminary experiments?
+theory?
+limitations of data?
+-->.
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
+In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.
+Cubic regression splines were used for both dimensions of the crossbasis function, with \_\_\_ <!--# same as number of lag steps -1 --> knots for the lag time and 3 knots for the drought response dimension to restrict the response to drought to bimodal at most complex.
+
+Growth
+
+Survival
+
+Flowering probability
+
 # Results

### 2021-01-06 analysis/wrangling/competition wrangling.Rmd
reorganized files to make R CMD check happier
calculate percentage of plants in database with coordinates

--- 2020-10-28 R/competition wrangling.Rmd
@@
 
-source(here("R", "utils.R"))
+library(HeliconiaDemography) #load functions in this compendium
 
@@
 
+What percentage of plants have coordinates?
+
+```{r}
+full_join(ha %>% select(-x_09, -y_09),
+          ha_coords,
+          by = c("plot", "row", "column", "tag_number", "year")) %>% 
+  group_by(ha_id_number) %>% 
+  summarize(has_coord = any(!is.na(x))) %>%  #true if there is a non-NA value for at least one year
+  summarize(percent_with_coords = sum(has_coord)/n())
+```
+
 # Calculate density

### 2021-01-07 analysis/paper/paper.Rmd

@@
 <!--# something about why SPEI-3 is appropriate.
---> Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->.
-We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
+--> Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+We compared the SPEI values calculated using this data source to <!--# others --> including <!--# TRMM, near-by weather stations, global SPEI --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
 

### 2021-01-08 analysis/paper/vital rate models.Rmd
change axes on plots

@@
   gam(shts_next ~ 
-        s(log_shts) + #does this *need* to be log # shoots?
-        # s(year, bs = "re") + #random effect of year
+        s(log_shts) + 
         s(plot, bs = "re") + #random effect of plot
@@
       data = demog_post_cf,
-      gamma = 1.2, #increases smoothness when above 1
+      # gamma = 1.2, #I maybe need to optimize this with cross-validation?
       method = "REML")
@@
 ```{r}
-plot(g_post_cf, pages=1)
+plot(g_post_cf, select = 1, main = "select = FALSE")
 ```
@@
       data = demog_post_1ha,
-      gamma = 1.2, #increases smoothness when above 1
+      # gamma = 1.2, #increases smoothness when above 1
       method = "REML")
@@
 
+```{r}
+plot(g_post_1ha, select = 1)
+```
+
 ### Get plot data
@@
 # exp(1.032814) # 2.8 shoots
+mean_size <- round(exp(ref_data$log_shts), 1)
 ```
@@
 pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
-```
-
-Lag labels
-
-```{r}
-# first column is February, the month of the census
-labs <- c("Feb", "Jan", paste0(month(12:1, label = TRUE), " (y-1)"), paste0(month(12:2, label = TRUE), " (y-2)"))
 ```
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:24, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+  ) +
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
   scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
-  labs(title = "Continuous Forest") +
+  labs(title = "Continuous Forest",
+       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
   theme_bw()
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
-  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:24, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+  ) +  scale_x_continuous("SPEI_3", expand = c(0,0)) +
   scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
-  labs(title = "1-ha Fragments") +
+  labs(title = "1-ha Fragments",
+       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
   theme_bw()
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "Continuous Forest", y = "shoots (t+1)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw()
 p_growth_cf_slice
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "1-ha Fragments", y = "shoots (t+1)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw()
 p_growth_1ha_slice
@@
 foo <- function(x) {
-  d <- 1:25
-  names(d)<-labs
   x <- as.numeric(x)
-  # s <- glue("lag {x} ({names(d[x])})")
-  s <- names(d[x])
-  return(s)
+  mon <- month(ymd("2020-2-01") - months(x), label = TRUE)
+  l <- glue("lag={x} ({mon})")
+  return(l)
 }
@@
   theme_bw()
+p_growth_cf_months
 ```
@@
       data = demog_post_cf,
-      gamma = 1.2, #increases smoothness when above 1
+      # gamma = 1.2, #increases smoothness when above 1
       method = "REML")
-anova(s_post_cf)
+```
+```{r}
+# gam.check(s_post_cf) #not sure what to expect from any of these plots
+```
+```{r}
+summary(s_post_cf)
 ```
@@
       data = demog_post_1ha,
-      gamma = 1.2, #increases smoothness when above 1
+      # gamma = 1.2, #increases smoothness when above 1
       method = "REML")
-anova(s_post_1ha)
+```
+```{r}
+summary(s_post_1ha)
 ```
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:24, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+  ) + 
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest") +
+  labs(title = "Continuous Forest",
+       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
   theme_bw()
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:24, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+  ) +
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments") +
+  labs(title = "1-ha Fragments",
+       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
   theme_bw()
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "Continuous Forest", y = "P(survival)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw()
 p_surv_cf_slice
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "1-ha Fragments", y = "P(survival)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw()
 p_surv_1ha_slice
@@
 ### Fit GAMs
+
 ```{r}
@@
 ```
+
+Around 0.05 percent of plants flowered.
 
@@
       data = demog_post_cf,
-      gamma = 1.2, #increases smoothness when above 1
+      # gamma = 1.2, #increases smoothness when above 1
       method = "REML")
@@
 ```{r}
+plot(f_post_cf, select = 1, trans = plogis)
+```
+
+Plants with exp(3) = 20 shoots have close to a 100% chance of flowering.
+
+```{r}
 f_post_1ha <-
@@
 summary(f_post_1ha)
+```
+
+```{r}
+plot(f_post_1ha, select = 1, trans = plogis)
 ```
@@
   geom_raster(interpolate = TRUE) +
-  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:24, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+  ) + 
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
@@
 ```
-Something is wrong.  Flowering probability should not be 1.
+
+Something is wrong.  Flowering probability should not be 1 for an average-sized plant.
 
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p_surv_cf_slice
+p_flowered_cf_slice
 ```
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p_surv_1ha_slice
+p_flowered_1ha_slice
 ```
@@
   plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
-  plot_layout(guides = "collect") & theme_talk
+  plot_layout(guides = "collect") & theme_talk & labs(caption = "")
 p_growth_spei
@@
   plot_annotation(title = "Effect of drought on survival") +
-  plot_layout(guides = "collect") & theme_talk
+  plot_layout(guides = "collect") & theme_talk & labs(caption="")
 p_surv_spei
@@
   plot_annotation(title = "Effect of drought on flowering") +
-  plot_layout(guides = "collect") & theme_talk
+  plot_layout(guides = "collect") & theme_talk & labs(caption = "")
 p_flowered_spei

### 2021-01-08 analysis/paper/vital rate models.Rmd
moved joining of demographic and spei datasets to wrangling script.

@@
 ```{r data, echo=TRUE}
-xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
-demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
-plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
-```
-
-```{r}
-length(unique(demog$ha_id_number))
-range(demog$year)
-2009-1998
-```
-
-# Join data sets
-
-## Prep precip data
-Get only rows for february (the month of the census), and year for joining
-
-```{r}
-xa2 <- 
-  xa %>%
-  filter(month(yearmonth) == 2) %>% 
-  mutate(year = year(yearmonth), .after =yearmonth) %>% 
-  separate(latlon, into = c("lat", "lon"), sep = "_")
-xa2
-```
-
-Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2
-
-```{r}
-unique(xa2$lat)
-mean(c(-2.25, -2.5))
-```
-So, from -2.25 to -2.5
-
-```{r}
-unique(xa2$lon)
-mean(c(-59.5, -59.75))
-mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
-mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
-```
-
-```{r}
-plot_coords
-```
-
-```{r}
-demog2 <-
-  demog %>% 
-  mutate(lon = case_when(
-    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
-    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
-    TRUE ~ NA_character_
-  ), .before = plot)
 demog_full <-
-  full_join(demog2, xa2, by = c("lon", "year")) %>% 
-  select(-lon, -lat, -yearmonth)
+  read_csv(here("analysis", "data", "derived_data", "ha_demog_spei.csv"), 
+           col_names = TRUE,
+           cols(plot = col_character(),
+                bdffp_reserve_no = col_character()))
 ```

### 2021-01-08 analysis/wrangling/1-weather_wrangling.R

@@ -1,0 +1,67 @@
+library(tidyverse)
+library(here)
+library(lubridate)
+library(tsibble)
+library(tsModel)
+library(SPEI)
+library(conflicted)
+conflict_prefer("lag", "dplyr")
+conflict_prefer("filter", "dplyr")
+
+# Set maximum lag in months
+maxlag = 24
+
+# Imputed rain gauge SPEI
+
+precip <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
+
+# Goal: a dataframe with rows for each year:site combination, columns with SPEI in jan, dec, nov, oct, ...jan_prev, dec_prev, etc. Say, 24 months.
+# Census occurrs late Jan, early Feb.
+
+df_spei <- 
+  precip %>% 
+  mutate(yearmonth = yearmonth(date)) %>% 
+  select(yearmonth, site, spei)
+
+spei_lag <-
+  df_spei %>%
+  group_by(site) %>%
+  arrange(site, yearmonth) %>%
+  mutate(spei_history = Lag(spei, 0:maxlag),
+         L = matrix(0:maxlag, nrow = n(), ncol = maxlag + 1, byrow = TRUE))
+
+#to join with demography, filter to get only history for february
+# filter(spei_lag, month(yearmonth) == 2)
+
+write_rds(spei_lag, here("analysis", "data", "derived_data", "spei_lags.rds"))
+
+# Repeat with gridded data from Xavier et al.
+xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
+
+xa_spei <-
+  xa %>% 
+  unite(latlon, lat, lon) %>% 
+  mutate(yearmonth = yearmonth(date)) %>% 
+  group_by(latlon, yearmonth) %>% 
+  summarize(precip = sum(precip), eto = sum(eto)) %>% 
+  mutate(cb = precip - eto) %>% 
+  as_tsibble(index = yearmonth, key = latlon) %>% 
+  mutate(spei = as.numeric(spei(cb, scale = 3)$fitted))
+
+
+xa_spei_lags <-
+  xa_spei %>%
+  group_by(latlon) %>% 
+  mutate(spei_history = Lag(spei, 0:maxlag),
+         L = matrix(0:maxlag, nrow = n(), ncol = maxlag + 1, byrow = TRUE))%>% 
[…]

### 2021-01-08 analysis/wrangling/2-demography_wrangling.R

@@ -1,0 +1,104 @@
+library(tidyverse)
+library(here)
+library(janitor)
+library(lubridate)
+library(tsibble)
+library(conflicted)
+conflict_prefer("filter", "dplyr")
+conflict_prefer("select", "dplyr")
+
+demog <- 
+  read_csv(here("analysis", "data", "raw_data", "Ha_survey_with_Zombies.csv"), col_names = TRUE,
+           cols(plot = col_character(),
+                bdffp_reserve_no = col_character(),
+                shts = col_integer(),
+                year = col_integer(),
+                infl = col_integer(),
+                tag_number = col_character(),
+                HA_ID_Number = col_character())) %>% 
+  clean_names()
+# one plant's x-coordinate entered with a comma.
+demog <- 
+  demog %>%
+  mutate(x_09 = ifelse(
+    ha_id_number == 5800,
+    2.0,
+    x_09
+  ))
+
+# When is a plant really dead?  It "dies" after the last non-`NA` value for ht.
+
+demog <-
+  demog %>% 
+  group_by(ha_id_number) %>% 
+  # Routine for figuring out last year of data collected
+  mutate(x = cumsum(ht > 0 & !is.na(ht))) %>%
+  mutate(n = 1:n()) %>% 
+  mutate(surv = if_else(
+    n == which.max(x == max(x)),
+    0L,
+    1L
+  )) %>% 
+  # remove rows after plant is dead
+  filter(row_number() <= which(surv == 0)[1]) %>% 
+  # If last year of data collected at that plot, it's not dead yet unless the notes say so.
+  group_by(plot) %>% 
+  mutate(surv = if_else(
+    year == max(year) & code_notes != "dead (2)",
+    NA_integer_,
+    surv)) %>% 
+  select(-n, -x) %>% 
+  ungroup()
+
+# count(demog, surv) # 2230 dead plants over course of study
+
+# add height and number shoots in the next year
+
+demog <- 
+  demog %>% 
+  group_by(ha_id_number) %>% 
+  mutate(ht_next = lead(ht),
+         shts_next = lead(shts)) %>% 
+  ungroup()
+
+# arrange columns
+demog <- 
+  demog %>%
+  select(ranch, bdffp_reserve_no, plot, habitat, #site level
+         row, column, x_09, y_09, ha_id_number, tag_number, #plot level
+         year,
[…]

### 2021-01-08 analysis/wrangling/2-demography_wrangling.R
changed survey data to .csv.  No reason for it to be .rds.

@@ -69,7 +69,7 @@ demog <-
          year,
          ht, ht_next, shts, shts_next, infl, surv, code_notes, code2) #plant level
 
-write_rds(demog, here("analysis", "data", "derived_data", "ha_survey.rds"))
+write_csv(demog, here("analysis", "data", "derived_data", "ha_survey.csv"))
 
 
 # combine demography with SPEI history ------------------------------------

### 2021-01-08 R/utils.R
moved creation of flwr and next year columns to demography-wrangling.r.  Also changed how dead plants are calculated (closes #2)

@@ -82,3 +82,44 @@ annotate_spei <- function(p, ...) {
   P2$layers <- c(P2$layers[(nlayers-3):nlayers], P2$layers[1:(nlayers-4)])
   return(P2)
 }
+
+
+
+#' Convert demographic survey data into binary survival
+#' 
+#' This function might only make sense to use if you continue surveying for
+#' plants long after you plan to count them as dead.  It is convenient if you
+#' have many NA entries after a plant has died (or is assumed dead) and you want
+#' to know which years it was alive and which it was dead.  You want to ignore
+#' NAs that are not trailing and you want to ignore trailing NAs < n
+#'
+#' @param x any ordered numeric vector such as height, size, or number of shoots, that is collected every observation unless plants are not detected.
+#' @param n number of sucessive, trailing missing survey points required to consider an organism dead.
+#'
+#' @return a binary vector with 1 = alive and 0 = assumed dead
+#' @export
+#' @examples
+#' #Assume plants dead with after not detected for 2 years
+#' dead  <- c(1, 3, NA, NA, 1, 2, 1, NA, NA)
+#' alive <- c(1, 3, NA, NA, 1, 2, 1, 3, NA)
+#' as_living(dead)
+#' as_living(alive)
+as_living <- function(x, n = 2) {
+  
+  #if at least last 2 values are NA
+  trail <- (length(x) - n + 1):length(x)
+  #exclude negative indexes
+  trail <- trail[trail >= 0]
+  
+  if(all(is.na(x))){
+    alive <- rep(NA_integer_, length(x))
+  } else if (all(is.na(x[trail]))) {
+    #find the last non-NA value
+    y <- cumsum(!is.na(x))
+    last_real <- which.max(y == max(y))
+    alive <- c(rep(1L, last_real), rep(0L, length(x) - last_real))
+  } else {
+    alive <- rep(1L, length(x))
+  }
+  return(alive)
+}

### 2021-01-08 analysis/wrangling/2-demography_wrangling.R

@@ -1,10 +1,19 @@
+
+# Wrangling demographic survey data ---------------------------------------
+#' Author: Eric R. Scott
+#' Creation Date: 2020-01-08
+
+
+# Load packages -----------------------------------------------------------
 library(tidyverse)
 library(here)
 library(janitor)
 library(conflicted)
+source(here("R", "utils.R"))
 conflict_prefer("filter", "dplyr")
 conflict_prefer("select", "dplyr")
 
+# Load data ---------------------------------------------------------------
 demog <- 
   read_csv(here("analysis", "data", "raw_data", "Ha_survey_with_Zombies.csv"), col_names = TRUE,
            cols(plot = col_character(),
@@ -14,43 +23,43 @@ demog <-
                 infl = col_integer(),
                 tag_number = col_character(),
                 HA_ID_Number = col_character())) %>% 
-  clean_names()
-# one plant's x-coordinate entered with a comma.
-demog <- 
-  demog %>%
-  mutate(x_09 = ifelse(
-    ha_id_number == 5800,
-    2.0,
-    x_09
-  ))
+  clean_names() 
 
-# When is a plant really dead?  It "dies" after the last non-`NA` value for ht.
+# A few entries don't read in because x_09 was entered with a comma or semicolon instead of a decimal.
+fails <- problems(demog)
+fails <-
+  fails %>% 
+  mutate(corrected = str_replace(actual, "[\\,;]", "\\."))
+
+demog$x_09[fails$row] <- fails$corrected
+
+# Survival ----------------------------------------------------------------
+#TODO: Change this so plants are only dead after 2 years of NAs or if marked dead.
 
 demog <-
   demog %>% 
   group_by(ha_id_number) %>% 
-  # Routine for figuring out last year of data collected
-  mutate(x = cumsum(ht > 0 & !is.na(ht))) %>%
-  mutate(n = 1:n()) %>% 
-  mutate(surv = if_else(
-    n == which.max(x == max(x)),
-    0L,
-    1L
-  )) %>% 
-  # remove rows after plant is dead
-  filter(row_number() <= which(surv == 0)[1]) %>% 
-  # If last year of data collected at that plot, it's not dead yet unless the notes say so.
-  group_by(plot) %>% 
-  mutate(surv = if_else(
-    year == max(year) & code_notes != "dead (2)",
-    NA_integer_,
-    surv)) %>% 
-  select(-n, -x) %>% 
+  arrange(ha_id_number, year) %>% 
[…]

### 2021-01-11 analysis/paper/vital rate models.Rmd

--- 2021-01-08 analysis/paper/vital rate models.Rmd
@@
 xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
-demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
+demog <-
+  read_csv(
+    here("analysis", "data", "derived_data", "ha_survey.csv"),
+    col_names = TRUE,
+    cols(plot = col_character(),
+         bdffp_reserve_no = col_character())
+  )
 plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
@@
 demog_full <- 
-  full_join(demog2, xa2, by = c("lon", "year")) %>% 
+  left_join(demog2, xa2, by = c("lon", "year")) %>% 
   select(-lon, -lat, -yearmonth)
@@
 
+There are plants with 0 shoots and 0 ht (not sure what that means yet), so to log-transform, I need to first add a small number
+
 ```{r}
@@
   demog_full %>% 
-  mutate(log_shts = log(shts), .after = shts) %>% 
-  mutate(log_shts_next = log(shts_next), .after = shts_next)
+  mutate(log_shts = log(shts + 0.0001), .after = shts) %>% 
+  mutate(log_shts_next = log(shts_next + 0.0001), .after = shts_next)
 ```
@@
   demog_full %>% 
-  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
-  mutate(flowered = as.numeric(infl > 0), .after = infl)
+  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
 # View(demog_full)
@@
 
-## Growth
+# Growth
 
@@
 
-### Fit GAMs
+## Fit GAMs
 
@@
 
-### Get plot data
+## Get plot data
 
@@
 
-### Plots
+## Plots
 
@@
 
-## Survival
+# Survival
 
-### Fit GAMs
+## Fit GAMs
 
@@
 
-## Flowering Probability
+# Flowering Probability
 
-### Fit GAMs
+## Fit GAMs
 
 ```{r}
-x <- demog_post_cf$flowered
+x <- demog_post_cf$flwr
 sum(x, na.rm = TRUE) / length(x)
@@
 f_post_cf <-
-  gam(flowered ~ 
+  gam(flwr ~ 
         s(log_shts) +
@@
 f_post_1ha <-
-  gam(flowered ~ 
+  gam(flwr ~ 
         s(log_shts) +
@@
 
-Something is wrong.  Flowering probability should not be 1 for an average-sized plant.
-
 ```{r}
@@
 ```
+very similar to growth
 
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "Continuous Forest", y = "P(flowering)") +
@@
   geom_line() +
-  scale_x_continuous("", breaks = 0:24, labels = labs) +
+  scale_x_continuous("lag(months)", breaks = 0:24) +
   labs(title = "1-ha Fragments", y = "P(flowering)") +

### 2021-01-12 analysis/paper/paper.Rmd
moved creation of flwr and next year columns to demography-wrangling.r.  Also changed how dead plants are calculated (closes #2)
worked on statistical methods

--- 2021-01-05 analysis/paper/paper.Rmd
@@
 These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
-The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al.
-2001) <!--# replace ref -->.
-The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al.
-1992) <!--# replace ref -->, and an understory dominated by stemless palms (Scariot 1999).
+The vegetation regenerating around fragments is cleared periodically to ensure their isolation [@bierregaardLessonsAmazoniaEcology2001].
+The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
 The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content.
@@
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X <!--# from data?
+bourcieri* [@bruna2004]; reproductive plants will have 1-X <!--# from data?
 --> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 
-Monthly precipitation and potential evapotranspiration records ( <!--# date range -->) were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method.
+Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
 SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
-We calculated 3-month SPEI using the `SPEI` package in \`r version\$version.string\` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-<!--# something about why SPEI-3 is appropriate.
---> Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->.
-We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).
+More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months.
+SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckee1993].
+We chose a 3 month time scale as most relevant for *H.
+acuminata* as plants do not access sources of water buffered over longer time scales such as ground water or flood water <!--# work on this sentence -->.
+We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database <!--# citation --> and TRMM <!--# citation --> and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1)<!--# ref to appendix 1 -->.
 
@@
 
-Growth
+To model growth, we used shoot number as a proxy for plant size.
+Shoot number is easy to measure accurately and, unlike plant height, does not change if a plant is knocked over by a treefall.
+<!--# I could consider an integrative measure like log(shts * ht) -->.
+Size in year $t+1$ was modeled as a function of the natural log of size in year $t$ and drought history (i.e.
+the crossbasis function) with plot as a random effect.
+We used a scaled t error family distribution with a log link since residuals were leptokurtic when a gaussian distribution was used.
 
-Survival
+For survival, plants were assumed dead the year after they were last observed if the plant was not observed for three or more years.
+Survival and flowering probability were similarly modeled as a function of size and drought history (i.e.
+the crossbasis function) with a random effect of plot, but with a binomial family error distribution and a logit link function.
 
-Flowering probability
+Statistical significance of crossbasis functions and covariates was determined by ....
+
+To better understand the shape of the crossbasis function, we calculated fitted values for an average sized plant given a range of possible SPEI values from \_\_\_ to \_\_\_ at different lag times from \_\_\_ to \_\_\_ months, holding all other SPEI values at the average across the dataset (\_\_\_).
+These results could then be plotted as heatmaps with the *ggplot2* package ( <!--# cite ggplot2 --> ).
+
+One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
+This limitation meant that we were unable to model a potential size by environment (i.e.
+drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
+To check for a size by environment interaction, we followed the methods of Tredennick et al.
+[-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package <!--# cite lme4 -->.
+Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
+For each vital rate, we included an additional random effect of year on either the intercept or both the intercept and slope of the relationship between plant size and the response.
+We tested the significance of the year by size interaction with a permutation <!--# not quite permutation, I think?
+Check trednnick et al.
+again for what they call it --> test, again following Treddnick et al.
+[-@tredennickSizebyenvironmentInteractionsNeglected2018].
 

### 2021-01-13 analysis/paper/paper.Rmd
implement tredennick year by size GLMM strategy.

@@
 For each vital rate, we included an additional random effect of year on either the intercept or both the intercept and slope of the relationship between plant size and the response.
-We tested the significance of the year by size interaction with a permutation <!--# not quite permutation, I think?
-Check trednnick et al.
-again for what they call it --> test, again following Treddnick et al.
-[-@tredennickSizebyenvironmentInteractionsNeglected2018].
+We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].
 
 # Results
+
+## Size by year interaction
+
+In continuous forest plots there was support for size by year interactions for growth (95% LR~null~ = 4.54, LR~obs~ = 45.69) and survival (95% LR~null~ = 3.91, LR~obs~ = 29.18) , but not for flowering (95% LR~null~ = 4.44, LR~obs~ = 1.77).
+However, in 1 ha fragments, we saw no evidence of size by year interaction.
+For growth 95% LR~null~ = 4.45, LR~obs~ = 4.35, for survival 95% LR~null~ = 4.47, LR~obs~ = 0.62, and for flowering probability 95% LR~null~ = 3.86, LR~obs~ = 2.82.
+This is consistent with the DLNMs in that the crossbasis functions for \_\_\_ and \_\_\_ were much flatter in fragments compared to continuous forest.
 

### 2021-01-13 analysis/supplementary-materials/size-environment_interaction.Rmd
implement tredennick year by size GLMM strategy.

---
title: "Size-environment interaction"
author: "Eric R. Scott"
date: "2020-01-11"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
bibliography: "`r here::here('analysis', 'paper', 'references.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lme4)
library(bbmle)
library(car)
library(conflicted)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Rather than explicitly modeling drought effects on *Heliconia*, model environmental stochasticity as a random effect of year. This allows me to test an interaction term with size to see if responses to environment (maybe drought) are size dependent. Inspired by Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018].

# Load Data

This is the same demography data used elsewhere, but without the SPEI matrix-column

```{r data, echo=TRUE}
demog <-
  read_csv(
    here("analysis", "data", "derived_data", "ha_survey.csv"),
    col_names = TRUE,
    cols(plot = col_character(),
         bdffp_reserve_no = col_character())
  )
demog <-
  demog %>%
  mutate(log_shts = log(shts + 0.0001),
         log_shts_next = log(shts_next + 0.0001))
```

Seedlings are likely to have different responses than post-seedlings, so I'll split the data that way. To start with a relatively simple model, I'll only use continuous forest for now.

```{r}
demog_post <- 
  demog %>% 
  filter(code_notes != "sdlg (1)" | is.na(code_notes))

demog_post_cf <-
  demog_post %>% 
  filter(habitat == "CF")

demog_post_1ha <-
  demog_post %>% 
  filter(habitat == "1-ha")
```

# Function
Likelihood ratio test for random effects from Tredennick et al.  Tests observed likelihood ratio between model with and without interaction against what would be expected under null.

```{r}
sy_lrt <- function(model_y, model_sy, nsim = 500) {
  ## Loop over many iterations, simulate, compute LL: generates null distribution
  lrt <- numeric(nsim)

  pb <- txtProgressBar(min = 1, max = nsim, style = 3)
  for(i in 1:nsim){
    gtmp <- model.frame(model_y) # make copy of gdata
    y_star <- simulate(model_y) # simulate model (null hypothesis)

    #which column is response (probably always the first one, but just in case)
    resp_col <- attr(attributes(gtmp)$terms, "response")
    gtmp[ , resp_col] <- as.numeric(y_star[,1])

    # Fit models with simulated data from the null hypothesis
    tmp_null <- suppressWarnings(suppressMessages(update(model_y, data = gtmp)))
    tmp_full <- suppressWarnings(suppressMessages(update(model_sy, data = gtmp)))

    # Calculate log likelihoods
    ll_null <- logLik(tmp_null)
    ll_full <- logLik(tmp_full)
    lrt[i] <- 2*as.numeric(ll_full - ll_null)
    setTxtProgressBar(pb, i)
  }

  ## Calculate upper 95th quantile as value to exceed, for one-sided test
  upper <- as.numeric(quantile(lrt, probs = 0.95))

  ## Calculuate observed log likelihoods
  ll_null <- logLik(model_y)
  ll_full <- logLik(model_sy)
  obs_lrt <- 2*as.numeric(ll_full - ll_null)

  print(glue::glue("
                   Observed LR: {round(obs_lrt, 3)}.
                   95th quantile expected under null: {round(upper, 3)}
                   Significant interaction: {obs_lrt > upper}"))
  invisible(list(sim_lrt = lrt, q95 = upper, obs_lrt))
}
```

# Analysis (CF)

## Growth

```{r}
growth_y <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_cf)
growth_sy <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_cf)
AICtab(growth_y, growth_sy)
```

```{r}
lrt_growth <- sy_lrt(growth_y, growth_sy, nsim = 500)
```
Upper = 4.539, observed = 45.689

```{r}
ecdf(lrt_growth$sim_lrt) -> growth_dist
2*(1-growth_dist(45.689))
```

## Survival

```{r}
surv_y <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
surv_sy <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
AICtab(surv_y, surv_sy)
```

```{r}
lrt_surv <- sy_lrt(surv_y, surv_sy, nsim = 500)
```
Upper = 3.913, observed = 29.181

```{r}
ecdf(lrt_surv$sim_lrt) -> surv_dist
2*(1-surv_dist(29.181))
```

## Flowering

```{r}
flwr_y <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
flwr_sy <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
AICtab(flwr_y, flwr_sy)
```

```{r}
lrt_flwr <- sy_lrt(flwr_y, flwr_sy, nsim = 500)
```
upper = 4.439, observed = 1.77

```{r}
ecdf(lrt_flwr$sim_lrt)->flwr_dist
2*(1-flwr_dist(1.77))
```

# Analysis (1-ha)

## Growth

```{r}
growth_y_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_1ha)
growth_sy_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_1ha)
AICtab(growth_y_1ha, growth_sy_1ha)
```

```{r}
lrt_growth_1ha <- sy_lrt(growth_y_1ha, growth_sy_1ha, nsim = 500)
```

## Survival

```{r}
surv_y_1ha <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
surv_sy_1ha <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
AICtab(surv_y_1ha, surv_sy_1ha)
```

```{r}
lrt_surv_1ha <- sy_lrt(surv_y_1ha, surv_sy_1ha, nsim = 500)
```

## Flowering

```{r}
flwr_y_1ha <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
flwr_sy_1ha <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
AICtab(flwr_y_1ha, flwr_sy_1ha)
```

```{r}

lrt_flwr_1ha <- sy_lrt(flwr_y_1ha, flwr_sy_1ha, nsim = 500)

```

### 2021-01-15 analysis/paper/vital rate models.Rmd
switch to maximum lag of 36 months

@@
 library(broom)
+library(gratia)
 
@@
 
-- 
-- ~~Remove seedlings from dataset, for now~~
-- ~~Separate models for CF and fragment, for now~~
-- Add model for probability of reproduction and other vital rates
-- Model seedlings separately?
-- Add random effects, interactions, other complexities
-- Try integrative measure of size.  Some paper you read did some formula with height and leaf number. Plantago?
+- Switch measure of plant size to shoots * height
+- ~~Change lag to 36 months (3 years)~~
+- Produce ridge plot
 
@@
 plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
-```
-
-```{r}
-length(unique(demog$ha_id_number))
-range(demog$year)
-2009-1998
 ```
@@
   mutate(log_shts = log(shts + 0.0001), .after = shts) %>% 
-  mutate(log_shts_next = log(shts_next + 0.0001), .after = shts_next)
+  mutate(log_shts_next = log(shts_next + 0.0001), .after = shts_next) %>% 
+  mutate(size = shts*ht, size_next = shts_next*ht_next)
 ```
@@
           bs = "cb",
-          k = c(3, 25),  #unimodal-ish response to drought, but one knot per month lag.
+          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
           xt = list(bs = "cr")),
-      family = scat(link = "log"),
+      family = poisson(link = "log"),
+      # family = scat(link = "log"), #like leptokurtic gaussian.
       data = demog_post_cf,
@@
 ```{r}
-gam.check(g_post_cf) 
+appraise(g_post_cf)
 ```
@@
           bs = "cb",
-          k = c(3, 25),  #unimodal response to drought, but one knot per month lag.
+          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
           xt = list(bs = "cr")),
-      family = scat(link = "log"),
+      # family = poisson(link = "log"),
+      # family = quasipoisson(link = "log"),
+      family = gaussian(link = "log"),
+      # family = scat(link = "log"), #like leptokurtic gaussian.
       data = demog_post_1ha,
@@
 ```{r}
-gam.check(g_post_1ha)
+appraise(g_post_1ha)
 ```
-
-Looks great
 
@@
     "lag (months)",
-    breaks = 0:24, 
+    breaks = 0:36, 
     expand = c(0,0),
@@
       name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) +
@@
       name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) +  scale_x_continuous("SPEI_3", expand = c(0,0)) +
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "Continuous Forest", y = "shoots (t+1)") +
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "1-ha Fragments", y = "shoots (t+1)") +
@@
           bs = "cb",
-          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
           xt = list(bs = "cr")),
@@
 # gam.check(s_post_cf) #not sure what to expect from any of these plots
+appraise(s_post_cf)
 ```
@@
       name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:24), label = TRUE))
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) + 
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "Continuous Forest", y = "P(survival)") +
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "1-ha Fragments", y = "P(survival)") +
@@
 ```{r}
-# gam.check(f_post_cf)
+appraise(f_post_cf)
 ```
@@
       method = "REML")
+```
+
+```{r}
+appraise(f_post_1ha)
 ```
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "Continuous Forest", y = "P(flowering)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw() 
 p_flowered_cf_slice
@@
   geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:24) +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
   labs(title = "1-ha Fragments", y = "P(flowering)") +
-  theme_linedraw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_linedraw()
 p_flowered_1ha_slice
@@
 p_growth_spei
-ggsave(here("growth_spei_xa.png"), p_growth_spei, height = 7, width = 14)
+ggsave(here("analysis", "figures", "growth_spei.png"), p_growth_spei, height = 7, width = 14)
 ```
@@
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
-ggsave(here("growth slice_xa.png"))
+ggsave(here("analysis", "figures", "growth slice.png"))
 ```
@@
 p_surv_spei
-ggsave(here("surv_spei_xa.png"), p_surv_spei, width = 14, height = 7)
+ggsave(here("analysis", "figures", "surv_spei.png"), p_surv_spei, width = 14, height = 7)
 ```
@@
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
-ggsave(here("survival slice_xa.png"))
+ggsave(here("analysis", "figures", "survival slice.png"))
 ```
@@
 p_flowered_spei
-ggsave(here("flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
+ggsave(here("analysis", "figures", "flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
 ```
@@
   plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
-ggsave(here("flowering slice.png"))
+ggsave(here("analysis", "figures", "flowering slice.png"))
 ```

### 2021-01-15 analysis/wrangling/1-weather_wrangling.R

@@ -9,7 +9,7 @@ conflict_prefer("lag", "dplyr")
 conflict_prefer("filter", "dplyr")
 
 # Set maximum lag in months
-maxlag = 24
+maxlag = 36
 
 # Imputed rain gauge SPEI
 

### 2021-01-15 analysis/paper/vital rate models.Rmd
switched to using shts x ht for size

@@
 
-There are plants with 0 shoots and 0 ht (not sure what that means yet), so to log-transform, I need to first add a small number
+There are plants with 0 shoots and 0 ht.  I will remove those, as having no above ground biomass is really not a size, but a different category.
+There's also a couple of plants that are probably data entry errors in height.
 
@@
   demog_full %>% 
-  mutate(log_shts = log(shts + 0.0001), .after = shts) %>% 
-  mutate(log_shts_next = log(shts_next + 0.0001), .after = shts_next) %>% 
-  mutate(size = shts*ht, size_next = shts_next*ht_next)
+  filter(ht < 200, ht_next < 200) %>% 
+  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
+  mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
 ```
@@
 g_post_cf <-
-  gam(shts_next ~ 
-        s(log_shts) + 
+  gam(log_size_next ~ 
+        s(log_size) + 
         s(plot, bs = "re") + #random effect of plot
@@
           xt = list(bs = "cr")),
-      family = poisson(link = "log"),
-      # family = scat(link = "log"), #like leptokurtic gaussian.
+      # family = gaussian(link = "identity"),
+      family = scat(link = "identity"), #like leptokurtic gaussian.
       data = demog_post_cf,
@@
 ```{r}
+AIC(g_post_cf)
+
 appraise(g_post_cf)
@@
 
-Suggests possibly more knots for s(log_shts), but it looks like a straight line to me
+AIC for... 
+Gaussian: 40909.27. looks leptokurtic
+scaled t: 37284.47.  Waaay better
 
@@
 ```
+
+Hmm. Make sure this doesn't predict negative sizes.
 
@@
 g_post_1ha <-
-  gam(shts_next ~ 
-        s(log_shts) +
+  gam(log_size_next ~ 
+        s(log_size) +
         s(plot, bs = "re") +
@@
           xt = list(bs = "cr")),
-      # family = poisson(link = "log"),
-      # family = quasipoisson(link = "log"),
-      family = gaussian(link = "log"),
-      # family = scat(link = "log"), #like leptokurtic gaussian.
+      family = scat(link = "identity"), #like leptokurtic gaussian.
       data = demog_post_1ha,
@@
   demog_post %>%
-  dplyr::summarize(log_shts = log(mean(shts, na.rm = TRUE)),  #log of mean shoots, not mean of log_shts
+  dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
                    plot = "newlevel")
 # exp(1.032814) # 2.8 shoots
-mean_size <- round(exp(ref_data$log_shts), 1)
+mean_size <- round(exp(ref_data$log_size), 1)
 ```
@@
 ```{r}
+#TODO: replace with ridge plot
 plotdf <-
@@
   gam(surv ~ 
-        s(log_shts) +
+        s(log_size) +
         s(plot, bs = "re") +
@@
 # gam.check(s_post_cf) #not sure what to expect from any of these plots
-appraise(s_post_cf)
+appraise(s_post_cf, method = "simulate")
 ```
@@
 ```
+wow, not significant!
 
@@
 ```
+not significant!  I should try select=TRUE here.
 
@@
 
-**Raster plots:**
+These are not significant, so probably don't need to actually make these plots.  I'm mostly just curious what a non-significant crossbasis function looks like.
 
@@
   gam(flwr ~ 
-        s(log_shts) +
+        s(log_size) +
         s(plot, bs = "re") +
@@
 
-Drought also has a significant effect on survival in both habitat types.
+These aren't significant.
 

### 2021-01-15 analysis/supplementary-materials/size-environment_interaction.Rmd
switched to using shts x ht for size

@@
 
+# TODO
+
+- ~~Test that random effect of year is significant for all models~~
+
 # Purpose
@@
   demog %>%
-  mutate(log_shts = log(shts + 0.0001),
-         log_shts_next = log(shts_next + 0.0001))
+  filter(ht < 200, ht_next < 200) %>% 
+  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
+  mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
 ```
@@
 ```{r}
-growth_y <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_cf)
-growth_sy <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_cf)
-AICtab(growth_y, growth_sy)
+growth_null <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_cf)
+growth_y <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_cf)
+growth_sy <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_cf)
+AICtab(growth_null, growth_y, growth_sy)
 ```
+Significant interaction.
 
 ```{r}
+#This takes a very long time and gives results qualitatively similar to comparing AIC
 lrt_growth <- sy_lrt(growth_y, growth_sy, nsim = 500)
 ```
-Upper = 4.539, observed = 45.689
 
@@
 ```{r}
-surv_y <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
-surv_sy <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
-AICtab(surv_y, surv_sy)
+surv_null <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
+surv_y <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
+surv_sy <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
+AICtab(surv_null, surv_y, surv_sy)
 ```
+no interaction
 
@@
 ```
-Upper = 3.913, observed = 29.181
 
@@
 ```{r}
-flwr_y <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
-flwr_sy <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
-AICtab(flwr_y, flwr_sy)
+flwr_null <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
+flwr_y <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
+flwr_sy <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
+AICtab(flwr_null, flwr_y, flwr_sy)
 ```
+No interaction
 
@@
 ```
-upper = 4.439, observed = 1.77
 
@@
 ```{r}
-growth_y_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_1ha)
-growth_sy_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_1ha)
-AICtab(growth_y_1ha, growth_sy_1ha)
+growth_null_1ha <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_1ha)
+growth_y_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_1ha)
+growth_sy_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_1ha)
+AICtab(growth_null_1ha, growth_y_1ha, growth_sy_1ha)
 ```
+significant interaction
 
@@
 ```{r}
-surv_y_1ha <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
-surv_sy_1ha <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
-AICtab(surv_y_1ha, surv_sy_1ha)
+surv_null_1ha <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
+surv_y_1ha <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
+surv_sy_1ha <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
+AICtab(surv_null_1ha, surv_y_1ha, surv_sy_1ha)
 ```
+No interaction or effect of year.
 
@@
 ```{r}
-flwr_y_1ha <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
-flwr_sy_1ha <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
-AICtab(flwr_y_1ha, flwr_sy_1ha)
+flwr_null_1ha <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
+flwr_y_1ha <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
+flwr_sy_1ha <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
+AICtab(flwr_null_1ha, flwr_y_1ha, flwr_sy_1ha)
 ```
+significant interaction?
 
 ```{r}
-
 lrt_flwr_1ha <- sy_lrt(flwr_y_1ha, flwr_sy_1ha, nsim = 500)
-
 ```

### 2021-01-15 analysis/supplementary-materials/size-environment_interaction.Rmd
filter out unrealisticly tiny plants

@@
   filter(ht < 200, ht_next < 200) %>% 
-  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
+  filter(shts > 0, shts_next > 0, ht >= 1, ht_next >= 1) %>% 
   mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))

### 2021-01-18 analysis/paper/paper.Rmd
draft of methods section complete

@@
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004]; reproductive plants will have 1-X <!--# from data?
---> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+bourcieri* [@bruna2004]; reproductive plants will have 1-7 flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months.
-SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckee1993].
-We chose a 3 month time scale as most relevant for *H.
-acuminata* as plants do not access sources of water buffered over longer time scales such as ground water or flood water <!--# work on this sentence -->.
+SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckeeRelationshipDroughtFrequency1993].
+A 3 month time scale for SPEI was deemed as most relevant for *H.
+acuminata* as it relies primarily on soil moisture and not water sources buffered over longer timescales (e.g.
+it lacks deep taproots to access groundwater).
 We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database <!--# citation --> and TRMM <!--# citation --> and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1)<!--# ref to appendix 1 -->.
+We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017] and TRMM [@trmm3b43] and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1).
 
@@
 This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
-Crossbasis functions were created using the *dlnm* package in R with a maximum lag of \_\_\_\_ <!--# how far back might drought impact heliconia?
--->months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-This particular maximum lag time was determined by ....
-<!--# preliminary experiments?
-theory?
-limitations of data?
--->.
+Crossbasis functions were created using the *dlnm* package in R with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
+This particular maximum lag time was chosen because previous experiments showed that *H.
+acuminata* individuals recovered from transplant shock after 36 months, so we expected at most 36 months lag time in the response to drought [@brunaExperimentalAssessmentHeliconia2002].
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
 In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.
-Cubic regression splines were used for both dimensions of the crossbasis function, with \_\_\_ <!--# same as number of lag steps -1 --> knots for the lag time and 3 knots for the drought response dimension to restrict the response to drought to bimodal at most complex.
 
-To model growth, we used shoot number as a proxy for plant size.
-Shoot number is easy to measure accurately and, unlike plant height, does not change if a plant is knocked over by a treefall.
-<!--# I could consider an integrative measure like log(shts * ht) -->.
-Size in year $t+1$ was modeled as a function of the natural log of size in year $t$ and drought history (i.e.
-the crossbasis function) with plot as a random effect.
-We used a scaled t error family distribution with a log link since residuals were leptokurtic when a gaussian distribution was used.
+The general form of the vital rate models was as follows:
 
-For survival, plants were assumed dead the year after they were last observed if the plant was not observed for three or more years.
-Survival and flowering probability were similarly modeled as a function of size and drought history (i.e.
-the crossbasis function) with a random effect of plot, but with a binomial family error distribution and a logit link function.
+$$
+g[\textrm{E}(y_{i})] = \alpha_j + f_1(x_i) + f_2(d_{i,t}, ..., d_{i,t-24}) \\
+\alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
+$$
 
-Statistical significance of crossbasis functions and covariates was determined by ....
+Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+2017 for details).
+Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex.
+A random effect of plot ID on the intercept is represented by $a_j$.
+For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$).
+A scaled t family was chosen because residuals were leptokurtic with a gaussian error structure.
+The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescence).
+For survival and flowering as a response, we used a binomial family error distribution with a logit link.
 
-To better understand the shape of the crossbasis function, we calculated fitted values for an average sized plant given a range of possible SPEI values from \_\_\_ to \_\_\_ at different lag times from \_\_\_ to \_\_\_ months, holding all other SPEI values at the average across the dataset (\_\_\_).
-These results could then be plotted as heatmaps with the *ggplot2* package ( <!--# cite ggplot2 --> ).
+------------------------------------------------------------------------
 
-One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
-This limitation meant that we were unable to model a potential size by environment (i.e.
-drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
-To check for a size by environment interaction, we followed the methods of Tredennick et al.
-[-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package <!--# cite lme4 -->.
-Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
-For each vital rate, we included an additional random effect of year on either the intercept or both the intercept and slope of the relationship between plant size and the response.
-We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].
+NOTE FOR EMILIO: In Gasparrini et al. 2017, the crossbasis function is defined like so:
+
+$$
+s(x_{i,t-l_0}, ...,x_{i,t-L}) = \sum_{l = l_0}^L f \cdot w(x_{i,t-l},l) 
+$$
+
+Where "the bi-dimensional dose-lag-response function $f \cdot w (x,l)$ is composed of two marginal functions: the standard dose-response function $f(x)$, and the additional lag-response function $w(l)$ that models the lag structure...".
+I like this summation notation because it clarifies that the function is two-dimensional, but I'm not sure it's mathematically appropriate to drop into the model equation.\
+The crossbasis function could also be represented by something a lot more abstract like $f_2(X_i, L)$ where $X_i$ is the SPEI history and $L$ is a vector (0, ..., 36) describing the lag structure of $X_i$.
+This is closer to how the model is written in R.
+Let's talk about the clearest way to write and describe these models.
+
+------------------------------------------------------------------------
+
+Statistical significance of crossbasis functions and covariates was determined by .... <!--# not sure yet! -->
 
@@
 
+## DLNMs
+
 ## Size by year interaction
+
+<!--# this all needs updating.  Probably going to use AIC instead of these p-values since I think they're doing the same thing. -->
 
@@
 
-metapopulation dynamics stuff goes here.
+Caveats:
+
+<!--# need to edit.  Moved from methods. -->
+
+One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
+This limitation meant that we were unable to model a potential size by environment (i.e.
+drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
+To check for a size by environment interaction, we followed the methods of Tredennick et al.
+[-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package [@batesFittingLinearMixedeffects2015].
+Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
+For each vital rate, we essentially replaced the crossbasis function in Eq 1.
+with a random effect of year on either the intercept or both the intercept and relationship with plant size.
+We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].
 

### 2021-01-18 analysis/supplementary-materials/appendix_SPEI.Rmd
draft of methods section complete

@@
 -   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
-
+    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
     -   Resolution: 0.25º x 0.25º

### 2021-01-19 analysis/paper/vital rate models.Rmd
switched gam() to bam() for speed.  Added year as a random effect.  Attempt at plotting covariates (but not correct yet).

@@
 
-- Switch measure of plant size to shoots * height
+- ~~Switch measure of plant size to shoots * height~~
 - ~~Change lag to 36 months (3 years)~~
 - Produce ridge plot
+- Heatmap for growth should be ∆size
+- try select = TRUE (after you understand it better? blog about it first)
 
@@
   demog_full %>% 
-  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
+  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
+  mutate(year_fac = as.factor(year))
 # View(demog_full)
@@
 
+# Speed up models with bam()
+
+I'm going to start fitting these with `bam()` to speed things up.  This sets up the paralellization.  CAUTION: you may need to change these settings for your machine.
+
+```{r}
+cl <- parallel::makeForkCluster()
+```
+
 # Growth
@@
 g_post_cf <-
-  gam(log_size_next ~ 
+  bam(log_size_next ~ 
         s(log_size) + 
+        s(year_fac, bs = "re") + 
         s(plot, bs = "re") + #random effect of plot
@@
       data = demog_post_cf,
-      # gamma = 1.2, #I maybe need to optimize this with cross-validation?
-      method = "REML")
+      method = "REML",
+      cluster = cl)
 ```
@@
 AIC(g_post_cf)
-
 appraise(g_post_cf)
@@
 ```{r}
-plot(g_post_cf, select = 1, main = "select = FALSE")
-```
-
-Hmm. Make sure this doesn't predict negative sizes.
-
-```{r}
 summary(g_post_cf)
 ```
-
 Remember, R-sq is probably so high because size this year is a very good predictor of size next year.
@@
 ```{r}
+draw(g_post_cf)
+```
+
+```{r}
 g_post_1ha <-
-  gam(log_size_next ~ 
+  bam(log_size_next ~ 
         s(log_size) +
+        s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
       data = demog_post_1ha,
-      # gamma = 1.2, #increases smoothness when above 1
-      method = "REML")
+      method = "REML",
+      cluster = cl)
 ```
@@
 ```
+```{r}
+draw(g_post_1ha)
+```
+
+## Covariate plots
+
+Effect of size with all else held average.
 
 ```{r}
-plot(g_post_1ha, select = 1)
+range(demog_post_cf$log_size)
+
+newdata <- 
+  tibble(log_size = seq(0, 7.8, length.out = 100),
+         plot = "newplot",
+         year_fac = "newyear", 
+         spei_history = mean(demog_post_cf$spei_history),
+         L = demog_post_cf$L[1:100, ])
+
+out <- augment(g_post_cf, newdata = newdata, se_fit = TRUE)
+
+covar_growth_cf <-
+  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
+  geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
+  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
+  theme_linedraw() +
+  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
+covar_growth_cf
+```
+```{r}
+range(demog_post_1ha$log_size)
+
+newdata <- 
+  tibble(log_size = seq(0, 7.6, length.out = 100),
+         plot = "newplot",
+         year_fac = "newyear",
+         spei_history = mean(demog_post_1ha$spei_history),
+         L = demog_post_cf$L[1:100, ])
+out <- augment(g_post_1ha, newdata = newdata, se_fit = TRUE)
+
+covar_growth_1ha <-
+  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
+  geom_point(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.1) +
+  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  # geom_rug(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.25) +
+  theme_linedraw() +
+  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
+covar_growth_1ha
 ```
@@
   dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
+                   year_fac = "newyear",
                    plot = "newlevel")
@@
 ```{r}
-pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data)
-pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
+pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
 ```
@@
 
-**Raster plots:**
+TODO: make color be ∆size (i.e. fitted - mean size from ref_data)
 
@@
   scale_x_continuous("SPEI_3", expand = c(0,0)) +
-  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
-  labs(title = "Continuous Forest",
-       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
+  scale_fill_viridis_c("size(t+1)", option = "A", limits = exp(shoot_lim)) +
+  labs(title = "Continuous Forest") +
   theme_bw()
@@
   scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
-  labs(title = "1-ha Fragments",
-       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
+  labs(title = "1-ha Fragments") +
   theme_bw()
@@
 **Slice through individual months**
+TODO: Change to ridge plots?
 
@@
 s_post_cf <-
-  gam(surv ~ 
+  bam(surv ~ 
         s(log_size) +
+        s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
       data = demog_post_cf,
-      # gamma = 1.2, #increases smoothness when above 1
-      method = "REML")
+      method = "REML",
+      cluster = cl)
 ```
@@
 # gam.check(s_post_cf) #not sure what to expect from any of these plots
-appraise(s_post_cf, method = "simulate")
+appraise(s_post_cf)
 ```
@@
 ```
-wow, not significant!
+
+```{r}
+draw(s_post_cf)
+```
 
@@
 s_post_1ha <-
-  gam(surv ~ 
+  bam(surv ~ 
         s(log_size) +
+        s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
 ```
+
 not significant!  I should try select=TRUE here.
+
+```{r}
+draw(s_post_1ha)
+```
+
+## Covariate plots
+
+Effect of size with all else held average.
+
+```{r}
+range(demog_post_cf$log_size)
+newdata <- 
+  tibble(log_size = seq(0, 7.8, length.out = 100),
+         plot = "newplot",
+         year_fac = "newyear",
+         spei_history = mean(demog_post_cf$spei_history),
+         L = demog_post_cf$L[1:100, ])
+out <- augment(s_post_cf, newdata = newdata, se_fit = TRUE)
+
+covar_s_cf <-
+  ggplot(out, aes(x = log_size, y = plogis(.fitted))) + 
+  # geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
+  theme_linedraw()
+covar_s_cf
+```
 
@@
 ```{r}
-pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
-pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
+pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
 ```
@@
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest",
-       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
+  labs(title = "Continuous Forest") +
   theme_bw()
@@
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments",
-       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
+  labs(title = "1-ha Fragments") +
   theme_bw()
@@
 f_post_cf <-
-  gam(flwr ~ 
+  bam(flwr ~ 
         s(log_size) +
+        s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
 ```
+these residuals are troubling
 
@@
 ```{r}
-plot(f_post_cf, select = 1, trans = plogis)
+draw(f_post_cf)
 ```
-
-Plants with exp(3) = 20 shoots have close to a 100% chance of flowering.
 
@@
 f_post_1ha <-
-  gam(flwr ~ 
+  bam(flwr ~ 
         s(log_size) +
+        s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
 ```{r}
-plot(f_post_1ha, select = 1, trans = plogis)
+draw(f_post_1ha)
 ```
@@
 ```{r}
-pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
-pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
+pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
 ```

### 2021-01-20 analysis/paper/vital rate models.Rmd
Changed growth plot to ∆size.  Worked on axis labels and exporting covariate plots. Closes #21

@@
 library(gratia)
+library(latex2exp)
 
@@
 ```{r}
-cl <- parallel::makeForkCluster()
+library(parallel)
+ncores <- detectCores()
+cl <- makeForkCluster(ncores - 2)
 ```
@@
 ```{r}
-range(demog_post_cf$log_size)
+sr <- range(demog_post_cf$log_size)
 
-newdata <- 
-  tibble(log_size = seq(0, 7.8, length.out = 100),
+newdata_covar_cf <- 
+  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
          plot = "newplot",
@@
 
-out <- augment(g_post_cf, newdata = newdata, se_fit = TRUE)
+out <- augment(g_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
 
@@
   theme_linedraw() +
-  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
+  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
 covar_growth_cf
+ggsave(here("analysis", "figures", "growth_cf_size.png"), covar_growth_cf)
 ```
 ```{r}
-range(demog_post_1ha$log_size)
-
-newdata <- 
-  tibble(log_size = seq(0, 7.6, length.out = 100),
+sr <- range(demog_post_1ha$log_size)
+newdata_covar_1ha <- 
+  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
          plot = "newplot",
@@
          L = demog_post_cf$L[1:100, ])
-out <- augment(g_post_1ha, newdata = newdata, se_fit = TRUE)
+
+out <- augment(g_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
 
@@
   theme_linedraw() +
-  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
+  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
+
 covar_growth_1ha
+ggsave(here("analysis", "figures", "growth_1ha_size.png"), covar_growth_1ha)
 ```
@@
                    plot = "newlevel")
-# exp(1.032814) # 2.8 shoots
-mean_size <- round(exp(ref_data$log_size), 1)
+
+mean_size <- exp(ref_data$log_size)
 ```
@@
 
+Calculate growth as change from year t to year t+1
+
+```{r}
+mean_size # size in year t for fitted values
+
+pred_g_post_cf <- 
+  pred_g_post_cf %>% 
+  mutate(e_fitted = exp(fitted),
+         d_fitted = e_fitted - mean_size)
+
+pred_g_post_1ha <-
+  pred_g_post_1ha %>% 
+  mutate(e_fitted = exp(fitted),
+         d_fitted = e_fitted - mean_size)
+```
+
 Shared color bar limits so scales can be combined.
@@
 ```{r}
-shoot_lim <- c(min(pred_g_post_1ha$fitted, pred_g_post_cf$fitted),
-               max(pred_g_post_1ha$fitted, pred_g_post_cf$fitted))
+shoot_lim <- c(min(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted),
+               max(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted))
 ```
@@
 ## Plots
-
-TODO: make color be ∆size (i.e. fitted - mean size from ref_data)
 
@@
   filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
+  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_growth_cf
+ggsave(here("analysis", "figures", "growth_cf_spei.png"), p_growth_cf)
+```
+
+```{r}
+p_growth_1ha <-
+  pred_g_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
   geom_raster(interpolate = TRUE) +
@@
   ) +
-  scale_x_continuous("SPEI_3", expand = c(0,0)) +
-  scale_fill_viridis_c("size(t+1)", option = "A", limits = exp(shoot_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_growth_cf
-```
-
-```{r}
-p_growth_1ha <-
-  pred_g_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +  scale_x_continuous("SPEI_3", expand = c(0,0)) +
-  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
   labs(title = "1-ha Fragments") +
@@
 p_growth_1ha
+ggsave(here("analysis", "figures", "growth_1ha_spei.png"), p_growth_1ha)
 ```
@@
 ```{r}
-p_growth_cf_slice <-
-  pred_g_post_cf %>% 
-  filter(nearest(x, -1.50)) %>% 
-  ggplot(aes(x = lag, y = exp(fitted))) +
-  geom_ribbon(aes(
-    ymin = exp(fitted - se.fit * 1.96),
-    ymax = exp(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "Continuous Forest", y = "shoots (t+1)") +
-  theme_linedraw()
-p_growth_cf_slice
+# p_growth_cf_slice <-
+#   pred_g_post_cf %>% 
+#   filter(nearest(x, -1.50)) %>% 
+#   ggplot(aes(x = lag, y = exp(fitted))) +
+#   geom_ribbon(aes(
+#     ymin = exp(fitted - se.fit * 1.96),
+#     ymax = exp(fitted + se.fit * 1.96)
+#   ), alpha = 0.3)+
+#   geom_line() +
+#   scale_x_continuous("lag(months)", breaks = 0:36) +
+#   labs(title = "Continuous Forest", y = "shoots (t+1)") +
+#   theme_linedraw()
+# p_growth_cf_slice
 ```
@@
 ```{r}
-p_growth_1ha_slice <-
-  pred_g_post_1ha %>% 
-  filter(nearest(x, -1.5)) %>% 
-  ggplot(aes(x = lag, y = exp(fitted))) +
-  geom_ribbon(aes(
-    ymin = exp(fitted - se.fit * 1.96),
-    ymax = exp(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "1-ha Fragments", y = "shoots (t+1)") +
-  theme_linedraw()
-p_growth_1ha_slice
+# p_growth_1ha_slice <-
+#   pred_g_post_1ha %>% 
+#   filter(nearest(x, -1.5)) %>% 
+#   ggplot(aes(x = lag, y = exp(fitted))) +
+#   geom_ribbon(aes(
+#     ymin = exp(fitted - se.fit * 1.96),
+#     ymax = exp(fitted + se.fit * 1.96)
+#   ), alpha = 0.3)+
+#   geom_line() +
+#   scale_x_continuous("lag(months)", breaks = 0:36) +
+#   labs(title = "1-ha Fragments", y = "shoots (t+1)") +
+#   theme_linedraw()
+# p_growth_1ha_slice
 ```
@@
 ```{r}
-#TODO: replace with ridge plot
-plotdf <-
-  pred_g_post_cf %>% 
-  mutate(lag = round(lag)) %>% 
-  group_by(x, lag) %>% 
-  summarize(across(everything(),  mean))
-
-foo <- function(x) {
-  x <- as.numeric(x)
-  mon <- month(ymd("2020-2-01") - months(x), label = TRUE)
-  l <- glue("lag={x} ({mon})")
-  return(l)
-}
-
-p_growth_cf_months <-
-  plotdf %>% 
-  ggplot(aes(x = x, y = exp(fitted))) +
-  geom_ribbon(aes(
-    ymin = exp(fitted - se.fit * 1.96),
-    ymax = exp(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("SPEI3") +
-  scale_y_continuous("shoots (t+1)") +
-  facet_wrap(~lag, labeller = as_labeller(foo)) +
-  theme_bw()
-p_growth_cf_months
+# #TODO: replace with ridge plot
+# plotdf <-
+#   pred_g_post_cf %>% 
+#   mutate(lag = round(lag)) %>% 
+#   group_by(x, lag) %>% 
+#   summarize(across(everything(),  mean))
+# 
+# foo <- function(x) {
+#   x <- as.numeric(x)
+#   mon <- month(ymd("2020-2-01") - months(x), label = TRUE)
+#   l <- glue("lag={x} ({mon})")
+#   return(l)
+# }
+# 
+# p_growth_cf_months <-
+#   plotdf %>% 
+#   ggplot(aes(x = x, y = exp(fitted))) +
+#   geom_ribbon(aes(
+#     ymin = exp(fitted - se.fit * 1.96),
+#     ymax = exp(fitted + se.fit * 1.96)
+#   ), alpha = 0.3)+
+#   geom_line() +
+#   scale_x_continuous("SPEI3") +
+#   scale_y_continuous("shoots (t+1)") +
+#   facet_wrap(~lag, labeller = as_labeller(foo)) +
+#   theme_bw()
+# p_growth_cf_months
 ```
@@
 ```{r}
-range(demog_post_cf$log_size)
-newdata <- 
-  tibble(log_size = seq(0, 7.8, length.out = 100),
-         plot = "newplot",
-         year_fac = "newyear",
-         spei_history = mean(demog_post_cf$spei_history),
-         L = demog_post_cf$L[1:100, ])
-out <- augment(s_post_cf, newdata = newdata, se_fit = TRUE)
+out <- augment(s_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
 
 covar_s_cf <-
-  ggplot(out, aes(x = log_size, y = plogis(.fitted))) + 
-  # geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
+  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
   geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
   geom_line(color = "blue") +
-  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
+  labs(x = TeX("size_t_"), y = "P(survival)") +
   theme_linedraw()
+
 covar_s_cf
+ggsave(here("analysis", "figures", "surv_cf_size.png"), covar_s_cf)
+```
+
+Uh...There might be something wrong with how I've determined survival.  I should revisit this.
+
+```{r}
+out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
+
+covar_s_1ha <-
+  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  labs(x = TeX("size_t_"), y = "P(survival)") +
+  theme_linedraw()
+
+covar_s_1ha
+ggsave(here("analysis", "figures", "surv_1ha_size.png"), covar_s_1ha)
 ```
@@
 ```{r}
+#mean survival
+x <- demog_post$surv
+sum(x, na.rm = TRUE) / length(x)
+```
+
+```{r}
 p_surv_cf <-
@@
     sec.axis = dup_axis(
-      name = "Month",
+      name = "month",
       labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) + 
-  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
@@
 p_surv_cf
+ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
 ```
@@
     sec.axis = dup_axis(
-      name = "Month",
+      name = "month",
       labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) +
-  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
   scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
@@
 p_surv_1ha
+ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
 ```
@@
 ```{r}
-x <- demog_post_cf$flwr
+x <- demog_post$flwr
 sum(x, na.rm = TRUE) / length(x)
@@
 
+Doesn't converge with random effect of year
+
 ```{r}
 appraise(f_post_cf)
+# concurvity(f_post_cf, full = FALSE)
 ```
@@
 
+## Covariate plots
+Effect of size with all else held average.
+
+```{r}
+out <- augment(f_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
+
+covar_f_cf <-
+  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_cf) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  labs(x = TeX("size_t_"), y = "P(flowering)") +
+  theme_linedraw()
+
+covar_f_cf
+ggsave(here("analysis", "figures", "flwr_cf_size.png"), covar_f_cf)
+```
+
+```{r}
+out <- augment(f_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
+
+covar_f_1ha <-
+  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_1ha) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  labs(x = TeX("size_t_"), y = "P(flowering)") +
+  theme_linedraw()
+
+covar_f_1ha
+ggsave(here("analysis", "figures", "flwr_1ha_size.png"), covar_f_1ha)
+```
 ## Get plot data
@@
     sec.axis = dup_axis(
-      name = "Month",
+      name = "month",
       labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
   ) + 
-  scale_x_continuous("SPEI_3", expand = c(0,0)) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
   scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
@@
 p_flowered_cf
+ggsave(here("analysis", "figures", "flwr_cf_spei.png"), p_flowered_cf)
 ```
@@
 p_flowered_1ha
+ggsave(here("analysis", "figures", "flwr_1ha_spei.png"), p_flowered_1ha)
 ```
@@
   (p_growth_cf | p_growth_1ha) +
-  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
+  plot_annotation(title = "Effect of drought on growth") +
   plot_layout(guides = "collect") & theme_talk & labs(caption = "")
 p_growth_spei
-ggsave(here("analysis", "figures", "growth_spei.png"), p_growth_spei, height = 7, width = 14)
+# ggsave(here("analysis", "figures", "growth_spei.png"), p_growth_spei, height = 7, width = 14)
 ```
-
-It looks like drought in September/October is bad for growth.
-
-Reminder that these are still fit with two different models and might not be entirely comparable.  Different penalties to the smooths make these hard to compare as well.
-
-Here's a slice through SPEI = -1.5 (severe drought)
-
-```{r}
-(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
-  p_growth_1ha_slice +
-  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
-ggsave(here("analysis", "figures", "growth slice.png"))
-```
-
-Again, you can see that drought immediately preceding the census (Jan of that year) and in the previous October are bad for growth, but dry Decembers and Septembers are good for growth.
 
@@
 p_surv_spei
-ggsave(here("analysis", "figures", "surv_spei.png"), p_surv_spei, width = 14, height = 7)
+# ggsave(here("analysis", "figures", "surv_spei.png"), p_surv_spei, width = 14, height = 7)
 ```
-
-The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.
-In continuous forest, wet conditions or very dry conditions in the previous August–October have a positive effect on survival.
-
-Here's a slice through SPEI = -1.5 (severe drought)
-
-```{r}
-((p_surv_cf_slice + theme(axis.text.x = element_blank())) / p_surv_1ha_slice) +
-  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
-ggsave(here("analysis", "figures", "survival slice.png"))
-```
-Pretty flat.
 
@@
 p_flowered_spei
-ggsave(here("analysis", "figures", "flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
+# ggsave(here("analysis", "figures", "flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
 ```
-
-The surface for continuous forest is perfectly flat (did I do something wrong with the color scale?) and for 1-ha fragments there is a strange (anomalous?) negative effect of high SPEI (wet conditions) in the summer 2 years ago.
-
-Here's a slice through SPEI = -1.5 (severe drought)
-
-```{r}
-((p_flowered_cf_slice + theme(axis.text.x = element_blank())) / p_flowered_1ha_slice) +
-  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
-ggsave(here("analysis", "figures", "flowering slice.png"))
-```

### 2021-01-25 analysis/paper/vital rate models.Rmd
calculate size_prev instead of size_next and switch models to use size_prev as covariate instead of size.  This means SPEI history starts at the month of the census when the response variables were recorded.

@@
 ## Prep precip data
+
 Get only rows for february (the month of the census), and year for joining
@@
 ```
+
 So, from -2.25 to -2.5
@@
   demog_full %>% 
-  filter(ht < 200, ht_next < 200) %>% 
-  filter(shts > 0, shts_next > 0, ht > 0, ht_next > 0) %>% 
-  mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
+  filter(ht < 200, ht_prev < 200) %>% 
+  filter(shts > 0, shts_prev > 0, ht > 0, ht_prev > 0) %>% 
+  mutate(size = shts*ht, size_prev = shts_prev * ht_prev, log_size = log(size), log_size_prev = log(size_prev))
 ```
@@
 
-Fit the GAM with crossbasis smooth for lagged drought
+Fit the GAM with crossbasis smooth for lagged drought.
+
+Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).
 
@@
 g_post_cf <-
-  bam(log_size_next ~ 
-        s(log_size) + 
-        s(year_fac, bs = "re") + 
+  bam(log_size ~ 
+        s(log_size_prev) + 
+        # s(year_fac, bs = "re") + 
         s(plot, bs = "re") + #random effect of plot
@@
       method = "REML",
+      select = TRUE,
       cluster = cl)
@@
 ```
+
 Remember, R-sq is probably so high because size this year is a very good predictor of size next year.
@@
 g_post_1ha <-
-  bam(log_size_next ~ 
-        s(log_size) +
-        s(year_fac, bs = "re") +
+  bam(log_size ~ 
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
       data = demog_post_1ha,
+      select = TRUE,
       method = "REML",
@@
 ```
+
 ```{r}
@@
 ```{r}
-sr <- range(demog_post_cf$log_size)
+sr <- range(demog_post_cf$log_size_prev)
 
 newdata_covar_cf <- 
-  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
+  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
          plot = "newplot",
@@
 covar_growth_cf <-
-  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
-  geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
+  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
+  geom_point(data = demog_post_cf, aes(x = size_prev, y = size), alpha = 0.1) +
   geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_abline(slope = 1, linetype = 2, color = "red") +
   geom_line(color = "blue") +
@@
 ```{r}
-sr <- range(demog_post_1ha$log_size)
+out %>%
+  filter(exp(log_size_prev) > exp(.fitted)) %>% # shinking plants
+  pull(log_size_prev) %>% min() %>% exp() #first size at which shrinking happens.
+```
+
+```{r}
+sr <- range(demog_post_1ha$log_size_prev)
 newdata_covar_1ha <- 
-  tibble(log_size = seq(sr[1], sr[2], length.out = 100),
+  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
          plot = "newplot",
@@
 covar_growth_1ha <-
-  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
-  geom_point(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.1) +
+  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
+  geom_point(data = demog_post_1ha, aes(x = size_prev, y = size), alpha = 0.1) +
   geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_abline(slope = 1, linetype = 2, color = "red") +
   geom_line(color = "blue") +
@@
 ggsave(here("analysis", "figures", "growth_1ha_size.png"), covar_growth_1ha)
+```
+```{r}
+out %>%
+  filter(exp(log_size_prev) > exp(.fitted)) %>% #shrinking plants
+  pull(log_size_prev) %>% min() %>% exp() #smallest size where plants start shrinking
 ```
@@
   demog_post %>%
-  dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
+  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
                    year_fac = "newyear",
@@
 
-mean_size <- exp(ref_data$log_size)
+mean_size <- exp(ref_data$log_size_prev)
 ```
@@
   bam(surv ~ 
-        s(log_size) +
-        s(year_fac, bs = "re") +
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
       method = "REML",
+      select = TRUE,
       cluster = cl)
 ```
+
 ```{r}
@@
 covar_s_cf <-
-  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
   stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
@@
 covar_s_1ha <-
-  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
   stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
@@
 
-Around 0.05 percent of plants flowered.
+Around 5 percent of plants flowered.
 
@@
   bam(flwr ~ 
-        s(log_size) +
-        s(year_fac, bs = "re") + 
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") + 
         s(plot, bs = "re") +
@@
 ```
+
 these residuals are troubling
@@
 
+This model changes **drastically** depending on whether year is included as a random effect.
+Does that mean ther random effect is important?
+Or is it because it is collinear / concurve with SPEI history?
+
 ```{r}
@@
   bam(flwr ~ 
-        s(log_size) +
-        s(year_fac, bs = "re") +
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
 ## Covariate plots
+
 Effect of size with all else held average.
@@
 covar_f_cf <-
-  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
   stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_cf) +
@@
 covar_f_1ha <-
-  ggplot(out, aes(x = exp(log_size), y = plogis(.fitted))) + 
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
   stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_1ha) +
@@
 ```
+
 ## Get plot data

### 2021-01-25 analysis/wrangling/2-demography_wrangling.R

@@ -62,14 +62,14 @@ demog <-
   mutate(infl = ifelse(is.na(infl) & (!is.na(shts) | !is.na(ht)), 0, infl)) %>% 
   mutate(flwr = ifelse(infl > 0, 1, 0), .after = infl)
 
-# Add year t+1 ------------------------------------------------------------
+# Add year t-1 ------------------------------------------------------------
 # add height and number shoots in the next year for each observation
 
 demog <- 
   demog %>% 
   group_by(ha_id_number) %>% 
-  mutate(ht_next = lead(ht),
-         shts_next = lead(shts)) %>% 
+  mutate(ht_prev = lag(ht),
+         shts_prev = lag(shts)) %>% 
   ungroup()
 
 # arrange columns
@@ -78,7 +78,7 @@ demog <-
   select(ranch, bdffp_reserve_no, plot, habitat, #site level
          row, column, x_09, y_09, ha_id_number, tag_number, #plot level
          year,
-         ht, ht_next, shts, shts_next, infl_num = infl, flwr, surv, code_notes, code2) #plant level
+         ht, ht_prev, shts, shts_prev, infl_num = infl, flwr, surv, code_notes, code2) #plant level
 
 
 write_csv(demog, here("analysis", "data", "derived_data", "ha_survey.csv"))

### 2021-01-26 analysis/paper/vital rate models.Rmd
use all data for survival, but subset data for growth and flowering to be only plants that survived (and therefore have size measurements).

@@
   demog_full %>% 
-  filter(ht < 200, ht_prev < 200) %>% 
-  filter(shts > 0, shts_prev > 0, ht > 0, ht_prev > 0) %>% 
+  filter(ht < 200 | is.na(ht), ht_prev < 200 | is.na(ht_prev)) %>%
+  filter(
+    # having no aboveground biomass is possible, but categorically different than just being small. Exclude plants with 0 shoots or 0 height
+    shts_prev > 0,
+    ht_prev > 0,
+    #for survival data, must include plants with NA for shts and height (i.e. dead plants)
+    shts > 0 | is.na(shts), 
+    ht > 0 | is.na(ht)
+  ) %>%
   mutate(size = shts*ht, size_prev = shts_prev * ht_prev, log_size = log(size), log_size_prev = log(size_prev))
@@
 
+# Inspect raw data
+
+```{r}
+demog_post %>% ggplot(aes(x = surv)) + geom_bar() + facet_wrap(~habitat)
+```
+```{r}
+demog_post %>% ggplot(aes(x = habitat, y = log_size)) + geom_boxplot()
+```
+```{r}
+demog_post %>% ggplot(aes(x = flwr)) + geom_bar() + facet_wrap(~habitat)
+```
 # Speed up models with bam()
@@
 
+# Survival
+`surv` is 1 in `year`s when a plant is alive and 0 in the first `year` when the plant is dead.
+These GAMs model the survival of a plant (over the past year) as a function of size in the previous census and SPEI values starting at february of the current year going back 36 months.
+## Fit GAMs
+
+```{r}
+s_post_cf <-
+  bam(surv ~ 
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") +
+        s(plot, bs = "re") +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_cf,
+      method = "REML",
+      select = TRUE,
+      cluster = cl)
+```
+
+```{r}
+# gam.check(s_post_cf) #not sure what to expect from any of these plots
+appraise(s_post_cf)
+```
+
+```{r}
+summary(s_post_cf)
+```
+
+```{r}
+draw(s_post_cf)
+```
+
+```{r}
+s_post_1ha <-
+  bam(surv ~ 
+        s(log_size_prev) +
+        # s(year_fac, bs = "re") +
+        s(plot, bs = "re") +
+        s(spei_history, L,
+          bs = "cb",
+          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
+          xt = list(bs = "cr")),
+      family = binomial,
+      data = demog_post_1ha,
+      method = "REML",
+      select = TRUE,
+      cluster = cl)
+```
+
+```{r}
+summary(s_post_1ha)
+```
+
+```{r}
+draw(s_post_1ha)
+```
+
+## Covariate plots
+
+Effect of size with all else held average.
+
+```{r}
+sr <- range(demog_post_cf$log_size_prev)
+
+newdata_covar_cf <- 
+  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
+         plot = "newplot",
+         year_fac = "newyear", 
+         spei_history = mean(demog_post_cf$spei_history),
+         L = demog_post_cf$L[1:100, ])
+```
+
+```{r}
+out <- augment(s_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
+
+covar_s_cf <-
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  labs(x = TeX("size_t_"), y = "P(survival)") +
+  theme_linedraw()
+
+covar_s_cf
+ggsave(here("analysis", "figures", "surv_cf_size.png"), covar_s_cf)
+```
+
+```{r}
+sr <- range(demog_post_1ha$log_size_prev)
+newdata_covar_1ha <- 
+  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
+         plot = "newplot",
+         year_fac = "newyear",
+         spei_history = mean(demog_post_1ha$spei_history),
+         L = demog_post_cf$L[1:100, ])
+
+out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
+
+covar_s_1ha <-
+  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
+  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
+  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
+  geom_line(color = "blue") +
+  labs(x = TeX("size_t_"), y = "P(survival)") +
+  theme_linedraw()
+
+covar_s_1ha
+ggsave(here("analysis", "figures", "surv_1ha_size.png"), covar_s_1ha)
+```
+
+## Get plot data
+
+```{r}
+# use mean size plant across both habitats so plots are comparable
+ref_data <-
+  demog_post %>%
+  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
+                   year_fac = "newyear",
+                   plot = "newlevel")
+
+mean_size <- exp(ref_data$log_size_prev)
+```
+
+This is pretty slow. It could probably be faster if I understood matrix math better.
+
+```{r}
+pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+```
+
+Shared color bar limits so scales can be combined.
+
+```{r}
+surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
+              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
+# plogis(surv_lim)
+```
+
+## Plots
+
+```{r}
+#mean survival
+x <- demog_post$surv
+sum(x, na.rm = TRUE) / length(x)
+```
+
+```{r}
+p_surv_cf <-
+  pred_s_post_cf %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) + 
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_surv_cf
+ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
+```
+
+```{r}
+p_surv_1ha <-
+  pred_s_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
+  labs(title = "1-ha Fragments") +
+  theme_bw()
+p_surv_1ha
+ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
+```
+
+**Slice through time at severe drought**
+
+```{r}
+p_surv_cf_slice <-
+  pred_s_post_cf %>% 
+  filter(nearest(x, -1.50)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
+  labs(title = "Continuous Forest", y = "P(survival)") +
+  theme_linedraw()
+p_surv_cf_slice
+```
+
+```{r}
+p_surv_1ha_slice <-
+  pred_s_post_1ha %>% 
+  filter(nearest(x, -1.5)) %>% 
+  ggplot(aes(x = lag, y = plogis(fitted))) +
+  geom_ribbon(aes(
+    ymin = plogis(fitted - se.fit * 1.96),
+    ymax = plogis(fitted + se.fit * 1.96)
+  ), alpha = 0.3)+
+  geom_line() +
+  scale_x_continuous("lag(months)", breaks = 0:36) +
+  labs(title = "1-ha Fragments", y = "P(survival)") +
+  theme_linedraw()
+p_surv_1ha_slice
+```
+
 # Growth
+For mature plants that are alive...
 
-Start with no random effects
-
-Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.
+```{r}
+demog2_post_cf <- demog_post_cf %>% filter(surv == 1, !is.na(log_size))
+demog2_post_1ha <- demog_post_1ha %>% filter(surv == 1, !is.na(log_size))
+```
 
@@
 Fit the GAM with crossbasis smooth for lagged drought.
+Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.
+Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).
 
-Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).
+The transformed response and identity link is actually better than a log-link, according to Ellner, Childs and Rees.
 
@@
       family = scat(link = "identity"), #like leptokurtic gaussian.
-      data = demog_post_cf,
+      data = demog2_post_cf,
       method = "REML",
@@
 AIC for... 
-Gaussian: 40909.27. looks leptokurtic
-scaled t: 37284.47.  Waaay better
+Gaussian: ~40909.27. looks leptokurtic
+scaled t: ~37284.47.  Waaay better
 
@@
       family = scat(link = "identity"), #like leptokurtic gaussian.
-      data = demog_post_1ha,
+      data = demog2_post_1ha,
       select = TRUE,
@@
 ```{r}
-sr <- range(demog_post_cf$log_size_prev)
-
-newdata_covar_cf <- 
-  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
-         plot = "newplot",
-         year_fac = "newyear", 
-         spei_history = mean(demog_post_cf$spei_history),
-         L = demog_post_cf$L[1:100, ])
-
 out <- augment(g_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
@@
 ```{r}
-sr <- range(demog_post_1ha$log_size_prev)
-newdata_covar_1ha <- 
-  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
-         plot = "newplot",
-         year_fac = "newyear",
-         spei_history = mean(demog_post_1ha$spei_history),
-         L = demog_post_cf$L[1:100, ])
-
 out <- augment(g_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
@@
 This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r}
-# use mean size plant across both habitats so plots are comparable
-ref_data <-
-  demog_post %>%
-  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
-                   year_fac = "newyear",
-                   plot = "newlevel")
-
-mean_size <- exp(ref_data$log_size_prev)
-```
 
@@
 
-# Survival
-
-## Fit GAMs
-
-```{r}
-s_post_cf <-
-  bam(surv ~ 
-        s(log_size_prev) +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog_post_cf,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-```{r}
-# gam.check(s_post_cf) #not sure what to expect from any of these plots
-appraise(s_post_cf)
-```
-
-```{r}
-summary(s_post_cf)
-```
-
-```{r}
-draw(s_post_cf)
-```
-
-```{r}
-s_post_1ha <-
-  bam(surv ~ 
-        s(log_size_prev) +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog_post_1ha,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-```{r}
-summary(s_post_1ha)
-```
-
-not significant!  I should try select=TRUE here.
-
-```{r}
-draw(s_post_1ha)
-```
-
-## Covariate plots
-
-Effect of size with all else held average.
-
-```{r}
-out <- augment(s_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
-
-covar_s_cf <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(survival)") +
-  theme_linedraw()
-
-covar_s_cf
-ggsave(here("analysis", "figures", "surv_cf_size.png"), covar_s_cf)
-```
-
-Uh...There might be something wrong with how I've determined survival.  I should revisit this.
-
-```{r}
-out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
-
-covar_s_1ha <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(survival)") +
-  theme_linedraw()
-
-covar_s_1ha
-ggsave(here("analysis", "figures", "surv_1ha_size.png"), covar_s_1ha)
-```
-
-## Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r}
-pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r}
-surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
-              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
-# plogis(surv_lim)
-```
-
-## Plots
-
-These are not significant, so probably don't need to actually make these plots.  I'm mostly just curious what a non-significant crossbasis function looks like.
-
-```{r}
-#mean survival
-x <- demog_post$surv
-sum(x, na.rm = TRUE) / length(x)
-```
-
-```{r}
-p_surv_cf <-
-  pred_s_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_surv_cf
-ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
-```
-
-```{r}
-p_surv_1ha <-
-  pred_s_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_surv_1ha
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
-```
-
-**Slice through time at severe drought**
-
-```{r}
-p_surv_cf_slice <-
-  pred_s_post_cf %>% 
-  filter(nearest(x, -1.50)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "Continuous Forest", y = "P(survival)") +
-  theme_linedraw()
-p_surv_cf_slice
-```
-
-```{r}
-p_surv_1ha_slice <-
-  pred_s_post_1ha %>% 
-  filter(nearest(x, -1.5)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "1-ha Fragments", y = "P(survival)") +
-  theme_linedraw()
-p_surv_1ha_slice
-```
-
 # Flowering Probability
@@
       family = binomial,
-      data = demog_post_cf,
+      data = demog2_post_cf,
       method = "REML",
@@
       family = binomial,
-      data = demog_post_1ha,
+      data = demog2_post_1ha,
       method = "REML",

### 2021-01-27 analysis/paper/vital rate models.Rmd
added some EDA

@@
 
+## Survival
+
 ```{r}
@@
 ```
+
+```{r}
+demog_post %>% 
+  group_by(habitat) %>% 
+  summarize(surv = sum(surv, na.rm = TRUE) / n())
+```
+
+Survival nearly identical for 1 ha and CF.
+
+```{r}
+survival <-
+  demog_post %>% 
+  filter(habitat !="10-ha") %>% 
+  group_by(year, habitat) %>% 
+  summarize(surv = sum(surv, na.rm = TRUE)/n()) %>% 
+  ggplot(aes(x = year, y = surv, color = habitat)) + geom_line()
+survival / annotate_spei(spei)
+```
+Looks like survival went down year after 2003 drought and durring 2007 drought.
+
+## Size / growth
+
 ```{r}
@@
 ```
+
+```{r}
+size <- demog_post %>% 
+  ggplot(aes(x = year, y = size, color = habitat)) +
+  stat_summary(geom = "pointrange", fun.data = "mean_cl_boot") +
+  stat_summary(geom = "line", fun = mean)
+
+dates <- range(demog_post$year) %>% paste0(., "-01-01") %>% ymd()
+
+spei <- ggplot(xa, aes(x = yearmonth, y = spei)) + 
+  geom_line(aes(group = latlon)) +
+  scale_x_yearmonth(limits = dates)
+size/annotate_spei(spei)
+```
+Average plant size shrank in 2003 drought, but not 2007 drought
+
+## Flowering
+
 ```{r}
@@
 ## Plots
-
-```{r}
-#mean survival
-x <- demog_post$surv
-sum(x, na.rm = TRUE) / length(x)
-```
 

### 2021-01-28 analysis/paper/paper.Rmd
mention separate models for CF and 1-ha
EB Edits to Intro

@@
 1.
-Organisms experience multiple sources of natural and anthropogenic stress simultaneously.
-However, the impacts of multiple sources of stress on population dynamics are not well understood.
-Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics.
+Organisms simultanously experience multiple forms of natural and anthropogenic disturbance.
+While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
+They could act in concert to exacerbate or dampen population dynamics.
 <!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].
@@
 2.
-Habitat fragmentation is one such source of environmental stress.
-Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest farther than 1 km from an edge has increased to 75% compared to 90% historically [@haddadHabitatFragmentationIts2015].
+Habitat fragmentation is a major source of environmental stress.
+**[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]**  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+Of particualar concern are the effects of fragmentation on species-rich tropical ecosystems.
+In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest >1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
 Fragmentation can impact plant population dynamics through ...
@@
 3.
-Simultaneously, fragmented tropical forests are experiencing increased drought stress as a result of climate change.
-The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
-Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
+In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change.
+For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
+Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
 
@@
 4.
-It has been hypothesized that drought and fragmentation could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
-Fragmentation results in changes in forest microclimate near edges including increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
-This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest.
-For example, drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
+It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+Fragmentation drives several changes in microclimate near forest edges, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
+This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
 
-5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.
+5.  Alternatively, the populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation.
 
@@
 
-Second, there is a growing literature on how tropical plants respond to severe droughts, but few studies have compared the responses of plants in continuous forest with those in fragments.
+Second, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.
 Laurance et al.
@@
 
-We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
-Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
+We investigated the effects of drought on the growth, survival, reproduction, and recruitment of a understory herb in an experimentally fragmented landscape in the Central Amazon.
+Specifically, we asked 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig.
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig.
 -- <!--# fig ref -->).
-These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
-The vegetation regenerating around fragments is cleared periodically to ensure their isolation [@bierregaardLessonsAmazoniaEcology2001].
+These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried.
+In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
+
 The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
@@
 
+Because DLNMs do not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+
 Statistical significance of crossbasis functions and covariates was determined by .... <!--# not sure yet! -->

### 2021-01-29 analysis/paper/paper.Rmd
erdit spei calcs parag

@@
 
-The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
-The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content.
-Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).
+The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
+The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
+There is a pronounced dry season from June to October **(Appendix?)**.
 
@@
 
-*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H.
-acuminata*---are found at lower densities in shaded forest understories.
-Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a].
+*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
+While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H.
+acuminata*---are usually found at lower densities and in the more shaded forest understory.
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
 In our sites *H.
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004]; reproductive plants will have 1-7 flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 
-In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H.
-acuminata*.
+In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
 The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
 Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
-Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf.
+Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
+We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
 Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
@@
 
-Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
-In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method.
-Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+The estimates of potential evapotranspiration were calculated using the Pennman-Monteith method, and then along with monthly precipitation were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
 SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
-More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months.
-SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckeeRelationshipDroughtFrequency1993].
-A 3 month time scale for SPEI was deemed as most relevant for *H.
-acuminata* as it relies primarily on soil moisture and not water sources buffered over longer timescales (e.g.
-it lacks deep taproots to access groundwater).
-We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017] and TRMM [@trmm3b43] and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1).
+Positive SPEI values indicate wetter than average months<!-- instead: "...indicate the specified time period is wetter than average." because time frame doesn't have to be a month, yes?
+-->, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (< -2) [@mckeeRelationshipDroughtFrequency1993].
+Because *H.
+acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).
 

### 2021-01-29 analysis/paper/paper.Rmd
edits, back to Eric

@@
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, called a crossbasis function.
-This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
-Crossbasis functions were created using the *dlnm* package in R with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-This particular maximum lag time was chosen because previous experiments showed that *H.
-acuminata* individuals recovered from transplant shock after 36 months, so we expected at most 36 months lag time in the response to drought [@brunaExperimentalAssessmentHeliconia2002].
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
+DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, referred to as the crossbasis function.
+This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
+Crossbasis functions were created using the *dlnm* package with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
+This particular maximum lag time in respont to drought was chosen because *H.
+acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
 In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.
@@
 For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$).
-A scaled t family was chosen because residuals were leptokurtic with a gaussian error structure.
-The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescence).
-For survival and flowering as a response, we used a binomial family error distribution with a logit link.
+A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
+The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences).
+When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.
 

### 2021-01-29 analysis/paper/paper.Rmd
work on results

--- 2021-01-28 analysis/paper/paper.Rmd
@@
 1.
-Organisms simultanously experience multiple forms of natural and anthropogenic disturbance.
-While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-They could act in concert to exacerbate or dampen population dynamics.
+Organisms experience multiple sources of natural and anthropogenic stress simultaneously.
+However, the impacts of multiple sources of stress on population dynamics are not well understood.
+Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics.
 <!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].
@@
 2.
-Habitat fragmentation is a major source of environmental stress.
-**[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]**  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-Of particualar concern are the effects of fragmentation on species-rich tropical ecosystems.
-In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest >1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
+Habitat fragmentation is one such source of environmental stress.
+Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest farther than 1 km from an edge has increased to 75% compared to 90% historically [@haddadHabitatFragmentationIts2015].
 Fragmentation can impact plant population dynamics through ...
@@
 3.
-In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change.
-For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
-Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
+Simultaneously, fragmented tropical forests are experiencing increased drought stress as a result of climate change.
+The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
+Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
 
@@
 4.
-It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
-Fragmentation drives several changes in microclimate near forest edges, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
-This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
-This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
+It has been hypothesized that drought and fragmentation could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+Fragmentation results in changes in forest microclimate near edges including increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest.
+For example, drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
 
-5.  Alternatively, the populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation.
+5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.
 
@@
 
-Second, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.
+Second, there is a growing literature on how tropical plants respond to severe droughts, but few studies have compared the responses of plants in continuous forest with those in fragments.
 Laurance et al.
@@
 
-We investigated the effects of drought on the growth, survival, reproduction, and recruitment of a understory herb in an experimentally fragmented landscape in the Central Amazon.
-Specifically, we asked 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
+We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.
+Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig.
+In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig.
 -- <!--# fig ref -->).
-These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried.
-In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
-
+These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried.
+The vegetation regenerating around fragments is cleared periodically to ensure their isolation [@bierregaardLessonsAmazoniaEcology2001].
 The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
@@
 
-## DLNMs
+Start with descriptive results of drought events---how many droughts, how severe, timing of droughts, etc.
+
+The size of plants significantly predicted their size in the following year in continuous forest (edf = 7.8, P \< 0.001 <!--# What stats to include?
+reference df and F also?
+Is this all better as a table?
+-->) and forest fragments (edf = 7.4, P \< 0.001).
+In continuous forest, plants under a size of 259 were predicted to grow from year t-1 to year t, and the fit line was below 1:1 for larger plants, indicating that larger plants tended to shrink, on average.
+In 1 ha fragments, the fit line followed a 1:1 line more closely, indicating that plants tended to stay the same size on average from year to year over the course of this study.
+<!--# something weird about backtransformation causing these strange results?
+--> Drought history also had a significant effect on plant size in both continuous forest (edf = 14.7, p \< 0.001) and fragments (edf = 17.8, p \< 0.001).
+<!--# Describe seasonality of drought effects / shape of crossbasis function -->
+
+Survival of post-seedling plants was very high (93.3%) and was higher, overall, in 1 ha fragments than in continuous forest.
+Survival was related to plant size (edf = , p = ) such that survival was essentially 1 for plants larger than size \_\_\_ in continuous forest and larger than size \_\_\_ in 1 ha fragments.
 

### 2021-01-29 analysis/paper/vital rate models.Rmd
work on results

@@
 ```{r}
+dates <- range(demog_post$year) %>% paste0(., "-01-01") %>% ymd()
+
+spei <- ggplot(xa, aes(x = yearmonth, y = spei)) + 
+  geom_line(aes(linetype = latlon)) +
+  scale_x_yearmonth(limits = dates)
+spei <- annotate_spei(spei)
 survival <-
@@
   ggplot(aes(x = year, y = surv, color = habitat)) + geom_line()
-survival / annotate_spei(spei)
+survival / spei
 ```
@@
 
-dates <- range(demog_post$year) %>% paste0(., "-01-01") %>% ymd()
-
-spei <- ggplot(xa, aes(x = yearmonth, y = spei)) + 
-  geom_line(aes(group = latlon)) +
-  scale_x_yearmonth(limits = dates)
-size/annotate_spei(spei)
+size/spei
 ```
@@
 ggsave(here("analysis", "figures", "surv_cf_size.png"), covar_s_cf)
+```
+```{r}
+demog_post_cf %>% filter(size_prev > 1500, surv == 0)
+
+demog %>% filter(ha_id_number == 5310) %>% View()
 ```

### 2021-02-04 analysis/paper/vital rate models.Rmd
test out drawing the basis function itself
work on figure 1, exploratory data visualizations

--- 2021-01-26 analysis/paper/vital rate models.Rmd
@@
 ```{r}
+demog_plotdf <-  
+  demog_post %>% 
+  filter(habitat !="10-ha") %>% 
+  select(ranch, bdffp_reserve_no, plot, habitat, year, ht, shts, size, surv, flwr) %>% 
+  mutate(date = paste(year, "-02-01") %>% ymd(),
+         yearmonth = yearmonth(date))
+```
+
+## SPEI
+
+```{r}
+dates <- range(demog_post$year) %>% paste0(., "-02-01") %>% ymd()
+census_dates <- min(demog_post$year)[1]:max(demog_post$year) %>% paste(., "-02-01") %>% ymd()
+
+spei <-
+  xa %>%
+  select(latlon, yearmonth, spei) %>%
+  filter(!is.na(spei)) %>% 
+  ggplot(aes(x = yearmonth, y = spei)) + 
+  geom_line(aes(group = latlon), alpha = 0.2) +
+  stat_summary(geom = "line", fun = "mean") +
+  scale_y_continuous("SPEI", expand = expansion(mult = c(0, 0.05), add = 0)) +
+  scale_x_yearmonth("Date", limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+spei <- annotate_spei(spei)
+spei
+```
+
+## Survival
+
+```{r}
 demog_post %>% ggplot(aes(x = surv)) + geom_bar() + facet_wrap(~habitat)
 ```
+
+```{r}
+demog_post %>% 
+  group_by(habitat) %>% 
+  summarize(surv = sum(surv, na.rm = TRUE) / n())
+```
+
+Survival nearly identical for 1 ha and CF.
+
+### Timeseries
+
+```{r}
+survival <-
+  demog_plotdf %>%  
+  ggplot(aes(x = yearmonth, y = surv, color = habitat)) +
+  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
+  stat_summary(geom = "pointrange", fatten = 1) +
+  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  scale_y_continuous("P(survived)") +
+  theme_bw() +
+  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+survival
+# (survival / spei) 
+```
+Looks like survival went down year after 2003 drought and during 2007 drought.
+
+## Size / growth
+
 ```{r}
@@
 ```{r}
+demog_post %>% 
+  group_by(habitat) %>% 
+  summarize(size = mean(size, na.rm = TRUE),
+            shts = mean(shts, na.rm = TRUE),
+            ht = mean(ht, na.rm = TRUE))
+```
+
+### Timeseries
+
+```{r}
+size <-
+  demog_plotdf %>% 
+  ggplot(aes(x = yearmonth, y = size, color = habitat)) +
+  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
+  stat_summary(geom = "pointrange", fatten = 1) +
+  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  scale_y_continuous("Size") +
+  theme_bw() +
+  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+size
+# size/survival/spei
+```
+Average plant size shrank in 2003 drought, but not 2007 drought
+
+## Flowering
+
+```{r}
 demog_post %>% ggplot(aes(x = flwr)) + geom_bar() + facet_wrap(~habitat)
 ```
+### Timeseries
+
+```{r}
+flowering <-
+  demog_plotdf %>% 
+  ggplot(aes(x = yearmonth, y = flwr, color = habitat)) + 
+  stat_summary(geom = "line", fun = "mean") +
+  stat_summary(geom = "pointrange", fatten = 1) +
+  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  scale_y_continuous("P(flowering)") +
+  theme_bw() +
+  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+flowering
+```
+```{r}
+demog_plotdf %>% filter(year == 2003) %>% group_by(habitat) %>% summarize(flwr = mean(flwr, na.rm = TRUE))
+```
+
+```{r}
+size/survival/flowering/spei +
+  plot_layout(guides = "collect") &
+  theme(plot.margin = margin(2,1,1,1)) &
+  guides(col = guide_legend(title = "Habitat"))
+ggsave(here("analysis", "figures", "eda.png"), height = 7, width = 10)
+```
+
 # Speed up models with bam()
@@
   bam(surv ~ 
-        s(log_size_prev) +
+        s(log_size_prev, bs = "cr") +
+        # s(year) +
         # s(year_fac, bs = "re") +
@@
 appraise(s_post_cf)
+# concurvity(s_post_cf, full = FALSE)
 ```
@@
 ```
+```{r}
+demog_post_cf %>% filter(size_prev > 1500, surv == 0)
+
+demog %>% filter(ha_id_number == 5310) %>% View()
+```
 
@@
 ```{r}
-draw(g_post_1ha)
+# draw(g_post_1ha)
+draw(g_post_1ha, select = 3, continuous_fill = scale_fill_viridis_c(option = "inferno"), contour = FALSE, fun = "exp")
+ggsave("function_smooth-cb.png")
 ```

### 2021-02-04 analysis/paper/paper.Rmd
note on intro
work on results

--- 2021-01-29 analysis/paper/paper.Rmd
@@
 Habitat fragmentation is a major source of environmental stress.
-**[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]**  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-Of particualar concern are the effects of fragmentation on species-rich tropical ecosystems.
-In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest >1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
+**[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]** <!--# fragmentation results in increaed mortality due to increased stress.
+fragmentation itself is not the stressor -->Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+Of particular concern are the effects of fragmentation on species-rich tropical ecosystems.
+In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
 Fragmentation can impact plant population dynamics through ...
@@
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** <!--# check data --> flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
+Plants that were not found for three consecutive surveys were considered dead.
 In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings.
@@
 SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
-Positive SPEI values indicate wetter than average months<!-- instead: "...indicate the specified time period is wetter than average." because time frame doesn't have to be a month, yes?
--->, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (< -2) [@mckeeRelationshipDroughtFrequency1993].
+Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
 Because *H.
@@
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, referred to as the crossbasis function.
+DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
 This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
-Crossbasis functions were created using the *dlnm* package with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-This particular maximum lag time in respont to drought was chosen because *H.
+Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
+We chose 36 months as a maximum lag because *H.
 acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
-In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.
 
@@
 $$
-g[\textrm{E}(y_{i})] = \alpha_j + f_1(x_i) + f_2(d_{i,t}, ..., d_{i,t-24}) \\
+g[\textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
 \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
@@
 
-Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+Where $s_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
 2017 for details).
-Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex.
+The crossbasis function, $s_2(\cdot)$ can also be written:
+
+$$
+s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
+$$
+
+Where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodsimonn.ChapterSmoothers] and Gasparrini et al.
+[-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
+Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex.
 A random effect of plot ID on the intercept is represented by $a_j$.
@@
 
-------------------------------------------------------------------------
+In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity <!--# refrence Wood 2006?
+-->.
+We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].
 
-NOTE FOR EMILIO: In Gasparrini et al. 2017, the crossbasis function is defined like so:
-
-$$
-s(x_{i,t-l_0}, ...,x_{i,t-L}) = \sum_{l = l_0}^L f \cdot w(x_{i,t-l},l) 
-$$
-
-Where "the bi-dimensional dose-lag-response function $f \cdot w (x,l)$ is composed of two marginal functions: the standard dose-response function $f(x)$, and the additional lag-response function $w(l)$ that models the lag structure...".
-I like this summation notation because it clarifies that the function is two-dimensional, but I'm not sure it's mathematically appropriate to drop into the model equation.\
-The crossbasis function could also be represented by something a lot more abstract like $f_2(X_i, L)$ where $X_i$ is the SPEI history and $L$ is a vector (0, ..., 36) describing the lag structure of $X_i$.
-This is closer to how the model is written in R.
-Let's talk about the clearest way to write and describe these models.
-
-------------------------------------------------------------------------
-
-Because DLNMs do not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
-
-Statistical significance of crossbasis functions and covariates was determined by .... <!--# not sure yet! -->
+Because the `dlnm` package does not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
 
@@
 
-## DLNMs
+Start with descriptive results of drought events---how many droughts, how severe, timing of droughts, etc.
+
+## Survival
+
+Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
+Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004.
+Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, will get exact numbers from data).
+The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
+
+### Continuous Forest
+
+size: p \< 0.001 (figure = surv_cf_size.png)
+
+plot: p \< 0.001
+
+SPEI: p \< 0.001 (figure = surv_cf_spei.png)
+
+Lags 6-0 (Aug--Feb) and 20-15 (Jun--Nov) show reduced survival with drought.
+Lags 14-7 (Dec - Jul) and 26-21 (Dec-May) show increased survival with drought.
+Lag 36-33 (Feb - May) shows quadratic relationship with increased survival in dry and wet conditions (this seems like it's gotta be spurious).
+
+### 1-ha fragment
+
+size: p \<0.001 (figure = surv_1ha_size.png)
+
+plot: p = 0.258
+
+SPEI: p \< 0.001 (figure = surv_1ha_spei.png)
+
+Crossbasis is very flat with essentially no effect of SPEI beyond 6 months lag. Recent SPEI has bimodal effect with both low and high values reducing survival.
+
+**Question**: Currently the figures are expected values for the average plant across both plots.
+This makes it look like survival is a lot higher in 1-ha fragments when in reality survival is exactly the same (0.95) because plants are smaller in 1-ha fragments.
+Should these plots instead be for an average sized plant within the habitat type?
+
+## Growth
+
+Plants were on average larger in continuous forest compared to 1-ha fragments.
+In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150.
+In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
+
+The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the dataset with mean sizes becoming more similar over time.
+
+**NOTE**: I considered adding a covariate of year to account for yearly trends in growth, but since there is also a trend in SPEI over the time of this experiment, they were concurve (wiggly collinearity), and shouldn't both be included in model.
+
+### Continuous Forest
+
+size: p \< 0.001
+
+plot: p \< 0.001
+
+SPEI: p \< 0.001
+
+Predicted results show only lagged effects of SPEI on growth (f(x) at lags 0 and 1 is flat).
+Wet conditions in preceding winter months improve growth (lags 4-2, Oct - Dec).
+Positive effect of drought on growth at lags 11-8 (Mar - Jun) and 34-24 (nearly the entirety of year t-2).
+
+### 1-ha fragment
+
+size: p \< 0.001
+
+plot: p \< 0.001
+
+SPEI: p \< 0.001
+
+Crossbasis looks very similar to continuous forest but more extreme.
+SPEI has a stronger effect on growth in fragments.
+Wet conditions at lag 0 predict the greatest shrinkage of plants in fragments (∆size = -75 at SPEI \>2), but at lags 5-2 (Sept-Dec) high SPEI has a less negative effect on plant size.
+We only expect plant growth (∆size \> 0) in years when SPEI was low at lags 34-32
+
+The size of plants significantly predicted their size in the following year in continuous forest (edf = 7.8, P \< 0.001 <!--# What stats to include?
+reference df and F also?
+Is this all better as a table?
+-->) and forest fragments (edf = 7.4, P \< 0.001).
+In continuous forest, plants under a size of 259 were predicted to grow from year t-1 to year t, and the fit line was below 1:1 for larger plants, indicating that larger plants tended to shrink, on average.
+In 1 ha fragments, the fit line followed a 1:1 line more closely, indicating that plants tended to stay the same size on average from year to year over the course of this study.
+<!--# something weird about backtransformation causing these strange results?
+--> Drought history also had a significant effect on plant size in both continuous forest (edf = 14.7, p \< 0.001) and fragments (edf = 17.8, p \< 0.001).
+
+## Flowering
+
+Probability of flowering increased with size and was generally higher in continuous forest compared to 1-ha fragments.
+The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest.
+The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.
+
+### Continuous forest
+
+size: p \< 0.001
+
+plot: p\<0.001
+
+SPEI: p \< 0.001
+
+Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov). Drought at lags 13-7 reduced flowering probability
 

### 2021-02-04 analysis/paper/paper.Rmd
set to sentence wrap for easier diffing in the future.

@@
 
-1.
-Organisms simultanously experience multiple forms of natural and anthropogenic disturbance.
+<!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->
+
+1.  Organisms simultanously experience multiple forms of natural and anthropogenic disturbance.
     While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
@@
 
-2.
-Habitat fragmentation is a major source of environmental stress.
+2.  Habitat fragmentation is a major source of environmental stress.
     **[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]** <!--# fragmentation results in increaed mortality due to increased stress.
@@
 In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
-Fragmentation can impact plant population dynamics through ...
-<!--# more about consequences of fragmentation pull from slides from Emilio-->
+    Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->
 
-3.
-In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change.
+3.  In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change.
     For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig.
--- <!--# fig ref -->).
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig. -- <!--# fig ref -->).
 These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried.
@@
 Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
-Because *H.
-acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+Because *H. acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
 We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
@@
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
-
 The general form of the vital rate models was as follows:
@@
 
-Crossbasis is very flat with essentially no effect of SPEI beyond 6 months lag. Recent SPEI has bimodal effect with both low and high values reducing survival.
+Crossbasis is very flat with essentially no effect of SPEI beyond 6 months lag.
+Recent SPEI has bimodal effect with both low and high values reducing survival.
 
@@
 
-Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov). Drought at lags 13-7 reduced flowering probability
+Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov).
+Drought at lags 13-7 reduced flowering probability
 
@@
 National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
-This is publication No.
----- in the BDFFP Technical Series.
+This is publication No. ---- in the BDFFP Technical Series.
 
@@
 NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM."
+PPS, which develop and compute the TMPA as a contribution to TRMM.
+"
 

### 2021-02-08 analysis/paper/vital rate models.Rmd
change 1-d smooths to use "cr" basis

@@
   bam(surv ~ 
-        s(log_size_prev) +
+        s(log_size_prev, bs = "cr") +
         # s(year_fac, bs = "re") +
@@
   bam(log_size ~ 
-        s(log_size_prev) + 
+        s(log_size_prev, bs = "cr") + 
         # s(year_fac, bs = "re") + 
@@
   bam(log_size ~ 
-        s(log_size_prev) +
+        s(log_size_prev, bs = "cr") +
         # s(year_fac, bs = "re") +
@@
   bam(flwr ~ 
-        s(log_size_prev) +
+        s(log_size_prev, bs = "cr") +
         # s(year_fac, bs = "re") + 
@@
   bam(flwr ~ 
-        s(log_size_prev) +
+        s(log_size_prev, bs = "cr") +
         # s(year_fac, bs = "re") +

### 2021-02-08 analysis/paper/vital rate models.Rmd
change to use `gratia` package to draw smooths instead of using cb_margeff() to draw marginal effects plots.

@@
 ```{r}
-draw(s_post_cf)
-```
-
-```{r}
 s_post_1ha <-
@@
 
-```{r}
-draw(s_post_1ha)
-```
-
 ## Covariate plots
 
-Effect of size with all else held average.
-
-```{r}
-sr <- range(demog_post_cf$log_size_prev)
-
-newdata_covar_cf <- 
-  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
-         plot = "newplot",
-         year_fac = "newyear", 
-         spei_history = mean(demog_post_cf$spei_history),
-         L = demog_post_cf$L[1:100, ])
-```
-
-```{r}
-out <- augment(s_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
-
-covar_s_cf <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_cf) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(survival)") +
-  theme_linedraw()
-
-covar_s_cf
-ggsave(here("analysis", "figures", "surv_cf_size.png"), covar_s_cf)
-```
-```{r}
-demog_post_cf %>% filter(size_prev > 1500, surv == 0)
-
-demog %>% filter(ha_id_number == 5310) %>% View()
-```
-
-```{r}
-sr <- range(demog_post_1ha$log_size_prev)
-newdata_covar_1ha <- 
-  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
-         plot = "newplot",
-         year_fac = "newyear",
-         spei_history = mean(demog_post_1ha$spei_history),
-         L = demog_post_cf$L[1:100, ])
-
-out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
-
-covar_s_1ha <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = surv), data = demog_post_1ha) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(survival)") +
-  theme_linedraw()
-
-covar_s_1ha
-ggsave(here("analysis", "figures", "surv_1ha_size.png"), covar_s_1ha)
-```
-
-## Get plot data
-
-```{r}
-# use mean size plant across both habitats so plots are comparable
-ref_data <-
-  demog_post %>%
-  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
-                   year_fac = "newyear",
-                   plot = "newlevel")
-
-mean_size <- exp(ref_data$log_size_prev)
-```
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r}
-pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r}
-surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
-              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
-# plogis(surv_lim)
-```
-
-## Plots
-
-```{r}
-p_surv_cf <-
-  pred_s_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_surv_cf
-ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
-```
-
-```{r}
-p_surv_1ha <-
-  pred_s_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_surv_1ha
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
-```
-
-**Slice through time at severe drought**
-
-```{r}
-p_surv_cf_slice <-
-  pred_s_post_cf %>% 
-  filter(nearest(x, -1.50)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "Continuous Forest", y = "P(survival)") +
-  theme_linedraw()
-p_surv_cf_slice
-```
-
-```{r}
-p_surv_1ha_slice <-
-  pred_s_post_1ha %>% 
-  filter(nearest(x, -1.5)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "1-ha Fragments", y = "P(survival)") +
-  theme_linedraw()
-p_surv_1ha_slice
+Plot smooth for size
+
+```{r}
+s_1ha_eval1 <-
+  smooth_estimates(s_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(s_post_1ha)[1])))
+
+ggplot(s_1ha_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  geom_rug(data = model.frame(s_post_1ha)) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(survival)", title = "1-ha fragments")
+ggsave(here("analysis", "figures", "surv_1ha_size.png"))
+```
+
+```{r}
+s_cf_eval1 <-
+  smooth_estimates(s_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(s_post_cf)[1])))
+
+ggplot(s_cf_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  geom_rug(data = model.frame(s_post_cf)) + 
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(survival)", title = "Continuous forest")
+ggsave(here("analysis", "figures", "surv_cf_size.png"))
+```
+
+## Crossbasis plot
+
+`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
+
+```{r}
+s_1ha_eval <-
+  evaluate_smooth(s_post_1ha, smooth = "spei_history", dist = 0.1) %>% 
+  mutate(est = plogis(est + coef(s_post_1ha)[1]))
+s_cf_eval <-
+  evaluate_smooth(s_post_cf, smooth = "spei_history", dist = 0.1) %>% 
+  mutate(est = plogis(est + coef(s_post_cf)[1]))
+
+surv_lims <- c(min(s_1ha_eval$est, s_cf_eval$est, na.rm = TRUE),
+                  max(s_1ha_eval$est, s_cf_eval$est, na.rm = TRUE))
+```
+
+```{r}
+ggplot(s_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "1ha fragments") +
+  theme_bw()
+ggsave(here("analysis", "figures", "surv_1ha_spei.png"))
+```
+```{r}
+ggplot(s_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) + 
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "Continuous forest") +
+  theme_bw()
+ggsave(here("analysis", "figures", "surv_cf_spei.png"))
 ```
@@
 ```{r}
-draw(g_post_cf)
-```
-
-```{r}
 g_post_1ha <-
@@
 
-```{r}
-# draw(g_post_1ha)
-draw(g_post_1ha, select = 3, continuous_fill = scale_fill_viridis_c(option = "inferno"), contour = FALSE, fun = "exp")
-ggsave("function_smooth-cb.png")
-```
-
 ## Covariate plots
 
-Effect of size with all else held average.
-
-```{r}
-out <- augment(g_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
-
-covar_growth_cf <-
-  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
-  geom_point(data = demog_post_cf, aes(x = size_prev, y = size), alpha = 0.1) +
-  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_abline(slope = 1, linetype = 2, color = "red") +
-  geom_line(color = "blue") +
-  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
-  theme_linedraw() +
-  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
-covar_growth_cf
-ggsave(here("analysis", "figures", "growth_cf_size.png"), covar_growth_cf)
-```
-```{r}
-out %>%
-  filter(exp(log_size_prev) > exp(.fitted)) %>% # shinking plants
-  pull(log_size_prev) %>% min() %>% exp() #first size at which shrinking happens.
-```
-
-```{r}
-out <- augment(g_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
-
-covar_growth_1ha <-
-  ggplot(out, aes(x = exp(log_size_prev), y = exp(.fitted))) + 
-  geom_point(data = demog_post_1ha, aes(x = size_prev, y = size), alpha = 0.1) +
-  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_abline(slope = 1, linetype = 2, color = "red") +
-  geom_line(color = "blue") +
-  # geom_rug(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.25) +
-  theme_linedraw() +
-  labs(x = TeX("size_t_"), y = TeX("size_{t+1}_"))
-
-covar_growth_1ha
-ggsave(here("analysis", "figures", "growth_1ha_size.png"), covar_growth_1ha)
-```
-```{r}
-out %>%
-  filter(exp(log_size_prev) > exp(.fitted)) %>% #shrinking plants
-  pull(log_size_prev) %>% min() %>% exp() #smallest size where plants start shrinking
-```
-
-## Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r}
-pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Calculate growth as change from year t to year t+1
-
-```{r}
-mean_size # size in year t for fitted values
-
-pred_g_post_cf <- 
-  pred_g_post_cf %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-
-pred_g_post_1ha <-
-  pred_g_post_1ha %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r}
-shoot_lim <- c(min(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted),
-               max(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted))
-```
-
-## Plots
-
-```{r}
-p_growth_cf <-
-  pred_g_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_growth_cf
-ggsave(here("analysis", "figures", "growth_cf_spei.png"), p_growth_cf)
-```
-
-```{r}
-p_growth_1ha <-
-  pred_g_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_growth_1ha
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"), p_growth_1ha)
-```
-
-**Slice through time at severe drought**
-
-```{r}
-# p_growth_cf_slice <-
-#   pred_g_post_cf %>% 
-#   filter(nearest(x, -1.50)) %>% 
-#   ggplot(aes(x = lag, y = exp(fitted))) +
-#   geom_ribbon(aes(
-#     ymin = exp(fitted - se.fit * 1.96),
-#     ymax = exp(fitted + se.fit * 1.96)
-#   ), alpha = 0.3)+
-#   geom_line() +
-#   scale_x_continuous("lag(months)", breaks = 0:36) +
-#   labs(title = "Continuous Forest", y = "shoots (t+1)") +
-#   theme_linedraw()
-# p_growth_cf_slice
-```
-
-```{r}
-# p_growth_1ha_slice <-
-#   pred_g_post_1ha %>% 
-#   filter(nearest(x, -1.5)) %>% 
-#   ggplot(aes(x = lag, y = exp(fitted))) +
-#   geom_ribbon(aes(
-#     ymin = exp(fitted - se.fit * 1.96),
-#     ymax = exp(fitted + se.fit * 1.96)
-#   ), alpha = 0.3)+
-#   geom_line() +
-#   scale_x_continuous("lag(months)", breaks = 0:36) +
-#   labs(title = "1-ha Fragments", y = "shoots (t+1)") +
-#   theme_linedraw()
-# p_growth_1ha_slice
-```
-
-**Slice through individual months**
-TODO: Change to ridge plots?
-
-I find this less useful, but just shows that relationships are slightly quadratic at most complicated.
-This would be more useful if there was a way to show which lines had "slopes" significantly different from 0 (i.e.
-don't totally overlap a flat line).
-
-```{r}
-# #TODO: replace with ridge plot
-# plotdf <-
-#   pred_g_post_cf %>% 
-#   mutate(lag = round(lag)) %>% 
-#   group_by(x, lag) %>% 
-#   summarize(across(everything(),  mean))
-# 
-# foo <- function(x) {
-#   x <- as.numeric(x)
-#   mon <- month(ymd("2020-2-01") - months(x), label = TRUE)
-#   l <- glue("lag={x} ({mon})")
-#   return(l)
-# }
-# 
-# p_growth_cf_months <-
-#   plotdf %>% 
-#   ggplot(aes(x = x, y = exp(fitted))) +
-#   geom_ribbon(aes(
-#     ymin = exp(fitted - se.fit * 1.96),
-#     ymax = exp(fitted + se.fit * 1.96)
-#   ), alpha = 0.3)+
-#   geom_line() +
-#   scale_x_continuous("SPEI3") +
-#   scale_y_continuous("shoots (t+1)") +
-#   facet_wrap(~lag, labeller = as_labeller(foo)) +
-#   theme_bw()
-# p_growth_cf_months
+```{r}
+g_1ha_eval1 <-
+  smooth_estimates(g_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~.x + coef(s_post_1ha)[1]))
+
+ggplot(g_1ha_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  # geom_abline(slope = 1, linetype = 2, color = "red") +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  geom_rug(data = model.frame(g_post_1ha)) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})"), title = "1-ha fragments")
+ggsave(here("analysis", "figures", "growth_1ha_size.png"))
+```
+```{r}
+g_cf_eval1 <-
+  smooth_estimates(g_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~.x + coef(s_post_cf)[1]))
+
+ggplot(g_cf_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  # geom_abline(slope = 1, linetype = 2, color = "red") +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  geom_rug(data = model.frame(g_post_cf)) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})"), title = "Continuous forest")
+ggsave(here("analysis", "figures", "growth_cf_size.png"))
+```
+
+## Crossbasis plot
+`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
+
+```{r}
+g_1ha_eval <-
+  evaluate_smooth(g_post_1ha, smooth = "spei_history", dist = 0.1) #%>% 
+  # mutate(est = est + coef(g_post_1ha)[1])
+g_cf_eval <-
+  evaluate_smooth(g_post_cf, smooth = "spei_history", dist = 0.1) # %>% 
+  # mutate(est = est + coef(g_post_cf)[1])
+
+grow_lims <- c(min(g_1ha_eval$est, g_cf_eval$est, na.rm = TRUE),
+               max(g_1ha_eval$est, g_cf_eval$est, na.rm = TRUE))
+```
+
+```{r}
+ggplot(g_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.1) +
+  scale_fill_viridis_c("∆Size (?)", option = "magma", limits = grow_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "1ha fragments") +
+  theme_bw()
+ggsave(here("analysis", "figures", "growth_1ha_spei.png"))
+```
+```{r}
+ggplot(g_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.1) +
+  scale_fill_viridis_c("∆Size (?)", option = "magma", limits = grow_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "Continuous forest") +
+  theme_bw()
+ggsave(here("analysis", "figures", "growth_cf_spei.png"))
 ```
@@
 
-```{r}
-draw(f_post_cf)
-```
-
 This model changes **drastically** depending on whether year is included as a random effect.
@@
 
-```{r}
-draw(f_post_1ha)
-```
-
 ## Covariate plots
 
-Effect of size with all else held average.
-
-```{r}
-out <- augment(f_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)
-
-covar_f_cf <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_cf) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(flowering)") +
-  theme_linedraw()
-
-covar_f_cf
-ggsave(here("analysis", "figures", "flwr_cf_size.png"), covar_f_cf)
-```
-
-```{r}
-out <- augment(f_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)
-
-covar_f_1ha <-
-  ggplot(out, aes(x = exp(log_size_prev), y = plogis(.fitted))) + 
-  stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = demog_post_1ha) +
-  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
-  geom_line(color = "blue") +
-  labs(x = TeX("size_t_"), y = "P(flowering)") +
-  theme_linedraw()
-
-covar_f_1ha
-ggsave(here("analysis", "figures", "flwr_1ha_size.png"), covar_f_1ha)
-```
-
-## Get plot data
+Plot smooth for size
+
+```{r}
+f_1ha_eval1 <-
+  smooth_estimates(f_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(f_post_1ha)[1])))
+
+ggplot(f_1ha_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  # stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = model.frame(f_post_1ha)) +
+  geom_rug(data = model.frame(f_post_1ha)) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(flowering)", title = "1-ha fragments")
+ggsave(here("analysis", "figures", "flwr_1ha_size.png"))
+```
+
+```{r}
+f_cf_eval1 <-
+  smooth_estimates(f_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
+  add_confint() %>% 
+  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(f_post_cf)[1])))
+
+ggplot(f_cf_eval1, aes(x = log_size_prev)) +
+  geom_line(aes(y = est)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
+  # stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = model.frame(f_post_cf)) +
+  geom_rug(data = model.frame(f_post_cf)) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(flowering)", title = "Continuous forest")
+ggsave(here("analysis", "figures", "flwr_cf_size.png"))
+```
+
+## Crossbasis plot
+
+`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
+
+```{r}
+f_1ha_eval <-
+  evaluate_smooth(f_post_1ha, smooth = "spei_history", dist = 0.1) %>% 
+  mutate(est = plogis(est + coef(f_post_1ha)[1]))
+f_cf_eval <-
+  evaluate_smooth(f_post_cf, smooth = "spei_history", dist = 0.1) %>% 
+  mutate(est = plogis(est + coef(f_post_cf)[1]))
+
+flwr_lims <- c(min(f_1ha_eval$est, f_cf_eval$est, na.rm = TRUE),
+               max(f_1ha_eval$est, f_cf_eval$est, na.rm = TRUE))
+```
+
+```{r}
+ggplot(f_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.001) +
+  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "1ha fragments") +
+  theme_bw()
+ggsave(here("analysis", "figures", "flwr_1ha_spei.png"))
+```
+```{r}
+ggplot(f_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.001) +
+  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) + 
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "Continuous forest") +
+  theme_bw()
+ggsave(here("analysis", "figures", "flwr_cf_spei.png"))
+```
+
+# Plots
+## Growth
+
+### Get plot data
 
@@
 
-```{r}
+```{r eval=FALSE}
+pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+```
+
+Calculate growth as change from year t to year t+1
+
+```{r eval=FALSE}
+mean_size # size in year t for fitted values
+
+pred_g_post_cf <- 
+  pred_g_post_cf %>% 
+  mutate(e_fitted = exp(fitted),
+         d_fitted = e_fitted - mean_size)
+
+pred_g_post_1ha <-
+  pred_g_post_1ha %>% 
+  mutate(e_fitted = exp(fitted),
+         d_fitted = e_fitted - mean_size)
+```
+
+Shared color bar limits so scales can be combined.
+
+```{r eval=FALSE}
+shoot_lim <- c(min(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted),
+               max(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted))
+```
+
+### Plots
+
+```{r eval=FALSE}
+p_growth_cf <-
+  pred_g_post_cf %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_growth_cf
+ggsave(here("analysis", "figures", "growth_cf_spei.png"), p_growth_cf)
+```
+
+```{r eval=FALSE}
+p_growth_1ha <-
+  pred_g_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "Month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
+  labs(title = "1-ha Fragments") +
+  theme_bw()
+p_growth_1ha
+ggsave(here("analysis", "figures", "growth_1ha_spei.png"), p_growth_1ha)
+```
+
+## Survival
+
+### Get plot data
+
+```{r eval=FALSE}
+# use mean size plant across both habitats so plots are comparable
+ref_data <-
+  demog_post %>%
+  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
+                   year_fac = "newyear",
+                   plot = "newlevel")
+
+mean_size <- exp(ref_data$log_size_prev)
+```
+
+This is pretty slow. It could probably be faster if I understood matrix math better.
+
+```{r eval=FALSE}
+pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
+```
+
+Shared color bar limits so scales can be combined.
+
+```{r eval=FALSE}
+surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
+              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
+# plogis(surv_lim)
+```
+
+### Plots
+
+```{r eval=FALSE}
+p_surv_cf <-
+  pred_s_post_cf %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) + 
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
+  labs(title = "Continuous Forest") +
+  theme_bw()
+p_surv_cf
+ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
+```
+
+```{r eval=FALSE}
+p_surv_1ha <-
+  pred_s_post_1ha %>%
+  filter(min_dist < 0.1) %>%
+  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
+  geom_raster(interpolate = TRUE) +
+  scale_y_continuous(
+    "lag (months)",
+    breaks = 0:36, 
+    expand = c(0,0),
+    sec.axis = dup_axis(
+      name = "month",
+      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
+  ) +
+  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
+  labs(title = "1-ha Fragments") +
+  theme_bw()
+p_surv_1ha
+ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
+```
+
+## Flowering
+### Get plot data
+
+This is pretty slow. It could probably be faster if I understood matrix math better.
+
+```{r eval=FALSE}
 pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
@@
 
-```{r}
+```{r eval=FALSE}
 flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
@@
 
-## Plots
-
-```{r}
+### Plots
+
+```{r eval=FALSE}
 p_flowered_cf <-
@@
 
-```{r}
+```{r eval=FALSE}
 p_flowered_1ha <-
@@
 very similar to growth
-
-**Slice through time at severe drought**
-
-```{r}
-p_flowered_cf_slice <-
-  pred_f_post_cf %>% 
-  filter(nearest(x, -1.50)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "Continuous Forest", y = "P(flowering)") +
-  theme_linedraw() 
-p_flowered_cf_slice
-```
-
-```{r}
-p_flowered_1ha_slice <-
-  pred_f_post_1ha %>% 
-  filter(nearest(x, -1.5)) %>% 
-  ggplot(aes(x = lag, y = plogis(fitted))) +
-  geom_ribbon(aes(
-    ymin = plogis(fitted - se.fit * 1.96),
-    ymax = plogis(fitted + se.fit * 1.96)
-  ), alpha = 0.3)+
-  geom_line() +
-  scale_x_continuous("lag(months)", breaks = 0:36) +
-  labs(title = "1-ha Fragments", y = "P(flowering)") +
-  theme_linedraw()
-p_flowered_1ha_slice
-```
-
-# Plots
-## Growth
-```{r}
-p_growth_spei <-
-  (p_growth_cf | p_growth_1ha) +
-  plot_annotation(title = "Effect of drought on growth") +
-  plot_layout(guides = "collect") & theme_talk & labs(caption = "")
-p_growth_spei
-# ggsave(here("analysis", "figures", "growth_spei.png"), p_growth_spei, height = 7, width = 14)
-```
-
-## Survival
-
-These aren't significant.
-
-```{r}
-p_surv_spei <- 
-  (p_surv_cf|p_surv_1ha) + 
-  plot_annotation(title = "Effect of drought on survival") +
-  plot_layout(guides = "collect") & theme_talk & labs(caption="")
-p_surv_spei
-# ggsave(here("analysis", "figures", "surv_spei.png"), p_surv_spei, width = 14, height = 7)
-```
-
-## Flowering
-
-```{r}
-p_flowered_spei <- 
-  (p_flowered_cf|p_flowered_1ha) + 
-  plot_annotation(title = "Effect of drought on flowering") +
-  plot_layout(guides = "collect") & theme_talk & labs(caption = "")
-p_flowered_spei
-# ggsave(here("analysis", "figures", "flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
-```

### 2021-02-09 analysis/paper/paper.Rmd
worked on survival results section

@@
 [-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
-However, in a temperate system...
-<!--# oak paper --> [@forner2020].
+However, in a temperate system... <!--# oak paper --> [@forner2020].
 
@@
 
-### Continuous Forest
+Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
+Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \_\_\_ <!--# might be good to plot both curves on same graph -->).
+Survival of the smallest plants was higher in 1-ha fragments, but the confidence intervals still overlapped the fitted values for continuous forest.
 
-size: p \< 0.001 (figure = surv_cf_size.png)
+Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
+For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \_\_\_).
+In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities.
+However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year, also effected survival probabilities.
+Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
+Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
+However, only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past, and although there were plants that did not survive this initial census, it may not be representative of a true survival process <!--# Emilio: let's talk about this.
+Maybe I should remove this year from the analysis for survival?
+-->
 
-plot: p \< 0.001
-
-SPEI: p \< 0.001 (figure = surv_cf_spei.png)
-
-Lags 6-0 (Aug--Feb) and 20-15 (Jun--Nov) show reduced survival with drought.
-Lags 14-7 (Dec - Jul) and 26-21 (Dec-May) show increased survival with drought.
-Lag 36-33 (Feb - May) shows quadratic relationship with increased survival in dry and wet conditions (this seems like it's gotta be spurious).
-
-### 1-ha fragment
-
-size: p \<0.001 (figure = surv_1ha_size.png)
-
-plot: p = 0.258
-
-SPEI: p \< 0.001 (figure = surv_1ha_spei.png)
-
-Crossbasis is very flat with essentially no effect of SPEI beyond 6 months lag.
-Recent SPEI has bimodal effect with both low and high values reducing survival.
-
-**Question**: Currently the figures are expected values for the average plant across both plots.
-This makes it look like survival is a lot higher in 1-ha fragments when in reality survival is exactly the same (0.95) because plants are smaller in 1-ha fragments.
-Should these plots instead be for an average sized plant within the habitat type?
+The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.258 <!--# double-check -->)
 
@@
 In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
+Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001) and the relationship was close to linear for both continuous forest (edf = ) and 1-ha fragments (edf = ).
 

### 2021-02-10 analysis/paper/paper.Rmd
finished draft of figures and added to written results.

@@
     bookdown::word_document2:
+      number_sections: no
       fig_caption: yes
@@
 )
+library(knitr)
+library(here)
 ```
@@
 [-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
-However, in a temperate system... <!--# oak paper --> [@forner2020].
+However, in a temperate system...
+<!--# oak paper --> [@forner2020].
 
@@
 $$
-g[\textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
+\begin{gather}
+    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
     \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j 
+\end{gather}
 $$
@@
 
+```{r}
+cap <- "Plots of means ± SE."
+```
+
+```{r eda, fig.cap=cap}
+knitr::include_graphics(here("analysis", "figures", "eda.png"))
+
+```
+
 ## Survival
 
+```{r}
+cap <- "Smooth effect of plant size in the previous census on survival modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
+```
+
+```{r surv-size, fig.cap=cap}
+include_graphics(here("analysis", "figures", "surv_both_size.png"))
+```
+
 Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
-Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004.
-Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, will get exact numbers from data).
+Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)).
+Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, <!--# will get exact numbers from data -->).
 The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
@@
 Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
-Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \_\_\_ <!--# might be good to plot both curves on same graph -->).
-Survival of the smallest plants was higher in 1-ha fragments, but the confidence intervals still overlapped the fitted values for continuous forest.
+Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \@ref(fig:surv-size)).
+Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals still overlapped the fitted values for continuous forest <!--# I wonder if this changes if I don't include both sources of uncertainty in draw.gam?
+-->.
+
+```{r}
+cap <- "Smooth effect of lagged SPEI on survival.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Contour lines correspond to a change of 0.002.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+
+```
+
+```{r surv-spei, fig.cap=cap}
+knitr::include_graphics(here("analysis", "figures", "surv_both_spei.png"))
+```
 
 Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
-For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \_\_\_).
-In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities.
+For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \@ref(fig:surv-spei)a).
+In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
 However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year, also effected survival probabilities.
@@
 In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
-Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001) and the relationship was close to linear for both continuous forest (edf = ) and 1-ha fragments (edf = ).
+The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).
 
-The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the dataset with mean sizes becoming more similar over time.
+```{r}
+cap <- "Smooth effect of plant size in year t on plant size in year t+1 modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+The dashed red line represents a 1:1 line for reference.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+Model intercepts were added to the fitted values of the smooths."
+```
+
+```{r growth-size, fig.cap=cap}
+include_graphics(here("analysis", "figures", "growth_both_size.png"))
+```
+
+Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001).
+The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:growth-size)).
+
+```{r}
+cap <- "Smooth effect of lagged SPEI on plant growth. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Contour lines correspond to a change of 0.1. "
+```
+
+<!--# this is the raw fitted values of the tensor product smooth.
+That is, I have not shifted this by adding the model intercepts like I did in the last figure.
+Not sure if the effect size should actually be interpreted as a additive change in size, but I think so maybe?
+If that's true, then the predicted effect of SPEI on size in year t+1 is extremely small compared to the effect of size in year t.
+-->
+
+```{r growth-spei, fig.cap=cap}
+include_graphics(here("analysis", "figures", "growth_both_spei.png"))
+```
+
+Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
+The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+Drought at lags 8--10 increase growth in both habitats.
+<!--# say a lot more about this here. -->
+
+The random effect of plot is significant in both habitat models (p \< 0.001).
 
@@
 
-### Continuous Forest
-
-size: p \< 0.001
-
-plot: p \< 0.001
-
-SPEI: p \< 0.001
-
-Predicted results show only lagged effects of SPEI on growth (f(x) at lags 0 and 1 is flat).
-Wet conditions in preceding winter months improve growth (lags 4-2, Oct - Dec).
-Positive effect of drought on growth at lags 11-8 (Mar - Jun) and 34-24 (nearly the entirety of year t-2).
-
-### 1-ha fragment
-
-size: p \< 0.001
-
-plot: p \< 0.001
-
-SPEI: p \< 0.001
-
-Crossbasis looks very similar to continuous forest but more extreme.
-SPEI has a stronger effect on growth in fragments.
-Wet conditions at lag 0 predict the greatest shrinkage of plants in fragments (∆size = -75 at SPEI \>2), but at lags 5-2 (Sept-Dec) high SPEI has a less negative effect on plant size.
-We only expect plant growth (∆size \> 0) in years when SPEI was low at lags 34-32
-
-The size of plants significantly predicted their size in the following year in continuous forest (edf = 7.8, P \< 0.001 <!--# What stats to include?
-reference df and F also?
-Is this all better as a table?
--->) and forest fragments (edf = 7.4, P \< 0.001).
-In continuous forest, plants under a size of 259 were predicted to grow from year t-1 to year t, and the fit line was below 1:1 for larger plants, indicating that larger plants tended to shrink, on average.
-In 1 ha fragments, the fit line followed a 1:1 line more closely, indicating that plants tended to stay the same size on average from year to year over the course of this study.
-<!--# something weird about backtransformation causing these strange results?
---> Drought history also had a significant effect on plant size in both continuous forest (edf = 14.7, p \< 0.001) and fragments (edf = 17.8, p \< 0.001).
-
 ## Flowering
 
-Probability of flowering increased with size and was generally higher in continuous forest compared to 1-ha fragments.
-The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest.
+Probability of flowering was generally higher in continuous forest compared to 1-ha fragments.
+The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest (Figure \@ref(fig:eda)c).
 The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.
 
-### Continuous forest
+```{r}
+cap <- "Smooth effect of plant size in the previous census on flowering probability modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
+```
 
-size: p \< 0.001
+```{r flwr-size, fig.cap=cap}
+include_graphics(here("analysis", "figures", "flwr_both_size.png"))
+```
 
-plot: p\<0.001
+Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
+Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
+Flowering probability was greater in 1-ha fragments for large plants compared to continuous forest.
+However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
+<!--# idea: plot vertical lines representing mean or median size in each habitat -->
 
-SPEI: p \< 0.001
+```{r}
+cap <- "Smooth effect of lagged SPEI on flowering probability.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Contour lines correspond to a change of 0.001.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+```
 
-Flowering probability was increased by recent drought (at lag = 0 with SPEI \< 1) and drought at lags 20-15 (Jun - Nov).
-Drought at lags 13-7 reduced flowering probability
+```{r flwr-spei, fig.cap=cap}
+include_graphics(here("analysis", "figures", "flwr_both_spei.png"))
+```
 
-## Size by year interaction
+Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
+Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
+The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
+In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun - Nov).
+Drought at lags 7--13 reduced flowering probability.
+The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
+That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
 
-<!--# this all needs updating.  Probably going to use AIC instead of these p-values since I think they're doing the same thing. -->
-
-In continuous forest plots there was support for size by year interactions for growth (95% LR~null~ = 4.54, LR~obs~ = 45.69) and survival (95% LR~null~ = 3.91, LR~obs~ = 29.18) , but not for flowering (95% LR~null~ = 4.44, LR~obs~ = 1.77).
-However, in 1 ha fragments, we saw no evidence of size by year interaction.
-For growth 95% LR~null~ = 4.45, LR~obs~ = 4.35, for survival 95% LR~null~ = 4.47, LR~obs~ = 0.62, and for flowering probability 95% LR~null~ = 3.86, LR~obs~ = 2.82.
-This is consistent with the DLNMs in that the crossbasis functions for \_\_\_ and \_\_\_ were much flatter in fragments compared to continuous forest.
+The random effect of plot also had a significant effect in both habitats (p \< 0.001).
 

### 2021-02-10 analysis/paper/vital rate models.Rmd
worked on survival results section

@@
 ```
+```{r}
+demog_post %>% filter(spei_history[,36] > 2) %>% pull(year) %>% unique()
+demog_post %>% filter(year == min(year)) %>% filter(surv == 0)
+```
 
@@
   plot_layout(guides = "collect") &
+  plot_annotation(tag_levels = "a", tag_suffix = ")") &
   theme(plot.margin = margin(2,1,1,1)) &
@@
 
+Combine and plot together
+
+```{r}
+bind_rows("1-ha" = s_1ha_eval1, "CF" = s_cf_eval1, .id = "habitat") %>% 
+  ggplot(aes(x = log_size_prev, color = habitat)) +
+  geom_line(aes(y = est)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, color = NULL, fill = habitat), alpha = 0.25) +
+  geom_rug(data = bind_rows(
+    "1-ha" = model.frame(s_post_1ha),
+    "CF" = model.frame(s_post_cf), 
+    .id = "habitat"),
+    color = "black") +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(survival)")
+ggsave(here("analysis", "figures", "surv_both_size.png"))
+```
+
 ## Crossbasis plot
@@
 
-```{r}
+TODO: It would be *very cool* if the colors could be the same, but the limits on the scale bar could be different.
+
+```{r}
+s_1ha_cbplot <-
   ggplot(s_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   theme_bw()
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"))
-```
-```{r}
+ggsave(here("analysis", "figures", "surv_1ha_spei.png"), s_1ha_cbplot)
+```
+```{r}
+s_cf_cbplot <- 
   ggplot(s_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   theme_bw()
-ggsave(here("analysis", "figures", "surv_cf_spei.png"))
+ggsave(here("analysis", "figures", "surv_cf_spei.png"), s_cf_cbplot)
+```
+
+combine
+```{r}
+s_1ha_cbplot + s_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
+ggsave(here("analysis", "figures", "surv_both_spei.png"))
 ```
@@
 
+Both together
+
+```{r}
+bind_rows("1-ha" = g_1ha_eval1, "CF" = g_cf_eval1, .id = "habitat") %>% 
+  ggplot(aes(x = log_size_prev)) +
+  geom_line(aes(y = est, color = habitat)) +
+  geom_abline(slope = 1, linetype = 2, color = "red") +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = habitat), alpha = 0.25) +
+  geom_rug(data = bind_rows(
+    "1-ha" = model.frame(g_post_1ha),
+    "CF" = model.frame(g_post_cf),
+    .id = "habitat")) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})"))
+ggsave(here("analysis", "figures", "growth_both_size.png"))
+```
+
 ## Crossbasis plot
@@
 ```{r}
+g_1ha_cbplot <- 
   ggplot(g_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   geom_contour(color = "black", binwidth = 0.1) +
-  scale_fill_viridis_c("∆Size (?)", option = "magma", limits = grow_lims) +
+  scale_fill_viridis_c("∆log(size) (?)", option = "magma", limits = grow_lims) +
   scale_y_continuous(
@@
   theme_bw()
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"))
-```
-```{r}
+ggsave(here("analysis", "figures", "growth_1ha_spei.png"), g_1ha_cbplot)
+```
+
+```{r}
+g_cf_cbplot <-
   ggplot(g_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   theme_bw()
-ggsave(here("analysis", "figures", "growth_cf_spei.png"))
+ggsave(here("analysis", "figures", "growth_cf_spei.png"), g_cf_cbplot)
+```
+
+combine
+
+```{r}
+g_1ha_cbplot + g_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
+ggsave(here("analysis", "figures", "growth_both_spei.png"))
 ```
@@
 
+combine and plot
+
+```{r}
+bind_rows("1-ha" = f_1ha_eval1, "CF" = f_cf_eval1, .id = "habitat") %>% 
+  ggplot(aes(x = log_size_prev)) +
+  geom_line(aes(y = est, color = habitat)) +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = habitat), alpha = 0.25) +
+  # stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = model.frame(f_post_cf)) +
+  geom_rug(data = bind_rows(
+    "1-ha" = model.frame(f_post_1ha),
+    "CF" = model.frame(f_post_cf),
+    .id = "habitat")) +
+  theme_bw() +
+  labs(x = TeX("$log(size_t)$"), y = "P(flowering)")
+ggsave(here("analysis", "figures", "flwr_both_size.png"))
+```
+
 ## Crossbasis plot
@@
 ```{r}
+f_1ha_cbplot <- 
   ggplot(f_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "flwr_1ha_spei.png"))
-```
-```{r}
+  labs(title = "1-ha fragments") +
+  theme_bw()
+ggsave(here("analysis", "figures", "flwr_1ha_spei.png"), f_1ha_cbplot)
+```
+```{r}
+f_cf_cbplot <- 
   ggplot(f_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
@@
   theme_bw()
-ggsave(here("analysis", "figures", "flwr_cf_spei.png"))
-```
-
-# Plots
+ggsave(here("analysis", "figures", "flwr_cf_spei.png"), f_cf_cbplot)
+```
+
+combine
+```{r}
+f_1ha_cbplot + f_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
+ggsave(here("analysis", "figures", "flwr_both_spei.png"))
+```
+
+# Predict Plots
 ## Growth

### 2021-02-11 analysis/paper/paper.Rmd
clarified seasonality of lag influence

@@
 In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
-However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year, also effected survival probabilities.
-Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
+However, SPEI at a lag time of 15--20 months, corresponding to June--October (the dry season) of the previous year <!--# i.e.
+not the most recent dry season, but the one before that.
+How can I clarify this?
+-->, also effected survival probabilities.
+Lower SPEI (drought) 15--20 months prior to census <!--# i.e. same dry season --> was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
 Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
@@
 
-The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.258 <!--# double-check -->)
+The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.257).
 
@@
 The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-Drought at lags 8--10 increase growth in both habitats.
-<!--# say a lot more about this here. -->
+Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats.
+
+<!--# There's a lot more going on in the 1-ha fragments.  Hurts my brain to look at it.  Will write another day. -->
 
 The random effect of plot is significant in both habitat models (p \< 0.001).
-
-**NOTE**: I considered adding a covariate of year to account for yearly trends in growth, but since there is also a trend in SPEI over the time of this experiment, they were concurve (wiggly collinearity), and shouldn't both be included in model.
 
@@
 The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
-In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun - Nov).
+In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, the dry season in the previous year).
 Drought at lags 7--13 reduced flowering probability.

### 2021-02-12 analysis/paper/EDA_figure_1.Rmd
moved figure 1 and data wrangling out to separate files.

--- 2021-02-08 analysis/paper/vital rate models.Rmd
@@
 ---
-title: "FLM for growth as function of SPEI"
+title: "Exploratory Data Visualizations (Fig. 1)"
 author: "Eric R. Scott"
-date: "Date Initiated"
+date: "2021-02-12"
 output: 
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 library(tidyverse)
 library(here)
-library(mgcv)
-library(dlnm)
-library(bbmle)
 library(lubridate)
-library(glue)
+library(tsibble)
 library(patchwork)
-library(sf)
-library(tsibble)
-library(broom)
-library(gratia)
-library(latex2exp)
+library(HeliconiaDemography) #load functions in current package.
+library(conflicted)
 
-library(conflicted)
 conflict_prefer("lag", "dplyr")
 conflict_prefer("filter", "dplyr")
-conflict_prefer("slice", "dplyr")
-
-library(HeliconiaDemography) #load functions in this compendium
-# devtools::load_all()
-```
-
-```{r}
-#define plotting themes
-theme_talk <- theme_bw() + theme(text = element_text(size = 18))
 ```
@@
 
-# TODO:
-
-- ~~Switch measure of plant size to shoots * height~~
-- ~~Change lag to 36 months (3 years)~~
-- Produce ridge plot
-- Heatmap for growth should be ∆size
-- try select = TRUE (after you understand it better? blog about it first)
-
 # Purpose
 
-Take a crack at modeling vital rates as a function of previous height and lagged, non-linear, SPEI.
+What is the goal of this notebook?
 
@@
 
-```{r data, echo=TRUE}
+```{r}
+demog_full <- read_rds(here("analysis", "data", "derived_data", "ha_spei_combined.rds"))
 xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
-demog <-
-  read_csv(
-    here("analysis", "data", "derived_data", "ha_survey.csv"),
-    col_names = TRUE,
-    cols(plot = col_character(),
-         bdffp_reserve_no = col_character())
-  )
-plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
-```
-
-# Join data sets
-
-## Prep precip data
-
-Get only rows for february (the month of the census), and year for joining
-
-```{r}
-xa2 <- 
-  xa %>%
-  filter(month(yearmonth) == 2) %>% 
-  mutate(year = year(yearmonth), .after =yearmonth) %>% 
-  separate(latlon, into = c("lat", "lon"), sep = "_")
-xa2
-```
-
-Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2
-
-```{r}
-unique(xa2$lat)
-mean(c(-2.25, -2.5))
-```
-
-So, from -2.25 to -2.5
-
-```{r}
-unique(xa2$lon)
-mean(c(-59.5, -59.75))
-mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
-mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
-```
-
-```{r}
-plot_coords
-```
-
-```{r}
-demog2 <-
-  demog %>% 
-  mutate(lon = case_when(
-    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
-    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
-    TRUE ~ NA_character_
-  ), .before = plot)
-demog_full <- 
-  left_join(demog2, xa2, by = c("lon", "year")) %>% 
-  select(-lon, -lat, -yearmonth)
 ```
@@
 
-`gam` wants factors!
-
-```{r}
-demog_full <-
-  demog_full %>% 
-  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
-  mutate(year_fac = as.factor(year))
-# View(demog_full)
-```
-
-# Splitting up data
-
-Because I can't model interactions with DLNMs, I can either include factors like habitat as predictors (different intercepts, same functional form, same data) or fit separate models (potentially different functional forms, but not using all the data).
+Only do post-seedlings
 
@@
   filter(code_notes != "sdlg (1)" | is.na(code_notes))
-
-demog_post_cf <-
-  demog_post %>% 
-  filter(habitat == "CF")
-
-demog_post_1ha <-
-  demog_post %>% 
-  filter(habitat == "1-ha")
-
-demog_sdlg <- 
-  demog_full %>% 
-  filter(code_notes == "sdlg (1)")
-
-demog_sdlg_cf <- 
-  demog_sdlg %>% 
-  filter(habitat == "CF")
-
-demog_sdlg_1ha <- 
-  demog_sdlg %>% 
-  filter(habitat == "1-ha")
 ```
@@
   filter(habitat !="10-ha") %>% 
-  select(ranch, bdffp_reserve_no, plot, habitat, year, ht, shts, size, surv, flwr) %>% 
+  select(ranch, bdffp_reserve_no, plot, habitat, year, ht, shts, shts_prev, size, surv, flwr) %>% 
   mutate(date = paste(year, "-02-01") %>% ymd(),
@@
 ```
-
 ## SPEI
@@
 ```
-
 ## Survival
@@
   demog_plotdf %>%  
+  filter(shts_prev >= 4) %>%
   ggplot(aes(x = yearmonth, y = surv, color = habitat)) +
@@
   demog_plotdf %>% 
+  filter(shts >= 4) %>% 
   ggplot(aes(x = yearmonth, y = flwr, color = habitat)) + 
@@
 ```
-
-# Speed up models with bam()
-
-I'm going to start fitting these with `bam()` to speed things up.  This sets up the paralellization.  CAUTION: you may need to change these settings for your machine.
-
-```{r}
-library(parallel)
-ncores <- detectCores()
-cl <- makeForkCluster(ncores - 2)
-```
-
-# Survival
-`surv` is 1 in `year`s when a plant is alive and 0 in the first `year` when the plant is dead.
-These GAMs model the survival of a plant (over the past year) as a function of size in the previous census and SPEI values starting at february of the current year going back 36 months.
-## Fit GAMs
-
-```{r}
-s_post_cf <-
-  bam(surv ~ 
-        s(log_size_prev, bs = "cr") +
-        # s(year) +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog_post_cf,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-```{r}
-# gam.check(s_post_cf) #not sure what to expect from any of these plots
-appraise(s_post_cf)
-# concurvity(s_post_cf, full = FALSE)
-```
-
-```{r}
-summary(s_post_cf)
-```
-
-```{r}
-s_post_1ha <-
-  bam(surv ~ 
-        s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog_post_1ha,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-```{r}
-summary(s_post_1ha)
-```
-
-## Covariate plots
-
-Plot smooth for size
-
-```{r}
-s_1ha_eval1 <-
-  smooth_estimates(s_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(s_post_1ha)[1])))
-
-ggplot(s_1ha_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  geom_rug(data = model.frame(s_post_1ha)) +
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = "P(survival)", title = "1-ha fragments")
-ggsave(here("analysis", "figures", "surv_1ha_size.png"))
-```
-
-```{r}
-s_cf_eval1 <-
-  smooth_estimates(s_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(s_post_cf)[1])))
-
-ggplot(s_cf_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  geom_rug(data = model.frame(s_post_cf)) + 
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = "P(survival)", title = "Continuous forest")
-ggsave(here("analysis", "figures", "surv_cf_size.png"))
-```
-
-## Crossbasis plot
-
-`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
-
-```{r}
-s_1ha_eval <-
-  evaluate_smooth(s_post_1ha, smooth = "spei_history", dist = 0.1) %>% 
-  mutate(est = plogis(est + coef(s_post_1ha)[1]))
-s_cf_eval <-
-  evaluate_smooth(s_post_cf, smooth = "spei_history", dist = 0.1) %>% 
-  mutate(est = plogis(est + coef(s_post_cf)[1]))
-
-surv_lims <- c(min(s_1ha_eval$est, s_cf_eval$est, na.rm = TRUE),
-                  max(s_1ha_eval$est, s_cf_eval$est, na.rm = TRUE))
-```
-
-```{r}
-ggplot(s_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.002) +
-  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"))
-```
-```{r}
-ggplot(s_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.002) +
-  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "surv_cf_spei.png"))
-```
-
-# Growth
-For mature plants that are alive...
-
-```{r}
-demog2_post_cf <- demog_post_cf %>% filter(surv == 1, !is.na(log_size))
-demog2_post_1ha <- demog_post_1ha %>% filter(surv == 1, !is.na(log_size))
-```
-
-## Fit GAMs
-
-Fit the GAM with crossbasis smooth for lagged drought.
-Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.
-Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).
-
-The transformed response and identity link is actually better than a log-link, according to Ellner, Childs and Rees.
-
-```{r}
-g_post_cf <-
-  bam(log_size ~ 
-        s(log_size_prev, bs = "cr") + 
-        # s(year_fac, bs = "re") + 
-        s(plot, bs = "re") + #random effect of plot
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
-          xt = list(bs = "cr")),
-      # family = gaussian(link = "identity"),
-      family = scat(link = "identity"), #like leptokurtic gaussian.
-      data = demog2_post_cf,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-Inspect the GAM
-
-```{r}
-AIC(g_post_cf)
-appraise(g_post_cf)
-```
-
-AIC for... 
-Gaussian: ~40909.27. looks leptokurtic
-scaled t: ~37284.47.  Waaay better
-
-```{r}
-summary(g_post_cf)
-```
-
-Remember, R-sq is probably so high because size this year is a very good predictor of size next year.
-
-```{r}
-g_post_1ha <-
-  bam(log_size ~ 
-        s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
-          xt = list(bs = "cr")),
-      family = scat(link = "identity"), #like leptokurtic gaussian.
-      data = demog2_post_1ha,
-      select = TRUE,
-      method = "REML",
-      cluster = cl)
-```
-
-```{r}
-appraise(g_post_1ha)
-```
-
-```{r}
-summary(g_post_1ha)
-```
-
-## Covariate plots
-
-```{r}
-g_1ha_eval1 <-
-  smooth_estimates(g_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~.x + coef(s_post_1ha)[1]))
-
-ggplot(g_1ha_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  # geom_abline(slope = 1, linetype = 2, color = "red") +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  geom_rug(data = model.frame(g_post_1ha)) +
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})"), title = "1-ha fragments")
-ggsave(here("analysis", "figures", "growth_1ha_size.png"))
-```
-```{r}
-g_cf_eval1 <-
-  smooth_estimates(g_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~.x + coef(s_post_cf)[1]))
-
-ggplot(g_cf_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  # geom_abline(slope = 1, linetype = 2, color = "red") +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  geom_rug(data = model.frame(g_post_cf)) +
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})"), title = "Continuous forest")
-ggsave(here("analysis", "figures", "growth_cf_size.png"))
-```
-
-## Crossbasis plot
-`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
-
-```{r}
-g_1ha_eval <-
-  evaluate_smooth(g_post_1ha, smooth = "spei_history", dist = 0.1) #%>% 
-  # mutate(est = est + coef(g_post_1ha)[1])
-g_cf_eval <-
-  evaluate_smooth(g_post_cf, smooth = "spei_history", dist = 0.1) # %>% 
-  # mutate(est = est + coef(g_post_cf)[1])
-
-grow_lims <- c(min(g_1ha_eval$est, g_cf_eval$est, na.rm = TRUE),
-               max(g_1ha_eval$est, g_cf_eval$est, na.rm = TRUE))
-```
-
-```{r}
-ggplot(g_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.1) +
-  scale_fill_viridis_c("∆Size (?)", option = "magma", limits = grow_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"))
-```
-```{r}
-ggplot(g_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.1) +
-  scale_fill_viridis_c("∆Size (?)", option = "magma", limits = grow_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "growth_cf_spei.png"))
-```
-
-# Flowering Probability
-
-## Fit GAMs
-
-```{r}
-x <- demog_post$flwr
-sum(x, na.rm = TRUE) / length(x)
-```
-
-Around 5 percent of plants flowered.
-
-```{r}
-f_post_cf <-
-  bam(flwr ~ 
-        s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") + 
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog2_post_cf,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-Doesn't converge with random effect of year
-
-```{r}
-appraise(f_post_cf)
-# concurvity(f_post_cf, full = FALSE)
-```
-
-these residuals are troubling
-
-```{r}
-summary(f_post_cf)
-```
-
-This model changes **drastically** depending on whether year is included as a random effect.
-Does that mean ther random effect is important?
-Or is it because it is collinear / concurve with SPEI history?
-
-```{r}
-f_post_1ha <-
-  bam(flwr ~ 
-        s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
-      family = binomial,
-      data = demog2_post_1ha,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-```{r}
-appraise(f_post_1ha)
-```
-
-```{r}
-summary(f_post_1ha)
-```
-
-## Covariate plots
-
-Plot smooth for size
-
-```{r}
-f_1ha_eval1 <-
-  smooth_estimates(f_post_1ha, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(f_post_1ha)[1])))
-
-ggplot(f_1ha_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  # stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = model.frame(f_post_1ha)) +
-  geom_rug(data = model.frame(f_post_1ha)) +
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = "P(flowering)", title = "1-ha fragments")
-ggsave(here("analysis", "figures", "flwr_1ha_size.png"))
-```
-
-```{r}
-f_cf_eval1 <-
-  smooth_estimates(f_post_cf, smooth = "log_size_prev", unconditional = TRUE) %>% 
-  add_confint() %>% 
-  mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + coef(f_post_cf)[1])))
-
-ggplot(f_cf_eval1, aes(x = log_size_prev)) +
-  geom_line(aes(y = est)) +
-  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.25) +
-  # stat_summary_bin(geom = "point", fun = mean, aes(y = flwr), data = model.frame(f_post_cf)) +
-  geom_rug(data = model.frame(f_post_cf)) +
-  theme_bw() +
-  labs(x = TeX("$log(size_t)$"), y = "P(flowering)", title = "Continuous forest")
-ggsave(here("analysis", "figures", "flwr_cf_size.png"))
-```
-
-## Crossbasis plot
-
-`dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
-
-```{r}
-f_1ha_eval <-
-  evaluate_smooth(f_post_1ha, smooth = "spei_history", dist = 0.1) %>% 
-  mutate(est = plogis(est + coef(f_post_1ha)[1]))
-f_cf_eval <-
-  evaluate_smooth(f_post_cf, smooth = "spei_history", dist = 0.1) %>% 
-  mutate(est = plogis(est + coef(f_post_cf)[1]))
-
-flwr_lims <- c(min(f_1ha_eval$est, f_cf_eval$est, na.rm = TRUE),
-               max(f_1ha_eval$est, f_cf_eval$est, na.rm = TRUE))
-```
-
-```{r}
-ggplot(f_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.001) +
-  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "flwr_1ha_spei.png"))
-```
-```{r}
-ggplot(f_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.001) +
-  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "flwr_cf_spei.png"))
-```
-
-# Plots
-## Growth
-
-### Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Calculate growth as change from year t to year t+1
-
-```{r eval=FALSE}
-mean_size # size in year t for fitted values
-
-pred_g_post_cf <- 
-  pred_g_post_cf %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-
-pred_g_post_1ha <-
-  pred_g_post_1ha %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-shoot_lim <- c(min(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted),
-               max(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted))
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_growth_cf <-
-  pred_g_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_growth_cf
-ggsave(here("analysis", "figures", "growth_cf_spei.png"), p_growth_cf)
-```
-
-```{r eval=FALSE}
-p_growth_1ha <-
-  pred_g_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_growth_1ha
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"), p_growth_1ha)
-```
-
-## Survival
-
-### Get plot data
-
-```{r eval=FALSE}
-# use mean size plant across both habitats so plots are comparable
-ref_data <-
-  demog_post %>%
-  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
-                   year_fac = "newyear",
-                   plot = "newlevel")
-
-mean_size <- exp(ref_data$log_size_prev)
-```
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
-              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
-# plogis(surv_lim)
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_surv_cf <-
-  pred_s_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_surv_cf
-ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
-```
-
-```{r eval=FALSE}
-p_surv_1ha <-
-  pred_s_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_surv_1ha
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
-```
-
-## Flowering
-### Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
-                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
-plogis(flowered_lim)
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_flowered_cf <-
-  pred_f_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_flowered_cf
-ggsave(here("analysis", "figures", "flwr_cf_spei.png"), p_flowered_cf)
-```
-
-```{r eval=FALSE}
-p_flowered_1ha <-
-  pred_f_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_flowered_1ha
-ggsave(here("analysis", "figures", "flwr_1ha_spei.png"), p_flowered_1ha)
-```
-very similar to growth

### 2021-02-12 analysis/paper/vital rate models.Rmd
moved figure 1 and data wrangling out to separate files.

@@
 library(patchwork)
-library(sf)
+# library(sf)
 library(tsibble)
@@
 # devtools::load_all()
-```
-
-```{r}
-#define plotting themes
-theme_talk <- theme_bw() + theme(text = element_text(size = 18))
 ```
@@
 
-```{r data, echo=TRUE}
-xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
-demog <-
-  read_csv(
-    here("analysis", "data", "derived_data", "ha_survey.csv"),
-    col_names = TRUE,
-    cols(plot = col_character(),
-         bdffp_reserve_no = col_character())
-  )
-plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
-```
-
-# Join data sets
-
-## Prep precip data
-
-Get only rows for february (the month of the census), and year for joining
-
 ```{r}
-xa2 <- 
-  xa %>%
-  filter(month(yearmonth) == 2) %>% 
-  mutate(year = year(yearmonth), .after =yearmonth) %>% 
-  separate(latlon, into = c("lat", "lon"), sep = "_")
-xa2
-```
-
-Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2
-
-```{r}
-unique(xa2$lat)
-mean(c(-2.25, -2.5))
-```
-
-So, from -2.25 to -2.5
-
-```{r}
-unique(xa2$lon)
-mean(c(-59.5, -59.75))
-mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
-mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
-```
-
-```{r}
-plot_coords
-```
-
-```{r}
-demog2 <-
-  demog %>% 
-  mutate(lon = case_when(
-    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
-    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
-    TRUE ~ NA_character_
-  ), .before = plot)
-demog_full <- 
-  left_join(demog2, xa2, by = c("lon", "year")) %>% 
-  select(-lon, -lat, -yearmonth)
+demog_full <- read_rds(here("analysis", "data", "derived_data", "ha_spei_combined.rds"))
 ```
@@
   filter(habitat == "1-ha")
-```
-
-# Inspect raw data
-
-```{r}
-demog_plotdf <-  
-  demog_post %>% 
-  filter(habitat !="10-ha") %>% 
-  select(ranch, bdffp_reserve_no, plot, habitat, year, ht, shts, size, surv, flwr) %>% 
-  mutate(date = paste(year, "-02-01") %>% ymd(),
-         yearmonth = yearmonth(date))
-```
-
-## SPEI
-
-```{r}
-dates <- range(demog_post$year) %>% paste0(., "-02-01") %>% ymd()
-census_dates <- min(demog_post$year)[1]:max(demog_post$year) %>% paste(., "-02-01") %>% ymd()
-
-spei <-
-  xa %>%
-  select(latlon, yearmonth, spei) %>%
-  filter(!is.na(spei)) %>% 
-  ggplot(aes(x = yearmonth, y = spei)) + 
-  geom_line(aes(group = latlon), alpha = 0.2) +
-  stat_summary(geom = "line", fun = "mean") +
-  scale_y_continuous("SPEI", expand = expansion(mult = c(0, 0.05), add = 0)) +
-  scale_x_yearmonth("Date", limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-spei <- annotate_spei(spei)
-spei
-```
-```{r}
-demog_post %>% filter(spei_history[,36] > 2) %>% pull(year) %>% unique()
-demog_post %>% filter(year == min(year)) %>% filter(surv == 0)
-```
-
-## Survival
-
-```{r}
-demog_post %>% ggplot(aes(x = surv)) + geom_bar() + facet_wrap(~habitat)
-```
-
-```{r}
-demog_post %>% 
-  group_by(habitat) %>% 
-  summarize(surv = sum(surv, na.rm = TRUE) / n())
-```
-
-Survival nearly identical for 1 ha and CF.
-
-### Timeseries
-
-```{r}
-survival <-
-  demog_plotdf %>%  
-  ggplot(aes(x = yearmonth, y = surv, color = habitat)) +
-  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
-  stat_summary(geom = "pointrange", fatten = 1) +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("P(survived)") +
-  theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
-survival
-# (survival / spei) 
-```
-Looks like survival went down year after 2003 drought and during 2007 drought.
-
-## Size / growth
-
-```{r}
-demog_post %>% ggplot(aes(x = habitat, y = log_size)) + geom_boxplot()
-```
-```{r}
-demog_post %>% 
-  group_by(habitat) %>% 
-  summarize(size = mean(size, na.rm = TRUE),
-            shts = mean(shts, na.rm = TRUE),
-            ht = mean(ht, na.rm = TRUE))
-```
-
-### Timeseries
-
-```{r}
-size <-
-  demog_plotdf %>% 
-  ggplot(aes(x = yearmonth, y = size, color = habitat)) +
-  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
-  stat_summary(geom = "pointrange", fatten = 1) +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("Size") +
-  theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
-size
-# size/survival/spei
-```
-Average plant size shrank in 2003 drought, but not 2007 drought
-
-## Flowering
-
-```{r}
-demog_post %>% ggplot(aes(x = flwr)) + geom_bar() + facet_wrap(~habitat)
-```
-### Timeseries
-
-```{r}
-flowering <-
-  demog_plotdf %>% 
-  ggplot(aes(x = yearmonth, y = flwr, color = habitat)) + 
-  stat_summary(geom = "line", fun = "mean") +
-  stat_summary(geom = "pointrange", fatten = 1) +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("P(flowering)") +
-  theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
-flowering
-```
-```{r}
-demog_plotdf %>% filter(year == 2003) %>% group_by(habitat) %>% summarize(flwr = mean(flwr, na.rm = TRUE))
-```
-
-```{r}
-size/survival/flowering/spei +
-  plot_layout(guides = "collect") &
-  plot_annotation(tag_levels = "a", tag_suffix = ")") &
-  theme(plot.margin = margin(2,1,1,1)) &
-  guides(col = guide_legend(title = "Habitat"))
-ggsave(here("analysis", "figures", "eda.png"), height = 7, width = 10)
 ```
@@
 s_1ha_cbplot + s_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "surv_both_spei.png"))
+ggsave(here("analysis", "figures", "surv_both_spei.png"), width = 10, height = 5)
 ```
@@
 g_1ha_cbplot + g_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "growth_both_spei.png"))
+ggsave(here("analysis", "figures", "growth_both_spei.png"), width = 10, height = 5)
 ```
@@
 f_1ha_cbplot + f_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "flwr_both_spei.png"))
+ggsave(here("analysis", "figures", "flwr_both_spei.png"), width = 10, height = 5)
 ```

### 2021-02-12 analysis/wrangling/3-combine-data.R

@@ -1,0 +1,65 @@
+# Combine demographic data with SPEI data ---------------------------------------
+#' Author: Eric R. Scott
+#' Creation Date: 2020-01-08
+
+
+# load packages -----------------------------------------------------------
+library(tidyverse)
+library(here)
+library(tsibble)
+
+# Load Data ---------------------------------------------------------------
+# SPEI data
+xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
+
+# cleaned demography data
+demog <-
+  read_csv(
+    here("analysis", "data", "derived_data", "ha_survey.csv"),
+    col_names = TRUE,
+    cols(plot = col_character(),
+         bdffp_reserve_no = col_character())
+  )
+
+# plot lat long estimates, for joining data
+plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
+
+
+# Prep SPEI data ----------------------------------------------------------
+
+# Only want rows starting with February, the month the census is conducted in.
+xa2 <- 
+  xa %>%
+  filter(month(yearmonth) == 2) %>% 
+  mutate(year = year(yearmonth), .after =yearmonth) %>% 
+  separate(latlon, into = c("lat", "lon"), sep = "_")
+
+# Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2
+# So, from -2.25 to -2.5
+
+unique(xa2$lon)
+mean(c(-59.5, -59.75))
+mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
+mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
+
+
+# Prep demography data ----------------------------------------------------
+# manually match up site names with lat long values in SPEI dataset.
+demog2 <-
+  demog %>% 
+  mutate(lon = case_when(
+    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
+    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
+    TRUE ~ NA_character_
+  ), .before = plot)
+
+
+# Join data ---------------------------------------------------------------
+demog_full <- 
+  left_join(demog2, xa2, by = c("lon", "year")) %>% 
+  select(-lon, -lat, -yearmonth)
+
+
[…]

### 2021-02-16 analysis/paper/EDA_figure_1.Rmd
added a survivorship curve for the 1998 cohort of plants

@@
 ```{r}
-demog_full <- read_rds(here("analysis", "data", "derived_data", "ha_spei_combined.rds"))
+demog <- read_rds(here("analysis", "data", "derived_data", "ha_spei_combined.rds"))
 xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
@@
 demog_full <-
-  demog_full %>% 
+  demog %>% 
   filter(ht < 200 | is.na(ht), ht_prev < 200 | is.na(ht_prev)) %>%
@@
 
+### Survivorship curve
+
+Emilio suggested an alternative to the timeseries plot---a survivorship curve for the 1998 cohort of plants.
+
+```{r paged.print=FALSE}
+cohort <-demog %>% filter(year == 1998) %>% pull(ha_id_number) %>% unique()
+
+surv_curve_df <-
+  demog %>% 
+  filter(ha_id_number %in% cohort) %>% 
+  filter(surv == 1, habitat != "10-ha") %>% 
+  group_by(year, habitat) %>% 
+  summarize(n = n()) %>%
+  group_by(habitat) %>% 
+  mutate(p_surv = n/first(n),
+         date = ymd(paste0(year, "-02-01")),
+         yearmonth = yearmonth(date))
+
+surv_curve <-
+  ggplot(surv_curve_df, aes(x = yearmonth, y = p_surv, color = habitat)) +
+  geom_step() +
+  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  scale_y_continuous("P(survived)") +
+  theme_bw() +
+  theme(
+    # axis.text.x = element_blank(),
+    # axis.title.x = element_blank(),
+    # axis.ticks.x = element_blank(),
+    panel.grid.minor.x = element_blank()
+  )
+surv_curve
+```
+
 ## Size / growth
@@
 
+# Combine
+
 ```{r}
-size/survival/flowering/spei +
+size/
+  # survival/
+  surv_curve/
+  flowering/
+  spei +
   plot_layout(guides = "collect") &

### 2021-02-16 analysis/paper/EDA_figure_1.Rmd
changed flowering to proportion (removed error bars)

@@
   theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+  theme(
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    panel.grid.minor.x = element_blank(),
+    legend.position = "none"
+  )
 survival
@@
 ```{r paged.print=FALSE}
-cohort <-demog %>% filter(year == 1998) %>% pull(ha_id_number) %>% unique()
+cohort <- demog %>% filter(year == 1998) %>% pull(ha_id_number) %>% unique()
 
@@
   theme(
-    # axis.text.x = element_blank(),
-    # axis.title.x = element_blank(),
-    # axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
     panel.grid.minor.x = element_blank()
@@
   theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+  theme(
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    panel.grid.minor.x = element_blank(),
+    legend.position = "none"
+  )
 size
@@
 ```{r}
-flowering <-
+flowering_df <-
   demog_plotdf %>% 
   filter(shts >= 4) %>% 
-  ggplot(aes(x = yearmonth, y = flwr, color = habitat)) + 
-  stat_summary(geom = "line", fun = "mean") +
-  stat_summary(geom = "pointrange", fatten = 1) +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("P(flowering)") +
+  group_by(yearmonth, habitat) %>% 
+  summarize(n = n(), flwr = sum(flwr == 1, na.rm = TRUE)) %>% 
+  mutate(prop_flwr = flwr/n)
+
+flowering <-
+  ggplot(flowering_df, aes(x = yearmonth, y = prop_flwr, color = habitat)) + 
+  geom_line() +
+    scale_x_yearmonth(
+      limits = dates,
+      breaks = census_dates,
+      date_minor_breaks = "1 month",
+      expand = expansion(mult = 0.02)
+    ) +
+    scale_y_continuous("P(flowering | shts ≥ 4)") +
   theme_bw() +
-  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid.minor.x = element_blank())
+  theme(
+    axis.text.x = element_blank(),
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    panel.grid.minor.x = element_blank()
+  )
 flowering

### 2021-02-17 analysis/paper/vital rate models.Rmd
removed old unused code (I forget I'm using git sometimes)

@@
 ```
-
-# Predict Plots
-## Growth
-
-### Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Calculate growth as change from year t to year t+1
-
-```{r eval=FALSE}
-mean_size # size in year t for fitted values
-
-pred_g_post_cf <- 
-  pred_g_post_cf %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-
-pred_g_post_1ha <-
-  pred_g_post_1ha %>% 
-  mutate(e_fitted = exp(fitted),
-         d_fitted = e_fitted - mean_size)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-shoot_lim <- c(min(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted),
-               max(pred_g_post_1ha$d_fitted, pred_g_post_cf$d_fitted))
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_growth_cf <-
-  pred_g_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_growth_cf
-ggsave(here("analysis", "figures", "growth_cf_spei.png"), p_growth_cf)
-```
-
-```{r eval=FALSE}
-p_growth_1ha <-
-  pred_g_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = d_fitted, fill = d_fitted)) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "Month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c(TeX("$\\Delta$ size"), option = "A", limits = shoot_lim) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_growth_1ha
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"), p_growth_1ha)
-```
-
-## Survival
-
-### Get plot data
-
-```{r eval=FALSE}
-# use mean size plant across both habitats so plots are comparable
-ref_data <-
-  demog_post %>%
-  dplyr::summarize(log_size_prev = log(mean(size_prev, na.rm = TRUE)),  #log of mean size, not mean of log_size
-                   year_fac = "newyear",
-                   plot = "newlevel")
-
-mean_size <- exp(ref_data$log_size_prev)
-```
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-surv_lim <- c(min(pred_s_post_1ha$fitted, pred_s_post_cf$fitted),
-              max(pred_s_post_1ha$fitted, pred_s_post_cf$fitted))
-# plogis(surv_lim)
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_surv_cf <-
-  pred_s_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_surv_cf
-ggsave(here("analysis", "figures", "surv_cf_spei.png"), p_surv_cf)
-```
-
-```{r eval=FALSE}
-p_surv_1ha <-
-  pred_s_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_surv_1ha
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"), p_surv_1ha)
-```
-
-## Flowering
-### Get plot data
-
-This is pretty slow. It could probably be faster if I understood matrix math better.
-
-```{r eval=FALSE}
-pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
-```
-
-Shared color bar limits so scales can be combined.
-
-```{r eval=FALSE}
-flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
-                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
-plogis(flowered_lim)
-```
-
-### Plots
-
-```{r eval=FALSE}
-p_flowered_cf <-
-  pred_f_post_cf %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
-  labs(title = "Continuous Forest") +
-  theme_bw()
-p_flowered_cf
-ggsave(here("analysis", "figures", "flwr_cf_spei.png"), p_flowered_cf)
-```
-
-```{r eval=FALSE}
-p_flowered_1ha <-
-  pred_f_post_1ha %>%
-  filter(min_dist < 0.1) %>%
-  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
-  geom_raster(interpolate = TRUE) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
-  labs(title = "1-ha Fragments") +
-  theme_bw()
-p_flowered_1ha
-ggsave(here("analysis", "figures", "flwr_1ha_spei.png"), p_flowered_1ha)
-```
-very similar to growth

### 2021-02-17 analysis/paper/vital rate models.Rmd
prototype of rotated crossbasis plot

@@
 ggsave(here("analysis", "figures", "surv_both_spei.png"), width = 10, height = 5)
+```
+Try rotating to put time on x-axis
+
+```{r}
+breaks <- seq(0,36, by = 2)
+ggplot(s_1ha_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
+  scale_x_reverse(
+    "lag (months)",
+    breaks = breaks,
+    expand = c(0, 0),
+    labels = function(x) -x,
+    sec.axis = dup_axis(name = "month",
+                        labels = month(ymd("2020-2-01") - months(breaks), label = TRUE))
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  # labs(title = "1ha fragments") +
+  theme_bw() +
+  theme(axis.text.x.top = element_text(angle = 45, vjust = 0.2, hjust = 0)) +
+  NULL
 ```

### 2021-02-17 analysis/paper/vital rate models.Rmd
changed diagnostic plots to quartile residuals for binomial models.

@@
         s(log_size_prev, bs = "cr") +
-        # s(year) +
-        # s(year_fac, bs = "re") +
-        s(plot, bs = "re") +
+        s(plot, bs = "re") + #random intercept
+        # s(plot, log_size_prev, bs = "re") + #random slope
         s(spei_history, L,
@@
 ```{r}
-# gam.check(s_post_cf) #not sure what to expect from any of these plots
-appraise(s_post_cf)
-# concurvity(s_post_cf, full = FALSE)
+# appraise(s_post_cf)
+#quantile residuals are better with discrete response variable.
+library(statmod)
+#histogram
+plot(density(qresid(s_post_cf)))
+#QQ plot
+qqnorm(qresid(s_post_cf)); qqline(qresid(s_post_cf))
+#Fitted vs. residuals
+scatter.smooth(predict(s_post_cf, type = "response"), qresid(s_post_cf), col = "grey")
 ```
@@
         s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") +
         s(plot, bs = "re") +
+        # s(plot, log_size_prev, bs = "re") +
         s(spei_history, L,
@@
       cluster = cl)
+```
+```{r}
+#histogram
+plot(density(qresid(s_post_1ha)))
+#QQ plot
+qqnorm(qresid(s_post_1ha)); qqline(qresid(s_post_1ha))
+#Fitted vs. residuals
+scatter.smooth(predict(s_post_1ha, type = "response"), qresid(s_post_1ha), col = "grey")
 ```
@@
 ```
+
+Emilio suggested plotting curves from all 1-ha plots just to visualize the variation.
+But the random effect of plot is not significant, so there is actually very little variation among plots in the model.
+
+```{r}
+new_data <-
+  tidyr::expand(model.frame(s_post_1ha),
+                nesting(plot),
+                log_size_prev = seq(min(log_size_prev), max(log_size_prev), length.out = 50),
+                spei_history = mean(spei_history), L = L)
+m1pred <- bind_cols(new_data,
+          as.data.frame(predict(s_post_1ha, newdata = new_data, se.fit = TRUE, type = "response")))
+ggplot(m1pred, aes(x = log_size_prev, y = fit, group = plot)) + geom_line()
+```
+
+The wide confidence intervals are probably due to lower sample size at this size of plant.
 
@@
         s(log_size_prev, bs = "cr") + 
-        # s(year_fac, bs = "re") + 
-        s(plot, bs = "re") + #random effect of plot
+        s(plot, bs = "re") + #random effect of plot on intercept
         s(spei_history, L,
@@
 ```{r}
-appraise(f_post_cf)
-# concurvity(f_post_cf, full = FALSE)
+#histogram
+plot(density(qresid(f_post_cf)))
+#QQ plot
+qqnorm(qresid(f_post_cf)); qqline(qresid(f_post_cf))
+#Fitted vs. residuals
+scatter.smooth(predict(f_post_cf, type = "response"), qresid(f_post_cf), col = "grey")
 ```
@@
 ```{r}
-appraise(f_post_1ha)
+#histogram
+plot(density(qresid(f_post_1ha)))
+#QQ plot
+qqnorm(qresid(f_post_1ha)); qqline(qresid(f_post_1ha))
+#Fitted vs. residuals
+scatter.smooth(predict(f_post_1ha, type = "response"), qresid(f_post_1ha), col = "grey")
 ```

### 2021-02-17 analysis/paper/vital rate models.Rmd
plotted difference between two crossbasis smooths.

@@
 breaks <- seq(0,36, by = 2)
+s_1ha_rotated <-
   ggplot(s_1ha_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
@@
   scale_y_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  # labs(title = "1ha fragments") +
+  labs(title = "1ha fragments") +
   theme_bw() +
@@
   NULL
+```
+```{r}
+s_cf_rotated <- 
+  ggplot(s_cf_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
+  scale_x_reverse(
+    "lag (months)",
+    breaks = breaks,
+    expand = c(0, 0),
+    labels = function(x) -x,
+    sec.axis = dup_axis(name = "month",
+                        labels = month(ymd("2020-2-01") - months(breaks), label = TRUE))
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "Continuous Forest") +
+  theme_bw() +
+  theme(axis.text.x.top = element_text(angle = 45, vjust = 0.2, hjust = 0)) +
+  NULL
+```
+
+Try plotting a raster of the difference between CF and 1-ha
+
+```{r}
+s_diff_rotated <- 
+  bind_cols(
+  s_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
+  s_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
+  ) %>%
+  mutate(diff = cf_est - frag_est) %>% 
+  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  scale_fill_viridis_c("∆Survival (CF – 1ha)", option = "plasma") +
+  scale_x_reverse(
+    "lag (months)",
+    breaks = breaks,
+    expand = c(0, 0),
+    labels = function(x) -x,
+    sec.axis = dup_axis(name = "month",
+                        labels = month(ymd("2020-2-01") - months(breaks), label = TRUE))
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  labs(title = "CF – 1-ha") +
+  theme_bw() +
+  theme(axis.text.x.top = element_text(angle = 45, vjust = 0.2, hjust = 0)) +
+  NULL
+```
+```{r}
+(s_cf_rotated |s_1ha_rotated) / (s_diff_rotated | plot_spacer()) + plot_layout(guides="collect")
 ```

### 2021-02-18 analysis/paper/EDA_figure_1.Rmd
added previous 36 months to SPEI figure

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
 library(tidyverse)
@@
     # having no aboveground biomass is possible, but categorically different than just being small. Exclude plants with 0 shoots or 0 height
-    shts_prev > 0,
-    ht_prev > 0,
+    shts_prev > 0 | is.na(shts_prev),
+    ht_prev > 0 | is.na(shts_prev),
     #for survival data, must include plants with NA for shts and height (i.e. dead plants)
@@
 ```
+
 ## SPEI
@@
 ```{r}
-dates <- range(demog_post$year) %>% paste0(., "-02-01") %>% ymd()
-census_dates <- min(demog_post$year)[1]:max(demog_post$year) %>% paste(., "-02-01") %>% ymd()
+dates <- 
+  range(demog_post$year) %>%
+  paste0(., "-02-01") %>% 
+  ymd()
+dates[1] <- dates[1] - months(36)
+
+census_dates <- seq(dates[1], dates[2], by = "year")
 
@@
 ```
+How many infloresences given flowering?
+
+```{r}
+demog %>% 
+  filter(flwr == 1) %>%
+  summarize(mean_infl = mean(infl_num),
+            median_infl = median(infl_num),
+            min = min(infl_num),
+            max = max(infl_num))
+```
+
 ### Timeseries

### 2021-02-18 analysis/paper/paper.Rmd
updated figures and results

@@
   These are the highlights
+editor_options:
+  markdown:
+    wrap: sentence
 ---
@@
 
+```{r}
+cap <- "This is just a placeholder figure, but a map or satelite image of BDFFP will go here."
+```
+
+```{r map, fig.cap=cap}
+include_graphics(here("analysis", "figures", "map.png"))
+```
+
 ## Study site
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig. -- <!--# fig ref -->).
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)).
 These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried.
@@
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** <!--# check data --> flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
-Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
+Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
 *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.
@@
 The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments <!--# are we including 10ha fragments?
+--> were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
@@
 Plants that were not found for three consecutive surveys were considered dead.
-In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings.
-There is a seven-fold range in H.
-acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings <!--# are we using these data?
+-->.
+There is a seven-fold range in *H.
+acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
@@
 Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
-The estimates of potential evapotranspiration were calculated using the Pennman-Monteith method, and then along with monthly precipitation were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
 SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
 Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
-Because *H. acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
-We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+SPEI can be calculated on different scales and we chose to calculate 3-month SPEI because *H.
+acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
 The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).
@@
 A random effect of plot ID on the intercept is represented by $a_j$.
-For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$).
+For plant growth the response was plant size (natural log of height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution.
 A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
@@
 
-In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity <!--# refrence Wood 2006?
--->.
+In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
 We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].
 
-Because the `dlnm` package does not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+Because the `dlnm` package does not currently allow modeling of factor by smooth interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
 
@@
 
-Start with descriptive results of drought events---how many droughts, how severe, timing of droughts, etc.
-
 ```{r}
-cap <- "Plots of means ± SE."
+cap <- "Timeseries of demographic parameters and drought ocurrence.
+Mean plant size ± standard error (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
+(b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
+c) shows the proportion of mature plants (with 4 or more shoots) flowering each census.
+In d), monthly 3-month SPEI is plotted.
+Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.
+Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
 ```
@@
 
+The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
+For example, the 1997 El Niño drought was one of the most severe droughts in the Amazon on record [@williamsonAmazonianTreeMortality2000] and correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
+The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season?-->
+
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with the most severe drought in the timeseries (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in the most severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
+
 ## Survival
@@
 cap <- "Smooth effect of plant size in the previous census on survival modeled by a cubic regression spline.
-95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
 The smooths for 1-ha fragments and continuous forest are fit in separate models.
@@
 Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
-Survival was lowest in 2004 which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)).
-Survival was lower in continuous forest compared to 1-ha fragments in this year (\~0.91 compared to \~0.94, <!--# will get exact numbers from data -->).
-The lowest survival for 1-ha fragment plots (about 0.93) was actually in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
+The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004.
+The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
 
 Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
-Survival probability of large plants approached 1, but this high survival rate was reached at a smaller size for plants in continuous forest compared to fragments (Figure \@ref(fig:surv-size)).
-Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals still overlapped the fitted values for continuous forest <!--# I wonder if this changes if I don't include both sources of uncertainty in draw.gam?
--->.
+Survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:surv-size)).
+Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
 
@@
 In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
-However, SPEI at a lag time of 15--20 months, corresponding to June--October (the dry season) of the previous year <!--# i.e.
-not the most recent dry season, but the one before that.
-How can I clarify this?
--->, also effected survival probabilities.
-Lower SPEI (drought) 15--20 months prior to census <!--# i.e. same dry season --> was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
+However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e. not the most recent dry season, but the one before that), also effected survival probabilities.
+Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
 Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
-However, only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past, and although there were plants that did not survive this initial census, it may not be representative of a true survival process <!--# Emilio: let's talk about this.
-Maybe I should remove this year from the analysis for survival?
--->
+However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
 
@@
 The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats.
+Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
 
@@
 The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.
+<!--# update to be only for large plants -->
 

### 2021-02-19 analysis/paper/vital rate models.Rmd
rotated crossbasis plots, added line for SPEI = 0, added season bar, made difference rasters.

--- 2021-02-17 analysis/paper/vital rate models.Rmd
@@
 
+### Setup for all crossbasis plots
+Ok, I want to add a lil strip on the bottom that shows dry and wet seasons.  Wet = November-may, dry = june-october.
+
+```{r}
+wet_xmaxs = c(3, 15, 27, 36)
+wet_xmins = c(0, 8, 20, 32)
+dry_xmaxs = c(8, 20, 32)
+dry_xmins = c(3, 15, 27)
+df <- tibble(x = 0:36)
+season_bar <- 
+  ggplot(df, aes(x = x, y = 1)) +
+  annotate(geom = "rect", xmin = wet_xmins, xmax = wet_xmaxs, ymin = 0.975, ymax = 1.025, fill = "black") +
+  annotate(geom = "rect", xmin = dry_xmins, xmax = dry_xmaxs, ymin = 0.975, ymax = 1.025, fill = "white") +
+  scale_y_continuous(expand = c(0,0)) +
+  scale_x_continuous(expand = c(0,0)) +
+  theme_void()
+```
+
+```{r}
+breaks <- seq(0,36, by = 2)
+```
+
+### Get plot data
 `dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
@@
 
-TODO: It would be *very cool* if the colors could be the same, but the limits on the scale bar could be different.
+### 1ha
 
 ```{r}
-s_1ha_cbplot <-
-  ggplot(s_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.002) +
-  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "surv_1ha_spei.png"), s_1ha_cbplot)
-```
-```{r}
-s_cf_cbplot <- 
-  ggplot(s_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
-  geom_raster() +
-  geom_contour(color = "black", binwidth = 0.002) +
-  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "surv_cf_spei.png"), s_cf_cbplot)
-```
-
-combine
-```{r}
-s_1ha_cbplot + s_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "surv_both_spei.png"), width = 10, height = 5)
-```
-Try rotating to put time on x-axis
-
-```{r}
-breaks <- seq(0,36, by = 2)
+s_1ha_rotated <-
   ggplot(s_1ha_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
@@
   geom_contour(color = "black", binwidth = 0.002) +
-  scale_fill_viridis_c("P(survival)", option = "magma", limits = surv_lims) +
-  scale_x_reverse(
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("P(survival)", option = "viridis", limits = surv_lims) +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  ) +
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0)))+ #leave room for season bar at bottom
+  labs(title = "1ha fragments") +
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
+```
+
+### CF
+
+```{r}
+s_cf_rotated <-
+  ggplot(s_cf_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("P(survival)", option = "viridis", limits = surv_lims) +
+  scale_x_continuous(
     "lag (months)",
     breaks = breaks,
-    expand = c(0, 0),
-    labels = function(x) -x,
-    sec.axis = dup_axis(name = "month",
-                        labels = month(ymd("2020-2-01") - months(breaks), label = TRUE))
+    expand = c(0, 0)
   )+
-  scale_y_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  # labs(title = "1ha fragments") +
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
+  labs(title = "Continuous Forest") +
   theme_bw() +
-  theme(axis.text.x.top = element_text(angle = 45, vjust = 0.2, hjust = 0)) +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
   NULL
+
+```
+
+### Combine
+
+```{r}
+(s_cf_rotated + theme(axis.title.x = element_blank())) / 
+  (s_1ha_rotated) + 
+  plot_layout(guides = "collect") + 
+  plot_annotation(tag_levels = "a", tag_suffix = ")") & 
+  labs(title = NULL) #remove plot titles
+ggsave(here("analysis", "figures", "surv_both_spei.png"), width = 7, height = 5)
+```
+
+### Difference
+Try plotting a raster of the difference between CF and 1-ha
+
+```{r}
+s_diff_rotated <- 
+  bind_cols(
+  s_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
+  s_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
+  ) %>%
+  mutate(diff = cf_est - frag_est) %>% 
+  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.002) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("∆Survival (CF – 1ha)", option = "viridis") +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
+  labs(title = "CF – 1-ha") +
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
+ggsave(here("analysis", "figures", "surv_diff_spei.png"),plot = s_diff_rotated, width = 7, height = 2.5)
 ```
@@
 ## Crossbasis plot
+
+### Get plot data
 `dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
@@
 g_1ha_eval <-
-  evaluate_smooth(g_post_1ha, smooth = "spei_history", dist = 0.1) #%>% 
-  # mutate(est = est + coef(g_post_1ha)[1])
+  evaluate_smooth(g_post_1ha, smooth = "spei_history", dist = 0.1) %>% 
+  mutate(est = est + coef(g_post_1ha)[1])
 g_cf_eval <-
-  evaluate_smooth(g_post_cf, smooth = "spei_history", dist = 0.1) # %>% 
-  # mutate(est = est + coef(g_post_cf)[1])
+  evaluate_smooth(g_post_cf, smooth = "spei_history", dist = 0.1)  %>% 
+  mutate(est = est + coef(g_post_cf)[1])
 
@@
 
+### 1ha
+
 ```{r}
-g_1ha_cbplot <- 
-  ggplot(g_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+g_1ha_rotated <-
+  ggplot(g_1ha_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
   geom_raster() +
-  geom_contour(color = "black", binwidth = 0.1) +
-  scale_fill_viridis_c("∆log(size) (?)", option = "magma", limits = grow_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  geom_contour(color = "black", binwidth = 0.05) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c(TeX("$log(size_{t+1})$"), option = "viridis", limits = grow_lims) +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0)))+ #leave room for season bar at bottom
   labs(title = "1ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "growth_1ha_spei.png"), g_1ha_cbplot)
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
 ```
 
+### CF
+
 ```{r}
-g_cf_cbplot <-
-  ggplot(g_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+g_cf_rotated <-
+  ggplot(g_cf_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
   geom_raster() +
-  geom_contour(color = "black", binwidth = 0.1) +
-  scale_fill_viridis_c("∆log(size) (?)", option = "magma", limits = grow_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  geom_contour(color = "black", binwidth = 0.05) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c(TeX("$log(size_{t+1})$"), option = "viridis", limits = grow_lims) +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0)))+ #leave room for season bar at bottom
   labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "growth_cf_spei.png"), g_cf_cbplot)
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
 ```
 
+### Combine
 combine
+```{r}
+(g_cf_rotated + theme(axis.title.x = element_blank())) / 
+  (g_1ha_rotated) + 
+  plot_layout(guides = "collect") + 
+  plot_annotation(tag_levels = "a", tag_suffix = ")") & 
+  labs(title = NULL) #remove plot titles
+ggsave(here("analysis", "figures", "growth_both_spei.png"), width = 7, height = 5)
+```
+### Difference
 
 ```{r}
-g_1ha_cbplot + g_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "growth_both_spei.png"), width = 10, height = 5)
+g_diff_rotated <- 
+  bind_cols(
+  g_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
+  g_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
+  ) %>%
+  mutate(diff = cf_est - frag_est) %>% 
+  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.05) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c(TeX("$\\Delta log(size_{t+1})$ (CF – 1ha)"), option = "viridis") +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
+  labs(title = "CF – 1-ha") +
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
+ggsave(here("analysis", "figures", "growth_diff_spei.png"),plot = g_diff_rotated, width = 7, height = 2.5)
 ```
@@
 ```
+```{r paged.print=FALSE}
+demog_full %>% filter(ht < 1)
+```
 
 ## Crossbasis plot
-
+### Get plot data
 `dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
@@
 
+### 1ha
+
 ```{r}
-f_1ha_cbplot <- 
-  ggplot(f_1ha_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+f_1ha_rotated <-
+  ggplot(f_1ha_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
   geom_raster() +
   geom_contour(color = "black", binwidth = 0.001) +
-  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) +
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
-  labs(title = "1-ha fragments") +
-  theme_bw()
-ggsave(here("analysis", "figures", "flwr_1ha_spei.png"), f_1ha_cbplot)
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("P(flowering)", option = "viridis", limits = flwr_lims) +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0)))+ #leave room for season bar at bottom
+  labs(title = "1ha fragments") +
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
 ```
+
+### CF
 ```{r}
-f_cf_cbplot <- 
-  ggplot(f_cf_eval, aes(x = spei_history, y = L, fill = est, z = est)) +
+f_cf_rotated <-
+  ggplot(f_cf_eval, aes(y = spei_history, x = L, fill = est, z = est)) +
   geom_raster() +
   geom_contour(color = "black", binwidth = 0.001) +
-  scale_fill_viridis_c("P(flowering)", option = "magma", limits = flwr_lims) +
-  scale_y_continuous(
-    "lag (months)",
-    breaks = 0:36, 
-    expand = c(0,0),
-    sec.axis = dup_axis(
-      name = "month",
-      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
-  ) + 
-  scale_x_continuous(TeX("SPEI_3_"), expand = c(0,0)) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("P(flowering)", option = "viridis", limits = flwr_lims) +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0)))+ #leave room for season bar at bottom
   labs(title = "Continuous forest") +
-  theme_bw()
-ggsave(here("analysis", "figures", "flwr_cf_spei.png"), f_cf_cbplot)
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
 ```
+### Combine
+combine
+`
+```{r}
+  (f_cf_rotated + theme(axis.title.x = element_blank())) / 
+  (f_1ha_rotated) + 
+  plot_layout(guides = "collect") + 
+  plot_annotation(tag_levels = "a", tag_suffix = ")") & 
+  labs(title = NULL) #remove plot titles
+ggsave(here("analysis", "figures", "flwr_both_spei.png"), width = 7, height = 5)
+```
+### Difference
 
-combine
 ```{r}
-f_1ha_cbplot + f_cf_cbplot + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a", tag_suffix = ")")
-ggsave(here("analysis", "figures", "flwr_both_spei.png"), width = 10, height = 5)
+f_diff_rotated <- 
+  bind_cols(
+  f_cf_eval %>% rename_with(.fn = ~glue("cf_{.}")),
+  f_1ha_eval %>% rename_with(.fn = ~glue("frag_{.}"))
+  ) %>%
+  mutate(diff = cf_est - frag_est) %>% 
+  ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
+  geom_raster() +
+  geom_contour(color = "black", binwidth = 0.001) +
+  geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
+  scale_fill_viridis_c("∆Flowering (CF – 1ha)", option = "viridis") +
+  scale_x_continuous(
+    "lag (months before census)",
+    breaks = breaks,
+    expand = c(0, 0)
+  )+
+  scale_y_continuous(TeX("SPEI_3_"), expand = expansion(mult = c(0.025, 0))) +
+  labs(title = "CF – 1-ha") +
+  theme_bw() +
+  annotation_custom(grob = ggplotGrob(season_bar), ymin = -Inf, ymax = min(s_cf_eval$spei_history), xmin = -Inf, xmax = Inf) +
+  NULL
+ggsave(here("analysis", "figures", "flwr_diff_spei.png"),plot = s_diff_rotated, width = 7, height = 2.5)
 ```

### 2021-02-19 analysis/paper/paper.Rmd
edits.  I hope sentence wrap "sticks" this time.

@@
 Plants that were not found for three consecutive surveys were considered dead.
-In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings <!--# are we using these data?
--->.
-There is a seven-fold range in *H.
-acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings <!--# are we using these data? -->.
+There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
@@
 ```{r}
-cap <- "Smooth effect of lagged SPEI on survival.
+cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
 Contour lines correspond to a change of 0.002.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
 
@@
 ```{r}
-cap <- "Smooth effect of lagged SPEI on plant growth. Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis. Contour lines correspond to a change of 0.1. "
+cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function.
+Contour lines correspond to a change of 0.05.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
 ```
-
-<!--# this is the raw fitted values of the tensor product smooth.
-That is, I have not shifted this by adding the model intercepts like I did in the last figure.
-Not sure if the effect size should actually be interpreted as a additive change in size, but I think so maybe?
-If that's true, then the predicted effect of SPEI on size in year t+1 is extremely small compared to the effect of size in year t.
--->
 
@@
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
 Contour lines correspond to a change of 0.001.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
 ```

### 2021-02-19 analysis/supplementary-materials/appendix_density_dependence.Rmd
Created appendix for SPEI data
sentence wrap.

@@
 ---
-title: "Density dependence"
+title: "Appendix 2: Density dependence"
 author: "Eric R. Scott"
@@
 csl: "../templates/ecology.csl" # Insert path for the bib-style
+editor_options:
+  markdown:
+    wrap: sentence
 ---
@@
 
+-   Use a tensor product with a monotonically decreasing P-spline for the distance dimension (see Wood book 5.3.6, something like `ssp <- s(x, bs = "ps", k = k); ssp$mono <- -1`
+
 # Methods
@@
 In 2009, plants were mapped to ....
-<!--# rubim study -->For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al.
-[-@teller2016].
-Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e.
-a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m.
+<!--# rubim study -->For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al. [-@teller2016].
+Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e. a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m.
 If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value.
@@
 
-We used the three transition years for which we had density data to test for effects of conspecific density on growth and survival of *H.
-acuminata*.
-We were unable to test for the effects of density dependence on probability of flowering, as surveys of reproduction were not conducted in these years <!--# or perhaps no plants flowered?
--->.
+We used the three transition years for which we had density data to test for effects of conspecific density on growth and survival of *H. acuminata*.
+We were unable to test for the effects of density dependence on probability of flowering, as surveys of reproduction were not conducted in these years <!--# or perhaps no plants flowered? -->.
 
@@
 We modeled density dependence using a penalized distributed lag non-linear model (DLNM) approach with the `dlnm` package in R [@gasparriniPenalizedFrameworkDistributed2017; @gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-This approach is similar to that of Teller et al.
-[-@teller2016], in that it makes no assumptions about the shape of the competition kernel.
+This approach is similar to that of Teller et al. [-@teller2016], in that it makes no assumptions about the shape of the competition kernel.
 Instead, it fits a two-dimensional spline allowing for a non-linear response to the intensity of conspecific density as well as the distance of conspecifics.
@@
 
-marginally significant size x conspecific density interaction (p = 0.046). No significant main effect of density.
+marginally significant size x conspecific density interaction (p = 0.046).
+No significant main effect of density.
 

### 2021-02-22 analysis/paper/EDA_figure_1.Rmd
updated figures and results
updates to results

@@
 
+When is survival lowest?
+
+```{r}
+demog_post %>% 
+  group_by(year, habitat) %>% 
+  summarize(surv = sum(surv, na.rm = TRUE) / n()) %>% 
+  pivot_wider(names_from = habitat, values_from = surv)
+```
+
 ### Timeseries
@@
   group_by(habitat) %>% 
-  summarize(size = mean(size, na.rm = TRUE),
-            shts = mean(shts, na.rm = TRUE),
-            ht = mean(ht, na.rm = TRUE))
+  summarize(across(
+    c(size, shts, ht),
+    list(mean = mean, sd = sd),
+    na.rm = TRUE,
+    .names = "{.col}_{.fn}"
+  ))
+
 ```
@@
   demog_plotdf %>% 
-  ggplot(aes(x = yearmonth, y = size, color = habitat)) +
-  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
-  stat_summary(geom = "pointrange", fatten = 1) +
+  ggplot(aes(x = yearmonth, y = log(size), color = habitat)) +
+    stat_summary(
+      geom = "line",
+      fun = "mean",
+      fun.args = list(na.rm = TRUE),
+      position = position_dodge(width = 100),
+      aes(group = habitat)
+    )+
+    stat_summary(
+      geom = "pointrange",
+      fatten = 1,
+      fun.data = "mean_sdl",
+      fun.args = list(mult = 1),
+      position = position_dodge(width = 100)
+    ) +
   scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("Size") +
+  scale_y_continuous("log(size)") +
   theme_bw() +
@@
     legend.position = "none"
-  )
-size
+  ) +
+  NULL
 # size/survival/spei
@@
 Average plant size shrank in 2003 drought, but not 2007 drought
+
+### Ridge plot
+```{r}
+library(ggridges)
+ggplot(demog_plotdf) +
+  geom_density_ridges(aes(x = log(size), y = as.factor(year), fill = habitat, color = habitat), alpha = 0.5)
+ggplot(demog_plotdf) +
+  geom_violin(aes(y = log(size), x = as.factor(year), fill = habitat), alpha = 0.5, draw_quantiles = c(0.5))
+
+```
 

### 2021-02-22 analysis/paper/paper.Rmd
updates to results

@@
 knitr::opts_chunk$set(
+	echo = FALSE,
+	fig.path = "../figures/",
+	message = FALSE,
+	warning = FALSE,
 	collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
 	comment = "#>",
-  fig.path = "../figures/",
 	dpi = 300
@@
 In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)).
-These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried.
+These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried.
 In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
@@
 
-Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+Precipitation and potential evapotranspiration data were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
 We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
@@
 Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
-SPEI can be calculated on different scales and we chose to calculate 3-month SPEI because *H.
-acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+SPEI can be calculated to represent different temporal scales of drought.
+We used 3-month SPEI because *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
 SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Appendix \_\_**).
 
@@
 cap <- "Timeseries of demographic parameters and drought ocurrence.
-Mean plant size ± standard error (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
+Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
 (b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
@@
 The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
-<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season?-->
+<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
@@
 
+```{r}
+cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
+```
+
+```{r surv-diff, fig.cap=cap}
+include_graphics(here("analysis", "figures", "surv_diff_spei.png"))
+```
+
+Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
+However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
+
 The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.257).
@@
 Plants were on average larger in continuous forest compared to 1-ha fragments.
-In continuous forest, plants had an average of 2.91 shoots and were on average 40.6 cm tall with an average size (shoots $\times$ height) of 150.
-In 1-ha fragments, plants had 2.53 shoots on average and a mean height of 36.2 cm and a mean size of 115.
+<!--# This is across the whole timeseries, which I'm not sure is actually that important.
+What summary statistics would be useful?
+-->In continuous forest, plants had an average of 2.9 ± 1.8 shoots (mean ± standard deviation) and were on average 40.6 ± 26.5 cm tall with an average size (shoots $\times$ height) of 150 ± 175.
+In 1-ha fragments, plants had 2.5 ± 1.5 shoots on average and a mean height of 36.2 ± 24.1 cm and a mean size of 112 ± 141.
 The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).
@@
 
-<!--# There's a lot more going on in the 1-ha fragments.  Hurts my brain to look at it.  Will write another day. -->
+```{r}
+cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
+```
+
+```{r growth-diff, fig.cap=cap}
+include_graphics(here("analysis", "figures", "growth_diff_spei.png"))
+```
+
+The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
+However, growth in continuous forests is higher by up to log(size) = 0.48 <!--# double-check this number --> when the current wet season is unusually wet (SPEI\>1).
 
@@
 Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
-Flowering probability was greater in 1-ha fragments for large plants compared to continuous forest.
+Flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest.
 However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
@@
 That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
+
+```{r}
+cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
+```
+
+```{r flwr-diff, fig.cap=cap}
+include_graphics(here("analysis", "figures", "flwr_diff_spei.png"))
+```
+
+Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
+The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.
 

### 2021-02-22 analysis/paper/vital rate models.Rmd
updates to results

@@
   NULL
-ggsave(here("analysis", "figures", "flwr_diff_spei.png"),plot = s_diff_rotated, width = 7, height = 2.5)
+ggsave(here("analysis", "figures", "flwr_diff_spei.png"),plot = f_diff_rotated, width = 7, height = 2.5)
 ```

### 2021-02-25 analysis/paper/paper.Rmd
revised intro a bit

@@
 
-1.  Organisms simultanously experience multiple forms of natural and anthropogenic disturbance.
+1.  Organisms simultaneously experience multiple forms of natural and anthropogenic disturbance.
     While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
     They could act in concert to exacerbate or dampen population dynamics.
-    <!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].
+    <!--# examples here? -->
 
-    -   Cite Pedro F. Quintana-Ascencio in intro---he will likely be a reviewer.
+2.
+Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+Of particular concern are the effects of fragmentation on species-rich tropical ecosystems.
+In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased from 90% to only 75% [@haddadHabitatFragmentationIts2015].
+    Fragmentation on its own can result in <!--# insert effects on plant populations -->
 
-    -   Cite Jennifer Williams
-
-    -   Emphasize demography.
-        We know about interactions on physiology and other stuff, but we don't know as much about how demography of plants is shaped by how populations respond to multiple stresses, increasingly anthropogenic.
-
-2.  Habitat fragmentation is a major source of environmental stress.
-    **[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]** <!--# fragmentation results in increaed mortality due to increased stress.
-fragmentation itself is not the stressor -->Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-    Of particular concern are the effects of fragmentation on species-rich tropical ecosystems.
-    In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015].
-    Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->
-
-3.  In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change.
+3.
+In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stresses that are increasing with climate change (CITE IPCC?).
     For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
 Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
-
-    -   Drought is bad because...
-
-    -   1997 ENSO drought [@williamsonAmazonianTreeMortality2000]
-
-    -   2005 ENSO drought [@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008]
-
-    -   2010 drought [@lewis2010AmazonDrought2011]
+Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].
 
@@
 This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
+Alternatively, plant populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
 
-5.  Alternatively, the populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation.
-
-    -   [@betts2019].
-
-6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->
-
+5.  The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.
     In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
@@
     This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
+    <!--# I'm not actually testing a size x climate or size x fragmentation interaction explicitly.  Maybe I should not emphasize the variation among size classes in the introduction -->
 
@@
 So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-
-<!--# transition better -->
 
@@
 [-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
-However, in a temperate system... <!--# oak paper --> [@forner2020].
+However, there are no similar studies on tropical understorey plants.
 
 Finally, the long-term data needed to test population-level hypotheses about climate change-fragmentation synergies are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
-However, without multi-decadal sampling one cannot capture enough droughts to quantify their ecological impact, nor the variation in vital rates needed to parameterize demographic models of long-lived species in changing climates [@morrisQuantitativeConservationBiology2002; @teller2016].
+Without multi-decadal sampling one cannot capture enough droughts to quantify their ecological impact, nor the variation in vital rates needed to parameterize demographic models of long-lived species in changing climates [@morrisQuantitativeConservationBiology2002; @teller2016].
 Long-term data are also needed because while some demographic effects of fragmentation or drought on can be detected immediately, others may take years to manifest.
+Lagged effects of climate on plant vital rates may, in fact, be the rule rather than the exception [@eversLaggedDormantSeason].
 
-<!--# more about the heliconia dataset here -->
-
-We investigated the effects of drought on the growth, survival, reproduction, and recruitment of a understory herb in an experimentally fragmented landscape in the Central Amazon.
+We investigated the effects of drought on the growth, survival, reproduction, and recruitment <!--# I haven't included recruitment.
+Should I?
+--> of a understory herb (*Heliconia acuminata*) in an experimentally fragmented landscape in the Central Amazon.
 Specifically, we asked 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
-2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
-3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?
+2) What lag times are most important in considering the effects of drought on vital rates?
+3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
 

### 2021-02-26 analysis/paper/EDA_figure_1.Rmd
used coord_cartesian to set limits instead of scale.  Dodged point was missing.

--- 2021-02-18 analysis/paper/EDA_figure_1.Rmd
@@
 
-There are plants with 0 shoots and 0 ht.  I will remove those, as having no above ground biomass is really not a size, but a different category.
+There are plants with 0 shoots and 0 ht.
+I will remove those, as having no above ground biomass is really not a size, but a different category.
 There's also a couple of plants that are probably data entry errors in height.
@@
 ```
+
 ## Survival
@@
 Survival nearly identical for 1 ha and CF.
+
+When is survival lowest?
+
+```{r}
+demog_post %>% 
+  group_by(year, habitat) %>% 
+  summarize(surv = sum(surv, na.rm = TRUE) / n()) %>% 
+  pivot_wider(names_from = habitat, values_from = surv)
+```
 
@@
 ```
+
 Looks like survival went down year after 2003 drought and during 2007 drought.
@@
   geom_step() +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
+  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.04)) +
   scale_y_continuous("P(survived)") +
@@
   demog_plotdf %>% 
-  ggplot(aes(x = yearmonth, y = size, color = habitat)) +
-  stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
-  stat_summary(geom = "pointrange", fatten = 1) +
-  scale_x_yearmonth(limits = dates, breaks = census_dates, date_minor_breaks ="1 month", expand = expansion(mult = 0.02)) +
-  scale_y_continuous("Size") +
+  ggplot(aes(x = yearmonth, y = log(size), color = habitat)) +
+    stat_summary(
+      geom = "line",
+      fun = "mean",
+      fun.args = list(na.rm = TRUE),
+      position = position_dodge(width = 100),
+      aes(group = habitat)
+    )+
+    stat_summary(
+      geom = "pointrange",
+      fatten = 1,
+      fun.data = "mean_sdl",
+      fun.args = list(mult = 1),
+      position = position_dodge(width = 100)
+    ) +
+  scale_x_yearmonth(
+    # limits = dates,
+    breaks = census_dates,
+    date_minor_breaks = "1 month",
+    expand = expansion(mult = 0.04)
+  ) +
+  scale_y_continuous("log(size)") +
+  coord_cartesian(xlim = dates) + #limits here so points don't get removed
   theme_bw() +
@@
     legend.position = "none"
-  )
-size
+  ) +
+  NULL
 # size/survival/spei
 ```
+
 Average plant size shrank in 2003 drought, but not 2007 drought
+
+### Ridge plot
+
+```{r}
+library(ggridges)
+ggplot(demog_plotdf) +
+  geom_density_ridges(aes(x = log(size), y = as.factor(year), fill = habitat, color = habitat), alpha = 0.5)
+ggplot(demog_plotdf) +
+  geom_violin(aes(y = log(size), x = as.factor(year), fill = habitat), alpha = 0.5, draw_quantiles = c(0.5))
+
+```
 
@@
 ```
+
 How many infloresences given flowering?
@@
       date_minor_breaks = "1 month",
-      expand = expansion(mult = 0.02)
+      expand = expansion(mult = 0.04)
     ) +

### 2021-02-26 analysis/paper/paper.Rmd
updated intro
new working title

@@
 ---
-title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
+title: "Delayed effects of climate lead to demographic divergence in a fragmented Amazonian landscape"
 author:
@@
 
-2.
-Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
+2.  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
     Of particular concern are the effects of fragmentation on species-rich tropical ecosystems.
@@
 
+\
 The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
@@
 The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
-<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
+<
+!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 

### 2021-03-01 analysis/paper/paper.Rmd
Changes in Methods and Introduction UP TO the "interaction" section.
saving here to make it easier to revert if unhappy with new intro focusing on hab diff with no frag x xlimate interaction).

--- 2021-02-25 analysis/paper/paper.Rmd
@@
 ---
-title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
+title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
 author:
@@
       email: scottericr@gmail.com
-      institute: [UF]
+      institute: [UF_WEC]
       correspondence: true
@@
       email: embruna@ufl.edu
-      intitute: [UF]
+      intitute: [UF_WEC,UF_LAS,BDFFP]
       correspondence: false
@@
 institute:
-  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
   - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
+  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 <!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->
+<!-- In response to this I deleted #1; not looking at interaciton of multiple effects and not studying population dyanmics per se in this one. Suggest alluding to this in Discussion instead. -->
 
-1.  Organisms simultaneously experience multiple forms of natural and anthropogenic disturbance.
-    While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-    They could act in concert to exacerbate or dampen population dynamics.
-    <!--# examples here? -->
+ <!-- [1].
+Although the local extinction of species from fragments is common [1], the demographic mechanisms underlying these extinctions are rarely known [2].
+However, it is often assumed – especially in the tropics – that these extinctions are driven by the dramatically altered abiotic conditions in fragments [3] reducing rates of reproduction or survivorship [4-6].
+This is why understanding the way in which landscape structure and environmental conditions interact to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].
+-->
 
-2.
-Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015].
-Of particular concern are the effects of fragmentation on species-rich tropical ecosystems.
-In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased from 90% to only 75% [@haddadHabitatFragmentationIts2015].
-    Fragmentation on its own can result in <!--# insert effects on plant populations -->
+Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
+Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
+Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*.
+However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, ultimately leading to local or regional population declines.
+This is thought to be particularly true in species-rich tropical forests, which continue to be cleared and fragmented at unprecedented rates.
+For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
 
-3.
-In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stresses that are increasing with climate change (CITE IPCC?).
-    For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].
-Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].
+<!-- CITATIONS_1: Haddad, others? -->
+<!-- CITATIONS_2: Bruna et al J Veg Sci 2009, others? -->
+<!-- CITATIONS_3: Bruna 1999 Nature, Laurance et al 1998 Ecology, Zartman 2015 Biotropica.  -->
+<!-- CITATION_4: WRI Report? -->
+<!-- CITATION_5: Brodie J et al 2012. TREE, Scheffers et al 2016. Science -->
+<!-- CITATION_6: Zeng et al 2008 Environmental Research Letters, Mora et al 2013 Nature -->
+
+Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*.
+These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] as well as increased drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
 Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)).
-These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried.
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure \@ref(fig:map).
 In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
@@
 
-*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H.
-acuminata*---are usually found at lower densities and in the more shaded forest understory.
+*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
+While some *Heliconia* species are at least partially self-compatible *[Kress Evolution 1983]*, pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
+data]*.
+
 In our sites *H.
 acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
-Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
-Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
-*Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.
+bourcieri* [@bruna2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
 
@@
 The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments <!--# are we including 10ha fragments?
---> were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
+This study uses data from 1-ha fragments and Continuous Forest sites.
+
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
@@
 Plants that were not found for three consecutive surveys were considered dead.
-In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings <!--# are we using these data? -->.
-There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+There is a seven-fold range in *H.
+acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
@@
 
-Precipitation and potential evapotranspiration data were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
 We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
@@
 Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
-SPEI can be calculated to represent different temporal scales of drought.
-We used 3-month SPEI because *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because -- given its shallow roots and rhizome -- *H.
+acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
 SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
@@
 
-## Statistical Modeling
-
+## Statistical Modeling of _______________ 
+<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
+-->
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
@@
 Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-We chose 36 months as a maximum lag because *H.
+We chose 36 months as a maximum lag because prior *H.
 acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
@@
 The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
-For example, the 1997 El Niño drought was one of the most severe droughts in the Amazon on record [@williamsonAmazonianTreeMortality2000] and correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
-The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
-<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
+For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
+The 2005 dry season (June--October) was also exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008]. <
+!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
@@
 
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].
+<!-- Didham RK, Kapos V, Ewers RM  2012. Rethinking the conceptual foundations of habitat fragmentation research. Oikos 121:161-170. -->
+<!-- Driscoll DA, Banks SC, Barton PS, Lindenmayer DB, Smith AL  2013. Conceptual domain of the matrix in fragmented landscapes. Trends in Ecology & Evolution 28:605-613. -->
+
+1.
+To set up the next paper: Organisms simultaneously experience multiple forms of disturbance.
+While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
+They could act in concert to exacerbate or dampen population dynamics.
+
 Caveats:

### 2021-03-01 analysis/paper/paper.Rmd
Changes in Methods and Introduction UP TO the "interaction" section.
saving here to make it easier to revert if unhappy with new intro focusing on hab diff with no frag x xlimate interaction).

@@
 
-<!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->
-<!-- In response to this I deleted #1; not looking at interaciton of multiple effects and not studying population dyanmics per se in this one. Suggest alluding to this in Discussion instead. -->
-
- <!-- [1].
-Although the local extinction of species from fragments is common [1], the demographic mechanisms underlying these extinctions are rarely known [2].
-However, it is often assumed – especially in the tropics – that these extinctions are driven by the dramatically altered abiotic conditions in fragments [3] reducing rates of reproduction or survivorship [4-6].
-This is why understanding the way in which landscape structure and environmental conditions interact to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].
--->
-
 Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
@@
 Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].
+These changes in forest structure can further exascerbate changes in microclimate near the edges of forest fragments, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
 
-4.
 It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
-    Fragmentation drives several changes in microclimate near forest edges, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+
     This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
@@
 
-5.  The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.
-    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
-    Large established adults could respond differently than seedlings.
-    This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
-    <!--# I'm not actually testing a size x climate or size x fragmentation interaction explicitly.  Maybe I should not emphasize the variation among size classes in the introduction -->
-
-    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
-Smaller/larger plants might be more susceptible to drought.
-So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.
 
@@
 
+Future directions:
+
+* There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+Smaller/larger plants might be more susceptible to drought.
+So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+
+* Synergistic effects of fragmentation and climate (need the stats to catch up with us)
+
 1.

### 2021-03-01 analysis/paper/paper.Rmd
Hardcore edit of the intro: tighter, more efficient.

@@
 Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*.
-However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, ultimately leading to local or regional population declines.
-This is thought to be particularly true in species-rich tropical forests, which continue to be cleared and fragmented at unprecedented rates.
+However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*.
+This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
@@
 <!-- CITATION_4: WRI Report? -->
+
+Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*.
+These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
+All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable - especially for plants, which are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]).
+Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
+
 <!-- CITATION_5: Brodie J et al 2012. TREE, Scheffers et al 2016. Science -->
@@
 
-Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*.
-These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] as well as increased drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
-Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].
-These changes in forest structure can further exascerbate changes in microclimate near the edges of forest fragments, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
+First, most studies of plants in fragments have focused on a single history-stage or process *(CITATIONS)*, making it challenging to draw broader demographic conclusions.
+Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.[@lauranceEffectsStrongDrought2001].
+Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
+These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @teller2016].
+They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest *(CITATION gagnon)*.
+Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason].
 
-It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
+We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
+Herbaceous plants represent up to 25% of plant diversity in tropical forests *(CITATION Gentry)*, are critical food and habitat for myriad species *(CITATION Snow)*, and are economically and culturally vital *(CITATION Nakazono, Ticktin)*.
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied *(CITATION Bruna, Gaoue)*.
+We addressed the following questions:
 
-    This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
-    This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
-    Alternatively, plant populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
+1) Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
+2) What lag times are most important in considering the effects of drought on vital rates?
+[Maybe: are there temporal delays effects in demographic respoinses to drought, and if so what lag times are most critical?]
+3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
 
-The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.
+<!-- Gentry AH, Emmons LH  1987. Geographical variation in fertility, phenology and composition of the understory of neotropical forests. Biotropica 19:216-217. -->
 
-To date, however, this hypothesis remains untested for three primary reasons.
-First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants.
-<!--# example of juvenile but not reproductive study with drought -->.
-<!--# example of reproductive but not juvenile study with drought -->.
-Additionally, while most studies of tropical plant demography focus on trees, little is known about the understory plants that comprise up to \_\_\_\_% of biomass in tropical forests and are ecologically important <!--# citations -->.
+<!-- Snow DW  1981. Tropical frugivorous birds and their food plants - a world survey. Biotropica 13:1-14. -->
 
-Second, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.
-Laurance et al.
-[-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments.
-However, there are no similar studies on tropical understorey plants.
-
-Finally, the long-term data needed to test population-level hypotheses about climate change-fragmentation synergies are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
-Without multi-decadal sampling one cannot capture enough droughts to quantify their ecological impact, nor the variation in vital rates needed to parameterize demographic models of long-lived species in changing climates [@morrisQuantitativeConservationBiology2002; @teller2016].
-Long-term data are also needed because while some demographic effects of fragmentation or drought on can be detected immediately, others may take years to manifest.
-Lagged effects of climate on plant vital rates may, in fact, be the rule rather than the exception [@eversLaggedDormantSeason].
-
-We investigated the effects of drought on the growth, survival, reproduction, and recruitment <!--# I haven't included recruitment.
-Should I?
---> of a understory herb (*Heliconia acuminata*) in an experimentally fragmented landscape in the Central Amazon.
-Specifically, we asked 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest?
-2) What lag times are most important in considering the effects of drought on vital rates?
-3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
+<!-- Nakazono EM, Bruna EM, Mesquita RCG  2004. Experiemental harvesting of the non-timber forest product Ischnosiphon polyphyllus in central Amazonia. Forest Ecology and Management 190:219-225. -->
 
@@
 
-Future directions:
+MAJOR TAKE-HOME MESSAGES
+
+1.
+
+2.
+
+3.
+
+Future directions/Caveats:
 
@@
 
-* Synergistic effects of fragmentation and climate (need the stats to catch up with us)
-
-1.
-To set up the next paper: Organisms simultaneously experience multiple forms of disturbance.
+* Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
 While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-They could act in concert to exacerbate or dampen population dynamics.
+They could act in concert to exacerbate or dampen population dynamics."
 

### 2021-03-01 analysis/supplementary-materials/appendix_SPEI.Rmd
Fix up figure S1

@@
 library(glue)
+library(colorspace)
+library(latex2exp)
 
@@
 bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
+#TODO edit to pull this from github or use bdffp rainfall data package
 
@@
 gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
+#TODO: add citations to this table as footnotes on `cause` or a new column.
 ```
@@
   geom_line(aes(color = dataset), size = 0.5) +
-  geom_segment(data = labs,
-               aes(xend = yearmonth, yend = spei -1)) +
-  geom_label(data = labs,
-             aes(x = yearmonth, y = spei-1, label = year, fill = drought_label)) +
-  scale_x_yearmonth(date_breaks = "2 years") +
-  scale_color_brewer(type = "qual", palette = "Set1") +
-  coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
-  theme_bw()
-annotate_spei(p, alpha = 0.7)
+  scale_x_yearmonth(
+    "Date",
+    date_breaks = "1 year",
+    # date_minor_breaks = "1 month"
+  ) +
+  scale_y_continuous(TeX("SPEI_3_"),
+                     breaks = seq(-2.5, 2.5, 0.5))+
+  scale_color_discrete_qualitative() +
+  coord_cartesian(xlim = c("1995-02-01", "2009-02-01")) +
+  labs(color = "Dataset", linetype = "Dataset") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+p
+annotate_spei(p)
 ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))

### 2021-03-01 analysis/paper/paper.Rmd
minor edit to 1st paragraph intro; added a few notes

@@
 
+<!-- Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->
+
 # Introduction
@@
 Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*.
-However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*.
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*.
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+Despite this, evidence linking fragmentation-related changes in environmental conditions with population-level demographic responses remains scarce.
 
@@
 These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
-All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable - especially for plants, which are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]).
+All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable in fragments - especially for plants, which unlike animals are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]).
 Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
@@
 
+<!-- Fig 1. (map) can go in the Supplement Materials -->
+
 The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
@@
 
+<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->
 ## Focal species

### 2021-03-01 analysis/paper/paper.Rmd
inserted citations and other minor edits

--- 2021-03-01 analysis/paper/paper.Rmd
@@
       email: embruna@ufl.edu
-      intitute: [UF_WEC,UF_LAS,BDFFP]
+      institute: [UF_WEC, UF_LAS, BDFFP]
       correspondence: false
@@
   - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
   - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
-  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
   - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
@@
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, message=FALSE}
 knitr::opts_chunk$set(
@@
 
+<!--# Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->
+
 # Introduction
@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*.
-However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*.
+Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others?
+-->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others?
+-->.
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
+-->.
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+Despite this, evidence linking fragmentation-related changes in environmental conditions with population-level demographic responses remains scarce.
 
-<!-- CITATIONS_1: Haddad, others? -->
-<!-- CITATIONS_2: Bruna et al J Veg Sci 2009, others? -->
-<!-- CITATIONS_3: Bruna 1999 Nature, Laurance et al 1998 Ecology, Zartman 2015 Biotropica.  -->
-<!-- CITATION_4: WRI Report? -->
+<!-- CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html-->
 
-Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*.
-These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
-All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable - especially for plants, which are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]).
+Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017].
+These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
+Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
 Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
 
-<!-- CITATION_5: Brodie J et al 2012. TREE, Scheffers et al 2016. Science -->
-<!-- CITATION_6: Zeng et al 2008 Environmental Research Letters, Mora et al 2013 Nature -->
-
 Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
-First, most studies of plants in fragments have focused on a single history-stage or process *(CITATIONS)*, making it challenging to draw broader demographic conclusions.
+First, most studies of plants in fragments have focused on a single history-stage or process ***(CITATIONS)***, making it challenging to draw broader demographic conclusions.
 Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.[@lauranceEffectsStrongDrought2001].
@@
 These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @teller2016].
-They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest *(CITATION gagnon)*.
-Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason].
+They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
+Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021].
 
 We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
-Herbaceous plants represent up to 25% of plant diversity in tropical forests *(CITATION Gentry)*, are critical food and habitat for myriad species *(CITATION Snow)*, and are economically and culturally vital *(CITATION Nakazono, Ticktin)*.
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied *(CITATION Bruna, Gaoue)*.
+Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
 We addressed the following questions:
@@
 2)  What lag times are most important in considering the effects of drought on vital rates?
-[Maybe: are there temporal delays effects in demographic respoinses to drought, and if so what lag times are most critical?]
+[Maybe: are there temporal delays effects in demographic responses to drought, and if so what lag times are most critical?]
 3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
@@
 
-```{r}
-cap <- "This is just a placeholder figure, but a map or satelite image of BDFFP will go here."
-```
-
-```{r map, fig.cap=cap}
-include_graphics(here("analysis", "figures", "map.png"))
-```
-
 ## Study site
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure \@ref(fig:map).
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure S\_ <!--# map can go in the supplemental materials -->).
 In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
@@
 
+<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->
+
 ## Focal species
@@
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
-While some *Heliconia* species are at least partially self-compatible *[Kress Evolution 1983]*, pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
+While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
 data]*.
@@
 The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
 This study uses data from 1-ha fragments and Continuous Forest sites.
@@
 Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
-SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because -- given its shallow roots and rhizome -- *H.
+SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H.
 acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
 SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Appendix \_\_**).
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
 
-## Statistical Modeling of _______________ 
+## Statistical Modeling of \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
 <!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
 -->
+
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
+DLNMs model potentially delayed effects of predictor variables (e.g.
+SPEI) by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
@@
 
-Where $s_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+Where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
 2017 for details).
@@
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
-The 2005 dry season (June--October) was also exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008]. <
-!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
+The 2005 dry season (June--October) was also exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
@@
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].
-<!-- Didham RK, Kapos V, Ewers RM  2012. Rethinking the conceptual foundations of habitat fragmentation research. Oikos 121:161-170. -->
-<!-- Driscoll DA, Banks SC, Barton PS, Lindenmayer DB, Smith AL  2013. Conceptual domain of the matrix in fragmented landscapes. Trends in Ecology & Evolution 28:605-613. -->
+<!-- Didham RK, Kapos V, Ewers RM  2012.
+Rethinking the conceptual foundations of habitat fragmentation research.
+Oikos 121:161-170.
+--> <!-- Driscoll DA, Banks SC, Barton PS, Lindenmayer DB, Smith AL  2013.
+Conceptual domain of the matrix in fragmented landscapes.
+Trends in Ecology & Evolution 28:605-613.
+-->
 
@@
 
-* There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
     Smaller/larger plants might be more susceptible to drought.
@@
 
-* Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
+-   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
 While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.

### 2021-03-03 analysis/paper/EDA_figure_1.Rmd
small revisions

@@
 spei
+```
+```{r}
+xa %>% select(-spei_history, -L) %>% ungroup() %>%  summarize(spei = mean(spei)) %>% filter(year(yearmonth) == 2003) %>% arrange(spei)
 ```
@@
     ) +
-    scale_y_continuous("P(flowering | shts ≥ 4)") +
+    scale_y_continuous("P(flowering)") +
   theme_bw() +
@@
 ```{r}
-demog_plotdf %>% filter(year == 2003) %>% group_by(habitat) %>% summarize(flwr = mean(flwr, na.rm = TRUE))
+demog_plotdf %>% filter(year == 2003, shts >= 4) %>% group_by(habitat) %>% summarize(flwr = mean(flwr, na.rm = TRUE))
+```
+```{r}
+demog_plotdf %>% 
+  group_by(habitat, year) %>% 
+  filter(!is.na(flwr)) %>% #removes both dead plants and plants with no data.
+  summarize(flwr = sum(flwr)/n()) %>% 
+  summarize(flwr_mean = mean(flwr),
+            flwr_sd = sd(flwr))
 ```

### 2021-03-03 analysis/paper/paper.Rmd
small revisions

@@
 Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
-There is a pronounced dry season from June to October **(Appendix?)**.
+There is a pronounced dry season from June to October **(Figure?)**.
 
@@
 
-## Statistical Modeling of \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+## Statistical Modeling of Vital Rates
 
@@
 \begin{gather}
-    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-24}) \\
+    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
     \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j 
@@
 (b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
-c) shows the proportion of mature plants (with 4 or more shoots) flowering each census.
+c) shows the proportion of plants with 4 or more shoots flowering each census.
 In d), monthly 3-month SPEI is plotted.
@@
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
-The 2005 dry season (June--October) was also exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+Our SPEI data show the 2005 dry season to be a moderate drought.
 <!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with the most severe drought in the timeseries (Figure \@ref(fig:eda)d).
-Two noticeable peaks in the proportion of flowering mature plants occur in the most severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with an severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
 
@@
 
-Overall survival rate across the entire dataset is 0.95 in both CF and 1-ha.
-The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004.
+Overall survival rate across the entire dataset was the same (0.95) in both CF and 1-ha.
+The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
 The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
@@
 
-Log transformed size in year t was a significant predictor of log transformed size in year t+1 (p \< 0.001).
+Size in year t was a significant predictor of size in year t+1 (p \< 0.001).
 The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:growth-size)).
@@
 The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
-However, growth in continuous forests is higher by up to log(size) = 0.48 <!--# double-check this number --> when the current wet season is unusually wet (SPEI\>1).
+However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
+This is due to reduced growth in fragments under wet conditions.
 
@@
 
-Probability of flowering was generally higher in continuous forest compared to 1-ha fragments.
-The exceptions to this are in 2006 and 2008 when the proportion of flowering plants was lower in continuous forest (Figure \@ref(fig:eda)c).
-The most striking difference between habitat types in flowering proportion was in 2003 when 9.7% of plants flowered in the continuous forest plots and only 2.7% flowered in fragment plots.
-<!--# update to be only for large plants -->
+The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
+In fact, the proportion of flowering plants was higher in continuous forest in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
+The most striking difference between habitat types in flowering proportion coincided with a severe drought in 2003 when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest and 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
+The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
+<!--# not sure how necessary this is -->
 
@@
 
-1.
+1.  Survival and growth rates are similar in CF and fragments, but more strongly effected by SPEI in fragments.
 
-2.
+2.  Flowering probability is higher in CF (overall and for averaged sized plants), but large plants have higher flowering probability in fragments than in CF.
 
 3.
+Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability (cite @nihadRelationshipStressFlowering2018 )
 

### 2021-03-03 analysis/paper/vital rate models.Rmd
small revisions

@@
   ) %>%
-  mutate(diff = cf_est - frag_est) %>% 
+  mutate(diff = cf_est - frag_est) %>% filter(diff == max(diff, na.rm = TRUE))
   ggplot(aes(x = cf_L, y = cf_spei_history, fill = diff, z = diff)) +
@@
         s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") + 
         s(plot, bs = "re") +
@@
           bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 35), 
           xt = list(bs = "cr")),
@@
 
-This model changes **drastically** depending on whether year is included as a random effect.
-Does that mean ther random effect is important?
-Or is it because it is collinear / concurve with SPEI history?
-
 ```{r}
@@
         s(log_size_prev, bs = "cr") +
-        # s(year_fac, bs = "re") +
         s(plot, bs = "re") +
@@
           bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 35),
           xt = list(bs = "cr")),
@@
 ```
+
+```{r}
+plogis(coef(f_post_1ha)[1])
+plogis(coef(f_post_cf)[1])
+```
+Overall probability of flowering higher in continuous forest (0.008) compared to 1ha fragments (0.003)
 

### 2021-03-03 analysis/supplementary-materials/appendix_SPEI.Rmd
Changed imputed BDFFP SPEI to only use BDFFP for precipitation and to use eto from the mean of the two EMBRAPA stations.

@@
 library(latex2exp)
+library(janitor)
+library(Amelia)
 
@@
 
-*Last compiled: `r Sys.Date()`*
-# TODO
+*Last compiled: `r Sys.Date()`* \# TODO
 
 -   Write
-- Fix colors and labels and such on fig S1
+-   ~~Fix colors and labels and such on fig S1~~
 -   Make a figure S2 that compares SPEI across BDFFP from east to west (3 panels)
-- Maybe add imputed data to this (or maybe not)
+-   ~~Maybe add imputed data to this (or maybe not)~~
+-   Move imputation code to this notebook. This requires finishing up the bdffp precip R package.
 
@@
 # Precip Sources
-xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
 sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
@@
 # Already calculated SPEI
-gspei <-
-  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
-             delim = ";") %>% 
-  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
-  mutate(yearmonth = yearmonth(date), date = NULL)
 
-bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
-#TODO edit to pull this from github or use bdffp rainfall data package
+# bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
+# #TODO edit to pull this from github or use bdffp rainfall data package
 
@@
 
-## Data Source Descriptions
+We compared 3-month SPEI data calculated from a number of sources for the BDFFP site including ground-based data from on-site rain gauges, two nearby weather stations, satellite products, and gridded land station based datasets.
 
-### BDFFP rain gauge data
-
-Daily rain gauge data collected from 1987--2010 at BDFFP camps.
-The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010.
-There we also a large number of observations that were accumulated rainfall over two or more days.
-We removed multi-day accumulations as it was impossible to know how rainfall was distributed over the prior days.
-Missing values were then imputed after averaging rainfall data by ranch.
-This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps.
-Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed.
-Multiple imputation produced \_\_\_ <!--# 10?
---> imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using potential evapotranspiration from Xavier et al.
-[-@xavierDailyGriddedMeteorological2016] to calculate climatic water balance <!--# why not use mean temperature from weather stations?
--->.
-Then, the <!--# number --> resulting SPEI datasets were averaged to produce a single dataset.
-
-### Meterological stations
-
--   EMBRAPA stations
-
-    -   Estacao Rio Preto da Eva: -2.65º, -59.65º
-
-    -   Estacao Manaus: -2.65º, -60.35º (note, this is not in Manaus city, but NW of it.)
-
-    -   Date Range: 1980-present
-
-    -   Unclear how evapotranspiration is calculated from the [information provided](https://www.embrapa.br/en/busca-de-solucoes-tecnologicas/-/produto-servico/6044/infoclima---informacoes-climaticas).
-
-### Ground-based gridded products
-
--   Daily gridded meteorological variables in Brazil [@xavierDailyGriddedMeteorological2016]
-
-    -   Resolution 0.25º x 0.25º
-    -   Includes precipitation and potential evapotranspiration
-    -   PET calculated by Pennman-Monteith method
-    -   Source: 3,625 ground-based weather stations across Brazil
-    -   Date range: 1980--2015
-
--   Global SPEI [@santiago2017]
-
-    -   Resolution: 0.5º x 0.5º
-
-    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
-
-    -   Source: Precipitation from CRU TS version 4.03, a gridded product based on land stations.
-
-    -   Date range: 1901--2018
-
-### Satelite and reanalysis products
-
--   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
-    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
-    -   Resolution: 0.25º x 0.25º
-    -   Includes: Monthly precip
-    -   Doesn't include PET, so I guess I'll take that from the ground stations??
-    -   Date range: 1998--2019
-
--   PERSIANN-CDR
-
-    -   only rainfall, but recommended by [@zhaoEvaluatingDroughtMonitoringUtility2019] for Amazonia.
-
-    -   0.25º resolution
-
-    -   1983--present
-
-# Xavier et al.
-
-```{r xa-spei}
-#aggregate monthly, then calculate climate water balance.
-xa_mon <-
-  xa %>% 
-  unite(latlon, lat, lon, sep = ", ") %>% 
-  mutate(yearmonth = yearmonth(date), .after = date) %>% 
-  group_by(yearmonth, latlon) %>% 
-  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
-  as_tsibble(index = yearmonth, key = latlon)
-
-xa_spei <-
-  xa_mon %>% 
-  mutate(cb = precip - eto) %>% 
-  group_by(latlon) %>%
-  group_nest() %>% 
-  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
-  unnest(c(xa_spei, data)) %>% 
-  select(-cb)
-```
-
-# EMBRAPA stations
+## EMBRAPA
 
@@
   select(yearmonth, station, st_spei) %>% 
-  pivot_wider(names_from= station, values_from = st_spei, names_prefix = "spei_") %>% 
+  pivot_wider(names_from = station,
+              values_from = st_spei,
+              names_prefix = "spei_") %>% 
   rowwise() %>% 
@@
 
-# TRMM
+Two weather stations within \_\_\_ km of BDFFP had data available through EMBRAPA <!--# what is this -->, Rio Preto da Eva to the East (-2.65º, -59.65º) and the Manaus station to the West (-2.65º, -60.35º).
+We downloaded daily precipitation and potential evapotranspiration data, aggregated it to monthly, and calculated 3 month SPEI for each station separately.
+We then averaged the SPEI values for two stations at each month to produce monthly estimates for the BDFFP site.
+
+## BDFFP rain gauge data
+
+```{r}
+# Read in data 
+bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv?token=AGB2K35OCIPUILNGOD6VXV3AH7XOU")
+
+# Prep data for imputation ------------------------------------------------
+
+# remove accumulations
+bdffp2 <-
+  bdffp_daily %>% 
+  filter(!flag %in% c("A", "U"))
+
+# spread and add rows for missing dates
+bdffp_wide <- 
+  bdffp2 %>% 
+  #complete all the dates
+  as_tsibble(key = site, index = date) %>% 
+  fill_gaps() %>% 
+  select(date, site, precip) %>% 
+  as_tibble() %>% 
+  pivot_wider(names_from = site, values_from = precip) %>% 
+  clean_names()
+
+# Cluster sites by ranch to reduce missingness
+bdffp_wide2 <-
+  bdffp_wide %>% 
+  rowwise() %>% 
+  mutate(colosso_clust = mean(c(colosso, florestal, cabo_frio, gaviao),
+                              na.rm = TRUE),
+         km_clust = mean(c(km37, km41),
+                         na.rm = TRUE)) %>% 
+  mutate(across(colosso_clust:km_clust, ~ifelse(is.nan(.x), NA, .x))) %>% 
+  select(-cabo_frio, -colosso, -florestal, -gaviao, -km37, -km41)
+
+# add additional representations of time for Amelia.
+full_wide <- 
+  bdffp_wide2 %>% 
+  mutate(
+    year = year(date),
+    doy = yday(date),
+    yearmonth = yearmonth(date),
+    .before = everything()
+  )
+```
+
+```{r}
+# Impute missing data -----------------------------------------------------
+# round(runif(1, 1, 1000))
+set.seed(937)
+all_cols <- colnames(select(full_wide, -year, -doy, -date))
+
+# histograms <-
+#   map(all_cols, ~{
+#     ggplot(full_wide, aes(full_wide[[.x]])) + geom_histogram() + labs(title = .x)
+#   })
+# histograms
+
+# variables with log-normal-ish distributions
+
+log_cols <-
+  all_cols[str_detect(all_cols, "precip") |
+             all_cols %in% c("km_clust", "colosso_clust", "porto_alegre", "dimona")]
+
+imp <- 
+  amelia(
+    as.data.frame(full_wide),
+    p2s = 0,
+    m = 10,
+    ts = "doy",
+    cs = "year",
+    intercs = TRUE,
+    polytime = 3,
+    logs = log_cols,
+    idvars = c("date"),
+    parallel = "multicore",
+    ncpus = detectCores() - 1,
+    empri = .01 * nrow(full_wide) #ridge penalty because of high degree of missingness
+  )
+# imp$imputations[[1]]
+```
+
+```{r}
+# Prep eto data for SPEI calculation
+# Aggregate by month
+imp_mon <-
+  map(imp$imputations, ~{
+    .x %>% 
+      rename_with(~paste0(., ".precip"), c(dimona, porto_alegre, colosso_clust, km_clust)) %>% 
+      select(date, ends_with(".precip")) %>% 
+      mutate(yearmonth = tsibble::yearmonth(date)) %>% 
+      group_by(yearmonth) %>% 
+      summarize(across(-date, ~sum(.x, na.rm = TRUE))) %>% 
+      #remove first and last month, as they aren't complete
+      slice(-1, -nrow(.))
+  })
+# Add mean ETO from EMBRAPA data
+imp_mon <-
+  map(imp_mon, ~ {
+    left_join(.x, embrapa_mon %>% summarize(eto = mean(eto)), by = "yearmonth")
+  }) %>% 
+  map(., ~{
+    mutate(.x, across(ends_with(".precip"),
+                      ~ .x - eto,
+                      .names = "{str_remove(col, '.precip')}.cb"))
+  })
+# imp_mon$imp1
+```
+
+```{r}
+# Calculate SPEI3 (i.e. 3-month window SPEI)
+
+imp_spei <-
+  map(imp_mon, ~{
+  .x %>% 
+    mutate(across(ends_with(".cb"), ~as.numeric(SPEI::spei(.x, scale = 3)$fitted),
+                  .names = "{str_remove(col, '.cb')}.spei")) %>%
+      select(yearmonth, ends_with(c(".precip", ".spi", ".spei")))
+}) 
+# imp_spei$imp1
+# Combine multiply imputed SPI and SPEI results by taking mean
+bdffp_spei <- 
+  imp_spei %>%
+  bind_rows(.id = "imp") %>% 
+  #replace any infinite values with NAs
+  mutate(across(ends_with(".spei"), ~ ifelse(is.infinite(.x), NA, .x))) %>% 
+  group_by(yearmonth) %>% 
+  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
+  #replace NaN's with NAs
+  mutate(across(ends_with(".spei"), ~ ifelse(is.nan(.x), NA, .x))) %>% 
+  rowwise() %>% 
+  mutate(bdffp_spei = mean(c_across(ends_with(".spei")))) %>% 
+  ungroup()
+```
+
+We obtained rain gauge data collected from 1987--2010 at BDFFP camps.
+The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010.
+There we also a large number of observations representing accumulated rainfall over two or more days.
+We removed multi-day accumulations as it was impossible to know how rainfall was distributed over the prior days.
+Missing values were then imputed after averaging rainfall data by ranch.
+This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps.
+Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed [@honakerAmeliaIIProgram2011].
+We produced 10 imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using the averaged potential evapotranspiration from the two EMBRAPA weather stations<!--# why not use mean temperature from weather stations?
+-->.
+Then, the 10 resulting SPEI datasets were averaged to produce a single dataset.
+
+## Xavier et al.
+
+```{r xa-spei}
+xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
+#aggregate monthly, then calculate climate water balance.
+xa_mon <-
+  xa %>% 
+  unite(latlon, lat, lon, sep = ", ") %>% 
+  mutate(yearmonth = yearmonth(date), .after = date) %>% 
+  group_by(yearmonth, latlon) %>% 
+  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
+  as_tsibble(index = yearmonth, key = latlon)
+
+xa_spei <-
+  xa_mon %>% 
+  mutate(cb = precip - eto) %>% 
+  group_by(latlon) %>%
+  group_nest() %>% 
+  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
+  unnest(c(xa_spei, data)) %>% 
+  select(-cb)
+```
+
+Gridded data from Xavier et al. [-@xavierDailyGriddedMeteorological2016] was used to calculate SPEI as described in the main text.
+For figure S1, we used an average of SPEI values across all grid cells.
+
+## TRMM
 
@@
 
-# Imputed data (averaged across for BDFFP)
+Monthly precipitation data from TRMM was used in combination with potential evapotranspiration data from Xavier et al. [-@xavierDailyGriddedMeteorological2016].
 
-```{r}
-bdffp_spei <- 
-  bdffp %>% group_by(date) %>% 
-  summarize(bdffp_spei = mean(spei)) %>% 
+## Global SPEI
+
+```{r global-spei}
+gspei <-
+  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
+             delim = ";") %>% 
+  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
   mutate(yearmonth = yearmonth(date), date = NULL)
 ```
+
+Additionally, we downloaded SPEI data from the Global SPEI dataset [@begueriaSbegueriaSPEIbaseVersion2017].
+These calculations were made using precipitation data from a gridded product based on land stations and evapotranspiration estimated with the Pennman-Montieth method.
 
@@
   full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
-  full_join(bdffp_spei) %>% 
+  full_join(bdffp_spei %>% select(yearmonth, bdffp_spei)) %>% 
   pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
@@
 # Table of Expected Droughts
+
 Annotate worst droughts.\
@@
 
+# Notes
+
+### Meterological stations
+
+-   EMBRAPA stations
+
+    -   Estacao Rio Preto da Eva: -2.65º, -59.65º
+
+    -   Estacao Manaus: -2.65º, -60.35º (note, this is not in Manaus city, but NW of it.)
+
+    -   Date Range: 1980-present
+
+    -   Unclear how evapotranspiration is calculated from the [information provided](https://www.embrapa.br/en/busca-de-solucoes-tecnologicas/-/produto-servico/6044/infoclima---informacoes-climaticas).
+
+### Ground-based gridded products
+
+-   Daily gridded meteorological variables in Brazil [@xavierDailyGriddedMeteorological2016]
+
+    -   Resolution 0.25º x 0.25º
+    -   Includes precipitation and potential evapotranspiration
+    -   PET calculated by Pennman-Monteith method
+    -   Source: 3,625 ground-based weather stations across Brazil
+    -   Date range: 1980--2015
+
+-   Global SPEI [@santiago2017]
+
+    -   Resolution: 0.5º x 0.5º
+
+    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
+
+    -   Source: Precipitation from CRU TS version 4.03, a gridded product based on land stations.
+
+    -   Date range: 1901--2018
+
+### Satelite and reanalysis products
+
+-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
+
+    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
+    -   Resolution: 0.25º x 0.25º
+    -   Includes: Monthly precip
+    -   Doesn't include PET, so I guess I'll take that from the ground stations??
+    -   Date range: 1998--2019
+
+-   PERSIANN-CDR
+
+    -   only rainfall, but recommended by [@zhaoEvaluatingDroughtMonitoringUtility2019] for Amazonia.
+
+    -   0.25º resolution
+
+    -   1983--present
+
 # References

### 2021-03-03 analysis/supplementary-materials/appendix_SPEI.Rmd
revised appendix so it knits correctly

@@
 knitr::opts_chunk$set(
-  include = FALSE,
+  include = TRUE,
   echo = FALSE,
@@
 
--   Write
--   ~~Fix colors and labels and such on fig S1~~
--   Make a figure S2 that compares SPEI across BDFFP from east to west (3 panels)
--   ~~Maybe add imputed data to this (or maybe not)~~
--   Move imputation code to this notebook. This requires finishing up the bdffp precip R package.
+-   Write brief text
+-   Possibly make a figure S2 that compares SPEI across BDFFP from east to west (3 panels), using all data with similar resolution
 
@@
 # Read in data 
-bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv?token=AGB2K35OCIPUILNGOD6VXV3AH7XOU")
-
+#Change to something like this once repo is public
+# bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv")
+bdffp_daily <- read_csv(here("analysis", "data", "raw_data", "daily_precip.csv"))
 # Prep data for imputation ------------------------------------------------
@@
 
-```{r}
+```{r imputation, cache=TRUE}
 # Impute missing data -----------------------------------------------------
@@
 
-# Table of Expected Droughts
+<!-- Annotate worst droughts.\ -->
+<!-- Expected: -->
 
-Annotate worst droughts.\
-Expected:
-
--   2004 & 2006: weak ENSO
--   2002 & 2009: moderate ENSO
--   1997: severe ENSO
--   2005 & 2010: severe SST
+<!-- -   2004 & 2006: weak ENSO -->
+<!-- -   2002 & 2009: moderate ENSO -->
+<!-- -   1997: severe ENSO -->
+<!-- -   2005 & 2010: severe SST -->
 
@@
   tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM, BDFFP))
-table
-gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
+# table
+# gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
 #TODO: add citations to this table as footnotes on `cause` or a new column.
+```
+
+```{r}
+cap <- "Comparison of SPEI values and drought categories across datasets for major drought events in the Amazon.
+Mild, moderate, severe, and extreme drought are represented by yellow, goldenrod, orange, and red, respectively.
+Cause indicates whether the drought was caused by the El Niño Southern Oscillation (ENSO) or Atlantic Sea Surface Temperatures (AMO).
+The lowest SPEI value for the year identified by each dataset is reported, along with the month that minimum value occurred in."
+```
+
+```{r spei-table, fig.cap=cap}
+knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
 ```
@@
 ```
+```{r}
+cap <- "Three-month SPEI at BDFFP calculated using 5 precipitation datasets.
+Drought categories mild, moderate, severe, and extreme drought are represented by yellow, goldenrod, orange, and red panels, respectively."
+```
 
-```{r fig.height=5, fig.width=10}
+```{r spei-plot, fig.cap=cap, fig.height=5, fig.width=10}
 labs <-
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-p
+# p
 annotate_spei(p)
-ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
+# ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
 ```
 
-# Notes
+<!-- # Notes -->
 
-### Meterological stations
+<!-- ### Meterological stations -->
 
--   EMBRAPA stations
+<!-- -   EMBRAPA stations -->
 
-    -   Estacao Rio Preto da Eva: -2.65º, -59.65º
+<!--     -   Estacao Rio Preto da Eva: -2.65º, -59.65º -->
 
-    -   Estacao Manaus: -2.65º, -60.35º (note, this is not in Manaus city, but NW of it.)
+<!--     -   Estacao Manaus: -2.65º, -60.35º (note, this is not in Manaus city, but NW of it.) -->
 
-    -   Date Range: 1980-present
+<!--     -   Date Range: 1980-present -->
 
-    -   Unclear how evapotranspiration is calculated from the [information provided](https://www.embrapa.br/en/busca-de-solucoes-tecnologicas/-/produto-servico/6044/infoclima---informacoes-climaticas).
+<!--     -   Unclear how evapotranspiration is calculated from the [information provided](https://www.embrapa.br/en/busca-de-solucoes-tecnologicas/-/produto-servico/6044/infoclima---informacoes-climaticas).
+-->
 
-### Ground-based gridded products
+<!-- ### Ground-based gridded products -->
 
--   Daily gridded meteorological variables in Brazil [@xavierDailyGriddedMeteorological2016]
+<!-- -   Daily gridded meteorological variables in Brazil [@xavierDailyGriddedMeteorological2016] -->
 
-    -   Resolution 0.25º x 0.25º
-    -   Includes precipitation and potential evapotranspiration
-    -   PET calculated by Pennman-Monteith method
-    -   Source: 3,625 ground-based weather stations across Brazil
-    -   Date range: 1980--2015
+<!--     -   Resolution 0.25º x 0.25º -->
+<!--     -   Includes precipitation and potential evapotranspiration -->
+<!--     -   PET calculated by Pennman-Monteith method -->
+<!--     -   Source: 3,625 ground-based weather stations across Brazil -->
+<!--     -   Date range: 1980--2015 -->
 
--   Global SPEI [@santiago2017]
+<!-- -   Global SPEI [@santiago2017] -->
 
-    -   Resolution: 0.5º x 0.5º
+<!--     -   Resolution: 0.5º x 0.5º -->
 
-    -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
+<!--     -   Provides SPEI calculated using monthly precipitation and potential evapotranspiration estimated with the Pennman-Monteith method, which is considered a superior method to Thornwaite, which only takes temperature into account.
+-->
 
-    -   Source: Precipitation from CRU TS version 4.03, a gridded product based on land stations.
+<!--     -   Source: Precipitation from CRU TS version 4.03, a gridded product based on land stations. -->
 
-    -   Date range: 1901--2018
+<!--     -   Date range: 1901--2018 -->
 
-### Satelite and reanalysis products
+<!-- ### Satelite and reanalysis products -->
 
--   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
+<!-- -   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? --> -->
 
-    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
-    -   Resolution: 0.25º x 0.25º
-    -   Includes: Monthly precip
-    -   Doesn't include PET, so I guess I'll take that from the ground stations??
-    -   Date range: 1998--2019
+<!--     -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7 -->
+<!--     -   Resolution: 0.25º x 0.25º -->
+<!--     -   Includes: Monthly precip -->
+<!--     -   Doesn't include PET, so I guess I'll take that from the ground stations?? -->
+<!--     -   Date range: 1998--2019 -->
 
--   PERSIANN-CDR
+<!-- -   PERSIANN-CDR -->
 
-    -   only rainfall, but recommended by [@zhaoEvaluatingDroughtMonitoringUtility2019] for Amazonia.
+<!--     -   only rainfall, but recommended by [@zhaoEvaluatingDroughtMonitoringUtility2019] for Amazonia. -->
 
-    -   0.25º resolution
+<!--     -   0.25º resolution -->
 
-    -   1983--present
+<!--     -   1983--present -->
 

### 2021-03-05 analysis/paper/vital rate models.Rmd
remove seedlings since not used

@@
   demog_post %>% 
-  filter(habitat == "1-ha")
-
-demog_sdlg <- 
-  demog_full %>% 
-  filter(code_notes == "sdlg (1)")
-
-demog_sdlg_cf <- 
-  demog_sdlg %>% 
-  filter(habitat == "CF")
-
-demog_sdlg_1ha <- 
-  demog_sdlg %>% 
   filter(habitat == "1-ha")
@@
 demog_full %>% filter(ht < 1)
+demog_full %>% 
+  # filter(log_size > 7) %>%
+  filter(shts >= 4) %>% 
+  group_by(habitat) %>% 
+  summarise(flower_prob = mean(flwr, na.rm = TRUE))
+
+coef(f_post_1ha)[1] %>% plogis()
 ```
+
+Opposite of what plot shows.
 

### 2021-03-07 analysis/paper/paper.Rmd
Suggestions from María

@@
       correspondence: true
-  - Maria Uriarte:
+  - María Uriarte:
       email: mu2126@columbia.edu
@@
 -->.
-This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
+This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999] <!--# also cite "Does forest
+fragmentation cause an increase in forest temperature" -->, which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
@@
 These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
+All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
 Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
@@
 First, most studies of plants in fragments have focused on a single history-stage or process ***(CITATIONS)***, making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments.[@lauranceEffectsStrongDrought2001].
+Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
@@
 They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
-Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021].
+Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021] <!--# add citations from María.Anderegg et al, Schwalm et al.
+Kannenberg et al.-->.
 
 We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
-Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
+<!--# add a sentence about "famous" amazonian droughts --> Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
 Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
-acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
@@
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@bruna2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 Plants that were not found for three consecutive surveys were considered dead.
-There is a seven-fold range in *H.
-acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
@@
 SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H.
-acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
+acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
 SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
@@
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs model potentially delayed effects of predictor variables (e.g.
-SPEI) by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
+They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
 This models a non-linear relationship between predictor and response (e.g.
@@
 We chose 36 months as a maximum lag because prior *H.
-acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
+acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
@@
 
-Where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
 2017 for details).
@@
 
-Where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
+where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
 These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodsimonn.ChapterSmoothers] and Gasparrini et al.
 [-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
-Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex.
-A random effect of plot ID on the intercept is represented by $a_j$.
+Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
+number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
+Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
+A random effect of plot ID on the intercept is represented by $a_j$, allowing for plot characteristics to influence average vital rates.
 For plant growth the response was plant size (natural log of height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution.
@@
 The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
-Our SPEI data show the 2005 dry season to be a moderate drought.
+Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
 <!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with an severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
 Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
@@
 The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to unusually wet weather (SPEI \>0.5), but with a moderate drought the preceding October.
+The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to wetter than average conditions (SPEI \> 0.5), but with a moderate drought the preceding October.
 
@@
 Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
-For 1-ha fragments, recent SPEI values had the strongest effect on survival with the highest survival near SPEI of 0 and decreasing survival in both drier and wetter conditions (Figure \@ref(fig:surv-spei)a).
-In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 producing slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
-However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e. not the most recent dry season, but the one before that), also effected survival probabilities.
+For 1-ha fragments, recent SPEI values (i.e.
+within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
+In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
+However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e.
+not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
 Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
@@
 
+1.  Consequences for demographers
+
+    1.  Delayed effects of environmental drivers on vital rates are rarely included in demographic models, but they may be the rule rather than the exception.
+        [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+        This study is another example of one where delayed effects of climate are important and sometimes better predictors of vital rates than more recent conditions (e.g.
+drought two dry seasons ago and flowering probability).
+
+    2.  Very few demographic studies have the data to fit models that include delayed effects.
+
+2.  Consequences for biology
+
+    1.  What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
+
+    2.
+Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
+        <!--# from model intercepts -->
+
+3.  Buffering
+
+    1.
+We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
+The effects of variation in SPEI were more extreme in fragments for survival and growth.
+We did not see this pattern with flowering probability, likely because flowering is overall much less likely in fragments, probably due to smaller plant size.
+
 MAJOR TAKE-HOME MESSAGES
@@
 Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability (cite @nihadRelationshipStressFlowering2018 )
+
+    1.
+In Manaus, canopy is as green or greener in dry season compared to wet season @guanPhotosyntheticSeasonalityGlobal2015.
+So probably direct rather than indirect effects of canopy openness (need to think about this still)
 

### 2021-03-08 analysis/paper/EDA_figure_1.Rmd
added weather normals figure

@@
          yearmonth = yearmonth(date))
+```
+
+## Seasonality
+
+Use INMET normals for Manaus from https://portal.inmet.gov.br/normais
+
+```{r}
+normals <-
+  tibble(
+  month = 1:12,
+  temp_mean = c(26.30, 26.30, 26.30, 26.40, 26.60, 26.70, 27.00, 27.60, 28.00, 28.00, 27.60, 26.90),
+  temp_min  = c(23.10, 23.10, 23.20, 23.20, 23.40, 23.00, 23.10, 23.40, 23.70, 23.90, 23.70, 23.50),
+  temp_max  = c(30.90, 30.80, 30.90, 31.00, 31.10, 31.40, 32.10, 33.10, 33.50, 33.40, 32.60, 31.70),
+  precip    = c(287.0, 295.1, 300.0, 319.0, 246.9, 118.3, 75.40, 64.30, 76.30, 104.1, 169.2, 245.6)
+)
+
+precip <-
+  ggplot(normals, aes(x = as.factor(month), y = precip)) + 
+  geom_col(fill = "darkblue") +
+  scale_y_continuous("Precipitation (mm)", expand = expansion(mult = c(0.05, 0.1))) +
+  theme_bw()
+
+temp <- 
+  ggplot(normals, aes(x = month)) +
+  geom_line(aes(y = temp_mean), color = "red") +
+  geom_line(aes(y = temp_min), color = "red", linetype = 2) +
+  geom_line(aes(y = temp_max), color = "red", linetype = 2) +
+  scale_x_continuous("Month", breaks = 1:12) +
+  scale_y_continuous("Temperature (ºC)", expand = expansion(mult = 0.3)) +
+  theme_bw()
+
+(precip + theme(axis.title.x = element_blank(), axis.text.x = element_blank()))/
+  (temp) + plot_annotation(tag_levels = "a", tag_suffix = ")")
+ggsave(here("analysis", "figures", "manaus_normals.png"))
 ```

### 2021-03-08 analysis/paper/paper.Rmd
added weather normals figure

@@
 
+```{r}
+cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+Precipitation (a) shows a marked dry seasons from June through October.
+Temperature (b) varies less throughout the year.
+Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
+Data from Brazilian National Institute of Meterology (INMET)."
+```
+
+```{r normals, fig.cap=cap}
+include_graphics(here("analysis", "figures", "manaus_normals.png"))
+```
+
 The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
@@
 Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
-There is a pronounced dry season from June to October **(Figure?)**.
+There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
 
@@
 The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
+<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->
 
@@
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g.
-SPEI) affect an outcome (e.g.
-growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g.
-between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].

### 2021-03-09 analysis/paper/vital rate models.Rmd
edits based on María's comments

@@
           bs = "cb",
-          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
+          k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
           xt = list(bs = "cr")),

### 2021-03-09 analysis/supplementary-materials/appendix_SPEI.Rmd
edits to SPEI to make more like knittable document

@@
 
-# Outline
+When looking for data to calculate SPEI for the BDFFP site, there were a number of possible options.
+A review of global precipitation datasets [@sunReviewGlobalPrecipitation2018] shows that ground-based products are more accurate compared to satellite products for Brazil specifically.
 
--   A review of global precip datasets (Sun et al. 2018) recommends ground-based products over satellite products for Brazil.
-
--   The global SPEI database is calculated from ground-based product (CRU) and a uses method for estimating evapotranspiration that takes more than just temperature into account.
-
--   There's also a ground-based product specifically for Brazil (Xavier et al).
-
--   We have rain gauge data that we can get SPEI from if we use imputation (informed by EMBRAPA station data)
-
-```{r read-data}
-# Precip Sources
-sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
-# trmm <-
-
-# Already calculated SPEI
-
-# bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
-# #TODO edit to pull this from github or use bdffp rainfall data package
-
-```
-
-# Methods
-
-We compared 3-month SPEI data calculated from a number of sources for the BDFFP site including ground-based data from on-site rain gauges, two nearby weather stations, satellite products, and gridded land station based datasets.
+We calculated 3-month SPEI data calculated from a number of sources for the BDFFP site including on-site rain gauges, nearby weather stations, a satellite product, and a gridded land-based product.
+We then compared the calculated SPEI values to each other and to reported qualitative drought severity from literature.
 
@@
 
-Two weather stations within \_\_\_ km of BDFFP had data available through EMBRAPA <!--# what is this -->, Rio Preto da Eva to the East (-2.65º, -59.65º) and the Manaus station to the West (-2.65º, -60.35º).
-We downloaded daily precipitation and potential evapotranspiration data, aggregated it to monthly, and calculated 3 month SPEI for each station separately.
+Two weather stations within \_\_\_ km of BDFFP had data available through EMBRAPA (*Empresa Brasileira de Pesquisa Agropecuária*): Rio Preto da Eva to the East (-2.65º, -59.65º) and the Manaus station to the West (-2.65º, -60.35º).
+We downloaded daily precipitation and potential evapotranspiration data through the *infoclima* portal (<https://www.cnpaf.embrapa.br/infoclima/>), aggregated it to monthly, and calculated 3 month SPEI for each station separately.
 We then averaged the SPEI values for two stations at each month to produce monthly estimates for the BDFFP site.
@@
 
-We obtained rain gauge data collected from 1987--2010 at BDFFP camps.
+We obtained rain gauge data collected from 1987--2010 at BDFFP camps <!--# cite eventual BDFFP rainfall package -->.
 The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010.
@@
 This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps.
-Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed [@honakerAmeliaIIProgram2011].
-We produced 10 imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using the averaged potential evapotranspiration from the two EMBRAPA weather stations<!--# why not use mean temperature from weather stations?
--->.
+Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, and seasonality [@honakerAmeliaIIProgram2011].
+We produced 10 imputed datasets which were then each aggregated monthly and used for calculating SPEI using potential evapotranspiration averaged from the two EMBRAPA weather stations.
 Then, the 10 resulting SPEI datasets were averaged to produce a single dataset.
@@
 
-Monthly precipitation data from TRMM was used in combination with potential evapotranspiration data from Xavier et al. [-@xavierDailyGriddedMeteorological2016].
+The Tropical Rainfall Measuring Mission (TRMM) is a satellite product of global precipitation data from 1998-2019 at a resolution of 0.25º x 0.25º <!--# make sentence better -->[@tropicalrainfallmeasuringmissiontrmmTRMMTMPA3B432011].
+Monthly precipitation data from the TRMM 3B43 product was used in combination with potential evapotranspiration data from Xavier et al.
+[-@xavierDailyGriddedMeteorological2016] to calculate SPEI at BDFFP.
 
@@
 
-Additionally, we downloaded SPEI data from the Global SPEI dataset [@begueriaSbegueriaSPEIbaseVersion2017].
+Lastly, we downloaded SPEI data from the Global SPEI dataset [@begueriaSbegueriaSPEIbaseVersion2017].
 These calculations were made using precipitation data from a gridded product based on land stations and evapotranspiration estimated with the Pennman-Montieth method.
@@
 <!-- Annotate worst droughts.\ -->
+
 <!-- Expected: -->
@@
 <!-- -   2004 & 2006: weak ENSO -->
+
 <!-- -   2002 & 2009: moderate ENSO -->
+
 <!-- -   1997: severe ENSO -->
+
 <!-- -   2005 & 2010: severe SST -->
@@
 <!--     -   Resolution 0.25º x 0.25º -->
+
 <!--     -   Includes precipitation and potential evapotranspiration -->
+
 <!--     -   PET calculated by Pennman-Monteith method -->
+
 <!--     -   Source: 3,625 ground-based weather stations across Brazil -->
+
 <!--     -   Date range: 1980--2015 -->
@@
 
-<!-- -   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? --> -->
+<!-- -   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
+
+--\>
 
 <!--     -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7 -->
+
 <!--     -   Resolution: 0.25º x 0.25º -->
+
 <!--     -   Includes: Monthly precip -->
+
 <!--     -   Doesn't include PET, so I guess I'll take that from the ground stations?? -->
+
 <!--     -   Date range: 1998--2019 -->

### 2021-03-10 doc/validate_models.Rmd
got Rmd reports working with knit button, running code chunks, and tar_make()

---
title: "Model validation plots"
author: "Eric R. Scott"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## target knits Rmds in their own session, so load libraries here.
source(here::here("packages.R"))
library(gratia)
```

```{r load-targets}
withr::with_dir(here(),{
  tar_load(c(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha))
  })
```

# Analysis

## Survival

Quantile residuals are more appropriate for binomial models.

### Continuous Forest

```{r}
summary(s_cf)
```

```{r echo=TRUE}
par(mfrow = c(2,2))
#histogram
plot(density(qresid(s_cf)))
#QQ plot
qqnorm(qresid(s_cf)); qqline(qresid(s_cf))
#Fitted vs. residuals
scatter.smooth(predict(s_cf, type = "response"), qresid(s_cf), col = "grey")
```
### Fragments

```{r}
summary(s_1ha)
```

```{r echo=TRUE}
par(mfrow = c(2,2))
#histogram
plot(density(qresid(s_1ha)))
#QQ plot
qqnorm(qresid(s_1ha)); qqline(qresid(s_1ha))
#Fitted vs. residuals
scatter.smooth(predict(s_1ha, type = "response"), qresid(s_1ha), col = "grey")
```

## Growth
### Continuous Forest
```{r}
summary(g_cf)
```

```{r}
gratia::appraise(g_cf)
```

### Fragments

```{r}
summary(g_1ha)
```

```{r}
appraise(g_1ha)
```

## Flowering

Quantile residuals are more appropriate for binomial models.

### Continuous Forest

```{r}
par(mfrow = c(2,2))
#histogram
plot(density(qresid(f_cf)))
#QQ plot
qqnorm(qresid(f_cf)); qqline(qresid(f_cf))
#Fitted vs. residuals
scatter.smooth(predict(f_cf, type = "response"), qresid(f_cf), col = "grey")
```

### Fragments

```{r}
par(mfrow = c(2,2))
#histogram
plot(density(qresid(f_1ha)))
#QQ plot
qqnorm(qresid(f_1ha)); qqline(qresid(f_1ha))
#Fitted vs. residuals
scatter.smooth(predict(f_1ha, type = "response"), qresid(f_1ha), col = "grey")
```

# Reproducibility

<details><summary>Reproducibility receipt</summary>

```{r}
## datetime
Sys.time()

## repository
if(requireNamespace('git2r', quietly = TRUE)) {
  git2r::repository()
} else {
  c(
    system2("git", args = c("log", "--name-status", "-1"), stdout = TRUE),
    system2("git", args = c("remote", "-v"), stdout = TRUE)
  )
}

## session info
sessionInfo()
```

</details>

### 2021-03-10 _targets.R

@@ -30,5 +30,7 @@ tar_plan(
   g_cf = fit_growth(model_data_cf),
   g_1ha = fit_growth(model_data_1ha),
   f_cf = fit_flwr(model_data_cf),
-  f_1ha = fit_flwr(model_data_1ha)
+  f_1ha = fit_flwr(model_data_1ha),
+  tar_render(validate_models, "doc/validate_models.Rmd"),
+  tar_render(model_summary, "doc/model_summary.Rmd")
 )

### 2021-03-10 packages.R

@@ -19,6 +19,7 @@ library(broom)
 library(gratia)
 library(latex2exp)
 library(parallel)
+library(statmod)
 
 conflict_prefer("filter", "dplyr")
 conflict_prefer("lag", "dplyr")

### 2021-03-10 analysis/paper/paper.Rmd
edits based on María's comments
working on discussion

@@
 ---
+ga---
 title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
@@
 1)  Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
+
 2)  What lag times are most important in considering the effects of drought on vital rates?
 [Maybe: are there temporal delays effects in demographic responses to drought, and if so what lag times are most critical?]
+
 3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
-
-<!-- Gentry AH, Emmons LH  1987. Geographical variation in fertility, phenology and composition of the understory of neotropical forests. Biotropica 19:216-217. -->
-
-<!-- Snow DW  1981. Tropical frugivorous birds and their food plants - a world survey. Biotropica 13:1-14. -->
-
-<!-- Nakazono EM, Bruna EM, Mesquita RCG  2004. Experiemental harvesting of the non-timber forest product Ischnosiphon polyphyllus in central Amazonia. Forest Ecology and Management 190:219-225. -->
 
@@
 
-    1.  Delayed effects of environmental drivers on vital rates are rarely included in demographic models, but they may be the rule rather than the exception.
-        [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-        This study is another example of one where delayed effects of climate are important and sometimes better predictors of vital rates than more recent conditions (e.g.
-drought two dry seasons ago and flowering probability).
+    1.  Very few demographic studies have the data required to fit models that include delayed effects.
 
-    2.  Very few demographic studies have the data to fit models that include delayed effects.
+    2.
+This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+        We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
+        Delayed effects of climate on plant vital rates may be the rule, rather than the exception, and distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
+        <!--# maybe the previous sentence goes in methods? -->
 
@@
 
+        1.  Previous work has shown flowering probability to be an important vital rate.
+            Bruna (CITATION) showed that since smaller plants are less likely to flower, a decrease in plant size can result in a significant reduction in flowering.
+            <!--# re-skim this paper and flesh out. -->
+
+    3.  Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
+
+        1.  The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
+
+        2.  Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
+            In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
+            Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H.
+acuminata*.
+
+        3.  More work to be done on flowering responses to drought x light x season interactions.
+
 3.  Buffering
 
-    1.
-We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
-The effects of variation in SPEI were more extreme in fragments for survival and growth.
-We did not see this pattern with flowering probability, likely because flowering is overall much less likely in fragments, probably due to smaller plant size.
-
-MAJOR TAKE-HOME MESSAGES
-
-1.  Survival and growth rates are similar in CF and fragments, but more strongly effected by SPEI in fragments.
-
-2.  Flowering probability is higher in CF (overall and for averaged sized plants), but large plants have higher flowering probability in fragments than in CF.
-
-3.
-Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability (cite @nihadRelationshipStressFlowering2018 )
-
-    1.
-In Manaus, canopy is as green or greener in dry season compared to wet season @guanPhotosyntheticSeasonalityGlobal2015.
-So probably direct rather than indirect effects of canopy openness (need to think about this still)
+    1.  We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
+    2.
+Mean survival and growth rates over the entire decade were similar in continuous forest and 1-ha fragments, but were more strongly affected by variation in SPEI in fragments.
+As climate becomes more variable, hot, and dry, drought frequency and severity is expected to increase in the Amazon (CITE IPCC?).
+Our results indicate that increasing drought frequency and severity will reduce *H.
+acuminata* growth and survival more in fragments compared to continuous forest.
+    3.  We did not see the same pattern with flowering probability, likely because flowering was overall much less likely in fragments where plants were smaller on average.
 

### 2021-03-11 analysis/paper/paper.Rmd
working on outline for discussion

@@
 -->.
-This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999] <!--# also cite "Does forest
-fragmentation cause an increase in forest temperature" -->, which continue to be cleared and fragmented at unprecedented rates.
+This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
@@
 
-<!-- CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html-->
+<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->
 
@@
 All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
-Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
+Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
 Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].
@@
 Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
-First, most studies of plants in fragments have focused on a single history-stage or process ***(CITATIONS)***, making it challenging to draw broader demographic conclusions.
+First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)***, making it challenging to draw broader demographic conclusions.
 Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
@@
 They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
-Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021] <!--# add citations from María.Anderegg et al, Schwalm et al.
-Kannenberg et al.-->.
+Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].
 
@@
 
-2)  What lag times are most important in considering the effects of drought on vital rates?
-[Maybe: are there temporal delays effects in demographic responses to drought, and if so what lag times are most critical?]
+2)  Are there delayed effects of drought on demographic responses, and if so what lag times are most critical?
 
@@
 
-Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].
-<!-- Didham RK, Kapos V, Ewers RM  2012.
-Rethinking the conceptual foundations of habitat fragmentation research.
-Oikos 121:161-170.
---> <!-- Driscoll DA, Banks SC, Barton PS, Lindenmayer DB, Smith AL  2013.
-Conceptual domain of the matrix in fragmented landscapes.
-Trends in Ecology & Evolution 28:605-613.
--->
+<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->
+
+\
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
 
@@
 
-        1.  Previous work has shown flowering probability to be an important vital rate.
-            Bruna (CITATION) showed that since smaller plants are less likely to flower, a decrease in plant size can result in a significant reduction in flowering.
-            <!--# re-skim this paper and flesh out. -->
+        1.
+Previous work has shown flowering probability to be an important vital rate.
+Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
 
@@
         3.  More work to be done on flowering responses to drought x light x season interactions.
+
+    4.  Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
+        Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
 

### 2021-03-15 analysis/paper/paper.Rmd
note to self
reshaping the discussion

@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@bruna2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 Plants that were not found for three consecutive surveys were considered dead.
-There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+There is a seven-fold range in *H.
+acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
@@
 We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
@@
 Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
-<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
+<!--#  I know I'm missing some literature here.
+Add more refs for more droughts.
+Any papers mentioning 2003 wet season drought?
+Also, Feb 2007?-->
 
@@
 \
+1.
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
 
-1.  Consequences for demographers
+2. Further evidence delayed effects may be the rule (emphasis on demographic modeling)
 
-    1.  Very few demographic studies have the data required to fit models that include delayed effects.
+      * This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+      * We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions. 
+      * Delayed effects of climate on plant vital rates may be the rule, rather than the exception. 
+      * distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
+<!--# maybe the previous sentence goes in methods?
+-->
 
-    2.
-This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-        We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
-        Delayed effects of climate on plant vital rates may be the rule, rather than the exception, and distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
-        <!--# maybe the previous sentence goes in methods? -->
+3.
+What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
+[this lays out what these patterns are]
 
-2.  Consequences for biology
+- Growth v Survivorship v Flowering
 
-    1.  What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
+- What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
 
-    2.
-Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
+ - Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
 <!--# from model intercepts -->
 
-        1.
-Previous work has shown flowering probability to be an important vital rate.
+- Previous work has shown flowering probability to be an important vital rate.
 Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
 
-    3.  Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
+- Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
 
-        1.  The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
+- The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
 
-        2.  Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
+4. Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]
+
+- I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year.
+Flowering is a big investment so perhaps flowering in one year means that you don’t flower the following one.
+If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth.
+It will also be easier to explain as you write the paper.
+**Emilio will look into physiological responses to light and H20 in the horticultural and ecological literature, eg.
+Westerband papers, esp if focusing on herbs** Basically anything on how more or less precip influences vital rates or how plants respond to them, esp if linked to changes in canopy openness, litterfall, flowering in ENSO years, Bruna transplant experiments, Heliconia germination in secondary forest, important to emphasize: results are often counter-intuitive.
+
+5. Ecological Consequences of these patterns: an eye towards a drier future
+
+- Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
 In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
@@
 
-        3.  More work to be done on flowering responses to drought x light x season interactions.
+-  More work to be done on flowering responses to drought x light x season interactions.
 
-    4.  Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
-        Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
+- Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
+Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al.
+1998.
 
-3.  Buffering
+6.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]
 
@@
     3.  We did not see the same pattern with flowering probability, likely because flowering was overall much less likely in fragments where plants were smaller on average.
-
-Future directions/Caveats:
-
--   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
-    Smaller/larger plants might be more susceptible to drought.
-    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-
--   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
-While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-They could act in concert to exacerbate or dampen population dynamics."
 
@@
 
-# Conclusion
+# Conclusions & Future Directions
+
+-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]).
+Smaller/larger plants might be more susceptible to drought.
+So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+
+-   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
+While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
+They could act in concert to exacerbate or dampen population dynamics."
+
+-  Very few demographic studies have the data required to fit models that include delayed effects.
+
+-  Looop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
 
@@
 National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
-This is publication No. ---- in the BDFFP Technical Series.
+This is publication No.
+---- in the BDFFP Technical Series.
 
@@
 NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM.
-"
+PPS, which develop and compute the TMPA as a contribution to TRMM. "
 

### 2021-03-15 analysis/paper/paper.Rmd
notes from convo with Emilio

--- 2021-03-11 analysis/paper/paper.Rmd
@@
 ---
-ga---
 title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
@@
   These are the highlights
+
 editor_options:
@@
     wrap: sentence
+    canonical: true
 ---
@@
 Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
-First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)***, making it challenging to draw broader demographic conclusions.
+First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)*** <!--# citations in proposal -->, making it challenging to draw broader demographic conclusions.
 Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
@@
 \
-Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].<!--# flesh out -->
 
@@
 
-        1.  The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
+        1.
+The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
+e.g.
+S.
+Joe Wright trees flowering after ENSO -->.
 
@@
         Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
+
+    5.  cite Esteban et al. 2021 for negative effects of large positive SPEI
 

### 2021-03-15 analysis/paper/paper.Rmd

--- 2021-03-15 analysis/paper/paper.Rmd
@@
 Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
-<!--#  I know I'm missing some literature here.
-Add more refs for more droughts.
-Any papers mentioning 2003 wet season drought?
-Also, Feb 2007?-->
+<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
@@
 
-\
 1.
@@
 
-      * This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-      * We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions. 
-      * Delayed effects of climate on plant vital rates may be the rule, rather than the exception. 
-      * distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
+    -   This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+    -   We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
+    -   Delayed effects of climate on plant vital rates may be the rule, rather than the exception.
+    -   distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
 <!--# maybe the previous sentence goes in methods?
@@
 
-3.
-What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
+3.  What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
     [this lays out what these patterns are]
@@
 -   I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year.
-Flowering is a big investment so perhaps flowering in one year means that you don’t flower the following one.
+    Flowering is a big investment so perhaps flowering in one year means that you don't flower the following one.
     If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth.
@@
 
-5. Ecological Consequences of these patterns: an eye towards a drier future
+-   cite Esteban et al. 2021 for negative effects of large positive SPEI
+
+-   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
+e.g.
+S.
+Joe Wright trees flowering after ENSO -->.
 
@@
 
+5.  Ecological Consequences of these patterns: an eye towards a drier future
+
 -   More work to be done on flowering responses to drought x light x season interactions.
@@
 -   Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
-Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al.
-1998.
+    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
 
@@
 NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM. "
+PPS, which develop and compute the TMPA as a contribution to TRMM.
+"
 

### 2021-03-15 analysis/paper/vital rate models.Rmd
add sentence wrap option to YAML

@@
     number_sections: yes
+editor_options:
+  markdown:
+    wrap: sentence
+    canonical: true
 ---
@@
 
-- ~~Switch measure of plant size to shoots * height~~
+-   ~~Switch measure of plant size to shoots \* height~~
 -   ~~Change lag to 36 months (3 years)~~
@@
 
-I'm going to start fitting these with `bam()` to speed things up.  This sets up the paralellization.  CAUTION: you may need to change these settings for your machine.
+I'm going to start fitting these with `bam()` to speed things up.
+This sets up the paralellization.
+CAUTION: you may need to change these settings for your machine.
 
@@
 # Survival
+
 `surv` is 1 in `year`s when a plant is alive and 0 in the first `year` when the plant is dead.
 These GAMs model the survival of a plant (over the past year) as a function of size in the previous census and SPEI values starting at february of the current year going back 36 months.
-## Fit GAMs
+\#\# Fit GAMs
 
@@
 ### Setup for all crossbasis plots
-Ok, I want to add a lil strip on the bottom that shows dry and wet seasons.  Wet = November-may, dry = june-october.
+
+Ok, I want to add a lil strip on the bottom that shows dry and wet seasons.
+Wet = November-may, dry = june-october.
 
@@
 ### Get plot data
+
 `dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
@@
 ### Difference
+
 Try plotting a raster of the difference between CF and 1-ha
@@
 # Growth
+
 For mature plants that are alive...
@@
 
-AIC for... 
-Gaussian: ~40909.27. looks leptokurtic
-scaled t: ~37284.47.  Waaay better
+AIC for... Gaussian: \~40909.27.
+looks leptokurtic scaled t: \~37284.47.
+Waaay better
 
@@
 ### Combine
+
 combine
+
 ```{r}
@@
 ```
+
 ### Difference
@@
 ```
+
 Overall probability of flowering higher in continuous forest (0.008) compared to 1ha fragments (0.003)
@@
 ```
+
 ```{r paged.print=FALSE}
@@
 ## Crossbasis plot
+
 ### Get plot data
+
 `dist` isn't working yet in `smooth_estimates`, but that's ok because I don't need CI for these plots (yet).
@@
 ### CF
+
 ```{r}
@@
 ```
+
 ### Combine
-combine
-`
+
+combine \`
+
 ```{r}

### 2021-03-15 analysis/supplementary-materials/appendix_SPEI.Rmd
add sentence wrap option to YAML

@@
 csl: "../templates/ecology.csl" # Insert path for the bib-style
+editor_options:
+  markdown:
+    wrap: sentence
+    canonical: true
 ---

### 2021-03-15 doc/paper.Rmd
added Rmd for main paper

---
title: "Untitled Draft"
author: "Report Author"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "../figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(knitr)
library(here)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot
  ))
})
```

### 2021-03-15 analysis/supplementary-materials/size-environment_interaction.Rmd

--- 2021-01-13 analysis/supplementary-materials/size-environment_interaction.Rmd
@@
 bibliography: "`r here::here('analysis', 'paper', 'references.bib')`"
+editor_options:
+  markdown:
+    wrap: sentence
+    canonical: true
 ---
@@
 
+# TODO
+
+-   ~~Test that random effect of year is significant for all models~~
+
 # Purpose
@@
 This allows me to test an interaction term with size to see if responses to environment (maybe drought) are size dependent.
-Inspired by Tredennick et al.
-[-@tredennickSizebyenvironmentInteractionsNeglected2018].
+Inspired by Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018].
 
@@
   )
-demog <-
-  demog %>%
-  mutate(log_shts = log(shts + 0.0001),
-         log_shts_next = log(shts_next + 0.0001))
+# demog <-
+#   demog %>%
+#   filter(ht_prev < 200, ht < 200) %>% 
+#   filter(shts_prev > 0, shts > 0, ht_prev >= 1, ht >= 1) %>% 
+#   mutate(size = shts*ht, size_next = shts_next*ht_next, log_size = log(size), log_size_next = log(size_next))
 ```
 
-Seedlings are likely to have different responses than post-seedlings, so I'll split the data that way. To start with a relatively simple model, I'll only use continuous forest for now.
+Seedlings are likely to have different responses than post-seedlings, so I'll split the data that way.
+To start with a relatively simple model, I'll only use continuous forest for now.
 
@@
 # Function
-Likelihood ratio test for random effects from Tredennick et al.  Tests observed likelihood ratio between model with and without interaction against what would be expected under null.
+
+Likelihood ratio test for random effects from Tredennick et al.
+Tests observed likelihood ratio between model with and without interaction against what would be expected under null.
 
@@
 ```{r}
-growth_y <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_cf)
-growth_sy <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_cf)
-AICtab(growth_y, growth_sy)
+growth_null <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_cf)
+growth_y <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_cf)
+growth_sy <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_cf)
+AICtab(growth_null, growth_y, growth_sy)
 ```
 
+Significant interaction.
+
 ```{r}
+#This takes a very long time and gives results qualitatively similar to comparing AIC
 lrt_growth <- sy_lrt(growth_y, growth_sy, nsim = 500)
 ```
-Upper = 4.539, observed = 45.689
 
@@
 ```{r}
-surv_y <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
-surv_sy <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
-AICtab(surv_y, surv_sy)
+surv_null <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
+surv_y <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
+surv_sy <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
+AICtab(surv_null, surv_y, surv_sy)
 ```
+
+no interaction
 
@@
 ```{r}
-flwr_y <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
-flwr_sy <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_cf)
-AICtab(flwr_y, flwr_sy)
+flwr_null <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_cf)
+flwr_y <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_cf)
+flwr_sy <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_cf)
+AICtab(flwr_null, flwr_y, flwr_sy)
 ```
+
+No interaction
 
@@
 ```{r}
-growth_y_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (1|year), data = demog_post_1ha)
-growth_sy_1ha <- lmer(log_shts_next ~ log_shts + (1|plot) + (log_shts|year), data = demog_post_1ha)
-AICtab(growth_y_1ha, growth_sy_1ha)
+growth_null_1ha <- lmer(log_size_next ~ log_size + (1|plot), data = demog_post_1ha)
+growth_y_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (1|year), data = demog_post_1ha)
+growth_sy_1ha <- lmer(log_size_next ~ log_size + (1|plot) + (log_size|year), data = demog_post_1ha)
+AICtab(growth_null_1ha, growth_y_1ha, growth_sy_1ha)
 ```
+
+significant interaction
 
@@
 ```{r}
-surv_y_1ha <- glmer(surv ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
-surv_sy_1ha <- glmer(surv ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
-AICtab(surv_y_1ha, surv_sy_1ha)
+surv_null_1ha <- glmer(surv ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
+surv_y_1ha <- glmer(surv ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
+surv_sy_1ha <- glmer(surv ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
+AICtab(surv_null_1ha, surv_y_1ha, surv_sy_1ha)
 ```
+
+No interaction or effect of year.
 
@@
 ```{r}
-flwr_y_1ha <- glmer(flwr ~ log_shts + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
-flwr_sy_1ha <- glmer(flwr ~ log_shts + (1|plot) + (log_shts|year), family = binomial, data = demog_post_1ha)
-AICtab(flwr_y_1ha, flwr_sy_1ha)
-```
+flwr_null_1ha <- glmer(flwr ~ log_size + (1|plot), family = binomial, data = demog_post_1ha)
+flwr_y_1ha <- glmer(flwr ~ log_size + (1|plot) + (1|year), family = binomial, data = demog_post_1ha)
+flwr_sy_1ha <- glmer(flwr ~ log_size + (1|plot) + (log_size|year), family = binomial, data = demog_post_1ha)
+AICtab(flwr_null_1ha, flwr_y_1ha, flwr_sy_1ha)
+``
 
-```{r}
+significant interaction?
 
 lrt_flwr_1ha <- sy_lrt(flwr_y_1ha, flwr_sy_1ha, nsim = 500)
-
 ```

### 2021-03-15 doc/paper.Rmd
moved paper over to being built by targets package.

@@
 ---
-title: "Untitled Draft"
-author: "Report Author"
+title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
+author:
+  - Eric R. Scott:
+      email: scottericr@gmail.com
+      institute: [UF_WEC]
+      correspondence: true
+  - María Uriarte:
+      email: mu2126@columbia.edu
+      institute: [Columbia]
+      correspondence: false
+  - Emilio M. Bruna:
+      email: embruna@ufl.edu
+      institute: [UF_WEC, UF_LAS, BDFFP]
+      correspondence: false
+
+institute:
+  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
+  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
+  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: html_document
+output: 
+    bookdown::word_document2:
+      number_sections: no
+      fig_caption: yes
+      reference_docx: "templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
+      - --lua-filter=templates/pagebreak.lua
+linkcolor: blue
 bibliography: references.bib
+csl: "templates/ecology.csl" # Insert path for the bib-style
+abstract: |
+  Text of abstract
+keywords: |
+  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
+highlights: |
+  These are the highlights
+
+editor_options:
+  markdown:
+    wrap: sentence
+    canonical: true
 ---
+
+<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+
+<!-- With the following code you can access and display values from the yml header above. -->
+
+Keywords: `r rmarkdown::metadata$keywords`
+
+Highlights: `r rmarkdown::metadata$highlights`
+
+<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
   tar_load(c(
-    normals_plot
+    normals_plot,
+    eda_plot,
+    s_covar_plot,
+    s_spei_plot,
+    s_spei_diff_plot,
+    g_covar_plot,
+    g_spei_plot,
+    g_spei_diff_plot,
+    f_covar_plot,
+    f_spei_plot,
+    f_spei_diff_plot
   ))
@@
 ```
+
+<!-- The actual document text starts here: -->
+
+<!--# Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->
+
+# Introduction
+
+Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
+Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
+Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others?
+-->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others?
+-->.
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
+-->.
+This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
+For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+Despite this, evidence linking fragmentation-related changes in environmental conditions with population-level demographic responses remains scarce.
+
+<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->
+
+Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017].
+These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
+Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
+Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+
+Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
+First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)*** <!--# citations in proposal -->, making it challenging to draw broader demographic conclusions.
+Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
+These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
+They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
+Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].
+
+We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
+<!--# add a sentence about "famous" amazonian droughts --> Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
+We addressed the following questions:
+
+1)  Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
+
+2)  Are there delayed effects of drought on demographic responses, and if so what lag times are most critical?
+
+3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
+
+# Methods
+
+## Study site
+
+The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure S\_ <!--# map can go in the supplemental materials -->).
+In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
+
+```{r}
+cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+Precipitation (a) shows a marked dry seasons from June through October.
+Temperature (b) varies less throughout the year.
+Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
+Data from Brazilian National Institute of Meterology (INMET)."
+```
+
+```{r normals, fig.cap=cap}
+normals_plot
+```
+
+The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
+The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
+There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
+
+<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->
+
+## Focal species
+
+*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
+While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
+data]*.
+
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
+
+## Demographic data collection
+
+In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
+The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
+This study uses data from 1-ha fragments and Continuous Forest sites.
+
+Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
+We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
+Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
+Plants that were not found for three consecutive surveys were considered dead.
+There is a seven-fold range in *H.
+acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+
+## Weather data
+
+Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
+SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
+Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
+SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H.
+acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
+SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
+<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->
+
+## Statistical Modeling of Vital Rates
+
+<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
+-->
+
+We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
+They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
+We chose 36 months as a maximum lag because prior *H.
+acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
+The general form of the vital rate models was as follows:
+
+$$
+\begin{gather}
+    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
+    \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j 
+\end{gather}
+$$
+
+where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+2017 for details).
+The crossbasis function, $s_2(\cdot)$ can also be written:
+
+$$
+s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
+$$
+
+where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodsimonn.ChapterSmoothers] and Gasparrini et al.
+[-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
+Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
+number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
+Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
+A random effect of plot ID on the intercept is represented by $a_j$, allowing for plot characteristics to influence average vital rates.
+For plant growth the response was plant size (natural log of height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution.
+A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
+The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences).
+When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.
+
+In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
+We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].
+
+Because the `dlnm` package does not currently allow modeling of factor by smooth interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+
+# Results
+
+```{r}
+cap <- "Timeseries of demographic parameters and drought ocurrence.
+Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
+(b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
+c) shows the proportion of plants with 4 or more shoots flowering each census.
+In d), monthly 3-month SPEI is plotted.
+Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.
+Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
+```
+
+```{r eda, fig.cap=cap, fig.height=7, fig.width=10}
+eda_plot
+```
+
+The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
+For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
+The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
+<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
+
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
+
+## Survival
+
+```{r}
+cap <- "Smooth effect of plant size in the previous census on survival modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
+```
+
+```{r surv-size, fig.cap=cap}
+s_covar_plot
+```
+
+Overall survival rate across the entire dataset was the same (0.95) in both CF and 1-ha.
+The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
+The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to wetter than average conditions (SPEI \> 0.5), but with a moderate drought the preceding October.
+
+Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
+Survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:surv-size)).
+Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
+
+```{r}
+cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
+Contour lines correspond to a change of 0.002.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+
+```
+
+```{r surv-spei, fig.cap=cap, fig.height=5, fig.width=7}
+s_spei_plot
+```
+
+Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
+For 1-ha fragments, recent SPEI values (i.e.
+within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
+In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
+However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e.
+not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
+Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
+Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
+However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
+
+```{r}
+cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
+```
+
+```{r surv-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+s_spei_diff_plot
+```
+
+Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
+However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
+
+The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.257).
+
+## Growth
+
+Plants were on average larger in continuous forest compared to 1-ha fragments.
+<!--# This is across the whole timeseries, which I'm not sure is actually that important.
+What summary statistics would be useful?
+-->In continuous forest, plants had an average of 2.9 ± 1.8 shoots (mean ± standard deviation) and were on average 40.6 ± 26.5 cm tall with an average size (shoots $\times$ height) of 150 ± 175.
+In 1-ha fragments, plants had 2.5 ± 1.5 shoots on average and a mean height of 36.2 ± 24.1 cm and a mean size of 112 ± 141.
+The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).
+
+```{r}
+cap <- "Smooth effect of plant size in year t on plant size in year t+1 modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+The dashed red line represents a 1:1 line for reference.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+Model intercepts were added to the fitted values of the smooths."
+```
+
+```{r growth-size, fig.cap=cap}
+g_covar_plot
+```
+
+Size in year t was a significant predictor of size in year t+1 (p \< 0.001).
+The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:growth-size)).
+
+```{r}
+cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function.
+Contour lines correspond to a change of 0.05.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+```
+
+```{r growth-spei, fig.cap=cap, fig.height=5, fig.width=7}
+g_spei_plot
+```
+
+Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
+The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
+
+```{r}
+cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
+```
+
+```{r growth-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+g_spei_diff_plot
+```
+
+The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
+However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
+This is due to reduced growth in fragments under wet conditions.
+
+The random effect of plot is significant in both habitat models (p \< 0.001).
+
+## Flowering
+
+The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
+In fact, the proportion of flowering plants was higher in continuous forest in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
+The most striking difference between habitat types in flowering proportion coincided with a severe drought in 2003 when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest and 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
+The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
+<!--# not sure how necessary this is -->
+
+```{r}
+cap <- "Smooth effect of plant size in the previous census on flowering probability modeled by a cubic regression spline.
+95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
+```
+
+```{r flwr-size, fig.cap=cap}
+f_covar_plot
+```
+
+Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
+Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
+Flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest.
+However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
+<!--# idea: plot vertical lines representing mean or median size in each habitat -->
+
+```{r}
+cap <- "Smooth effect of lagged SPEI on flowering probability.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
+Contour lines correspond to a change of 0.001.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+```
+
+```{r flwr-spei, fig.cap=cap, fig.height=5, fig.width=7}
+f_spei_plot
+```
+
+Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
+Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
+The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
+In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, the dry season in the previous year).
+Drought at lags 7--13 reduced flowering probability.
+The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
+That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
+
+```{r}
+cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
+```
+
+```{r flwr-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+f_spei_diff_plot
+```
+
+Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
+The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.
+
+The random effect of plot also had a significant effect in both habitats (p \< 0.001).
+
+# Discussion
+
+<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->
+
+1.
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+
+2.  Further evidence delayed effects may be the rule (emphasis on demographic modeling)
+
+    -   This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+    -   We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
+    -   Delayed effects of climate on plant vital rates may be the rule, rather than the exception.
+    -   distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
+<!--# maybe the previous sentence goes in methods?
+-->
+
+3.  What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
+    [this lays out what these patterns are]
+
+-   Growth v Survivorship v Flowering
+
+-   What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
+
+-   Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
+    <!--# from model intercepts -->
+
+-   Previous work has shown flowering probability to be an important vital rate.
+    Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
+
+-   Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
+
+-   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
+
+4.  Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]
+
+-   I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year.
+    Flowering is a big investment so perhaps flowering in one year means that you don't flower the following one.
+    If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth.
+    It will also be easier to explain as you write the paper.
+    **Emilio will look into physiological responses to light and H20 in the horticultural and ecological literature, eg.
+Westerband papers, esp if focusing on herbs** Basically anything on how more or less precip influences vital rates or how plants respond to them, esp if linked to changes in canopy openness, litterfall, flowering in ENSO years, Bruna transplant experiments, Heliconia germination in secondary forest, important to emphasize: results are often counter-intuitive.
+
+-   cite Esteban et al. 2021 for negative effects of large positive SPEI
+
+-   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
+e.g.
+S.
+Joe Wright trees flowering after ENSO -->.
+
+-   Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
+    In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
+    Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H.
+acuminata*.
+
+5.  Ecological Consequences of these patterns: an eye towards a drier future
+
+-   More work to be done on flowering responses to drought x light x season interactions.
+
+-   Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
+    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
+
+6.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]
+
+    1.  We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
+    2.
+Mean survival and growth rates over the entire decade were similar in continuous forest and 1-ha fragments, but were more strongly affected by variation in SPEI in fragments.
+As climate becomes more variable, hot, and dry, drought frequency and severity is expected to increase in the Amazon (CITE IPCC?).
+Our results indicate that increasing drought frequency and severity will reduce *H.
+acuminata* growth and survival more in fragments compared to continuous forest.
+    3.  We did not see the same pattern with flowering probability, likely because flowering was overall much less likely in fragments where plants were smaller on average.
+
+Caveats:
+
+<!--# need to edit.  Moved from methods. -->
+
+One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
+This limitation meant that we were unable to model a potential size by environment (i.e.
+drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
+To check for a size by environment interaction, we followed the methods of Tredennick et al.
+[-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package [@batesFittingLinearMixedeffects2015].
+Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
+For each vital rate, we essentially replaced the crossbasis function in Eq 1.
+with a random effect of year on either the intercept or both the intercept and relationship with plant size.
+We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].
+
+# Conclusions & Future Directions
+
+-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
+    Smaller/larger plants might be more susceptible to drought.
+    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+
+-   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
+While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
+They could act in concert to exacerbate or dampen population dynamics."
+
+-   Very few demographic studies have the data required to fit models that include delayed effects.
+
+-   Looop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
+
+# Acknowledgments
+
+We thank ----, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
+We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
+Financial support was provided by the U.S.
+National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
+This is publication No. ---- in the BDFFP Technical Series.
+
+If TRMM data is used, they suggest: "The TMPA data were provided by the\
+NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
+PPS, which develop and compute the TMPA as a contribution to TRMM.
+"
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2021-03-15 R/normals.R

@@ -17,7 +17,7 @@ normals_data <- function() {
 }
 
 
-plot_normals <- function(normals, save = TRUE) {
+plot_normals <- function(normals) {
   precip <-
     ggplot(normals, aes(x = as.factor(month), y = precip)) + 
     geom_col(fill = "darkblue") +
@@ -36,9 +36,5 @@ plot_normals <- function(normals, save = TRUE) {
   p <- 
     (precip + theme(axis.title.x = element_blank(), axis.text.x = element_blank()))/
     (temp) + plot_annotation(tag_levels = "a", tag_suffix = ")")
-  
-  if(save == TRUE) {
-    ggsave(here("analysis", "figures", "manaus_normals.png"), p)
-  }
   return(p)
 }

### 2021-03-15 R/plot_eda.R

@@ -167,13 +167,11 @@ plot_eda_flwr <- function(demog_post, date_lims) {
   flowering
 }
 
-plot_eda_combine <- function(..., save = TRUE) {
+plot_eda_combine <- function(...) {
   p <- wrap_plots(..., ncol = 1) +
     plot_layout(guides = "collect") &
     plot_annotation(tag_levels = "a", tag_suffix = ")") &
     theme(plot.margin = margin(2,1,1,1)) &
     guides(col = guide_legend(title = "Habitat"))
-  if(save == TRUE) {
-    ggsave(here("analysis", "figures", "eda.png"), height = 7, width = 10)
-  }
+  return(p)
 }

### 2021-03-15 _targets.R

@@ -56,20 +56,20 @@ tar_plan(
   ## Survival
   s_covar_plot = plot_covar_smooth(frag_model = s_1ha, cf_model = s_cf, covar = "log_size_prev") + 
                   labs(x = TeX("$log(size_t)$"), y = "P(survival)"),
-  s_cb_plot = plot_cb_2panel(s_cf, s_1ha, binwidth = 0.002, response_lab = "P(survival)"),
-  s_cb_diff_plot = plot_cb_diff(s_cf, s_1ha, binwidth = 0.002, response_lab = "∆P(survival) (CF – 1ha)"),
+  s_spei_plot = plot_cb_2panel(s_cf, s_1ha, binwidth = 0.002, response_lab = "P(survival)"),
+  s_spei_diff_plot = plot_cb_diff(s_cf, s_1ha, binwidth = 0.002, response_lab = "∆P(survival) (CF – 1ha)"),
 
   ## Growth
   g_covar_plot = plot_covar_smooth(frag_model = g_1ha, cf_model = g_cf, covar = "log_size_prev") +
                   labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})")),
-  g_cb_plot = plot_cb_2panel(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$log(size_{t+1})$")),
-  g_cb_diff_plot = plot_cb_diff(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$\\Delta log(size_{t+1})$ (CF – 1ha)")),
+  g_spei_plot = plot_cb_2panel(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$log(size_{t+1})$")),
+  g_spei_diff_plot = plot_cb_diff(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$\\Delta log(size_{t+1})$ (CF – 1ha)")),
 
   ## Flowering
   f_covar_plot = plot_covar_smooth(frag_model = f_1ha, cf_model = f_cf, covar = "log_size_prev") +
                   labs(x = TeX("$log(size_t)$"), y = "P(flowering)"),
-  f_cb_plot = plot_cb_2panel(f_cf, f_1ha, binwidth = 0.001, response_lab = "P(flowering)"),
[…]

### 2021-03-15 doc/validate_data.Rmd
add data validation with pointblank

---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(pointblank)
library(here)
# Use `validate_rmd()` here to set options for the
# pointblank validation workflow within R Markdown documents
```

```{r}
library(targets)
withr::with_dir(here(), {
  tar_load(c(
    xa_done,
    demog_done
  ))
})
```

# Demographic data

`pointblank` validation on tidied demographic data

`demog_done`: Tidied dataset on *Heliconia acuminata* demography.
More details [here](https://github.com/embruna/HeliconiaDataPaper).

-   `ranch` (factor): The name of the ranch the plot is in
-   `bdffp_reserve_no` (factor): The officall BDFFP reserve number
-   `plot` (factor): A plot ID
-   `habitat` (factor): "cf" = continuous forest, "1-ha" = one hectare fragment
-   `ha_id_number` (factor): unique plant identifier number
-   `year` (integer): survey year
-   `ht` (numeric): height in survey year (cm)
-   `ht_prev` (numeric): height in previous year (cm)
-   `shts` (integer): number of shoots in survey year
-   `shts_prev` (integer): number of shoots in previous year
-   `size` (numeric): shoots \* height
-   `size_prev` (numeric): shoots \* height in previous year
-   `log_size` (numeric): log(shoots \* height)
-   `log_size_prev` (numeric): log(shoots \* height) in previous year
-   `infl_num` (integer): number of inflorescences (if plant flowered)
-   `flwr` (integer): binary. Did plant flower in survey year?
-   `surv` (integer): binary. Did plant survive?
-   `code_notes` (character): possible values include: "plant missing" meaning the plant could not be found in this survey; "dead" meaning the plant is confirmed dead in this survey; "ULY" meaning\_\_\_\_\_\_\_\_; and "tag missing" meaning the plant was missing a tag and given a new one (this does not affect the `ha_id_number` as these take into account plants given new tags)
-   `code2` (character): possible value is "initial.tag.yr" indicating the beginning of data for that `ha_id_number`. Could be the first year of the survey, or could represent recruitment into the plot.

```{r}
demog_done
```

```{r processed}
create_agent(
  demog_done,
  actions = action_levels(warn_at = 1, stop_at = 0.2)
) %>% 
  col_is_character(vars(row, code_notes, code2)) %>% 
  col_is_factor(vars(ranch, bdffp_reserve_no, plot, habitat, ha_id_number)) %>% 
  col_is_numeric(vars(ht, ht_prev, size, size_prev, log_size, log_size_prev)) %>%
  col_is_integer(vars(year, shts, shts_prev, infl_num, surv, flwr)) %>% 
  col_vals_in_set(vars(surv, flwr), c(0,1, NA)) %>% 
  col_vals_in_set(vars(column), 1:10) %>% 
  col_vals_in_set(vars(row), LETTERS[1:10]) %>% 
  col_vals_in_set(vars(habitat), c("1-ha", "CF", "10-ha")) %>% 
  col_vals_between(vars(shts, shts_prev), 1, 20, na_pass = TRUE) %>% 
  col_vals_between(vars(ht, ht_prev), 0, 200, na_pass = TRUE) %>% 
  rows_distinct() %>% 
  interrogate()
```

### 2021-03-15 R/demog.R

@@ -51,14 +51,15 @@ add_size <- function(data) {
 tidy_demog <- function(data) {
   data %>% 
     #add binary column for flowering
-    mutate(infl = ifelse(is.na(infl) & (!is.na(shts) | !is.na(ht)), 0, infl)) %>% 
-    mutate(flwr = ifelse(infl > 0, 1, 0), .after = infl) %>% 
+    mutate(infl = if_else(is.na(infl) & (!is.na(shts) | !is.na(ht)), 0L, infl)) %>% 
+    mutate(flwr = if_else(infl > 0, 1L, 0L), .after = infl) %>% 
     #create factors
     mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
     mutate(year_fac = as.factor(year)) %>% 
     # arrange columns
     select(ranch, bdffp_reserve_no, plot, habitat, #site level
-           row, column, x_09, y_09, ha_id_number, tag_number, #plot level
+           # row, column, x_09, y_09, 
+           ha_id_number, #plot level
            year,
            ht, ht_prev, shts, shts_prev, size, size_prev, log_size, log_size_prev, infl_num = infl, flwr, surv, code_notes, code2)
 }

### 2021-03-15 _targets.R

@@ -28,6 +28,9 @@ tar_plan(
   model_data_cf = filter(model_data, habitat == "CF"),
   model_data_1ha = filter(model_data, habitat == "1-ha"),
   
+  # Data validation
+  tar_render(validate_data, "doc/validate_data.Rmd"),
+  
   # Fit demographic models
   s_cf = fit_surv(model_data_cf),
   s_1ha = fit_surv(model_data_1ha),

### 2021-03-16 doc/SPEI_appendix.Rmd
port SPEI appendix to targets package style

--- 2021-03-09 analysis/supplementary-materials/appendix_SPEI.Rmd
@@
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      reference_docx: "templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
+      - --lua-filter=templates/pagebreak.lua
 bibliography: references.bib
-csl: "../templates/ecology.csl" # Insert path for the bib-style
+csl: "templates/ecology.csl" # Insert path for the bib-style
+editor_options:
+  markdown:
+    wrap: sentence
+    canonical: true
 ---
 
-```{r setup, include=FALSE, echo=FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(
@@
   dpi = 300)
-library(tidyverse)
-library(here)
-library(conflicted)
-library(SPEI)
-library(tsibble)
-library(lubridate)
-library(glue)
-library(colorspace)
-library(latex2exp)
-library(janitor)
-library(Amelia)
-
-library(HeliconiaDemography)
-
-conflict_prefer("lag", "dplyr")
-conflict_prefer("filter", "dplyr")
-conflict_prefer("select", "dplyr")
+source(here::here("packages.R"))
 ```
@@
 ```{r embrapa-spei}
-rpde <-
-  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv")) %>% 
-  select(date = Data,
-         precip = Precipitacao,
-         eto = Evatotranspiracao,
-  )
-manaus <-
-  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Manaus_1980-01-01_2016-12-31.csv"))%>% 
-  select(date = Data,
-         precip = Precipitacao,
-         eto = Evatotranspiracao,
-  )
-
-embrapa_mon <-
-  bind_rows("rpde" = rpde, "manaus" = manaus, .id = "station") %>% 
-  mutate(yearmonth = yearmonth(date), .after = date) %>% 
-  group_by(yearmonth, station) %>% 
-  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
-  as_tsibble(index = yearmonth, key = station)
-
-embrapa_spei <-
-  embrapa_mon %>% 
-  mutate(cb = precip - eto) %>% 
-  group_by(station) %>% 
-  group_nest() %>% 
-  mutate(st_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
-  unnest(c(st_spei, data)) %>% 
-  select(-cb)
-
-embrapa_wide <-
-  embrapa_spei %>% 
-  select(yearmonth, station, st_spei) %>% 
-  pivot_wider(names_from = station,
-              values_from = st_spei,
-              names_prefix = "spei_") %>% 
-  rowwise() %>% 
-  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
-  ungroup()
+withr::with_dir(here(),{
+  tar_load(embrapa_wide)
+  tar_load(embrapa_mon)
+})
 ```
@@
 ```{r}
-# Read in data 
-#Change to something like this once repo is public
-# bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv")
-bdffp_daily <- read_csv(here("analysis", "data", "raw_data", "daily_precip.csv"))
-# Prep data for imputation ------------------------------------------------
-
-# remove accumulations
-bdffp2 <-
-  bdffp_daily %>% 
-  filter(!flag %in% c("A", "U"))
-
-# spread and add rows for missing dates
-bdffp_wide <- 
-  bdffp2 %>% 
-  #complete all the dates
-  as_tsibble(key = site, index = date) %>% 
-  fill_gaps() %>% 
-  select(date, site, precip) %>% 
-  as_tibble() %>% 
-  pivot_wider(names_from = site, values_from = precip) %>% 
-  clean_names()
-
-# Cluster sites by ranch to reduce missingness
-bdffp_wide2 <-
-  bdffp_wide %>% 
-  rowwise() %>% 
-  mutate(colosso_clust = mean(c(colosso, florestal, cabo_frio, gaviao),
-                              na.rm = TRUE),
-         km_clust = mean(c(km37, km41),
-                         na.rm = TRUE)) %>% 
-  mutate(across(colosso_clust:km_clust, ~ifelse(is.nan(.x), NA, .x))) %>% 
-  select(-cabo_frio, -colosso, -florestal, -gaviao, -km37, -km41)
-
-# add additional representations of time for Amelia.
-full_wide <- 
-  bdffp_wide2 %>% 
-  mutate(
-    year = year(date),
-    doy = yday(date),
-    yearmonth = yearmonth(date),
-    .before = everything()
-  )
-```
-
-```{r imputation, cache=TRUE}
-# Impute missing data -----------------------------------------------------
-# round(runif(1, 1, 1000))
-set.seed(937)
-all_cols <- colnames(select(full_wide, -year, -doy, -date))
-
-# histograms <-
-#   map(all_cols, ~{
-#     ggplot(full_wide, aes(full_wide[[.x]])) + geom_histogram() + labs(title = .x)
-#   })
-# histograms
-
-# variables with log-normal-ish distributions
-
-log_cols <-
-  all_cols[str_detect(all_cols, "precip") |
-             all_cols %in% c("km_clust", "colosso_clust", "porto_alegre", "dimona")]
-
-imp <- 
-  amelia(
-    as.data.frame(full_wide),
-    p2s = 0,
-    m = 10,
-    ts = "doy",
-    cs = "year",
-    intercs = TRUE,
-    polytime = 3,
-    logs = log_cols,
-    idvars = c("date"),
-    parallel = "multicore",
-    ncpus = detectCores() - 1,
-    empri = .01 * nrow(full_wide) #ridge penalty because of high degree of missingness
-  )
-# imp$imputations[[1]]
-```
-
-```{r}
-# Prep eto data for SPEI calculation
-# Aggregate by month
-imp_mon <-
-  map(imp$imputations, ~{
-    .x %>% 
-      rename_with(~paste0(., ".precip"), c(dimona, porto_alegre, colosso_clust, km_clust)) %>% 
-      select(date, ends_with(".precip")) %>% 
-      mutate(yearmonth = tsibble::yearmonth(date)) %>% 
-      group_by(yearmonth) %>% 
-      summarize(across(-date, ~sum(.x, na.rm = TRUE))) %>% 
-      #remove first and last month, as they aren't complete
-      slice(-1, -nrow(.))
+withr::with_dir(here::here(), {
+  tar_load(bdffp_spei)
 })
-# Add mean ETO from EMBRAPA data
-imp_mon <-
-  map(imp_mon, ~ {
-    left_join(.x, embrapa_mon %>% summarize(eto = mean(eto)), by = "yearmonth")
-  }) %>% 
-  map(., ~{
-    mutate(.x, across(ends_with(".precip"),
-                      ~ .x - eto,
-                      .names = "{str_remove(col, '.precip')}.cb"))
-  })
-# imp_mon$imp1
-```
-
-```{r}
-# Calculate SPEI3 (i.e. 3-month window SPEI)
-
-imp_spei <-
-  map(imp_mon, ~{
-  .x %>% 
-    mutate(across(ends_with(".cb"), ~as.numeric(SPEI::spei(.x, scale = 3)$fitted),
-                  .names = "{str_remove(col, '.cb')}.spei")) %>%
-      select(yearmonth, ends_with(c(".precip", ".spi", ".spei")))
-}) 
-# imp_spei$imp1
-# Combine multiply imputed SPI and SPEI results by taking mean
-bdffp_spei <- 
-  imp_spei %>%
-  bind_rows(.id = "imp") %>% 
-  #replace any infinite values with NAs
-  mutate(across(ends_with(".spei"), ~ ifelse(is.infinite(.x), NA, .x))) %>% 
-  group_by(yearmonth) %>% 
-  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
-  #replace NaN's with NAs
-  mutate(across(ends_with(".spei"), ~ ifelse(is.nan(.x), NA, .x))) %>% 
-  rowwise() %>% 
-  mutate(bdffp_spei = mean(c_across(ends_with(".spei")))) %>% 
-  ungroup()
 ```
@@
 ```{r xa-spei}
-xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
-#aggregate monthly, then calculate climate water balance.
-xa_mon <-
-  xa %>% 
-  unite(latlon, lat, lon, sep = ", ") %>% 
-  mutate(yearmonth = yearmonth(date), .after = date) %>% 
-  group_by(yearmonth, latlon) %>% 
-  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
-  as_tsibble(index = yearmonth, key = latlon)
-
-xa_spei <-
-  xa_mon %>% 
-  mutate(cb = precip - eto) %>% 
-  group_by(latlon) %>%
-  group_nest() %>% 
-  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
-  unnest(c(xa_spei, data)) %>% 
-  select(-cb)
+withr::with_dir(here::here(), {
+  tar_load(xa_spei)
+})
 ```
@@
 ```{r trmm-spei}
-trmm <- read_csv(here("analysis", "data", "raw_data", "trmm.csv"))
-trmm2 <-
-  trmm %>% 
-  unite(latlon, lat, long, sep = ", ") %>% 
-  mutate(yearmonth = yearmonth(date), .after = date)
-# fortunately xa_mon has same latlon, so can join eto
-
-unique(xa_mon$latlon) %in% unique(trmm2$latlon)
-
-trmm_ts <- 
-  left_join(trmm2, xa_mon %>% select(yearmonth, latlon, eto), by = c("yearmonth", "latlon")) %>% 
-  #I guess I only need grid cells that are also in the Xavier et al. dataset?
-  filter(!is.na(eto)) %>% 
-  as_tsibble(index = yearmonth, key = latlon)
-
-trmm_spei <- 
-  trmm_ts %>% 
-  mutate(cb = precip - eto) %>% 
-  group_by(latlon) %>% 
-  group_nest() %>% 
-  mutate(trmm_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
-  unnest(c(trmm_spei, data)) %>% 
-  select(-cb)
+withr::with_dir(here::here(),{
+  tar_load(trmm_spei)
+})
 ```
@@
 ```{r global-spei}
-gspei <-
-  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
-             delim = ";") %>% 
-  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
-  mutate(yearmonth = yearmonth(date), date = NULL)
+withr::with_dir(here::here(), {
+  tar_load(gspei)
+})
 ```
@@
   xa_spei %>% 
+  as_tibble() %>% 
   group_by(yearmonth) %>% 
   #Summarize across grid cells
-  summarize(xa_spei = mean(xa_spei))
+  summarize(xa_spei = mean(spei))
 
@@
   full_join(bdffp_spei %>% select(yearmonth, bdffp_spei)) %>% 
-  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
+  pivot_longer(c(-yearmonth), names_to = "dataset", values_to = "spei") %>% 
   filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
@@
 ```{r spei-table, fig.cap=cap}
-knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
+# knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
+table
 ```

### 2021-03-16 R/embrapa.R

@@ -1,0 +1,41 @@
+tidy_embrapa <- function(rpde, manaus) {
+ rpde <- 
+   rpde %>% 
+    select(date = Data,
+           precip = Precipitacao,
+           eto = Evatotranspiracao,
+    )
+ manaus <-
+   manaus %>% 
+   select(date = Data,
+          precip = Precipitacao,
+          eto = Evatotranspiracao,
+   )
+
+ bind_rows("rpde" = rpde, "manaus" = manaus, .id = "station") %>% 
+   mutate(yearmonth = yearmonth(date), .after = date) %>% 
+   group_by(yearmonth, station) %>% 
+   summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
+   as_tsibble(index = yearmonth, key = station)
+}
+
+calc_spei_embrapa <- function(embrapa_mon) {
+  embrapa_spei <-
+    embrapa_mon %>% 
+    #calculate climate balance
+    mutate(cb = precip - eto) %>% 
+    group_by(station) %>% 
+    group_nest() %>% 
+    mutate(st_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
+    unnest(c(st_spei, data)) %>% 
+    select(-cb) 
+  
+  embrapa_spei %>% 
+    select(yearmonth, station, st_spei) %>% 
+    pivot_wider(names_from = station,
+                values_from = st_spei,
+                names_prefix = "spei_") %>% 
+    rowwise() %>% 
+    mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
+    ungroup()
+}

### 2021-03-16 R/gspei.R

@@ -1,0 +1,5 @@
+read_tidy_gspei <- function(path) {
+  read_delim(path, delim = ";") %>% 
+    rename(date = "days since 1900-1-1", g_spei = spei) %>% 
+    mutate(yearmonth = yearmonth(date), date = NULL)
+}

### 2021-03-16 R/trmm.R

@@ -1,0 +1,26 @@
+read_tidy_trmm <- function(path) {
+  read_csv(path) %>% 
+    unite(latlon, lat, long, sep = "_") %>% 
+    mutate(yearmonth = yearmonth(date), .after = date)
+}
+
+calc_spei_trmm <- function(trmm, xa_spei) {
+  # # fortunately xavier et al data has same resolution and overlap in latlon, so can join eto
+  
+  # unique(xa_spei$latlon) %in% unique(trmm2$latlon)
+  
+  trmm_ts <- 
+    left_join(trmm, xa_spei %>% select(yearmonth, latlon, eto), by = c("yearmonth", "latlon")) %>% 
+    #I guess I only need grid cells that are also in the Xavier et al. dataset?
+    filter(!is.na(eto)) %>% 
+    as_tsibble(index = yearmonth, key = latlon)
+  
+  #calculate SPEI
+  trmm_ts %>% 
+    mutate(cb = precip - eto) %>% 
+    group_by(latlon) %>% 
+    group_nest() %>% 
+    mutate(trmm_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
+    unnest(c(trmm_spei, data)) %>% 
+    select(-cb)
+}

### 2021-03-16 _targets.R

@@ -16,7 +16,35 @@ tar_plan(
   xa_raw = read_csv(xa_file),
   xa_spei = calc_spei_xa(xa_raw),
   xa_lag = lag_spei(xa_spei, maxlag),
+
   
+  
+  # SPEI supplemental
+  tar_target(rpde_file,
+             here("data", "supplemental", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv"),
+             format = "file"),
+  tar_target(manaus_file,
+             here("data", "supplemental", "Estacao_Manaus_1980-01-01_2016-12-31.csv"),
+             format = "file"),
+  rpde = read_csv(rpde_file),
+  manaus = read_csv(manaus_file),
+  embrapa_mon = tidy_embrapa(rpde,  manaus),
+  embrapa_wide = calc_spei_embrapa(embrapa_mon),
+  tar_target(trmm_file, here("data", "supplemental", "trmm.csv"), format = "file"),
+  trmm = read_tidy_trmm(trmm_file),
+  tar_render(SPEI_appendix, "doc/SPEI_appendix.Rmd"),
+  tar_target(gspei_file, 
+             here("data", "supplemental", "global_spei_-59.75_-2.25.csv"),
+             format = "file"),
+  gspei = read_tidy_gspei(gspei_file),
+  trmm_spei  = calc_spei_trmm(trmm, xa_spei),
+  tar_target(bdffp_file,
+             here("data", "supplemental", "daily_precip.csv"),
+             format = "file"),
+  bdffp_daily = read_csv(bdffp_file),
+  bdffp_full = prep_bdffp(bdffp_daily),
+  bdffp_imputations = impute_bdffp(bdffp_full),
+  bdffp_spei = calc_spei_bdffp(bdffp_imputations, embrapa_mon),
   # Prep demographic data
   tar_target(demog_file, here("data", "Ha_survey_with_Zombies.csv"), format = "file"),
   demog_raw = read_fix_demog(demog_file),

### 2021-03-16 analysis/wrangling/validation.R

@@ -1,26 +1,0 @@
-library(here)
-library(tidyverse)
-library(pointblank)
-
-ha <-
-  read_csv(
-    here("analysis", "data", "derived_data", "ha_survey.csv"),
-    col_names = TRUE,
-    cols(plot = col_character(),
-         bdffp_reserve_no = col_character())
-  )
-
-ha %>% 
-  create_agent(
-    actions = action_levels(warn_at = 0.1, stop_at = 0.2)
-  ) %>% 
-  col_is_character(vars(ranch, bdffp_reserve_no, plot, habitat, row, code_notes, code2)) %>% 
-  col_is_numeric(vars(ha_id_number, tag_number, year, ht, ht_next, shts, shts_next, infl_num, surv, flwr)) %>% 
-  col_vals_in_set(vars(surv, flwr), c(0,1, NA)) %>% 
-  col_vals_in_set(vars(column), 1:10) %>% 
-  col_vals_in_set(vars(row), LETTERS[1:10]) %>% 
-  col_vals_in_set(vars(habitat), c("1-ha", "CF", "10-ha")) %>% 
-  col_vals_between(vars(shts, shts_next), 1, 20, na_pass = TRUE) %>% 
-  col_vals_between(vars(ht, ht_next), 0, 200, na_pass = TRUE) %>% 
-  rows_distinct() %>% 
-  interrogate()

### 2021-03-16 packages.R

@@ -20,7 +20,11 @@ library(gratia)
 library(latex2exp)
 library(parallel)
 library(statmod)
+library(rmarkdown)
+library(colorspace)
+library(Amelia)
 
+conflict_prefer("select", "dplyr")
 conflict_prefer("filter", "dplyr")
 conflict_prefer("lag", "dplyr")
-library(rmarkdown)
+conflict_prefer("slice", "dplyr")

### 2021-03-16 README.Rmd
remove non-applicable info from README, now that this is no longer a package.

--- 2020-11-09 README.Rmd
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) 
-[![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/run-tests/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) <!-- [![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/R-CMD-check/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions) -->
 
@@
 
-> Authors, (YYYY). *`r Title`*. Name of journal/book <https://doi.org/xxx/xxx>
+> Authors, (YYYY).
+> *`r Title`*.
+> Name of journal/book <https://doi.org/xxx/xxx>
 
-Our pre-print is online here:
+<!-- Our pre-print is online here: -->
 
-> Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). *Compendium of R code and data for `r Title`*. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
-
-## Contents
-
-The **R** directory contains functions used in the analysis.
-
-The **analysis** directory contains:
-
--   [:file\_folder: wrangling](/analysis/wrangling): .R and .Rmd files with code related to data wrangling (i.e. converting raw data into cleaned data)
--   [:file\_folder: paper](/analysis/paper): R Markdown source document for manuscript.
-Includes code to reproduce the figures and tables generated by the analysis.
-It also has a rendered version, `paper.docx`, suitable for reading (the code is replaced by figures and tables in this file)
--   [:file\_folder: data](/analysis/data): Data used in the analysis.
--   [:file\_folder: figures](/analysis/figures): Plots and other illustrations
--   [:file\_folder: supplementary-materials](/analysis/supplementary-materials): Supplementary materials including notes and other documents prepared and collected during the analysis.
+> Authors, (`r format(Sys.Date(), "%Y")`).
+> *Compendium of R code and data for `r Title`*.
+> Accessed `r format(Sys.Date(), "%d %b %Y")`.
+> Online at <https://doi.org/xxx/xxx>
 
@@
 
-You can download the compendium as a zip from from this URL: [master.zip](/archive/master.zip). After unzipping:
+You can download the compendium as a zip from from this URL: [master.zip](/archive/master.zip).
+After unzipping:
 
 -   open the `.Rproj` file in RStudio
--   run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
--   finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
+-   something about `renv`
+-   something about `targets::tar_make()` and `targets::tar_visnetwork()` <!-- -   run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
+--> <!-- -   finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console -->
 
@@
 
-**Text and figures :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
+<!-- **Text and figures :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/) -->
 
-**Code :** See the [DESCRIPTION](DESCRIPTION) file
+<!-- <!-- **Code :** See the [DESCRIPTION](DESCRIPTION) file -->
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+<!-- **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse -->

### 2021-03-16 README.Rmd
change name of action and add badge to README

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) <!-- [![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/R-CMD-check/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions) -->
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) [![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml)
 

### 2021-03-16 README.Rmd
remove docker file and binder badge (doesn't work with private repos anyway)

--- 2021-03-16 README.Rmd
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) <!-- [![R build status](https://github.com/BrunaLab/HeliconiaDemography/workflows/R-CMD-check/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions) -->
+[![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml)
 

### 2021-03-16 doc/SPEI_appendix.Rmd
move figure path

@@
   comment = "#>",
-  fig.path = "",
+  fig.path = "figures/",
   dpi = 300)
@@
   pivot_wider(names_from = dataset, values_from = c(spei, month)) %>% 
-  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), .after = "year") %>% 
+  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), 
+             .after = "year") %>% 
   mutate(`Global SPEI` = glue("{round(spei_g, 2)} ({month_g})"),
@@
   tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM, BDFFP))
-# table
-# gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
 #TODO: add citations to this table as footnotes on `cause` or a new column.
@@
     date_breaks = "1 year",
-    # date_minor_breaks = "1 month"
   ) +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# p
+
 annotate_spei(p)
-# ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
 ```

### 2021-03-16 doc/paper.Rmd
move figure path

@@
 	echo = FALSE,
-	fig.path = "../figures/",
+	fig.path = "figures/",
 	message = FALSE,
@@
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].

### 2021-03-22 doc/paper.Rmd
work on discussion outline

@@
 
+<!--# consider boxplots or median ± IQR for panel a) -->
+
 The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
@@
 
+<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->
+
 Size in year t was a significant predictor of size in year t+1 (p \< 0.001).
@@
 The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
-In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, the dry season in the previous year).
+In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, two dry seasons ago).
 Drought at lags 7--13 reduced flowering probability.
@@
 
-    -   This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-    -   We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
-    -   Delayed effects of climate on plant vital rates may be the rule, rather than the exception.
-    -   distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
-<!--# maybe the previous sentence goes in methods?
--->
+    This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+    We found evidence for delayed effects of drought which, in our models, are sometimes stronger predictors of vital rates than more recent conditions.
+    Distributed lag non-linear models offer an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
 
-3.  What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
+3.
+What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat ~~and life-history stage~~ <!--# we didn't really do this.
+This implies a size * SPEI interaction -->.
     [this lays out what these patterns are]
 
--   Growth v Survivorship v Flowering
+    -   Survival at mean SPEI was similar in fragments and continuous forest for all plant sizes.
+        Plants were smaller in forest fragments, but grew slightly faster than plants in continuous forest, closing the size gap over the decade of data.
+
+    -   However, the effect of varying SPEI on survival and growth was much greater in fragments.
+
+    -   At all sizes, variation in SPEI had a greater impact on both growth and survival.
+        Both unusually wet and unusually dry conditions resulted in reductions in survival in fragments, with lagged effects of up to around 15 months.
+        Similarly for plant size, extremes in both directions in the census month caused reductions in growth in fragments.
+        However, the lagged effects of SPEI on plant size depended on season with the general pattern of unusually dry wet seasons and unusually wet dry seasons being beneficial to plant growth.
+        Together, this could possibly be explained by reduced microclimate buffering in forest fragments.
+        More research... <!--# add more, possibly move this whole paragraph. -->
 
     -   What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
+
+    -   Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
+        <!--# combine with first bullet point. -->
 
@@
 
--   Previous work has shown flowering probability to be an important vital rate.
-    Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
-
     -   Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
@@
     -   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
+
+    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
+        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
+        <!--# double-check against models and edit. -->
 
@@
 
-5.  Ecological Consequences of these patterns: an eye towards a drier future
+5.  Ecological Consequences of these patterns: an eye towards a drier/wetter future
 

### 2021-03-22 doc/cost_of_reproduction.Rmd
added cost of reproduciton supplemental

---
title: "Model ANOVA tables with flowering in prev year as covariate"
author: "Eric R. Scott"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
## target knits Rmds in their own session, so load libraries here.
source(here::here("packages.R"))
```

```{r load-targets}
withr::with_dir(here(),{
  tar_load(c(s_cf_flwr, s_1ha_flwr, g_cf_flwr, g_1ha_flwr, f_cf_flwr, f_1ha_flwr))
  })
```

# Analysis

## Survival

### Continuous

```{r}
summary(s_cf_flwr)
```

```{r}
draw(s_cf_flwr)
```

### Fragments

```{r}
summary(s_1ha_flwr)
```

```{r}
draw(s_1ha_flwr)
```

## Growth

### Continuous

```{r}
summary(g_cf_flwr)
```

```{r}
draw(g_cf_flwr)
```

### Fragments

```{r}
summary(g_1ha_flwr)
```

## Flowering

### Continuous

```{r}
summary(f_cf_flwr)
```

```{r}
draw(f_cf_flwr)
```

### Fragments

```{r}
summary(f_1ha_flwr)
```

```{r}
draw(f_1ha_flwr)
```

## Reproducibility

<details>

<summary>

Reproducibility receipt

</summary>

```{r}
## datetime
Sys.time()

## repository
if(requireNamespace('git2r', quietly = TRUE)) {
  git2r::repository()
} else {
  c(
    system2("git", args = c("log", "--name-status", "-1"), stdout = TRUE),
    system2("git", args = c("remote", "-v"), stdout = TRUE)
  )
}

## session info
sessionInfo()
```

</details>

### 2021-03-22 R/demog.R

@@ -53,13 +53,17 @@ tidy_demog <- function(data) {
     #add binary column for flowering
     mutate(infl = if_else(is.na(infl) & (!is.na(shts) | !is.na(ht)), 0L, infl)) %>% 
     mutate(flwr = if_else(infl > 0, 1L, 0L), .after = infl) %>% 
+    mutate(flwr_prev = lag(flwr)) %>% 
     #create factors
-    mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
+    mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number, flwr_prev), as.factor)) %>% 
     mutate(year_fac = as.factor(year)) %>% 
     # arrange columns
     select(ranch, bdffp_reserve_no, plot, habitat, #site level
            # row, column, x_09, y_09, 
            ha_id_number, #plot level
            year,
-           ht, ht_prev, shts, shts_prev, size, size_prev, log_size, log_size_prev, infl_num = infl, flwr, surv, code_notes, code2)
+           ht, ht_prev, shts, shts_prev,
+           size, size_prev, log_size, log_size_prev,
+           infl_num = infl, flwr, flwr_prev, surv,
+           code_notes, code2)
 }

### 2021-03-22 R/fit_models.R

@@ -1,15 +1,21 @@
 
-fit_surv <- function(data) {
+fit_surv <- function(data, flwr_prev = FALSE) {
   ncores <- detectCores()
   cl <- makeForkCluster(ncores - 2)
-  bam(surv ~ 
-        s(log_size_prev, bs = "cr") +
-        s(plot, bs = "re") + #random intercept
-        # s(plot, log_size_prev, bs = "re") + #random slope
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
-          xt = list(bs = "cr")),
+  
+  f <- surv ~ 
+    s(log_size_prev, bs = "cr") +
+    s(plot, bs = "re") + #random intercept
+    s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
+      xt = list(bs = "cr"))
+  
+  if (flwr_prev == TRUE) {
+    f <- update(f, .~. + flwr_prev)
+  }
+  
+  bam(f,
       family = binomial,
       data = data,
       method = "REML",
@@ -17,18 +23,26 @@ fit_surv <- function(data) {
       cluster = cl)
 }
 
-fit_growth <- function(data){
+fit_growth <- function(data, flwr_prev = FALSE){
   ncores <- detectCores()
   cl <- makeForkCluster(ncores - 2)
   # use only living plants
   data2 <- data %>% filter(surv == 1, !is.na(log_size))
-  bam(log_size ~ 
-        s(log_size_prev, bs = "cr") + 
-        s(plot, bs = "re") + #random effect of plot on intercept
-        s(spei_history, L,
-          bs = "cb",
-          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
-          xt = list(bs = "cr")),
+  
+  f <- log_size ~ 
+    # flwr_prev +
+    s(log_size_prev, bs = "cr") + 
+    s(plot, bs = "re") + #random effect of plot on intercept
+    s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
+      xt = list(bs = "cr"))
+  
+  if (flwr_prev == TRUE) {
+    f <- update(f, .~. + flwr_prev)
+  }
+  
+  bam(f,
       # family = gaussian(link = "identity"),
[…]

### 2021-03-22 _targets.R

@@ -67,9 +67,19 @@ tar_plan(
   f_cf = fit_flwr(model_data_cf),
   f_1ha = fit_flwr(model_data_1ha),
   
+  ## With flowering in previous year as covariate
+  
+  s_cf_flwr = fit_surv(model_data_cf, flwr_prev = TRUE),
+  s_1ha_flwr= fit_surv(model_data_1ha, flwr_prev = TRUE),
+  g_cf_flwr = fit_growth(model_data_cf, flwr_prev = TRUE),
+  g_1ha_flwr= fit_growth(model_data_1ha, flwr_prev = TRUE),
+  f_cf_flwr = fit_flwr(model_data_cf, flwr_prev = TRUE),
+  f_1ha_flwr= fit_flwr(model_data_1ha, flwr_prev = TRUE),
+  
   # Validate and summarize results
   tar_render(validate_models, "doc/validate_models.Rmd"),
   tar_render(model_summary, "doc/model_summary.Rmd"),
+  tar_render(cost_of_reproduction, "doc/cost_of_reproduction.Rmd"),
   
   # Descriptive / Exploratory Data Analysis Figures
   normals = normals_data(),
@@ -92,6 +102,7 @@ tar_plan(
 
   ## Growth
   g_covar_plot = plot_covar_smooth(frag_model = g_1ha, cf_model = g_cf, covar = "log_size_prev") +
+                  geom_abline(slope = 1, color = "red", linetype = 2) +
                   labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})")),
   g_spei_plot = plot_cb_2panel(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$log(size_{t+1})$")),
   g_spei_diff_plot = plot_cb_diff(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$\\Delta log(size_{t+1})$ (CF – 1ha)")),

### 2021-03-22 doc/paper.Rmd
papers/bullet points to consider adding to the discussion
added new references to .bib

@@
 
+#### Citations / information to allocate to appropriate paragraphs
+
+-    @criley1994 : under hort conditions, shoots that flower in spring emerge in June/July the previous year.
+    \~6 months after shoot emergence that the developing infloresence is susceptible to adverse effects.
+
+-   @dowd (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production by Heliconia (one step past flaowering, so might be worth citing to say that there are other ways reprod.
+can be affected
+
+-   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimeová2018.
+Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiro2021
+
+-   @westerband2017: *Heliconia* grew more at low than high precipitation.
+    Leaf life span and biomass allocation did not differ among treatments.
+    The effect of environmental variability was dependent on early life conditions as well as precipitation, suggesting that generalist species may experience high fitness as forest environments become more variable by maintaining high growth at the expense of survival.
+
+-   the observed effects on reproduction can have effects beyond demography, eg.
+    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortes2013, there can also be drought effects eg @sifres2006)
+
+-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stiles1975
+
+-   @wright1992: seasonal drought.
+    experimental irrigation influences survival and growth of tropical herbs
+
+-   Effects of precip extremes on understory plants can be also be indirect:
+
+    -   high precip = lots of clouds, less photosynthesis, less or slower growth
+
+    -   low precip = trees lose leaves or die, more light gets through or gaps open.
+        This can be positive (more light = more growth) but not always ( @westerband2015) there is a threshold that can lead to photoinhibition ( @he1996, @mulkey1991).
+        Drought and herbivory effects can be additive, at least in seedlings ( @barton2020)
+
+-   @horvitz1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT
+
+-   @horvitz2002 Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
+    Survival of large stages was not generally affected by any of the factors.
+    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
+    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
+    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
+    For reproductives, it was additionally positively affected by current fruit production.
+
+-   @rundel1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
+    Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
+    See also @rundel2020 \
+
+-    @stiles1975?: drought resulted in less flowering in Heliconia
+
+-   @mulkey1996 and @bonal2016 will have goods stuff; haven't looked them over yet\
+
 <!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->

### 2021-03-24 doc/paper.Rmd
edded refs to discussion
fixed up a few citations

@@
 
--   @dowd (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production by Heliconia (one step past flaowering, so might be worth citing to say that there are other ways reprod.
+-   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production by Heliconia (one step past flaowering, so might be worth citing to say that there are other ways reprod.
 can be affected
 
--   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimeová2018.
+-   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesova2018.
     Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiro2021
@@
 -   the observed effects on reproduction can have effects beyond demography, eg.
-    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortes2013, there can also be drought effects eg @sifres2006)
+    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortes2013, there can also be drought effects eg @sifres2007)
 
@@
     Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
-    See also @rundel2020 \
+    See also @rundel2020\
 
--    @stiles1975?: drought resulted in less flowering in Heliconia
+-   @stiles1975?:
+    drought resulted in less flowering in Heliconia
 
@@
 National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
-This is publication No. ---- in the BDFFP Technical Series.
+This is publication No. ------ in the BDFFP Technical Series.
 

### 2021-03-24 doc/paper.Rmd
work on discussion

@@
 
-1.
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-
-2.  Further evidence delayed effects may be the rule (emphasis on demographic modeling)
 
@@
 
-3.
-What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat ~~and life-history stage~~ <!--# we didn't really do this.
-This implies a size * SPEI interaction -->.
-    [this lays out what these patterns are]
-
-    -   Survival at mean SPEI was similar in fragments and continuous forest for all plant sizes.
-        Plants were smaller in forest fragments, but grew slightly faster than plants in continuous forest, closing the size gap over the decade of data.
-
-    -   However, the effect of varying SPEI on survival and growth was much greater in fragments.
-
-    -   At all sizes, variation in SPEI had a greater impact on both growth and survival.
+In addition to finding delayed effects of SPEI on demography, we found that these effects differed between forest fragments and continuous forest habitat.
+Overall survival of plants was similar in fragments and continuous forest for all plant sizes.
+Although plants were smaller in forest fragments, they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
+However, at all sizes, variation in SPEI had a greater impact on both growth and survival in fragments compared to continuous forest.
 Both unusually wet and unusually dry conditions resulted in reductions in survival in fragments, with lagged effects of up to around 15 months.
 Similarly for plant size, extremes in both directions in the census month caused reductions in growth in fragments.
-        However, the lagged effects of SPEI on plant size depended on season with the general pattern of unusually dry wet seasons and unusually wet dry seasons being beneficial to plant growth.
-        Together, this could possibly be explained by reduced microclimate buffering in forest fragments.
-        More research... <!--# add more, possibly move this whole paragraph. -->
+However, the lagged effects of SPEI on plant size depended on season with the general pattern of extremes (unusually wet wet seasons and unusually dry dry seasons) being detrimental to plant growth.
+
+Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
+Average flowering probability of plants in continuous forest is equivalent to flowering once every 125 years compared to once every 288 years in 1-ha fragments (assuming average plant size and SPEI).
+<!--# this doesn't make sense.
+Would be better to pick a large plant size and plug in to both --> We also see delayed effects of drought in flowering with increased flowering in years when plants experienced a drought two dry seasons ago (15--20 months prior to census).
 
@@
 
-    -   Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
-        <!--# combine with first bullet point. -->
-
-    -   Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
-        <!--# from model intercepts -->
-
 -   Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
 
-    -   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].
-
-    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
-        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
-        <!--# double-check against models and edit. -->
-
-4.  Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]
+1.  Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]
 
@@
 
-5.  Ecological Consequences of these patterns: an eye towards a drier/wetter future
+    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
+        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
+        <!--# double-check against models and edit. -->
+
+2.  Ecological Consequences of these patterns: an eye towards a drier/wetter future
 

### 2021-03-24 doc/paper.Rmd
added notes on wright 2006

@@
 
--   @mulkey1996 and @bonal2016 will have goods stuff; haven't looked them over yet\
+-   @mulkey1996 and @bonal2016 will have goods stuff; haven't looked them over yet
+
+-   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
+    Might be due to increases solar irradiance.
+    Maybe evidence of lagged effects?
+    Table 2 is inconsistent with text, so not sure...
+
+-   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
+    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
+    So rather than bet-hedging, maybe extra dry dry seasons are actually just really prime conditions for *Heliconia* reproduction and recruitment.
+
+-   \
 

### 2021-03-31 doc/paper.Rmd
Working on discussion.

@@
 
+There was a pronounced threshold in flowering probability corresponding with a log size of about 5 (Figure \@ref(fig:flwr-size)).
 The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
@@
 ```
+
+<!--# instead of rug plot, add second panel or margin with box plots or violin plots of size at both habitats -->
 
@@
 
--   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production by Heliconia (one step past flaowering, so might be worth citing to say that there are other ways reprod.
+-   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
 can be affected
@@
     (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
-    So rather than bet-hedging, maybe extra dry dry seasons are actually just really prime conditions for *Heliconia* reproduction and recruitment.
+    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.
 
--   \
+-   cite Esteban et al. 2021 for negative effects of large positive SPEI\
 
@@
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+<!--# more here -->
 
 This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-We found evidence for delayed effects of drought which, in our models, are sometimes stronger predictors of vital rates than more recent conditions.
-Distributed lag non-linear models offer an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
+We found evidence for delayed effects of drought which are sometimes stronger predictors of vital rates than more recent conditions.
+The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
+This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
 
-In addition to finding delayed effects of SPEI on demography, we found that these effects differed between forest fragments and continuous forest habitat.
-Overall survival of plants was similar in fragments and continuous forest for all plant sizes.
-Although plants were smaller in forest fragments, they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
-However, at all sizes, variation in SPEI had a greater impact on both growth and survival in fragments compared to continuous forest.
-Both unusually wet and unusually dry conditions resulted in reductions in survival in fragments, with lagged effects of up to around 15 months.
-Similarly for plant size, extremes in both directions in the census month caused reductions in growth in fragments.
-However, the lagged effects of SPEI on plant size depended on season with the general pattern of extremes (unusually wet wet seasons and unusually dry dry seasons) being detrimental to plant growth.
+```{r flwr-prob-quartile, include=FALSE}
 
-Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
-Average flowering probability of plants in continuous forest is equivalent to flowering once every 125 years compared to once every 288 years in 1-ha fragments (assuming average plant size and SPEI).
-<!--# this doesn't make sense.
-Would be better to pick a large plant size and plug in to both --> We also see delayed effects of drought in flowering with increased flowering in years when plants experienced a drought two dry seasons ago (15--20 months prior to census).
+#figure out flowering probability of largest quartile of plants in each habitat
+tar_load(f_cf)
+tar_load(f_1ha)
 
--   What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].
+newdf_cf <-
+  f_cf$model %>% #use model data frame
+  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
+  filter(quartiles == 4)
 
--   Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.
+newdf_1ha <-
+  f_1ha$model %>% 
+  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
+  filter(quartiles == 4)
 
-1.  Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]
+#equivalent to flowering once every __ years
+1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
+1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
+```
 
-    -   I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year.
-        Flowering is a big investment so perhaps flowering in one year means that you don't flower the following one.
-        If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth.
-        It will also be easier to explain as you write the paper.
-        **Emilio will look into physiological responses to light and H20 in the horticultural and ecological literature, eg.
-Westerband papers, esp if focusing on herbs** Basically anything on how more or less precip influences vital rates or how plants respond to them, esp if linked to changes in canopy openness, litterfall, flowering in ENSO years, Bruna transplant experiments, Heliconia germination in secondary forest, important to emphasize: results are often counter-intuitive.
+*Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
+Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations?
+-->.
+Flowering probability for the largest plants is actually higher in forest fragments; however, because there are fewer large plants in fragments overall flowering probability is much lower.
+The average fitted probability of flowering for plants in the upper quartile for size is 0.125 in continuous forest and 0.094 in forest fragments, equivalent to flowering once every 7.97 years and 10.63 years, respectively.
+Seemingly small differences in vital rates such as these can be extremely important as they compound over many years, especially in long-lived organisms[@gagnonGrowthUnderstoryHerb2011].
 
-    -   cite Esteban et al. 2021 for negative effects of large positive SPEI
+We found that the effects of variation in SPEI differed between forest fragments and continuous forest habitat.
+At mean SPEI, overall survival of plants was similar in fragments and continuous forest, but extremes in SPEI, both positive (wet) and negative (dry) had much greater detrimental effects on survival in habitat fragments.
+Similarly variation in SPEI had a greater impact on plant size in fragments compared to continuous forest with a general trend of unusually wet wet seasons and unusually dry dry seasons having a greater negative impact in fragments.
+Flowering probability showed a different pattern with a greater effect of variation in SPEI in continuous forests.
+However, with few plants large enough to flower, it is likely we simply did not detect the full effects of variation in SPEI on flowering probability in forest fragments with our models.
+Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC?
+-->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
 
-    -   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
-e.g.
-S.
-Joe Wright trees flowering after ENSO -->.
-
-    -   Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
-        In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
-        Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H.
-acuminata*.
-
-    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
-        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
-        <!--# double-check against models and edit. -->
-
-2.  Ecological Consequences of these patterns: an eye towards a drier/wetter future
-
--   More work to be done on flowering responses to drought x light x season interactions.
-
--   Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
-    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
-
-6.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]
+1.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]
 
@@
 
-Caveats:
+Past climate conditions were important for survival, size, and flowering probability.
+Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
+Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
+Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
+For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
+Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
+Might end up integrating with a later paragraph.
+-->.
 
-<!--# need to edit.  Moved from methods. -->
+There are several possible non-mutually-exclusive explanations for delayed effects of SPEI on vital rates.
+First, delayed effects may represent a demographic delay since we do not know precisely when since the previous survey the plant died or changed size.
+For example, if extreme drought in the dry season before the census increases plant mortality in that season, this would show up as a delayed effect in our models (e.g.
+in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
+This explanation for delayed effects applies only to plant size and survival, as flowering occurs very close to the yearly demographic census (Jan--Apr).
 
-One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
-This limitation meant that we were unable to model a potential size by environment (i.e.
-drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
-To check for a size by environment interaction, we followed the methods of Tredennick et al.
-[-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package [@batesFittingLinearMixedeffects2015].
-Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
-For each vital rate, we essentially replaced the crossbasis function in Eq 1.
-with a random effect of year on either the intercept or both the intercept and relationship with plant size.
-We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].
+Second, these delayed effects may represent a physiological delayed effect of environmental conditions.
+Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@criley1994].
+<!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H.
+acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.
+"Differentiation of the apical meristem" ?
+--> likely happens many months earlier.
+The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
+e.g.
+S.
+Joe Wright trees flowering after ENSO -->.
+We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
+However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
+Our models show that drought two dry seasons prior to the census increase flowering probability.
+A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.
+
+Finally, delayed effects could be mediated by indirect effects and ecological interactions.
+<!--# I'm not sure if this is really an explanation for the *delay* per se.
+Need to think about this.
+-->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration (<!--# citations -->).
+In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
+<!--# Heliconia grow better at edges, where there's more light -->Here, the delayed effect could be a combined delay of the response of canopy trees to drought and the response of *H.
+acuminata* to increased insolation.
+
+-   Ecological Consequences of these patterns: an eye towards a drier/wetter future
+
+-   Address how all but the smallest sizes are shrinking on average over the period of this study <!--# this is not true.
+A bug previously made it look like all plants were shrinking.
+Keeping this because citations -->(Figure \@ref(fig:growth-size)).
+    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
 
@@
     So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
+    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.
 

### 2021-04-01 doc/paper.Rmd
revise intro slightly

@@
 These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
+<!--# Additionally, extreme wet episodes are becoming as severe and frequent as droughts in parts of the Amazon, and can also have negative effects on plants [@estebanOtherSideDroughts2021].
+-->All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
 Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
@@
 Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
-First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)*** <!--# citations in proposal -->, making it challenging to draw broader demographic conclusions.
+First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
 Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
@@
 We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
-<!--# add a sentence about "famous" amazonian droughts --> Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
+This time series includes "famous" Amazonian droughts in 1997 [@mcphadenChildProdigy1997981999] and 2005[@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008] that were called "once in a century" climate events.
+Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
 Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
@@
 There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
-
-<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->
 
@@
 
--   @horvitz2002 Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
+-   @horvitz2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
     Survival of large stages was not generally affected by any of the factors.

### 2021-04-01 doc/paper.Rmd
worked on buffering paragraph

@@
 Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
-We addressed the following questions:
+We addressed the following questions:\
+\
+<
+!--# Gaoue =?
+Orou G.
+Gaoue, Carol C.
+Horvitz, Ulrich K.
+Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,https://doi.org/10.1016/j.ecolmodel.2019.03.007.
+-->
 
@@
 *Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
+Size regression became more likely as plants increased in size, which is typical of long-lived forest understory plants <!--# citations -->.
 Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations?
@@
 
-1.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]
-
-    1.  We also found evidence that the reduction of microclimatic buffering by fragmentation has demographic consequences for understory plants.
-    2.
-Mean survival and growth rates over the entire decade were similar in continuous forest and 1-ha fragments, but were more strongly affected by variation in SPEI in fragments.
-As climate becomes more variable, hot, and dry, drought frequency and severity is expected to increase in the Amazon (CITE IPCC?).
-Our results indicate that increasing drought frequency and severity will reduce *H.
-acuminata* growth and survival more in fragments compared to continuous forest.
-    3.  We did not see the same pattern with flowering probability, likely because flowering was overall much less likely in fragments where plants were smaller on average.
+Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
+A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
+It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H.
+acuminata* growth and survival in fragments compared to continuous forest.
+It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
+Bruna et al.
+[-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H.
+accuminata* plants transplanted near edges was reduced relative to continuous forest.
+A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
 

### 2021-04-01 doc/paper.Rmd
un-gitingore paper.docx for emilio

@@
 This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
+<!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->
 
@@
 
--   Address how all but the smallest sizes are shrinking on average over the period of this study <!--# this is not true.
-A bug previously made it look like all plants were shrinking.
-Keeping this because citations -->(Figure \@ref(fig:growth-size)).
-    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.
-
 # Conclusions & Future Directions

### 2021-04-01 doc/paper.Rmd
minor terminology note

--- 2021-04-01 doc/paper.Rmd
@@
 \
-<
-!--# Gaoue =?
+\<
+!--\# Gaoue =
+?
 Orou G.
@@
 Horvitz, Ulrich K.
-Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,https://doi.org/10.1016/j.ecolmodel.2019.03.007.
--->
+Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
+--
+\>
 
@@
 
+#### Vocab
+
+Delayed effects, time lag, the lag...avoid lagged effects
+
 #### Citations / information to allocate to appropriate paragraphs
@@
 It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
-Bruna et al.
-[-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H.
-accuminata* plants transplanted near edges was reduced relative to continuous forest.
+Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
 A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.

### 2021-04-06 README.Rmd
update readme

@@
 
-The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
-Binder uses rocker-project.org Docker images to ensure a consistent and reproducible computational environment.
-These Docker images can also be used locally.
+<!-- The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
+-->
+
+<!-- Binder uses rocker-project.org Docker images to ensure a consistent and reproducible computational environment. -->
+
+<!-- These Docker images can also be used locally. -->
 
 You can download the compendium as a zip from from this URL: [master.zip](/archive/master.zip).
-After unzipping:
 
--   open the `.Rproj` file in RStudio
--   something about `renv`
--   something about `targets::tar_make()` and `targets::tar_visnetwork()` <!-- -   run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the [DESCRIPTION](/DESCRIPTION) file).
---> <!-- -   finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console -->
+To run the compendium and reproduce all outputs:
+
+-   Open the `.Rproj` file in RStudio
+-   Install the `renv` package if it's not already installed and run `renv::restore()` to install all package dependencies.
+(I'm not actually sure that this step is necessar.
+Should auto-load because of .Rprofile.)
+-   Run `targets::tar_make()` to run all code and produce all outputs.
 

### 2021-04-06 doc/paper.Rmd
switch to four-panel plot for size covariate with size distribution at bottom.

@@
     eda_plot,
-    s_covar_plot,
     s_spei_plot,
     s_spei_diff_plot,
-    g_covar_plot,
     g_spei_plot,
     g_spei_diff_plot,
-    f_covar_plot,
     f_spei_plot,
-    f_spei_diff_plot
+    f_spei_diff_plot,
+    size_plot
   ))
@@
 ```{r}
-cap <- "Smooth effect of plant size in the previous census on survival modeled by a cubic regression spline.
+cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
+Panel d) shows the smoothed density of different plant sizes in the two habitat types.
 95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
 The smooths for 1-ha fragments and continuous forest are fit in separate models.
-The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.
+In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
 ```
 
-```{r surv-size, fig.cap=cap}
-s_covar_plot
+```{r size-plot, fig.cap=cap, fig.height=8, fig.width=5}
+size_plot
 ```
@@
 Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
-Survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:surv-size)).
+Survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a).
 Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
@@
 
-```{r}
-cap <- "Smooth effect of plant size in year t on plant size in year t+1 modeled by a cubic regression spline.
-95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
-The dashed red line represents a 1:1 line for reference.
-The smooths for 1-ha fragments and continuous forest are fit in separate models.
-Model intercepts were added to the fitted values of the smooths."
-```
-
-```{r growth-size, fig.cap=cap}
-g_covar_plot
-```
-
 <!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->
@@
 Size in year t was a significant predictor of size in year t+1 (p \< 0.001).
-The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:growth-size)).
+The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:size-plot)b).
 
@@
 
-```{r}
-cap <- "Smooth effect of plant size in the previous census on flowering probability modeled by a cubic regression spline.
-95% confidence intervals are shown and include uncertainty in the model intercept as well as uncertainty due to smoothness selection.
-The smooths for 1-ha fragments and continuous forest are fit in separate models.
-The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale"
-```
-
-```{r flwr-size, fig.cap=cap}
-f_covar_plot
-```
-
-<!--# instead of rug plot, add second panel or margin with box plots or violin plots of size at both habitats -->
-
 Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
-Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:flwr-size)).
+Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:size-plot)c).
 Flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest.
-However, because mean plant size is larger in continuous forest, overall flowering probability was greater.
-<!--# idea: plot vertical lines representing mean or median size in each habitat -->
+However, because mean plant size is larger in continuous forest (Figure \@ref(fig:size-plot)d), overall flowering probability was greater.
 
@@
 # Discussion
-
-#### Vocab
-
-Delayed effects, time lag, the lag...avoid lagged effects
 

### 2021-04-06 R/plot_smooths.R

@@ -1,3 +1,5 @@
+
+
 #' Evaluate a smooth with confidence intervals and backtransform
 #'
 #' Evaluates a smooth, adds the model intercept, confidence intervals, and
@@ -16,7 +18,6 @@ my_eval_smooth <- function(model, smooth, ...) {
     add_confint() %>% 
     mutate(across(c(est, lower_ci, upper_ci), ~linkinv(.x + coef(model)[1])))
 }
-
 
 
 #' Plots smooth covariates from two models
@@ -39,6 +40,35 @@ plot_covar_smooth <- function(cf_model, frag_model, covar) {
     theme_bw()
   p
 }
+
+make_size_plot <- function(s, g, f, model_data) {
+  
+  d <-
+    ggplot(model_data, aes(x = log_size_prev, fill = habitat, color = habitat)) +
+    geom_density(alpha = 0.5) +
+    labs(y = "Density", x = TeX("$log(size_t)$")) +
+    theme_bw() +
+    theme(legend.position = "none")
+  
+  top <-
+    s /
+    g /
+    f & 
+    theme(axis.title.x = element_blank(),
+          axis.text.x = element_blank(),
+          axis.ticks.x = element_blank())
+  
+  top /
+    d +
+    plot_layout(guides = "collect", ) +
+    plot_annotation(tag_levels = "a", tag_suffix = ")") &
+    theme(plot.margin = margin()) &
+    # set color for all panels
+    scale_color_discrete_qualitative(palette = "Set 2",
+                                     aesthetics = c("color", "fill"))
+}
+
+
 
 #' Makes a little bar indicating wet and dry seasons.
 #'

### 2021-04-06 doc/paper.Rmd
fix citations and figure refs.  Closes #35

@@
     f_spei_diff_plot,
-    size_plot
+    size_plot,
+    f_cf,
+    f_1ha
   ))
@@
 While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
-These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
 While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
@@
 SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmmTMPA3B432011]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
 <!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->
@@
 where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
-These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodsimonn.ChapterSmoothers] and Gasparrini et al.
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al.
 [-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
@@
 
-There was a pronounced threshold in flowering probability corresponding with a log size of about 5 (Figure \@ref(fig:flwr-size)).
+There was a pronounced threshold in flowering probability corresponding with a log size of about 5 (Figure \@ref(fig:size-plot)c).
 The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
@@
 
--   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
 can be affected
 
--   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesova2018.
-    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiro2021
+-   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesovaBelowgroundPlantFunctional2018.
+    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021
 
--   @westerband2017: *Heliconia* grew more at low than high precipitation.
+-   @westerbandEarlyLifeConditions2017: *Heliconia* grew more at low than high precipitation.
     Leaf life span and biomass allocation did not differ among treatments.
@@
 -   the observed effects on reproduction can have effects beyond demography, eg.
-    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortes2013, there can also be drought effects eg @sifres2007)
+    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
 
--   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stiles1975
+-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
 
--   @wright1992: seasonal drought.
+-   @wrightSeasonalDroughtSoil1992: seasonal drought.
     experimental irrigation influences survival and growth of tropical herbs
@@
     -   low precip = trees lose leaves or die, more light gets through or gaps open.
-        This can be positive (more light = more growth) but not always ( @westerband2015) there is a threshold that can lead to photoinhibition ( @he1996, @mulkey1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @barton2020)
+        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020)
 
--   @horvitz1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT
+-   @horvitzDemographicCostReproduction1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT
 
--   @horvitz2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
+-   @horvitzEffectsPlantSize2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
     Survival of large stages was not generally affected by any of the factors.
@@
 
--   @rundel1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
+-   @rundelStructuralPhysiologicalAdaptation1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
     Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
-    See also @rundel2020\
+    See also @rundelFunctionalTraitsBroadleaved2020\
 
@@
 
--   @mulkey1996 and @bonal2016 will have goods stuff; haven't looked them over yet
+-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
 
@@
 We found evidence for delayed effects of drought which are sometimes stronger predictors of vital rates than more recent conditions.
-The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
+The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
 This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
@@
 #figure out flowering probability of largest quartile of plants in each habitat
-tar_load(f_cf)
-tar_load(f_1ha)
-
+library(tidyverse)
 newdf_cf <-
@@
 Second, these delayed effects may represent a physiological delayed effect of environmental conditions.
-Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@criley1994].
+Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994].
 <!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H.
@@
 
--   Looop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
+-   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
 

### 2021-04-07 doc/paper.Rmd
fix multi-line equation knitting to word.  Closes #30.
added notes about Janssen et al. 2021

@@
 $$
-\begin{gather}
-    g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
-    \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j 
-\end{gather}
+\begin{split}
+& g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
+& \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot }j 
+\end{split}
 $$
@@
 
--   cite Esteban et al. 2021 for negative effects of large positive SPEI\
+-   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
+
+-   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+    \
 
@@
 Need to think about this.
--->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration (<!--# citations -->).
-In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+-->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
 This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
-<!--# Heliconia grow better at edges, where there's more light -->Here, the delayed effect could be a combined delay of the response of canopy trees to drought and the response of *H.
-acuminata* to increased insolation.
+<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
+Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
+Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
+Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.
 
+-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
+    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
+    Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
+    Dry season droughts result in green-up, but not wet season droughts.
+    During drought, old leaves are shed and new leaves are maintained.
+    2-5 months for newly flushed leaves to fully expand.
+
+```{=html}
+<!-- -->
+```
 -   Ecological Consequences of these patterns: an eye towards a drier/wetter future

### 2021-04-09 doc/paper.Rmd
add a few notes

@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 
--   @stiles1975?:
-    drought resulted in less flowering in Heliconia
+-   @stiles1975?: drought resulted in less flowering in Heliconia
 
@@
 -   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
-    \
+
+-   refs in @leitoldNinoDroughtIncreased2018 peak in tree mortality can happen up to 3 years after drought event in experimental studies.
+    Observational studies show mortality happening in same year as 2007 and 2010 droughts.
+
+-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
+    Similar increase in continuous and (large) fragments.
+    Increase in turnover more variable in fragments.
+    Coarse woody debris increased 62% during drought.\
 
@@
 It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
-Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
+Bruna et al.
+[-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H.
+accuminata* plants transplanted near edges was reduced relative to continuous forest.
 A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
@@
 
-```{=html}
-<!-- -->
-```
 -   Ecological Consequences of these patterns: an eye towards a drier/wetter future

### 2021-04-12 doc/paper.Rmd
Edits to Intro

@@
 -->.
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
 -->.
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
-For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Despite this, evidence linking fragmentation-related changes in environmental conditions with population-level demographic responses remains scarce.
+For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.
 
@@
 
-Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017].
-These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-<!--# Additionally, extreme wet episodes are becoming as severe and frequent as droughts in parts of the Amazon, and can also have negative effects on plants [@estebanOtherSideDroughts2021].
--->All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
-Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
-Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
+For instance, <!--#parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
+In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].
 
@@
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
-These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
+These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
 They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
@@
 
-We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
-This time series includes "famous" Amazonian droughts in 1997 [@mcphadenChildProdigy1997981999] and 2005[@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008] that were called "once in a century" climate events.
 Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
-We addressed the following questions:\
-\
-\<
-!--\# Gaoue =
-?
-Orou G.
-Gaoue, Carol C.
-Horvitz, Ulrich K.
-Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
---
-\>
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
+This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\
 
@@
 
-2)  Are there delayed effects of drought on demographic responses, and if so what lag times are most critical?
+2)  Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
 
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H.
-acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
+#### Vocab
+
+This seems tacked on here and perhaps better in the discussion: populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+
+\< !--
+\# Gaoue = ?
+Orou G.
+Gaoue, Carol C.
+Horvitz, Ulrich K.
+Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
+-- \>
+
+Delayed effects, time lag, the lag...avoid lagged effects
+
 #### Citations / information to allocate to appropriate paragraphs
@@
 
--   @stiles1975?: drought resulted in less flowering in Heliconia
+-   @stiles1975?
+    drought resulted in less flowering in Heliconia
 

### 2021-04-12 doc/paper.Rmd
Edits Methods

@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
-While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl.
+While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare *[unpubl.
 data]*.
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 Plants that were not found for three consecutive surveys were considered dead.
-There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+There is a seven-fold range in *H.
+acuminata* density in the forest plots, with density in several of the continuous forest plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
 
-## Weather data
+## Climate data
 
@@
 
-We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
+To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
-We chose 36 months as a maximum lag because prior *H.
-acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+We chose 36 months as a maximum lag because prior transplant experiments with *H.
+acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
+
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
 The general form of the vital rate models was as follows:
@@
 Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
-A random effect of plot ID on the intercept is represented by $a_j$, allowing for plot characteristics to influence average vital rates.
-For plant growth the response was plant size (natural log of height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution.
+
+To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
+We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
+This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
 A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
-The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences).
-When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.
+For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
+We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .
 
@@
 
-Because the `dlnm` package does not currently allow modeling of factor by smooth interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+The `dlnm` package does not currently allow modeling of factor by smooth interactions.
+This means we cannot \_\_\_\_\_\_\_\_\_\_\_\_\_\_.
+We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
 
@@
 
-The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
+Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\

### 2021-04-12 doc/paper.Rmd
Results changes

@@
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+**THIS NEEDS TO BE BETTER LINKED TO THE INFO ABOVE:** Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
 Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
 
-## Survival
+## Survival, growth, and flowering in continuous forest vs.  fragments
+
+**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, *Heliconia acuminata* individuals had similarly high survival rates in both CF and 1-ha (0.95).
+The proportion of surviving plants was lowest in 2004 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
+The lowest survival for 1-ha fragment plots (0.93) was for the 2005-2006 transition year, which corresponded to wetter than average conditions (SPEI \> 0.5) but with a moderate drought the preceding October.
+(october 2005?).
+Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
+While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
+However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
+
+<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap?
+-->
+
+Plants were in continuous forest were larger on average than those in 1-ha fragments.
+<!--# This is across the whole timeseries, which I'm not sure is actually that important.
+What summary statistics would be useful?
+--> Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
+Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
+Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were significantly larger than those in forest fragments (150 ± 175 SD vs.
+112 ± 141 SD, respectively).
+This disparity in plant size was most notable in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
+As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
+While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
+
+<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->
+
+While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
+This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
+This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c).
+Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
+Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs.
+only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
+The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
+<!--# not sure how necessary this is. EB: same-->
+
+<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->
+
+## Delayed effects of drought on demographic vital rates
 
@@
 ```
-
-Overall survival rate across the entire dataset was the same (0.95) in both CF and 1-ha.
-The proportion of surviving plants across both habitat types was lowest in 2004 (0.92) which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (0.93) was in the 2006 census corresponding to wetter than average conditions (SPEI \> 0.5), but with a moderate drought the preceding October.
-
-Log transformed plant size had a significant effect on survival in both continuous forest and 1-ha fragments (both p \< 0.001).
-Survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a).
-Survival of the smallest plants was higher in 1-ha fragments, but the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
 
@@
 
-The random effect of plot was significant for continuous forest ( p \< 0.001), but not for 1-ha fragments (p = 0.257).
-
-## Growth
-
-Plants were on average larger in continuous forest compared to 1-ha fragments.
-<!--# This is across the whole timeseries, which I'm not sure is actually that important.
-What summary statistics would be useful?
--->In continuous forest, plants had an average of 2.9 ± 1.8 shoots (mean ± standard deviation) and were on average 40.6 ± 26.5 cm tall with an average size (shoots $\times$ height) of 150 ± 175.
-In 1-ha fragments, plants had 2.5 ± 1.5 shoots on average and a mean height of 36.2 ± 24.1 cm and a mean size of 112 ± 141.
-The size difference between plants in continuous forest and 1-ha fragments was more dramatic early on in the timeseries with mean sizes becoming more similar over time (Figure \@ref(fig:eda)a).
-
-<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->
-
-Size in year t was a significant predictor of size in year t+1 (p \< 0.001).
-The 95% confidence intervals for the effect of size in year t in fragments and continuous forest overlapped for most sizes, but was slightly higher in continuous forest for mid-sized plants (Figure \@ref(fig:size-plot)b).
-
 ```{r}
@@
 This is due to reduced growth in fragments under wet conditions.
-
-The random effect of plot is significant in both habitat models (p \< 0.001).
-
-## Flowering
-
-There was a pronounced threshold in flowering probability corresponding with a log size of about 5 (Figure \@ref(fig:size-plot)c).
-The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
-In fact, the proportion of flowering plants was higher in continuous forest in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
-The most striking difference between habitat types in flowering proportion coincided with a severe drought in 2003 when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest and 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
-The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
-<!--# not sure how necessary this is -->
-
-Plant size had a significant effect on flowering probability in both habitats (p \< 0.001).
-Flowering probability increased rapidly with log-transformed plant size greater than 5 in both habitats (Figure \@ref(fig:size-plot)c).
-Flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest.
-However, because mean plant size is larger in continuous forest (Figure \@ref(fig:size-plot)d), overall flowering probability was greater.
 
@@
 
-The random effect of plot also had a significant effect in both habitats (p \< 0.001).
+FInally, with the exception of growth in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
 

### 2021-04-14 doc/paper.Rmd
knit paper
a little work on the intro (opening paragraph, effect of frag on vital rates)

@@
 
+Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern
+
 Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H.
-acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g.
-SPEI) affect an outcome (e.g.
-growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g.
-between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
@@
 acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
+
+<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
 
@@
 
-**THIS NEEDS TO BE BETTER LINKED TO THE INFO ABOVE:** Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
-
 ## Survival, growth, and flowering in continuous forest vs. fragments
 
-**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, *Heliconia acuminata* individuals had similarly high survival rates in both CF and 1-ha (0.95).
-The proportion of surviving plants was lowest in 2004 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan--Mar of 2004 (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (0.93) was for the 2005-2006 transition year, which corresponded to wetter than average conditions (SPEI \> 0.5) but with a moderate drought the preceding October.
-(october 2005?).
+**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
+The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
+The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN.
 Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
@@
 
-Plants were in continuous forest were larger on average than those in 1-ha fragments.
-<!--# This is across the whole timeseries, which I'm not sure is actually that important.
-What summary statistics would be useful?
---> Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
+<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->
+
+Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
 Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were significantly larger than those in forest fragments (150 ± 175 SD vs.
+Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
-This disparity in plant size was most notable in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
-As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
+This disparity in plant size was most notable in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
 While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
@@
 This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
-This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c).
+This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
 Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
@@
 <!--# not sure how necessary this is. EB: same-->
+
+<!-- This paragraphis talkng about size, but not really growth -->
 
@@
 
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
+
 #### Vocab
@@
 
-Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-<!--# more here -->
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+<!--# more here -->.
+Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
+Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales.
+Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+Most Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
+
+# Temporal variation in demographic responses to forest fragmentation
+
+Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
+In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
+While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher in fragments.
+The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
+However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs.
+0.094 in FF).
+These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
+However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
+every 10.63 years in continuous forest.
+Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].
+
+# Delayed effects of climate on demographic vital rates
 
@@
 
-*Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
-Size regression became more likely as plants increased in size, which is typical of long-lived forest understory plants <!--# citations -->.
-Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations?
--->.
-Flowering probability for the largest plants is actually higher in forest fragments; however, because there are fewer large plants in fragments overall flowering probability is much lower.
-The average fitted probability of flowering for plants in the upper quartile for size is 0.125 in continuous forest and 0.094 in forest fragments, equivalent to flowering once every 7.97 years and 10.63 years, respectively.
-Seemingly small differences in vital rates such as these can be extremely important as they compound over many years, especially in long-lived organisms[@gagnonGrowthUnderstoryHerb2011].
-
 We found that the effects of variation in SPEI differed between forest fragments and continuous forest habitat.

### 2021-04-19 doc/choose_size_var.Rmd
add appendix exploring which size proxy is best

---
title: "Untitled Draft"
author: "Report Author"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, paged.print = FALSE)
## target knits Rmds in their own session, so load libraries here.
source(here::here("packages.R"))
withr::with_dir(here::here(), {
  targets::tar_load(model_data)
})
library(performance)
```

\#TODO: The last part with the transplant experiment data is probably enough.
Probably delete the GAM comparisons and make the last bit work with `targets` by adding that data file as a `target`.

## What proxy of size is best?

The goal is to determine whether plant height, number of shoots, or their product (shoots \* height) is a good proxy for plant biomass.
First, using the 10 year demographic dataset, we can see how vital rate models fit using each of these as a predictor perform.

## Set up

Create columns for log-transformed size variables.

```{r}
df <- 
  model_data %>% 
  mutate(log_ht_prev = log(ht_prev),
         log_ht = log(ht),
         log_shts_prev = log(shts_prev),
         log_shts = log(shts))
```

Set up parallel computation

```{r}
ncores <- detectCores()
cl <- makeForkCluster(ncores - 2)
```

### Build formulas

```{r}
s_size <- surv ~ 
  s(log_size_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))

s_ht <- surv ~ 
  s(log_ht_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))

s_shts <- surv ~ 
  s(log_shts_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))
```

## Survival

```{r}
m_size <- 
  bam(s_size,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_ht <- 
  bam(s_ht,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_shts <- 
  bam(s_shts,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

### Compare performance

```{r}
compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
```

log(size) has lowest AIC, similar R2 to height.
Number of shoots is worst fit to survival.

## Flowering

```{r}
f_ht <- update(s_ht, flwr ~ .)
f_size <- update(s_size, flwr ~ .)
f_shts <- update(s_shts, flwr ~ .)
```

Use only live plants for flowering probability.

```{r}
df2 <- df %>% filter(surv == 1)
```

```{r}
m_size <- 
  bam(f_size,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_ht <- 
  bam(f_ht,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_shts <- 
  bam(f_shts,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

### Compare performance

```{r}
compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
```

## Data approach

Which is a better predictor of leaf area, stem number, height, or size (stem number\* height)?

Using data from transplant experiment (CITATION)

```{r}
library(readxl)
library(janitor)
la <- read_excel(here("data", "HA-la-stems-ht.xlsx")) %>% clean_names()
la <- 
  la %>%
  rename(leaf_area = leaf_area_feb_99,
         leaves = leaves_feb_99,
         ht = height_feb_99,
         shts = stems_feb_99) %>% 
  mutate(size = shts * ht) %>% 
  mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
  mutate(log_leaves_ht = log(leaves * ht))

la$exp %>% unique()
```

```{r}
ggplot(la, aes(x = ht, y = leaf_area)) + geom_point()
ggplot(la, aes(x = shts, y = leaf_area)) + geom_point()
ggplot(la, aes(x = size, y = leaf_area)) + geom_point()
ggplot(la, aes(x = leaves, y = leaf_area)) + geom_point()
```

I can already tell size is going to be the best

```{r}
hist(la$leaf_area)
hist(la$log_leaf_area)
```

I'll use log_leaf_area as response

```{r}
m_ht <- lm(log_leaf_area ~ log_ht, data = la)
m_shts <- lm(log_leaf_area ~ log_shts, data = la)
m_size <- lm(log_leaf_area ~ log_size, data = la)
```

Options we don't have (because we don't have leaf number), but might be better proxies of plant size:

```{r}
m_leaves <- lm(log_leaf_area ~ log_leaves, data = la)
m_leaves_ht <- lm(log_leaf_area ~ log_leaves_ht, data = la)
m_leaves_shts <- lm(log_leaf_area ~ I(log(leaves*shts)), data = la)
```

```{r}
AIC(m_leaves_ht)
```

```{r}
compare_performance(m_ht, m_shts, m_size, m_leaves, m_leaves_ht,m_leaves_shts, metrics = "common", rank = TRUE)
```

## Reproducibility

<details>

<summary>

Reproducibility receipt

</summary>

```{r}
## datetime
Sys.time()

## repository
if(requireNamespace('git2r', quietly = TRUE)) {
  git2r::repository()
} else {
  c(
    system2("git", args = c("log", "--name-status", "-1"), stdout = TRUE),
    system2("git", args = c("remote", "-v"), stdout = TRUE)
  )
}

## session info
sessionInfo()
```

</details>

### 2021-04-19 _targets.R

@@ -17,6 +17,7 @@ tar_plan(
   xa_spei = calc_spei_xa(xa_raw),
   xa_lag = lag_spei(xa_spei, maxlag),
 
+  tar_render(choose_size_var, "doc/choose_size_var.Rmd"),
   
   
   # SPEI supplemental

### 2021-04-19 doc/choose_size_var.Rmd
pare appendix down to essentials

@@
 ---
-title: "Untitled Draft"
-author: "Report Author"
+title: "Choice of plant size proxy"
+author: "Eric R. Scott"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 knitr::opts_chunk$set(echo = FALSE, paged.print = FALSE)
-## target knits Rmds in their own session, so load libraries here.
-source(here::here("packages.R"))
+
 withr::with_dir(here::here(), {
-  targets::tar_load(model_data)
+  targets::tar_load(la_data)
 })
+
+library(tidyverse)
 library(performance)
@@
 
-\#TODO: The last part with the transplant experiment data is probably enough.
-Probably delete the GAM comparisons and make the last bit work with `targets` by adding that data file as a `target`.
-
 ## What proxy of size is best?
 
-The goal is to determine whether plant height, number of shoots, or their product (shoots \* height) is a good proxy for plant biomass.
-First, using the 10 year demographic dataset, we can see how vital rate models fit using each of these as a predictor perform.
+In our 10 year demographic dataset we have yearly measurements of plant height (height to tip of tallest leaf) and shoot number.
+Height $\times$ shoot number is a possibility for an integrative proxy of plant size.
 
-## Set up
-
-Create columns for log-transformed size variables.
+Using data from transplant experiments (CITATION), we show that plant size (height $\times$ shoot number) best explains variation in total plant leaf area.
 
 ```{r}
-df <- 
-  model_data %>% 
-  mutate(log_ht_prev = log(ht_prev),
-         log_ht = log(ht),
-         log_shts_prev = log(shts_prev),
-         log_shts = log(shts))
+la <- 
+  la_data %>% 
+  mutate(size = shts * ht) %>% 
+  #log transform all size variables
+  mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
+  filter(!is.na(leaf_area) & !is.na(ht) & !is.na(shts)) %>% 
+  filter(shts > 0)
 ```
-
-Set up parallel computation
-
-```{r}
-ncores <- detectCores()
-cl <- makeForkCluster(ncores - 2)
-```
-
-### Build formulas
-
-```{r}
-s_size <- surv ~ 
-  s(log_size_prev, bs = "cr") +
-  s(plot, bs = "re") + #random intercept
-  s(spei_history, L,
-    bs = "cb",
-    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
-    xt = list(bs = "cr"))
-
-s_ht <- surv ~ 
-  s(log_ht_prev, bs = "cr") +
-  s(plot, bs = "re") + #random intercept
-  s(spei_history, L,
-    bs = "cb",
-    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
-    xt = list(bs = "cr"))
-
-s_shts <- surv ~ 
-  s(log_shts_prev, bs = "cr") +
-  s(plot, bs = "re") + #random intercept
-  s(spei_history, L,
-    bs = "cb",
-    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
-    xt = list(bs = "cr"))
-```
-
-## Survival
-
-```{r}
-m_size <- 
-  bam(s_size,
-      family = binomial,
-      data = df,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-
-m_ht <- 
-  bam(s_ht,
-      family = binomial,
-      data = df,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-
-m_shts <- 
-  bam(s_shts,
-      family = binomial,
-      data = df,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-### Compare performance
-
-```{r}
-compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
-```
-
-log(size) has lowest AIC, similar R2 to height.
-Number of shoots is worst fit to survival.
-
-## Flowering
-
-```{r}
-f_ht <- update(s_ht, flwr ~ .)
-f_size <- update(s_size, flwr ~ .)
-f_shts <- update(s_shts, flwr ~ .)
-```
-
-Use only live plants for flowering probability.
-
-```{r}
-df2 <- df %>% filter(surv == 1)
-```
-
-```{r}
-m_size <- 
-  bam(f_size,
-      family = binomial,
-      data = df2,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-
-m_ht <- 
-  bam(f_ht,
-      family = binomial,
-      data = df2,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-
-m_shts <- 
-  bam(f_shts,
-      family = binomial,
-      data = df2,
-      method = "REML",
-      select = TRUE,
-      cluster = cl)
-```
-
-### Compare performance
-
-```{r}
-compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
-```
-
-## Data approach
-
-Which is a better predictor of leaf area, stem number, height, or size (stem number\* height)?
-
-Using data from transplant experiment (CITATION)
-
-```{r}
-library(readxl)
-library(janitor)
-la <- read_excel(here("data", "HA-la-stems-ht.xlsx")) %>% clean_names()
-la <- 
-  la %>%
-  rename(leaf_area = leaf_area_feb_99,
-         leaves = leaves_feb_99,
-         ht = height_feb_99,
-         shts = stems_feb_99) %>% 
-  mutate(size = shts * ht) %>% 
-  mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
-  mutate(log_leaves_ht = log(leaves * ht))
-
-la$exp %>% unique()
-```
-
-```{r}
-ggplot(la, aes(x = ht, y = leaf_area)) + geom_point()
-ggplot(la, aes(x = shts, y = leaf_area)) + geom_point()
-ggplot(la, aes(x = size, y = leaf_area)) + geom_point()
-ggplot(la, aes(x = leaves, y = leaf_area)) + geom_point()
-```
-
-I can already tell size is going to be the best
 
@@
 hist(la$log_leaf_area)
+#use log(leaf area) as response to better meet normality assumptions.
 ```
-
-I'll use log_leaf_area as response
 
@@
 
-Options we don't have (because we don't have leaf number), but might be better proxies of plant size:
-
 ```{r}
-m_leaves <- lm(log_leaf_area ~ log_leaves, data = la)
-m_leaves_ht <- lm(log_leaf_area ~ log_leaves_ht, data = la)
-m_leaves_shts <- lm(log_leaf_area ~ I(log(leaves*shts)), data = la)
+check_model(m_ht)
+check_model(m_shts)
+check_model(m_size)
 ```
@@
 ```{r}
-AIC(m_leaves_ht)
+compare_performance(m_ht, m_shts, m_size, metrics = "common", rank = TRUE)
 ```
@@
 ```{r}
-compare_performance(m_ht, m_shts, m_size, m_leaves, m_leaves_ht,m_leaves_shts, metrics = "common", rank = TRUE)
+plot_fit <- function(m) {
+  df <- broom::augment(m, se_fit = TRUE)
+
+  pred <- colnames(df)[2]
+  resp <- colnames(df)[1]
+
+  p <- ggplot(df, aes_string(x = pred)) +
+    geom_point(aes_string(y = resp)) + 
+    geom_ribbon(aes(ymin = .fitted - .se.fit, ymax = .fitted + .se.fit), alpha = 0.4) +
+    geom_line(aes(y = .fitted))
+  return(p)
+}
 ```
+
+```{r}
+plot_fit(m_ht) + labs(x = "log(height)", y = "log(leaf area)")
+plot_fit(m_shts) + labs(x = "log(shoot number)", y = "log(leaf area)")
+plot_fit(m_size) + labs(x = "log(shoots * height)", y = "log(leaf area)")
+```
+
+Using the log of size (shoot number $\times$ height) as a predictor gives the highest $R^2$ and lowest RMSE of the models as well as the lowest AIC indicating that it results in the best explanatory power and model fit for predicting log(leaf area) as a response.
 

### 2021-04-19 _targets.R

@@ -17,7 +17,10 @@ tar_plan(
   xa_spei = calc_spei_xa(xa_raw),
   xa_lag = lag_spei(xa_spei, maxlag),
 
-  tar_render(choose_size_var, "doc/choose_size_var.Rmd"),
+  # Choice of size supplemental
+  tar_target(la_file, here("data", "HA-la-stems-ht.xlsx"), format = "file"),
+  la_data = read_tidy_la(la_file),
+    tar_render(choose_size_var, "doc/choose_size_var.Rmd"),
   
   
   # SPEI supplemental

### 2021-04-26 doc/paper.Rmd
more discussion edits - looking good

@@
 Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
-For instance, <!--#parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
 These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
@@
 112 ± 141 SD, respectively).
-This disparity in plant size was most notable in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
+This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
 While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
@@
 Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Most Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
+Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
+
+Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+Using a decadal dataset of plant demography, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
 
@@
 In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
-While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher in fragments.
+While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
 The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
@@
 
-This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-We found evidence for delayed effects of drought which are sometimes stronger predictors of vital rates than more recent conditions.
-The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
-This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
-<!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->
+Precipitation and other climatic factors are known to have large effects on plant growth, survival, or reproduction.
+Despite this, models of population dynamics have rarely included the effects of environmental drivers.
+This has largely due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+
+However, while it appeared that delayed effects of climate on demographic vital rates were ubiquitous, the extent to which they vary spatially or with habitat remained an open question [@eversLaggedDormantSeason2021].
+Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
+For example, [delete this phrase?
+the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
+However, the magnitude of these declines was far greater in forest fragments.
+Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
+In contrast, variation in SPEI had a far stronger effect on populations in continuous forests, though it is is important to recognize that because there are fewer plants in fragments above the threshold-size for flowering it is possible we lacked the statistical power to detect the full effects of SPEI in fragments.
+
+Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC?
+-->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
 
@@
 
-We found that the effects of variation in SPEI differed between forest fragments and continuous forest habitat.
-At mean SPEI, overall survival of plants was similar in fragments and continuous forest, but extremes in SPEI, both positive (wet) and negative (dry) had much greater detrimental effects on survival in habitat fragments.
-Similarly variation in SPEI had a greater impact on plant size in fragments compared to continuous forest with a general trend of unusually wet wet seasons and unusually dry dry seasons having a greater negative impact in fragments.
-Flowering probability showed a different pattern with a greater effect of variation in SPEI in continuous forests.
-However, with few plants large enough to flower, it is likely we simply did not detect the full effects of variation in SPEI on flowering probability in forest fragments with our models.
-Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC?
--->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
-
 Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
@@
 
--   Very few demographic studies have the data required to fit models that include delayed effects.
+-    few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
+    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
+    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->
+
+-
 

### 2021-04-26 doc/paper.Rmd
continued edits - moved the mechanistic paragraphs up.

@@
 Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
-In contrast, variation in SPEI had a far stronger effect on populations in continuous forests, though it is is important to recognize that because there are fewer plants in fragments above the threshold-size for flowering it is possible we lacked the statistical power to detect the full effects of SPEI in fragments.
+In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than frragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
 
-Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC?
+There are several, non-mutually-exclusive explanations for the effects of SPEI on vital rates could take months or even years to manifest.
+First, these delayed effects may represent a physiological delayed effect of environmental conditions.
+Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994].
+<!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H.
+acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.
+"Differentiation of the apical meristem" ?
+--> likely happens many months earlier.
+The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
+e.g.
+S.
+Joe Wright trees flowering after ENSO -->.
+We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
+However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
+Our models show that drought two dry seasons prior to the census increase flowering probability.
+A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.
+
+Second, delayed effects could be mediated by indirect effects and ecological interactions.
+<!--# I'm not sure if this is really an explanation for the *delay* per se.
+Need to think about this.
+-->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
+This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
+<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
+Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
+Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
+Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.
+
+Finally, it could be an artefact of the timing of responses vis-à-vis the census date.
+If extreme drought in the dry season before the census increases plant mortality in that season, for example, this would show up as a delayed effect in our models (e.g.
+in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
+This explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
+
+# Implications of delayed effects of climate on demography
+
+Both extreme values and variation in SPEI are predicted to increase as climate change progresses ( <!--# cite IPCC?
 -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
@@
 -->.
-
-There are several possible non-mutually-exclusive explanations for delayed effects of SPEI on vital rates.
-First, delayed effects may represent a demographic delay since we do not know precisely when since the previous survey the plant died or changed size.
-For example, if extreme drought in the dry season before the census increases plant mortality in that season, this would show up as a delayed effect in our models (e.g.
-in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
-This explanation for delayed effects applies only to plant size and survival, as flowering occurs very close to the yearly demographic census (Jan--Apr).
-
-Second, these delayed effects may represent a physiological delayed effect of environmental conditions.
-Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994].
-<!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H.
-acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.
-"Differentiation of the apical meristem" ?
---> likely happens many months earlier.
-The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
-e.g.
-S.
-Joe Wright trees flowering after ENSO -->.
-We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
-However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
-Our models show that drought two dry seasons prior to the census increase flowering probability.
-A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.
-
-Finally, delayed effects could be mediated by indirect effects and ecological interactions.
-<!--# I'm not sure if this is really an explanation for the *delay* per se.
-Need to think about this.
--->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
-This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
-<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
-Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
-Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
-Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.
 

### 2021-04-29 doc/paper.Rmd
almost done with discussion

@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H.
+acuminata*---are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
@@
 
--   @criley1994 : under hort conditions, shoots that flower in spring emerge in June/July the previous year.
-    \~6 months after shoot emergence that the developing infloresence is susceptible to adverse effects.
-
 -   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
@@
 
--   @westerbandEarlyLifeConditions2017: *Heliconia* grew more at low than high precipitation.
-    Leaf life span and biomass allocation did not differ among treatments.
-    The effect of environmental variability was dependent on early life conditions as well as precipitation, suggesting that generalist species may experience high fitness as forest environments become more variable by maintaining high growth at the expense of survival.
-
 -   the observed effects on reproduction can have effects beyond demography, eg.
@@
 -   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
-
--   @wrightSeasonalDroughtSoil1992: seasonal drought.
-    experimental irrigation influences survival and growth of tropical herbs
 
@@
         This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020)
-
--   @horvitzDemographicCostReproduction1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT
-
--   @horvitzEffectsPlantSize2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
-    Survival of large stages was not generally affected by any of the factors.
-    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
-    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
-    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
-    For reproductives, it was additionally positively affected by current fruit production.
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
 
@@
 
--   @stiles1975?
-    drought resulted in less flowering in Heliconia
-
--   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
+    @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
 
@@
 
-Precipitation and other climatic factors are known to have large effects on plant growth, survival, or reproduction.
+Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992].
+For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought.
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
-This has largely due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
@@
 
-However, while it appeared that delayed effects of climate on demographic vital rates were ubiquitous, the extent to which they vary spatially or with habitat remained an open question [@eversLaggedDormantSeason2021].
+While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021].
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
@@
 Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
-In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than frragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
+In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
 
-There are several, non-mutually-exclusive explanations for the effects of SPEI on vital rates could take months or even years to manifest.
-First, these delayed effects may represent a physiological delayed effect of environmental conditions.
-Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994].
-<!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H.
-acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.
-"Differentiation of the apical meristem" ?
---> likely happens many months earlier.
-The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation.
-e.g.
-S.
-Joe Wright trees flowering after ENSO -->.
-We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
-However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
-Our models show that drought two dry seasons prior to the census increase flowering probability.
-A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.
+There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
+Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
+Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering.
+While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988].
+In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.
 
-Second, delayed effects could be mediated by indirect effects and ecological interactions.
-<!--# I'm not sure if this is really an explanation for the *delay* per se.
-Need to think about this.
--->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
-This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
-<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
-Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
-Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
-Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.
+Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
+For example, the mortality rate of tropical trees can spike in the year following a severe drought.
+The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
+Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
+Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
+Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.
 
-Finally, it could be an artefact of the timing of responses vis-à-vis the census date.
-If extreme drought in the dry season before the census increases plant mortality in that season, for example, this would show up as a delayed effect in our models (e.g.
-in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
-This explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
+Finally, demographic delays could be an artefact of the timing of responses vis-à-vis the census date.
+If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g.
+in Figure \@ref(fig:surv-spei)b).
+In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
+[consequence of demography on an annual scale and weather on a monthly / seasonal scale.
+This problem not unique to us, same for everyone.
+Mismatched scale at which analyses operate]
 
@@
 
--
+-   More experiments like westerband that try and see how contiions today influence growth in future years Generalizations about the direction and magnitude of delayed demographic responses will continue to prove elusive.
+    Without experiments designed greater understanding of the complex ways in which light, environmental conditions, and other ecological conditions interactions to influence plant vital rates (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002),
 

### 2021-04-29 doc/SPEI_appendix.Rmd

@@
 ```{r}
+citations <- tribble(~year, ~Reference,
+                     2005, "Marengo et al. 2008; Zeng et al. 2008",
+                     2010, "Lewis et al. 2011",
+                     1997, "McPhaden 1999",
+                     2002, NA,
+                     2009, NA,
+                     2004, NA,
+                     2006, NA) 
+```
+
+```{r}
 spei_table <-
@@
   ungroup()
+
+spei_table <- left_join(spei_table, citations)
 ```

### 2021-04-30 doc/paper.Rmd
line breaks at end of sentence
hyphens, n-dashes, m-dashes

@@
 -->.
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
 -->.
@@
 Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
-For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
+For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
 In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].
 
-Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern
+<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->
 
@@
 The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
-Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
 There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
@@
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
-The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
+The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
 This study uses data from 1-ha fragments and Continuous Forest sites.
@@
 
-<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
+<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
 -->
@@
 
-<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
+<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
 
@@
 where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
-These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al.
-[-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see @woodChapterSmoothers and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
@@
 For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
-We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .
+We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
 
 In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
-We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].
+We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in @marraPracticalVariableSelection2011.
 
@@
 Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
-Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales.
+Using a decadal dataset of plant demography, we show that some---but not all---results from short-term studies can be extrapolated to longer times scales.
 Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
@@
 0.094 in FF).
-These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
+These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
 However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
@@
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
-This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+This has largely been due to the challenge---both ecologically and statistically---of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
 
@@
 For example, [delete this phrase?
-the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
+the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 <!--# weeks? --> prior to the appearance of infloresences.
 Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
@@
 The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
-Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
+Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
 Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
@@
 acuminata* growth and survival in fragments compared to continuous forest.
-It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
-Bruna et al.
-[-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H.
-accuminata* plants transplanted near edges was reduced relative to continuous forest.
+It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1-ha fragments.
+Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
 A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.

### 2021-04-30 R/plot_eda.R
fix typo in argument name

@@ -66,7 +66,7 @@ eda_plot_df <- function(demog_post) {
 #' @param demog_post demography data
 #' @param dates_lims x-axis limits
 #'
-plot_eda_surv_ts <- function(demog_post, dates_lims) {
+plot_eda_surv_ts <- function(demog_post, date_lims) {
   
   date_breaks <- seq(date_lims[1], date_lims[2], by = "year")
   

### 2021-04-30 _targets.R
Edits to results, survival, growth and flowering (exploratory data analysis section)

@@ -91,6 +91,7 @@ tar_plan(
   plot_dates = get_plot_daterange(model_data, maxlag),
   
   eda_spei = plot_eda_spei(xa_lag, plot_dates),
+  eda_surv_ts = plot_eda_surv_ts(model_data, plot_dates),
   eda_surv = plot_eda_surv_cohort(demog_done, plot_dates),
   eda_size = plot_eda_size(model_data, plot_dates),
   eda_flwr = plot_eda_flwr(model_data, plot_dates),

### 2021-04-30 doc/paper.Rmd
Edits to results, survival, growth and flowering (exploratory data analysis section)

@@
 library(targets)
+library(tidyverse)
+library(tsibble)
+library(lubridate)
 library(knitr)
 library(here)
+library(patchwork)
 ```
@@
     f_cf,
-    f_1ha
+    f_1ha,
+    model_data,
+    xa_spei
   ))
@@
 This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
@@
 
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
 The general form of the vital rate models was as follows:
@@
 
-**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
+```{r include=FALSE}
+summary_stats <-
+  model_data %>%
+  group_by(habitat) %>%
+  summarize(across(c(ht, shts, size, flwr, surv),
+                   .fns = list(mean = mean, sd = sd),
+                   na.rm = TRUE)) %>%
+  right_join(model_data %>% count(habitat)) %>%
+  mutate(across(ends_with("_mean") | ends_with("_sd"),
+                                ~round(.x, 2)))
+summary_stats <- 
+  full_join(
+  summary_stats,
+  model_data %>%
+    count(habitat, plot) %>%
+    group_by(habitat) %>%
+    summarize(n_mean = mean(n), n_sd = sd(n)) %>%
+    mutate(across(where(is.numeric), round))
+)
+summary_stats
+
+by_year <-
+  model_data %>% 
+  group_by(year, habitat) %>% 
+  summarize(prop_surv = mean(surv)) %>% 
+    arrange(prop_surv)
+by_year
+
+xa_spei %>% ungroup() %>%
+  summarize(across(everything(), mean)) %>%
+  filter(year(yearmonth)==2005)
+
+```
+
+Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
+When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
 The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN.
+The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
 Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
@@
 
-<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap?
--->
-
-<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->
-
 Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
-Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs.
+Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
+Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
@@
 
-While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
+While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
 This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
-This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
-Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
+This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
+Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
 Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs.
@@
 <!-- This paragraphis talkng about size, but not really growth -->
-
-<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->
 

### 2021-04-30 doc/paper.Rmd
fixed a couple typos in statistics

@@
 Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
-The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
+The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
 
@@
 
-FInally, with the exception of growth in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
+FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
 

### 2021-05-03 doc/paper.Rmd
change CSL to Global Change Biology

@@
 bibliography: references.bib
-csl: "templates/ecology.csl" # Insert path for the bib-style
+csl: "templates/global-change-biology.csl" # Insert path for the bib-style
 abstract: |

### 2021-05-03 doc/paper.Rmd
fix citations

@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others?
--->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others?
--->.
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
--->.
+Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
-For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
 Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.
 
-<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->
+Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
+While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
+Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @bruna2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@laurance2001; @opdam2004, @selwood2015].
 
-Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
-For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
-In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].
-
-<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->
-
-Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
+Whether plant populations in tropical forest fragments are more susceptible to drought remains unclear for three primary reasons.
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
@@
 Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009] <!--# Emilio had written "Gaoue" as potential reference, but the on I have in Zotero is about trees, not understory plants -->.
+We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
 This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\
@@
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
-
-#### Vocab
-
-This seems tacked on here and perhaps better in the discussion: populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
-
-\< !--
-\# Gaoue = ?
-Orou G.
-Gaoue, Carol C.
-Horvitz, Ulrich K.
-Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
--- \>
-
-Delayed effects, time lag, the lag...avoid lagged effects
-
-#### Citations / information to allocate to appropriate paragraphs
-
--   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
-can be affected
-
--   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesovaBelowgroundPlantFunctional2018.
-    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021
-
--   the observed effects on reproduction can have effects beyond demography, eg.
-    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
-
--   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
-
--   Effects of precip extremes on understory plants can be also be indirect:
-
-    -   high precip = lots of clouds, less photosynthesis, less or slower growth
-
-    -   low precip = trees lose leaves or die, more light gets through or gaps open.
-        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
-
--   @rundelStructuralPhysiologicalAdaptation1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
-    Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
-    See also @rundelFunctionalTraitsBroadleaved2020\
-
-    @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
-
--   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
-    Might be due to increases solar irradiance.
-    Maybe evidence of lagged effects?
-    Table 2 is inconsistent with text, so not sure...
-
--   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
-    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
-    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.
-
--   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
-
--   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
-
--   refs in @leitoldNinoDroughtIncreased2018 peak in tree mortality can happen up to 3 years after drought event in experimental studies.
-    Observational studies show mortality happening in same year as 2007 and 2010 droughts.
-
--   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
-    Similar increase in continuous and (large) fragments.
-    Increase in turnover more variable in fragments.
-    Coarse woody debris increased 62% during drought.\
-
-<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->
-
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
@@
 0.094 in FF).
+
 These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
-However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
+However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
 every 10.63 years in continuous forest.
@@
 
-Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992].
-For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought.
+Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
+These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
+There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
-This has largely been due to the challenge---both ecologically and statistically---of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
@@
 
-While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021].
+However, while it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, [delete this phrase?
-the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
+For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
-The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 <!--# weeks? --> prior to the appearance of infloresences.
-Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
-Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering.
-While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988].
-In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@evers2021], and hence be shaped by climatic events at any point in that physiological window.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
+Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
+Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
+While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
+In fact, flowering in one year is associated with increased reproduction and growth in the next.
+**(is this an acceptable colloquial way to interpret the covariate effect?)**.
 
-Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
-For example, the mortality rate of tropical trees can spike in the year following a severe drought.
-The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
-Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
-Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
-Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.
+Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
+The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimeová2018].
+This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@bruna2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
+This would also be consistent with the results of prior experiments, in which the growth rates of *H.
+acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@bruna2005].
+
+Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
+For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
+This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
+Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
+\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 
@@
 In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
-[consequence of demography on an annual scale and weather on a monthly / seasonal scale.
-This problem not unique to us, same for everyone.
-Mismatched scale at which analyses operate]
+This is possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
+While the very slow growth and extremely low mortality rates of *H.
+acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.
 
-# Implications of delayed effects of climate on demography
+***THIS INFO PROB NEEDS TO BE WORKED IN SOMEWHERE ABOVE***
 
-Both extreme values and variation in SPEI are predicted to increase as climate change progresses ( <!--# cite IPCC?
--->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
+
+Past climate conditions were important for survival, size, and flowering probability.
+Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
+Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
+Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
+For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
+Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
+Might end up integrating with a later paragraph.
+-->.
 
@@
 
-Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
-A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H.
-acuminata* growth and survival in fragments compared to continuous forest.
-It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1-ha fragments.
-Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
-A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
-
-Past climate conditions were important for survival, size, and flowering probability.
-Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
-Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
-Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
-For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
-Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
-Might end up integrating with a later paragraph.
--->.
-
--   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
-    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
-    Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
-    Dry season droughts result in green-up, but not wet season droughts.
-    During drought, old leaves are shed and new leaves are maintained.
-    2-5 months for newly flushed leaves to fully expand.
-
--   Ecological Consequences of these patterns: an eye towards a drier/wetter future
-
 # Conclusions & Future Directions
 
--   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
-    Smaller/larger plants might be more susceptible to drought.
-    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.
+Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
+Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@laurance2014].
+However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @bruna2005a; @markewitz2010; @westerband2015; @westerbandEarlyLifeConditions2017)
 
--   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
-While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-They could act in concert to exacerbate or dampen population dynamics."
+Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms (cite `dlnm`).
+They also don't allow for modeling interactions between \_\_\_\_ and plant size; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019], the ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
 
--   few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
-    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
-    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->
-
--   More experiments like westerband that try and see how contiions today influence growth in future years Generalizations about the direction and magnitude of delayed demographic responses will continue to prove elusive.
-    Without experiments designed greater understanding of the complex ways in which light, environmental conditions, and other ecological conditions interactions to influence plant vital rates (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002),
-
--   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
+Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @evers2021].
+Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@bruna2005].
+Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
 
@@
 
+### Left-over text
+
+-   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+-   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is.
+It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
+-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
+-   eg.
+gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
+-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
+Estimates green-up, literfall, growth, etc.
+for central and southern america including the Amazon.
+Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
+Dry season droughts result in green-up, but not wet season droughts.
+During drought, old leaves are shed and new leaves are maintained.
+2-5 months for newly flushed leaves to fully expand
+
+#### Citations / information could allocate to appropriate paragraphs
+
+-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+can be affected
+
+-   Effects of precip extremes on understory plants can be also be indirect:
+
+    -   high precip = lots of clouds, less photosynthesis, less or slower growth
+
+    -   low precip = trees lose leaves or die, more light gets through or gaps open.
+        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
+
+-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
+
+-   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
+    Might be due to increases solar irradiance.
+    Maybe evidence of lagged effects?
+    Table 2 is inconsistent with text, so not sure...
+
+-   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
+
+-   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+
+-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
+    Similar increase in continuous and (large) fragments.
+    Increase in turnover more variable in fragments.
+    Coarse woody debris increased 62% during drought.
+
+\newpage
+
 ### Colophon

### 2021-05-03 doc/paper.Rmd
fix citations

--- 2021-05-03 doc/paper.Rmd
@@
 
-\< !--
-\# Gaoue = ?
+Gaoue = ?
 Orou G.
@@
 Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
--- \>
+
+-   Eric: I don't think so. It was a reference supporting that understory plants are understudied.
 
@@
         This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020 )
 
@@
     Coarse woody debris increased 62% during drought.\
-
-<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->
 
@@
 Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
-Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
+Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare [sensu @westerband2015; @westerbandEarlyLifeConditions2017; @horvitzEffectsPlantSize2002; @gagnonGrowthUnderstoryHerb2011]<!--# what is this "sensu" in reference to?
+The definition of vital rates?
+of "rare"?
+--> .
 Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

### 2021-05-03 doc/paper.Rmd
edits

--- 2021-05-03 doc/paper.Rmd
@@
 bibliography: references.bib
-csl: "templates/global-change-biology.csl" # Insert path for the bib-style
+csl: "templates/ecology.csl" # Insert path for the bib-style
 abstract: |
@@
 library(targets)
-library(tidyverse)
-library(tsibble)
-library(lubridate)
 library(knitr)
 library(here)
-library(patchwork)
 ```
@@
     f_cf,
-    f_1ha,
-    model_data,
-    xa_spei
+    f_1ha
   ))
@@
 -->.
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
 -->.
@@
 Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
-For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
+For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
 In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].
 
-<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->
+Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern
 
@@
 The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
-Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
 There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
@@
 In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
-The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
+The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
 This study uses data from 1-ha fragments and Continuous Forest sites.
@@
 
-<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
+<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
 -->
@@
 This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
@@
 
-<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
+<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
 
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
 The general form of the vital rate models was as follows:
@@
 where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
-These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see @woodChapterSmoothers and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al.
+[-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
@@
 For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
-We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
+We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .
 
 In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
-We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in @marraPracticalVariableSelection2011.
+We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].
 
@@
 
-```{r include=FALSE}
-summary_stats <-
-  model_data %>%
-  group_by(habitat) %>%
-  summarize(across(c(ht, shts, size, flwr, surv),
-                   .fns = list(mean = mean, sd = sd),
-                   na.rm = TRUE)) %>%
-  right_join(model_data %>% count(habitat)) %>%
-  mutate(across(ends_with("_mean") | ends_with("_sd"),
-                                ~round(.x, 2)))
-summary_stats <- 
-  full_join(
-  summary_stats,
-  model_data %>%
-    count(habitat, plot) %>%
-    group_by(habitat) %>%
-    summarize(n_mean = mean(n), n_sd = sd(n)) %>%
-    mutate(across(where(is.numeric), round))
-)
-summary_stats
-
-by_year <-
-  model_data %>% 
-  group_by(year, habitat) %>% 
-  summarize(prop_surv = mean(surv)) %>% 
-    arrange(prop_surv)
-by_year
-
-xa_spei %>% ungroup() %>%
-  summarize(across(everything(), mean)) %>%
-  filter(year(yearmonth)==2005)
-
-```
-
-Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
-When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
+**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
 The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
+The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN.
 Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
@@
 
+<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap?
+-->
+
+<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->
+
 Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
-Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
+Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
+Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
@@
 
-While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
+While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
 This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
-This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
-Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
+This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
+Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
 Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs.
@@
 <!-- This paragraphis talkng about size, but not really growth -->
+
+<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->
 
@@
 Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
-The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
+The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
 
@@
 
-FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
+FInally, with the exception of growth in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
 
@@
     Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021
-
--   the observed effects on reproduction can have effects beyond demography, eg.
-    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
-
--   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
 
@@
 Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
-Using a decadal dataset of plant demography, we show that some---but not all---results from short-term studies can be extrapolated to longer times scales.
+Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales.
 Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
@@
 0.094 in FF).
-These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
-However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
+
+These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
+However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
 every 10.63 years in continuous forest.
@@
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
-This has largely been due to the challenge---both ecologically and statistically---of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
 
@@
 For example, [delete this phrase?
-the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
+the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
-The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 <!--# weeks? --> prior to the appearance of infloresences.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
 Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
-Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering.
-While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988].
-In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.
+Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
+While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
+In fact, flowering in one year is associated with increased reproduction and growth in the next.
+**(is this an acceptable colloquial way to interpret the covariate effect?)**.
 
@@
 The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
-Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
+Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
 Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
@@
 In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
-[consequence of demography on an annual scale and weather on a monthly / seasonal scale.
-This problem not unique to us, same for everyone.
-Mismatched scale at which analyses operate]
+This is possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
+While the very slow growth and extremely low mortality rates of *H.
+acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.
 
@@
 acuminata* growth and survival in fragments compared to continuous forest.
-It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1-ha fragments.
+It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
 Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
@@
 -   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
-    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
+Estimates green-up, literfall, growth, etc.
+for central and southern america including the Amazon.
 Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
@@
 During drought, old leaves are shed and new leaves are maintained.
-    2-5 months for newly flushed leaves to fully expand.
+2-5 months for newly flushed leaves to fully expand
+
+Effects beyond Ha
+
+-   the observed effects on reproduction can have effects beyond demography, eg.
+    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
+
+-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
+
+# Conclusions & Future Directions
 
 -   Ecological Consequences of these patterns: an eye towards a drier/wetter future
-
-# Conclusions & Future Directions
 

### 2021-05-04 doc/paper.Rmd
added citations, continued editing discussion

@@
 
-Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern
+Several studies have demonstrated that climate change can have major effects on the population dynamics of species in the temperate zone (Doak and Morris 2010, Sletvold et al.
+2013, Selwood et al.
+2015, Williams et al.
+2015).
+In contrast, almost nothing is known about the consequences of changing climates for the population dynamics of tropical species.
+Mild droughts have been shown to increase the growth rates of tropical trees, perhaps due to reductions in cloud cover (Condit et al. 2004).
+However, in severe
+drought years growth can be extremely low while mortality can be sharply elevated, with this increased mortality persisting for several years (Phillips et al. 2010).
+The observed effects of mild droughts on reproduction are more complex, but include decreases in the abundance of pollinators (Stiles 1992, Jevanandam et al.
+2013) and a boom in drought-year fruit production followed by severe post-drought "famine" (Wright et al.
+1999, Pau et al.
+2013).
+In addition, while some studies have shown negligible effects of droughts on seedling recruitment and survivorship (Connell and Green 2000, Engelbrecht et al.
+2002), others have shown survivorship decreases significantly during droughts (Edwards and Krockenberger 2006).
+These results, coupled with evidence that the growth and survivorship of plants can be lower in
+fragments (Laurance et al.
+1998, Bruna 2002, Zartman et al.
+2015) and that droughts can exacerbate this disparity (Laurance et al.
+2011), has led to the ***hypothesis that as drought frequency or intensity increases, the viability of plant populations in fragments will decrease relative to that of populations in continuous forest***(Laurance and Williamson 2001, Opdam and Wascher 2004, Selwood et al.
+2015).
 
-Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
+Whether plant populations in tropical forest fragments are more susceptible to drought remains unclear for three primary reasons.
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
@@
 
-#### Vocab
-
-This seems tacked on here and perhaps better in the discussion: populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+#### Vocab: Delayed effects, time lag, the lag...avoid lagged effects
 
@@
 Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
--- \>
-
-Delayed effects, time lag, the lag...avoid lagged effects
+\>
 
@@
 can be affected
-
--   Lots of great stuff about the importance of rhizomes, including for drought resistance, in @klimesovaBelowgroundPlantFunctional2018.
-    Rhizome resistance to variability in environmental conditions also a conclusion in this study on invasion by another zingiberales ( @pinheiroEffectRhizomeExposure2021
 
@@
 
--   @rundelStructuralPhysiologicalAdaptation1998 Heliconia appear to allocate much of their above-ground biomass to leaf tissues and have a relatively low investment in support tissues.
-    Investment in leaf form vs structure interpreted as a trade-off between economy and protection against tissue loss from falling branches; presence of below-ground rhizomes in Heliconia may be key.
-    See also @rundelFunctionalTraitsBroadleaved2020\
-
-    @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
+-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
 
@@
 
--   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
-    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
-    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.
-
 -   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
@@
 -   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
-
--   refs in @leitoldNinoDroughtIncreased2018 peak in tree mortality can happen up to 3 years after drought event in experimental studies.
-    Observational studies show mortality happening in same year as 2007 and 2010 droughts.
 
@@
 
-Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992].
-For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought.
+Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
@@
 
-While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021].
+However, while it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, [delete this phrase?
-the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
+For example, although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
-The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@evers2021], and hence be shaped by climatic events at any point in that physiological window.
 For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
-Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
+Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
 Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
@@
 
-Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
-For example, the mortality rate of tropical trees can spike in the year following a severe drought.
-The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
-Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
-Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
-Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.
+Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
+The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimeová2018].
+This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@bruna2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
+This would also be consistent with the results of prior experiments, in which the growth rates of *H.
+acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@bruna2005].
+
+Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
+For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
+This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
+Similar delayed changes in the local environment could also influence the foraging behavior of *H.
+acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**]** it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 
@@
 
-# Implications of delayed effects of climate on demography
+*THIS PARAGRAPH HAS INFO THAT NEEDS TO BE WORKED IN SOMEWHERE ABOVE*
 
-Both extreme values and variation in SPEI are predicted to increase as climate change progresses ( <!--# cite IPCC?
--->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.
+Past climate conditions were important for survival, size, and flowering probability.
+Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
+Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
+Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
+For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
+Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
+Might end up integrating with a later paragraph.
+-->.
 
@@
 
-Reduced microclimate buffering could possibly explain the greater effects of variation in SPEI in fragments compared to continuous forest.
-A major consequence of forest fragmentation is reported alteration in microclimate near forest edges [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-It is possible that increases in temperature, evaporative drying, wind speed, and solar radiation commonly found near forest edges exacerbated the effects of extremes in SPEI on *H.
-acuminata* growth and survival in fragments compared to continuous forest.
-It is also possible that the microclimate experienced by plants in continuous forest plots was simply more stable compared to that of 1 ha fragments.
-Bruna et al. [-@brunaExperimentalAssessmentHeliconia2002] found that the growth of transplanted *H. accuminata* plants transplanted near edges was reduced relative to continuous forest.
-A greater understanding of the microclimatic changes in fragments during extreme climate events and their effects on understory plants is needed, but we believe differences in microclimate buffering are a likely explanation for the observed differences between fragments and continuous forest.
-
-Past climate conditions were important for survival, size, and flowering probability.
-Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
-Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
-Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
-For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
-Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
-Might end up integrating with a later paragraph.
--->.
-
--   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
-Estimates green-up, literfall, growth, etc.
-for central and southern america including the Amazon.
-Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
-Dry season droughts result in green-up, but not wet season droughts.
-During drought, old leaves are shed and new leaves are maintained.
-2-5 months for newly flushed leaves to fully expand
-
-Effects beyond Ha
-
--   the observed effects on reproduction can have effects beyond demography, eg.
-    gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
-
--   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
-
 # Conclusions & Future Directions
 
--   Ecological Consequences of these patterns: an eye towards a drier/wetter future
+Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
+Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@laurance2014].
+However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @bruna2005a; @markewitz2010; @westerband2015; @westerbandEarlyLifeConditions2017)
 
--   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
-    Smaller/larger plants might be more susceptible to drought.
-    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
-    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.
+Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms (cite `dlnm`).
+They also don't allow for modeling interactions between \_\_\_\_ and plant size; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019], the ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
 
--   Synergistic effects of fragmentation and climate (need the stats to catch up with us) "Organisms simultaneously experience multiple forms of disturbance.
-While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
-They could act in concert to exacerbate or dampen population dynamics."
+Finally, no analytical approcah for testign for climate lags can compensate for a lack of data (eveers tenhumberg)and long-term demographic data collected in fragmented landscapes [@bruna2005; @laurance1998].
+Unfortunately, both of these are rare [@brunaHabitatFragmentationPlant2009], suggesting generalizations regarding the causes and consequences of delayed effects will continue to prove elusive.
 
--   few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
+-   few demographic studies have the data required for modeling lagged effects .
     This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
@@
 
--   More experiments like westerband that try and see how contiions today influence growth in future years Generalizations about the direction and magnitude of delayed demographic responses will continue to prove elusive.
-    Without experiments designed greater understanding of the complex ways in which light, environmental conditions, and other ecological conditions interactions to influence plant vital rates (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002),
+-   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio
 
--   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
+-
 
@@
 
+### Left-over text
+
+-   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+-   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is.
+It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
+-   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
+-   eg.
+gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
+-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
+Estimates green-up, literfall, growth, etc.
+for central and southern america including the Amazon.
+Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
+Dry season droughts result in green-up, but not wet season droughts.
+During drought, old leaves are shed and new leaves are maintained.
+2-5 months for newly flushed leaves to fully expand
+
+\newpage
+
 ### Colophon

### 2021-05-04 doc/paper.Rmd

@@
 Mild droughts have been shown to increase the growth rates of tropical trees, perhaps due to reductions in cloud cover (Condit et al. 2004).
-However, in severe
-drought years growth can be extremely low while mortality can be sharply elevated, with this increased mortality persisting for several years (Phillips et al. 2010).
+However, in severe drought years growth can be extremely low while mortality can be sharply elevated, with this increased mortality persisting for several years (Phillips et al. 2010).
 The observed effects of mild droughts on reproduction are more complex, but include decreases in the abundance of pollinators (Stiles 1992, Jevanandam et al.
@@
 2002), others have shown survivorship decreases significantly during droughts (Edwards and Krockenberger 2006).
-These results, coupled with evidence that the growth and survivorship of plants can be lower in
-fragments (Laurance et al.
+These results, coupled with evidence that the growth and survivorship of plants can be lower in fragments (Laurance et al.
 1998, Bruna 2002, Zartman et al.
@@
 acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
-While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**]** it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 

### 2021-05-04 doc/paper.Rmd
edits - handing off to Eric

@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others?
--->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others?
--->.
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others?
--->.
+Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
-For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
+For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
 Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.
 
-<!--# CITATION_4: WRI Report? Or INPE, see here: https://rainforests.mongabay.com/amazon/deforestation_calculations.html or this: https://mapbiomas.org/en -->
-
-Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
-For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
-In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].
-
-Several studies have demonstrated that climate change can have major effects on the population dynamics of species in the temperate zone (Doak and Morris 2010, Sletvold et al.
-2013, Selwood et al.
-2015, Williams et al.
-2015).
-In contrast, almost nothing is known about the consequences of changing climates for the population dynamics of tropical species.
-Mild droughts have been shown to increase the growth rates of tropical trees, perhaps due to reductions in cloud cover (Condit et al. 2004).
-However, in severe drought years growth can be extremely low while mortality can be sharply elevated, with this increased mortality persisting for several years (Phillips et al. 2010).
-The observed effects of mild droughts on reproduction are more complex, but include decreases in the abundance of pollinators (Stiles 1992, Jevanandam et al.
-2013) and a boom in drought-year fruit production followed by severe post-drought "famine" (Wright et al.
-1999, Pau et al.
-2013).
-In addition, while some studies have shown negligible effects of droughts on seedling recruitment and survivorship (Connell and Green 2000, Engelbrecht et al.
-2002), others have shown survivorship decreases significantly during droughts (Edwards and Krockenberger 2006).
-These results, coupled with evidence that the growth and survivorship of plants can be lower in fragments (Laurance et al.
-1998, Bruna 2002, Zartman et al.
-2015) and that droughts can exacerbate this disparity (Laurance et al.
-2011), has led to the ***hypothesis that as drought frequency or intensity increases, the viability of plant populations in fragments will decrease relative to that of populations in continuous forest***(Laurance and Williamson 2001, Opdam and Wascher 2004, Selwood et al.
-2015).
+Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
+While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
+Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @bruna2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@laurance2001; @opdam2004, @selwood2015].
 
@@
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
-
-#### Vocab: Delayed effects, time lag, the lag...avoid lagged effects
-
-\< !--
-\# Gaoue = ?
-Orou G.
-Gaoue, Carol C.
-Horvitz, Ulrich K.
-Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
-\>
-
-#### Citations / information to allocate to appropriate paragraphs
-
--   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
-can be affected
-
--   Effects of precip extremes on understory plants can be also be indirect:
-
-    -   high precip = lots of clouds, less photosynthesis, less or slower growth
-
-    -   low precip = trees lose leaves or die, more light gets through or gaps open.
-        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
-
--   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
-
--   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
-    Might be due to increases solar irradiance.
-    Maybe evidence of lagged effects?
-    Table 2 is inconsistent with text, so not sure...
-
--   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
-
--   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
-
--   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
-    Similar increase in continuous and (large) fragments.
-    Increase in turnover more variable in fragments.
-    Coarse woody debris increased 62% during drought.\
-
-<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->
-
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
@@
 Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
+These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
+There is also evidence that the effects can persist for multiple years [ @phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
@@
 
-*THIS PARAGRAPH HAS INFO THAT NEEDS TO BE WORKED IN SOMEWHERE ABOVE*
+***THIS INFO PROB NEEDS TO BE WORKED IN SOMEWHERE ABOVE***
+
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
 
@@
 
-Finally, no analytical approcah for testign for climate lags can compensate for a lack of data (eveers tenhumberg)and long-term demographic data collected in fragmented landscapes [@bruna2005; @laurance1998].
-Unfortunately, both of these are rare [@brunaHabitatFragmentationPlant2009], suggesting generalizations regarding the causes and consequences of delayed effects will continue to prove elusive.
-
--   few demographic studies have the data required for modeling lagged effects .
-    This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.
-    <!--# Or something like "some of these long-term studies exist and are prime for doing these kinds of analyses" -->
-
--   Loop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio
-
--
+Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @evers2021].
+Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@bruna2005].
+Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
 
@@
 
+#### Citations / information could allocate to appropriate paragraphs
+
+-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+can be affected
+
+-   Effects of precip extremes on understory plants can be also be indirect:
+
+    -   high precip = lots of clouds, less photosynthesis, less or slower growth
+
+    -   low precip = trees lose leaves or die, more light gets through or gaps open.
+        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
+
+-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
+
+-   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
+    Might be due to increases solar irradiance.
+    Maybe evidence of lagged effects?
+    Table 2 is inconsistent with text, so not sure...
+
+-   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
+
+-   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+
+-   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
+    Similar increase in continuous and (large) fragments.
+    Increase in turnover more variable in fragments.
+    Coarse woody debris increased 62% during drought.
+
 \newpage

### 2021-05-04 doc/paper.Rmd
line breaks at sentences

@@
 These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [ @phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
 
@@
 This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
-Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
-While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
+\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 

### 2021-05-04 doc/paper.Rmd
remove duplicate citations

@@
 Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
@@
 These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @bruna2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@laurance2001; @opdam2004, @selwood2015].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @brunaExperimentalAssessmentHeliconia2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].
 
@@
 In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 
-<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
+<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
 
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
 The general form of the vital rate models was as follows:
@@
 
-These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
+These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
@@
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
 
@@
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
+For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
-The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@evers2021], and hence be shaped by climatic events at any point in that physiological window.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
 For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
@@
 Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
-While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
+While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
 In fact, flowering in one year is associated with increased reproduction and growth in the next.
@@
 Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
-The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimeová2018].
-This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@bruna2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
+The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
+This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
 This would also be consistent with the results of prior experiments, in which the growth rates of *H.
-acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@bruna2005].
+acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@brunaCompensatoryResponsesUnderstory2005].
 
-Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
-For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
+Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021].
+For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
 This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
-Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
+Similar delayed changes in the local environment could also influence the foraging behavior of *H.
+acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
 While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @bruna2005a; @markewitz2010; @westerband2015; @westerbandEarlyLifeConditions2017)
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017)
 
@@
 
-Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @evers2021].
-Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@bruna2005].
+Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @eversLaggedDormantSeason2021].
+Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
 Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

### 2021-05-04 doc/paper.Rmd
fix broken citations

@@
 Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
+acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
 While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
@@
         This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20
+        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020
 

### 2021-05-04 doc/paper.Rmd
change caption objects to unique names

@@
 	comment = "#>",
+	# fig.show = "hide" #hide all figures until the end
 	dpi = 300
@@
 ```{r}
-cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
 Precipitation (a) shows a marked dry seasons from June through October.
@@
 
-```{r normals, fig.cap=cap}
+```{r normals, fig.cap=normals_cap}
 normals_plot
@@
 ```{r}
-cap <- "Timeseries of demographic parameters and drought ocurrence.
+eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
 Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
@@
 
-```{r eda, fig.cap=cap, fig.height=7, fig.width=10}
+```{r eda, fig.cap=eda_cap, fig.height=7, fig.width=10}
 eda_plot
@@
 ```{r}
-cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
+size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
 Panel d) shows the smoothed density of different plant sizes in the two habitat types.
@@
 
-```{r size-plot, fig.cap=cap, fig.height=8, fig.width=5}
+```{r size-plot, fig.cap=size_cap, fig.height=8, fig.width=5}
 size_plot
@@
 ```{r}
-cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
+s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
@@
 
-```{r surv-spei, fig.cap=cap, fig.height=5, fig.width=7}
+```{r surv-spei, fig.cap=s_spei_cap, fig.height=5, fig.width=7}
 s_spei_plot
@@
 ```{r}
-cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
+s_spei_diff_cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
 ```
 
-```{r surv-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+```{r surv-diff, fig.cap=s_spei_diff_cap, fig.height=2.5, fig.width=7}
 s_spei_diff_plot
@@
 ```{r}
-cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
+g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
@@
 
-```{r growth-spei, fig.cap=cap, fig.height=5, fig.width=7}
+```{r growth-spei, fig.cap=g_spei_cap, fig.height=5, fig.width=7}
 g_spei_plot
@@
 ```{r}
-cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
+g_spei_diff_cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
 ```
 
-```{r growth-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+```{r growth-diff, fig.cap=g_spei_diff_cap, fig.height=2.5, fig.width=7}
 g_spei_diff_plot
@@
 ```{r}
-cap <- "Smooth effect of lagged SPEI on flowering probability.
+f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability.
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
@@
 
-```{r flwr-spei, fig.cap=cap, fig.height=5, fig.width=7}
+```{r flwr-spei, fig.cap=f_spei_cap, fig.height=5, fig.width=7}
 f_spei_plot
@@
 ```{r}
-cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
+f_spei_diff_cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
 ```
 
-```{r flwr-diff, fig.cap=cap, fig.height=2.5, fig.width=7}
+```{r flwr-diff, fig.cap=f_spei_diff_cap, fig.height=2.5, fig.width=7}
 f_spei_diff_plot
@@
 
-# Temporal variation in demographic responses to forest fragmentation
+## Temporal variation in demographic responses to forest fragmentation
 
@@
 
-# Delayed effects of climate on demographic vital rates
+## Delayed effects of climate on demographic vital rates
 
@@
 
-# Conclusions & Future Directions
+## Conclusions & Future Directions
 
@@
 
+# Figures
+
+```{r}
+
+```
+
 # References
@@
 
-### Left-over text
+# Left-over text
 
@@
 
-#### Citations / information could allocate to appropriate paragraphs
+# Citations / information could allocate to appropriate paragraphs
 

### 2021-05-04 doc/paper.Rmd
fixed some typos

--- 2021-05-04 doc/paper.Rmd
@@
 Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
+However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
 This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
@@
 
-Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
-While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
+Climate change can alter the demography of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
+While the consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about how climate influences the demography of species in Amazonia and other tropical biomes [@paniw2021].
 Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
 These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @brunaExperimentalAssessmentHeliconia2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @bruna2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@laurance2001; @opdam2004, @selwood2015].
 
@@
 Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009] <!--# Emilio had written "Gaoue" as potential reference, but the on I have in Zotero is about trees, not understory plants -->.
-We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
 This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
@@
 To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
@@
 
-These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
+These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
@@
 These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+There is also evidence that the effects can persist for multiple years [ @phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
 
@@
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
 
@@
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
+For example, although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
 However, the magnitude of these declines was far greater in forest fragments.
@@
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
-The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
+The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@evers2021], and hence be shaped by climatic events at any point in that physiological window.
 For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
@@
 Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
-While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
+While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
 In fact, flowering in one year is associated with increased reproduction and growth in the next.
@@
 Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
-The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
-This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
+The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimeová2018].
+This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@bruna2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
 This would also be consistent with the results of prior experiments, in which the growth rates of *H.
-acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@brunaCompensatoryResponsesUnderstory2005].
+acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@bruna2005].
 
-Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021].
-For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
+Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
+For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
 This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
 Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
-While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
-\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017)
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @bruna2005a; @markewitz2010; @westerband2015; @westerbandEarlyLifeConditions2017)
 
@@
 
-Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @eversLaggedDormantSeason2021].
-Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
+Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @evers2021].
+Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@bruna2005].
 Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

### 2021-05-04 doc/paper.Rmd
tweak

@@
     bookdown::word_document2:
+      linenumbers: yes
       number_sections: no
@@
 Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
-This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\
-
-1)  Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
-
-2)  Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
-
-3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@bruna2009].
+We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
+This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions: (1) Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
+(2) Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
+(3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
 
@@
 
-```{r size-plot, fig.cap=cap, fig.height=8, fig.width=5}
+```{r size-plot, fig.cap=cap, fig.height=5, fig.width=7}
 size_plot

### 2021-05-04 doc/paper.Rmd
fixed some references

--- 2021-05-04 doc/paper.Rmd
@@
 	comment = "#>",
-	# fig.show = "hide" #hide all figures until the end
+	fig.show = "hide", #don't knit figures for now.
 	dpi = 300
@@
 While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
-Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015a].
+Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015].
 These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
@@
 
-```{r eda, fig.cap=eda_cap, fig.height=7, fig.width=10}
+```{r eda, fig.cap=eda_cap}
 eda_plot
@@
 
-```{r size-plot, fig.cap=size_cap, fig.height=8, fig.width=5}
+```{r size-plot, fig.cap=size_cap, fig.height=6, fig.width=5}
 size_plot
@@
 Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
-These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
+These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
 There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
@@
 Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
-Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@laurance2014].
+Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.

### 2021-05-04 doc/paper.Rmd
more fixing of citations

@@
 These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-These results, coupled with evidence that growth and survivorship are already lower in fragments [@laurance1998; @brunaExperimentalAssessmentHeliconia2002; @zartman2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].
 
@@
 where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
-These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see @woodChapterSmoothers and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
+These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
@@
 These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
 
@@
 
-Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumberg2018; @eversLaggedDormantSeason2021].
+Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021].
 Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].

### 2021-05-05 doc/paper.Rmd
Moved figures and captions to end

@@
 	comment = "#>",
-	fig.show = "hide", #don't knit figures for now.
 	dpi = 300
@@
 In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
-
-```{r}
-normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
-Precipitation (a) shows a marked dry seasons from June through October.
-Temperature (b) varies less throughout the year.
-Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
-Data from Brazilian National Institute of Meterology (INMET)."
-```
-
-```{r normals, fig.cap=normals_cap}
-normals_plot
-```
 
@@
 
-```{r}
-eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
-Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
-(b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
-c) shows the proportion of plants with 4 or more shoots flowering each census.
-In d), monthly 3-month SPEI is plotted.
-Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.
-Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
-```
-
-```{r eda, fig.cap=eda_cap}
-eda_plot
-```
-
-<!--# consider boxplots or median ± IQR for panel a) -->
-
 Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
@@
 
-```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
-Panel d) shows the smoothed density of different plant sizes in the two habitat types.
-95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
-The smooths for 1-ha fragments and continuous forest are fit in separate models.
-The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.
-In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
-```
-
-```{r size-plot, fig.cap=size_cap, fig.height=6, fig.width=5}
-size_plot
-```
-
-```{r}
-s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Contour lines correspond to a change of 0.002.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
-
-```
-
-```{r surv-spei, fig.cap=s_spei_cap, fig.height=5, fig.width=7}
-s_spei_plot
-```
-
 Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
@@
 
-```{r}
-s_spei_diff_cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
-```
-
-```{r surv-diff, fig.cap=s_spei_diff_cap, fig.height=2.5, fig.width=7}
-s_spei_diff_plot
-```
-
 Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
 However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
-
-```{r}
-g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function.
-Contour lines correspond to a change of 0.05.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
-```
-
-```{r growth-spei, fig.cap=g_spei_cap, fig.height=5, fig.width=7}
-g_spei_plot
-```
 
@@
 
-```{r}
-g_spei_diff_cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
-```
-
-```{r growth-diff, fig.cap=g_spei_diff_cap, fig.height=2.5, fig.width=7}
-g_spei_diff_plot
-```
-
 The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
@@
 This is due to reduced growth in fragments under wet conditions.
-
-```{r}
-f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability.
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Contour lines correspond to a change of 0.001.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
-```
-
-```{r flwr-spei, fig.cap=f_spei_cap, fig.height=5, fig.width=7}
-f_spei_plot
-```
 
@@
 That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
-
-```{r}
-f_spei_diff_cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
-```
-
-```{r flwr-diff, fig.cap=f_spei_diff_cap, fig.height=2.5, fig.width=7}
-f_spei_diff_plot
-```
 
@@
 ```{r}
+normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+Precipitation (a) shows a marked dry seasons from June through October.
+Temperature (b) varies less throughout the year.
+Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
+Data from Brazilian National Institute of Meterology (INMET)."
+```
 
+```{r normals, fig.cap=normals_cap}
+normals_plot
+```
+
+```{r}
+eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
+Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
+(b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
+c) shows the proportion of plants with 4 or more shoots flowering each census.
+In d), monthly 3-month SPEI is plotted.
+Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.
+Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
+```
+
+```{r eda, fig.cap=eda_cap}
+eda_plot
+```
+
+```{r}
+size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
+Panel d) shows the smoothed density of different plant sizes in the two habitat types.
+95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
+The smooths for 1-ha fragments and continuous forest are fit in separate models.
+The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.
+In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
+```
+
+```{r size-plot, fig.cap=size_cap, fig.height=6, fig.width=5}
+size_plot
+```
+
+```{r}
+s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
+Contour lines correspond to a change of 0.002.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+
+```
+
+```{r surv-spei, fig.cap=s_spei_cap, fig.height=5, fig.width=7}
+s_spei_plot
+```
+
+```{r}
+s_spei_diff_cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
+```
+
+```{r surv-diff, fig.cap=s_spei_diff_cap, fig.height=2.5, fig.width=7}
+s_spei_diff_plot
+```
+
+```{r}
+g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function.
+Contour lines correspond to a change of 0.05.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+```
+
+```{r growth-spei, fig.cap=g_spei_cap, fig.height=5, fig.width=7}
+g_spei_plot
+```
+
+```{r}
+g_spei_diff_cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
+```
+
+```{r growth-diff, fig.cap=g_spei_diff_cap, fig.height=2.5, fig.width=7}
+g_spei_diff_plot
+```
+
+```{r}
+f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability.
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
+Contour lines correspond to a change of 0.001.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+```
+
+```{r flwr-spei, fig.cap=f_spei_cap, fig.height=5, fig.width=7}
+f_spei_plot
+```
+
+```{r}
+f_spei_diff_cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
+```
+
+```{r flwr-diff, fig.cap=f_spei_diff_cap, fig.height=2.5, fig.width=7}
+f_spei_diff_plot
 ```

### 2021-05-05 doc/paper.Rmd
First pass of revisions

@@
 keywords: |
-  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
+  forest fragmentation; demographic vital rates; *Heliconia acuminata*; drought; ENSO; delayed effects; distributed lag non-linear models
 highlights: |
@@
 
-Highlights: `r rmarkdown::metadata$highlights`
+<!-- Highlights: `r rmarkdown::metadata$highlights` -->
 
@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although the local extinction of species from individual fragments can be common [@haddadHabitatFragmentationIts2015 @fletcher2018], the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
-However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009].
-This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
+Although species in forest fragments can have depressed rates of reproduction or survivorship [\@zartmanDroughtImpactsSurvivorship2015; \@brunaSeedGerminationRainforest1999; \@lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
+However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
+The demographic consequences of habitat fragmentation are thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.
+Despite this, evidence linking changes in population-level demographic responses with changes in environmental conditions remains scarce.
 
-Climate change can alter the demographic of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015 @holyoak2016].
+Climate change has been shown to alter the demography of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015; @holyoak2016].
 While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
-Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013 @duffy2015].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, for which an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particualrly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004, @selwood2015].
+Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
+These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable.
+Plant populations may be particularly sensitive as an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].
 
@@
 The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
-In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure S\_ <!--# map can go in the supplemental materials -->).
+In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried [@bierregaardBiologicalDynamicsTropical1992].
 In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H.
+acuminata,* are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
-Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
-Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}=2$), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
 Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
@@
 
-In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
-The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
+In 1997--1998 we established a series of 5000 m^2^ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
+The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
 Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
-This study uses data from 1-ha fragments and Continuous Forest sites.
+This study uses data from 1-ha fragments and continuous forest sites.
 
-Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
+Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering.
 We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
@@
 
-Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
+Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
 We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
@@
 acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
-SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmmTMPA3B432011]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
-<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->
+Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspirtaion of the previous three months.
+SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
+<!--# add this as a "child document" or whatever -->
 
 ## Statistical Modeling of Vital Rates
-
-<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).
--->
 
@@
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
-The general form of the vital rate models was as follows:
+The general form of the vital rate models was as follows: <!--# add flowering covariate? -->
 
@@
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
-number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
+number of lagged SPEI values for each observation) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex.
 Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
+Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.
 
@@
 The `dlnm` package does not currently allow modeling of factor by smooth interactions.
-This means we cannot \_\_\_\_\_\_\_\_\_\_\_\_\_\_.
+This means we could not include habitat as an interaction term.
 We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
@@
 Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
-For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
+For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
 The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
 Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
-<!--#  I know I'm missing some literature here. Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->
 
@@
 
-Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
+Our dataset comprised `r format(summary_stats$n[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
 When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
-The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
-Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
+The proportion of surviving plants was lowest in the 2003--2004 transition year ($P\_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
+The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
+Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1).
 While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
@@
 Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
-Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
+Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
 Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
-This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
+This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
+Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$).
 While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
-
-<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->
 
@@
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
-This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
-This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
+This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# This requires a supplemental figure, probably.
+A plot of shoots vs.
+log(size) -->.
 Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
@@
 only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
-The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
-<!--# not sure how necessary this is. EB: same-->
-
-<!-- This paragraphis talkng about size, but not really growth -->
 
@@
 within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
-In continuous forest, recent SPEI had a weaker effect with only SPEI values less than -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
+In continuous forest, recent SPEI had a weaker effect with only SPEI \< -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
 However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e.
@@
 However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
-
 Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
@@
 Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
-
 The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
@@
 Drought at lags 7--13 reduced flowering probability.
-The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
+The shape of the crossbasis function was similar in 1-ha fragments, but substantially muted (Figure \@ref(fig:flwr-spei)a).
 That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
-
 Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
@@
 
-FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).
+FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly among plots (p \< 0.001 for the random effect of plot).
+<!--# add sentence about flwr_prev covariate -->
 
@@
 Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-<!--# more here -->.
 Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
@@
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
-every 10.63 years in continuous forest.
+every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
 Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].
@@
 Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
-These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].
+These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
+There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; <!--# emilio had "see also", but not sure why -->@pau2013].
 
 Despite this, models of population dynamics have rarely included the effects of environmental drivers.
-This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
+This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.
+Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
 
@@
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival both habitat types.
-However, the magnitude of these declines was far greater in forest fragments.
-Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
+For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types, the magnitude of these declines was far greater in forest fragments.
+Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
 In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences.
-Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994].
-Interestingly, we observed the opposite effect - drought conditions increased the probability of flowering two years later.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of infloresences.
+Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
+Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
 While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
 In fact, flowering in one year is associated with increased reproduction and growth in the next.
-**(is this an acceptable colloquial way to interpret the covariate effect?)**.
+<!--# need to add to methods and results and double-check this is correct -->
 
-Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [sensu @pumisutapon2012].
+Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [ <!--# emilio had "sensu" here, but not sure what definition that refers to --> @pumisutapon2012].
 The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
-This may be why drought led to delayed increases in growth - by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
+This may be why drought led to delayed increases in growth---by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
 This would also be consistent with the results of prior experiments, in which the growth rates of *H.
@@
 acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
-While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
-\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, <!--# question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?
+--> it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+<!--# hmm..; not sure this is the best way to end this paragraph -->
 
-Finally, demographic delays could be an artefact of the timing of responses vis-à-vis the census date.
+Finally, demographic delays could be an artefact of the timing of responses in relation to the census date.
 If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g.
@@
 ***THIS INFO PROB NEEDS TO BE WORKED IN SOMEWHERE ABOVE***
-
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought (-1.5 \> SPEI \> -2) during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-Two noticeable peaks in the proportion of flowering mature plants occur in severe drought years (2003 and 2007) (Figure \@ref(fig:eda)c).
 
@@
 
-Climate models for the Amazon predict a future of extremes - increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
+Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
 Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [sensu @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017)
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [ <!--# Emilio had "sensu", but not sure what definition it refered to --> @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
 
-Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms (cite `dlnm`).
-They also don't allow for modeling interactions between \_\_\_\_ and plant size; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019], the ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
+Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms.
+This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
+The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
 
@@
 NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM.
-"
+PPS, which develop and compute the TMPA as a contribution to TRMM. "<
+!--# this can go in supplemental -->
 
@@
 Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
-(b) shows the survival curve for the 1998 cohort of plants in both fragments and continuous forest.
+(b) shows the survival curve for the 1998 cohort of plants in fragments and continuous forest.
 c) shows the proportion of plants with 4 or more shoots flowering each census.
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by a cubic regression splines.
+size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by cubic regression splines.
 Panel d) shows the smoothed density of different plant sizes in the two habitat types.

### 2021-05-05 doc/paper.Rmd
argh.  fix line breaks

@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although species in forest fragments can have depressed rates of reproduction or survivorship [\@zartmanDroughtImpactsSurvivorship2015; \@brunaSeedGerminationRainforest1999; \@lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
+Although species in forest fragments can have depressed rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
 However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H.
-acuminata,* are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; <!--# emilio had "see also", but not sure why -->@pau2013].
+There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; <!--# emilio had "see also", but not sure why --> @pau2013].
 
@@
 
-Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [ <!--# emilio had "sensu" here, but not sure what definition that refers to --> @pumisutapon2012].
+Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [<!--# emilio had "sensu" here, but not sure what definition that refers to --> @pumisutapon2012].
 The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [ <!--# Emilio had "sensu", but not sure what definition it refered to --> @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [<!--# Emilio had "sensu", but not sure what definition it refered to --> @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
 
@@
 NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM. "<
-!--# this can go in supplemental -->
+PPS, which develop and compute the TMPA as a contribution to TRMM. "\<
+!--\# this can go in supplemental --
+\>
 

### 2021-05-05 doc/paper.Rmd
switched to use models with covariate of flowering in previous year.  Closes #38

@@
 \begin{split}
-& g[ \textrm{E}(y_{i})] = \alpha_j + s_1(x_i) + s_2(d_{i,t}, ..., d_{i,t-36}) \\
+& g[ \textrm{E}(y_{i})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
 & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot }j 
@@
 
-where $s_1(x_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+where $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
 2017 for details).
@@
 We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
+We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$.
 
@@
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
-Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1).
+Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
 While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
@@
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$).
+As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
 While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
@@
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
-This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# This requires a supplemental figure, probably.
+This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# This requires a supplemental figure, probably.
 A plot of shoots vs.
@@
 
-Drought history had a significant effect on survival in both habitats (P \< 0.001), but the shape of the crossbasis surface differed.
+Drought history had a significant effect on survival in both habitats ($p < 0.001$), but the shape of the crossbasis surface differed.
 For 1-ha fragments, recent SPEI values (i.e.
@@
 
-Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
-The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+Drought history also had a significant effect on plant size in both habitats ($p < 0.001$).
+The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
 Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
@@
 
-Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
+Drought history had a significant effect on flowering probability in both habitat types ($p < 0.001$).
 Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
-The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
+The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.1, continuous forest edf = 10.3).
 In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, two dry seasons ago).
@@
 
-FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly among plots (p \< 0.001 for the random effect of plot).
-<!--# add sentence about flwr_prev covariate -->
+FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
+We did not see evidence for a cost of reproduction.
+Flowering in the previous year was a significant predictor of growth ($p = 0.048$ for continuous forest, $p = 0.031$ for fragments) and flowering ($p < 0.001$ in both continuous forest and fragments), but the coefficients were positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
 
@@
 In fact, flowering in one year is associated with increased reproduction and growth in the next.
-<!--# need to add to methods and results and double-check this is correct -->
 

### 2021-05-05 _targets.R

@@ -64,27 +64,18 @@ tar_plan(
   tar_render(validate_data, "doc/validate_data.Rmd"),
   
   # Fit demographic models
-  s_cf = fit_surv(model_data_cf),
-  s_1ha = fit_surv(model_data_1ha),
-  g_cf = fit_growth(model_data_cf),
-  g_1ha = fit_growth(model_data_1ha),
-  f_cf = fit_flwr(model_data_cf),
-  f_1ha = fit_flwr(model_data_1ha),
   
-  ## With flowering in previous year as covariate
-  
-  s_cf_flwr = fit_surv(model_data_cf, flwr_prev = TRUE),
-  s_1ha_flwr= fit_surv(model_data_1ha, flwr_prev = TRUE),
-  g_cf_flwr = fit_growth(model_data_cf, flwr_prev = TRUE),
-  g_1ha_flwr= fit_growth(model_data_1ha, flwr_prev = TRUE),
-  f_cf_flwr = fit_flwr(model_data_cf, flwr_prev = TRUE),
-  f_1ha_flwr= fit_flwr(model_data_1ha, flwr_prev = TRUE),
+  s_cf = fit_surv(model_data_cf, flwr_prev = TRUE),
+  s_1ha = fit_surv(model_data_1ha, flwr_prev = TRUE),
+  g_cf  = fit_growth(model_data_cf, flwr_prev = TRUE),
+  g_1ha = fit_growth(model_data_1ha, flwr_prev = TRUE),
+  f_cf  = fit_flwr(model_data_cf, flwr_prev = TRUE),
+  f_1ha = fit_flwr(model_data_1ha, flwr_prev = TRUE),
   
   # Validate and summarize results
   tar_render(validate_models, "doc/validate_models.Rmd"),
   tar_render(model_summary, "doc/model_summary.Rmd"),
-  tar_render(cost_of_reproduction, "doc/cost_of_reproduction.Rmd"),
-  
+
   # Descriptive / Exploratory Data Analysis Figures
   normals = normals_data(),
   normals_plot = plot_normals(normals),

### 2021-05-06 doc/paper.Rmd

@@
 Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
-Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009] <!--# Emilio had written "Gaoue" as potential reference, but the on I have in Zotero is about trees, not understory plants -->.
+Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009].
 We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
-This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\
-
-1)  Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
-
-2)  Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
-
-3)  Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
+This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions: (1) Does drought increase or decrease the growth, survival, and fertility of plant populations in continuous forest?
+(2) Are there delayed effects of drought on demographic vital rates, and if so what lag times are most critical?
+(3) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest?
 
@@
 
-<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->
-
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
-The general form of the vital rate models was as follows: <!--# add flowering covariate? -->
+The general form of the vital rate models was as follows:
 

### 2021-05-06 doc/paper.Rmd
add "sensu" back

@@
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
-This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
+This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size <!--# reference supplemental about plant size? -->.
 A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
@@
 
-Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [<!--# emilio had "sensu" here, but not sure what definition that refers to --> @pumisutapon2012].
+Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [*sensu* @pumisutapon2012].
 The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [<!--# Emilio had "sensu", but not sure what definition it refered to --> @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
+Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
 

### 2021-05-06 doc/SPEI_appendix.Rmd

@@
 source(here::here("packages.R"))
+source(here::here("R", "utils.R"))
 ```
 
-*Last compiled: `r Sys.Date()`* \# TODO
+*Last compiled: `r Sys.Date()`*
+
+# TODO
 
 -   Write brief text
--   Possibly make a figure S2 that compares SPEI across BDFFP from east to west (3 panels), using all data with similar resolution
+-   Convert table to `kableExtra` so it knits in word and pdf
 
@@
 # knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
-table
+table #only works when knit to HTML
 ```

### 2021-05-06 doc/SPEI_appendix.Rmd
migrate `gt` table to `flextable`.  Now knits to Word.  Closes #39

@@
 
+# Normals
+
+```{r}
+normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+Precipitation (a) shows a marked dry seasons from June through October.
+Temperature (b) varies less throughout the year.
+Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
+Data from Brazilian National Institute of Meterology (INMET)."
+```
+
+```{r normals, fig.cap=normals_cap}
+withr::with_dir(here::here(), {
+  tar_load(normals_plot)
+})
+normals_plot & theme(text = element_text(size = 10))
+```
+
 *Last compiled: `r Sys.Date()`*
+
+If TRMM data is used, they suggest: "The TMPA data were provided by the\
+NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
+PPS, which develop and compute the TMPA as a contribution to TRMM. "\<
+!--\# this can go in supplemental --
+\>
 
@@
          BDFFP = glue("{round(spei_bdffp, 2)} ({month_bdffp})")) %>% 
+  mutate(TRMM = ifelse(TRMM == "NA (NA)", "–", TRMM)) %>% 
   ungroup()
@@
 spei_table <- left_join(spei_table, citations)
-```
-
-```{r}
-library(gt)
-table <- 
-  gt(spei_table) %>% 
-  tab_style(
-    style = list(
-      cell_fill(color = "#ffffb2")
-    ),
-    locations = list(
-      cells_body(
-        columns = vars(`Global SPEI`),
-        rows = spei_g < 0 & spei_g >= -1
-      ),
-      cells_body(
-        columns = vars(EMBRAPA),
-        rows = spei_st < 0 & spei_st >= -1
-      ),
-      cells_body(
-        columns = vars(`Xavier et al.`),
-        rows = spei_xa < 0 & spei_xa >= -1
-      ),
-      cells_body(
-        columns = vars(TRMM),
-        rows = spei_trmm < 0 & spei_trmm >= -1
-      ),
-      cells_body(
-        columns = vars(BDFFP),
-        rows = spei_bdffp < 0 & spei_bdffp >= -1
-      )
-    )
-  ) %>% 
-  tab_style(
-    style = list(
-      cell_fill(color = "#fecc5c")
-    ),
-    locations = list(
-      cells_body(
-        columns = vars(`Global SPEI`),
-        rows = spei_g < -1 & spei_g >= -1.5
-      ),
-      cells_body(
-        columns = vars(EMBRAPA),
-        rows = spei_st < -1 & spei_st >= -1.5
-      ),
-      cells_body(
-        columns = vars(`Xavier et al.`),
-        rows = spei_xa < -1 & spei_xa >= -1.5
-      ),
-      cells_body(
-        columns = vars(TRMM),
-        rows = spei_trmm < -1 & spei_trmm >= -1.5
-      ),
-      cells_body(
-        columns = vars(BDFFP),
-        rows = spei_bdffp < -1 & spei_bdffp >= -1.5
-      )
-    )
-  ) %>% 
-  tab_style(
-    style = list(
-      cell_fill(color = "#fd8d3c")
-    ),
-    locations = list(
-      cells_body(
-        columns = vars(`Global SPEI`),
-        rows = spei_g < -1.5 & spei_g >= -2
-      ),
-      cells_body(
-        columns = vars(EMBRAPA),
-        rows = spei_st < -1.5 & spei_st >= -2
-      ),
-      cells_body(
-        columns = vars(`Xavier et al.`),
-        rows = spei_xa < -1.5 & spei_xa >= -2
-      ),
-      cells_body(
-        columns = vars(TRMM),
-        rows = spei_trmm < -1.5 & spei_trmm >= -2
-      ),
-      cells_body(
-        columns = vars(BDFFP),
-        rows = spei_bdffp < -1.5 & spei_bdffp >= -2
-      )
-    )
-  ) %>% 
-  tab_style(
-    style = list(
-      cell_fill(color = "#e31a1c")
-    ),
-    locations = list(
-      cells_body(
-        columns = vars(`Global SPEI`),
-        rows = spei_g < -2
-      ),
-      cells_body(
-        columns = vars(EMBRAPA),
-        rows = spei_st < -2
-      ),
-      cells_body(
-        columns = vars(`Xavier et al.`),
-        rows = spei_xa < -2
-      ),
-      cells_body(
-        columns = vars(TRMM),
-        rows = spei_trmm < -2
-      ),
-      cells_body(
-        columns = vars(BDFFP),
-        rows = spei_bdffp < -2
-      )
-    )
-  ) %>% 
-  cols_hide(starts_with(c("spei_", "month_"))) %>% 
-  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM, BDFFP))
-#TODO: add citations to this table as footnotes on `cause` or a new column.
 ```
@@
 
+```{r}
+library(flextable)
+set_flextable_defaults(big.mark = "")
+table <- 
+  spei_table %>% 
+  # select(-starts_with("month_"), -starts_with("spei_")) %>% 
+  flextable(
+    col_keys = c("year", "cause", "Global SPEI", "Xavier et al.", "EMBRAPA", "TRMM", "BDFFP")
+  ) %>% 
+  add_header_row(colwidths = c(1,1, 5),
+                 values = c("", "", "Lowest SPEI for the year (month)")) %>% 
+  align(i = 1, part = "header", align = "center") %>% 
+  #extreme drought
+  bg(~ spei_g < -2, ~`Global SPEI`, bg = "#e31a1c") %>% 
+  bg(~ spei_xa < -2, ~ `Xavier et al.`, bg = "#e31a1c") %>% 
+  bg(~ spei_st < -2, ~ EMBRAPA, bg = "#e31a1c") %>% 
+  bg(~spei_trmm < -2, ~TRMM, bg = "#e31a1c") %>% 
+  bg(~ spei_bdffp < -2, ~BDFFP, bg = "#e31a1c") %>%
+
+  #severe drought
+  bg(~ between(spei_g,     -2, -1.5), ~ `Global SPEI`, bg = "#fd8d3c") %>% 
+  bg(~ between(spei_xa,    -2, -1.5), ~ `Xavier et al.`, bg = "#fd8d3c") %>% 
+  bg(~ between(spei_st,    -2, -1.5), ~ EMBRAPA, bg = "#fd8d3c") %>% 
+  bg(~ between(spei_trmm,  -2, -1.5), ~ TRMM, bg = "#fd8d3c") %>% 
+  bg(~ between(spei_bdffp, -2, -1.5), ~ BDFFP, bg = "#fd8d3c") %>%
+
+  #moderate drought
+  bg(~ between(spei_g,     -1.5, -1), ~ `Global SPEI`, bg = "#fecc5c") %>% 
+  bg(~ between(spei_xa,    -1.5, -1), ~ `Xavier et al.`, bg = "#fecc5c") %>% 
+  bg(~ between(spei_st,    -1.5, -1), ~ EMBRAPA, bg = "#fecc5c") %>% 
+  bg(~ between(spei_trmm,  -1.5, -1), ~ TRMM, bg = "#fecc5c") %>% 
+  bg(~ between(spei_bdffp, -1.5, -1), ~ BDFFP, bg = "#fecc5c") %>%
+
+  #mild drought
+  bg(~ between(spei_g,     -1, 0), ~ `Global SPEI`, bg = "#ffffb2") %>% 
+  bg(~ between(spei_xa,    -1, 0), ~ `Xavier et al.`, bg = "#ffffb2") %>% 
+  bg(~ between(spei_st,    -1, 0), ~ EMBRAPA, bg = "#ffffb2") %>% 
+  bg(~ between(spei_trmm,  -1, 0), ~ TRMM, bg = "#ffffb2") %>% 
+  bg(~ between(spei_bdffp, -1, 0), ~ BDFFP, bg = "#ffffb2")
+```
+
 ```{r spei-table, fig.cap=cap}
-# knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
-table #only works when knit to HTML
+table %>%
+  set_table_properties(layout = "autofit") %>%
+  align(align = "center", j = 3:6) %>%
+  border_remove() %>% 
+  hline_bottom() %>% 
+  hline_top() %>% 
+  hline(j = 3:6, part = "header") %>% 
+  footnote(j = 2, i = c(1, 4, 7), value = as_paragraph(
+    c("McPhaden 1999",
+      "Marengo et al. 2008; Zeng et al. 2008",
+      "Lewis et al. 2011")),
+    ref_symbols = c("a", "b", "c"),
+    part = "body", inline = TRUE
+  )
 ```

### 2021-05-07 doc/paper.Rmd
first pass at Maria's revisions.  Mostly changes to text and references and figure formatting

@@
 However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
-The demographic consequences of habitat fragmentation are thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
+The demographic consequences of habitat fragmentation are thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], which continue to be cleared and fragmented at unprecedented rates.
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Despite this, evidence linking changes in population-level demographic responses with changes in environmental conditions remains scarce.
+Despite this, evidence linking changes in population-level demographic responses with altered environmental conditions remains scarce.
 
@@
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivelmuelbert2019; @gonzalez-mDivergingFunctionalStrategies2020], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory.
+While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H.
+acuminata,* are found at lower densities in the darker and cooler forest understory.
 These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
@@
 data]*.
+<!--# maria: not clear why this is relevant <!--#  --> --\>
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 This study uses data from 1-ha fragments and continuous forest sites.
+<!--# maria: justify why 10ha plots omitted or don't say they exist -->
 
@@
 acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
-Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspirtaion of the previous three months.
+Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
 SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
@@
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
-acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
+acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
 
 The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
-The general form of the vital rate models was as follows:
+The general form of the vital rate ($y$) models was as follows:
 
@@
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
-number of lagged SPEI values for each observation) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex.
-Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
+number of lagged SPEI values for each observation with 36 months as a maximum lag) and 3 knots for the drought response dimension to restrict the shape of the fitted response to drought to bimodal when most complex.
+Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
 Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.
@@
 A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
-For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
+Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
 We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
@@
                    .fns = list(mean = mean, sd = sd),
-                   na.rm = TRUE)) %>%
-  right_join(model_data %>% count(habitat)) %>%
+                   na.rm = TRUE),
+            n_plants = length(unique(ha_id_number))) %>%
+  # right_join(model_data %>% count(habitat)) %>%
   mutate(across(ends_with("_mean") | ends_with("_sd"),
@@
 
-Our dataset comprised `r format(summary_stats$n[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
+Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
 When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
@@
 Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
+Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
@@
 Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
-These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@conditTropicalForestDynamics2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
+These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
 There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; <!--# emilio had "see also", but not sure why --> @pau2013].
 
-Despite this, models of population dynamics have rarely included the effects of environmental drivers.
+Despite this knowledge, models of plant population dynamics have rarely included the effects of environmental drivers.
 This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
-To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+To address this challenge, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
 Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
 
-However, while it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
+While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
@@
 
+<!--# maria suggests moving normals figure to appendix -->
+
 ```{r}
@@
 ```
+
+<!--# standard deviation not shown in a), but mentioned in caption. Why? -->
+
+<!--# Maria would prefer survival rates for all plants rather than tracking 1998 cohort survivorship -->
 
@@
 In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
+
 ```

### 2021-05-07 R/plot_smooths.R

@@ -126,7 +126,7 @@ plot_spei_heatmap <-
     
     ggplot(eval_df, aes_string(y = "spei_history", x = "L")) +
       geom_raster(aes_string(fill = "est")) +
-      geom_contour(aes_string(z = "est"), color = "black", binwidth = binwidth) +
+      geom_contour(aes_string(z = "est"), color = "black", binwidth = binwidth, size = 0.3) +
       geom_hline(aes(yintercept = 0), color = "grey", linetype = 2) +
       scale_fill_viridis_c(response_lab, option = "viridis", limits = fill_lims) +
       scale_x_continuous("lag (months before census)",

### 2021-05-07 doc/paper.Rmd
change figure size and font size to be more consistent

@@
 ```{r normals, fig.cap=normals_cap}
-normals_plot
+normals_plot & theme(text = element_text(size = 10))
 ```
@@
 
-```{r eda, fig.cap=eda_cap}
-eda_plot
+```{r eda, fig.cap=eda_cap, fig.width=7, fig.height=7}
+eda_plot & theme(text = element_text(size = 10))
 ```
@@
 
-```{r size-plot, fig.cap=size_cap, fig.height=6, fig.width=5}
-size_plot
+```{r size-plot, fig.cap=size_cap, fig.width=7, fig.height=7}
+size_plot & theme(text = element_text(size = 10))
 ```
@@
 ```{r surv-spei, fig.cap=s_spei_cap, fig.height=5, fig.width=7}
-s_spei_plot
+s_spei_plot & theme(text = element_text(size = 10))
 ```
@@
 ```{r surv-diff, fig.cap=s_spei_diff_cap, fig.height=2.5, fig.width=7}
-s_spei_diff_plot
+s_spei_diff_plot  & theme(text = element_text(size = 10))
 ```
@@
 ```{r growth-spei, fig.cap=g_spei_cap, fig.height=5, fig.width=7}
-g_spei_plot
+g_spei_plot & theme(text = element_text(size = 10))
 ```
@@
 ```{r growth-diff, fig.cap=g_spei_diff_cap, fig.height=2.5, fig.width=7}
-g_spei_diff_plot
+g_spei_diff_plot & theme(text = element_text(size = 10))
 ```
@@
 ```{r flwr-spei, fig.cap=f_spei_cap, fig.height=5, fig.width=7}
-f_spei_plot
+f_spei_plot & theme(text = element_text(size = 10))
 ```
@@
 ```{r flwr-diff, fig.cap=f_spei_diff_cap, fig.height=2.5, fig.width=7}
-f_spei_diff_plot
+f_spei_diff_plot & theme(text = element_text(size = 10))
 ```

### 2021-05-07 doc/paper.Rmd
Combine difference in smooths into third panel of main figure for each vital rate. Not super happy with guide alignment, but closes #43 for now.

@@
     s_spei_plot,
-    s_spei_diff_plot,
     g_spei_plot,
-    g_spei_diff_plot,
     f_spei_plot,
-    f_spei_diff_plot,
     size_plot,
@@
 However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
-Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-diff)).
+Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
 However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
@@
 Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
-The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-diff).
+The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-spei)c.
 However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
@@
 That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
-Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-diff)).
+Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-spei)c).
 The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.
@@
 s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
+Panel c) shows the difference between panels a) and b).
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
@@
 
-```{r surv-spei, fig.cap=s_spei_cap, fig.height=5, fig.width=7}
+```{r surv-spei, fig.cap=s_spei_cap, fig.height=7.5, fig.width=7}
 s_spei_plot & theme(text = element_text(size = 10))
@@
 ```{r}
-s_spei_diff_cap <- "The difference between predicted effects of SPEI history on survival in continuous forest and 1-ha fragments."
-```
-
-```{r surv-diff, fig.cap=s_spei_diff_cap, fig.height=2.5, fig.width=7}
-s_spei_diff_plot  & theme(text = element_text(size = 10))
-```
-
-```{r}
 g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
+Panel c) shows the difference between panels a) and b).
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
@@
 
-```{r growth-spei, fig.cap=g_spei_cap, fig.height=5, fig.width=7}
+```{r growth-spei, fig.cap=g_spei_cap, fig.height=7.5, fig.width=7}
 g_spei_plot & theme(text = element_text(size = 10))
@@
 ```{r}
-g_spei_diff_cap <- "The difference between predicted effects of SPEI history on plant size in continuous forest and 1-ha fragments."
-```
-
-```{r growth-diff, fig.cap=g_spei_diff_cap, fig.height=2.5, fig.width=7}
-g_spei_diff_plot & theme(text = element_text(size = 10))
-```
-
-```{r}
-f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability.
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability for a) continuous forest and b) 1-ha fragments.
+Panel c) shows the difference between panels a) and b).Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
 Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
@@
 
-```{r flwr-spei, fig.cap=f_spei_cap, fig.height=5, fig.width=7}
+```{r flwr-spei, fig.cap=f_spei_cap, fig.height=7.5, fig.width=7}
 f_spei_plot & theme(text = element_text(size = 10))
-```
-
-```{r}
-f_spei_diff_cap <- "The difference between predicted effects of SPEI history on flowering probability in continuous forest and 1-ha fragments."
-```
-
-```{r flwr-diff, fig.cap=f_spei_diff_cap, fig.height=2.5, fig.width=7}
-f_spei_diff_plot & theme(text = element_text(size = 10))
 ```

### 2021-05-07 R/plot_smooths.R

@@ -152,7 +152,7 @@ plot_spei_heatmap <-
 #' @param response_lab response label (for color bar)
 #' @param binwidth binwidth for contour lines
 #'
-plot_cb_2panel <-
+plot_cb_3panel <-
   function(cf_model, frag_model, smooth = "spei_history", response_lab, binwidth) {
     df_cf <- my_eval_smooth(cf_model, smooth, dist = 0.1)
     df_frag <- my_eval_smooth(frag_model, smooth, dist = 0.1)
@@ -165,49 +165,37 @@ plot_cb_2panel <-
         df_frag,
         binwidth = binwidth,
         fill_lims = fill_lims,
-        response_lab = response_lab
+        response_lab = TeX(response_lab)
       )
     cf_plot <-
       plot_spei_heatmap(
         df_cf,
         binwidth = binwidth,
         fill_lims = fill_lims,
-        response_lab = response_lab
+        response_lab = TeX(response_lab)
       )
     
-    (cf_plot + theme(axis.title.x = element_blank())) /
-      (frag_plot) +
-      plot_layout(guides = "collect") +
-      plot_annotation(tag_levels = "a", tag_suffix = ")") &
-      labs(title = NULL) #remove plot titles
+    diff_df <-
+      bind_cols(
+        df_cf %>% rename_with(.fn = ~glue("cf_{.}")),
+        df_frag %>% rename_with(.fn = ~glue("frag_{.}"))
+      ) %>%
+      mutate(est = cf_est - frag_est) %>%
+      select(est, L = cf_L, spei_history = cf_spei_history)
     
+    diff_plot <-
+      plot_spei_heatmap(
+        diff_df,
+        fill_lims = NULL,
+        binwidth = binwidth,
+        response_lab = ""
+      ) +
+      scale_fill_distiller(TeX(glue::glue("$\\Delta${response_lab} (CF-1ha)")), palette = "PuOr")
+    
+    cf_plot / 
+      (frag_plot + theme(legend.position = "none")) /
+      diff_plot + plot_layout(guides = "keep") &
+      theme(legend.justification = "left") &
+      plot_annotation(tag_levels = "a", tag_suffix = ")") &
+      theme(legend.justification = "left")
   }
-
-#' Plot difference between evaluated crossbasis smooths in two habitats
-#'
-#' @param cf_model model object for continuous forest
[…]

### 2021-05-07 _targets.R

@@ -93,22 +93,19 @@ tar_plan(
   ## Survival
   s_covar_plot = plot_covar_smooth(frag_model = s_1ha, cf_model = s_cf, covar = "log_size_prev") + 
                   labs(x = TeX("$log(size_t)$"), y = "P(survival)"),
-  s_spei_plot = plot_cb_2panel(s_cf, s_1ha, binwidth = 0.002, response_lab = "P(survival)"),
-  s_spei_diff_plot = plot_cb_diff(s_cf, s_1ha, binwidth = 0.002, response_lab = "∆P(survival) (CF – 1ha)"),
+  s_spei_plot = plot_cb_3panel(s_cf, s_1ha, binwidth = 0.002, response_lab = "P(survival)"),
 
   ## Growth
   g_covar_plot = plot_covar_smooth(frag_model = g_1ha, cf_model = g_cf, covar = "log_size_prev") +
                   geom_abline(slope = 1, color = "grey50") +
                   labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})")),
-  g_spei_plot = plot_cb_2panel(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$log(size_{t+1})$")),
-  g_spei_diff_plot = plot_cb_diff(g_cf, g_1ha, binwidth = 0.05, response_lab = TeX("$\\Delta log(size_{t+1})$ (CF – 1ha)")),
+  g_spei_plot = plot_cb_3panel(g_cf, g_1ha, binwidth = 0.05, response_lab = "$log(size_{t+1})$"),
 
   ## Flowering
   f_covar_plot = plot_covar_smooth(frag_model = f_1ha, cf_model = f_cf, covar = "log_size_prev") +
                   labs(x = TeX("$log(size_t)$"), y = "P(flowering)"),
-  f_spei_plot = plot_cb_2panel(f_cf, f_1ha, binwidth = 0.001, response_lab = "P(flowering)"),
-  f_spei_diff_plot = plot_cb_diff(f_cf, f_1ha, binwidth = 0.001, response_lab = "∆Flowering (CF – 1ha)"),
-  
+  f_spei_plot = plot_cb_3panel(f_cf, f_1ha, binwidth = 0.001, response_lab = "P(flowering)"),
+
   ## Size covariate
   
   size_plot = make_size_plot(s = s_covar_plot, g = g_covar_plot, f = f_covar_plot, model_data = model_data),

### 2021-05-10 doc/paper.Rmd
got equation numbering and in-text references working using pandoc-crossref

@@
       - --lua-filter=templates/pagebreak.lua
+      - -Fpandoc-crossref
 linkcolor: blue
@@
   \end{split}
-$$
+$${\#eq:eqn1}
 
@@
 s_2(d_{i,t}, ..., d_{i,t-24}) = \sum_{l = l_0}^L f \cdot w(d_{i,t-l},l) 
-$$
+$${\#eq:eqn2}
 
@@
 
-To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
+To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.

### 2021-05-11 doc/paper.Rmd
update eda figure caption

@@
 eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
-Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (red) than continuous forest (blue).
-(b) shows the survival curve for the 1998 cohort of plants in fragments and continuous forest.
-c) shows the proportion of plants with 4 or more shoots flowering each census.
-In d), monthly 3-month SPEI is plotted.
-Light grey lines represent values from different grid cells encompassing BDFFP and the dark line is the site mean.
+Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed).
+The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census.
+The proportion of reproductive sized plants (c) is on average low and fluctuates substantially year to year.
+The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years.
+Monthly 3-month SPEI is shown (d) with grey lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
 Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
 ```
-
-<!--# standard deviation not shown in a), but mentioned in caption. Why? -->
-
-<!--# Maria would prefer survival rates for all plants rather than tracking 1998 cohort survivorship -->
 

### 2021-05-11 doc/SPEI_appendix.Rmd
move normals plot to supplemental

--- 2021-05-06 doc/SPEI_appendix.Rmd
@@
 source(here::here("R", "utils.R"))
+```
+
+# Normals
+
+```{r}
+normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
+Precipitation (a) shows a marked dry seasons from June through October.
+Temperature (b) varies less throughout the year.
+Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
+Data from Brazilian National Institute of Meterology (INMET)."
+```
+
+```{r normals, fig.cap=normals_cap}
+withr::with_dir(here::here(), {
+  tar_load(normals_plot)
+})
+normals_plot & theme(text = element_text(size = 10))
 ```

### 2021-05-11 doc/paper.Rmd
move normals plot to supplemental

@@
   tar_load(c(
-    normals_plot,
     eda_plot,
@@
 Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
-There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).
+There is a pronounced dry season from June to October (Figure S1 <!--# update figure number manually -->).
 
@@
 ```{r}
-normals_cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010.
-Precipitation (a) shows a marked dry seasons from June through October.
-Temperature (b) varies less throughout the year.
-Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
-Data from Brazilian National Institute of Meterology (INMET)."
-```
-
-```{r normals, fig.cap=normals_cap}
-normals_plot & theme(text = element_text(size = 10))
-```
-
-<!--# maria suggests moving normals figure to appendix -->
-
-```{r}
 eda_cap <- "Timeseries of demographic parameters and drought ocurrence.

### 2021-05-11 doc/paper.Rmd
create stand-alone plot of survivorship curve

@@
 ```{r load-targets}
-withr::with_dir(here(), {
+withr::with_dir(here::here(), {
   tar_load(c(
     eda_plot,
+    surv_curve,
     s_spei_plot,
@@
 ```{r}
+surv_curve_cap <- "Survivorship of plants labeled in the first survey year, 1998, which comprise 49% of the plants in the full dataset.
+After 10 years, 79.7% (1629/2055) of plants in continuous forest survived and 72.4% (393/543) of plants in fragments survived."
+```
+
+```{r surv_curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
+surv_curve + theme(text = element_text(size = 10))
+```
+
+```{r}
 size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by cubic regression splines.

### 2021-05-11 R/plot_eda.R

@@ -102,9 +102,7 @@ plot_eda_surv_ts <- function(demog_post, date_lims) {
 #' @param demog demography data
 #' @param date_lims x-axis limits
 #'
-plot_eda_surv_cohort <- function(demog, date_lims) {
-  
-  date_breaks <- seq(date_lims[1], date_lims[2], by = "year")
+plot_eda_surv_cohort <- function(demog) {
   
   cohort <-
     demog %>% 
@@ -124,21 +122,22 @@ plot_eda_surv_cohort <- function(demog, date_lims) {
            yearmonth = yearmonth(date))
   
   surv_curve <-
-    ggplot(surv_curve_df, aes(x = yearmonth, y = p_surv, color = habitat, linetype = habitat)) +
+    ggplot(surv_curve_df, aes(x = date, y = p_surv, color = habitat, linetype = habitat)) +
     geom_step() +
-    scale_x_yearmonth(
-      limits = date_lims,
-      breaks = date_breaks,
+    scale_x_date(
+      "Census Year",
+      date_breaks = "1 year",
+      date_labels = "%Y",
       date_minor_breaks = "1 month",
       expand = expansion(mult = 0.04)
     ) +
     scale_y_continuous("P(survived)") +
+    scale_color_manual(values = c("#E66101", "#5E3C99"),
+                        aesthetics = c("color", "fill")) +
+    guides(col = guide_legend(title = "Habitat"),
+           linetype = guide_legend(title = "Habitat")) +
     theme_classic() +
-    theme(
-      axis.text.x = element_blank(),
-      axis.title.x = element_blank(),
-      axis.ticks.x = element_blank()
-    )
+    theme(axis.text.x = element_text(angle = 45, hjust = 1))
   surv_curve
 }
 

### 2021-05-11 _targets.R

@@ -83,10 +83,11 @@ tar_plan(
   
   eda_spei = plot_eda_spei(xa_lag, plot_dates),
   eda_surv_ts = plot_eda_surv_ts(model_data, plot_dates),
-  eda_surv = plot_eda_surv_cohort(demog_done, plot_dates),
   eda_size = plot_eda_size(model_data, plot_dates),
   eda_flwr = plot_eda_flwr(model_data, plot_dates),
   eda_plot = plot_eda_combine(eda_size, eda_surv_ts, eda_flwr, eda_spei),
+  
+  surv_curve = plot_eda_surv_cohort(demog_done),
   
   # Model output figures
 

### 2021-05-11 doc/paper.Rmd
update text to reflect new figures.  Closes #45

@@
 Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
-When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
-The proportion of surviving plants was lowest in the 2003--2004 transition year ($P\_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
-The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
+When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
+The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
+The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
 Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
@@
 
+```{r}
+#data for eda plot panel c
+model_data %>% 
+  filter(size >= 168) %>% 
+    group_by(year, habitat) %>% 
+    summarize(n = n(), flwr = sum(flwr == 1, na.rm = TRUE)) %>% 
+    mutate(prop_flwr = flwr/n)
+```
+
 While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
-This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# This requires a supplemental figure, probably.
+This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c) <!--# , which corresponds roughly to ____ shoots --This requires a supplemental figure, probably.
 A plot of shoots vs.
 log(size) -->.
-Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
-Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs.
-only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
+Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
+Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
+only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).
 

### 2021-05-11 doc/paper.Rmd
fix citation

@@
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivelmuelbert2019; @gonzalez-mDivergingFunctionalStrategies2020], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].

### 2021-05-12 doc/paper.Rmd
final edits to text. adds a paragraph about microclimate buffering to discussion.

@@
 Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although species in forest fragments can have depressed rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
+Although <!--# plant?
+-->species in forest fragments can have depressed rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
 However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
@@
 data]*.
-<!--# maria: not clear why this is relevant <!--#  --> --\>
+<!--# maria: not clear why this is relevant <!--#  -->
 
@@
 We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
-We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$.
+We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
 
@@
 
-While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
@@
 not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
-Lower SPEI (drought) 15--20 months prior to census was associated with reduced survival and higher SPEI at these lag times was associated with higher survival.
+SPEI \< -1 (drought) 15--20 months prior to census was associated with reduced survival and SPEI \> 1 at these lag times was associated with higher survival (i.e.
+a nearly linear relationship between SPEI and survival probability).
 Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
@@
 Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
-The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-spei)c.
+The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-spei)c).
 However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
@@
 
+<!--# new paragraph below.  What do you think Emilio? -->
+
+Fragmentation of forests diminishes their ability to buffer changes in microclimate due to edge effects [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
+This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. accuminata* in forest fragments in this study.
+We show that the effects of variation in SPEI on survival and growth are more extreme in fragments compared to continuous forest.
+Additionally, the direction of the SPEI effects appear to be seasonal---that is, extreme dry conditions in dry seasons and extreme wet conditions in wet seasons are most detrimental to growth and survival.
+The ability of continuous forest habitat to buffer these extremes may be responsible for the less extreme effects of variation in SPEI in continuous forest.
+More experimental work is needed to establish alterations in microclimate or impaired microclimate buffering in extreme weather events as a causal mechanism for reduced growth and survival of plant individuals in forest fragments.
+
 ## Delayed effects of climate on demographic vital rates
@@
 These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; <!--# emilio had "see also", but not sure why --> @pau2013].
+There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].
 
@@
 For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
-This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
+This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018].
 Similar delayed changes in the local environment could also influence the foraging behavior of *H.
 acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
-While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, <!--# question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?
---> it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
-<!--# hmm..; not sure this is the best way to end this paragraph -->
+While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
 
-Finally, demographic delays could be an artefact of the timing of responses in relation to the census date.
+Finally, demographic delays could be an artifact of the timing of responses in relation to the census date.
 If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g.
@@
 acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.
-
-***THIS INFO PROB NEEDS TO BE WORKED IN SOMEWHERE ABOVE***
-
-Past climate conditions were important for survival, size, and flowering probability.
-Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
-Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
-Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
-For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph?
-Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.
-Might end up integrating with a later paragraph.
--->.
 
@@
 
-We thank ----, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Collin Edwards, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
 We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
@@
 This is publication No. ------ in the BDFFP Technical Series.
-
-If TRMM data is used, they suggest: "The TMPA data were provided by the\
-NASA/Goddard Space Flight Center's Mesoscale Atmospheric Processes Laboratory and\
-PPS, which develop and compute the TMPA as a contribution to TRMM. "\<
-!--\# this can go in supplemental --
-\>
 

### 2021-05-12 doc/paper.Rmd
remove code chunk

@@
 
-```{r}
+```{r include=FALSE}
 #data for eda plot panel c

### 2021-05-12 doc/paper.Rmd
edits (title, cites, intro, 1st paragraph of discussion)

@@
 ---
-title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
+title: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
 author:
@@
 abstract: |
-  Text of abstract
+  Text of abstract. 34,281 year x individual measurments! 
 keywords: |
-  forest fragmentation; demographic vital rates; *Heliconia acuminata*; drought; ENSO; delayed effects; distributed lag non-linear models
+  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
 highlights: |
@@
 
-Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
-Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
-Although <!--# plant?
--->species in forest fragments can have depressed rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
-However, it is often assumed that these detrimental effects are caused by dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
-The demographic consequences of habitat fragmentation are thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], which continue to be cleared and fragmented at unprecedented rates.
+Tropical deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015; @alroy2017].
+Deforestation also results in landscapes where the remaining forest can be highly fragmented, with forest remnants of different sizes embedded in a matrix of often highly-contrasting habitat [@taubert2018; @bianchi2013].
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Despite this, evidence linking changes in population-level demographic responses with altered environmental conditions remains scarce.
+Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006a; @dasilva2000].
+Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999 ; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
+Despite the fact that this mechanism is thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions remains scarce.
 
-Climate change has been shown to alter the demography of species in temperate zones [@doak2010; @sletvold2005; @selwood2015; @williams2015], and these effects can be exacerbated in fragmented landscapes [@oliver2015; @holyoak2016].
+Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
 While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
 Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
-These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable.
-Plant populations may be particularly sensitive as an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-These results, coupled with evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].
+Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].
 
-Whether plant populations in tropical forest fragments are more susceptible to drought remains unclear for three primary reasons.
+Whether the demography of plant populations in tropical forest fragments is more susceptible to drought and other climatic extremes remains unclear for three primary reasons.
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarte2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
 These data are critical not simply because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
-They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
+They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [*e.g.,* @gagnonGrowthUnderstoryHerb2011].
 Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].
@@
 The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
-Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
+Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall ranges from 1900--2300 mm.
 There is a pronounced dry season from June to October (Figure S1 <!--# update figure number manually -->).
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in disturbed areas or treefall gaps, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory.
-These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991].
-While some *Heliconia* species are at least partially self-compatible [@kressSelfincompatibilitySystemsCentral1983], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare *[unpubl.
-data]*.
-<!--# maria: not clear why this is relevant <!--#  -->
+While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H.
+acuminata,* are found at lower densities in the darker and cooler forest understory [@rundel2020].
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996d; @bruna2004].
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 
-Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering.
+Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
+Plants that were not found for three consecutive surveys were considered dead.
 We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
-Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
-Plants that were not found for three consecutive surveys were considered dead.
-There is a seven-fold range in *H.
-acuminata* density in the forest plots, with density in several of the continuous forest plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
+For additional details on the plots, survey methods, and H.
+acuminata populations see @brunaHabitatFragmentationDemographic2002.
 
@@
 
-Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
+Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
+Plots in CF had on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
 When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
@@
 
-Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
-Using a decadal dataset of plant demography, we show that some---but not all---results from short-term studies can be extrapolated to longer times scales.
-Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
-
-Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-Using a decadal dataset of plant demography, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
+Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
+Here we support the emerging generalization that the effects of climate drivers on demographic vital rates can be delayed [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+We also found that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
+Consequently, the hypothesized synergies between climate and fragmentation may be far more complex and pervasive than previously thought.
 
@@
 
+Our results indicate that some---but not all---results from short-term demographic studies can be extrapolated to longer times scales.
 Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
@@
 These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
-However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs.
+However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
+They translate into flowering once every 7.97 years in fragments vs.
 every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
-Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].
+While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundel2020a], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
+data*).
+Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations.
+[@gagnonGrowthUnderstoryHerb2011].
 
@@
 Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
+acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996d] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
 While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

### 2021-05-13 doc/paper.Rmd
edit of results

@@
 Plots in CF had on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
+
 When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
@@
 112 ± 141 SD, respectively).
-This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
+<!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average).
+-->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
@@
 
-Drought history had a significant effect on survival in both habitats ($p < 0.001$), but the shape of the crossbasis surface differed.
-For 1-ha fragments, recent SPEI values (i.e.
-within the past 0--12 months) had the strongest effect on survival, with the highest survival near SPEI of 0 and decreasing survival when plants were exposed to either drier or wetter conditions over the prior 0--12 months (Figure \@ref(fig:surv-spei)a).
-In continuous forest, recent SPEI had a weaker effect with only SPEI \< -1 leading to slightly reduced survival probabilities (Figure \@ref(fig:surv-spei)b).
-However, SPEI at a lag time of 15--20 months, corresponding to June--October of the previous year (i.e.
-not the most recent dry season, but the one before that), also affected survival probabilities in continuous forest.
-SPEI \< -1 (drought) 15--20 months prior to census was associated with reduced survival and SPEI \> 1 at these lag times was associated with higher survival (i.e.
-a nearly linear relationship between SPEI and survival probability).
-Plants in both continuous forest and 1-ha fragments showed an increase in survival probability with very high (wet) SPEI values at a lag time of 36 months.
-However, it should be noted that only the first year of census data (1999) had an SPEI value greater than 2 within 36 months in the past.
-Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
+Drought history had a significant effect on the survival ($p < 0.001$), growth ($p < 0.001$), and flowering ($p < 0.001$) of plants in both habitats.
+Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.
+
+<!--# what do you think of this wording?
+Trying to walk people through the big result by breaking into two sentences telling them what they are about to see --> For 1-ha fragments, recent SPEI values (i.e.
+within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
+<!--# what do you think?
+trying to put the results in more ecological terms) -->In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional -- survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
+Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census).
+Drought 15--20 months prior to a census (i.e., SPEI \< -1) was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, was a nearly linear relationship between SPEI and survival probability).
+Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
+It should be noted, however, that only the first year of census data (1999) met these conditions.
+<!--# can you walk me through these next two sentences?
+I'm not sure if this applies to the 36 mop lag or the whole time-series -->Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
 However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
 
-Drought history also had a significant effect on plant size in both habitats ($p < 0.001$).
-The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.
-The difference between the effects of SPEI history in continuous forest and 1-ha fragments shows that growth is equal or slightly higher in continuous forest for average conditions (SPEI = 0) over all lag periods (Figure \@ref(fig:growth-spei)c).
-However, growth in continuous forests is higher by up to log(size) = 0.57 when the current wet season is unusually wet (SPEI\>2).
-This is due to reduced growth in fragments under wet conditions.
+While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the surface for 1-ha fragments indicated more complex responses in some situations (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
+However, when the current wet season is unusually wet (SPEI\>2), plants in continuous forests were larger by up to log(size) = 0.57 due to reduced growth in fragments.
+In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, but the magnitude of the response was greater in 1-ha fragments.
 
-Drought history had a significant effect on flowering probability in both habitat types ($p < 0.001$).
-Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
-The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.1, continuous forest edf = 10.3).
-In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, two dry seasons ago).
-Drought at lags 7--13 reduced flowering probability.
-The shape of the crossbasis function was similar in 1-ha fragments, but substantially muted (Figure \@ref(fig:flwr-spei)a).
-That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
+Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
+The effects of SPEI on flowering probability in continuous forest were also more complex (1-ha edf = 8.1, continuous forest edf = 10.3); recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
+However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?  --> reduced flowering probability.
+The responses in 1-ha fragments were substantially more muted, as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
+In other words, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
 Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-spei)c).
@@
 
+***ALTERNATIVE SUMMARY OF THE FLOWERING RESULTS?*** Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
+The responses in 1-ha fragments were also more muted (1-ha edf = 8.1, continuous forest edf = 10.3), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
+This led to some important inter-habitat differences in plant responses to prior droughts: recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
+However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?
+--> reduced flowering probability in continuous forest far more than it did in fragments (Figure \@ref(fig:flwr-spei)c).
+The greatest similarity between forest and fragments occurs when there is strong drought 8--12 months prior, or precipitation 15--20 months prior is extremely high.
+
 FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
-We did not see evidence for a cost of reproduction.
-Flowering in the previous year was a significant predictor of growth ($p = 0.048$ for continuous forest, $p = 0.031$ for fragments) and flowering ($p < 0.001$ in both continuous forest and fragments), but the coefficients were positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
+There was no evidence for a cost of reproduction --- in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
+However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
+This suggests that....<!--# not surew what to make of this lastsentence. Interpret for reader?   -->
 

### 2021-05-13 doc/paper.Rmd
added one citation

@@
 
-<!--# what do you think of this wording?
-Trying to walk people through the big result by breaking into two sentences telling them what they are about to see --> For 1-ha fragments, recent SPEI values (i.e.
+<!--# what do you think of this wording? Trying to walk people through the big result by breaking into two sentences telling them what they are about to see -->
+
+For 1-ha fragments, recent SPEI values (i.e.
 within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
@@
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
-They translate into flowering once every 7.97 years in fragments vs.
-every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
+They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
 While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundel2020a], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments are needed to determine if habitat-specific differences in environmental conditions and local ecological neighborhood interact to delay the demographic responses of plants to climatic variability and the mechanisms underlying these responses [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandInteractionsPlantSize2015; @westerbandEarlyLifeConditions2017]
-
+Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability.
 Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms.

### 2021-05-13 doc/supplemental.Rmd
combine supplemental into one doc

---
title: "Supplement to 'Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments'"
author: 
  - Eric R. Scott
  - María Uriarte
  - Emilio M. Bruna
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: no
      number_sections: no
      fig_caption: yes
      # reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
nocite: |
  @mcphadenChildProdigy1997981999, marengoDroughtAmazonia20052008, zengCausesImpacts20052008, lewis2010AmazonDrought2011
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

withr::with_dir(here::here(), {
  targets::tar_load(c(
    normals_plot,
    embrapa_wide,
    xa_spei,
    trmm_spei,
    gspei
    ))
})

library(tidyverse)
library(patchwork)
library(flextable)
library(tsibble)
library(lubridate)
library(latex2exp)
library(colorspace)
library(glue)
source(here::here("R", "utils.R"))
```

# BDFFP plot map

Map here

Map caption here

\pagebreak

# Climate Normals for BDFFP

```{r normals}
normals_plot & theme(text = element_text(size = 10))
```

Figure S1.
Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981--2010.
Precipitation (a) shows a marked dry seasons from June through October.
Temperature (b) varies less throughout the year.
Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively.
Data from Brazilian National Institute of Meterology (INMET).
<!--# citation needed? -->

\pagebreak

# Comparison of SPEI calculated from different data sources

```{r}
# combine data
xa_df <-
  xa_spei %>% 
  as_tibble() %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(xa_spei = mean(spei))

trmm_df <-
  trmm_spei %>% 
  group_by(yearmonth) %>% 
  summarize(trmm_spei = mean(trmm_spei))

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  pivot_longer(c(-yearmonth), names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

```{r}
# data for plot
spei_plot <- 
  comb %>% 
  group_by(year) %>% 
  mutate(min_mon = spei == min(spei, na.rm = TRUE)) %>% 
  mutate(dataset = case_when(
    dataset == "g_spei"  ~ "Global SPEI",   
    dataset == "xa_spei" ~ "Xavier et al.",
    dataset == "st_spei" ~ "EMBRAPA",
    dataset == "trmm_spei" ~ "TRMM"
  ))
```

```{r spei-plot, fig.width=7, fig.height=3.5}

p <-
  ggplot(spei_plot, aes(x = yearmonth, y = spei)) +
  geom_line(aes(color = dataset), size = 0.5) +
  scale_x_yearmonth(
    "Date",
    date_breaks = "1 year",
  ) +
  scale_y_continuous(TeX("SPEI_3_"),
                     breaks = seq(-2.5, 2.5, 0.5))+
  scale_color_discrete_qualitative() +
  coord_cartesian(xlim = c("1995-02-01", "2009-02-01")) +
  labs(color = "Dataset", linetype = "Dataset") +
  theme_bw() +
  theme(text = element_text(size = 11), axis.text.x = element_text(angle = 45, hjust = 1))

annotate_spei(p)
```

Figure S2.
Three-month SPEI at BDFFP calculated using 4 datasets.
"EMBRAPA" (red) is the average of SPEI calculated using precipitation and evapotranspiration from two weather stations near BDFFP---Estacao Rio Preto da Eva ( -2.65º, -59.65º) and Estacao Manaus (-2.65º, -60.35º).
The data was accessed from the EMBRAPA Infoclima data portal (<https://www.cnpaf.embrapa.br/infoclima/>).
"Global SPEI" (green) is from the SPEI global database[@vicente-serranoNewGlobalGridded2010; @begueriaMultiscalarGlobalDrought2010].
"TRMM" (blue) SPEI uses precipitation data from TRMM [@trmmTMPA3B432011] and evapotranspiration data from @xavierDailyGriddedMeteorological2016 to calculate 3-month SPEI.
Note that TRMM data only goes back to 1998.
@xavierDailyGriddedMeteorological2016 is the SPEI source described in the main text.
Drought categories mild, moderate, severe, and extreme drought are represented by yellow, goldenrod, orange, and red panels, respectively.

\pagebreak

```{r}
citations <- tribble(~year, ~Reference,
                     2005, "Marengo et al. 2008; Zeng et al. 2008",
                     2010, "Lewis et al. 2011",
                     1997, "McPhaden 1999",
                     2002, NA,
                     2009, NA,
                     2004, NA,
                     2006, NA) 
```

```{r}
spei_table <-
  comb %>% 
  group_by(year, dataset) %>% 
  mutate(min = min(spei)) %>% 
  filter(spei == min) %>% 
  mutate(month = month(yearmonth, label = TRUE)) %>% 
  # summarize(min = min, month = month) %>%
  filter(year %in% c(2002, 2004, 2006, 2009, 1997, 2005, 2010)) %>% 
  select(year, dataset, spei, month) %>% 
  mutate(dataset = str_remove(dataset, "_spei")) %>% 
  pivot_wider(names_from = dataset, values_from = c(spei, month)) %>% 
  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), 
             .after = "year") %>% 
  mutate(`Global SPEI` = glue("{round(spei_g, 2)} ({month_g})"),
         `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
         EMBRAPA = glue("{round(spei_st, 2)} ({month_st})"),
         TRMM = glue("{round(spei_trmm, 2)} ({month_trmm})")) %>% 
  mutate(TRMM = if_else(TRMM == "NA (NA)", glue("–"), TRMM)) %>% 
  ungroup()

spei_table <- left_join(spei_table, citations)
```

```{r}
set_flextable_defaults(big.mark = "")
table <- 
  spei_table %>% 
  # select(-starts_with("month_"), -starts_with("spei_")) %>% 
  flextable(
    col_keys = c("year", "cause", "Global SPEI", "Xavier et al.", "EMBRAPA", "TRMM", "BDFFP")
  ) %>% 
  add_header_row(colwidths = c(1,1, 5),
                 values = c("", "", "Lowest SPEI for the year (month)")) %>% 
  align(i = 1, part = "header", align = "center") %>% 
  #extreme drought
  bg(~ spei_g < -2, ~`Global SPEI`, bg = "#e31a1c") %>% 
  bg(~ spei_xa < -2, ~ `Xavier et al.`, bg = "#e31a1c") %>% 
  bg(~ spei_st < -2, ~ EMBRAPA, bg = "#e31a1c") %>% 
  bg(~spei_trmm < -2, ~TRMM, bg = "#e31a1c") %>%

  #severe drought
  bg(~ between(spei_g,     -2, -1.5), ~ `Global SPEI`, bg = "#fd8d3c") %>% 
  bg(~ between(spei_xa,    -2, -1.5), ~ `Xavier et al.`, bg = "#fd8d3c") %>% 
  bg(~ between(spei_st,    -2, -1.5), ~ EMBRAPA, bg = "#fd8d3c") %>% 
  bg(~ between(spei_trmm,  -2, -1.5), ~ TRMM, bg = "#fd8d3c") %>%

  #moderate drought
  bg(~ between(spei_g,     -1.5, -1), ~ `Global SPEI`, bg = "#fecc5c") %>% 
  bg(~ between(spei_xa,    -1.5, -1), ~ `Xavier et al.`, bg = "#fecc5c") %>% 
  bg(~ between(spei_st,    -1.5, -1), ~ EMBRAPA, bg = "#fecc5c") %>% 
  bg(~ between(spei_trmm,  -1.5, -1), ~ TRMM, bg = "#fecc5c") %>%

  #mild drought
  bg(~ between(spei_g,     -1, 0), ~ `Global SPEI`, bg = "#ffffb2") %>% 
  bg(~ between(spei_xa,    -1, 0), ~ `Xavier et al.`, bg = "#ffffb2") %>% 
  bg(~ between(spei_st,    -1, 0), ~ EMBRAPA, bg = "#ffffb2") %>% 
  bg(~ between(spei_trmm,  -1, 0), ~ TRMM, bg = "#ffffb2") 
```

Table S1.
Comparison of SPEI values and drought categories across datasets for major drought events in the Amazon.
Mild, moderate, severe, and extreme drought are represented by yellow, goldenrod, orange, and red, respectively.
Cause indicates whether the drought was caused by the El Niño Southern Oscillation (ENSO) or Atlantic Sea Surface Temperatures (AMO).
The lowest SPEI value for the year identified by each dataset is reported, along with the month that minimum value occurred in.
Datasets are described in the legend for Figure S2.

```{r spei-table}
table %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", j = 3:6) %>%
  border_remove() %>% 
  hline_bottom() %>% 
  hline_top() %>% 
  hline(j = 3:6, part = "header") %>% 
  footnote(j = 2, i = c(1, 4, 7), value = as_paragraph(
    c("McPhaden 1999",
      "Marengo et al. 2008; Zeng et al. 2008",
      "Lewis et al. 2011")),
    ref_symbols = c("a", "b", "c"),
    part = "body", inline = TRUE
  )
```

\pagebreak

# Choice of variable for plant size

include contents of `choose_size_var.Rmd` here.
Ask Emilio about citaiton for dataset.

# References

::: {#refs}
:::

### 2021-05-13 _targets.R
combine model validation and model summaries

@@ -69,7 +69,6 @@ tar_plan(
   
   # Validate and summarize results
   tar_render(validate_models, "doc/validate_models.Rmd"),
-  tar_render(model_summary, "doc/model_summary.Rmd"),
 
   # Descriptive / Exploratory Data Analysis Figures
   normals = normals_data(),

### 2021-05-13 doc/supplemental.Rmd
finalize supplemental doc

@@
     trmm_spei,
-    gspei
+    gspei,
+    la_data
     ))
@@
 library(glue)
+library(performance)
 source(here::here("R", "utils.R"))
@@
 
+Figure S1.
 Map caption here
@@
 
-Figure S1.
+Figure S2.
 Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981--2010.
@@
 
-Figure S2.
+Figure S3.
 Three-month SPEI at BDFFP calculated using 4 datasets.
@@
 
+```{r}
+m_ht <-   lm(log_leaf_area ~ log_ht,   data = la_data)
+m_shts <- lm(log_leaf_area ~ log_shts, data = la_data)
+m_size <- lm(log_leaf_area ~ log_size, data = la_data)
+```
+
+```{r include=FALSE}
+check_model(m_ht)
+check_model(m_shts)
+check_model(m_size)
+```
+
+Table S2.
+Using data from transplant experiments of *Heliconia acuminata* [@brunaExperimentalAssessmentHeliconia2002 <!--# and doi for unpublished data -->] we fit linear regressions with log(leaf area) as the response variable and either log(height), log(shoot number) or log(shoot number x height) as the predictor.
+Using the log of size (shoot number $\times$ height) as a predictor gives the lowest AIC, BIC, root mean squared error (RMSE) and highest $R^2$ indicating that it is the best proxy for predicting log(leaf area) as a response.
+
+```{r size-table}
+compare_performance(m_ht, m_shts, m_size, metrics = "common", rank = TRUE) %>%
+  select(-Model) %>% 
+  add_column(
+    "Model" = c(
+      "log(leaf area) ~ log(shoot number x height)",
+      "log(leaf area) ~ log(height)",
+      "log(leaf area) ~ log(shoot number)"
+    )
+  ) %>%
+  mutate(dAIC = AIC - min(AIC), dBIC = BIC- min(BIC)) %>% 
+  mutate(across(c(dAIC, dBIC, R2), ~round(.x, 2)), RMSE = round(RMSE, 3)) %>% 
+  select(Model, dAIC, dBIC, R2, RMSE) %>% 
+  flextable() %>% 
+  set_table_properties(layout = "autofit") 
+```
+
+\pagebreak
+
 # References

### 2021-05-13 R/leaf_area.R

@@ -5,5 +5,10 @@ read_tidy_la <- function(file) {
    dplyr::rename(leaf_area = leaf_area_feb_99,
                  leaves = leaves_feb_99,
                  ht = height_feb_99,
-                 shts = stems_feb_99)
+                 shts = stems_feb_99) %>% 
+   mutate(size = shts * ht) %>% 
+   #log transform all size variables
+   mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
+   filter(!is.na(leaf_area) & !is.na(ht) & !is.na(shts)) %>% 
+   filter(shts > 0)
 }

### 2021-05-13 _targets.R

@@ -16,33 +16,6 @@ tar_plan(
   xa_raw = read_csv(xa_file),
   xa_spei = calc_spei_xa(xa_raw),
   xa_lag = lag_spei(xa_spei, maxlag),
-
-  # Choice of size supplemental
-  tar_target(la_file, here("data", "HA-la-stems-ht.xlsx"), format = "file"),
-  la_data = read_tidy_la(la_file),
-    tar_render(choose_size_var, "doc/choose_size_var.Rmd"),
-  
-  
-  # SPEI supplemental
-  tar_target(rpde_file,
-             here("data", "supplemental", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv"),
-             format = "file"),
-  tar_target(manaus_file,
-             here("data", "supplemental", "Estacao_Manaus_1980-01-01_2016-12-31.csv"),
-             format = "file"),
-  rpde = read_csv(rpde_file),
-  manaus = read_csv(manaus_file),
-  embrapa_mon = tidy_embrapa(rpde,  manaus),
-  embrapa_wide = calc_spei_embrapa(embrapa_mon),
-  tar_target(trmm_file, here("data", "supplemental", "trmm.csv"), format = "file"),
-  trmm = read_tidy_trmm(trmm_file),
-  tar_target(gspei_file, 
-             here("data", "supplemental", "global_spei_-59.75_-2.25.csv"),
-             format = "file"),
-  gspei = read_tidy_gspei(gspei_file),
-  trmm_spei  = calc_spei_trmm(trmm, xa_spei),
-  
-  tar_render(supplemental, "doc/supplemental.Rmd"),
   
   # Prep demographic data
   tar_target(demog_file, here("data", "Ha_survey_with_Zombies.csv"), format = "file"),
@@ -106,6 +79,29 @@ tar_plan(
   size_plot = make_size_plot(s = s_covar_plot, g = g_covar_plot, f = f_covar_plot, model_data = model_data),
   
   # Main text
-  tar_render(paper, "doc/paper.Rmd")
+  tar_render(paper, "doc/paper.Rmd"),
+  
+  # Supplemental
+  tar_target(rpde_file,
+             here("data", "supplemental", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv"),
+             format = "file"),
+  tar_target(manaus_file,
[…]

### 2021-05-13 packages.R

@@ -25,6 +25,7 @@ library(colorspace)
 library(Amelia)
 library(readxl)
 library(Hmisc)
+library(qqplotr)
 
 conflict_prefer("select", "dplyr")
 conflict_prefer("filter", "dplyr")

### 2021-05-13 doc/validate_models.Rmd

@@
 ## target knits Rmds in their own session, so load libraries here.
-source(here::here("packages.R"))
+library(targets)
+library(statmod)
+library(tidyverse)
 library(gratia)
@@
 ```{r load-targets}
-withr::with_dir(here(),{
+withr::with_dir(here::here(),{
   tar_load(c(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha))
@@
 
+```{r}
+draw(s_cf)
+```
+
 ```{r echo=TRUE}
@@
 ```
+
 ### Fragments
@@
 summary(s_1ha)
+```
+
+```{r}
+draw(s_1ha)
 ```
@@
 ## Growth
+
 ### Continuous Forest
+
 ```{r}
 summary(g_cf)
+```
+
+```{r}
+draw(g_cf)
 ```
@@
 ```{r}
+draw(g_1ha)
+```
+
+```{r}
 appraise(g_1ha)
@@
 ### Continuous Forest
+
+```{r}
+summary(f_cf)
+```
+
+```{r}
+draw(f_cf)
+```
 
@@
 ```{r}
+summary(f_1ha)
+```
+
+```{r}
+draw(f_1ha)
+```
+
+```{r}
 par(mfrow = c(2,2))
@@
 
-<details><summary>Reproducibility receipt</summary>
+<details>
+
+<summary>
+
+Reproducibility receipt
+
+</summary>
 

### 2021-05-13 doc/paper.Rmd
fixed duplicate citations

@@
 For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006a; @dasilva2000].
+Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
 Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999 ; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
@@
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarte2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
@@
 While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H.
-acuminata,* are found at lower densities in the darker and cooler forest understory [@rundel2020].
-These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996d; @bruna2004].
+acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020].
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
 
@@
 They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
-While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundel2020a], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
+While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
 data*).
@@
 Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996d] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
+acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
 While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

### 2021-05-14 doc/paper.Rmd
1st paragraph of intro

@@
 
+<!--# Insert page break -->
+
 # Introduction
@@
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
-Here we support the emerging generalization that the effects of climate drivers on demographic vital rates can be delayed [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-We also found that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.
-Consequently, the hypothesized synergies between climate and fragmentation may be far more complex and pervasive than previously thought.
+Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
+This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdam2004a; @selwood2015a; @laurance2001] are likely to be pervasive, but also far more complex than previously thought.
 
@@
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
-They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
+They translate into flowering once every 7.97 years in fragments vs.
+every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
 While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.

### 2021-05-14 doc/paper.Rmd
spellcheck (and sentence wrap)

@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H.
+While many *Heliconia* species grow in large patches in treefall gaps and other disturbed areas, others, such as *H.
 acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020].
-These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
+These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
 
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
+Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
 While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
@@
 However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
-They translate into flowering once every 7.97 years in fragments vs.
-every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
+They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
 While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
@@
 Fragmentation of forests diminishes their ability to buffer changes in microclimate due to edge effects [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. accuminata* in forest fragments in this study.
+This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. acuminata* in forest fragments in this study.
 We show that the effects of variation in SPEI on survival and growth are more extreme in fragments compared to continuous forest.
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of infloresences.
-Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of inflorescences.
+Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
 Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
@@
 Estimates green-up, literfall, growth, etc.
-for central and southern america including the Amazon.
-Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
+for central and southern America including the Amazon.
+Drought results in anomalously high leaf flushing at start of drought followed by increased litterfall during the drought.
 Dry season droughts result in green-up, but not wet season droughts.
@@
 
--   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
 can be affected

### 2021-05-14 doc/paper.Rmd
fix duplicate references

@@
 
-<!--# Insert page break -->
+\pagebreak
 
@@
 Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].
+This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwood2015].
 
@@
 We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
-This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdam2004a; @selwood2015a; @laurance2001] are likely to be pervasive, but also far more complex than previously thought.
+This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwood2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.
 

### 2021-05-17 doc/paper.Rmd
em dashes

@@
 Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
-Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
 This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwood2015].
@@
 FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
-There was no evidence for a cost of reproduction --- in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
+There was no evidence for a cost of reproduction---in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
 However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.

### 2021-05-17 doc/paper.Rmd
add refs to supplemental and spellcheck

--- 2021-05-14 doc/paper.Rmd
@@
 abstract: |
-  Text of abstract. 34,281 year x individual measurments! 
+  Text of abstract. 34,281 year x individual measurements! 
 keywords: |
@@
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
+In the current study we used the natural logarithm of the product of shoot number and height as a proxy for plant size.
+Preliminary analysis showed that this was a better predictor of total leaf area than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
 Plants that were not found for three consecutive surveys were considered dead.
 We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
-For additional details on the plots, survey methods, and H.
-acuminata populations see @brunaHabitatFragmentationDemographic2002.
+For additional details on the plots, survey methods, and *H. acuminata* populations see @brunaHabitatFragmentationDemographic2002.
 
@@
 SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
-<!--# add this as a "child document" or whatever -->
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (Figure S2; Table S1 <!--# update crossrefs manually -->).
 
@@
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
-This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size <!--# reference supplemental about plant size? -->.
-A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
+This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
 Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
@@
 
-Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
+Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
@@
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
-This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c) <!--# , which corresponds roughly to ____ shoots --This requires a supplemental figure, probably.
-A plot of shoots vs.
-log(size) -->.
+This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
 Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
@@
 ```{r}
-eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
+eda_cap <- "Timeseries of demographic parameters and drought occurrence.
 Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed).
@@
 The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years.
-Monthly 3-month SPEI is shown (d) with grey lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
+Monthly 3-month SPEI is shown (d) with gray lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
 Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by cubic regression splines.
+size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probability, modeled by cubic regression splines.
 Panel d) shows the smoothed density of different plant sizes in the two habitat types.

### 2021-05-17 R/plot_smooths.R
changes size figure to show smooths as linear predictors (not back-transformed and without intercept added).  Adresses #51.

@@ -28,9 +28,9 @@ my_eval_smooth <- function(model, smooth, ...) {
 #' @param ... other arguments passed to gratia::smooth_estimates()
 #'
 plot_covar_smooth <- function(cf_model, frag_model, covar) {
-  cf   <- my_eval_smooth(cf_model, covar, unconditional = TRUE)
-  frag <- my_eval_smooth(frag_model, covar, unconditional = TRUE)
-  
+
+  cf <- gratia::smooth_estimates(cf_model, covar, unconditional = TRUE) %>% add_confint()
+  frag <- gratia::smooth_estimates(frag_model, covar, unconditional = TRUE) %>% add_confint()
   data <- bind_rows("1-ha" = frag, "CF" = cf, .id = "habitat")
   
   p <- 
@@ -73,11 +73,9 @@ make_size_plot <- function(s, g, f, model_data) {
     plot_annotation(tag_levels = "a", tag_suffix = ")") &
     theme(plot.margin = margin()) &
     # set color for all panels
-    scale_color_manual(values = c("#E66101", "#5E3C99"),
+    scale_color_manual("Habitat", values = c("#E66101", "#5E3C99"),
                        aesthetics = c("color", "fill")) &
-    guides(col = guide_legend(title = "Habitat"),
-           linetype = guide_legend(title = "Habitat"),
-           fill = guide_legend(title = "Habitat"))
+    guides(linetype = guide_legend(title = "Habitat"))
     
 }
 

### 2021-05-17 _targets.R

@@ -32,7 +32,6 @@ tar_plan(
   tar_render(validate_data, "doc/validate_data.Rmd"),
   
   # Fit demographic models
-  
   s_cf = fit_surv(model_data_cf, flwr_prev = TRUE),
   s_1ha = fit_surv(model_data_1ha, flwr_prev = TRUE),
   g_cf  = fit_growth(model_data_cf, flwr_prev = TRUE),
@@ -60,18 +59,17 @@ tar_plan(
 
   ## Survival
   s_covar_plot = plot_covar_smooth(frag_model = s_1ha, cf_model = s_cf, covar = "log_size_prev") + 
-                  labs(x = TeX("$log(size_t)$"), y = "P(survival)"),
+                  labs(x = TeX("$log(size_t)$"), y = "Effect [survival]"),
   s_spei_plot = plot_cb_3panel(s_cf, s_1ha, binwidth = 0.002, response_lab = "P(survival)"),
 
   ## Growth
   g_covar_plot = plot_covar_smooth(frag_model = g_1ha, cf_model = g_cf, covar = "log_size_prev") +
-                  geom_abline(slope = 1, color = "grey50") +
-                  labs(x = TeX("$log(size_t)$"), y = TeX("$log(size_{t+1})")),
+                  labs(x = TeX("$log(size_t)$"), y = TeX("Effect \\[$log(size_{t+1})\\]")),
   g_spei_plot = plot_cb_3panel(g_cf, g_1ha, binwidth = 0.05, response_lab = "$log(size_{t+1})$"),
 
   ## Flowering
   f_covar_plot = plot_covar_smooth(frag_model = f_1ha, cf_model = f_cf, covar = "log_size_prev") +
-                  labs(x = TeX("$log(size_t)$"), y = "P(flowering)"),
+                  labs(x = TeX("$log(size_t)$"), y = "Effect [flowering]"),
   f_spei_plot = plot_cb_3panel(f_cf, f_1ha, binwidth = 0.001, response_lab = "P(flowering)"),
 
   ## Size covariate

### 2021-05-17 doc/paper.Rmd
edits including change to figure caption that addresses #51

@@
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
-In the current study we used the natural logarithm of the product of shoot number and height as a proxy for plant size.
-Preliminary analysis showed that this was a better predictor of total leaf area than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
+In this study, we used the product of shoot number and plant height as our measure of plant size.
+Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
 Plants that were not found for three consecutive surveys were considered dead.
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probability, modeled by cubic regression splines.
-Panel d) shows the smoothed density of different plant sizes in the two habitat types.
-95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
+size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s_1_(z_i_) in eq.
+1.
+95% confidence intervals are shown and include uncertainty in the intercept and uncertainty due to smoothness selection.
 The smooths for 1-ha fragments and continuous forest are fit in separate models.
-The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.
-In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
-
+Panel d) shows the distribution of plant sizes in the two habitat types."
 ```

### 2021-05-17 doc/paper.Rmd
re-order panels in size smooth plot

@@
 Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
-While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
+While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.
 However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
@@
 As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
-While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).
+While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).
 

### 2021-05-17 R/plot_smooths.R

@@ -60,8 +60,8 @@ make_size_plot <- function(s, g, f, model_data) {
     theme(legend.position = "none")
   
   top <-
-    s /
     g /
+    s /
     f & 
     theme(axis.title.x = element_blank(),
           axis.text.x = element_blank(),

### 2021-05-17 doc/paper.Rmd
better now that we discussed the 2 paragraphs

--- 2021-05-17 doc/paper.Rmd
@@
 abstract: |
-  Text of abstract. 34,281 year x individual measurements! 
+  Text of abstract. 34,281 year x individual measurments! 
 keywords: |
@@
 
-\pagebreak
+<!--# Insert page break -->
 
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in treefall gaps and other disturbed areas, others, such as *H.
+While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H.
 acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020].
-These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
+These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
 
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
-In the current study we used the natural logarithm of the product of shoot number and height as a proxy for plant size.
-Preliminary analysis showed that this was a better predictor of total leaf area than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
 Plants that were not found for three consecutive surveys were considered dead.
 We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
-For additional details on the plots, survey methods, and *H. acuminata* populations see @brunaHabitatFragmentationDemographic2002.
+For additional details on the location of plots, survey methods, and *H. acuminata* population structure see @brunaHabitatFragmentationDemographic2002.
 
@@
 SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (Figure S2; Table S1 <!--# update crossrefs manually -->).
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
+<!--# add this as a "child document" or whatever -->
 
@@
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
-This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
+This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size <!--# reference supplemental about plant size? -->.
+A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
 Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
@@
 
-Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
+Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
@@
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
+Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
 While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
@@
 <!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average).
--->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
+-->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
@@
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
-This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
+This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c) <!--# , which corresponds roughly to ____ shoots --This requires a supplemental figure, probably.
+A plot of shoots vs.
+log(size) -->.
 Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
@@
 
-Our results indicate that some---but not all---results from short-term demographic studies can be extrapolated to longer times scales.
-Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
-In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
-While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
-The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
-However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs.
-0.094 in FF).
+Many studies investigating the biological consequences of habitat fragmentation on plant growth, survival, and reproduction comprise short-term ([\<]{.ul}3 year) experiments and observations.
+Our results underscore the difficulty in extrapolating long-term trends from such short-term studies, particularly when studying long-lived organisms or when the responses of interest can vary with size or age.
+For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1997-2002 (i.e., higher survival in continuous forest), 2002-2005 (i.e., higher survival in fragments), or 2004-2007 (i.e, no clear effect of fragmentation; \@ref(fig:eda)**X**).
+It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative fo continuous forest (**Fig 2**), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@bruna2003b].
 
-These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
-However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
-They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
-While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl.
-data*).
-Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations.
-[@gagnonGrowthUnderstoryHerb2011].
+Similarly, conclusions regarding the effects of fragmentation on flowering---which is also both rare and size-dependent [@brooks2019]---would also differ based on the year in which they were investigated.
+This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or how population genetic structure [@uriarte2010; @uriarte2011; @cortes2013].
+Conclusions based on short-term observations of temporally variable vital rates could lead to conservation and management practices that are ineffective or even counterproductive, especially when when failing to consider how the consequences of this variation might be modulated by organismal life history [@morris2008].
 
-<!--# new paragraph below.  What do you think Emilio? -->
-
-Fragmentation of forests diminishes their ability to buffer changes in microclimate due to edge effects [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. acuminata* in forest fragments in this study.
-We show that the effects of variation in SPEI on survival and growth are more extreme in fragments compared to continuous forest.
-Additionally, the direction of the SPEI effects appear to be seasonal---that is, extreme dry conditions in dry seasons and extreme wet conditions in wet seasons are most detrimental to growth and survival.
-The ability of continuous forest habitat to buffer these extremes may be responsible for the less extreme effects of variation in SPEI in continuous forest.
-More experimental work is needed to establish alterations in microclimate or impaired microclimate buffering in extreme weather events as a causal mechanism for reduced growth and survival of plant individuals in forest fragments.
+It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (**Fig X**).
+Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments ---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
+While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999;@ewers2013].
+We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@laurance2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
 
@@
 
-Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
+Climate anomalies are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
 These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
@@
 
-Despite this knowledge, models of plant population dynamics have rarely included the effects of environmental drivers.
+Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015a; @tenhumberg2018; @molowny-horas2017).
 This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
-To address this challenge, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
+To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
 Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
@@
 Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types, the magnitude of these declines was far greater in forest fragments.
+For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types.
+However, the magnitude of these declines was far greater in forest fragments.
 Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
-In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
+In contrast, variation in SPEI had far stronger effects on the probability of flowering in continuous forest than fragments.
+These results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
 
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of inflorescences.
-Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of infloresences.
+Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
 Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
@@
 Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021].
-For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
-This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018].
-Similar delayed changes in the local environment could also influence the foraging behavior of *H.
-acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
-While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
+For example, topical trees may not die until three or more years after a drought [@crileyYearProductionHigh1994].
+When they finally do, the resulting leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018], triggering a boom in the growth and flowering of understory plants [@bruna2005a].
+Similar delayed changes in the local environment could also influence the foraging behavior of a plant's pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996], seed dispersers [@uriarteDisentanglingDriversReduced2011], or herbivores [@scotta].
+While more work is needed to explain why the (delayed) effects of SPEI on *H.
+acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis motivated by recent intriguing results from other systems **(CITATIONS)** is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.
 
@@
 Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
+More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.
 
@@
 ```{r}
-eda_cap <- "Timeseries of demographic parameters and drought occurrence.
+eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
 Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed).
@@
 The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years.
-Monthly 3-month SPEI is shown (d) with gray lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
+Monthly 3-month SPEI is shown (d) with grey lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
 Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probability, modeled by cubic regression splines.
+size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by cubic regression splines.
 Panel d) shows the smoothed density of different plant sizes in the two habitat types.
@@
 Estimates green-up, literfall, growth, etc.
-for central and southern America including the Amazon.
-Drought results in anomalously high leaf flushing at start of drought followed by increased litterfall during the drought.
+for central and southern america including the Amazon.
+Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
 Dry season droughts result in green-up, but not wet season droughts.
@@
 
--   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
 can be affected

### 2021-05-17 doc/paper.Rmd
changes exploratory plot panel a to fold-change. addresses #52

--- 2021-05-17 doc/paper.Rmd
@@
       correspondence: false
-
 institute:
@@
 eda_cap <- "Timeseries of demographic parameters and drought occurrence.
-Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed).
+Mean fold-change in plant plant size (log2(size_t+1_ / size_t)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed).
+In most years plants grew on average with a notable exception in 2003 where plants regressed in size on average in both habitats (fold-change < 0).
+Error bars in a) represent standard deviation.
 The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census.

### 2021-05-17 R/plot_eda.R

@@ -147,6 +147,7 @@ plot_eda_size_foldchange <- function(demog_post, date_lims) {
   size <-
     demog_plotdf %>% 
     ggplot(aes(x = yearmonth, y = log2_growth, color = habitat, linetype = habitat)) +
+    geom_hline(yintercept = 0, color = "grey50") +
     stat_summary(
       geom = "line",
       fun = "mean",

### 2021-05-17 _targets.R

@@ -49,7 +49,7 @@ tar_plan(
   
   eda_spei = plot_eda_spei(xa_lag, plot_dates),
   eda_surv_ts = plot_eda_surv_ts(model_data, plot_dates),
-  eda_size = plot_eda_size(model_data, plot_dates),
+  eda_size = plot_eda_size_foldchange(model_data, plot_dates),
   eda_flwr = plot_eda_flwr(model_data, plot_dates),
   eda_plot = plot_eda_combine(eda_size, eda_surv_ts, eda_flwr, eda_spei),
   

### 2021-05-17 doc/paper.Rmd
switch to sentence wrap for ease of merging

--- 2021-05-17 doc/paper.Rmd
@@
 Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments ---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
-While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999;@ewers2013].
+While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999; @ewers2013].
 We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@laurance2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.

### 2021-05-17 doc/paper.Rmd
trying to get diffs to work.

--- 2021-05-17 doc/paper.Rmd
@@
       correspondence: false
+
 institute:

### 2021-05-17 doc/paper.Rmd

--- 2021-05-17 doc/paper.Rmd
@@
 *Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
-While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H.
+While many *Heliconia* species grow in large patches in treefall gaps and other disturbed areas, others, such as *H.
 acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020].
-These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
+These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
 
@@
 Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
+In this study, we used the product of shoot number and plant height as our measure of plant size.
+Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
 Plants that were not found for three consecutive surveys were considered dead.
@@
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
-While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
+Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
+While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.
 However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
@@
 While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999; @ewers2013].
-We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@laurance2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
+We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
 
@@
 ```{r}
-eda_cap <- "Timeseries of demographic parameters and drought ocurrence.
-Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed).
+eda_cap <- "Timeseries of demographic parameters and drought occurrence.
+Mean fold-change in plant plant size (log2(size_t+1_ / size_t)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed).
+In most years plants grew on average with a notable exception in 2003 where plants regressed in size on average in both habitats (fold-change < 0).
+Error bars in a) represent standard deviation.
 The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census.
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census on a) survival, b) log(size), and c) flowering probabiility, modeled by cubic regression splines.
-Panel d) shows the smoothed density of different plant sizes in the two habitat types.
-95% confidence intervals are shown and include uncertainty in the model intercept and uncertainty due to smoothness selection.
+size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s_1_(z_i_) in eq.
+1.
+95% confidence intervals are shown and include uncertainty in the intercept and uncertainty due to smoothness selection.
 The smooths for 1-ha fragments and continuous forest are fit in separate models.
-The model intercepts were added to the fitted values of the smooth and values are back-transformed to the response scale.
-In b), a 1:1 line is shown in dashed red to indicate the line of no growth."
-
+Panel d) shows the distribution of plant sizes in the two habitat types."
 ```

### 2021-05-17 doc/paper.Rmd
fix duplicate citations

@@
 
-Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
+Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
 While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
@@
 Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].
+This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
 
@@
 We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
-This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdam2004a; @selwood2015a; @laurance2001] are likely to be pervasive, but also far more complex than previously thought.
+This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.
 
@@
 For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1997-2002 (i.e., higher survival in continuous forest), 2002-2005 (i.e., higher survival in fragments), or 2004-2007 (i.e, no clear effect of fragmentation; \@ref(fig:eda)**X**).
-It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative fo continuous forest (**Fig 2**), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@bruna2003b].
+It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative fo continuous forest (**Fig 2**), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@brunaArePlantPopulations2003].
 
-Similarly, conclusions regarding the effects of fragmentation on flowering---which is also both rare and size-dependent [@brooks2019]---would also differ based on the year in which they were investigated.
-This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or how population genetic structure [@uriarte2010; @uriarte2011; @cortes2013].
-Conclusions based on short-term observations of temporally variable vital rates could lead to conservation and management practices that are ineffective or even counterproductive, especially when when failing to consider how the consequences of this variation might be modulated by organismal life history [@morris2008].
+Similarly, conclusions regarding the effects of fragmentation on flowering---which is also both rare and size-dependent [@brooksStatisticalModelingPatterns2019]---would also differ based on the year in which they were investigated.
+This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or how population genetic structure [@uriarteEffectsForestFragmentation2010; @uriarteDisentanglingDriversReduced2011; @cortesLowPlantDensity2013].
+Conclusions based on short-term observations of temporally variable vital rates could lead to conservation and management practices that are ineffective or even counterproductive, especially when when failing to consider how the consequences of this variation might be modulated by organismal life history [@morrisLongevityCanBuffer2008].
 
@@
 Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments ---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
-While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999; @ewers2013].
+While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
 We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
@@
 
-Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015a; @tenhumberg2018; @molowny-horas2017).
+Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015; @tenhumbergTimelaggedEffectsWeather2018; @molowny-horasChangesNaturalDynamics2017).
 This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
@@
 For example, topical trees may not die until three or more years after a drought [@crileyYearProductionHigh1994].
-When they finally do, the resulting leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018], triggering a boom in the growth and flowering of understory plants [@bruna2005a].
-Similar delayed changes in the local environment could also influence the foraging behavior of a plant's pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996], seed dispersers [@uriarteDisentanglingDriversReduced2011], or herbivores [@scotta].
+When they finally do, the resulting leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018], triggering a boom in the growth and flowering of understory plants [@brunaDemographicEffectsHabitat2005].
+Similar delayed changes in the local environment could also influence the foraging behavior of a plant's pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996], seed dispersers [@uriarteDisentanglingDriversReduced2011], or herbivores [@scottDifferingNonlinearLagged2021].
 While more work is needed to explain why the (delayed) effects of SPEI on *H.

### 2021-05-24 doc/paper.Rmd
fix line wrap for better merging
minor edits that I am confident in

@@
 
-<!--# Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->
-
 \pagebreak
@@
 The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
+Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest.
 This study uses data from 1-ha fragments and continuous forest sites.
@@
 Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were consequently substantially larger than those in forest fragments (150 ± 175 SD vs.
+Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were on average substantially larger than those in forest fragments (150 ± 175 SD vs.
 112 ± 141 SD, respectively).
-<!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average).
--->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
+The distribution of sizes of individuals was somewhat bimodal in both habitats and the major difference in plant size was in the larger of the two modes being smaller in fragments (Figure \@ref(fig:size-plot)d).
+This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).

### 2021-05-24 doc/paper.Rmd
incorporated Emilio's edits to results

@@
 
-<!--# what do you think of this wording? Trying to walk people through the big result by breaking into two sentences telling them what they are about to see -->
-
-For 1-ha fragments, recent SPEI values (i.e.
-within the past 0--12 months) had the strongest effect on survival <!--# Is another way to say this "For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with..."  -->., with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
-<!--# what do you think?
-trying to put the results in more ecological terms) -->In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional -- survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
-Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census).
-Drought 15--20 months prior to a census (i.e., SPEI \< -1) was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, was a nearly linear relationship between SPEI and survival probability).
+For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
+In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional---survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
+Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census) and 32--36 months.
+Drought (i.e., SPEI \< -1) 15--20 months prior to a census was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, there was a nearly linear relationship between SPEI and survival probability).
 Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
 It should be noted, however, that only the first year of census data (1999) met these conditions.
-<!--# can you walk me through these next two sentences?
-I'm not sure if this applies to the 36 mop lag or the whole time-series -->Subtracting the estimated contribution of SPEI history to survival in fragments from that in continuous forest shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and fragments (Figure \@ref(fig:surv-spei)c).
-However, under extreme conditions, survival probability is higher in continuous forest by up to 0.02.
+We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
+This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
+However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.
 
@@
 Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
-The effects of SPEI on flowering probability in continuous forest were also more complex (1-ha edf = 8.1, continuous forest edf = 10.3); recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
-However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?  --> reduced flowering probability.
-The responses in 1-ha fragments were substantially more muted, as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
-In other words, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.
-Plotting the difference between the crossbasis smooths for continuous forest and 1-ha fragments revealed that flowering probability is higher in continuous forest compared to 1-ha fragments and is increased by recent drought and drought during the dry season 15--20 months prior to the census (Figure \@ref(fig:flwr-spei)c).
-The least difference between habitat types occurs when there is strong drought at about 8--12 months lag or very wet conditions at 15--20 months lag.
-
-***ALTERNATIVE SUMMARY OF THE FLOWERING RESULTS?*** Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
 The responses in 1-ha fragments were also more muted (1-ha edf = 8.1, continuous forest edf = 10.3), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
-This led to some important inter-habitat differences in plant responses to prior droughts: recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
-However, drought at lags 7--13 <!--# this corresponds to what - the dry and rainy 1 year ago?
---> reduced flowering probability in continuous forest far more than it did in fragments (Figure \@ref(fig:flwr-spei)c).
-The greatest similarity between forest and fragments occurs when there is strong drought 8--12 months prior, or precipitation 15--20 months prior is extremely high.
+This led to some important inter-habitat differences in plant responses to prior droughts.
+Recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
+However, drought at lags 7--13 (the end of the rainy season one year prior) reduced flowering probability in continuous forest far more than it did in fragments (Figure \@ref(fig:flwr-spei)c).
 
 FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
-There was no evidence for a cost of reproduction --- in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
+There was no evidence for a cost of reproduction---in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
 However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
-This suggests that....<!--# not surew what to make of this lastsentence. Interpret for reader?   -->
 

### 2021-05-24 doc/paper.Rmd
edits to discussion

@@
 We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+<!--# I'd like to cite nothdurft et al.
+Possibly as an example of a three-way interaction.
+Theirs is temperature x precip x lag, but could be used for SPEI x size x lag *if* we had plant size on the same time scale as SPEI.
+Maybe a discussion sentence.
+-->
 
@@
 Our results underscore the difficulty in extrapolating long-term trends from such short-term studies, particularly when studying long-lived organisms or when the responses of interest can vary with size or age.
-For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1997-2002 (i.e., higher survival in continuous forest), 2002-2005 (i.e., higher survival in fragments), or 2004-2007 (i.e, no clear effect of fragmentation; \@ref(fig:eda)**X**).
-It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative fo continuous forest (**Fig 2**), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@brunaArePlantPopulations2003].
+For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1999--2002 (i.e., higher survival in continuous forest), 2002--2005 (i.e., higher survival in fragments), or 2004--2007 (i.e, no clear effect of fragmentation; Figure \@ref(fig:eda)b).
+It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative to continuous forest (Figure \@ref(fig:surv-curve)), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@brunaArePlantPopulations2003].
 
 Similarly, conclusions regarding the effects of fragmentation on flowering---which is also both rare and size-dependent [@brooksStatisticalModelingPatterns2019]---would also differ based on the year in which they were investigated.
-This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or how population genetic structure [@uriarteEffectsForestFragmentation2010; @uriarteDisentanglingDriversReduced2011; @cortesLowPlantDensity2013].
+This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or population genetic structure [@uriarteEffectsForestFragmentation2010; @uriarteDisentanglingDriversReduced2011; @cortesLowPlantDensity2013].
 Conclusions based on short-term observations of temporally variable vital rates could lead to conservation and management practices that are ineffective or even counterproductive, especially when when failing to consider how the consequences of this variation might be modulated by organismal life history [@morrisLongevityCanBuffer2008].
 
-It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (**Fig X**).
-Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments ---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
-While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
+It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (Figures \@ref(fig:surv-spei), \@ref(fig:growth-spei)).
+Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
+While intact forest and its canopy buffer populations from climatic extremes, populations in fragments---especially near edges with high contrast matrix---likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
 We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of inflorescences.
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences.
 Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
@@
 While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
-In fact, flowering in one year is associated with increased reproduction and growth in the next.
+In fact, flowering in one year is associated with increased reproduction and growth in the next <!--# this may change once individual-level random effects are incorporated -->.
 
-Demographic responses will also be delayed if abiotic stress causes plants to invest in below-ground rhizomes [*sensu* @pumisutapon2012].
-The carbohydrates stored in rhizomes allow *Heliconia* to regenerate above-ground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
+Demographic responses will also be delayed if abiotic stress causes plants to invest in belowground rhizomes [*sensu* @pumisutapon2012].
+The carbohydrates stored in rhizomes allow *Heliconia* to regenerate aboveground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
 This may be why drought led to delayed increases in growth---by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
@@
 While more work is needed to explain why the (delayed) effects of SPEI on *H.
-acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis motivated by recent intriguing results from other systems **(CITATIONS)** is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.
+acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis motivated by recent intriguing results from other systems **(CITATIONS)** <!--#  --> is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.
 
@@
 In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
-This is possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
+This possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
 While the very slow growth and extremely low mortality rates of *H.
@@
 
-We thank Collin Edwards, ----, ----, and --- anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Collin Edwards, \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
 We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
@@
 National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
-This is publication No. ------ in the BDFFP Technical Series.
+This is publication No. \_\_\_ in the BDFFP Technical Series.
 

### 2021-05-24 doc/validate_models.Rmd
addresses #56

@@
 withr::with_dir(here::here(),{
-  tar_load(c(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha))
+  tar_load(c(
+    s_cf,
+    s_1ha,
+    g_cf,
+    g_1ha,
+    f_cf,
+    f_1ha,
+    s_cf_sub,
+    g_cf_sub,
+    f_cf_sub
+    ))
   })
@@
 Quantile residuals are more appropriate for binomial models.
+### Fragments
+
+```{r}
+summary(s_1ha)
+```
+
+```{r}
+draw(s_1ha)
+```
+
+```{r echo=TRUE}
+par(mfrow = c(2,2))
+#histogram
+plot(density(qresid(s_1ha)))
+#QQ plot
+qqnorm(qresid(s_1ha)); qqline(qresid(s_1ha))
+#Fitted vs. residuals
+scatter.smooth(predict(s_1ha, type = "response"), qresid(s_1ha), col = "grey")
+```
 
@@
 
-### Fragments
+### Continuous forest subsample
+
+To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
+For survival, edf is actually higher in CF with a larger sample size.
+With subsample, edf are more similar.
 
 ```{r}
-summary(s_1ha)
+summary(s_1ha)$edf[3]
+summary(s_cf)$edf[3]
+summary(s_cf_sub)$edf[3]
 ```
@@
 ```{r}
-draw(s_1ha)
+draw(s_cf_sub)
 ```
-
-```{r echo=TRUE}
-par(mfrow = c(2,2))
-#histogram
-plot(density(qresid(s_1ha)))
-#QQ plot
-qqnorm(qresid(s_1ha)); qqline(qresid(s_1ha))
-#Fitted vs. residuals
-scatter.smooth(predict(s_1ha, type = "response"), qresid(s_1ha), col = "grey")
-```
+Surface still looks different in a similar way though.
 
 ## Growth
-
-### Continuous Forest
-
-```{r}
-summary(g_cf)
-```
-
-```{r}
-draw(g_cf)
-```
-
-```{r}
-gratia::appraise(g_cf)
-```
 
@@
 
+### Continuous Forest
+
+```{r}
+summary(g_cf)
+```
+
+```{r}
+draw(g_cf)
+```
+
+```{r}
+gratia::appraise(g_cf)
+```
+### Continuous Forest subsample
+
+To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
+
+```{r}
+summary(g_1ha)$edf[3]
+summary(g_cf)$edf[3]
+summary(g_cf_sub)$edf[3]
+```
+edf of subsample is similar to full dataset, despite differences in sample size.
+
+```{r}
+draw(g_cf_sub)
+```
+Crossbasis surface looks nearly identical.
+
 ## Flowering
@@
 
+### Fragments
+
+```{r}
+summary(f_1ha)
+```
+
+```{r}
+draw(f_1ha)
+```
+
+```{r}
+par(mfrow = c(2,2))
+#histogram
+plot(density(qresid(f_1ha)))
+#QQ plot
+qqnorm(qresid(f_1ha)); qqline(qresid(f_1ha))
+#Fitted vs. residuals
+scatter.smooth(predict(f_1ha, type = "response"), qresid(f_1ha), col = "grey")
+```
 ### Continuous Forest
@@
 
-### Fragments
+### Continuous Forest subsample
+To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
+For flowering, edf is actually slightly higher in CF with a larger sample size.
+With subsample, edf are more similar.
 
 ```{r}
-summary(f_1ha)
+summary(f_1ha)$edf[3]
+summary(f_cf)$edf[3]
+summary(f_cf_sub)$edf[3]
 ```
@@
 ```{r}
-draw(f_1ha)
+draw(f_cf_sub)
 ```
-
-```{r}
-par(mfrow = c(2,2))
-#histogram
-plot(density(qresid(f_1ha)))
-#QQ plot
-qqnorm(qresid(f_1ha)); qqline(qresid(f_1ha))
-#Fitted vs. residuals
-scatter.smooth(predict(f_1ha, type = "response"), qresid(f_1ha), col = "grey")
-```
+Crossbasis surface looks extremely similar.
 

### 2021-05-24 R/filter_data.R

@@ -21,3 +21,10 @@ filter_data <- function(data) {
     #only use post-seedlings for this
     filter(code_notes != "sdlg (1)" | is.na(code_notes))
 }
+
+subset_cf <- function(model_data_cf, n_plants = 1010) {
+  set.seed(123)
+  plant_ids <- unique(model_data_cf$ha_id_number)
+  plant_sample <- sample(plant_ids, n_plants)
+  model_data_cf %>% filter(ha_id_number %in% plant_sample)
+}

### 2021-05-24 _targets.R

@@ -27,6 +27,7 @@ tar_plan(
   model_data = filter_data(demog_spei),
   model_data_cf = filter(model_data, habitat == "CF"),
   model_data_1ha = filter(model_data, habitat == "1-ha"),
+  model_data_cf_sub = subset_cf(model_data),
   
   # Data validation
   tar_render(validate_data, "doc/validate_data.Rmd"),
@@ -40,6 +41,10 @@ tar_plan(
   f_1ha = fit_flwr(model_data_1ha, flwr_prev = TRUE),
   
   # Validate and summarize results
+  ### Check for edf differences due to sample size
+  g_cf_sub = fit_growth(model_data_cf_sub, flwr_prev = TRUE),
+  f_cf_sub = fit_flwr(model_data_cf_sub, flwr_prev = TRUE),
+  s_cf_sub = fit_surv(model_data_cf_sub, flwr_prev = TRUE),
   tar_render(validate_models, "doc/validate_models.Rmd"),
 
   # Descriptive / Exploratory Data Analysis Figures

### 2021-05-24 doc/paper.Rmd
edits before sending to Maria

@@
 Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
-Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999 ; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
+Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
 Despite the fact that this mechanism is thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions remains scarce.
@@
 Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
-Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
+Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+This increased sensitivity to climate fluctuations, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
 
@@
 Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall ranges from 1900--2300 mm.
-There is a pronounced dry season from June to October (Figure S1 <!--# update figure number manually -->).
+There is a pronounced dry season from June to October (Figure S1<!--# update figure number manually -->).
 
@@
 These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
-
 In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
@@
 Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}=2$), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
-Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
+Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
 
@@
 Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
-SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
-The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (Figure S2; Table S1 <!--# update crossrefs manually -->).
+SPEI calculations were made using the `SPEI` package[@begueriaSPEICalculationStandardised2017].
+The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought sometimes differed (Figure S2; Table S1 <!--# update crossrefs manually -->).
 
@@
 We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
-<!--# I'd like to cite nothdurft et al.
-Possibly as an example of a three-way interaction.
-Theirs is temperature x precip x lag, but could be used for SPEI x size x lag *if* we had plant size on the same time scale as SPEI.
-Maybe a discussion sentence.
--->
+
+All analyses were conducted in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020].
 
@@
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
-The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
+The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the southwestern Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
 Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).
@@
 
+model_data %>% 
+  group_by(habitat, plot) %>% 
+  summarize(n_plants = length(unique(ha_id_number))) %>% 
+  group_by(habitat) %>% 
+  summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants))
+
 by_year <-
@@
 Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
-Plots in CF had on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` SD; 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` SD).
+Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).
 
@@
 This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
-Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
+Even though flowering probability of large plants was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive size threshold in forest fragments (Figure \@ref(fig:size-plot)d).
 Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
@@
 
-Drought history had a significant effect on the survival ($p < 0.001$), growth ($p < 0.001$), and flowering ($p < 0.001$) of plants in both habitats.
+Drought history had a significant ($p < 0.001$) effect on the survival, growth, and flowering of plants in both habitats.
 Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.
@@
 
-While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the surface for 1-ha fragments indicated more complex responses in some situations (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
 For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
@@
 
-FInally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
-There was no evidence for a cost of reproduction---in both forest and fragments, flowering in the previous year was a significant predictor of growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha).
-However, the coefficients were always positive, indicating that flowering in the previous year was positively related to both size and flowering probability in the current year.
+Finally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
+We found no evidence for a cost of reproduction.
+In both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
 
@@
 Our results underscore the difficulty in extrapolating long-term trends from such short-term studies, particularly when studying long-lived organisms or when the responses of interest can vary with size or age.
-For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1999--2002 (i.e., higher survival in continuous forest), 2002--2005 (i.e., higher survival in fragments), or 2004--2007 (i.e, no clear effect of fragmentation; Figure \@ref(fig:eda)b).
+For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1999--2002 (i.e., higher survival in continuous forest), 2002--2005 (i.e., higher survival in fragments), or 2004--2007 (i.e, no clear effect of fragmentation) (Figure \@ref(fig:eda)b).
 It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative to continuous forest (Figure \@ref(fig:surv-curve)), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@brunaArePlantPopulations2003].
@@
 Climate anomalies are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
-These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
+These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation [@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
 There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].
@@
 The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
-For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences.
-Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
+For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences [@crileyYearProductionHigh1994].
+Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots.
 Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
@@
 eda_cap <- "Timeseries of demographic parameters and drought occurrence.
-Mean fold-change in plant plant size (log2(size_t+1_ / size_t)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed).
+Mean fold-change in plant plant size (log2(size~t+1~ / size~t~)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed).
 In most years plants grew on average with a notable exception in 2003 where plants regressed in size on average in both habitats (fold-change < 0).
@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s_1_(z_i_) in eq.
+size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s~1~(z~i~) in eq.
 1.

### 2021-05-25 doc/paper.Rmd
add in final citation

@@
 While more work is needed to explain why the (delayed) effects of SPEI on *H.
-acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis motivated by recent intriguing results from other systems **(CITATIONS)** <!--#  --> is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.
+acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis, motivated by recent intriguing results from other systems [@sapsfordChickenEggWhich2017], is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.
 

### 2021-05-25 doc/paper.Rmd
justify omitting 10-ha plots

@@
 Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest.
-This study uses data from 1-ha fragments and continuous forest sites.
-<!--# maria: justify why 10ha plots omitted or don't say they exist -->
+This study uses data only from 1-ha fragments and continuous forest sites because these represent extremes and because there are only three 10 ha fragment plots.
 

### 2021-05-25 README.Rmd
got equation numbering and in-text references working using pandoc-crossref

@@
 )
-# Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
 ```
@@
 
-This repository contains the data and code for our paper:
+This repository contains the data and code for a manuscript written for submission to Global Change Biology tentatively titled: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
 
-> Authors, (YYYY).
-> *`r Title`*.
-> Name of journal/book <https://doi.org/xxx/xxx>
-
-<!-- Our pre-print is online here: -->
-
-<!-- > Authors, (YYYY). *`r Title`*. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
-
-### How to cite
-
-Please cite this compendium as:
-
-> Authors, (`r format(Sys.Date(), "%Y")`).
-> *Compendium of R code and data for `r Title`*.
-> Accessed `r format(Sys.Date(), "%d %b %Y")`.
-> Online at <https://doi.org/xxx/xxx>
-
-## How to run in your browser or download and run locally
+## How to run
 
@@
 To work with the compendium, you will need installed on your computer the [R software](https://cloud.r-project.org/) itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
-
-<!-- The simplest way to explore the text, code and data is to click on [binder](https://mybinder.org/v2/gh/BrunaLab/HeliconiaDemography/master?urlpath=rstudio) to open an instance of RStudio in your browser, which will have the compendium files ready to work with.
--->
-
-<!-- Binder uses rocker-project.org Docker images to ensure a consistent and reproducible computational environment. -->
-
-<!-- These Docker images can also be used locally. -->
 
@@
 
--   Open the `.Rproj` file in RStudio
--   Install the `renv` package if it's not already installed and run `renv::restore()` to install all package dependencies.
-(I'm not actually sure that this step is necessar.
-Should auto-load because of .Rprofile.)
--   Run `targets::tar_make()` to run all code and produce all outputs.
+-   Open the project in RStudio by double-clicking the `HeliconiaDemography.Rproj` file.
+-   Install the [`renv` package](https://rstudio.github.io/renv/articles/renv.html) if it's not already installed and run `renv::restore()` to install all package dependencies.
+-   Install the [`targets` package](https://docs.ropensci.org/targets/).
+-   Run `targets::tar_make()` from the R console to run all code and produce all outputs.
 
+## Notes
+
+-   Numbering of and in-text references to equations in `paper.Rmd` is handled by [`pandoc-crossref`](https://github.com/lierdakil/pandoc-crossref), which may not be part of the standard RStudio pandoc installation.
+You might need to install it for the main text to knit correctly to Word.
+
+<!--
 ### Licenses
-
+-->
 <!-- **Text and figures :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/) -->

### 2021-05-25 doc/paper.Rmd
fix figure reference and comment out unused text

@@
 
-```{r surv_curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
+```{r surv-curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
 surv_curve + theme(text = element_text(size = 10))
@@
 
-# Left-over text
+<!-- # Left-over text -->
 
--   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
--   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is.
+<!-- -   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
+-->
+
+<!-- -   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is.
 It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
--   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975
--   eg.
-gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007)
--   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
+-->
+
+<!-- -   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975 -->
+
+<!-- -   eg.
+gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007) -->
+
+<!-- -   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
 Estimates green-up, literfall, growth, etc.
@@
 During drought, old leaves are shed and new leaves are maintained.
-2-5 months for newly flushed leaves to fully expand
+2-5 months for newly flushed leaves to fully expand -->
 
-# Citations / information could allocate to appropriate paragraphs
+<!-- # Citations / information could allocate to appropriate paragraphs -->
 
--   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
-can be affected
+<!-- -   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
+can be affected -->
 
--   Effects of precip extremes on understory plants can be also be indirect:
+<!-- -   Effects of precip extremes on understory plants can be also be indirect: -->
 
-    -   high precip = lots of clouds, less photosynthesis, less or slower growth
+<!--     -   high precip = lots of clouds, less photosynthesis, less or slower growth -->
 
-    -   low precip = trees lose leaves or die, more light gets through or gaps open.
-        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
-        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020
+<!--     -   low precip = trees lose leaves or die, more light gets through or gaps open. -->
 
--   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet
+<!--         This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
+-->
 
--   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas.
-    Might be due to increases solar irradiance.
-    Maybe evidence of lagged effects?
-    Table 2 is inconsistent with text, so not sure...
+<!--         Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020 -->
 
--   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI
+<!-- -   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet -->
 
--   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+<!-- -   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas. -->
 
--   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought.
-    Similar increase in continuous and (large) fragments.
-    Increase in turnover more variable in fragments.
-    Coarse woody debris increased 62% during drought.
+<!--     Might be due to increases solar irradiance. -->
 
-\newpage
+<!--     Maybe evidence of lagged effects? -->
+
+<!--     Table 2 is inconsistent with text, so not sure... -->
+
+<!-- -   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI -->
+
+<!-- -   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
+-->
+
+<!-- -   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought. -->
+
+<!--     Similar increase in continuous and (large) fragments. -->
+
+<!--     Increase in turnover more variable in fragments. -->
+
+<!--     Coarse woody debris increased 62% during drought. -->
+
+<!-- \newpage -->
 

### 2021-06-08 doc/paper.Rmd
update methods and results text with individual level random effects models. #61

@@
 We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
+Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
+Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
+1 A random effect of plant ID was not included in survival models since each plant only dies once.
 
@@
 
-While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.9 for 1-ha fragments; edf = 12.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
 For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
@@
 Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
-The responses in 1-ha fragments were also more muted (1-ha edf = 8.1, continuous forest edf = 10.3), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
+The responses in 1-ha fragments were also more muted (1-ha edf = 8.3, continuous forest edf = 10.6), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
 This led to some important inter-habitat differences in plant responses to prior droughts.
@@
 
-Finally, with the exception of survival in 1-ha fragments ($p = 0.252$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.001$ for the random effect of plot).
+Finally, with the exception of survival in 1-ha fragments ($p = 0.253$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.01$ for the random effect of plot).
 We found no evidence for a cost of reproduction.
-In both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$, 1-ha: $p = 0.031$) and flowering ($p < 0.001$ for both CF and 1-ha) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
+In both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flowering (CF: $p < 0.001$; 1-ha: $p = 0.004$) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
+The random effect of plant ID was also significant for flowering probability (CF: $p < 0.001$; 1-ha: $p=0.003$) indicating significant individual-level variation in flowering probability.
 

### 2021-06-08 doc/paper.Rmd
added figures from hipergator run with ind-level random effects. #61

@@
 ```{r size-plot, fig.cap=size_cap, fig.width=7, fig.height=7}
-size_plot & theme(text = element_text(size = 10))
+# size_plot & theme(text = element_text(size = 10))
+include_graphics(here("doc", "figures", "size_plot_ufhpc.png"))
 ```
@@
 ```{r flwr-spei, fig.cap=f_spei_cap, fig.height=7.5, fig.width=7}
-f_spei_plot & theme(text = element_text(size = 10))
+# f_spei_plot & theme(text = element_text(size = 10))
+include_graphics(here("doc", "figures", "f_spei_plot_ufhpc.png"))
 ```

### 2021-06-14 doc/paper.Rmd

@@
 
-Climate anomalies are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
+Climate anomalies are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stilesEcologyFloweringPhenology1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
 These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation [@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].

### 2021-06-14 doc/paper.Rmd
only load targets when needed. (#64)

@@
   tar_load(c(
-    eda_plot,
-    surv_curve,
-    s_spei_plot,
-    g_spei_plot,
-    f_spei_plot,
-    size_plot,
-    f_cf,
-    f_1ha,
+    # eda_plot,
+    # surv_curve,
+    # s_spei_plot,
+    # g_spei_plot,
+    # f_spei_plot,
+    # size_plot,
     model_data,
@@
 
-```{r flwr-prob-quartile, include=FALSE}
-
-#figure out flowering probability of largest quartile of plants in each habitat
-library(tidyverse)
-newdf_cf <-
-  f_cf$model %>% #use model data frame
-  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
-  filter(quartiles == 4)
-
-newdf_1ha <-
-  f_1ha$model %>% 
-  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
-  filter(quartiles == 4)
-
-#equivalent to flowering once every __ years
-1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
-1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
-```
-
 ## Conclusions & Future Directions
@@
 ```{r eda, fig.cap=eda_cap, fig.width=7, fig.height=7}
-eda_plot & theme(text = element_text(size = 10))
+tar_read(eda_plot) & theme(text = element_text(size = 10))
 ```
@@
 ```{r surv-curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
-surv_curve + theme(text = element_text(size = 10))
+tar_read(surv_curve) + theme(text = element_text(size = 10))
 ```
@@
 ```{r size-plot, fig.cap=size_cap, fig.width=7, fig.height=7}
-# size_plot & theme(text = element_text(size = 10))
-include_graphics(here("doc", "figures", "size_plot_ufhpc.png"))
+tar_read(size_plot) & theme(text = element_text(size = 10))
 ```
@@
 ```{r surv-spei, fig.cap=s_spei_cap, fig.height=7.5, fig.width=7}
-s_spei_plot & theme(text = element_text(size = 10))
+tar_read(s_spei_plot) & theme(text = element_text(size = 10))
 ```
@@
 ```{r growth-spei, fig.cap=g_spei_cap, fig.height=7.5, fig.width=7}
-g_spei_plot & theme(text = element_text(size = 10))
+tar_read(g_spei_plot) & theme(text = element_text(size = 10))
 ```
@@
 ```{r flwr-spei, fig.cap=f_spei_cap, fig.height=7.5, fig.width=7}
-# f_spei_plot & theme(text = element_text(size = 10))
-include_graphics(here("doc", "figures", "f_spei_plot_ufhpc.png"))
+tar_read(f_spei_plot) & theme(text = element_text(size = 10))
 ```

### 2021-06-14 doc/validate_models.Rmd
only load targets when needed. (#64)

@@
 
-```{r load-targets}
-withr::with_dir(here::here(),{
-  tar_load(c(
-    s_cf,
-    s_1ha,
-    g_cf,
-    g_1ha,
-    f_cf,
-    f_1ha,
-    s_cf_sub,
-    g_cf_sub,
-    f_cf_sub
-    ))
-  })
-```
-
 # Analysis
@@
 ### Fragments
+```{r}
+withr::with_dir(here::here(),{
+  tar_load(c(s_1ha, s_cf, s_cf_sub))
+})
+```
 
 ```{r}
+
 summary(s_1ha)
@@
 Surface still looks different in a similar way though.
+```{r}
+rm(s_1ha, s_cf, s_cf_sub)
+```
 
 ## Growth
+
+```{r}
+withr::with_dir(here::here(),{
+  tar_load(c(g_1ha, g_cf, g_cf_sub))
+})
+```
 
@@
 
+```{r}
+rm(g_1ha, g_cf, g_cf_sub)
+```
+
 ## Flowering
@@
 Quantile residuals are more appropriate for binomial models.
+
+```{r}
+withr::with_dir(here::here(),{
+  tar_load(c(f_1ha, f_cf, f_cf_sub))
+})
+```
 
@@
 
+### Flowering prob of largest plants
+
+Figure out flowering probability of largest quartile of plants in each habitat
+
+```{r}
+newdf_cf <-
+  f_cf$model %>% #use model data frame
+  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
+  filter(quartiles == 4)
+
+newdf_1ha <-
+  f_1ha$model %>% 
+  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
+  filter(quartiles == 4)
+```
+
+```{r}
+#equivalent to flowering once every __ years
+1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
+1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
+```
+
+```{r}
+rm(f_1ha, f_cf, f_cf_sub)
+```
+
 # Reproducibility

### 2021-06-14 doc/paper.Rmd
add placeholders for additional required sections

@@
 title: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
+shorttitle: "Delayed effects of climate on vital rates"
 author:
@@
 
+# Data Availability Statement
+
+All data and code used in this study are available at [insert zenodo link].
+
 # Figures

### 2021-06-14 doc/paper.Rmd
added abstract

@@
 abstract: |
-  Text of abstract. 34,281 year x individual measurements! 
+  Deforestation is a major threat to species biodiversity in the Amazon rainforest.
+Deforestation results in loss of habitat, but also often leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
+Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood.
+It is often hypothesized that the dramatically altered environmental conditions in fragments are driving declines in reproduction, recruitment, or survivorship.
+In addition to deforestation, the Amazon is facing changes in precipitation events due to climate change with some regions predicted to face increasingly intense and more frequent droughts and other parts of the basin predicted to see more frequent and severe high precipitation events.
+These changes can have detrimental consequences for plant survival and reproduction.
+Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear.
+Most studies of plants in fragments are relatively short (1–6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest.
+Even fewer of these studies take into account the possibility of delayed effects of climate despite the prevalence of such delayed effects on plant demographic vital rates in studies that look for them.
+Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (Heliconia acuminata, Heliconiaceae).
+We modeled potentially delayed and non-linear effects of climate on plant vital rates without *a priori* assumptions about climate windows using distributed lag non-linear models.
+  We found that delayed detrimental effects of extremes in precipitation on survival were more intense in forest fragments compared to continuous forest.
+The size transitions of plants in forest fragments was also more influenced by climate compared to continuous forest, with the direction of the effect depending on the lag time and season.
+Flowering probability was size dependent and much lower in forest fragments due to smaller average plant size.
+In continuous forest, flowering probability increased during droughts and when there was drought two dry seasons prior.
+The complex delayed effects of climate and habitat fragmentation on H.
+acuminata vital rates points to the importance of long term demography experiments in understanding the effects of anthropogenic change on plant populations.
 keywords: |
@@
 Keywords: `r rmarkdown::metadata$keywords`
+
+Running title: `r rmarkdown::metadata$shorttile`
 

### 2021-06-15 _targets.R
run all models on HiperGator sucessfully

@@ -1,9 +1,15 @@
 ## Set options for using tar_make_clustermq()
+## Uncomment and edit these lines to run model targets on a cluster computer via SSH
 options(
   clustermq.scheduler = "ssh",
   clustermq.ssh.host = "ericscott@hpg.rc.ufl.edu", # use your user and host
   clustermq.ssh.log = "~/cmq_ssh.log" # log for easier debugging
 )
+
+## Uncomment these lines to run locally on multiple cores
+# options(
+#   clustermq.schedule = "multicore"
+# )
 
 ## Load your packages, e.g. library(targets).
 source("./packages.R")
@@ -41,7 +47,10 @@ tar_plan(
   tar_target(s_1ha, fit_surv(model_data_1ha), deployment = "worker"),
   tar_target(g_cf, fit_growth(model_data_cf), deployment = "worker"),
   tar_target(g_1ha, fit_growth(model_data_1ha), deployment = "worker"),
-  tar_target(f_cf, fit_flwr(model_data_cf), deployment = "worker"),
+  
+  #### NOTE: The f_cf target takes ~2 hrs to run on a single core on the cluster
+  #### with ind_raneff = TRUE.
+  tar_target(f_cf, fit_flwr(model_data_cf, ind_raneff = TRUE), deployment = "worker"),
   tar_target(f_1ha, fit_flwr(model_data_1ha, ind_raneff = TRUE), deployment = "worker"),
   
   # Validate and summarize results

### 2021-06-17 doc/paper.Rmd
added methods ref to 85% CI overlap
test out 84% CI inline code

@@
 We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
+Significant main effects of habitat type were assessed by looking for overlap in the 84% confidence intervals of model intercepts (84% CIs of two samples drawn from the same population overlap about 95% of the time [@paytonOverlappingConfidenceIntervals2003]).
 
@@
 
-When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
 The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
+Survival probability did not differ significantly between continuous forest and 1 ha fragments (estimate and 84% CI were `r pull_ci(tar_read(s_cf))` , and `r pull_ci(tar_read(s_1ha))` for continuous forest and 1 ha fragments, respectively).
 Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).

### 2021-06-17 R/pull_ci.R

@@ -1,0 +1,15 @@
+#' Get CI around parametric terms of gam
+#'
+#' @param model a gam
+#' @param conf.level confidence level. Default is 0.84
+#'
+#' @return formatted estimate [lower, upper]
+#'
+pull_ci <- function(model, conf.level = 0.84) {
+  linkinv <- model$family$linkinv
+  ci <-
+    broom::tidy(model, conf.int = TRUE, parametric = TRUE, conf.level = conf.level) %>%
+    mutate(across(c(estimate, conf.low, conf.high), ~linkinv(.x) %>% round(2))) %>%
+    mutate(ci = glue::glue("{estimate} [{conf.low}, {conf.high}]"))
+  return(ci$ci[1])
+}

### 2021-06-17 R/pull_ci.R
refactored pull_ci()

@@ -5,11 +5,17 @@
 #'
 #' @return formatted estimate [lower, upper]
 #'
-pull_ci <- function(model, conf.level = 0.84) {
+pull_ci <- function(model, conf.level = 0.84, digits = 3, type = c("response", "link")) {
   linkinv <- model$family$linkinv
-  ci <-
-    broom::tidy(model, conf.int = TRUE, parametric = TRUE, conf.level = conf.level) %>%
-    mutate(across(c(estimate, conf.low, conf.high), ~linkinv(.x) %>% round(2))) %>%
-    mutate(ci = glue::glue("{estimate} [{conf.low}, {conf.high}]"))
-  return(ci$ci[1])
+  type <- match.arg(type)
+  intercept <- coef(model)[1]
+  ci <- confint.default(model, parm = "(Intercept)", level = conf.level)
+  if (type == "response") {
+    intercept <- linkinv(intercept)
+    ci <- linkinv(ci)
+  }
+  intercept <- round(intercept, digits)
+  ci <- round(ci, digits)
+  ci_string <- glue::glue("{intercept} [{ci[1]}, {ci[2]}]")
+  return(ci_string)
 }

### 2021-06-17 doc/paper.Rmd
added text about 84% CI overlap of intercepts

@@
 summary_stats
-
 model_data %>% 
@@
 
+```{r}
+#get 84% CI around intercepts
+withr::with_dir(here::here(), {
+  tar_load(c(s_cf, s_1ha))
+})
+source(here::here("R", "pull_ci.R"))
+pull_ci(s_cf)
+pull_ci(s_1ha)
+rm(s_cf, s_1ha)
+```
+
 The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-Survival probability did not differ significantly between continuous forest and 1 ha fragments (estimate and 84% CI were `r pull_ci(tar_read(s_cf))` , and `r pull_ci(tar_read(s_1ha))` for continuous forest and 1 ha fragments, respectively).
+Survival probability did not differ significantly between continuous forest and 1 ha fragments (intercept estimate and 84% CI were 0.972 [0.967, 0.976] and 0.966 [0.961, 0.971] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
 Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
@@
 However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
+
+```{r}
+#get 84% CI around intercepts
+withr::with_dir(here::here(), {
+  tar_load(c(g_cf, g_1ha))
+})
+pull_ci(g_cf, digits = 2)
+pull_ci(g_1ha, digits = 2)
+rm(g_cf, g_1ha)
+```
 
@@
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+However, the main effect of habitat on log(size) was not significant (intercept and 84% CIs were 4.41 [4.38, 4.44] and 4.36 [4.22, 4.5] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
 As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
@@
 
+```{r}
+#get 84% CI around intercepts
+withr::with_dir(here::here(), {
+  tar_load(c(f_cf, f_1ha))
+})
+pull_ci(f_cf, digits = 4)
+pull_ci(f_1ha, digits = 4)
+rm(f_cf, f_1ha)
+```
+
 While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
@@
 only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).
+However, the main effect of habitat on flowering probability was not significant (intercept and 84% CIs were 0.0059 [0.0042, 0.0082] and 0.0033 [0.0016, 0.007] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
 

### 2021-06-17 doc/paper.Rmd
edited results for survival

@@
 
-For 1-ha fragments, SPEI in the preceding 12 months had the strongest effect on survival, with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)a).
-In contrast, the effect of recent SPEI in continuous forest was both weaker and unidirectional---survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)b).
+For 1-ha fragments, SPEI in the preceding 13 months had a significant effect on survival, with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
+Wet conditions (SPEI \> 1) in the preceeding 23 months had a significant negative effect on survival in 1 ha fragments (Figure \@ref(fig:surv-spei)b).
+In contrast, the effect of recent SPEI in continuous forest was weaker and significant for a shorter window of only the preceding 6 months (Figure \@ref(fig:surv-spei)a).
+The effect of SPEI in continuous forest also differed in that it was unidirectional---survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)a).
 Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census) and 32--36 months.
-Drought (i.e., SPEI \< -1) 15--20 months prior to a census was associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was associated with higher survival (put another way, there was a nearly linear relationship between SPEI and survival probability).
-Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 36 months.
+Drought (i.e., SPEI \< -1) 15--20 months prior to a census was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival (put another way, there was a nearly linear relationship between SPEI and survival probability).
+Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
 It should be noted, however, that only the first year of census data (1999) met these conditions.
 We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
-This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
+This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c) <!--# is this where I should mention the overlap in 84%CI around the intercepts?
+-->.
 However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.

### 2021-06-17 doc/paper.Rmd
add 95%ci to results for flowering

@@
 
-Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)).
-The responses in 1-ha fragments were also more muted (1-ha edf = 8.3, continuous forest edf = 10.6), as indicated by the shape of the crossbasis function (Figure \@ref(fig:flwr-spei)a).
+Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CI of intercepts overlaps).
+The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).
 This led to some important inter-habitat differences in plant responses to prior droughts.
-Recent drought (i.e., at lag = 0 with SPEI \< -1) increased the probability of flowering in continuous forest, as did drought at lags 15--20 (i.e., two dry seasons prior).
-However, drought at lags 7--13 (the end of the rainy season one year prior) reduced flowering probability in continuous forest far more than it did in fragments (Figure \@ref(fig:flwr-spei)c).
+In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20 ) and in the wet season 34--36 months prior all increased the probability of flowering.
+However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
+In Fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
+The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
 
 Finally, with the exception of survival in 1-ha fragments ($p = 0.253$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.01$ for the random effect of plot).
-We found no evidence for a cost of reproduction.
-In both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flowering (CF: $p < 0.001$; 1-ha: $p = 0.004$) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
+We found no evidence for a cost of reproduction: in both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flowering (CF: $p < 0.001$; 1-ha: $p = 0.004$) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
 The random effect of plant ID was also significant for flowering probability (CF: $p < 0.001$; 1-ha: $p=0.003$) indicating significant individual-level variation in flowering probability.

### 2021-06-17 doc/paper.Rmd
uninculde code chunks

@@
 
-```{r}
+```{r include=FALSE}
 #get 84% CI around intercepts

### 2021-06-18 README.Rmd
run all models on HiperGator sucessfully
update readme

@@
 -   Install the [`targets` package](https://docs.ropensci.org/targets/).
--   Run `targets::tar_make()` from the R console to run all code and produce all outputs.
+-   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
 
-## Notes
+**NOTE**: The full pipeline will take a *very long* time to run.
+I suggest switching off individual-level random effects (more details in comments in `_targets.R`) if you want to run the project on your own computer.
+Without individual-level random effects the pipeline should run relatively quickly (< 1hr, I think).
+
+## Additional Dependencies
 

### 2021-06-20 doc/paper.Rmd
refactored summary code, revised writing about 84%CIs

@@
 
-Running title: `r rmarkdown::metadata$shorttile`
+Running title: `r rmarkdown::metadata$shorttitle`
 
@@
 ```{r include=FALSE}
-summary_stats <-
+means_sds <-
   model_data %>%
   group_by(habitat) %>%
-  summarize(across(c(ht, shts, size, flwr, surv),
+  summarize(
+    across(c(ht, shts, size, surv, flwr),
            .fns = list(mean = mean, sd = sd),
            na.rm = TRUE),
-            n_plants = length(unique(ha_id_number))) %>%
-  # right_join(model_data %>% count(habitat)) %>%
+    n_plants = length(unique(ha_id_number))
+  ) %>%
   mutate(across(ends_with("_mean") | ends_with("_sd"),
                 ~round(.x, 2)))
-summary_stats <- 
-  full_join(
-  summary_stats,
-  model_data %>%
+
+counts <- model_data %>%
   count(habitat, plot) %>%
@@
   mutate(across(where(is.numeric), round))
+
+summary_stats <- 
+  full_join(
+    means_sds,
+    counts
   )
-summary_stats
+
+```
+
+```{r include=FALSE}
+plants_per_plot <-
   model_data %>% 
@@
   summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants))
+plants_per_plot
+```
 
-by_year <-
+```{r include=FALSE}
+surv_by_year_habitat <-
   model_data %>% 
@@
     arrange(prop_surv)
-by_year
+surv_by_year_habitat
 
-xa_spei %>% ungroup() %>%
+surv_by_year <-
+  model_data %>% 
+  group_by(year) %>% 
+  summarize(prop_surv = mean(surv)) %>% 
+    arrange(prop_surv)
+surv_by_year
+```
+
+```{r include=FALSE}
+spei_2005 <- 
+  xa_spei %>%
+  ungroup() %>%
   summarize(across(everything(), mean)) %>%
   filter(year(yearmonth)==2005)
-
+spei_2005
 ```
@@
 
-The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
+*Survival:* The proportion of surviving plants across all plots was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d); it was also preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-Survival probability did not differ significantly between continuous forest and 1 ha fragments (intercept estimate and 84% CI were 0.972 [0.967, 0.976] and 0.966 [0.961, 0.971] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
-Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
+The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
+When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv}=0.95$; Figure \@ref(fig:eda)b), with 85% CI of model intercepts overlapping, indicating no significant difference between the two habitat types.
+However, survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
 While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.
-However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.
 
@@
 
-Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
+*Growth:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
 Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
 Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were on average substantially larger than those in forest fragments (150 ± 175 SD vs.
-112 ± 141 SD, respectively).
+112 ± 141 SD, respectively), however, size was did not differ significantly between habitats (overlap in 84% CI of model intercepts).
 The distribution of sizes of individuals was somewhat bimodal in both habitats and the major difference in plant size was in the larger of the two modes being smaller in fragments (Figure \@ref(fig:size-plot)d).
@@
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-However, the main effect of habitat on log(size) was not significant (intercept and 84% CIs were 4.41 [4.38, 4.44] and 4.36 [4.22, 4.5] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
 As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
@@
   filter(size >= 168) %>% 
+    filter(surv == 1) %>% 
     group_by(year, habitat) %>% 
@@
 
-While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
+*Flowering:* While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), although this difference did not reach statistical significance (84% CIs of model intercepts overlapped).
 This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
@@
 only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).
-However, the main effect of habitat on flowering probability was not significant (intercept and 84% CIs were 0.0059 [0.0042, 0.0082] and 0.0033 [0.0016, 0.007] for continuous forest and 1 ha fragments, respectively) <!--# not sure if this is where this goes -->.
 

### 2021-06-20 doc/paper.Rmd
typo

@@
 We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
-This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c) <!--# is this where I should mention the overlap in 84%CI around the intercepts?
--->.
+This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
 However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.
@@
 
-Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CI of intercepts overlaps).
+Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
 The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).

### 2021-06-20 doc/paper.Rmd
moved abstact out of YAML into body

@@
 csl: "templates/global-change-biology.csl" # Insert path for the bib-style
-abstract: |
-  Deforestation is a major threat to species biodiversity in the Amazon rainforest.
-Deforestation results in loss of habitat, but also often leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
-Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood.
-It is often hypothesized that the dramatically altered environmental conditions in fragments are driving declines in reproduction, recruitment, or survivorship.
-In addition to deforestation, the Amazon is facing changes in precipitation events due to climate change with some regions predicted to face increasingly intense and more frequent droughts and other parts of the basin predicted to see more frequent and severe high precipitation events.
-These changes can have detrimental consequences for plant survival and reproduction.
-Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear.
-Most studies of plants in fragments are relatively short (1–6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest.
-Even fewer of these studies take into account the possibility of delayed effects of climate despite the prevalence of such delayed effects on plant demographic vital rates in studies that look for them.
-Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (Heliconia acuminata, Heliconiaceae).
-We modeled potentially delayed and non-linear effects of climate on plant vital rates without *a priori* assumptions about climate windows using distributed lag non-linear models.
-  We found that delayed detrimental effects of extremes in precipitation on survival were more intense in forest fragments compared to continuous forest.
-The size transitions of plants in forest fragments was also more influenced by climate compared to continuous forest, with the direction of the effect depending on the lag time and season.
-Flowering probability was size dependent and much lower in forest fragments due to smaller average plant size.
-In continuous forest, flowering probability increased during droughts and when there was drought two dry seasons prior.
-The complex delayed effects of climate and habitat fragmentation on H.
-acuminata vital rates points to the importance of long term demography experiments in understanding the effects of anthropogenic change on plant populations.
+# abstract: |
 keywords: |
@@
 <!-- The actual document text starts here: -->
+
+# Abstract
+
+Deforestation is a major threat to species biodiversity in the Amazon rainforest.
+Deforestation results in loss of habitat, but also often leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
+Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood.
+It is often hypothesized that the dramatically altered environmental conditions in fragments are driving declines in reproduction, recruitment, or survivorship.
+In addition to deforestation, the Amazon is facing changes in precipitation events due to climate change with some regions predicted to face increasingly intense and more frequent droughts and other parts of the basin predicted to see more frequent and severe high precipitation events.
+These changes can have detrimental consequences for plant survival and reproduction.
+Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear.
+Most studies of plants in fragments are relatively short (1--6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest.
+Even fewer of these studies take into account the possibility of delayed effects of climate despite the prevalence of such delayed effects on plant demographic vital rates in studies that look for them.
+Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (Heliconia acuminata, Heliconiaceae).
+We modeled potentially delayed and non-linear effects of climate on plant vital rates without *a priori* assumptions about climate windows using distributed lag non-linear models.
+
+We found that delayed detrimental effects of extremes in precipitation on survival were more intense in forest fragments compared to continuous forest.
+The size transitions of plants in forest fragments was also more influenced by climate compared to continuous forest, with the direction of the effect depending on the lag time and season.
+Flowering probability was size dependent and much lower in forest fragments due to smaller average plant size.
+In continuous forest, flowering probability increased during droughts and when there was drought two dry seasons prior.
+The complex delayed effects of climate and habitat fragmentation on H.
+acuminata vital rates points to the importance of long term demography experiments in understanding the effects of anthropogenic change on plant populations.
 

### 2021-06-20 doc/paper.Rmd
edit figure captions

@@
 
+\pagebreak
+
 # Abstract
 
-Deforestation is a major threat to species biodiversity in the Amazon rainforest.
-Deforestation results in loss of habitat, but also often leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
+Deforestation is a major threat to biodiversity in the Amazon, partly because it leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
 Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood.
-It is often hypothesized that the dramatically altered environmental conditions in fragments are driving declines in reproduction, recruitment, or survivorship.
-In addition to deforestation, the Amazon is facing changes in precipitation events due to climate change with some regions predicted to face increasingly intense and more frequent droughts and other parts of the basin predicted to see more frequent and severe high precipitation events.
-These changes can have detrimental consequences for plant survival and reproduction.
+It is often hypothesized that the altered environmental conditions in fragments drive declines in reproduction, recruitment, or survivorship.
+The Amazon basin, in addition to experiencing continuing fragmentation, is warming and experiencing changes in precipitation leading to altered frequency and intensity of droughts and unusually wet periods.
 Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear.
 Most studies of plants in fragments are relatively short (1--6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest.
-Even fewer of these studies take into account the possibility of delayed effects of climate despite the prevalence of such delayed effects on plant demographic vital rates in studies that look for them.
-Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (Heliconia acuminata, Heliconiaceae).
-We modeled potentially delayed and non-linear effects of climate on plant vital rates without *a priori* assumptions about climate windows using distributed lag non-linear models.
-
-We found that delayed detrimental effects of extremes in precipitation on survival were more intense in forest fragments compared to continuous forest.
-The size transitions of plants in forest fragments was also more influenced by climate compared to continuous forest, with the direction of the effect depending on the lag time and season.
-Flowering probability was size dependent and much lower in forest fragments due to smaller average plant size.
-In continuous forest, flowering probability increased during droughts and when there was drought two dry seasons prior.
-The complex delayed effects of climate and habitat fragmentation on H.
-acuminata vital rates points to the importance of long term demography experiments in understanding the effects of anthropogenic change on plant populations.
+Even fewer studies consider delayed effects of climate despite the importance of such delayed effects for plant demographic vital rates in studies that look for them.
+Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (*Heliconia acuminata*, Heliconiaceae).
+We used distributed lag non-linear models to understand the delayed effects of temperature and precipitation on survival, growth, and flowering.
+We detected delayed effects of climate up to 36 months.
+Drought two dry seasons prior to the February census decreased survival and increased flowering probability while drought in the wet season a year prior to the census decreased flowering probability and increased growth.
+The effects of extremes in precipitation on survival and growth were more intense in forest fragments compared to continuous forest.
+The complex delayed effects of climate and habitat fragmentation on *H.
+acuminata* vital rates points to the importance of long-term demography experiments in understanding the effects of anthropogenic change on plant populations.
 
@@
 
-```{r include=FALSE}
+```{r eval=FALSE, include=FALSE}
 #get 84% CI around intercepts
@@
 Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Contour lines correspond to a change of 0.002.
+Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
 The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
@@
 Model intercepts were added to fitted values of the crossbasis function.
-Contour lines correspond to a change of 0.05.
+Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
 The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
@@
 Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Contour lines correspond to a change of 0.001.
+Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
 The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."

### 2021-06-22 doc/paper.Rmd
through results (survival, lag)

@@
 ---
-title: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
+title: "Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments"
 shorttitle: "Delayed effects of climate on vital rates"
@@
 
-Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
-While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
+Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with effects that are more pronounced on fragmented populations [@oliver2015; @holyoak2016].
+While the demographic consequences are expected to be similarly severe in the tropics [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for species found in tropical biomes [@paniw2021].
 Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
@@
 
-Whether the demography of plant populations in tropical forest fragments is more susceptible to drought and other climatic extremes remains unclear for three primary reasons.
+Whether the demography of plant populations in tropical forest fragments is more susceptible to climatic extremes remains unclear for three primary reasons.
 First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
-Similarly, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
+Second, there is a growing literature on how tropical plants respond to droughts [@esquivel-muelbertCompositionalResponseAmazon2019; @gonzalez-mDivergingFunctionalStrategies2020; @uriarteImpactsClimateVariability2016], but few studies have compared the responses of plants in continuous forest with those of plants in forest fragments [@lauranceEffectsStrongDrought2001].
 Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
@@
 They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [*e.g.,* @gagnonGrowthUnderstoryHerb2011].
-Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].
+Indeed, lagged responses of demographic vital rates to climate may be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].
 
@@
 These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
-In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H.
+acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
+bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 
-In 1997--1998 we established a series of 5000 m^2^ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
-The plots are located in continuous forest ($n=6$), 10 ha fragments ($n=3$), and 1-ha fragments ($n=4$), with distance between plots ranging from 500 m--41 km.
-Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest.
-This study uses data only from 1-ha fragments and continuous forest sites because these represent extremes and because there are only three 10 ha fragment plots.
+```{r include=FALSE}
+means_sds <-
+  model_data %>%
+  group_by(habitat) %>%
+  summarize(
+    across(c(ht, shts, size, surv, flwr),
+           .fns = list(mean = mean, sd = sd),
+           na.rm = TRUE),
+    n_plants = length(unique(ha_id_number))
+  ) %>%
+  mutate(across(ends_with("_mean") | ends_with("_sd"),
+                ~round(.x, 2)))
 
-Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
+counts <- model_data %>%
+  count(habitat, plot) %>%
+  group_by(habitat) %>%
+  summarize(n_mean = mean(n), n_sd = sd(n)) %>%
+  mutate(across(where(is.numeric), round))
+
+summary_stats <- 
+  full_join(
+    means_sds,
+    counts
+  )
+
+```
+
+```{r include=FALSE}
+plants_per_plot <-
+  model_data %>% 
+  group_by(habitat, plot) %>% 
+  summarize(n_plants = length(unique(ha_id_number))) %>% 
+  group_by(habitat) %>% 
+  summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants))
+plants_per_plot
+```
+
+This study uses data collected in four 1-ha fragment reserves and six continuous forest sites.
+In 1997--1998 we established a 5000 m^2^ plots ($50 \times 100$m) in each of these sites in which we marked and measured all *Heliconia acuminata*; plots in 1-ha fragments were on one randomly selected half of the fragment, while plots in continuous forest were located 500--4000 m from the borders of secondary and mature forest.
+The distance between plots ranged from 500 m--41 km.
+Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
+Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).
+
+Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
 In this study, we used the product of shoot number and plant height as our measure of plant size.
@@
 Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
-SPEI calculations were made using the `SPEI` package[@begueriaSPEICalculationStandardised2017].
+SPEI calculations were made using the `SPEI` package [@begueriaSPEICalculationStandardised2017].
 The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought sometimes differed (Figure S2; Table S1 <!--# update crossrefs manually -->).
@@
 
-To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
+To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs; @gasparriniPenalizedFrameworkDistributed2017].
 DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
@@
 
-The `dlnm` package does not currently allow modeling of factor by smooth interactions.
-This means we could not include habitat as an interaction term.
+The `dlnm` package does not currently allow the modeling of interaction terms, which means we could not asses the interaction of habitat type and lagged effects.
 We therefore fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
-Significant main effects of habitat type were assessed by looking for overlap in the 84% confidence intervals of model intercepts (84% CIs of two samples drawn from the same population overlap about 95% of the time [@paytonOverlappingConfidenceIntervals2003]).
+Significant main effects of habitat type were assessed by looking for overlap in the 84% confidence intervals of model intercepts; the 84% CIs of two samples drawn from the same population overlap about 95% of the time [@paytonOverlappingConfidenceIntervals2003].
 
@@
 
-Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
+The meteorological droughts in our focal region indicated by SPEI are generally consistent with those reported in the literature (Table S1).
 For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).
@@
 ## Survival, growth, and flowering in continuous forest vs. fragments
-
-```{r include=FALSE}
-means_sds <-
-  model_data %>%
-  group_by(habitat) %>%
-  summarize(
-    across(c(ht, shts, size, surv, flwr),
-           .fns = list(mean = mean, sd = sd),
-           na.rm = TRUE),
-    n_plants = length(unique(ha_id_number))
-  ) %>%
-  mutate(across(ends_with("_mean") | ends_with("_sd"),
-                ~round(.x, 2)))
-
-counts <- model_data %>%
-  count(habitat, plot) %>%
-  group_by(habitat) %>%
-  summarize(n_mean = mean(n), n_sd = sd(n)) %>%
-  mutate(across(where(is.numeric), round))
-
-summary_stats <- 
-  full_join(
-    means_sds,
-    counts
-  )
-
-```
-
-```{r include=FALSE}
-plants_per_plot <-
-  model_data %>% 
-  group_by(habitat, plot) %>% 
-  summarize(n_plants = length(unique(ha_id_number))) %>% 
-  group_by(habitat) %>% 
-  summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants))
-plants_per_plot
-```
 
@@
 
-Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
-Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).
-
 ```{r eval=FALSE, include=FALSE}
@@
 
-*Survival:* The proportion of surviving plants across all plots was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d); it was also preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
+*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
+This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was apreceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
 The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
-When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv}=0.95$; Figure \@ref(fig:eda)b), with 85% CI of model intercepts overlapping, indicating no significant difference between the two habitat types.
-However, survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
-While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), the survival of the smallest plants was higher in 1-ha fragments.
+When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
+However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
+The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.
 
@@
 *Growth:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
-Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
-Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were on average substantially larger than those in forest fragments (150 ± 175 SD vs.
-112 ± 141 SD, respectively), however, size was did not differ significantly between habitats (overlap in 84% CI of model intercepts).
-The distribution of sizes of individuals was somewhat bimodal in both habitats and the major difference in plant size was in the larger of the two modes being smaller in fragments (Figure \@ref(fig:size-plot)d).
-This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
+Plants in 1-ha fragments had on average `r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were `r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
+Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 34% larger than those in forest fragments (150 ± 175 SD vs.
+112 ± 141 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
+This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
+
 Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
@@
 
-*Flowering:* While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), although this difference did not reach statistical significance (84% CIs of model intercepts overlapped).
-This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
-Even though flowering probability of large plants was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive size threshold in forest fragments (Figure \@ref(fig:size-plot)d).
-Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
+*Flowering:* The overall proportion of plants that flowered was very low.
+While it was almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
+The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
+Despite the flowering probability of large plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
+The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
 only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).
@@
 
-For 1-ha fragments, SPEI in the preceding 13 months had a significant effect on survival, with the highest survival near SPEI of 0 and mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
-Wet conditions (SPEI \> 1) in the preceeding 23 months had a significant negative effect on survival in 1 ha fragments (Figure \@ref(fig:surv-spei)b).
-In contrast, the effect of recent SPEI in continuous forest was weaker and significant for a shorter window of only the preceding 6 months (Figure \@ref(fig:surv-spei)a).
-The effect of SPEI in continuous forest also differed in that it was unidirectional---survival probability was reduced only when SPEI was \< -1 (i.e., mild-severe drought), and even then only slightly (Figure \@ref(fig:surv-spei)a).
-Instead the most pronounced effects on the survival of plants in continuous forest were of SPEI at a lag time of 15--20 months (i.e. two dry seasons prior to a census) and 32--36 months.
-Drought (i.e., SPEI \< -1) 15--20 months prior to a census was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival (put another way, there was a nearly linear relationship between SPEI and survival probability).
+*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 13 months.
+The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
+Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
+In contrast, the effect of recent SPEI in continuous forest was weaker, with only the 6 months preceding a census having a significant effect on survival (Figure \@ref(fig:surv-spei)a).
+These short-term effects of SPEI on survival were also unidirectional---the probability of survival declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
+In contrast, the most pronounced negative effects of SPEI on the survival of plants in continuous forest were at lag times of 15--20 months and 32--36 months.
+Drought 15--20 months prior to a census (i.e.
+two dry seasons prior to a census) was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival.
 Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
@@
 
-While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+*Growth:* While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
 For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
@@
 
-Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
+*Flowering:* Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
 The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).

### 2021-06-22 doc/paper.Rmd
results and flowering

@@
 
-*Growth:* While the fitted crossbasis functions for fragments and continuous forest showed generally similar patterns of drought effects on growth (i.e, trends in plant size), the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
-For example, when SPEI = 0 (i.e., average conditions), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
-However, when the current wet season is unusually wet (SPEI\>2), plants in continuous forests were larger by up to log(size) = 0.57 due to reduced growth in fragments.
-In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, but the magnitude of the response was greater in 1-ha fragments.
+*Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
+However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
+However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
+In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a m,ore pronounced response in 1-ha fragments.
 
@@
 However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
-In Fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
+In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
 The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
+We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flower again (CF: $p < 0.001$; 1-ha: $p = 0.004$).
+The random effect of plant ID on flowering probability was also significant (CF: $p < 0.001$; 1-ha: $p=0.003$), indicating significant individual-level variation in flowering probability.
 
 Finally, with the exception of survival in 1-ha fragments ($p = 0.253$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.01$ for the random effect of plot).
-We found no evidence for a cost of reproduction: in both forest and fragments, flowering in the previous year was significantly positively related to growth (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flowering (CF: $p < 0.001$; 1-ha: $p = 0.004$) meaning that plants which had flowered in the previous year were more likely to be larger and flower again, on average.
-The random effect of plant ID was also significant for flowering probability (CF: $p < 0.001$; 1-ha: $p=0.003$) indicating significant individual-level variation in flowering probability.
 

### 2021-06-22 doc/paper.Rmd
data availibility statement
legend 1 & 2

@@
 
-All data and code used in this study are available at [insert zenodo link].
+Data used in this study are available at the Dryad Digital Repository; R code for analyses and visualizations are archived at Zenodo [links included upon acceptance].
 
@@
 ```{r}
-eda_cap <- "Timeseries of demographic parameters and drought occurrence.
-Mean fold-change in plant plant size (log2(size~t+1~ / size~t~)) (a) shows that plant growth varied by year and habitat (1 ha fragments in orange, solid; continuous forest in blue, dashed).
-In most years plants grew on average with a notable exception in 2003 where plants regressed in size on average in both habitats (fold-change < 0).
-Error bars in a) represent standard deviation.
-The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census.
-The proportion of reproductive sized plants (c) is on average low and fluctuates substantially year to year.
-The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years.
-Monthly 3-month SPEI is shown (d) with gray lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean.
-Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
+eda_cap <- "(a-c) Time series of H.
+acuminata vital rates in 1-ha fragments (solid orange lines) and continuous forest (dashed blue lines) and (d) drought occurrence in the study region.
+(a) mean fold-change in plant plant size (log2(size~t+1~ / size~t~)) varies by year and habitat.
+On average, plants grew in most years with the notable exception in 2003, in which on average plants regressed in size in both habitats (i.e., fold-change < 0).
+Error bars represent the standard deviation.
+(b) The proportion of plants surviving from one transition year to the next varied from 0.98 (CF in 1998-1999) to 0.91 (CF in 2003-2004).
+(c) The proportion of H.
+acuminata above the size threshold for reproduction that flowered each year is on average low but variable.
+The size threshold is determined by the upper 90th percentile size of flowering plants across all years.
+(d) Monthly 3-month SPEI for our study region.
+Gray lines represent values from different grid cells encompassing BDFFP; the dark line represents the site mean.
+Colored stripes represent drought intensity: yellow = mild, orange = moderate, dark orange = severe, red = extreme."
 ```
@@
 ```{r}
-surv_curve_cap <- "Survivorship of plants labeled in the first survey year, 1998, which comprise 49% of the plants in the full dataset.
-After 10 years, 79.7% (1629/2055) of plants in continuous forest survived and 72.4% (393/543) of plants in fragments survived."
+surv_curve_cap <- "Survivorship curve for plants marked in the 1998 survey year; these plants comprise 49% of those in the complete demographic dataset.
+The percentage of these plants that were still alive ten years later was 79.7% (1629/2055) in continuous forest vs.
+72.4% (393/543) in 1-ha fragments."
 ```

### 2021-06-22 doc/paper.Rmd
fig legends 4-6

@@
 ```{r}
-size_cap <- "Smooth effect of plant size in the previous census from models for a) survival, b) log(size), and c) flowering probability, corresponding to the additive term s~1~(z~i~) in eq.
+size_cap <- "Smooth effect from models of plant size in the previous census on (a) survival, (b) log(size), and (c) flowering probability; these values correspond to the additive term s~1~(z~i~) in eq.
 1.
-95% confidence intervals are shown and include uncertainty in the intercept and uncertainty due to smoothness selection.
-The smooths for 1-ha fragments and continuous forest are fit in separate models.
-Panel d) shows the distribution of plant sizes in the two habitat types."
+The bands depicting the 95% confidence interval include uncertainty in the intercept and uncertainty due to smoothness selection; the smooths for 1-ha fragments and continuous forest are fit in separate models.
+(d) Plant size distribution by habitat type (solid line = 1-ha fragments, dashed line = Continuous Forest)."
 ```
@@
 ```{r}
-s_spei_cap <- "Smooth effect of lagged SPEI on survival in a) continuous forest and b) 1-ha fragments.
-Panel c) shows the difference between panels a) and b).
+s_spei_cap <- "Smooth effect of lagged SPEI on H.
+acuminata survival in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October).
 Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
 
@@
 ```{r}
-g_spei_cap <- "Smooth effect of lagged SPEI on plant growth for a) continuous forest and b) 1-ha fragments.
-Panel c) shows the difference between panels a) and b).
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function.
-Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+g_spei_cap <- "Smooth effect of lagged SPEI on H.
+acuminata growth in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October)."
 ```
@@
 ```{r}
-f_spei_cap <- "Smooth effect of lagged SPEI on flowering probability for a) continuous forest and b) 1-ha fragments.
-Panel c) shows the difference between panels a) and b).Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale.
-Outlines show regions where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October)."
+f_spei_cap <- "Smooth effect of lagged SPEI on H.
+acuminata flowering probability in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October).
+Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
+Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
 ```

### 2021-06-22 doc/paper.Rmd
intro smooth out, added a new citation

@@
 
-Tropical deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015; @alroy2017].
-Deforestation also results in landscapes where the remaining forest can be highly fragmented, with forest remnants of different sizes embedded in a matrix of often highly-contrasting habitat [@taubert2018; @bianchi2013].
-For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
-Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
-Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
-Despite the fact that this mechanism is thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions remains scarce.
+The expansion of agriculture and other human activities is a primary driver of deforestation in the tropics [@haddadHabitatFragmentationIts2015; @alroy2017].
+It also results in landscapes where the remaining forest can be highly fragmented, with patches of different sizes embedded in a matrix of often contrasting habitat [@taubert2018; @bianchi2013].
+This fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
+Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in tropical forest fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] drive declines in plant reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
+Despite the prevalence of this hypothesis [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions in fragments remains scarce.
 
-Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with effects that are more pronounced on fragmented populations [@oliver2015; @holyoak2016].
-While the demographic consequences are expected to be similarly severe in the tropics [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for species found in tropical biomes [@paniw2021].
-Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
-Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
-This increased sensitivity to climate fluctuations, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
+Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], and that the effects of climate change could be more pronounced when habitat is fragmented [@oliver2015; @holyoak2016].
+While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
+Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
 
@@
 
-Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
-Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
+Over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation [@broadbentForestFragmentationEdge2008].
+Climate models predict a future of extremes for these forests---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
+Our results support the hypothesis that populations in Amazonian forest fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.

### 2021-06-22 doc/paper.Rmd
linebreaks
figure caption minor edits

@@
 These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].
-In our sites *H.
-acuminata* is pollinated by *Phaeothornis superciliosus* and *P.
-bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
+In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
 Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}=1.1$ flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
@@
 ```{r}
-eda_cap <- "(a-c) Time series of H.
-acuminata vital rates in 1-ha fragments (solid orange lines) and continuous forest (dashed blue lines) and (d) drought occurrence in the study region.
-(a) mean fold-change in plant plant size (log2(size~t+1~ / size~t~)) varies by year and habitat.
+eda_cap <- "(a-c) Time series of *H.
+acuminata* vital rates in 1-ha fragments (solid orange lines) and continuous forest (dashed blue lines) and (d) drought occurrence in the study region.
+(a) Mean fold-change in plant plant size (log2(size~t+1~ / size~t~)) varies by year and habitat.
 On average, plants grew in most years with the notable exception in 2003, in which on average plants regressed in size in both habitats (i.e., fold-change < 0).
@@
 (b) The proportion of plants surviving from one transition year to the next varied from 0.98 (CF in 1998-1999) to 0.91 (CF in 2003-2004).
-(c) The proportion of H.
-acuminata above the size threshold for reproduction that flowered each year is on average low but variable.
+(c) The proportion of *H.
+acuminata* above the size threshold for reproduction that flowered each year is on average low but variable.
 The size threshold is determined by the upper 90th percentile size of flowering plants across all years.
@@
 ```{r}
-s_spei_cap <- "Smooth effect of lagged SPEI on H.
-acuminata survival in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+s_spei_cap <- "Smooth effect of lagged SPEI on *H.
+acuminata* survival in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
 Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
@@
 ```{r}
-g_spei_cap <- "Smooth effect of lagged SPEI on H.
-acuminata growth in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+g_spei_cap <- "Smooth effect of lagged SPEI on *H.
+acuminata* growth in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
 Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
@@
 ```{r}
-f_spei_cap <- "Smooth effect of lagged SPEI on H.
-acuminata flowering probability in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+f_spei_cap <- "Smooth effect of lagged SPEI on *H.
+acuminata* flowering probability in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
 Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.

### 2021-06-22 doc/paper.Rmd
minor edits to results and discussion

@@
 *Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
-This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was apreceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
+This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
@@
 The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
-Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
+Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also had a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
 In contrast, the effect of recent SPEI in continuous forest was weaker, with only the 6 months preceding a census having a significant effect on survival (Figure \@ref(fig:surv-spei)a).
@@
 However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
-In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a m,ore pronounced response in 1-ha fragments.
+In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a more pronounced response in 1-ha fragments.
 

### 2021-06-22 doc/paper.Rmd
minor edit to methods

@@
 
-To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs; @gasparriniPenalizedFrameworkDistributed2017].
+To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs, @gasparriniPenalizedFrameworkDistributed2017].
 DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.

### 2021-06-23 doc/paper.Rmd
added citations to r packages

@@
 
-All analyses were conducted in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020].
+All analyses were conducted in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020] using the `targets` package for workflow management [@landauTargetsPackageDynamic2021].
+Figures were created with the aid of the `gratia`, `ggplot2`, and `patchwork` packages [@wickhamGgplot2ElegantGraphics2016; @simpsonGratiaGracefulGgplot2021; @pedersenPatchworkComposerPlots2020].
 

### 2021-06-23 doc/paper.Rmd
miscalculation of percentage difference

@@
   mutate(across(ends_with("_mean") | ends_with("_sd"),
-                ~round(.x, 2)))
+                ~round(.x, 3)))
 
@@
 *Flowering:* The overall proportion of plants that flowered was very low.
-While it was almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+While it was 27% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
@@
 Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
-It should be noted, however, that only the first year of census data (1999) met these conditions.
+It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.
 We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.

### 2021-06-23 doc/paper.Rmd
cut abstract down to 300 words.  Closes #66

@@
 
-Deforestation is a major threat to biodiversity in the Amazon, partly because it leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
+Deforestation often results in landscapes where remaining forest habitat is highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix.
 Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood.
-It is often hypothesized that the altered environmental conditions in fragments drive declines in reproduction, recruitment, or survivorship.
+It is often hypothesized that altered environmental conditions in fragments drive declines in reproduction, recruitment, or survivorship.
 The Amazon basin, in addition to experiencing continuing fragmentation, is warming and experiencing changes in precipitation leading to altered frequency and intensity of droughts and unusually wet periods.
@@
 Most studies of plants in fragments are relatively short (1--6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest.
-Even fewer studies consider delayed effects of climate despite the importance of such delayed effects for plant demographic vital rates in studies that look for them.
+Even fewer studies consider delayed effects of climate on demographic vital rates despite the importance of delayed effects in studies that consider them.
 Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (*Heliconia acuminata*, Heliconiaceae).
@@
 Drought two dry seasons prior to the February census decreased survival and increased flowering probability while drought in the wet season a year prior to the census decreased flowering probability and increased growth.
-The effects of extremes in precipitation on survival and growth were more intense in forest fragments compared to continuous forest.
-The complex delayed effects of climate and habitat fragmentation on *H.
-acuminata* vital rates points to the importance of long-term demography experiments in understanding the effects of anthropogenic change on plant populations.
+The effects of extremes in precipitation on survival and growth were more pronounced in forest fragments compared to continuous forest.
+The complex delayed effects of climate and habitat fragmentation in our study point to the importance of long-term demography experiments in understanding the effects of anthropogenic change on plant populations.
 

### 2021-06-24 doc/paper.Rmd
remove reproducibility report, high res figures

@@
 	fig.path = "figures/",
+	dev = "tiff",
 	message = FALSE,
@@
 
-### Colophon
+<!-- ### Colophon -->
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->
 
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
+<!-- ```{r colophon, cache = FALSE} -->
 
-The current Git commit details are:
+<!-- # which R packages and versions? -->
 
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```
+<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
+
+<!-- ``` -->
+
+<!-- The current Git commit details are: -->
+
+<!-- ```{r} -->
+
+<!-- # what commit is this file at?  -->
+
+<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
+
+<!-- ``` -->

### 2021-06-24 doc/paper.Rmd
save .tiff in addition to embedded .png

@@
 	fig.path = "figures/",
-	dev = "tiff",
+	dev = "png",
+	dpi = 300,
 	message = FALSE,
@@
 	collapse = TRUE,
-	comment = "#>",
-	dpi = 300
+	comment = "#>"
 )
@@
   tar_load(c(
-    # eda_plot,
-    # surv_curve,
-    # s_spei_plot,
-    # g_spei_plot,
-    # f_spei_plot,
-    # size_plot,
     model_data,
@@
 
-```{r eda, fig.cap=eda_cap, fig.width=7, fig.height=7}
+```{r eda, fig.cap=eda_cap, fig.height=7, fig.width=7}
+tar_read(eda_plot) & theme(text = element_text(size = 10))
+```
+
+<!-- high res version -->
+
+```{r eda_hires, fig.height=7, fig.width=7, dev='tiff', include=FALSE}
 tar_read(eda_plot) & theme(text = element_text(size = 10))
@@
 
-```{r surv-curve, fig.cap=surv_curve_cap, fig.width=7, fig.height=3.5}
+```{r surv-curve, fig.cap=surv_curve_cap, fig.height=3.5, fig.width=7}
+tar_read(surv_curve) + theme(text = element_text(size = 10))
+```
+
+<!-- high res version -->
+
+```{r surv-curve_hires, fig.height=3.5, fig.width=7, dev="tiff", include=FALSE}
 tar_read(surv_curve) + theme(text = element_text(size = 10))
@@
 ```{r size-plot, fig.cap=size_cap, fig.width=7, fig.height=7}
+tar_read(size_plot) & theme(text = element_text(size = 10))
+```
+
+<!-- high res version -->
+
+```{r size-plot_hires, fig.width=7, fig.height=7, dev='tiff', include=FALSE}
 tar_read(size_plot) & theme(text = element_text(size = 10))
@@
 
+<!-- high res version-->
+
+```{r surv-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
+tar_read(s_spei_plot) & theme(text = element_text(size = 10))
+```
+
 ```{r}
@@
 ```{r growth-spei, fig.cap=g_spei_cap, fig.height=7.5, fig.width=7}
+tar_read(g_spei_plot) & theme(text = element_text(size = 10))
+```
+
+<!-- high res version -->
+
+```{r growth-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
 tar_read(g_spei_plot) & theme(text = element_text(size = 10))
@@
 
+<!-- high res version-->
+
+```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
+tar_read(f_spei_plot) & theme(text = element_text(size = 10))
+```
+
 # References
@@
 
-\newpage
+<!-- \newpage -->
 

### 2021-06-25 doc/validate_data.Rmd
update data validation

@@
 ---
-title: "Untitled"
+title: "Demographic data validation"
 output: html_document
@@
     xa_done,
-    demog_done
+    model_data
   ))
@@
 
-`demog_done`: Tidied dataset on *Heliconia acuminata* demography.
+`model_data`: Tidied dataset on *Heliconia acuminata* demography.
 More details [here](https://github.com/embruna/HeliconiaDataPaper).
@@
 ```{r}
-demog_done
+model_data
 ```
@@
 create_agent(
-  demog_done,
+  model_data,
   actions = action_levels(warn_at = 1, stop_at = 0.2)
 ) %>% 
-  col_is_character(vars(row, code_notes, code2)) %>% 
+  col_is_character(vars(code_notes, code2)) %>% 
   col_is_factor(vars(ranch, bdffp_reserve_no, plot, habitat, ha_id_number)) %>% 
@@
   col_vals_in_set(vars(surv, flwr), c(0,1, NA)) %>% 
-  col_vals_in_set(vars(column), 1:10) %>% 
-  col_vals_in_set(vars(row), LETTERS[1:10]) %>% 
   col_vals_in_set(vars(habitat), c("1-ha", "CF", "10-ha")) %>% 

### 2021-06-28 doc/supplemental.Rmd
Can't get reference_docx to work with flextable package

@@
 ---
-title: "Supplement to 'Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments'"
+title: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
 author: 
@@
     bookdown::word_document2:
-      linenumbers: no
+      linenumbers: yes
       number_sections: no
       fig_caption: yes
-      # reference_docx: "templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-      - --lua-filter=templates/pagebreak.lua
+      # reference_docx: "template.docx" # Insert path for the DOCX file
 linkcolor: blue
@@
                       warning = FALSE)
+```
 
-withr::with_dir(here::here(), {
-  targets::tar_load(c(
-    normals_plot,
-    embrapa_wide,
-    xa_spei,
-    trmm_spei,
-    gspei,
-    la_data
-    ))
-})
-
+```{r packages, include=FALSE}
 library(tidyverse)
@@
 
-# BDFFP plot map
+```{r load-targets, include=FALSE}
+withr::with_dir(here::here(), {
+  targets::tar_load(c(
+    normals_plot,
+    embrapa_wide,
+    xa_spei,
+    trmm_spei,
+    gspei,
+    la_data
+    ))
+})
 
-Map here
-
-Figure S1.
-Map caption here
-
-\pagebreak
+```
 
@@
 
-Figure S2.
+Figure S1.
 Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981--2010.
@@
 
-Figure S3.
+Figure S2.
 Three-month SPEI at BDFFP calculated using 4 datasets.
@@
 
-include contents of `choose_size_var.Rmd` here.
-Ask Emilio about citaiton for dataset.
+<!-- include contents of `choose_size_var.Rmd` here. -->
+
+<!-- Ask Emilio about citaiton for dataset. -->
 

### 2021-06-28 doc/paper.Rmd
add conflict of interest statement
higher resolution high-res figures

@@
 This is publication No. \_\_\_ in the BDFFP Technical Series.
+The authors delcare no conflicts of interest.
 
@@
 
-```{r eda_hires, fig.height=7, fig.width=7, dev='tiff', include=FALSE}
+```{r eda_hires, fig.height=7, fig.width=7, dev='tiff', include=FALSE, dpi=600}
 tar_read(eda_plot) & theme(text = element_text(size = 10))
@@
 
-```{r surv-curve_hires, fig.height=3.5, fig.width=7, dev="tiff", include=FALSE}
+```{r surv-curve_hires, fig.height=3.5, fig.width=7, dev="tiff", dpi=600, include=FALSE}
 tar_read(surv_curve) + theme(text = element_text(size = 10))
@@
 
-```{r size-plot_hires, fig.width=7, fig.height=7, dev='tiff', include=FALSE}
+```{r size-plot_hires, fig.width=7, fig.height=7, dev='tiff', dpi=600, include=FALSE}
 tar_read(size_plot) & theme(text = element_text(size = 10))
@@
 
-```{r surv-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
+```{r surv-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
 tar_read(s_spei_plot) & theme(text = element_text(size = 10))
@@
 
-```{r growth-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
+```{r growth-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
 tar_read(g_spei_plot) & theme(text = element_text(size = 10))
@@
 
-```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', include=FALSE}
+```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
 tar_read(f_spei_plot) & theme(text = element_text(size = 10))

### 2021-06-28 doc/paper.Rmd
pdf versions of figures

@@
 
+```{r eda_vect, fig.height=7, fig.width=7, dev='pdf', include=FALSE}
+tar_read(eda_plot) & theme(text = element_text(size = 10))
+```
+
 ```{r}
@@
 ```{r surv-curve_hires, fig.height=3.5, fig.width=7, dev="tiff", dpi=600, include=FALSE}
+tar_read(surv_curve) + theme(text = element_text(size = 10))
+```
+
+```{r surv-curve_vect, fig.height=3.5, fig.width=7, dev="pdf", include=FALSE}
 tar_read(surv_curve) + theme(text = element_text(size = 10))
@@
 ```{r size-plot_hires, fig.width=7, fig.height=7, dev='tiff', dpi=600, include=FALSE}
+tar_read(size_plot) & theme(text = element_text(size = 10))
+```
+
+```{r size-plot_vect, fig.width=7, fig.height=7, dev='pdf', include=FALSE}
 tar_read(size_plot) & theme(text = element_text(size = 10))
@@
 
+```{r surv-spei_vect, fig.height=7.5, fig.width=7, dev='pdf', include=FALSE}
+tar_read(s_spei_plot) & theme(text = element_text(size = 10))
+```
+
 ```{r}
@@
 ```{r growth-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
+tar_read(g_spei_plot) & theme(text = element_text(size = 10))
+```
+
+```{r growth-spei_vect, fig.height=7.5, fig.width=7, dev='pdf', include=FALSE}
 tar_read(g_spei_plot) & theme(text = element_text(size = 10))
@@
 ```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
+tar_read(f_spei_plot) & theme(text = element_text(size = 10))
+```
+
+```{r flwr-spei_vect, fig.height=7.5, fig.width=7, dev='pdf', include=FALSE}
 tar_read(f_spei_plot) & theme(text = element_text(size = 10))

### 2021-06-28 doc/paper.Rmd
minor corrections spotted in proofs

@@
 
-Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
+Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
 In this study, we used the product of shoot number and plant height as our measure of plant size.
@@
   & g[ \textrm{E}(y_{i})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
-  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot }j 
+  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot j} 
   \end{split}
@@
 Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
-1 A random effect of plant ID was not included in survival models since each plant only dies once.
+A random effect of plant ID was not included in survival models since each plant only dies once.
 
@@
 This led to some important inter-habitat differences in plant responses to prior droughts.
-In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20 ) and in the wet season 34--36 months prior all increased the probability of flowering.
+In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
 However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
@@
 However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated.
-Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability.
+Multi-factorial, multi-season experiments [*sensu* @aguirre2021; @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability.
 Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms.

### 2021-06-29 doc/supplemental.Rmd
added reference to data

@@
                       message = FALSE,
-                      warning = FALSE)
+                      warning = FALSE,
+                      dpi = 300)
 ```
@@
 
-```{r normals}
+```{r normals, fig.height=4, fig.width=5.5}
 normals_plot & theme(text = element_text(size = 10))
@@
 
-<!-- include contents of `choose_size_var.Rmd` here. -->
-
-<!-- Ask Emilio about citaiton for dataset. -->
-
 ```{r}
@@
 Table S2.
-Using data from transplant experiments of *Heliconia acuminata* [@brunaExperimentalAssessmentHeliconia2002 <!--# and doi for unpublished data -->] we fit linear regressions with log(leaf area) as the response variable and either log(height), log(shoot number) or log(shoot number x height) as the predictor.
+Using data from transplant experiments of *Heliconia acuminata* [@brunaExperimentalAssessmentHeliconia2002; @brunaLeafNumberLeaf2021] we fit linear regressions with log(leaf area) as the response variable and either log(height), log(shoot number) or log(shoot number x height) as the predictor.
 Using the log of size (shoot number $\times$ height) as a predictor gives the lowest AIC, BIC, root mean squared error (RMSE) and highest $R^2$ indicating that it is the best proxy for predicting log(leaf area) as a response.

### 2021-06-30 doc/validate_data.Rmd
revise data validation
removed `code2` column as its no longer in raw data

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
 
@@
 -   `code_notes` (character): possible values include: "plant missing" meaning the plant could not be found in this survey; "dead" meaning the plant is confirmed dead in this survey; "ULY" meaning\_\_\_\_\_\_\_\_; and "tag missing" meaning the plant was missing a tag and given a new one (this does not affect the `ha_id_number` as these take into account plants given new tags)
--   `code2` (character): possible value is "initial.tag.yr" indicating the beginning of data for that `ha_id_number`.
-Could be the first year of the survey, or could represent recruitment into the plot.
 
@@
 ) %>% 
-  col_is_character(vars(code_notes, code2)) %>% 
+  col_is_character(vars(code_notes)) %>% 
   col_is_factor(vars(ranch, bdffp_reserve_no, plot, habitat, ha_id_number)) %>% 

### 2021-06-30 R/wrangle_demography.R

@@ -80,5 +80,5 @@ wrangle_demog <- function(data, n_years = 3) {
            ht, ht_prev, shts, shts_prev,
            size, size_prev, log_size, log_size_prev,
            infl_num = infl, flwr, flwr_prev, surv,
-           code_notes, code2)
+           code_notes)
 }

### 2021-08-09 doc/validate_data.Rmd

@@
   col_is_numeric(vars(ht, ht_prev, size, size_prev, log_size, log_size_prev)) %>%
-  col_is_integer(vars(year, shts, shts_prev, infl_num, surv, flwr)) %>% 
+  col_is_integer(vars(year, shts, shts_prev, infl, infl_prev, surv, flwr, flwr_prev)) %>% 
   col_vals_in_set(vars(surv, flwr), c(0,1, NA)) %>% 

### 2021-08-09 doc/paper.Rmd
edit abstract (closes #84)

@@
 Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (*Heliconia acuminata*, Heliconiaceae).
-We used distributed lag non-linear models to understand the delayed effects of temperature and precipitation on survival, growth, and flowering.
+We used distributed lag non-linear models to understand the delayed effects of standardized precipitation evapotranspiration index (SPEI) on survival, growth, and flowering.
 We detected delayed effects of climate up to 36 months.
@@
 
-We thank Collin Edwards, \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Collin Edwards, Andrew Mercadante, Ellie McDaniel, \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
 We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.

### 2021-08-09 review/Response to reviews.Rmd
Clarify death condition.  Closes #91

---
title: "Response to Reviews"
author: "Eric Scott"
date: "8/9/2021"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
---

# Reviewer 1

## General comments

>Scott and colleagues present the results of an interesting series of analyses designed to disentangle the lagged effects of climate and forest fragmentation on the demographic performance of a tropical understory plant. As ecologists seek to understand not just how changes in mean climate, but also changes in extremes, will influence population persistence, studies such as this that capitalize on long time series are critical to making progress. The authors find that the effects of swings in precipitation (both drought and particularly wet years) affect survival, growth and flowering, but that these effects are both lagged and differ by whether plants were in continuous forest or the smallest fragment size (1 ha). Overall, it’s a well written paper, and an application of a relatively new modeling approach to address the questions. The results highlight both the challenges with understanding the complex factors driving population persistence and also how critical it will be to grapple with them. Below I provide suggestions for improvement, particularly aimed to make the new modeling approach more accessible.

## Specific comments

>1. By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words ‘extreme precipitation’ – this of course makes sense later that the results are focused both on the effects of drought and extreme wet. The introduction could set up the reader better to understand both. Since drought is well covered, adding background and hypotheses for why extreme wet would have a negative effect on vital rates would help.

>2. In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest. I wondered if the authors thought this was too obvious (edge effects, etc), but I think it’d be helpful to spell this out more clearly. L466 – 467 offer a new hypothesis, but briefly revisiting the ‘old’ evidence/hypotheses also seems important.

>3. One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone. As written in the methods and results, I found the implementation very difficult to follow. A slightly longer summary of the methods would help. In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit. While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit. One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn. Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).

>4. Related to above, I think it’s important to see more of the raw data than are presented in Fig. 1. (I’m thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs). This could be an appendix, but it would help with interpreting Figures 3 – 6.

>5.  I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust. In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census. While the authors mention this, what about adding something short about when using this type of analysis is critical and when it might not be (should all demographers adopt this approach? Why or why not?)

>6. I was curious about why the paper did not include analyses of recruitment. Perhaps it is beyond the scope of this paper, but I wondered how new recruits responded to drought and wet also, both in number, as well as how big they were in the first year.

>7. I had a hard time linking these statements in the results to the figures.
- L272-273 – How do you see disparities in plant size here? I thought this was about growth?
- L277-278 - How does this show the effect of size on growth is greatest for mid-size plants? More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion. More precise language would help.
- L331 – 332 – How can you see in this figure that the response is ‘more muted as indicated by the shape of the crossbasis function’?

>8. L442-443 – What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.

## Figures:

>Fig 2 – This was only cited in the discussion as far as I can tell. Bring it back into the results?

>Fig 3A – The results mention something about survival approaching 1 – I wasn’t sure if that was supposed to be visible here.

>Fig 3 – I mentioned this earlier, but I wasn’t really clear on how to interpret effect size. I see the explanation in L 539 in the legend, but I don’t really know what that means. Since this is such an important result of the paper, it would help the reader if they were walked through this idea. What is a negative vs. positive effect size in relation to plant size (does this compare to the median size)? I am likely missing something here. (I think the challenge for me is that I kept wanting the figures to fit into a more standard demographic paper – like the probability of survival based on size, but then I realized something different was being shown here.)

>Figs 4 – 6 – Where the isocline circles are large, does that mean it’s a very large time span? I was a bit confused when describing the results of Fig 4b (L296-297), for example, the lag is described as the preceding 13 months in the 1 ha fragment, rather than 0 – 13 months (although now I see maybe the word preceding was meant to indicate this whole range). A little bit more explanation in the legends would help these figures be more intuitive.

# Reviewer 2
## General comments
>The manuscript by Scott et al analyzed an eleven year long demographic data set on the herb, Heliconia acuminata in the Central Amazon to evaluate the effect of drought and fragmentation on survival growth and fecundity. The authors also employ distributed lag non-linear models to explore delayed effects of drought. This work is an important contribution adding to the sparce literature documenting the role of weather variables on plant vital rates. The manuscript is generally well written. I hope the authors will find my comments helpful in improving this already very good manuscript.

## Specific Comments:

>1) The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.

>2) Line 51-52. I find this sentence confusing. What specific demographic mechanisms are you referring to? Do you mean that high mortality rate lead to extinction and we don’t know what causes high mortality rates?

>3) Line 146. Can you explain what you mean with “the height of each plant to the tallest leaf”? If the tallest leaf is growing more horizontally the height of the plant would be shorter than the length of the largest leaf?

>4) Lines 251ff: I don’t see a reference of Fig 2 in the result section.

>5) Line 277-278: I don’t see in Figure 3a that the impact of size on growth was greatest for mid-sized
plants in continuous forest. It looks to me that the 1-ha and CF lines are extremely close together and not statistically different. Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.

>6) Lines 287-289. While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments). I feel that the reporting of the results is a bit biased towards concluding that plants do worse in fragments compared to CF, especially since none of the result reported so far are significant.

>7) Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don’t know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant. Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution. I would like to see at least a justification for why representing the difference figures are meaningful.

>8) Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also. I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a. What springs to my mind is that the area that is significant is much larger in in fragments compared to CF. It seems that in CF only a few months are significant. Surprisingly they fall not consistently in the dry or wet season.

>9) Line 323-325: I don’t know why the growth of plants in fragments would affect the size of plant in CF.

>10) Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).

>11) Line 370: There are no confidence intervals in Fig 2, so we don’t know if the difference is significant.

>12) Lines 383 ff: see comment 7

>13) Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.

>14) 424-427: I think this is exaggerating since the differences between habitat types is not significant. 15) You may also consider arguing that an 11-year long time series is not sufficient to detect delayed effects (see Tenhumberg et al. 2018)

>16) Lines 437-438 change “starting expand” with “starting to expand”

>17) Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?

>18) Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.

>19) Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.

>20) Lines 468 ff: I suggest deleting this paragraph. It doesn’t really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next. Distributed Lag Non-Linear Models account for the exact timing of the weather events.

>21) The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H. acuminata. For instance, why does too much rain reduce survival? Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past? Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)? Etc.

>22) Lines 484-486: I don’t think this study strongly support that plants do worse in fragments than CF. 23) Line 499: This is confusing. If you measure size every year than you calculate growth as the change in size from one year to the next, don’t you?

# Reviewer 3

## General comments

>This is a really nice addition to the growing literature on lagged climate effects on plant demography, and also expands the tool kit for these studies. The combination of experimentally generated fragmentation with long term demographic data is also unique and very valuable. The paper is very clear and concise, with a few minor exceptions discussed in detail below. The one surprising omission was that the authors didn't use an Integral Projection Model to put the vital rates together into a metric of population growth potential. I enjoyed reading this.

## Specific comments

>Line 152 -- "Plants not found for 3 consecutive seasons ..." when were they considered to have died? first year they were missed or last? How many plants were not found in an average survey, and why? Does Heliconia show whole plant dormancy?

```{r eval=FALSE, include=FALSE}
n_miss <- 
  model_data %>%
  group_by(year) %>%
  summarize(n_obs = n(), n_miss = sum(is.na(shts)), p_miss = n_miss/n_obs) %>% 
  slice(-1) %>%  #remove first year when all plants are found
  pull(p_miss)
1 - range(n_miss)

yrs_missing <- 
  model_data %>% 
  group_by(ha_id_number) %>% 
  filter(!is.na(last(shts))) %>% 
  summarize(n_obs = n(), n_miss = sum(is.na(shts))) %>% 
  filter(n_miss > 0) %>% 
  pull(n_miss)

quantile(yrs_missing, c(0.25, 0.5, 0.75, 0.95, 0.99))
```

We have edited to clarify: *"After the initial census year, between 80% and 97% of marked plants were found. Of plants that had missing values for some years, but were found again in a subsequent year, 95% had 2 or fewer years of missing values. Therefore, plants that were not found for three consecutive surveys, and no subsequent survey, were considered to have died in the transition year after their last observation."*

Some of the missing values in the data do represent *Heliconia* plants that lost all above-ground tissue but re-sprouted in successive years, but other missing values may simply be imperfect "recapture".

>Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames? You describe why not using longer smooths, but not shorter.

>Line 191 - 194 (equation and description) for clarity you should describe what the index variables i and j are indexing, and I think clearer to have y_ij for ith plant in jth plot. Indexes on the random effect description suggest there is a separate variance estimated for each plot? d_it suggests that each plant has a separate SPEI estimate? I think it is probably aggregated at a higher level than individual? I would reorganize the following paragraph so that all the descriptive text for the components of (1) occurs before the digression on the details of the crossbasis function.

>Line 205 - I'm not sure "bimodal" is correct here -- a cubic function can have 2 local extrema, is that what you mean?

>Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.

>Line 211 - random effect -- how did you deal with this in the predictions? set to zero in design matrix, Choose a particular value, sample from the posterior? The code would tell me but not available.

>Line 253 - fig 1a shows mean change in size, not mean size?

>Figure 2 isn't referenced in this section, not used until discussion. Maybe move to supplement or add a sentence in results?

>Line 260 -- you mention the CI of the model estimated coefficients, but these aren't provided anywhere that I could find. I think a table based on the summary of the two fitted models discussed here (and elsewhere too) should be added. And report the variance of the random effect.

>Figure 4 and 5 -- very nice way to display the confidence limit. I did have to spend a while looking at them and doing a bit of head scratching to figure out which side the significant values were on, especially for the ones that are along the middle where the difference in color is not large. I wonder if a diverging color scheme with values close to the mean set to white would make this easier to see? Or if you could add some hatching or change the alpha transparency on one side to highlight those regions more clearly.

>Line 413 -- "our expansion of this approach ..." I'm not sure that the expansion is the penalty that prevents over fitting -- both Teller et al and Tenhumberg et al used the penalized spline function in mgcv. What IS an expansion is the non-linear cross basis function. Both Teller et al and Tenhumberg et al used (essentially) linear cross basis functions, so using dlnm is super-cool and interesting.

>Line 471 -- this demographic delay left me a bit confused for a while until I stopped looking at figure 4b and started looking at 4a. Then it made a lot of sense. However, wouldn't it be that the demographic effect takes place AFTER the census for an echo to appear at previous lags?

### 2021-08-09 doc/paper.Rmd
Clarify death condition.  Closes #91

@@
 Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
-Plants that were not found for three consecutive surveys were considered dead.
+After the initial census year, between 80% and 97% of marked plants were found.
+Of plants that had missing values for some years, but were found again in a subsequent year, 95% had 2 or fewer years of missing values.
+Therefore, plants that were not found for three consecutive surveys, and no subsequent survey, were considered to have died in the transition year after their last observation.
 We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
@@
 For example, the flowering shoots of *Heliconia chartacea* begin to develop 6--10 months prior to the appearance of inflorescences [@crileyYearProductionHigh1994].
-Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots.
+Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting to expand, leads to the aborted production of flowering shoots.
 Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.

### 2021-08-09 review/Response to reviews.Rmd
response to reviewer for #84
visual editor formatting

@@
     reference_docx: template.docx
+bibliography: references.bib
 ---
@@
 
->Scott and colleagues present the results of an interesting series of analyses designed to disentangle the lagged effects of climate and forest fragmentation on the demographic performance of a tropical understory plant.
-As ecologists seek to understand not just how changes in mean climate, but also changes in extremes, will influence population persistence, studies such as this that capitalize on long time series are critical to making progress.
-The authors find that the effects of swings in precipitation (both drought and particularly wet years) affect survival, growth and flowering, but that these effects are both lagged and differ by whether plants were in continuous forest or the smallest fragment size (1 ha).
-Overall, it’s a well written paper, and an application of a relatively new modeling approach to address the questions.
-The results highlight both the challenges with understanding the complex factors driving population persistence and also how critical it will be to grapple with them.
-Below I provide suggestions for improvement, particularly aimed to make the new modeling approach more accessible.
+> Scott and colleagues present the results of an interesting series of analyses designed to disentangle the lagged effects of climate and forest fragmentation on the demographic performance of a tropical understory plant.
+> As ecologists seek to understand not just how changes in mean climate, but also changes in extremes, will influence population persistence, studies such as this that capitalize on long time series are critical to making progress.
+> The authors find that the effects of swings in precipitation (both drought and particularly wet years) affect survival, growth and flowering, but that these effects are both lagged and differ by whether plants were in continuous forest or the smallest fragment size (1 ha).
+> Overall, it's a well written paper, and an application of a relatively new modeling approach to address the questions.
+> The results highlight both the challenges with understanding the complex factors driving population persistence and also how critical it will be to grapple with them.
+> Below I provide suggestions for improvement, particularly aimed to make the new modeling approach more accessible.
 
@@
 
->1.
-By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words ‘extreme precipitation’ – this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
+> 1.
+By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
 The introduction could set up the reader better to understand both.
@@
 
->2.
+> 2.
 In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest.
-I wondered if the authors thought this was too obvious (edge effects, etc), but I think it’d be helpful to spell this out more clearly.
-L466 – 467 offer a new hypothesis, but briefly revisiting the ‘old’ evidence/hypotheses also seems important.
+I wondered if the authors thought this was too obvious (edge effects, etc), but I think it'd be helpful to spell this out more clearly.
+L466 -- 467 offer a new hypothesis, but briefly revisiting the 'old' evidence/hypotheses also seems important.
 
->3.
+> 3.
 One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
@@
 
->4.
-Related to above, I think it’s important to see more of the raw data than are presented in Fig.
+> 4.
+Related to above, I think it's important to see more of the raw data than are presented in Fig.
 1.
-(I’m thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs).
-This could be an appendix, but it would help with interpreting Figures 3 – 6.
+(I'm thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs).
+This could be an appendix, but it would help with interpreting Figures 3 -- 6.
 
->5.
+> 5.
 I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.
@@
 
->6.
+> 6.
 I was curious about why the paper did not include analyses of recruitment.
@@
 
->7. I had a hard time linking these statements in the results to the figures.
-- L272-273 – How do you see disparities in plant size here? I thought this was about growth?
-- L277-278 - How does this show the effect of size on growth is greatest for mid-size plants?
+> 7.  I had a hard time linking these statements in the results to the figures.
+>
+> -   L272-273 -- How do you see disparities in plant size here? I thought this was about growth?
+> -   L277-278 - How does this show the effect of size on growth is greatest for mid-size plants?
 More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
 More precise language would help.
-- L331 – 332 – How can you see in this figure that the response is ‘more muted as indicated by the shape of the crossbasis function’?
+> -   L331 -- 332 -- How can you see in this figure that the response is 'more muted as indicated by the shape of the crossbasis function'?
 
->8. L442-443 – What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
+> 8.  L442-443 -- What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
 
@@
 
->Fig 2 – This was only cited in the discussion as far as I can tell. Bring it back into the results?
+> Fig 2 -- This was only cited in the discussion as far as I can tell.
+> Bring it back into the results?
 
->Fig 3A – The results mention something about survival approaching 1 – I wasn’t sure if that was supposed to be visible here.
+> Fig 3A -- The results mention something about survival approaching 1 -- I wasn't sure if that was supposed to be visible here.
 
->Fig 3 – I mentioned this earlier, but I wasn’t really clear on how to interpret effect size.
-I see the explanation in L 539 in the legend, but I don’t really know what that means.
-Since this is such an important result of the paper, it would help the reader if they were walked through this idea.
-What is a negative vs.
-positive effect size in relation to plant size (does this compare to the median size)?
-I am likely missing something here.
-(I think the challenge for me is that I kept wanting the figures to fit into a more standard demographic paper – like the probability of survival based on size, but then I realized something different was being shown here.)
+> Fig 3 -- I mentioned this earlier, but I wasn't really clear on how to interpret effect size.
+> I see the explanation in L 539 in the legend, but I don't really know what that means.
+> Since this is such an important result of the paper, it would help the reader if they were walked through this idea.
+> What is a negative vs. positive effect size in relation to plant size (does this compare to the median size)?
+> I am likely missing something here.
+> (I think the challenge for me is that I kept wanting the figures to fit into a more standard demographic paper -- like the probability of survival based on size, but then I realized something different was being shown here.)
 
->Figs 4 – 6 – Where the isocline circles are large, does that mean it’s a very large time span?
-I was a bit confused when describing the results of Fig 4b (L296-297), for example, the lag is described as the preceding 13 months in the 1 ha fragment, rather than 0 – 13 months (although now I see maybe the word preceding was meant to indicate this whole range).
-A little bit more explanation in the legends would help these figures be more intuitive.
+> Figs 4 -- 6 -- Where the isocline circles are large, does that mean it's a very large time span?
+> I was a bit confused when describing the results of Fig 4b (L296-297), for example, the lag is described as the preceding 13 months in the 1 ha fragment, rather than 0 -- 13 months (although now I see maybe the word preceding was meant to indicate this whole range).
+> A little bit more explanation in the legends would help these figures be more intuitive.
 
 # Reviewer 2
+
 ## General comments
->The manuscript by Scott et al analyzed an eleven year long demographic data set on the herb, Heliconia acuminata in the Central Amazon to evaluate the effect of drought and fragmentation on survival growth and fecundity.
-The authors also employ distributed lag non-linear models to explore delayed effects of drought.
-This work is an important contribution adding to the sparce literature documenting the role of weather variables on plant vital rates.
-The manuscript is generally well written.
-I hope the authors will find my comments helpful in improving this already very good manuscript.
+
+> The manuscript by Scott et al analyzed an eleven year long demographic data set on the herb, Heliconia acuminata in the Central Amazon to evaluate the effect of drought and fragmentation on survival growth and fecundity.
+> The authors also employ distributed lag non-linear models to explore delayed effects of drought.
+> This work is an important contribution adding to the sparce literature documenting the role of weather variables on plant vital rates.
+> The manuscript is generally well written.
+> I hope the authors will find my comments helpful in improving this already very good manuscript.
 
@@
 
->1) The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.
+> 1)  The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.
 
->2) Line 51-52.
+We have edited the abstract.
+Now the sentence to which the reviewer refers reads: *"We used distributed lag non-linear models to understand the delayed effects of standardized precipitation evapotranspiration index (SPEI) on survival, growth, and flowering."*
+
+> 2)  Line 51-52.
 I find this sentence confusing.
 What specific demographic mechanisms are you referring to?
-Do you mean that high mortality rate lead to extinction and we don’t know what causes high mortality rates?
+Do you mean that high mortality rate lead to extinction and we don't know what causes high mortality rates?
 
->3) Line 146.
-Can you explain what you mean with “the height of each plant to the tallest leaf”?
+@brunaHabitatFragmentationPlant2009 discusses the lack of knowledge of what demographic vital rates (e.g. survival, reproduction, recruitment) contribute most to population declines in fragments.
+We have edited this line to clarify: " <!--# edit abstract? -->
+
+> 3)  Line 146.
+Can you explain what you mean with "the height of each plant to the tallest leaf"?
 If the tallest leaf is growing more horizontally the height of the plant would be shorter than the length of the largest leaf?
 
->4) Lines 251ff: I don’t see a reference of Fig 2 in the result section.
+> 4)  Lines 251ff: I don't see a reference of Fig 2 in the result section.
 
->5) Line 277-278: I don’t see in Figure 3a that the impact of size on growth was greatest for mid-sized
-plants in continuous forest.
+> 5)  Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
 It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
@@
 
->6) Lines 287-289.
+> 6)  Lines 287-289.
 While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
@@
 
->7) Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don’t know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
+> 7)  Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
 Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
@@
 
->8) Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
+> 8)  Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
 I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a.
@@
 
->9) Line 323-325: I don’t know why the growth of plants in fragments would affect the size of plant in CF.
+> 9)  Line 323-325: I don't know why the growth of plants in fragments would affect the size of plant in CF.
 
->10) Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).
+> 10) Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).
 
->11) Line 370: There are no confidence intervals in Fig 2, so we don’t know if the difference is significant.
+> 11) Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.
 
->12) Lines 383 ff: see comment 7
+> 12) Lines 383 ff: see comment 7
 
->13) Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.
+> 13) Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.
 
->14) 424-427: I think this is exaggerating since the differences between habitat types is not significant.
+> 14) 424-427: I think this is exaggerating since the differences between habitat types is not significant.
 15) You may also consider arguing that an 11-year long time series is not sufficient to detect delayed effects (see Tenhumberg et al.
@@
 
->16) Lines 437-438 change “starting expand” with “starting to expand”
+> 16) Lines 437-438 change "starting expand" with "starting to expand"
 
->17) Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?
+> 17) Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?
 
->18) Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.
+> 18) Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.
 
->19) Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
+> 19) Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
 
->20) Lines 468 ff: I suggest deleting this paragraph.
-It doesn’t really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
+> 20) Lines 468 ff: I suggest deleting this paragraph.
+It doesn't really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
 Distributed Lag Non-Linear Models account for the exact timing of the weather events.
 
->21) The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
+> 21) The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
 acuminata.
@@
 
->22) Lines 484-486: I don’t think this study strongly support that plants do worse in fragments than CF.
+> 22) Lines 484-486: I don't think this study strongly support that plants do worse in fragments than CF.
 23) Line 499: This is confusing.
-If you measure size every year than you calculate growth as the change in size from one year to the next, don’t you?
+If you measure size every year than you calculate growth as the change in size from one year to the next, don't you?
 
@@
 
->This is a really nice addition to the growing literature on lagged climate effects on plant demography, and also expands the tool kit for these studies.
-The combination of experimentally generated fragmentation with long term demographic data is also unique and very valuable.
-The paper is very clear and concise, with a few minor exceptions discussed in detail below.
-The one surprising omission was that the authors didn't use an Integral Projection Model to put the vital rates together into a metric of population growth potential.
-I enjoyed reading this.
+> This is a really nice addition to the growing literature on lagged climate effects on plant demography, and also expands the tool kit for these studies.
+> The combination of experimentally generated fragmentation with long term demographic data is also unique and very valuable.
+> The paper is very clear and concise, with a few minor exceptions discussed in detail below.
+> The one surprising omission was that the authors didn't use an Integral Projection Model to put the vital rates together into a metric of population growth potential.
+> I enjoyed reading this.
 
@@
 
->Line 152 -- "Plants not found for 3 consecutive seasons ..." when were they considered to have died?
-first year they were missed or last?
-How many plants were not found in an average survey, and why?
-Does Heliconia show whole plant dormancy?
+> Line 152 -- "Plants not found for 3 consecutive seasons ..." when were they considered to have died?
+> first year they were missed or last?
+> How many plants were not found in an average survey, and why?
+> Does Heliconia show whole plant dormancy?
 
@@
 
->Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames? You describe why not using longer smooths, but not shorter.
+> Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames?
+> You describe why not using longer smooths, but not shorter.
 
->Line 191 - 194 (equation and description) for clarity you should describe what the index variables i and j are indexing, and I think clearer to have y_ij for ith plant in jth plot.
-Indexes on the random effect description suggest there is a separate variance estimated for each plot?
-d_it suggests that each plant has a separate SPEI estimate?
-I think it is probably aggregated at a higher level than individual?
-I would reorganize the following paragraph so that all the descriptive text for the components of (1) occurs before the digression on the details of the crossbasis function.
+> Line 191 - 194 (equation and description) for clarity you should describe what the index variables i and j are indexing, and I think clearer to have y_ij for ith plant in jth plot.
+> Indexes on the random effect description suggest there is a separate variance estimated for each plot?
+> d_it suggests that each plant has a separate SPEI estimate?
+> I think it is probably aggregated at a higher level than individual?
+> I would reorganize the following paragraph so that all the descriptive text for the components of (1) occurs before the digression on the details of the crossbasis function.
 
->Line 205 - I'm not sure "bimodal" is correct here -- a cubic function can have 2 local extrema, is that what you mean?
+> Line 205 - I'm not sure "bimodal" is correct here -- a cubic function can have 2 local extrema, is that what you mean?
 
->Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.
+> Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.
 
->Line 211 - random effect -- how did you deal with this in the predictions?
-set to zero in design matrix, Choose a particular value, sample from the posterior?
-The code would tell me but not available.
+> Line 211 - random effect -- how did you deal with this in the predictions?
+> set to zero in design matrix, Choose a particular value, sample from the posterior?
+> The code would tell me but not available.
 
->Line 253 - fig 1a shows mean change in size, not mean size?
+> Line 253 - fig 1a shows mean change in size, not mean size?
 
->Figure 2 isn't referenced in this section, not used until discussion. Maybe move to supplement or add a sentence in results?
+> Figure 2 isn't referenced in this section, not used until discussion.
+> Maybe move to supplement or add a sentence in results?
 
->Line 260 -- you mention the CI of the model estimated coefficients, but these aren't provided anywhere that I could find.
-I think a table based on the summary of the two fitted models discussed here (and elsewhere too) should be added.
-And report the variance of the random effect.
+> Line 260 -- you mention the CI of the model estimated coefficients, but these aren't provided anywhere that I could find.
+> I think a table based on the summary of the two fitted models discussed here (and elsewhere too) should be added.
+> And report the variance of the random effect.
 
->Figure 4 and 5 -- very nice way to display the confidence limit.
-I did have to spend a while looking at them and doing a bit of head scratching to figure out which side the significant values were on, especially for the ones that are along the middle where the difference in color is not large.
-I wonder if a diverging color scheme with values close to the mean set to white would make this easier to see?
-Or if you could add some hatching or change the alpha transparency on one side to highlight those regions more clearly.
+> Figure 4 and 5 -- very nice way to display the confidence limit.
+> I did have to spend a while looking at them and doing a bit of head scratching to figure out which side the significant values were on, especially for the ones that are along the middle where the difference in color is not large.
+> I wonder if a diverging color scheme with values close to the mean set to white would make this easier to see?
+> Or if you could add some hatching or change the alpha transparency on one side to highlight those regions more clearly.
 
->Line 413 -- "our expansion of this approach ..." I'm not sure that the expansion is the penalty that prevents over fitting -- both Teller et al and Tenhumberg et al used the penalized spline function in mgcv.
-What IS an expansion is the non-linear cross basis function.
-Both Teller et al and Tenhumberg et al used (essentially) linear cross basis functions, so using dlnm is super-cool and interesting.
+> Line 413 -- "our expansion of this approach ..." I'm not sure that the expansion is the penalty that prevents over fitting -- both Teller et al and Tenhumberg et al used the penalized spline function in mgcv.
+> What IS an expansion is the non-linear cross basis function.
+> Both Teller et al and Tenhumberg et al used (essentially) linear cross basis functions, so using dlnm is super-cool and interesting.
 
->Line 471 -- this demographic delay left me a bit confused for a while until I stopped looking at figure 4b and started looking at 4a.
-Then it made a lot of sense.
-However, wouldn't it be that the demographic effect takes place AFTER the census for an echo to appear at previous lags?
+> Line 471 -- this demographic delay left me a bit confused for a while until I stopped looking at figure 4b and started looking at 4a.
+> Then it made a lot of sense.
+> However, wouldn't it be that the demographic effect takes place AFTER the census for an echo to appear at previous lags?

### 2021-08-10 doc/validate_models.Rmd
add checks for adequate knots (#92)

@@
 library(gratia)
+library(mgcv)
+library(dlnm)
 ```
+
+# TODO
+
+- increase knots for some models (https://github.com/BrunaLab/HeliconiaDemography/issues/92)
+- check diagnostics for individual-level random effects.  Doesn't look good, but not sure how to interpret or what to do.
 
@@
 Quantile residuals are more appropriate for binomial models.
+
 ### Fragments
+
 ```{r}
@@
 ```{r}
-
 summary(s_1ha)
@@
 ```{r echo=TRUE}
+s_1ha_res <- qresid(s_1ha)
 par(mfrow = c(2,2))
 #histogram
-plot(density(qresid(s_1ha)))
+plot(density(s_1ha_res))
 #QQ plot
-qqnorm(qresid(s_1ha)); qqline(qresid(s_1ha))
+qqnorm(s_1ha_res); qqline(s_1ha_res)
 #Fitted vs. residuals
-scatter.smooth(predict(s_1ha, type = "response"), qresid(s_1ha), col = "grey")
+scatter.smooth(predict(s_1ha, type = "response"), s_1ha_res, col = "grey")
+```
+```{r}
+# gam.check(s_1ha)
+```
+
+Basis dimension checking with `gam.check()` doesn't appear to work for `dlnm` crossbasis smooths.
+Instead I'll use a method from `?choose.k` to check for adequate knots.
+Unfortunately, `bs = "cs"` doesn't work with `dlnm`, so I'll use `select = TRUE` instead to reduce chance of overfitting.
+
+```{r}
+# looking for near zero edf
+gam(s_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(s_1ha)) #fine
+
+gam(s_1ha_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(s_1ha)) #fine
 ```
@@
 ```{r echo=TRUE}
+s_cf_res <- qresid(s_cf)
 par(mfrow = c(2,2))
 #histogram
-plot(density(qresid(s_cf)))
+plot(density(s_cf_res))
 #QQ plot
-qqnorm(qresid(s_cf)); qqline(qresid(s_cf))
+qqnorm(s_cf_res); qqline(s_cf_res)
 #Fitted vs. residuals
-scatter.smooth(predict(s_cf, type = "response"), qresid(s_cf), col = "grey")
+scatter.smooth(predict(s_cf, type = "response"), s_cf_res, col = "grey")
+```
+
+```{r}
+# looking for near zero edf
+gam(s_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(s_cf)) #fine
+
+gam(s_cf_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(s_cf)) 
 ```
@@
 Surface still looks different in a similar way though.
+
 ```{r}
@@
 ```
+```{r}
+g_1ha_res <- residuals(g_1ha)
+# looking for near zero edf
+gam(g_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(g_1ha)) #fine
+
+gam(g_1ha_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(g_1ha)) #fine
+```
 
@@
 ```
+
+```{r}
+g_cf_res <- residuals(g_cf)
+# looking for near zero edf
+gam(g_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(g_cf)) #fine
+
+gam(g_cf_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(g_cf)) #fine
+```
+
 ### Continuous Forest subsample
@@
 ```{r}
+f_1ha_res <- qresid(f_1ha)
 par(mfrow = c(2,2))
 #histogram
-plot(density(qresid(f_1ha)))
+plot(density(f_1ha_res))
 #QQ plot
-qqnorm(qresid(f_1ha)); qqline(qresid(f_1ha))
+qqnorm(f_1ha_res); qqline(f_1ha_res)
 #Fitted vs. residuals
-scatter.smooth(predict(f_1ha, type = "response"), qresid(f_1ha), col = "grey")
+scatter.smooth(predict(f_1ha, type = "response"), f_1ha_res, col = "grey")
 ```
+
+```{r}
+# looking for near zero edf
+gam(f_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(f_1ha)) #fine
+
+gam(f_1ha_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(f_1ha)) #fine
+```
+
 ### Continuous Forest
@@
 ```{r}
+f_cf_res <- qresid(f_cf)
 par(mfrow = c(2,2))
 #histogram
-plot(density(qresid(f_cf)))
+plot(density(f_cf_res))
 #QQ plot
-qqnorm(qresid(f_cf)); qqline(qresid(f_cf))
+qqnorm(f_cf_res); qqline(f_cf_res)
 #Fitted vs. residuals
-scatter.smooth(predict(f_cf, type = "response"), qresid(f_cf), col = "grey")
+scatter.smooth(predict(f_cf, type = "response"), f_cf_res, col = "grey")
+```
+
+```{r}
+# looking for near zero edf
+gam(f_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
+    data = model.frame(f_cf)) #fine
+
+gam(f_cf_res ~ s(spei_history, L,
+      bs = "cb",
+      k = c(3, 35),
+      xt = list(bs = "cr")),
+    gamma = 1.4,
+    select = TRUE,
+    data = model.frame(f_cf)) #fine
 ```
@@
 ### Continuous Forest subsample
+
 To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.

### 2021-08-10 review/Response to reviews.Rmd
added section numbering as cross-ref-able headers

@@
     reference_docx: template.docx
+    number_sections: true
 bibliography: references.bib
@@
 
-## General comments
+**General comments**
 
@@
 
-## Specific comments
+## {#R1-1}
 
-> 1.
-By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
+> By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
 The introduction could set up the reader better to understand both.
@@
 
-> 2.
-In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest.
+Testing testing
+
+## {#R1-2}
+
+> In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest.
 I wondered if the authors thought this was too obvious (edge effects, etc), but I think it'd be helpful to spell this out more clearly.
@@
 
-> 3.
-One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
+## {#R1-3}
+
+> One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
 As written in the methods and results, I found the implementation very difficult to follow.
@@
 
-> 4.
-Related to above, I think it's important to see more of the raw data than are presented in Fig.
+Testing testing (see also our response to \@ref(R1-2)).
+
+## {#R1-4}
+
+> Related to above, I think it's important to see more of the raw data than are presented in Fig.
 1.
@@
 
-> 5.
-I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.
+## {#R1-5}
+
+> I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.
 In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.
@@
 
-> 6.
-I was curious about why the paper did not include analyses of recruitment.
+## {#R1-6}
+
+> I was curious about why the paper did not include analyses of recruitment.
 Perhaps it is beyond the scope of this paper, but I wondered how new recruits responded to drought and wet also, both in number, as well as how big they were in the first year.
 
-> 7.  I had a hard time linking these statements in the results to the figures.
+## {#R1-7}
+
+> I had a hard time linking these statements in the results to the figures.
 >
-> -   L272-273 -- How do you see disparities in plant size here? I thought this was about growth?
-> -   L277-278 - How does this show the effect of size on growth is greatest for mid-size plants?
+> L272--273---How do you see disparities in plant size here? I thought this was about growth?
+>
+> L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
 More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
 More precise language would help.
-> -   L331 -- 332 -- How can you see in this figure that the response is 'more muted as indicated by the shape of the crossbasis function'?
+>
+> L331--332---How can you see in this figure that the response is 'more muted as indicated by the shape of the crossbasis function'?
 
-> 8.  L442-443 -- What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
+## {#R1-8}
 
-## Figures:
+> L442-443 -- What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
+
+## Figures: {#R1-figs}
 
@@
 
-## Specific Comments:
+## {#R2-1}
 
-> 1)  The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.
+> The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.
 
@@
 
-> 2)  Line 51-52.
+## {#R2-2}
+
+> Line 51-52.
 I find this sentence confusing.
@@
 
-> 3)  Line 146.
+## {#R2-3}
+
+> Line 146.
 Can you explain what you mean with "the height of each plant to the tallest leaf"?
@@
 
-> 4)  Lines 251ff: I don't see a reference of Fig 2 in the result section.
+## {#R2-4}
 
-> 5)  Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
+> Lines 251ff: I don't see a reference of Fig 2 in the result section.
+
+## {#R2-5}
+
+> Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
 It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
@@
 
-> 6)  Lines 287-289.
+## {#R2-6}
+
+> Lines 287-289.
 While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
@@
 
-> 7)  Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
+## {#R2-7}
+
+> Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
 Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
@@
 
-> 8)  Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
+## {#R2-8}
+
+> Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
 I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a.
@@
 
-> 9)  Line 323-325: I don't know why the growth of plants in fragments would affect the size of plant in CF.
+## {R2-9}
 
-> 10) Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).
+> Line 323-325: I don't know why the growth of plants in fragments would affect the size of plant in CF.
 
-> 11) Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.
+## {#R2-10}
 
-> 12) Lines 383 ff: see comment 7
+> Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).
 
-> 13) Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.
+## {#R2-11}
 
-> 14) 424-427: I think this is exaggerating since the differences between habitat types is not significant.
-15) You may also consider arguing that an 11-year long time series is not sufficient to detect delayed effects (see Tenhumberg et al.
-2018)
+> Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.
 
-> 16) Lines 437-438 change "starting expand" with "starting to expand"
+## {#R2-12}
 
-> 17) Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?
+> Lines 383 ff: see comment 7
 
-> 18) Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.
+## {#R2-13}
 
-> 19) Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
+> Line 412: I think it would be very informative if you were to expand on this and explain how exactly your method differs from what others have used to detect delayed effects.
 
-> 20) Lines 468 ff: I suggest deleting this paragraph.
+## {#R2-14}
+
+> 424-427: I think this is exaggerating since the differences between habitat types is not significant.
+
+## {#R2-15}
+
+> You may also consider arguing that an 11-year long time series is not sufficient to detect delayed
+effects (see Tenhumberg et al. 2018)
+
+## {#R2-16}
+
+> Lines 437-438 change "starting expand" with "starting to expand"
+
+## {#R2-17}
+
+> Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?
+
+## {#R2-18}
+
+> Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.
+
+## {#R2-19}
+
+> Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
+
+## {#R2-20}
+
+> Lines 468 ff: I suggest deleting this paragraph.
 It doesn't really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
@@
 
-> 21) The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
+## {#R2-21}
+
+> The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
 acuminata.
@@
 
-> 22) Lines 484-486: I don't think this study strongly support that plants do worse in fragments than CF.
+## {#R2-22}
+
+> Lines 484-486: I don't think this study strongly support that plants do worse in fragments than CF.
 23) Line 499: This is confusing.
@@
 
-## Specific comments
+## {#R3-1}
 
@@
 
+## {#R3-2}
+
 > Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames?
 > You describe why not using longer smooths, but not shorter.
+
+## {#R3-3}
 
@@
 
+## {#R3-4}
+
 > Line 205 - I'm not sure "bimodal" is correct here -- a cubic function can have 2 local extrema, is that what you mean?
 
+## {#R3-5}
+
 > Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.
+
+## {#R3-6}
 
@@
 
+## {#R3-7}
+
 > Line 253 - fig 1a shows mean change in size, not mean size?
+
+## {#R3-8}
 
@@
 
+## {#R3-9}
+
 > Line 260 -- you mention the CI of the model estimated coefficients, but these aren't provided anywhere that I could find.
@@
 > And report the variance of the random effect.
+
+## {#R3-10}
 
@@
 
+## {#R3-11}
+
 > Line 413 -- "our expansion of this approach ..." I'm not sure that the expansion is the penalty that prevents over fitting -- both Teller et al and Tenhumberg et al used the penalized spline function in mgcv.
@@
 
+## {#R3-12}
+
 > Line 471 -- this demographic delay left me a bit confused for a while until I stopped looking at figure 4b and started looking at 4a.

### 2021-08-10 review/Response to reviews.Rmd
Addressed R1-1 (issue #75)

@@
 > By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
-The introduction could set up the reader better to understand both.
-Since drought is well covered, adding background and hypotheses for why extreme wet would have a negative effect on vital rates would help.
+> The introduction could set up the reader better to understand both.
+> Since drought is well covered, adding background and hypotheses for why extreme wet would have a negative effect on vital rates would help.
 
-Testing testing
+We edited a sentence in the introduction that cites @estebanOtherSideDroughts2021 to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
+*"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction...".* Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon.
 
@@
 > In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest.
-I wondered if the authors thought this was too obvious (edge effects, etc), but I think it'd be helpful to spell this out more clearly.
-L466 -- 467 offer a new hypothesis, but briefly revisiting the 'old' evidence/hypotheses also seems important.
+> I wondered if the authors thought this was too obvious (edge effects, etc), but I think it'd be helpful to spell this out more clearly.
+> L466 -- 467 offer a new hypothesis, but briefly revisiting the 'old' evidence/hypotheses also seems important.
 
@@
 > One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
-As written in the methods and results, I found the implementation very difficult to follow.
-A slightly longer summary of the methods would help.
-In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
-While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit.
-One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn.
-Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).
+> As written in the methods and results, I found the implementation very difficult to follow.
+> A slightly longer summary of the methods would help.
+> In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
+> While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit.
+> One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn.
+> Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).
 
@@
 > Related to above, I think it's important to see more of the raw data than are presented in Fig.
-1.
-(I'm thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs).
-This could be an appendix, but it would help with interpreting Figures 3 -- 6.
+> 1.
+> (I'm thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs).
+> This could be an appendix, but it would help with interpreting Figures 3 -- 6.
 
@@
 > I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.
-In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.
-While the authors mention this, what about adding something short about when using this type of analysis is critical and when it might not be (should all demographers adopt this approach?
+> In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.
+> While the authors mention this, what about adding something short about when using this type of analysis is critical and when it might not be (should all demographers adopt this approach?
 Why or why not?)
@@
 > I was curious about why the paper did not include analyses of recruitment.
-Perhaps it is beyond the scope of this paper, but I wondered how new recruits responded to drought and wet also, both in number, as well as how big they were in the first year.
+> Perhaps it is beyond the scope of this paper, but I wondered how new recruits responded to drought and wet also, both in number, as well as how big they were in the first year.
 
@@
 >
-> L272--273---How do you see disparities in plant size here? I thought this was about growth?
+> L272--273---How do you see disparities in plant size here?
+> I thought this was about growth?
 >
 > L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
-More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
-More precise language would help.
+> More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
+> More precise language would help.
 >
@@
 
-> L442-443 -- What is a mechanism for why plants might flower more after drought? (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
+> L442-443 -- What is a mechanism for why plants might flower more after drought?
+> (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
 
@@
 > Line 51-52.
-I find this sentence confusing.
-What specific demographic mechanisms are you referring to?
-Do you mean that high mortality rate lead to extinction and we don't know what causes high mortality rates?
+> I find this sentence confusing.
+> What specific demographic mechanisms are you referring to?
+> Do you mean that high mortality rate lead to extinction and we don't know what causes high mortality rates?
 
@@
 > Line 146.
-Can you explain what you mean with "the height of each plant to the tallest leaf"?
-If the tallest leaf is growing more horizontally the height of the plant would be shorter than the length of the largest leaf?
+> Can you explain what you mean with "the height of each plant to the tallest leaf"?
+> If the tallest leaf is growing more horizontally the height of the plant would be shorter than the length of the largest leaf?
 
@@
 > Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
-It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
-Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.
+> It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
+> Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.
 
@@
 > Lines 287-289.
-While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
-I feel that the reporting of the results is a bit biased towards concluding that plants do worse in fragments compared to CF, especially since none of the result reported so far are significant.
+> While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
+> I feel that the reporting of the results is a bit biased towards concluding that plants do worse in fragments compared to CF, especially since none of the result reported so far are significant.
 
@@
 > Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
-Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
-I would like to see at least a justification for why representing the difference figures are meaningful.
+> Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
+> I would like to see at least a justification for why representing the difference figures are meaningful.
 
@@
 > Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
-I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a.
-What springs to my mind is that the area that is significant is much larger in in fragments compared to CF.
-It seems that in CF only a few months are significant.
-Surprisingly they fall not consistently in the dry or wet season.
+> I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a.
+> What springs to my mind is that the area that is significant is much larger in in fragments compared to CF.
+> It seems that in CF only a few months are significant.
+> Surprisingly they fall not consistently in the dry or wet season.
 
@@
 
-> You may also consider arguing that an 11-year long time series is not sufficient to detect delayed
-effects (see Tenhumberg et al. 2018)
+> You may also consider arguing that an 11-year long time series is not sufficient to detect delayed effects (see Tenhumberg et al. 2018)
 
@@
 
-> Lines 451-453: I think herbivory is this is very different from drought. Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
+> Lines 451-453: I think herbivory is this is very different from drought.
+> Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
 
@@
 > Lines 468 ff: I suggest deleting this paragraph.
-It doesn't really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
-Distributed Lag Non-Linear Models account for the exact timing of the weather events.
+> It doesn't really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
+> Distributed Lag Non-Linear Models account for the exact timing of the weather events.
 
@@
 > The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
-acuminata.
-For instance, why does too much rain reduce survival?
-Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past?
+> acuminata.
+> For instance, why does too much rain reduce survival?
+> Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past?
 Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)?
-Etc.
+> Etc.
 
@@
 > Lines 484-486: I don't think this study strongly support that plants do worse in fragments than CF.
-23) Line 499: This is confusing.
-If you measure size every year than you calculate growth as the change in size from one year to the next, don't you?
+
+##  {#R2-23}
+
+> Line 499: This is confusing.
+> If you measure size every year than you calculate growth as the change in size from one year to the next, don't you?
 

### 2021-08-10 doc/paper.Rmd
Addressed R1-1 (issue #75)

@@
 Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], and that the effects of climate change could be more pronounced when habitat is fragmented [@oliver2015; @holyoak2016].
-While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
-Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
+<!--# explain why R1.2 --> While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
+Climate models predict increases in the areas of the Amazon effected by severe drought and unusual wetness by the year 2100 [@duffyProjectionsFutureMeteorological2015].
+Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events (both drought and extreme wet) reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
 This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

### 2021-08-11 doc/validate_models.Rmd
switch to binned residuals plot using arm package

@@
 library(dlnm)
+library(arm)
 ```
@@
 qqnorm(s_1ha_res); qqline(s_1ha_res)
-#Fitted vs. residuals
-scatter.smooth(predict(s_1ha, type = "response"), s_1ha_res, col = "grey")
+#Fitted vs. residuals--binned residuals plot
+binnedplot(fitted(s_1ha), residuals(s_1ha, type = "response"))
 ```
@@
 qqnorm(s_cf_res); qqline(s_cf_res)
-#Fitted vs. residuals
-scatter.smooth(predict(s_cf, type = "response"), s_cf_res, col = "grey")
+#Fitted vs. residuals--binned residuals plot
+binnedplot(fitted(s_cf), residuals(s_cf, type = "response"))
 ```
@@
 qqnorm(f_1ha_res); qqline(f_1ha_res)
-#Fitted vs. residuals
-scatter.smooth(predict(f_1ha, type = "response"), f_1ha_res, col = "grey")
+#Fitted vs. residuals--binned residuals plot
+binnedplot(fitted(f_1ha), residuals(f_1ha, type = "response"))
 ```
@@
 qqnorm(f_cf_res); qqline(f_cf_res)
-#Fitted vs. residuals
-scatter.smooth(predict(f_cf, type = "response"), f_cf_res, col = "grey")
+#Fitted vs. residuals--binned residuals plot
+binnedplot(fitted(f_cf), residuals(f_cf, type = "response"))
 ```

### 2021-08-11 doc/validate_models.Rmd
re-organize validation

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: html_document
+output: 
+  html_document:
+    toc: yes
+    toc_float: yes
 ---
@@
 
-# TODO
+**TODO**
 
@@
 
-# Analysis
+# Survival
 
-## Survival
-
-Quantile residuals are more appropriate for binomial models.
-
-### Fragments
-
-```{r}
+```{r, echo=FALSE}
 withr::with_dir(here::here(),{
@@
 
-```{r}
-summary(s_1ha)
-```
+## Forest Fragments
 
-```{r}
-draw(s_1ha)
-```
+### Diagnostics
 
@@
 ```
+
+### Basis dimensions
+
 ```{r}
@@
 Basis dimension checking with `gam.check()` doesn't appear to work for `dlnm` crossbasis smooths.
-Instead I'll use a method from `?choose.k` to check for adequate knots.
+Instead I'll use a method described in the help file `?choose.k` to check for adequate knots.
 Unfortunately, `bs = "cs"` doesn't work with `dlnm`, so I'll use `select = TRUE` instead to reduce chance of overfitting.
@@
 
-### Continuous Forest
+### Summary
 
 ```{r}
-summary(s_cf)
+summary(s_1ha)
 ```
@@
 ```{r}
-draw(s_cf)
+draw(s_1ha)
 ```
+
+## Continuous Forest
+
+### Diagnostics
 
@@
 ```
+### Basis dimensions
 
@@
 
-### Continuous forest subsample
+### Summary
 
-To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
-For survival, edf is actually higher in CF with a larger sample size.
-With subsample, edf are more similar.
+```{r}
+summary(s_cf)
+```
+
+```{r}
+draw(s_cf)
+```
+
+## Effect of sample size
+
+Here I use a random sub-sample to check that differences between continuous forest and fragments are not purely due to sample size differences, particularly differences in the complexity of the crossbasis smooth.
 
@@
 
-```{r}
+```{r echo=FALSE}
 rm(s_1ha, s_cf, s_cf_sub)
@@
 
-## Growth
+# Growth
 
-```{r}
+```{r echo=FALSE}
 withr::with_dir(here::here(),{
@@
 
-### Fragments
+## Fragments
 
-```{r}
-summary(g_1ha)
-```
-
-```{r}
-draw(g_1ha)
-```
+### Diagnostics
 
@@
 
-### Continuous Forest
+### Summary
 
 ```{r}
-summary(g_cf)
+summary(g_1ha)
 ```
@@
 ```{r}
-draw(g_cf)
+draw(g_1ha)
 ```
+
+## Continuous Forest
+
+### Diagnostics
 
@@
 ```
+
+### Basis dimensions
 
@@
 
-### Continuous Forest subsample
+### Summary
+
+```{r}
+summary(g_cf)
+```
+
+```{r}
+draw(g_cf)
+```
+
+## Continuous Forest subsample
 
@@
 
-```{r}
+```{r echo=FALSE}
 rm(g_1ha, g_cf, g_cf_sub)
@@
 
-## Flowering
+# Flowering
 
-Quantile residuals are more appropriate for binomial models.
-
-```{r}
+```{r echo=FALSE}
 withr::with_dir(here::here(),{
@@
 
-### Fragments
+## Fragments
 
-```{r}
-summary(f_1ha)
-```
-
-```{r}
-draw(f_1ha)
-```
+### Diagnostics
 
@@
 
-### Continuous Forest
+### Summary
 
 ```{r}
-summary(f_cf)
+summary(f_1ha)
 ```
@@
 ```{r}
-draw(f_cf)
+draw(f_1ha)
 ```
+
+## Continuous Forest
+
+### Diagnostics
 
@@
 
-### Continuous Forest subsample
+### Summary
+
+```{r}
+summary(f_cf)
+```
+
+```{r}
+draw(f_cf)
+```
+
+## Continuous Forest subsample
 
@@
 Crossbasis surface looks extremely similar.
-
-### Flowering prob of largest plants
-
-Figure out flowering probability of largest quartile of plants in each habitat
-
-```{r}
-newdf_cf <-
-  f_cf$model %>% #use model data frame
-  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
-  filter(quartiles == 4)
-
-newdf_1ha <-
-  f_1ha$model %>% 
-  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
-  filter(quartiles == 4)
-```
-
-```{r}
-#equivalent to flowering once every __ years
-1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
-1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
-```
-
-```{r}
-rm(f_1ha, f_cf, f_cf_sub)
-```
 

### 2021-08-11 review/Response to reviews.Rmd
started response to one of the comments in #87

@@
 
-## {R2-9}
+##  {#R2-9}
 
@@
 
+What we show in the figures are not predictions (i.e. results of `predict()` in R), but evaluated smooths generated with the `gratia` package (equivalent to `mgcv::plot.gam()`).
+For figures 4, 5, and 6 we have added the model intercepts to those evaluated smooths and back-transformed (when applicable) the estimates.
+These plots, therefore, do not include the additive effects of covariates or random effects.
+We think that this is an intuitive way to display the partial effects of SPEI.
+<!--# Edit methods or figure captions to clarify? -->
+
+We apologize for not including a link to the code for the project.
+All analysis code can be found here: [<https://github.com/BrunaLab/HeliconiaDemography>](https://github.com/BrunaLab/HeliconiaDemography){.uri}.
+It is so rare for a reviewer to make suggestions based on analysis code, and we are very grateful for these comments!
+
 ##  {#R3-7}

### 2021-08-11 review/Response to reviews.Rmd
addressed confusion with size/growth.  Closes #89

@@
 > I thought this was about growth?
->
+
+Models which predict size in year t+1 as a function of size in year t, such as ours, are often described as growth functions because they can be used to project change in size (i.e. growth).
+However, we agree that it would be more clear for this section heading to be "size" rather than "growth"
+
 > L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
@@
 > More precise language would help.
->
+
+We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest." <!--# are there other edits we can make to clarify?
+A sentence introducing the concept that size t+1 ~ size t *is* growth?
+-->
+
 > L331--332---How can you see in this figure that the response is 'more muted as indicated by the shape of the crossbasis function'?
@@
 
+We clarified the reference to the figure: "...was preceded by a drop in average plant size in the 2002-2003 transition year (i.e. negative growth in Figure 1a)." Also see our response to \@ref(R1-7).
+
 ##  {#R3-8}

### 2021-08-11 doc/paper.Rmd
addressed confusion with size/growth.  Closes #89

@@
 *Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
-This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
+This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e.
+negative growth in Figure \@ref(fig:eda)a).
 The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
@@
 
-*Growth:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
+*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
 Plants in 1-ha fragments had on average `r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were `r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
@@
 As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
-While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).
+While the effect was generally similar across size classes and habitat types, the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).
 

### 2021-08-11 review/Response to reviews.Rmd
closes #89 for reals

@@
 
+That's correct.
+This line meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
+We've edited this sentence to read: "The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g.
+monthly) as climate drivers."
+
 # Reviewer 3

### 2021-08-11 doc/paper.Rmd
closes #89 for reals

@@
 This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
-The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
+The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g. monthly) as climate drivers.
 

### 2021-08-11 doc/paper.Rmd
updated descriptive statistics related text. #72

@@
   summarize(prop_surv = mean(surv)) %>% 
-    arrange(prop_surv)
+    arrange(habitat, prop_surv)
 surv_by_year_habitat
@@
 surv_by_year
+
+model_data %>% summarize(mean(surv))
 ```
@@
 
-*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
+*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
 This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e.
 negative growth in Figure \@ref(fig:eda)a).
-The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
-The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
 When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
@@
 
-*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
-Plants in 1-ha fragments had on average `r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were `r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
-Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 34% larger than those in forest fragments (150 ± 175 SD vs.
-112 ± 141 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
+```{r include=FALSE}
+summary_stats
+```
+
+*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 39.8 cm tall (± 26.2 SD).
+Plants in 1-ha fragments had on average `r format((1 - (2.45/2.85))*100, digits = 3)`% fewer shoots (2.5 ± 1.4 SD) and were `r format((1 - (35.57/39.75))*100, digits = 3)`% shorter (35.6 cm ± 23.8 SD).
+Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 32% larger than those in forest fragments (144 ± 176 SD vs.
+109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
 This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
@@
 
+```{r include=FALSE}
+summary_stats
+```
+
 *Flowering:* The overall proportion of plants that flowered was very low.
-While it was 27% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+While it was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
@@
 The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
-only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).
+17% in 1-ha fragments (Figure \@ref(fig:eda)c).
+
+```{r include=FALSE}
+repro_size <- model_data %>% filter(flwr == 1) %>% pull(size) %>% quantile(.1, na.rm = TRUE)
+
+model_data %>%
+  filter(size >= repro_size) %>% 
+  group_by(habitat, year) %>%
+  summarize(prop_flwr = sum(flwr, na.rm = TRUE)/n()) %>% 
+    arrange(year) %>% 
+  pivot_wider(names_from = habitat, values_from = prop_flwr) %>% 
+  mutate(diff = `1-ha` - CF) %>% 
+  arrange(desc(abs(diff)))
+```
 

### 2021-08-11 R/plot_eda.R

@@ -229,10 +229,10 @@ plot_eda_size <- function(demog_post, date_lims) {
 #' @param data demography data
 #' @param date_lims x-axis limits
 #' @param repro_size a size cutoff for plants that are considered reproductive.
-#'  The default, 168, corresponds to the upper 90th percentile of the size of
+#'  The default, 165, corresponds to the upper 90th percentile of the size of
 #'  all flowering plants in the dataset.
 #'  
-plot_eda_flwr <- function(data, date_lims, repro_size = 168) {
+plot_eda_flwr <- function(data, date_lims, repro_size = 165) {
   demog_plotdf <- eda_plot_df(data)
   date_breaks <- seq(date_lims[1], date_lims[2], by = "year")
   

### 2021-08-12 doc/paper.Rmd
added paragraph explaining partial effects plots.

@@
 
+To visualize results, we plotted partial effects plots for the one-dimensional smooth function of log size in the previous year and the two-dimensional smooth function of lagged SPEI (the crossbasis function) for each model.
+To make these plots more interpretable, we added the model intercept and (when applicable) back-transformed the y-axis to be on the scale of the response (e.g.
+a probability for survival and flowering).
+
 All analyses were conducted in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020] using the `targets` package for workflow management [@landauTargetsPackageDynamic2021].
@@
 *Flowering:* Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
-The responses in 1-ha fragments were also more muted as indicated by the shape of the crossbasis function (1-ha edf = 8.3, continuous forest edf = 10.6) (Figure \@ref(fig:flwr-spei)).
+SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
 This led to some important inter-habitat differences in plant responses to prior droughts.

### 2021-08-12 review/Response to reviews.Rmd
Explained partial effects plots to address part of #87

@@
 
+Our point was actually that SPEI had less of an impact on flowering probability in fragments.
+We edited this sentence to read: "SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure 6b)."
+
 ##  {#R1-8}
@@
 > Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.
+
+<!--# I think this will become more clear if I switch to plotting these as back-transformed. See comments in R1-figs -->
 
@@
 For figures 4, 5, and 6 we have added the model intercepts to those evaluated smooths and back-transformed (when applicable) the estimates.
-These plots, therefore, do not include the additive effects of covariates or random effects.
-We think that this is an intuitive way to display the partial effects of SPEI.
-<!--# Edit methods or figure captions to clarify? -->
+These plots, therefore, do not include the additive effects of other smooths, parametric covariates, or random effects.
+We think that this is an intuitive way to display the results.
+We added a paragraph to the methods explaining our strategy for visualizing the GAM results.
 
 We apologize for not including a link to the code for the project.
-All analysis code can be found here: [<https://github.com/BrunaLab/HeliconiaDemography>](https://github.com/BrunaLab/HeliconiaDemography){.uri}.
+All analysis code can be found here: [\<https://github.com/BrunaLab/HeliconiaDemography\>](https://github.com/BrunaLab/HeliconiaDemography){.uri}.
 It is so rare for a reviewer to make suggestions based on analysis code, and we are very grateful for these comments!

### 2021-08-12 review/Response to reviews.Rmd
improve (technical) explanation of models.  Closes #87

@@
 > I think it is probably aggregated at a higher level than individual?
+
+We thank the reviewer for the very helpful comments on our model equation.
+We have changed $y_i$ to $y_{ij}$ and added an explanation of the indexing as suggested.
+We have also corrected the error in the random effect equation so that only one variance is estimated.
+$d_{it}$ is intended to indicate that each plant observation is associated with a measure of SPEI for the past 36 months.
+The SPEI history varies among plants spatially (there are three grid cells in the weather data within BDFFP, so different plots experience different climate) and temporally (observations from all years are pooled).
+
 > I would reorganize the following paragraph so that all the descriptive text for the components of (1) occurs before the digression on the details of the crossbasis function.
+
+We moved the explanation of covariates and random effects earlier as suggested.
 

### 2021-08-12 doc/paper.Rmd
improve (technical) explanation of models.  Closes #87

@@
 DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
-They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
+<!--# DLNMs solve the problems created by including all possible climate windows in a single model---namely collinearity of predictors and the arbitrary choice of window length -->They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
 This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+<!--# The shape of the function can be penalized so that it is only as complex as the data supports, thus reducing effective degrees of freedom --> Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
@@
   \begin{split}
-  & g[ \textrm{E}(y_{i})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
-  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot j} 
+  & g[ \textrm{E}(y_{ij})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
+  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2), \;\textrm{for plot j} 
   \end{split}
@@
 
-where $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
+where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al.
 2017 for details).
+To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
+We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
+Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
+Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
+A random effect of plant ID was not included in survival models since each plant only dies once.
 The crossbasis function, $s_2(\cdot)$ can also be written:
@@
 Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
-Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.
+<!--# reference supplemental model diagnostics here --> Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.
 
-To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
@@
 We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
-We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
-Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
-Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
-A random effect of plant ID was not included in survival models since each plant only dies once.
 

### 2021-08-13 doc/validate_models.Rmd
add gam.check with no plot #92
track html outputs

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
 ## target knits Rmds in their own session, so load libraries here.
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, eval=TRUE}
 withr::with_dir(here::here(),{
@@
 
-```{r echo=TRUE}
+```{r echo=TRUE, eval=TRUE}
 s_1ha_res <- qresid(s_1ha)
@@
 
-```{r}
-# gam.check(s_1ha)
+```{r eval=TRUE}
+pdf(NULL)
+gam.check(s_1ha)
+dev.off()
 ```
@@
 
-```{r}
+```{r eval=TRUE}
 # looking for near zero edf
@@
 
-```{r}
+```{r eval=TRUE}
 summary(s_1ha)
@@
 
-```{r}
+```{r eval=TRUE}
 draw(s_1ha)
@@
 ### Basis dimensions
+
+```{r}
+pdf(NULL)
+gam.check(s_cf)
+dev.off()
+```
 
@@
 ```{r}
+pdf(NULL)
+gam.check(g_1ha)
+dev.off()
+```
+
+```{r}
 g_1ha_res <- residuals(g_1ha)
@@
 ### Basis dimensions
+
+```{r}
+pdf(NULL)
+gam.check(g_cf)
+dev.off()
+```
 
@@
 ```{r}
+pdf(NULL)
+gam.check(f_1ha)
+dev.off()
+```
+
+```{r}
 # looking for near zero edf
@@
 ### Basis dimensions
+
+```{r}
+pdf(NULL)
+gam.check(f_cf)
+dev.off()
+```
 

### 2021-08-13 review/Response to reviews.Rmd
add crossref
added some notes and crossrefs

@@
 > A slightly longer summary of the methods would help.
+
+...see also our response to \@ref(R3-3)
+
 > In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
@@
 
-Testing testing (see also our response to \@ref(R1-2)).
+<!--# make a table I guess? -->
 
@@
 
+<!--# figure should be back-transformed -->
+
 > Fig 3 -- I mentioned this earlier, but I wasn't really clear on how to interpret effect size.
@@
 > (I think the challenge for me is that I kept wanting the figures to fit into a more standard demographic paper -- like the probability of survival based on size, but then I realized something different was being shown here.)
+
+<!--# figure should be back-transformed -->
 
@@
 
-<!--# I think this will become more clear if I switch to plotting these as back-transformed. See comments in R1-figs -->
+<!--# I think this will become more clear if I switch to plotting these as back-transformed. --> ... see also our response to \@ref(R1-figs).
 
@@
 
+<!--# need to check the line to figure out the source of confusion, but we did test this with 84%CI overlap -->
+
 ##  {#R2-11}
@@
 > Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.
+
+<!--# this is not a plot that can have CIs.  The line doesn't imply statistical significance either.  Not sure how to respond -->
 
@@
 
+...see also our response to \@ref(R1-5).
+<!--# R1-5 is asking for more discussion on limitations of DLNMs -->
+
 ##  {#R2-14}
@@
 > 424-427: I think this is exaggerating since the differences between habitat types is not significant.
+
+...see also our response to \@ref(R2-10)
 
@@
 
+...see also our response to \@ref(R1-5) and \@ref(R2-13).
+
 ##  {#R2-16}
@@
 
+We have made this correction.
+
 ##  {#R2-17}
@@
 > Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?
+
+see our response to \@ref(R1-8).
 
@@
 > Distributed Lag Non-Linear Models account for the exact timing of the weather events.
+
+<!--# not sure how to respond.  I think there must be some underlying confusion.  Other reviewers like this explanation -->
 
@@
 
+<!--# yes, but might change number of knots, so might require more work revising this -->
+
 ##  {#R3-5}
@@
 
-What we show in the figures are not predictions (i.e. results of `predict()` in R), but evaluated smooths generated with the `gratia` package (equivalent to `mgcv::plot.gam()`).
-For figures 4, 5, and 6 we have added the model intercepts to those evaluated smooths and back-transformed (when applicable) the estimates.
+What we show in the figures are not predictions (i.e. results of `predict()` in R), but partial effects plots generated with the `gratia` package (equivalent to `mgcv::plot.gam()`).
+For figures 4, 5, and 6 we have added the model intercepts to evaluated smooths and back-transformed (when applicable) the estimates.
 These plots, therefore, do not include the additive effects of other smooths, parametric covariates, or random effects.
@@
 We added a paragraph to the methods explaining our strategy for visualizing the GAM results.
+
+<!--# see also response to comments where we decide to backtransform univariate smooths -->
 
@@
 
+...see also our response to \@ref(R1-figs)
+
 ##  {#R3-9}
@@
 > And report the variance of the random effect.
+
+...see also our response to \@ref(R1-3)
 

### 2021-08-13 review/Response to reviews.Rmd
explain exclude.too.far (#94)

@@
 > This could be an appendix, but it would help with interpreting Figures 3 -- 6.
+
+One way we approached this is by excluding areas of the heatmap that are far from data, equivalent to not drawing regression lines beyond the range of data.
+We implemented this with the `exclude.too.far()` function in the `mgcv` package.
+Displaying raw data or residuals on the heatmaps was too messy and not informative in a 2d plot.
+Instead, we have explored "slices" of these heatmaps in a supplement <!--# TODO, make that supplement! -->.
 
@@
 
+<!--# just the height of the plant -->
+
 ##  {#R2-4}
@@
 
-<!--# I think this will become more clear if I switch to plotting these as back-transformed. --> ... see also our response to \@ref(R1-figs).
+<!--# I think this will become more clear if I switch to plotting these as back-transformed. -->
+
+... see also our response to \@ref(R1-figs).
 

### 2021-08-16 doc/validate_models.Rmd
switch to k.check() for no plotting

@@
 ```{r eval=TRUE}
-pdf(NULL)
-gam.check(s_1ha)
-dev.off()
+k.check(s_1ha)
 ```
@@
 gam(s_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(s_1ha)) #fine
+    data = model.frame(s_1ha)) %>%
+  edf() #fine
 
@@
     select = TRUE,
-    data = model.frame(s_1ha)) #fine
+    data = model.frame(s_1ha)) %>%
+  edf()#fine
 ```
@@
 ```{r}
-pdf(NULL)
-gam.check(s_cf)
-dev.off()
+k.check(s_cf)
 ```
@@
 ```{r}
-pdf(NULL)
-gam.check(g_1ha)
-dev.off()
+k.check(g_1ha)
 ```
@@
 ```{r}
-pdf(NULL)
-gam.check(g_cf)
-dev.off()
+k.check(g_cf)
 ```
@@
 ```{r}
-pdf(NULL)
-gam.check(f_1ha)
-dev.off()
+k.check(f_1ha)
 ```
@@
 ```{r}
-pdf(NULL)
-gam.check(f_cf)
-dev.off()
+k.check(f_cf)
 ```

### 2021-08-18 doc/validate_models.Rmd
add supplemental to choose correct knot number #92

@@
 library(arm)
+source(here::here("R", "utils.R"))
 ```
@@
 
-- increase knots for some models (https://github.com/BrunaLab/HeliconiaDemography/issues/92)
+- check that # of knots are adequate
 - check diagnostics for individual-level random effects.  Doesn't look good, but not sure how to interpret or what to do.
@@
 ```{r echo=TRUE, eval=TRUE}
-s_1ha_res <- qresid(s_1ha)
+s_1ha_qres <- qresid(s_1ha)
 par(mfrow = c(2,2))
 #histogram
-plot(density(s_1ha_res))
+plot(density(s_1ha_qres))
 #QQ plot
-qqnorm(s_1ha_res); qqline(s_1ha_res)
+qqnorm(s_1ha_qres); qqline(s_1ha_qres)
 #Fitted vs. residuals--binned residuals plot
@@
 ```{r eval=TRUE}
-# looking for near zero edf
-gam(s_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(s_1ha)) %>%
-  edf() #fine
+check_res_edf(s_1ha)
+```
 
-gam(s_1ha_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(s_1ha)) %>%
-  edf()#fine
+The crossbasis smooth possibly needs more knots.
+
+```{r}
+# plot_lag_slice(s_1ha, "spei_history", lag =  0) + labs(title = "surv, 1ha")
 ```
+
+Here you can see how the CIs hardly overlap 0, but it could be an artifact of the line not being allowed to be wiggly enough.
 
@@
 ```{r echo=TRUE}
-s_cf_res <- qresid(s_cf)
+s_cf_qres <- qresid(s_cf)
 par(mfrow = c(2,2))
 #histogram
-plot(density(s_cf_res))
+plot(density(s_cf_qres))
 #QQ plot
-qqnorm(s_cf_res); qqline(s_cf_res)
+qqnorm(s_cf_qres); qqline(s_cf_qres)
 #Fitted vs. residuals--binned residuals plot
@@
 ```
+
 ### Basis dimensions
@@
 # looking for near zero edf
-gam(s_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(s_cf)) #fine
-
-gam(s_cf_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(s_cf)) 
+check_res_edf(s_cf)
 ```
@@
 ```{r}
-g_1ha_res <- residuals(g_1ha)
-# looking for near zero edf
-gam(g_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(g_1ha)) #fine
-
-gam(g_1ha_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(g_1ha)) #fine
+check_res_edf(g_1ha)
 ```
@@
 ```{r}
-g_cf_res <- residuals(g_cf)
-# looking for near zero edf
-gam(g_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(g_cf)) #fine
+check_res_edf(g_cf)
+```
 
-gam(g_cf_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(g_cf)) #fine
+```{r}
+# plot_lag_slice(g_1ha, "spei_history", lag = 33)
 ```
+
+hmm... CIs are **super** narrow here, but effect size is **tiny**
 
@@
 ```{r}
-f_1ha_res <- qresid(f_1ha)
+f_1ha_qres <- qresid(f_1ha)
 par(mfrow = c(2,2))
 #histogram
-plot(density(f_1ha_res))
+plot(density(f_1ha_qres))
 #QQ plot
-qqnorm(f_1ha_res); qqline(f_1ha_res)
+qqnorm(f_1ha_qres); qqline(f_1ha_qres)
 #Fitted vs. residuals--binned residuals plot
@@
 # looking for near zero edf
-gam(f_1ha_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(f_1ha)) #fine
-
-gam(f_1ha_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(f_1ha)) #fine
+check_res_edf(f_1ha)
 ```
@@
 ```{r}
-f_cf_res <- qresid(f_cf)
+f_cf_qres <- qresid(f_cf)
 par(mfrow = c(2,2))
 #histogram
-plot(density(f_cf_res))
+plot(density(f_cf_qres))
 #QQ plot
-qqnorm(f_cf_res); qqline(f_cf_res)
+qqnorm(f_cf_qres); qqline(f_cf_qres)
 #Fitted vs. residuals--binned residuals plot
@@
 # looking for near zero edf
-gam(f_cf_res ~ s(log_size_prev, k = 10, bs = "cs"), gamma = 1.4,
-    data = model.frame(f_cf)) #fine
-
-gam(f_cf_res ~ s(spei_history, L,
-      bs = "cb",
-      k = c(3, 35),
-      xt = list(bs = "cr")),
-    gamma = 1.4,
-    select = TRUE,
-    data = model.frame(f_cf)) #fine
+check_res_edf(f_cf)
 ```

### 2021-08-18 R/utils.R

@@ -84,3 +84,22 @@ as_living <- function(x, n = 2) {
 nearest <- function(x, val) {
   abs(x - val) == min(abs(x - val))
 }
+
+
+
+#' Check crossbasis for adequate knots
+#'
+#' Implements a strategy described in detail in `?mgcv::choose.k` for checking
+#' if there are adequate knots.  Basically looking for any pattern in residuals
+#'
+#' @param model a gam with a crossbasis smooth described by `s(spei_history, L,
+#'   bs = "cb")`
+#'   
+check_res_edf <- function(model) {
+  res <- residuals(model)
+  mgcv::gam(res ~ te(spei_history, L, k = c(20, 35), bs = "cs"),
+      gamma = 1.4,
+      data = model.frame(model)) %>%
+    gratia::edf() %>% 
+    dplyr::mutate(edf = round(edf, 1))
+}

### 2021-08-18 _targets.R

@@ -1,16 +1,16 @@
 ## Set options for using tar_make_clustermq()
 ## Uncomment and edit these lines to run model targets on a cluster computer via SSH
-# options(
-#   clustermq.scheduler = "ssh",
-#   clustermq.ssh.host = "ericscott@hpg.rc.ufl.edu", # use your user and host
-#   clustermq.ssh.log = "~/cmq_ssh.log", # log for easier debugging
-#   clustermq.worker.timeout = 2400
-# )
-
-## Uncomment these lines to run locally on multiple cores
 options(
-  clustermq.schedule = "multicore"
+  clustermq.scheduler = "ssh",
+  clustermq.ssh.host = "ericscott@hpg.rc.ufl.edu", # use your user and host
+  clustermq.ssh.log = "~/cmq_ssh.log", # log for easier debugging
+  clustermq.worker.timeout = 2400
 )
+
+# ## Uncomment these lines to run locally on multiple cores
+# options(
+#   clustermq.schedule = "multicore"
+# )
 
 ## Load your packages, e.g. library(targets).
 source("./packages.R")
@@ -43,6 +43,9 @@ tar_plan(
   
   # Data validation
   tar_render(validate_data, "doc/validate_data.Rmd", deployment = "main"),
+  
+  # Select number of knots for GAMs
+  tar_render(knot_choice, "doc/knot_choice.Rmd", deployment = "main"),
   
   # Fit demographic models
   tar_target(s_cf,  fit_surv(model_data_cf),    deployment = "worker"),

### 2021-08-18 review/Response to reviews.Rmd
clarified height (#82)

@@
 
-<!--# just the height of the plant -->
+This is just the height of the plant from the ground to the tallest leaf leaf tip, not the tip of the youngest leaf.
+The reviewer is correct that the height of a plant growing more horizontally (for example if it was knocked over by a branchfall) could be shorter than the length from the base to the end of the tallest leaf.
+We edited the description to read: "...the height of each plant (i.e. distance from the ground to the tallest leaf tip)"
 

### 2021-08-18 doc/paper.Rmd
clarified height (#82)

@@
 
-Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
+Each plot was subdivided into 50 quadrats ($10 \times 10$m) to simplify annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant (i.e.
+distance from the ground to the tallest leaf tip), and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
 In this study, we used the product of shoot number and plant height as our measure of plant size.

### 2021-08-18 review/Response to reviews.Rmd
explain why no IPM (closes #81)

@@
 > Perhaps it is beyond the scope of this paper, but I wondered how new recruits responded to drought and wet also, both in number, as well as how big they were in the first year.
+
+The reviewer is correct that recruitment is an important aspect of demography that could be effected by SPEI and fragmentation.
+However, we believed the analyses in this paper were complicated enough to not include additional analyses.
+We intend to follow up this paper with analyses of recruitment and integrating all vital rate models with an integral projection model.
+We simply felt doing all of this (describing and interpreting DLNMs, modeling recruitment, and integrating all into an IPM) would be too much for a single paper.
 
@@
 
+We thought it would be too much to describe and interpret DLNMs for the vital rates, model recruitment, *and* integrate all those models into an integral projection model.
+We though the description and interpretation of the DLNMs for survival, size, and flowering were complex and interesting enough to stand on their own.
+We do intend to follow up with an IPM in a future publication.
+See also our response to \@ref(R1-6).
+
 ##  {#R3-1}

### 2021-08-23 review/Response to reviews.Rmd
closes #92 (but does not update results yet)

@@
 
+This comment was especially helpful and made us re-think limiting the number of "knots" to 3 in the SPEI dimension.
+The term *k* is probably more accurately described as the "maximum basis complexity" or "maximum basis dimension".
+We had thought that by limiting k to 3, we were simply limiting the complexity of the smooth in the SPEI dimension.
+However, after taking the reviewer's suggestions we realized that there were other unexpected consequences.
+We increased *k* for the SPEI dimension for all models and for other smooths for some models.
+This resulted in some changes in the the shape of the crossbasis smooth and which areas of it were statistically significant.
+We have updated the methods and results sections accordingly.
+<!--# TODO: update results section! -->\
+\
+We have also now included a supplemental document with model validation, including checks that *k* is adequately large
+.
+
 ##  {#R3-6}

### 2021-08-23 README.Rmd
closes #92 (but does not update results yet)

@@
 
+**Using HPC**:
+Also, you may need to install packages on the remote HPC if using `tar_make_clustermq()` with the SSH option.
+Unfortunately, I'm not sure if there is an easy way to do this using `renv`.
+You should, however, make sure the same versions of the packages are installed locally and on the HPC, especially `gratia` which I'm using a development version of that has important bugs fixed, but not new bugs introduced.
+
 ## Additional Dependencies

### 2021-08-23 doc/paper.Rmd
closes #92 (but does not update results yet)

@@
 These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
-Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e.
-number of lagged SPEI values for each observation with 36 months as a maximum lag) and 3 knots for the drought response dimension to restrict the shape of the fitted response to drought to bimodal when most complex.
-Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
+Penalized cubic regression splines were used for both marginal bases of the crossbasis function.
+We increased *k*, the maximum basis dimension, as necessary to be sure it was adequate.
+Because of penalization, the maximum basis dimension is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
 <!--# reference supplemental model diagnostics here --> Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.

### 2021-08-23 R/fit_models.R

@@ -8,7 +8,7 @@
 #' @param k vector of knots to pass to `s(log_size_prev)` (default 10), and
 #'   `s(spei_history, L)` (default 3, 35).
 #'   
-fit_surv <- function(data, k = c(10, 3, 35)) {
+fit_surv <- function(data, k = c(10, 10, 15)) {
   stopifnot(length(k) == 3L) #check that 3 values for k are supplied
   data2 <- data %>% mutate(flwr_prev = factor(flwr_prev))
   f <- surv ~ 
@@ -37,7 +37,7 @@ fit_surv <- function(data, k = c(10, 3, 35)) {
 #' @param k vector of knots to pass to `s(log_size_prev)` (default 10), and
 #'   `s(spei_history, L)` (default 3, 35).
 #'
-fit_growth <- function(data, k = c(10, 3, 35)){
+fit_growth <- function(data, k = c(10, 10, 15)){
   stopifnot(length(k) == 3L) #check that 3 values for k are supplied
   # use only living plants
   data2 <- data %>% dplyr::filter(surv == 1, !is.na(log_size)) %>% 
@@ -69,7 +69,7 @@ fit_growth <- function(data, k = c(10, 3, 35)){
 #' @param data prepared model data
 #' @param ind_raneff logical; include individual-level random effect (i.e. plant ID as a random effect)?
 #' 
-fit_flwr <- function(data, k = c(10, 3, 35), ind_raneff = FALSE) {
+fit_flwr <- function(data, k = c(10, 10, 15), ind_raneff = FALSE) {
   stopifnot(length(k) == 3L) #check that 3 values for k are supplied
   # use only living plants
   data2 <- data %>% dplyr::filter(surv == 1, !is.na(log_size)) %>%

### 2021-08-23 _targets.R

@@ -9,14 +9,14 @@ options(
 
 # ## Uncomment these lines to run locally on multiple cores
 # options(
-#   clustermq.schedule = "multicore"
+#   clustermq.scheduler = "multiprocess"
 # )
 
 ## Load your packages, e.g. library(targets).
 source("./packages.R")
 
 ## Load your R files
-lapply(list.files("./R", full.names = TRUE), source)
+lapply(list.files("./R", full.names = TRUE), source) 
 
 ## Set options
 options(tidyverse.quiet = TRUE)
@@ -48,24 +48,22 @@ tar_plan(
   tar_render(knot_choice, "doc/knot_choice.Rmd", deployment = "main"),
   
   # Fit demographic models
-  tar_target(s_cf,  fit_surv(model_data_cf, k = c(10,15,15)),    deployment = "worker"),
-  tar_target(s_1ha, fit_surv(model_data_1ha, k = c(10,10,10)),   deployment = "worker"),
-  tar_target(g_cf,  fit_growth(model_data_cf, k = c(25,5,15)),  deployment = "worker"),
-  tar_target(g_1ha, fit_growth(model_data_1ha, k = c(10,5,15)), deployment = "worker"),
+  tar_target(s_cf,  fit_surv(model_data_cf, k = c(10,15,15))),
+  tar_target(s_1ha, fit_surv(model_data_1ha, k = c(10,10,10))),
+  tar_target(g_cf,  fit_growth(model_data_cf, k = c(25,5,15))),
+  tar_target(g_1ha, fit_growth(model_data_1ha, k = c(10,5,15))),
   
   #### NOTE: The f_cf target takes ~20 hrs to run on a single core on a MacBook
   #### pro and results in a 816MB object when ind_raneff = TRUE.  Set to FALSE
   #### for testing
-  tar_target(f_cf, fit_flwr(model_data_cf, k = c(10,15,15), ind_raneff = TRUE),
-             deployment = "worker"),
-  tar_target(f_1ha, fit_flwr(model_data_1ha, k = c(10,10,18), ind_raneff = TRUE),
-             deployment = "worker"),
+  tar_target(f_cf, fit_flwr(model_data_cf, k = c(10,15,15), ind_raneff = FALSE)),
+  tar_target(f_1ha, fit_flwr(model_data_1ha, k = c(10,10,18), ind_raneff = TRUE)),
   
   # Validate and summarize results
[…]

### 2021-08-23 review/Response to reviews.Rmd
added more detail about fitting model.  Closes #77, I think.

@@
 > A slightly longer summary of the methods would help.
+
+We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages.
+<!--# add line reference before submiting! --> We are not entirely sure if this is the type of description the reviewer was looking for.
+If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
 

### 2021-08-23 doc/paper.Rmd
added more detail about fitting model.  Closes #77, I think.

@@
 acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.
-
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
 The general form of the vital rate ($y$) models was as follows:
@@
 <!--# reference supplemental model diagnostics here --> Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.
+
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
+The crossbasis smooth was coded as:
+
+    s(spei_history, L, bs = "cb", xt = list(bs = "cr"))
+
+Where `spei_history` is a $n \times m$ matrix with each row ($n$) containing the climate history of a single plant in a year and $m=36$ columns representing the past weather conditions starting at the month of the census and going backwards in time 36 months.
+`L` is an $n \times m$ matrix with columns describing the lag structure, i.e. the integers 0 through 36.
+The spline basis, `"cb"` is provided by the `dlnm` package, and we specified the marginal bases as cubic regression splines with the `xt` argument.
+For the full code used to fit the models, see <!--# insert citation to Zenodo archive -->.
 

### 2021-08-23 review/Response to reviews.Rmd
go back to back-transformed marginal effects plots (#83, #95)

@@
 Displaying raw data or residuals on the heatmaps was too messy and not informative in a 2d plot.
-Instead, we have explored "slices" of these heatmaps in a supplement <!--# TODO, make that supplement! -->.
+\
+\
+... see also our response to \@ref(R1-figs).
 
@@
 
-<!--# figure should be back-transformed -->
+This was referring to a previous iteration of the figure where the y-axis was on the response scale.
+We have reverted the figure to this version (see response below) and this should be obvious now.
 
@@
 
-<!--# figure should be back-transformed -->
+These plots were the evaluated smooths on the link scale.
+We agree that they were difficult to interpret in that form, so we have added the model intercept and back-transformed them to be on their respective response scales (e.g.
+probability for flowering and survival).
+This is equivalent to the marginal effect of log(size_t)---that is, without considering the additive effect of lagged SPEI or the random intercepts for different plots and for plants that had not flowered in the previous year (the reference level for the `flwr_prev` factor we included as a parametric covariate).
+We have not clarified that in the figure caption.
 

### 2021-08-23 doc/paper.Rmd
go back to back-transformed marginal effects plots (#83, #95)

@@
 ```{r}
-size_cap <- "Smooth effect from models of plant size in the previous census on (a) survival, (b) log(size), and (c) flowering probability; these values correspond to the additive term s~1~(z~i~) in eq.
-1.
+size_cap <- "Marginal effect of plant size in the previous census on (a) survival, (b) log(size), and (c) flowering probability.
+Plots a--c were created by evaluating the smooth functions of $log(size_t)$ at observed values, adding the model intercepts, and back-transforming (when appropriate) to the response scale (e.g.
+probability for a and c).
 The bands depicting the 95% confidence interval include uncertainty in the intercept and uncertainty due to smoothness selection; the smooths for 1-ha fragments and continuous forest are fit in separate models.

### 2021-08-23 R/plot_smooths.R

@@ -8,10 +8,23 @@
 #'
 plot_covar_smooth <- function(cf_model, frag_model, covar) {
 
-  cf <- gratia::smooth_estimates(cf_model, covar, unconditional = TRUE) %>%
-    add_confint()
-  frag <- gratia::smooth_estimates(frag_model, covar, unconditional = TRUE) %>% add_confint()
-  data <- bind_rows("1-ha" = frag, "CF" = cf, .id = "habitat")
+  cf <- 
+    gratia::smooth_estimates(cf_model, covar, unconditional = TRUE) %>%
+    add_confint() %>% 
+    add_column(intercept = coef(cf_model)[1])
+  
+  frag <- 
+    gratia::smooth_estimates(frag_model, covar, unconditional = TRUE) %>%
+    add_confint() %>% 
+    add_column(intercept = coef(frag_model)[1])
+  
+  linkinv <- cf_model$family$linkinv
+  
+  data <-
+    bind_rows("1-ha" = frag, "CF" = cf, .id = "habitat") %>% 
+    # add intercept and backtransform to response scale
+    mutate(across(c(est, lower_ci, upper_ci), ~linkinv(.x + intercept)))
+    
   
   p <- 
     ggplot(data, aes_string(x = covar, color = "habitat", linetype = "habitat")) +

### 2021-08-23 _targets.R

@@ -102,13 +102,13 @@ tar_plan(
   ## Size covariate
   s_covar_plot = plot_covar_smooth(frag_model = s_1ha, cf_model = s_cf,
                                    covar = "log_size_prev") + 
-    labs(x = TeX("$log(size_t)$"), y = "Effect [survival]"),
+    labs(x = TeX("$log(size_t)$"), y = "P(survival)"),
   g_covar_plot = plot_covar_smooth(frag_model = g_1ha, cf_model = g_cf,
                                    covar = "log_size_prev") +
-    labs(x = TeX("$log(size_t)$"), y = TeX("Effect \\[$log(size_{t+1})\\]")),
+    labs(x = TeX("$log(size_t)$"), y = TeX("log(size_{t+1})")),
   f_covar_plot = plot_covar_smooth(frag_model = f_1ha, cf_model = f_cf,
                                    covar = "log_size_prev") +
-    labs(x = TeX("$log(size_t)$"), y = "Effect [flowering]"),
+    labs(x = TeX("$log(size_t)$"), y = "P(flowering)"),
   size_plot = make_size_plot(
     s = s_covar_plot,
     g = g_covar_plot,

### 2021-08-24 review/Response to reviews.Rmd
edits to figure captions (#83, closes #95)

@@
 > A little bit more explanation in the legends would help these figures be more intuitive.
+
+We've revised the figure captions to clarify: *"Outlines show combinations of SPEI and lag time that have a significant effect on survival, defined as areas where the 95% confidence interval around the response does not overlap the intercept."*
 
@@
 
+In the updated version of figure 4 (see response to \@ref(R3-5)), a smaller area of the plot is significantly different from the intercept, and heatmap area far from observed data is grayed out.
+We believe these two things make it easier to see the outlines around significant values, but we are open to trying other modifications of the plot if the reviewers think it is still difficult to interpret in this way.
+
 ##  {#R3-11}

### 2021-08-24 doc/paper.Rmd
edits to figure captions (#83, closes #95)

@@
 ```{r}
-s_spei_cap <- "Smooth effect of lagged SPEI on *H.
+s_spei_cap <- "Marginal effect of lagged SPEI on *H.
 acuminata* survival in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
-Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October).
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+Outlines show combinations of SPEI and lag time that have a significant effect on survival, defined as areas where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October).
+For a and b, the model intercepts were added to the evaluated crossbasis smooths and values were back-transformed to the response scale (i.e.
+probabilities).
+Areas of the fitted smooth far from observed values (i.e.
+combinations of lag time and SPEI) are shown in grey."
 
@@
 ```{r}
-g_spei_cap <- "Smooth effect of lagged SPEI on *H.
-acuminata* growth in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
-Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October)."
+g_spei_cap <- "Marginal effect of lagged SPEI on *H.
+acuminata* size in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
+Outlines show combinations of SPEI and lag time that have a significant effect on plant size, defined as areas where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October).
+For a and b, the model intercepts were added to the evaluated crossbasis smooths.
+Areas of the fitted smooth far from observed values (i.e.
+combinations of lag time and SPEI) are shown in grey."
 ```
@@
 ```{r}
-f_spei_cap <- "Smooth effect of lagged SPEI on *H.
+f_spei_cap <- "Marginal effect of lagged SPEI on *H.
 acuminata* flowering probability in (a) continuous forest, (b) 1-ha fragments, (c) and the difference between the two.
-Outlines show regions where the effect of SPEI is significant, defined as those where the 95% confidence interval around the response does not overlap the intercept.
-The bar on the bottom of each panel indicates the wet seasons (black, November–May) and dry seasons (white, June–October).
-Surface is modeled as a crossbasis function with cubic regression splines for each marginal basis.
-Model intercepts were added to fitted values of the crossbasis function and back-transformed to the response scale."
+Outlines show combinations of SPEI and lag time that have a significant effect on probability of flowering, defined as areas where the 95% confidence interval around the response does not overlap the intercept.
+The bar on the bottom of each panel indicates wet seasons (black, November–May) and dry seasons (white, June–October).
+For a and b, the model intercepts were added to the evaluated crossbasis smooths and values were back-transformed to the response scale (i.e.
+probabilities).
+Areas of the fitted smooth far from observed values (i.e.
+combinations of lag time and SPEI) are shown in grey."
 ```

### 2021-08-24 review/Response to reviews.Rmd
add more detail on limiations of DLNMs and related methods (#80)

@@
 If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
-
-...see also our response to \@ref(R3-3)
+See also our response to \@ref(R3-3).
 
@@
 
-One way we approached this is by excluding areas of the heatmap that are far from data, equivalent to not drawing regression lines beyond the range of data.
+We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs), but showing the raw data made the plots visually cluttered, in our opinion.
+Another way we approached this comment is by excluding areas of the heatmaps in figures 4-6 that are far from observed data, equivalent to not drawing regression lines beyond the range of data.
 We implemented this with the `exclude.too.far()` function in the `mgcv` package.
-Displaying raw data or residuals on the heatmaps was too messy and not informative in a 2d plot.
-\
-\
-... see also our response to \@ref(R1-figs).
+Displaying raw data on the heatmaps was too messy and not informative in a 2d plot.
 
@@
 > I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.
+
+<!--# The following isn't in the main text yet.  Just ideas. -->
+
+The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored.
+@tellerLinkingDemographyDrivers2016 used simulations to show that 20--25 years of data were required to detect lagged effects using their method, but this cutoff depends on the effect size and the number of climate extremes in the dataset.
+Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required, but this is unpublished currently <!--# cite Aldo Compagnoni's ESA talk -->.
+
 > In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.
+
+On the contrary, averaging weather conditions over an entire year likely obscures ways plants perceive environmental variation.
+For example, a drought in the wet season may have a different effect than a drought in the dry season.
+This differential response to season- or month-level climate data is observable in yearly plant census.
+Collapsing finer temporal scale environmental drivers to yearly averages (or a few means of arbitrarily chosen climate windows) is exactly what we avoid with this method.
+
 > While the authors mention this, what about adding something short about when using this type of analysis is critical and when it might not be (should all demographers adopt this approach?
 Why or why not?)
+
+We have added additional detail to the conclusions recommending that all demographers consider lagged effects when possible in light of recent research (including this study).
+<!--# insert line ref -->
 
@@
 > Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.
-
-<!--# I think this will become more clear if I switch to plotting these as back-transformed. -->
 

### 2021-08-24 doc/paper.Rmd
add more detail on limiations of DLNMs and related methods (#80)

@@
 
-Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021].
+Recent research, including our own presented here, shows that lagged effects of climate drivers on demography may be the norm [@eversLaggedDormantSeason2021].
+As such, we suggest that all demographers investigate the possibility of lagged environmental effects, ideally using methods that do not require an *a priori* choice of lag times such as DLNMs, automated algorithms for choosing critical weather windows[@vandepolIdentifyingBestClimatic2016], or Bayesian methods that incorporate "ecological memory" [@ogleQuantifyingEcologicalMemory2015].
+
+Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->.
 Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].

### 2021-08-24 review/Response to reviews.Rmd
misc comment (#82)

@@
 
-... see also our response to \@ref(R1-figs).
+We thank the reviewer for pointing out this mistake.
+We have edited this line to read "Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats." See also our response to \@ref(R1-figs).
 

### 2021-08-24 doc/paper.Rmd
misc comment (#82)

@@
 })
-pull_ci(g_cf, digits = 2)
-pull_ci(g_1ha, digits = 2)
+pull_ci(g_cf, digits = 3)
+pull_ci(g_1ha, digits = 3)
 rm(g_cf, g_1ha)
@@
 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
-This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
+Plants were significantly larger in continuous forest (no overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
 
@@
 As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
-While the effect was generally similar across size classes and habitat types, the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)a).
+Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats (Figure \@ref(fig:size-plot)a).
 

### 2021-08-24 doc/paper.Rmd
effect of habitat on flowering is now significant (#72)

@@
 *Flowering:* The overall proportion of plants that flowered was very low.
-While it was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
+Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
+`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (84% CIs of model intercepts did not overlap).
 The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
-Despite the flowering probability of large plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
+Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
 The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
@@
 ```{r include=FALSE}
-repro_size <- model_data %>% filter(flwr == 1) %>% pull(size) %>% quantile(.1, na.rm = TRUE)
+repro_size <- model_data %>%
+  filter(flwr == 1) %>%
+  pull(size) %>%
+  quantile(.1, na.rm = TRUE)
 

### 2021-08-24 _targets.R
re-ran pipeline

@@ -3,8 +3,8 @@
 options(
   clustermq.scheduler = "ssh",
   clustermq.ssh.host = "ericscott@hpg.rc.ufl.edu", # use your user and host
-  clustermq.ssh.log = "~/cmq_ssh.log", # log for easier debugging
-  clustermq.worker.timeout = 2400
+  clustermq.ssh.log = "~/cmq_ssh.log" # log for easier debugging
+  # clustermq.worker.timeout = 2400
 )
 
 # ## Uncomment these lines to run locally on multiple cores

### 2021-08-26 review/Response to reviews.Rmd
add 84%CI table (#78, #85)

@@
 
-<!--# make a table I guess? -->
+<!--# include plots with raw data?  Would need to be supplemental probably, because messy. -->
 
@@
 
+<!--# need to double-check and revise --> The difference in flowering probability between cointinuous forests and fragments is in fact significant.
+We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.
+
 ##  {#R2-7}
@@
 > I would like to see at least a justification for why representing the difference figures are meaningful.
+
+<!--# We *could* add outlines to the difference plots indicating areas significantly different from 0, I think?  Gratia has a difference smooth plot--could look into that. -->
 
@@
 
-<!--# need to check the line to figure out the source of confusion, but we did test this with 84%CI overlap -->
+While we could not test for an interaction between habitat and lagged SPEI (i.e.
+whether the shape of the crossbasis smooth was *significantly* different between habitats), we did test for a main effect of habitat by looking for overlap in 84% confidence intervals of model intercepts (see also response to \@ref(R2-6)).
+However, to reduce confusion here, we have changed "significantly" to "substantially" so as not to imply statistical significance for all of "magnitude, direction and lag time".
 
@@
 > Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.
-
-<!--# this is not a plot that can have CIs.  The line doesn't imply statistical significance either.  Not sure how to respond -->
 

### 2021-08-26 doc/paper.Rmd
add 84%CI table (#78, #85)

@@
 
-```{r eval=FALSE, include=FALSE}
-#get 84% CI around intercepts
-withr::with_dir(here::here(), {
-  tar_load(c(s_cf, s_1ha))
-})
-source(here::here("R", "pull_ci.R"))
-pull_ci(s_cf)
-pull_ci(s_1ha)
-rm(s_cf, s_1ha)
-```
-
 *Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
@@
 negative growth in Figure \@ref(fig:eda)a).
-When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
+When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
 However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
 The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.
-
-```{r eval=FALSE, include=FALSE}
-#get 84% CI around intercepts
-withr::with_dir(here::here(), {
-  tar_load(c(g_cf, g_1ha))
-})
-pull_ci(g_cf, digits = 3)
-pull_ci(g_1ha, digits = 3)
-rm(g_cf, g_1ha)
-```
 
@@
 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
-Plants were significantly larger in continuous forest (no overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
+Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
 
@@
 
-```{r eval=FALSE, include=FALSE}
-#get 84% CI around intercepts
-withr::with_dir(here::here(), {
-  tar_load(c(f_cf, f_1ha))
-})
-pull_ci(f_cf, digits = 4)
-pull_ci(f_1ha, digits = 4)
-rm(f_cf, f_1ha)
-```
-
 ```{r include=FALSE}
@@
 Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (84% CIs of model intercepts did not overlap).
+`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
 The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
@@
 Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
+We also found that the delayed responses of populations in fragments can differ substantially in magnitude, direction, and lag time from those of populations in continuous forest.
 This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.
@@
 Data used in this study are available at the Dryad Digital Repository; R code for analyses and visualizations are archived at Zenodo [links included upon acceptance].
+
+# Tables
+
+```{r}
+ci_cap <- "Model intercepts and their 84% confidence intervals.
+Non-overlapping confidence intervals are bolded and can be interpreted as a significant difference in the vital rate between continuous forest and 1 ha fragments."
+```
+
+```{r tab.cap=ci_cap, label="intercept-table"}
+tar_read(intercept_table)
+```
 

### 2021-08-26 doc/supplemental.Rmd
add 84%CI table (#78, #85)

@@
       fig_caption: yes
-      # reference_docx: "template.docx" # Insert path for the DOCX file
+      reference_docx: "templates/template.docx" # Insert path for the DOCX file
 linkcolor: blue

### 2021-08-26 R/tables.R

@@ -1,0 +1,126 @@
+#' Get CI around parametric terms of gam
+#'
+#' @param model a gam
+#' @param conf.level confidence level. Default is 0.84
+#'
+#' @return formatted estimate [lower, upper]
+#'
+pull_ci <- function(model, conf.level = 0.84, digits = 3, type = c("response", "link")) {
+  linkinv <- model$family$linkinv
+  type <- match.arg(type)
+  intercept <- coef(model)[1]
+  ci <- confint.default(model, parm = "(Intercept)", level = conf.level)
+  if (type == "response") {
+    intercept <- linkinv(intercept)
+    ci <- linkinv(ci)
+  }
+  intercept <- round(intercept, digits)
+  ci <- round(ci, digits)
+  ci_string <- glue::glue("{intercept} [{ci[1]}, {ci[2]}]")
+  return(ci_string)
+}
+
+
+
+#' Make results table showing model intercepts ± confidence limits
+#' 
+#' Shows intercepts on a response scale with 84% confidence intervals.
+#'
+#' @param s_cf survival in CF model
+#' @param s_1ha survival in FF model
+#' @param g_cf size in CF model
+#' @param g_1ha size in FF model
+#' @param f_cf flowering in CF model
+#' @param f_1ha flowering in FF model
+#' 
+make_intercept_table <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
+  expand_grid(`vital rate` = c("survival", "size", "flowering"), 
+              habitat = c("CF", "1 ha")) %>% 
+    add_column(intercept = map_chr(list(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha), pull_ci)) %>% 
+    flextable() %>% 
+    autofit() %>% 
+    merge_v(j = 1) %>% 
+    valign(j = 1, valign = "top") %>% 
+    align(j = 3, align = "right", part = "all") %>% 
+    bold(j = 3, i = c(3:6)) %>% 
+    fix_border_issues()
+}
+
+#' Helper function for formatting EDF/DF
+#' 
+#' Degrees of freedom are integers and should be displayed with no decimal
+#' places while EDF can have decimal places.  Needed this helper to format the
+#' DF/EDF column correctly with `flextable`.
+#'
[…]

### 2021-08-26 _targets.R

@@ -111,6 +111,12 @@ tar_plan(
     f = f_covar_plot,
     model_data = model_data
   ), 
+  
+  # Tables
+  intercept_table = make_intercept_table(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
+  s_results = make_results_table(s_cf, s_1ha),
+  g_results = make_results_table(g_cf, g_1ha),
+  f_results = make_results_table(f_cf, f_1ha),
 
   # Main text
   tar_render(paper, "doc/paper.Rmd", deployment = "main"),

### 2021-08-26 packages.R

@@ -24,6 +24,8 @@ library(colorspace)
 library(readxl)
 library(Hmisc)
 library(qqplotr)
+library(flextable)
+library(scales)
 
 conflict_prefer("select", "dplyr")
 conflict_prefer("filter", "dplyr")
@@ -31,3 +33,4 @@ conflict_prefer("lag", "dplyr")
 conflict_prefer("slice", "dplyr")
 conflict_prefer("summarize", "dplyr")
 conflict_prefer("Lag", "tsModel")
+conflict_prefer("compose", "flextable")

### 2021-08-26 review/Response to reviews.Rmd
changed surv curve to percentage (#85)

@@
 
+Figure 2 shows the survivorship curve of plants tagged in 1998.
+These are exact counts and therefore do not have confidence intervals.
+They do not require statistics to interpret if they are different.
+We think the confusion may have come from the axis scale and label ("P(survived)") which we intended to be interpreted as "proportion of plants surviving" but may have been interpreted as "probability of survival".
+We have changed the scale to a percentage and the axis label to "% surviving".
+
 ##  {#R2-12}

### 2021-08-26 R/plot_eda.R

@@ -116,7 +116,7 @@ plot_eda_surv_cohort <- function(demog) {
     group_by(year, habitat) %>% 
     summarize(n = n()) %>%
     group_by(habitat) %>% 
-    mutate(p_surv = n/first(n),
+    mutate(p_surv = n/first(n)*100,
            date = ymd(paste0(year, "-02-01")),
            yearmonth = yearmonth(date))
   
@@ -130,7 +130,7 @@ plot_eda_surv_cohort <- function(demog) {
       date_minor_breaks = "1 month",
       expand = expansion(mult = 0.04)
     ) +
-    scale_y_continuous("P(survived)") +
+    scale_y_continuous("% surviving") +
     scale_color_manual(values = c("#E66101", "#5E3C99"),
                         aesthetics = c("color", "fill")) +
     guides(col = guide_legend(title = "Habitat"),

### 2021-08-26 review/Response to reviews.Rmd
add results table.  closes #78

@@
 
-<!--# include plots with raw data?  Would need to be supplemental probably, because messy. -->
+Because model selection was performed automatically by shrinkage using the `select = TRUE` option in the `gam()` function, we cannot show relative model fits (i.e., we did not perform model selection and can't show dAIC).
+However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term.
+We have replaced many of the parenthetical p-values reported in the results section with references to this table.
 
@@
 However, to reduce confusion here, we have changed "significantly" to "substantially" so as not to imply statistical significance for all of "magnitude, direction and lag time".
+See also our response to \@ref(R2-7).
 
@@
 
+<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values. Not sure what else to say about this-->
+
 ##  {#R2-13}
@@
 
-...see also our response to \@ref(R2-10)
+<!--# the magnitude *was* greater, interaction not tested -->...see also our response to \@ref(R2-10)
 

### 2021-08-26 doc/paper.Rmd
add results table.  closes #78

@@
 When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
-However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
-The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.
+However, survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
+The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1-ha fragments.
 
@@
 `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
-The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
+The observed disparity was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
 Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
@@
 
-Drought history had a significant ($p < 0.001$) effect on the survival, growth, and flowering of plants in both habitats.
+Drought history had a significant effect on the survival, growth, and flowering of plants in both habitats (Table \@ref(tab:results-table)).
 Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.
@@
 *Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
-However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
+However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)) <!--# this is no longer true.
+need to update -->.
 For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
@@
 
-*Flowering:* Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
+*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
 SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
@@
 The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
-We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flower again (CF: $p < 0.001$; 1-ha: $p = 0.004$).
-The random effect of plant ID on flowering probability was also significant (CF: $p < 0.001$; 1-ha: $p=0.003$), indicating significant individual-level variation in flowering probability.
+We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
+The random effect of plant ID on flowering probability was also significant (Table \@ref(tab:results-table)), indicating significant individual-level variation in flowering probability.
+<!--# individual level random effects need to be removed? -->
 
-Finally, with the exception of survival in 1-ha fragments ($p = 0.253$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.01$ for the random effect of plot).
+Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).
 
@@
 
-```{r tab.cap=ci_cap, label="intercept-table"}
-tar_read(intercept_table)
+```{r label="intercept-table"}
+tar_read(intercept_table) %>% set_caption(ci_cap, style = "Image Caption")
+```
+
+<br>
+
+```{r}
+result_cap <- 'Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H.
+acuminata* in continuous forest and 1 ha fragment plots.
+The adjusted R^2^ is reported as a measure of model fit.
+The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowered"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)").
+Degrees of freedom are reported for the parametric term, and estimated degrees of freedom (edf) are reproted for smooths.
+Larger values for edf indicate more complex smooths and when edf is zero the term is effectively dropped from the model.
+The test statistic reported is $\\chi^2$ for survival and flowering and $F$ for log(size~t+1~).'
+```
+
+<!--# add something about how p-values are appropriate for random effects?  see help files from mgcv for references -->
+
+```{r label="results-table"}
+tar_read(results_table) %>% set_caption(result_cap, style = "Image Caption")
 ```

### 2021-08-26 R/tables.R

@@ -34,6 +34,7 @@ pull_ci <- function(model, conf.level = 0.84, digits = 3, type = c("response", "
 #' @param f_1ha flowering in FF model
 #' 
 make_intercept_table <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
+  set_flextable_defaults(font.family = "Cambria", font.size = 10)
   expand_grid(`vital rate` = c("survival", "size", "flowering"), 
               habitat = c("CF", "1 ha")) %>% 
     add_column(intercept = map_chr(list(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha), pull_ci)) %>% 
@@ -69,7 +70,7 @@ int_or_round <- function(x, digits = 2) {
 }
 
 
-make_results_table <- function(cf_model, ff_model) {
+make_results_df <- function(cf_model, ff_model) {
   ff <- summary(ff_model)
   ff_param <- ff$pTerms.table
   ff_smooth <- ff$s.table
@@ -96,31 +97,41 @@ make_results_table <- function(cf_model, ff_model) {
   df <- 
     bind_rows("CF" = cf_table, "1 ha" = ff_table, .id = "habitat") %>% 
     mutate(`p-value` = scales::pvalue(`p-value`)) %>% 
-    mutate("(e)df" = int_or_round(`(e)df`, 2)) 
-  
-  ft <- 
-    flextable(df) %>% 
-    merge_v(j = c(1,2)) %>% 
-    valign(j = c(1,2), valign = "top") %>% 
-    align(j = 4, align = "right", part = "all") %>% 
-    padding(i = 4, padding.bottom = 12) %>% 
-    colformat_double(j = c(2,5), digits = 2) %>% 
-    fix_border_issues() %>% 
-    flextable::compose(j = 2, part = "header", value = as_paragraph("R", as_sup("2")))
-  
-  if(names(df)[5] == "Chi.sq") {
-    ft %>% flextable::compose(j = 5, part = "header", value = as_paragraph("χ", as_sup("2")))
-  } else {
-    ft
-  }
+    mutate("(e)df" = int_or_round(`(e)df`, 2)) %>% 
+    mutate(term = case_when(
+      term == "flwr_prev"          ~ "flowered",
+      term == "s(log_size_prev)"   ~ "s(log(size_t))",
+      term == "s(spei_history,L)"  ~ "s(SPEI, lag)",
+      TRUE                         ~ term
[…]

### 2021-08-26 _targets.R

@@ -114,9 +114,7 @@ tar_plan(
   
   # Tables
   intercept_table = make_intercept_table(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
-  s_results = make_results_table(s_cf, s_1ha),
-  g_results = make_results_table(g_cf, g_1ha),
-  f_results = make_results_table(f_cf, f_1ha),
+  results_table = make_results_table(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
 
   # Main text
   tar_render(paper, "doc/paper.Rmd", deployment = "main"),

### 2021-08-26 doc/paper.Rmd
move flextable code to paper.Rmd so easier to change to kable if needed.

@@
 ```{r}
+withr::with_dir(here::here(), {
+  targets::tar_load(c(
+    intercept_df,
+    results_df
+  ))
+})
+flextable::set_flextable_defaults(font.family = "Cambria", font.size = 10)
+```
+
+```{r}
 ci_cap <- "Model intercepts and their 84% confidence intervals.
@@
 ```{r label="intercept-table"}
-tar_read(intercept_table) %>% set_caption(ci_cap, style = "Image Caption")
+flextable(intercept_df) %>% 
+  autofit() %>% 
+  merge_v(j = 1) %>% 
+  valign(j = 1, valign = "top") %>% 
+  align(j = 3, align = "right", part = "all") %>% 
+  bold(j = 3, i = c(3:6)) %>% 
+  fix_border_issues() %>%
+  set_caption(ci_cap, style = "Image Caption")
 ```
@@
 ```{r label="results-table"}
-tar_read(results_table) %>% set_caption(result_cap, style = "Image Caption")
+flextable(results_df) %>% 
+  merge_v(j = c(2)) %>% 
+  valign(j = c(2), valign = "top") %>% 
+  align(j = c(5, 7), align = "right", part = "all") %>%
+  colformat_double(j = c(3,6), digits = 2) %>% 
+  fix_border_issues() %>% 
+  flextable::compose(j = 3, part = "header",
+                     value = as_paragraph("R", as_sup("2"))) %>% 
+  flextable::compose(j = 4, i = c(4,9,15,20,26,31),
+                     value = as_paragraph("s(log(size", as_sub("t"), "))")) %>%
+  flextable::compose(j = 1, i = 12, 
+                     value = as_paragraph("log(size", as_sub("t+1"), ")")) %>% 
+  autofit() %>% 
+  set_caption(result_cap, style = "Image Caption")
 ```

### 2021-08-26 R/tables.R

@@ -33,18 +33,10 @@ pull_ci <- function(model, conf.level = 0.84, digits = 3, type = c("response", "
 #' @param f_cf flowering in CF model
 #' @param f_1ha flowering in FF model
 #' 
-make_intercept_table <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
-  set_flextable_defaults(font.family = "Cambria", font.size = 10)
+make_intercept_df <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
   expand_grid(`vital rate` = c("survival", "size", "flowering"), 
               habitat = c("CF", "1 ha")) %>% 
-    add_column(intercept = map_chr(list(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha), pull_ci)) %>% 
-    flextable() %>% 
-    autofit() %>% 
-    merge_v(j = 1) %>% 
-    valign(j = 1, valign = "top") %>% 
-    align(j = 3, align = "right", part = "all") %>% 
-    bold(j = 3, i = c(3:6)) %>% 
-    fix_border_issues()
+    add_column(intercept = map_chr(list(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha), pull_ci))
 }
 
 #' Helper function for formatting EDF/DF
@@ -70,7 +62,7 @@ int_or_round <- function(x, digits = 2) {
 }
 
 
-make_results_df <- function(cf_model, ff_model) {
+make_results <- function(cf_model, ff_model) {
   ff <- summary(ff_model)
   ff_param <- ff$pTerms.table
   ff_smooth <- ff$s.table
@@ -108,30 +100,18 @@ make_results_df <- function(cf_model, ff_model) {
   df
 }
 
-make_results_table <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
+make_results_df <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
   cf_list <- list(s_cf, g_cf, f_cf)
   ff_list <- list(s_1ha, g_1ha, f_1ha)
   
   df_list <- 
-    purrr::map2(cf_list, ff_list, ~make_results_df(.x, .y))
+    purrr::map2(cf_list, ff_list, ~make_results(.x, .y))
   
   vrates <- list("survival", "size_next", "flowering")
   
-  df <- map2_df(vrates, df_list, ~bind_rows(tibble(`vital rate` = .x), .y))
[…]

### 2021-08-26 _targets.R

@@ -113,8 +113,8 @@ tar_plan(
   ), 
   
   # Tables
-  intercept_table = make_intercept_table(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
-  results_table = make_results_table(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
+  intercept_df = make_intercept_df(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
+  results_df = make_results_df(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha),
 
   # Main text
   tar_render(paper, "doc/paper.Rmd", deployment = "main"),

### 2021-08-26 review/Response to reviews.Rmd
remove individual-level random effects (closes #96)

@@
 ---
+
+Cascading effects of increasing `k`:
+
+-   Changed shape of crossbasis functions (edf changes, now not always more complex in fragments)
+
+-   Changes in which lags are significant
+
+-   Individual-level random effects impossible to fit due to computational limitations.
+    Using a sub-sample, checked with and without invididual-level random effects.
+    Changes r2, but very minimal changes to shape of smooths, edfs, and significance of smooths.
+    Dropped from flowering model.
 

### 2021-08-26 doc/paper.Rmd
remove individual-level random effects (closes #96)

@@
 We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
-Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
-Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
 A random effect of plant ID was not included in survival models since each plant only dies once.
+Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
+Preliminary modeling with data subsets showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.
+
 The crossbasis function, $s_2(\cdot)$ can also be written:
@@
 We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
-The random effect of plant ID on flowering probability was also significant (Table \@ref(tab:results-table)), indicating significant individual-level variation in flowering probability.
-<!--# individual level random effects need to be removed? -->
-
 Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).
@@
 While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
-In fact, flowering in one year is associated with increased reproduction and growth in the next <!--# this may change once individual-level random effects are incorporated -->.
+In fact, flowering in one year is associated with increased reproduction and growth in the next.
 

### 2021-08-27 review/Response to reviews.Rmd
track word docs in review/

@@
 
-The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored.
+(Something I'm considering adding to the discussion):
+
+"The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored.
 @tellerLinkingDemographyDrivers2016 used simulations to show that 20--25 years of data were required to detect lagged effects using their method, but this cutoff depends on the effect size and the number of climate extremes in the dataset.
-Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required, but this is unpublished currently <!--# cite Aldo Compagnoni's ESA talk -->.
+Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required (Aldo Compagnoni's ESA talk)."
 
@@
 > Bring it back into the results?
+
+... see also response to \@ref(R3-8).
 
@@
 
-<!--# need to double-check and revise --> The difference in flowering probability between cointinuous forests and fragments is in fact significant.
+<!--# need to double-check and revise -->
+
+The difference in flowering probability between cointinuous forests and fragments is in fact significant.
 We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.
@@
 
-<!--# the magnitude *was* greater, interaction not tested -->...see also our response to \@ref(R2-10)
+<!--# the magnitude *was* greater, interaction not tested -->
+
+...see also our response to \@ref(R2-10)
 

### 2021-08-27 doc/paper.Rmd
update results section with new results.  Closes #72

@@
 Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
-Preliminary modeling with data subsets showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.
+Fitting models with a data subset with and without individual-level random effects showed that excluding individual-level random effects reduced model R^2^, but did not affect the shape or statistical significance of the smooths.
 
@@
 
-## Survival, growth, and flowering in continuous forest vs. fragments
+## Survival, size, and flowering in continuous forest vs. fragments
 
@@
 
-*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
+*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year (93% survival in both continuous forest and forest fragments).
 This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e.
 negative growth in Figure \@ref(fig:eda)a).
-When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
-However, survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
-The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1-ha fragments.
+There was no significant difference in survival between continuous forest and 1 ha fragments (Table \@ref(tab:intercept-table)).
+Survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
+The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1 ha fragments (although 95% confidence intervals overlap at all sizes).
 
@@
 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
-Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).
+Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time.
 
-Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
-As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
-Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats (Figure \@ref(fig:size-plot)a).
+Mean plant size dropped dramatically in 2003 in both habitat types (negative fold-change in Figure \@ref(fig:eda)a), corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
+As with survival, size in year t was a significant predictor of size in year t+1 (Table \@ref(tab:results-table)).
+Although plants were significantly larger in continuous forest compared to fragments (Table \@ref(tab:intercept-table)), the effect of size in year t on size in year t+1 is nearly identical in the two habitats (Figure \@ref(fig:size-plot)a).
 
@@
 *Flowering:* The overall proportion of plants that flowered was very low.
-Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs.
-`r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
-The observed disparity was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
+The proportion of plants flowering was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]*100`% vs. `r summary_stats$flwr_mean[1]*100`%, respectively).
+The intercepts for the continuous forest and 1 ha fragment models were significantly different, indicating a main effect of habitat on flowering probability (Table \@ref(tab:intercept-table)).
+The observed disparity in proportion of flowering plants was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
 Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
@@
 
-*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 13 months.
+*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 0--16 months.
 The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
-Wet conditions in the preceding 23 months (i.e., SPEI \> 1) also had a significant negative effect on survival in fragments (Figure \@ref(fig:surv-spei)b).
-In contrast, the effect of recent SPEI in continuous forest was weaker, with only the 6 months preceding a census having a significant effect on survival (Figure \@ref(fig:surv-spei)a).
-These short-term effects of SPEI on survival were also unidirectional---the probability of survival declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
-In contrast, the most pronounced negative effects of SPEI on the survival of plants in continuous forest were at lag times of 15--20 months and 32--36 months.
-Drought 15--20 months prior to a census (i.e.
-two dry seasons prior to a census) was significantly associated with reduced survival, while high precipitation (i.e., SPEI \> 1) was significantly associated with higher survival.
+In contrast, only recent conditions (0--4 month delay) has a significant effect on survival in continuous forest (Figure \@ref(fig:surv-spei)a).
+The short-term effects of SPEI on survival in continuous forest were also unidirectional---the probability of survival was highest in wet conditions and declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
 Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
@@
 This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
-However, under extreme conditions, survival probability is higher in continuous forest by up to 0.025.
+However, under extreme conditions, survival probability is higher in continuous forest by up to 0.042.
 
-*Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
-However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)) <!--# this is no longer true.
-need to update -->.
-For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
-However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
-In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a more pronounced response in 1-ha fragments.
+*Size:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
+Under all conditions, plant size is greater in continuous forest (Figure \@ref(fig:growth-spei)c).
+Decreasing SPEI at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats.
+In continuous forest, but not 1 ha fragments, SPEI at lags of 22--25 months had a significant effect on plant size in the opposite direction, with wet conditions at that lag time resulting in larger predicted sizes.
+At longer lags of 26--36 months positive effects of drought on plant size are predicted for both continuous forest and 1 ha fragments by our models, although the area that reaches statistical significance is larger for continuous forest.
 
-*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
-SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
+*Flowering:* Overall, the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
+Recent and past SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
 This led to some important inter-habitat differences in plant responses to prior droughts.
-In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
-However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
-In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
-The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
-We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
+In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1) and droughts two dry seasons prior (lags 15--20) increased the probability of flowering.
+The shape of the crossbasis smooth for 1 ha fragments suggested that moderate drought (-1.5 \<SPEI\< -1) in the previous 0--18 months and at lags of 28--36 months significantly increased flowering probability slightly (the highest probability estimated is 0.004; Figure \@ref(fig:flwr-spei)b).
+The effects of drought on flowering probability appeared stronger in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
+We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were significantly more likely to be larger and flower again.
 Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).
@@
 Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->.
-Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
+<!--# possibly good place to mention limitations of interactions -->Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
 Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

### 2021-08-27 review/Response to reviews.Rmd
add reference to fig 2

@@
     Dropped from flowering model.
+
+For coverletter:
+
+we identified \_\_\_ main suggestions
+
+1.  more in depth explanation of model (both technical and for broad audience)
+2.
+something about changing declaritive statements about habitat:SPEI,Lag interaction.
+Added caveats about limitations of DLNM approach.
+Expanded on this in future directions to draw attention to the importance of including interactions for demography.
+3.  R\_ suggested increasing k, important suggestion. While it didn't alter the conclusions, it did require considerable work and further evaluation of results. new appendix, review rewrite
 
@@
 
-On the contrary, averaging weather conditions over an entire year likely obscures ways plants perceive environmental variation.
+Counter-intuitively, averaging weather conditions over an entire year may obscure ways plants perceive environmental variation.
 For example, a drought in the wet season may have a different effect than a drought in the dry season.
@@
 
-... see also response to \@ref(R3-8).
+We have added a sentence about this figure in the results.
 
@@
 
-<!--# We *could* add outlines to the difference plots indicating areas significantly different from 0, I think?  Gratia has a difference smooth plot--could look into that. -->
+Because we were not able to test for a significant interaction between habitat and lagged SPEI, we did not intend to suggest that our results implied statistically significant differences in the shape of the crossbasis functions between fragments and continuous forest.
+We have made changes throughout the manuscript that clarify this and remove some of the stronger declarative language about differences in the effects of SPEI between continuous forest and forest fragments.
+<!--# still need to do this -->
 
@@
 > Line 323-325: I don't know why the growth of plants in fragments would affect the size of plant in CF.
+
+This sentence meant to describe the difference between CF and fragments, not the absolute size in fragments.
+It has been re-worded to clarify.
 
@@
 
-<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values. Not sure what else to say about this-->
+<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values.
+Not sure what else to say about this--> See also our response to \@ref(R2-7).
 
@@
 
-...see also our response to \@ref(R2-10)
+...see also our response to \@ref(R2-7)
 
@@
 
+see also our response to \@ref(R2-7).
+
 ##  {#R2-23}
@@
 That's correct.
-This line meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
+This sentence meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
 We've edited this sentence to read: "The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g.
@@
 
-...see also our response to \@ref(R1-figs)
+See our response to \@ref(R1-figs).
 

### 2021-08-27 doc/paper.Rmd
add reference to fig 2

@@
 There was no significant difference in survival between continuous forest and 1 ha fragments (Table \@ref(tab:intercept-table)).
+Although differences in survival probability are small and not statistically significant, these slight differences can be seen to compound over time in the survivorship of plants labeled in the first year of the surveys (Figure \@ref(fig:surv-curve)).
 Survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
@@
 This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
-However, under extreme conditions, survival probability is higher in continuous forest by up to 0.042.
+However, under extreme conditions, survival probability is higher in continuous forest compared to fragments by up to 0.042.
 

### 2021-08-27 R/plot_eda.R
refactor plots and bring axis ticks back

@@ -3,12 +3,12 @@
 #' The range is the first date of the survey minus the maximum lag for the
 #' models to the last date of the survey data.
 #'
-#' @param demog_post 
+#' @param data 
 #' @param maxlag 
 #'
-get_plot_daterange <- function(demog_post, maxlag) {
+get_plot_daterange <- function(data, maxlag) {
   dates <-
-    range(demog_post$year) %>%
+    range(data$year) %>%
     paste0(., "-02-01") %>%
     ymd()
   dates[1] <- dates[1] - months(maxlag)
@@ -31,7 +31,7 @@ plot_eda_spei <- function(xa, date_lims) {
   spei <-
     xa %>%
     select(latlon, yearmonth, spei) %>%
-    filter(!is.na(spei)) %>% 
+    dplyr::filter(!is.na(spei)) %>% 
     ggplot(aes(x = yearmonth, y = spei)) + 
     geom_line(aes(group = latlon), alpha = 0.2) +
     stat_summary(geom = "line", fun = "mean") +
@@ -45,18 +45,6 @@ plot_eda_spei <- function(xa, date_lims) {
   
 }
 
-#' Create data frame for plotting demographic data
-#'
-#' @param demog_post demography data filtered to only have post-seedlings
-#'
-eda_plot_df <- function(demog_post) {
-  demog_plotdf <-  
-    demog_post %>% 
-    filter(habitat !="10-ha") %>% 
-    mutate(date = paste(year, "-02-01") %>% ymd(),
-           yearmonth = yearmonth(date))
-  demog_plotdf
-}
 
 #' Survival time series plot
 #' 
@@ -65,33 +53,34 @@ eda_plot_df <- function(demog_post) {
 #' @param demog_post demography data
 #' @param dates_lims x-axis limits
 #'
-plot_eda_surv_ts <- function(demog_post, date_lims) {
+plot_eda_surv_ts <- function(data, date_lims) {
   
   date_breaks <- seq(date_lims[1], date_lims[2], by = "year")
   
-  demog_plotdf <- eda_plot_df(demog_post)
-  
-  survival <-
-    demog_plotdf %>%  
-    # filter(shts_prev >= 4) %>%
-    ggplot(aes(x = yearmonth, y = surv, color = habitat, linetype = habitat)) +
-    stat_summary(geom = "line", fun = "mean", fun.args = list(na.rm = TRUE)) +
-    # stat_summary(geom = "pointrange", fatten = 1) + #not sure plot-level SD makes sense
[…]

### 2021-08-27 R/plot_smooths.R

@@ -57,8 +57,7 @@ make_size_plot <- function(s, g, f, model_data) {
     s /
     f & 
     theme(axis.title.x = element_blank(),
-          axis.text.x = element_blank(),
-          axis.ticks.x = element_blank())
+          axis.text.x = element_blank())
   
   top /
     d +

### 2021-08-27 review/Response to reviews.Rmd
Close #85

@@
 > Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
-> I would like to see at least a justification for why representing the difference figures are meaningful.
 
@@
 We have made changes throughout the manuscript that clarify this and remove some of the stronger declarative language about differences in the effects of SPEI between continuous forest and forest fragments.
-<!--# still need to do this -->
 
-##  {#R2-8}
+> I would like to see at least a justification for why representing the difference figures are meaningful.
+
+We believe the difference figures (panel c in figure 4,5, and 6) are helpful in showing combinations of SPEI and lag time where the marginal effects on vital rates are most divergent between continuous forest and fragments.
+For example, at a lag of 0, wet conditions have a positive effect on survival in continuous forest and a negative effect on survival in fragments, leading to a large predicted difference in survival between habitats under those conditions.
 
@@
 
-<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values.
-Not sure what else to say about this--> See also our response to \@ref(R2-7).
+See our response to \@ref(R2-7).
 
@@
 
-<!--# the magnitude *was* greater, interaction not tested -->
-
-...see also our response to \@ref(R2-7)
+See our response to \@ref(R2-7).
 
@@
 
-see also our response to \@ref(R2-7).
+See our response to \@ref(R2-7).
 

### 2021-08-27 doc/paper.Rmd
Close #85

@@
 Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-We also found that the delayed responses of populations in fragments can differ substantially in magnitude, direction, and lag time from those of populations in continuous forest.
-This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.
+We also found different magnitudes, directions, and lag times of climate effects in fragments and continuous forests.
+This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] may be important in this system in a way far more complex than previously thought.
 
@@
 
-It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (Figures \@ref(fig:surv-spei), \@ref(fig:growth-spei)).
-Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
+It is important to emphasize, however, that the range of the estimated response to SPEI for survival and growth was greater in fragments compared with continuous forest (Figures \@ref(fig:surv-spei), \@ref(fig:growth-spei)).
+This suggests that extremes in SPEI may be more detrimental in forest fragments compared with continuous forest.
 While intact forest and its canopy buffer populations from climatic extremes, populations in fragments---especially near edges with high contrast matrix---likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.
+We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying the observed reduced plant growth and survival in forest fragments.
 
@@
 While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
-Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
-For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types.
-However, the magnitude of these declines was far greater in forest fragments.
-Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
-In contrast, variation in SPEI had far stronger effects on the probability of flowering in continuous forest than fragments.
+Our results suggest that they may, with habitat-specific differences in how environmental conditions influenced future vital rates.
+For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in forest fragments while extreme wet conditions significantly increased survival in continuous forest.
+Similarly, the marginal effects of SPEI on plant size were greater in fragments, suggesting a more pronounced effect of extreme climates in fragments.
+In contrast, variation in SPEI corresponded to a greater range of flowering probabilities in continuous forest than fragments.
 These results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

### 2021-08-27 doc/paper.Rmd
fix table

@@
 flextable(results_df) %>% 
-  merge_v(j = c(2)) %>% 
-  valign(j = c(2), valign = "top") %>% 
+  merge_v(j = c(1,2,3)) %>% 
+  valign(j = c(1,2,3), valign = "top") %>% 
   align(j = c(5, 7), align = "right", part = "all") %>%
@@
   flextable::compose(j = 3, part = "header",
-                     value = as_paragraph("R", as_sup("2"))) %>% 
-  flextable::compose(j = 4, i = c(4,9,15,20,26,31),
+                     value = as_paragraph("model R", as_sup("2"))) %>% 
+  flextable::compose(j = 4, i = c(2,6,10,14,18,22),
                      value = as_paragraph("s(log(size", as_sub("t"), "))")) %>%
-  flextable::compose(j = 1, i = 12, 
+  flextable::compose(j = 1, i = 9, 
                      value = as_paragraph("log(size", as_sub("t+1"), ")")) %>% 

### 2021-08-27 R/tables.R

@@ -74,18 +74,24 @@ make_results <- function(cf_model, ff_model) {
   cf_r2 <- cf$r.sq
   
   ff_table <- bind_rows(
-    ff_param %>% as_tibble(rownames = "term") %>% rename(`(e)df` = df),
-    ff_smooth %>% as_tibble(rownames = "term") %>%  select(- Ref.df) %>% rename(`(e)df` = edf)
-  ) 
-  ff_table <- bind_rows(tibble(R2 = ff_r2), ff_table)
+    ff_param %>% as_tibble(rownames = "term") %>% 
+      rename(`(e)df` = df),
+    ff_smooth %>% as_tibble(rownames = "term") %>%
+      dplyr::select(- Ref.df) %>% 
+      rename(`(e)df` = edf)
+  ) %>% 
+    add_column(R2 = ff_r2, .before = "term")
 
   
   cf_table <- bind_rows(
-    cf_param %>% as_tibble(rownames = "term") %>% rename(`(e)df` = df),
-    cf_smooth %>% as_tibble(rownames = "term") %>%  select(- Ref.df) %>% rename(`(e)df` = edf)
-  ) 
-  cf_table <- bind_rows(tibble(R2 = cf_r2), cf_table)
-  
+    cf_param %>% as_tibble(rownames = "term") %>%
+      rename(`(e)df` = df),
+    cf_smooth %>% as_tibble(rownames = "term") %>% 
+      dplyr::select(- Ref.df) %>%
+      rename(`(e)df` = edf)
+  ) %>% 
+    add_column(R2 = cf_r2, .before = "term")
+
   df <- 
     bind_rows("CF" = cf_table, "1 ha" = ff_table, .id = "habitat") %>% 
     mutate(`p-value` = scales::pvalue(`p-value`)) %>% 
@@ -104,12 +110,9 @@ make_results_df <- function(s_cf, s_1ha, g_cf, g_1ha, f_cf, f_1ha) {
   cf_list <- list(s_cf, g_cf, f_cf)
   ff_list <- list(s_1ha, g_1ha, f_1ha)
   
-  df_list <- 
-    purrr::map2(cf_list, ff_list, ~make_results(.x, .y))
-  
-  vrates <- list("survival", "size_next", "flowering")
-  
- map2_df(vrates, df_list, ~bind_rows(tibble(`vital rate` = .x), .y))
+  purrr::map2(cf_list, ff_list, ~make_results(.x, .y)) %>% 
[…]

### 2021-08-27 doc/validate_models.Rmd
finish model validation supplement

@@
 ```
-
-**TODO**
-
-- check that # of knots are adequate
-- check diagnostics for individual-level random effects.  Doesn't look good, but not sure how to interpret or what to do.
 
@@
 
-Basis dimension checking with `gam.check()` doesn't appear to work for `dlnm` crossbasis smooths.
-Instead I'll use a method described in the help file `?choose.k` to check for adequate knots.
-Unfortunately, `bs = "cs"` doesn't work with `dlnm`, so I'll use `select = TRUE` instead to reduce chance of overfitting.
+`k` is adequate for `log_size_prev`.
+Unfortunately `k.check()` doesn't work for data with matrices, so I'll use an alternative approach detailed in `?mgcv::choose.k` for the crossbasis smooth.
+Briefly, it involves looking for pattern in residuals by extracting residuals and fitting a GAM to them with the smooth of interest with an increased value for `k`.
+If there is no unexplained pattern in residuals, that smooth should have an edf near zero.
 
@@
 ```
-
-The crossbasis smooth possibly needs more knots.
-
-```{r}
-# plot_lag_slice(s_1ha, "spei_history", lag =  0) + labs(title = "surv, 1ha")
-```
-
-Here you can see how the CIs hardly overlap 0, but it could be an artifact of the line not being allowed to be wiggly enough.
-
-### Summary
-
-```{r eval=TRUE}
-summary(s_1ha)
-```
-
-```{r eval=TRUE}
-draw(s_1ha)
-```
+Adequate k for crossbasis.
 
@@
 
+Adequate knots for `log_size_prev`
+
 ```{r}
-# looking for near zero edf
 check_res_edf(s_cf)
 ```
-
-### Summary
-
-```{r}
-summary(s_cf)
-```
-
-```{r}
-draw(s_cf)
-```
+Adequate knots for crossbasis smooth.
 
@@
 
-Here I use a random sub-sample to check that differences between continuous forest and fragments are not purely due to sample size differences, particularly differences in the complexity of the crossbasis smooth.
+Here I use a random sub-sample to check that differences between continuous forest and fragments are not purely due to sample size differences, particularly differences in the complexity and shape of the crossbasis smooth.
 
 ```{r}
-summary(s_1ha)$edf[3]
-summary(s_cf)$edf[3]
-summary(s_cf_sub)$edf[3]
+summary(s_1ha)$edf[2]
+summary(s_cf)$edf[2]
+summary(s_cf_sub)$edf[2]
 ```
 
+The crossbasis is much smoother
+
 ```{r}
-draw(s_cf_sub)
+draw(s_cf, select = "s(spei_history,L)", n_contour = 5)
+draw(s_cf_sub, select = "s(spei_history,L)", n_contour = 5)
 ```
-Surface still looks different in a similar way though.
 
@@
 
+Adequate k for `log_size_prev`.
+
 ```{r}
@@
 
-### Summary
-
-```{r}
-summary(g_1ha)
-```
-
-```{r}
-draw(g_1ha)
-```
+Adequate k for crossbasis.
 
@@
 
+Adequate k for `log_size_prev`
+
 ```{r}
@@
 
-```{r}
-# plot_lag_slice(g_1ha, "spei_history", lag = 33)
-```
+Adequate k for crossbasis smooth.
 
-hmm... CIs are **super** narrow here, but effect size is **tiny**
-
-### Summary
-
-```{r}
-summary(g_cf)
-```
-
-```{r}
-draw(g_cf)
-```
-
-## Continuous Forest subsample
+## Effect of sample size
 
@@
 ```{r}
-summary(g_1ha)$edf[3]
-summary(g_cf)$edf[3]
-summary(g_cf_sub)$edf[3]
+summary(g_1ha)$edf[2]
+summary(g_cf)$edf[2]
+summary(g_cf_sub)$edf[2]
 ```
-edf of subsample is similar to full dataset, despite differences in sample size.
+
+edf is similar for the sub-sample
 
 ```{r}
-draw(g_cf_sub)
+draw(g_cf, select = "s(spei_history,L)", n_contour = 5)
+draw(g_cf_sub, select = "s(spei_history,L)", n_contour = 5)
 ```
-Crossbasis surface looks nearly identical.
+
+The shape of the smooth is also similar.
 
@@
 
-### Summary
-
-```{r}
-summary(f_1ha)
-```
-
-```{r}
-draw(f_1ha)
-```
+Adequate k for crossbasis smooth.
 
@@
 ```
+Adequate k for crossbasis.
 
-### Summary
+## Effect of sample size
+
+To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
 
 ```{r}
-summary(f_cf)
+summary(f_1ha)$edf[2]
+summary(f_cf)$edf[2]
+summary(f_cf_sub)$edf[2]
 ```
 
+Similar edf for sub-sample
+
 ```{r}
-draw(f_cf)
+draw(f_cf, select = "s(spei_history,L)", n_contour = 5)
+draw(f_cf_sub, select = "s(spei_history,L)", n_contour = 5)
 ```
 
-## Continuous Forest subsample
-
-To check that differences are not purely due to sample size differences, particularly that lower edf in continuous forests is due to higher sample size.
-For flowering, edf is actually slightly higher in CF with a larger sample size.
-With subsample, edf are more similar.
-
-```{r}
-summary(f_1ha)$edf[3]
-summary(f_cf)$edf[3]
-summary(f_cf_sub)$edf[3]
-```
-
-```{r}
-draw(f_cf_sub)
-```
-Crossbasis surface looks extremely similar.
+Similar, but essentially linear in SPEI dimension.
 

### 2021-08-27 R/subset_cf.R

@@ -1,7 +1,7 @@
 
 subset_cf <- function(model_data_cf, n_plants = 1375) {
-  set.seed(123)
+  set.seed(342)
   plant_ids <- unique(model_data_cf$ha_id_number)
   plant_sample <- sample(plant_ids, n_plants)
-  model_data_cf %>% filter(ha_id_number %in% plant_sample)
+  model_data_cf %>% dplyr::filter(ha_id_number %in% plant_sample)
 }

### 2021-08-27 _targets.R

@@ -57,9 +57,9 @@ tar_plan(
   
   # Validate and summarize results
   ### Check for edf differences due to sample size
-  tar_target(g_cf_sub, fit_growth(model_data_cf_sub)),
-  tar_target(f_cf_sub, fit_flwr(model_data_cf_sub)),
-  tar_target(s_cf_sub, fit_surv(model_data_cf_sub)),
+  tar_target(g_cf_sub, fit_growth(model_data_cf_sub, k = c(25,5,15))),
+  tar_target(f_cf_sub, fit_flwr(model_data_cf_sub, k = c(10,15,15))),
+  tar_target(s_cf_sub, fit_surv(model_data_cf_sub, k = c(10,15,15))),
   tar_render(validate_models, "doc/validate_models.Rmd", deployment = "main"),
 
   # Descriptive / Exploratory Data Analysis Figures

### 2021-08-30 _targets.R
re-organization

@@ -14,13 +14,14 @@ options(
 
 ## Load your packages, e.g. library(targets).
 source("./packages.R")
+source("./conflicts.R")
 
 ## Load your R files
 lapply(list.files("./R", full.names = TRUE), source) 
 
 ## Set options
 options(tidyverse.quiet = TRUE)
-# tar_option_set(deployment = "main") #run targets locally by default
+tar_option_set()
 
 ## tar_plan supports drake-style targets and also tar_target()
 tar_plan(
@@ -28,7 +29,7 @@ tar_plan(
   maxlag = 36,
   tar_target(xa_file, here("data", "xavier_daily_0.25x0.25.csv"), format = "file"),
   xa_raw  = read_csv(xa_file),
-  xa_spei = calc_spei_xa(xa_raw),
+  xa_spei = calc_spei_xa(xa_raw, scale = 3),
   xa_lag  = lag_spei(xa_spei, maxlag),
   
   # Prep demographic data
@@ -45,7 +46,7 @@ tar_plan(
   tar_render(validate_data, "doc/validate_data.Rmd", deployment = "main"),
   
   # Select number of knots for GAMs
-  tar_render(knot_choice, "doc/knot_choice.Rmd", deployment = "main"),
+  # tar_render(knot_choice, "doc/knot_choice.Rmd", deployment = "main"),
   
   # Fit demographic models
   tar_target(s_cf,  fit_surv(model_data_cf, k = c(10,15,15))),

### 2021-08-30 conflicts.R

@@ -1,0 +1,9 @@
+library(conflicted)
+
+conflict_prefer("select", "dplyr")
+conflict_prefer("filter", "dplyr")
+conflict_prefer("lag", "dplyr")
+conflict_prefer("slice", "dplyr")
+conflict_prefer("summarize", "dplyr")
+conflict_prefer("Lag", "tsModel")
+conflict_prefer("compose", "flextable")

### 2021-08-30 packages.R

@@ -1,5 +1,4 @@
 ## library() calls go here
-library(conflicted)
 library(dotenv)
 library(targets)
 library(tarchetypes)
@@ -26,11 +25,3 @@ library(Hmisc)
 library(qqplotr)
 library(flextable)
 library(scales)
-
-conflict_prefer("select", "dplyr")
-conflict_prefer("filter", "dplyr")
-conflict_prefer("lag", "dplyr")
-conflict_prefer("slice", "dplyr")
-conflict_prefer("summarize", "dplyr")
-conflict_prefer("Lag", "tsModel")
-conflict_prefer("compose", "flextable")

### 2021-08-30 review/Response to reviews.Rmd
responses to line-specific comments (closes #82)

@@
 
-@brunaHabitatFragmentationPlant2009 discusses the lack of knowledge of what demographic vital rates (e.g. survival, reproduction, recruitment) contribute most to population declines in fragments.
-We have edited this line to clarify: " <!--# edit abstract? -->
+The reference, @brunaHabitatFragmentationPlant2009, discusses the lack of knowledge of what demographic vital rates (e.g.
+survival, reproduction, recruitment) in which life stages (e.g.
+small or large individuals) contribute most to population declines in fragments.
+We have re-ordered this sentence to emphasize the hypothesis that fragmentation alters survival, growth, or reproduction with the caveat that few studies have determined which of these vital rate changes are driving extinctions.
 
@@
 > While this statement is true, there are also years where flowering is lower in CF (e.g., 2002 similar values as following the drought in 2003 in fragments).
+
+We thank the reviewer for pointing this out.
+In fact, with updated data, 2006 actually shows the largest disparity in proportion of large plants flowering with a greater proportion flowering in 1 ha fragments.
+However, the total number of flowering plants in fragments was always lower due to 1) lower plant density in fragments and 2) fewer large reproductive-sized plants in fragments.
+We have clarified this in the revisions.
+
 > I feel that the reporting of the results is a bit biased towards concluding that plants do worse in fragments compared to CF, especially since none of the result reported so far are significant.
 
-<!--# need to double-check and revise -->
-
-The difference in flowering probability between cointinuous forests and fragments is in fact significant.
-We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.
+We have added Table 1 to clarify what statistically significant results we found between habitats.
+The table shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat.
+The difference in flowering probability between continuous forests and fragments is in fact significant, although the difference is slight.
 
@@
 > Surprisingly they fall not consistently in the dry or wet season.
+
+We see how this could be confusing and have taken the reviewers suggestion to include the 32--36 month lag when describing lags that are significant for each habitat.
 
@@
 
+<!--# Emilio: I feel like I don't have a great response to this one.
+Is there some edit to the manuscript that would clarify?
+I don't think "let the model do the smoothing" accurately explains what would happen if we switched to SPEI_1 -->\
+\
+The choice of scale for the SPEI calculation is indeed difficult, and contributes to the definition of drought (i.e. is drought defined as dry conditions over the previous 1, 2, 3 or more months?)
+. A thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper
+.
+
 ##  {#R3-3}

### 2021-08-30 doc/paper.Rmd
responses to line-specific comments (closes #82)

@@
 This fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000].
-Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in tropical forest fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] drive declines in plant reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998].
+It is often hypothesized that the dramatically altered environmental conditions in tropical forest fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] drive declines in plant reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998], although the demographic mechanisms responsible for these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009].
 Despite the prevalence of this hypothesis [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions in fragments remains scarce.
@@
 Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
-The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs.
-17% in 1-ha fragments (Figure \@ref(fig:eda)c).
+The most striking differences between habitat types coincided with relatively extreme weather conditions.
+In 2003, during the onset of a severe drought, the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1 ha fragments.
+In 2006, following a wet spring and winter (SPEI \> 1), the trend was reversed when only 14% of reproductive-sized plants flowered in continuous forest and 26% flowered in 1 ha fragments (Figure \@ref(fig:eda)c).
+It is important to keep in mind that 1 ha fragment plots have a lower density of plants on average and fewer reproductive-sized plants compared to continuous forest plots, so total numbers of flowering plants was always lower in fragments.
 
@@
 model_data %>%
-  filter(size >= repro_size) %>% 
+  # filter(size >= repro_size) %>% 
   group_by(habitat, year) %>%
-  summarize(prop_flwr = sum(flwr, na.rm = TRUE)/n()) %>% 
+  summarize(total_flwr = sum(flwr, na.rm = TRUE),
+            prop_flwr = sum(flwr, na.rm = TRUE)/n()) %>% 
     arrange(year) %>% 
@@
 The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
-In contrast, only recent conditions (0--4 month delay) has a significant effect on survival in continuous forest (Figure \@ref(fig:surv-spei)a).
+Additionally, a positive, lagged effect of high SPEI at 32--36 months lag reached statistical significance.
+There was less of a delayed effect of SPEI on survival in continuous forest with only the preceding 0--4 months and 32--36 months reaching statistical significance (Figure \@ref(fig:surv-spei)a).
 The short-term effects of SPEI on survival in continuous forest were also unidirectional---the probability of survival was highest in wet conditions and declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
-Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
+Plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
 It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.

### 2021-08-30 doc/paper.Rmd
adresses #76

@@
 
-Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], and that the effects of climate change could be more pronounced when habitat is fragmented [@oliver2015; @holyoak2016].
-<!--# explain why R1.2 --> While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
+Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015].
+While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021].
 Climate models predict increases in the areas of the Amazon effected by severe drought and unusual wetness by the year 2100 [@duffyProjectionsFutureMeteorological2015].
@@
 This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].
+
+The simultaneous pressures of climate change and habitat fragmentation may also result in worse than additive impacts on demography [@oliver2015; @holyoak2016].
+For example, if fragments have a reduced capacity to buffer changes in microclimate [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999] or if fragment connectivity is climate dependent [@honnayPossibleEffectsHabitat2002], populations experiencing habitat fragmentation will fare worse under climate change.
 

### 2021-08-30 review/Response to reviews.Rmd
adresses #76
Working on #86

@@
 > L466 -- 467 offer a new hypothesis, but briefly revisiting the 'old' evidence/hypotheses also seems important.
+
+At the reviewer's suggestion, we have added a short paragraph to the introduction explaining why the pressures of habitat fragmentation and climate change might interact (i.e. be more than additive).
 
@@
 > (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.
+
+We are a little hesitant to speculate too much here, as not much is known about the optimal microclimate conditions for flowering in *H.
+acuminata*, but we have revised this paragraph (see also our response to @ref(R2-17).
+It now includes the following possible explanation: *"Drought conditions could be favorable for H.
+accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts[@pauCloudsTemperatureDrive2013],* *and the effects could be delayed due to the development time of inflorescences."*
 
@@
 
-see our response to \@ref(R1-8).
+We intended this paragraph to focus on one possible explanation for delayed effects---delays due to development times.
+We believe that the bet-hedging argument was confusing for two reasons: 1) it assumed that drought (as we measured it with SPEI) is detrimental to *H.
+accuminata*, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
+We have therefore revised the example in this paragraph to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
+See also our response to \@ref(R1-8).
 
@@
 
-<!--# not sure how to respond.  I think there must be some underlying confusion.  Other reviewers like this explanation -->
+We agree with the reviewer that the exact timing of mortality within a season doesn't matter for population dynamics.
+This is likely obvious to demographers who are used to working with yearly census data and finer scale climate data.
+However, we think this paragraph is appropriate to follow up with the three before it---a delayed effect indicated by a DLNM plot like those in Figures 4, 5 and 6 could be due to 1) developmental delays, 2) buffering by below-ground storage tissues, 3) indirect effects, or 4) **none of the above.** We want to be sure to point out that in the case of survival especially, a "delayed effect" of SPEI 5 months in the past could simply mean that a plant died *during* a drought 5 months in the past (no actual delay between adverse conditions and death).
+Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
+We have revised this paragraph to emphasize the above points.
 
@@
 Is there some edit to the manuscript that would clarify?
-I don't think "let the model do the smoothing" accurately explains what would happen if we switched to SPEI_1 -->\
+I don't think "let the model do the smoothing" accurately explains what would happen if we switched to SPEI_1 -->
+
+\
 \
@@
 > However, wouldn't it be that the demographic effect takes place AFTER the census for an echo to appear at previous lags?
+
+We We have revised this paragraph to clarify.
+See our response to \@ref(R2-20).

### 2021-08-30 doc/paper.Rmd
Working on #86

@@
 Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting to expand, leads to the aborted production of flowering shoots.
-Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
-While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988].
-In fact, flowering in one year is associated with increased reproduction and growth in the next.
+Our results showed delayed increases in flowering probability after drought.
+Drought conditions could be favorable for *H.
+accuminata* flowering due to increased temperatures or decreased cloudiness associated with droughts[@pauCloudsTemperatureDrive2013], and the effects could be delayed due to the development time of inflorescences.
+A better understanding of the developmental effects of climate on *Heliconia* is needed to determine if this is the best explanation for lagged effects of drought on flowering.
 
@@
 
-Finally, demographic delays could be an artifact of the timing of responses in relation to the census date.
-If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g.
-in Figure \@ref(fig:surv-spei)b).
-In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
+Finally, demographic delays could be be demographic only, rather than due to physiological delays or indirect effects.
+For example, if a severe month-long drought 6 months prior to a census caused increased mortality, it would appear in our models as a delayed effect, even if plants died immediately within that month, because mortality was only captured once per year.
+In our case, this potential explanation for delayed effects applies only to plant size and survival, as the flowering season coincided with the yearly census.
+It also only applies to lags of up to 12 months since mortality, size, and flowering are recorded yearly.
 This possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
 While the very slow growth and extremely low mortality rates of *H.
-acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated.
+acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated to better understand delayed effects of climate.
+<!--# might need some other edits here to address why it doesn't matter for demography which of these explanations is true.  Might be a better sentence than the previous sentence -->
 

### 2021-08-31 review/Response to reviews.Rmd
Discuss pros and cons of DLNMs more. Closes #80.

@@
 
-<!--# The following isn't in the main text yet.  Just ideas. -->
-
-(Something I'm considering adding to the discussion):
-
-"The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored.
-@tellerLinkingDemographyDrivers2016 used simulations to show that 20--25 years of data were required to detect lagged effects using their method, but this cutoff depends on the effect size and the number of climate extremes in the dataset.
-Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required (Aldo Compagnoni's ESA talk)."
+The main limitations of this method are of data (see also reviewer 2's comment \@ref(R2-15)).
+We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment.
 
@@
 Models which predict size in year t+1 as a function of size in year t, such as ours, are often described as growth functions because they can be used to project change in size (i.e. growth).
-However, we agree that it would be more clear for this section heading to be "size" rather than "growth"
+However, we agree that it would be more clear for this section heading to be "size" rather than "growth".
 
@@
 
-We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest." <!--# are there other edits we can make to clarify?
-A sentence introducing the concept that size t+1 ~ size t *is* growth?
--->
+We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest."
 
@@
 It now includes the following possible explanation: *"Drought conditions could be favorable for H.
-accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts[@pauCloudsTemperatureDrive2013],* *and the effects could be delayed due to the development time of inflorescences."*
+accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts*[@pauCloudsTemperatureDrive2013]*,* *and the effects could be delayed due to the development time of inflorescences."*
 
@@
 
-...see also our response to \@ref(R1-5).
-<!--# R1-5 is asking for more discussion on limitations of DLNMs -->
+As reviewer 3 pointed out (\@ref(R3-11)), our extension to previous approaches is the use of a 2-dimensional spline to allow for fitting non-linear responses to the environment.
+We have edited this sentence accordingly
 
@@
 
-...see also our response to \@ref(R1-5) and \@ref(R2-13).
+We have added additional discussion on the limitations of DLNMs for detecting delayed effects to the discussion.
+See our responses to \@ref(R1-5).
 
@@
 
+<!--# wouldn't this reduce growth?  Not what we saw in our results, so not sure how this is relevant? -->
+
 ##  {#R2-19}
@@
 > Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.
+
+<!--# but we just said drought also removes tissue in the previous sentence.  Am I missing something? -->
 
@@
 
+While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality.
+
 ##  {#R2-22}
@@
 
+Although the differences in plant size and flowering are relatively small, they are statistically significant (see Table 1 in revised manuscript), and we believe they are demographically significant over long time periods.
 See our response to \@ref(R2-7).
@@
 
-<!--# yes, but might change number of knots, so might require more work revising this -->
+This is what we meant, and is no longer relevant as we have increased the number of knots.
+See response to \@ref(R3-5).
 
@@
 This resulted in some changes in the the shape of the crossbasis smooth and which areas of it were statistically significant.
-We have updated the methods and results sections accordingly.
-<!--# TODO: update results section! -->\
+We have updated the methods and results sections accordingly.\
 \
@@
 
-...see also our response to \@ref(R1-3)
+We have now included these in Table 1.
+See also our response to \@ref(R1-3).
 
@@
 
+We thank the reviewer for the positive comment.
+This is correct---our expansion was not the fitting of lagged effects with a spline, but the use of a 2 dimensional spline to *also* fit non-linear reaction norms.
+We have edited this sentence accordingly.
+
 ##  {#R3-12}

### 2021-08-31 doc/paper.Rmd
Discuss pros and cons of DLNMs more. Closes #80.

@@
 This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
-To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
-Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
+To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021; @ogleQuantifyingEcologicalMemory2015].
+Our extension of this approach allows for modeling of non-linear responses to the environment in addition to an unbiased way of identifying these delayed effects without overfitting.
+Our results are consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
 
@@
 
-Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021] <!--# add citation to Aldo's ESA talk -->.
-<!--# possibly good place to mention limitations of interactions -->Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
+Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021].
+@tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data.
+Compagnoni et al.
+( <!--# cite ms in prep?
+cite ESA talk?
+--> ) showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather.
+Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
 Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

### 2021-09-01 review/Response to reviews.Rmd
finishing touches on response to reviewers

@@
 
+We are delighted that the reviewer found the manuscript engaging and valuable.
+Our responses to specific suggestions follow.
+
 ##  {#R1-1}
@@
 
-We edited a sentence in the introduction that cites @estebanOtherSideDroughts2021 to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
+We edited a sentence in the introduction that cites Esteban et al. (2021) to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
 *"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction...".* Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon.
@@
 
-Because model selection was performed automatically by shrinkage using the `select = TRUE` option in the `gam()` function, we cannot show relative model fits (i.e., we did not perform model selection and can't show dAIC).
-However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term.
+Rather than comparing the fit of multiple models with different predictors, we took advantage of the `select = TRUE` option in the `gam()` function, which adds an additional penalty to smooth terms that effectively removes them from the model if they are not supported by data.
+Therefore we cannot show something like a dAIC table.
+However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term (Table 2).
 We have replaced many of the parenthetical p-values reported in the results section with references to this table.
@@
 
-We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs), but showing the raw data made the plots visually cluttered, in our opinion.
+We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs)), but showing the raw data made the plots visually cluttered, in our opinion.
 Another way we approached this comment is by excluding areas of the heatmaps in figures 4-6 that are far from observed data, equivalent to not drawing regression lines beyond the range of data.
@@
 Collapsing finer temporal scale environmental drivers to yearly averages (or a few means of arbitrarily chosen climate windows) is exactly what we avoid with this method.
+The caveat to this is that without finer scale demographic data, one may not be able to discern the cause of an apparent delayed effect, which is why we expand on several possible mechanisms for delayed effects in detail in the discussion.
 
@@
 We are a little hesitant to speculate too much here, as not much is known about the optimal microclimate conditions for flowering in *H.
-acuminata*, but we have revised this paragraph (see also our response to @ref(R2-17).
+acuminata*, but we have revised this paragraph (see also our response to \@ref(R2-17)).
 It now includes the following possible explanation: *"Drought conditions could be favorable for H.
-accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts*[@pauCloudsTemperatureDrive2013]*,* *and the effects could be delayed due to the development time of inflorescences."*
+accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts (Pau et al.
+2013),* *and the effects could be delayed due to the development time of inflorescences."*
 
@@
 This was referring to a previous iteration of the figure where the y-axis was on the response scale.
-We have reverted the figure to this version (see response below) and this should be obvious now.
+We have reverted the figure to this version (see response below) and this should be more obvious now.
 
@@
 This is equivalent to the marginal effect of log(size_t)---that is, without considering the additive effect of lagged SPEI or the random intercepts for different plots and for plants that had not flowered in the previous year (the reference level for the `flwr_prev` factor we included as a parametric covariate).
-We have not clarified that in the figure caption.
+We have now clarified that in the figure caption.
 
@@
 
+We are glad to hear the reviewer saw this work as important and well written.
+We address specific concerns and suggestions below.
+
 ##  {#R2-1}
@@
 
-The reference, @brunaHabitatFragmentationPlant2009, discusses the lack of knowledge of what demographic vital rates (e.g.
+The reference, Bruna et al.
+(2009), discusses the lack of knowledge of what demographic vital rates (e.g.
 survival, reproduction, recruitment) in which life stages (e.g.
@@
 > Lines 251ff: I don't see a reference of Fig 2 in the result section.
+
+We have added a sentence to the results that refers to this figure.
 
@@
 As reviewer 3 pointed out (\@ref(R3-11)), our extension to previous approaches is the use of a 2-dimensional spline to allow for fitting non-linear responses to the environment.
-We have edited this sentence accordingly
+We have edited this sentence accordingly.
 
@@
 We have added additional discussion on the limitations of DLNMs for detecting delayed effects to the discussion.
-See our responses to \@ref(R1-5).
+See also our responses to \@ref(R1-5).
 
@@
 We believe that the bet-hedging argument was confusing for two reasons: 1) it assumed that drought (as we measured it with SPEI) is detrimental to *H.
-accuminata*, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
+accuminata* survival, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
 We have therefore revised the example in this paragraph to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
@@
 
-<!--# wouldn't this reduce growth?  Not what we saw in our results, so not sure how this is relevant? -->
+We thank the reviewer for the suggestion---in fact, effects of drought on maintenance costs is precisely what we were trying to communicate in this line.
+After shedding shoots and leaves during drought, maintenance costs may be reduced, and plants might invest more in rhizomes, possibly allowing plants to recover after drought to a larger size.
+We are unaware of literature investigating the belowground effects of drought on *Heliconia*, but this is consistent with results in Bruna et al. 2002 (see response to \@ref(R2-19)).
 
@@
 
-<!--# but we just said drought also removes tissue in the previous sentence.  Am I missing something? -->
+We agree with the reviewer and have replaced this sentence with a more in-depth explanation of Bruna et al. 2002.
 
@@
 
-While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality.
+While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
+We have added a paragraph to the discussion regarding effects of wet conditions on tropical plants.
 
@@
 
-<!--# Emilio: I feel like I don't have a great response to this one.
-Is there some edit to the manuscript that would clarify?
-I don't think "let the model do the smoothing" accurately explains what would happen if we switched to SPEI_1 -->
-
 \
-\
-The choice of scale for the SPEI calculation is indeed difficult, and contributes to the definition of drought (i.e. is drought defined as dry conditions over the previous 1, 2, 3 or more months?)
-. A thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper
-.
+The choice of scale for the SPEI calculation is indeed difficult, and is related to the operational definition of drought (i.e.
+is drought defined as dry conditions over the previous 1, 2, 3 or more months?).
+We are not certain that the reviewer's suggestion to allow the GAM to "do the smoothing" would have the same effect as changing the scale parameter when calculating SPEI.
+SPEI is calculated using a sliding window, while GAMs use penalization to achieve smoothing.
+We believe a thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper.
 
@@
 This is what we meant, and is no longer relevant as we have increased the number of knots.
-See response to \@ref(R3-5).
+See our response to \@ref(R3-5).
 

### 2021-09-01 doc/paper.Rmd
fixed table captions by switching to pander.  Closes #97

@@
 library(patchwork)
+library(pander)
 ```
@@
 
+Interestingly, we found significant detrimental effects of unusually wet conditions at some lag times for all vital rates except survival in continuous forest and flowering probability in 1 ha fragments.
+Although the literature focuses primarily on the effects of droughts in the tropics, some research shows that unusually wet conditions can also be detrimental to tropical plants [@estebanOtherSideDroughts2021].
+Wet conditions could be associated with increased cloud cover which could reduce photosynthetic activity and decrease growth, or with wind storms which could increase tree- and branch-fall---a source of mortality for understory plants.
+Waterlogged soils can also decrease growth and increase mortality [@parentOverviewPlantResponses2008].
+Finally, cool temperatures associated with unusually wet conditions may decrease flowering [@pauCloudsTemperatureDrive2013].
+
 There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
@@
 The carbohydrates stored in rhizomes allow *Heliconia* to regenerate aboveground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
-This may be why drought led to delayed increases in growth---by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
-This would also be consistent with the results of prior experiments, in which the growth rates of *H.
-acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@brunaCompensatoryResponsesUnderstory2005].
+This may be why drought led to delayed increases in growth---by shedding shoots and leaves and investing in rhizomes, plants could be generating proportionately more buds with which to regenerate when conditions improve.
+This is consistent with the results of prior experiments in which *H.
+acuminata* plants transplanted into hotter, drier fragments lost and then recovered far more leaf area than control plants[@brunaExperimentalAssessmentHeliconia2002].
 
@@
 
-Finally, demographic delays could be be demographic only, rather than due to physiological delays or indirect effects.
-For example, if a severe month-long drought 6 months prior to a census caused increased mortality, it would appear in our models as a delayed effect, even if plants died immediately within that month, because mortality was only captured once per year.
-In our case, this potential explanation for delayed effects applies only to plant size and survival, as the flowering season coincided with the yearly census.
+Finally it is important to clarify the distinction between "physiological delays" and "observational delays".
+For example, a Fall drought could reduce cold tolerance and therefore overwinter survival.
+Alternatively the drought could kill plants immediately.
+In both cases, if mortality is recorded in the Spring, both would appear as a delayed effect of the drought, but only the first case is a truly delayed physiological response to drought, the apparent delayed response in the second case is due to the timing of the response in relation to the census date.
+In our case, this potential explanation of "observational delays" applies only to plant size and survival, as the flowering season coincided with the yearly census.
 It also only applies to lags of up to 12 months since mortality, size, and flowering are recorded yearly.
 This possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
-While the very slow growth and extremely low mortality rates of *H.
-acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated to better understand delayed effects of climate.
-<!--# might need some other edits here to address why it doesn't matter for demography which of these explanations is true.  Might be a better sentence than the previous sentence -->
+For some demographic models and follow-up experiments, knowing the precise timing of mortality might be critical; for others less so.
+To disentangle possible mechanisms for observed delayed effects, it may be necessary to conduct demographic surveys at the same temporal scale at which climate is aggregated.
 
@@
 @tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data.
-Compagnoni et al.
-( <!--# cite ms in prep?
-cite ESA talk?
---> ) showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather.
+@compagnoni2021 showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather.
 Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
@@
 
-We thank Collin Edwards, Andrew Mercadante, Ellie McDaniel, \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Collin Edwards, Andrew Mercadante, Ellie McDaniel and 3 anonymous reviewers for helpful discussions and comments on the manuscript.
 We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
@@
 
+\pagebreak
+
 # Tables
@@
 })
-flextable::set_flextable_defaults(font.family = "Cambria", font.size = 10)
 ```
 
-```{r}
-ci_cap <- "Model intercepts and their 84% confidence intervals.
+```{r intercept-table, results='asis'}
+intercept_df[c(2,4,6), 1] <- ""
+emphasize.strong.cells(as.matrix(data.frame(row = 3:6, col = rep(3,4))))
+pandoc.table(
+  intercept_df,
+  justify = c("left", "left", "right"),
+  full_width = FALSE,
+  caption = "(\\#tab:intercept-table) Model intercepts and their 84% confidence intervals.
 Non-overlapping confidence intervals are bolded and can be interpreted as a significant difference in the vital rate between continuous forest and 1 ha fragments."
-```
-
-```{r label="intercept-table"}
-flextable(intercept_df) %>% 
-  autofit() %>% 
-  merge_v(j = 1) %>% 
-  valign(j = 1, valign = "top") %>% 
-  align(j = 3, align = "right", part = "all") %>% 
-  bold(j = 3, i = c(3:6)) %>% 
-  fix_border_issues() %>%
-  set_caption(ci_cap, style = "Image Caption")
+)
 ```
@@
 
-```{r}
-result_cap <- 'Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H.
+```{r results-table, results='asis'}
+results_tab <- 
+  results_df %>% 
+  mutate(across(c(R2, `test statistic`),
+                ~round(.x, 2) %>% 
+                  prettyNum(big.mark = ","))) %>% 
+  mutate(term = case_when(
+    term == "s(log(size_t))" ~ "s(log(size~t~))",
+    term == "flowered" ~ "flowering~t~",
+    TRUE ~ term
+  ),
+  `vital rate` = case_when(
+    `vital rate` == "size_next" ~ "log(size~t+1~)", 
+    `vital rate` == "flowering" ~ "flowering~t+1~",
+    TRUE ~ `vital rate`)) %>% 
+  rename("R^2^" = R2)
+results_tab[c(2:8, 10:16, 18:24), 1] <- ""
+results_tab[c(2:4, 6:8, 10:12, 14:16, 18:20, 22:24), 2:3] <- ""
+
+pandoc.table(
+  results_tab,
+  full_width = FALSE,
+  split.table = 100,
+  justify = c("left", "left", "right", "left", "right", "right", "right"),
+  caption = '(\\#tab:results-table) Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H.
 acuminata* in continuous forest and 1 ha fragment plots.
 The adjusted R^2^ is reported as a measure of model fit.
-The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowered"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)").
+The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowering~t~"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)").
 Degrees of freedom are reported for the parametric term, and estimated degrees of freedom (edf) are reproted for smooths.
@@
 The test statistic reported is $\\chi^2$ for survival and flowering and $F$ for log(size~t+1~).'
-```
-
-<!--# add something about how p-values are appropriate for random effects?  see help files from mgcv for references -->
-
-```{r label="results-table"}
-flextable(results_df) %>% 
-  merge_v(j = c(1,2,3)) %>% 
-  valign(j = c(1,2,3), valign = "top") %>% 
-  align(j = c(5, 7), align = "right", part = "all") %>%
-  colformat_double(j = c(3,6), digits = 2) %>% 
-  fix_border_issues() %>% 
-  flextable::compose(j = 3, part = "header",
-                     value = as_paragraph("model R", as_sup("2"))) %>% 
-  flextable::compose(j = 4, i = c(2,6,10,14,18,22),
-                     value = as_paragraph("s(log(size", as_sub("t"), "))")) %>%
-  flextable::compose(j = 1, i = 9, 
-                     value = as_paragraph("log(size", as_sub("t+1"), ")")) %>% 
-  autofit() %>% 
-  set_caption(result_cap, style = "Image Caption")
+)
 ```

### 2021-09-01 packages.R

@@ -23,5 +23,5 @@ library(colorspace)
 library(readxl)
 library(Hmisc)
 library(qqplotr)
-library(flextable)
+library(pander)
 library(scales)

### 2021-09-01 doc/paper.Rmd
remove old text graveyard

@@
 
-<!-- # Left-over text -->
-
-<!-- -   populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].
--->
-
-<!-- -   While the alteration of microclimate near forest edges is one of the fundamental results to have emerged from the study of fragmented landscapes [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013], little is known regarding how variable it is.
-It We hypothesize that the microclimate in continuous forest plots is more stable than that in 1 ha fragments, and that this stability will demographically buffers populations in forest interiors [@brunaExperimentalAssessmentHeliconia2002].
--->
-
-<!-- -   There can be cascading effects on other species (hispine beetles that depend on Heliconia flowers/fruits, hummingbirds and seed dispersers ( @stilesEcologyFloweringPhenology1975 -->
-
-<!-- -   eg.
-gene flow and genetic structure among and between plots already shaped by fragmentation ( @cortesLowPlantDensity2013, there can also be drought effects eg @sifresGeneticStructureLycopersicon2007) -->
-
-<!-- -   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
-Estimates green-up, literfall, growth, etc.
-for central and southern America including the Amazon.
-Drought results in anomalously high leaf flushing at start of drought followed by increased litterfall during the drought.
-Dry season droughts result in green-up, but not wet season droughts.
-During drought, old leaves are shed and new leaves are maintained.
-2-5 months for newly flushed leaves to fully expand -->
-
-<!-- # Citations / information could allocate to appropriate paragraphs -->
-
-<!-- -   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod.
-can be affected -->
-
-<!-- -   Effects of precip extremes on understory plants can be also be indirect: -->
-
-<!--     -   high precip = lots of clouds, less photosynthesis, less or slower growth -->
-
-<!--     -   low precip = trees lose leaves or die, more light gets through or gaps open. -->
-
-<!--         This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
--->
-
-<!--         Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020 -->
-
-<!-- -   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet -->
-
-<!-- -   @wrightSeasonalNinoLonger2006: Increase in flower and seed production after El Niño events for trees and lianas. -->
-
-<!--     Might be due to increases solar irradiance. -->
-
-<!--     Maybe evidence of lagged effects? -->
-
-<!--     Table 2 is inconsistent with text, so not sure... -->
-
-<!-- -   [@estebanOtherSideDroughts2021] negative effects of large positive SPEI -->
-
-<!-- -   In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
--->
-
-<!-- -   @leitoldNinoDroughtIncreased2018 increased canopy turnover during drought. -->
-
-<!--     Similar increase in continuous and (large) fragments. -->
-
-<!--     Increase in turnover more variable in fragments. -->
-
-<!--     Coarse woody debris increased 62% during drought. -->
-
-<!-- \newpage -->
-
 <!-- ### Colophon -->

### 2021-09-01 doc/supplemental.Rmd
add raw data plot supplement. closes #79

@@
 
+# Size plots with raw data
+
+```{r}
+withr::with_dir(here::here(), {
+  targets::tar_load(c(
+    s_cf,
+    s_1ha,
+    g_cf,
+    g_1ha,
+    f_cf,
+    f_1ha,
+  ))
+})
+```
+
+```{r}
+surv_curves <- 
+  map_df(list(CF = s_cf, "1-ha" = s_1ha), ~{
+    smooth_estimates(.x, smooth = "s(log_size_prev)", unconditional = TRUE, overal_uncertainty = TRUE) %>% add_confint() %>% 
+      add_column(intercept = coef(.x)[1]) %>% 
+      mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + intercept)))
+  },
+  .id = "habitat")
+
+raw_df <- bind_rows("1-ha" = model.frame(s_1ha), "CF" = model.frame(s_cf), .id = "habitat")
+
+s <- 
+  ggplot(surv_curves, aes(x = log_size_prev, y = est)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.4) +
+  geom_violin(data = filter(raw_df, surv == 1),
+              aes(y = 1.05),
+              draw_quantiles = c(0.5),
+              width = 0.1) +
+  geom_violin(data = filter(raw_df, surv == 0),
+              aes(y = -0.05),
+              draw_quantiles = c(0.5),
+              width = 0.1) +
+  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
+  facet_wrap(~habitat) +
+  theme_bw() +
+  labs(y = "P(survival)", x = "log(size_t)")
+```
+
+```{r}
+flwr_curves <- 
+  map_df(list(CF = f_cf, "1-ha" = f_1ha), ~{
+    smooth_estimates(.x, smooth = "s(log_size_prev)", unconditional = TRUE, overal_uncertainty = TRUE) %>% add_confint() %>% 
+      add_column(intercept = coef(.x)[1]) %>% 
+      mutate(across(c(est, lower_ci, upper_ci), ~plogis(.x + intercept)))
+  },
+  .id = "habitat")
+
+raw_df <- bind_rows("1-ha" = model.frame(f_1ha), "CF" = model.frame(f_cf), .id = "habitat")
+
+f <- 
+  ggplot(flwr_curves, aes(x = log_size_prev, y = est)) +
+  geom_line() +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.4) +
+  geom_violin(data = filter(raw_df, flwr == 1),
+              aes(y = 1.05),
+              draw_quantiles = c(0.5),
+              width = 0.1) +
+  geom_violin(data = filter(raw_df, flwr == 0),
+              aes(y = -0.05),
+              draw_quantiles = c(0.5),
+              width = 0.1) +
+  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
+  facet_wrap(~habitat) +
+  theme_bw() +
+  labs(y = "P(flowering)", x = "log(size_t)")
+```
+
+```{r}
+library(ggpointdensity)
+size_curves <- 
+  map_df(list(CF = g_cf, "1-ha" = g_1ha), ~{
+    smooth_estimates(.x, smooth = "s(log_size_prev)", unconditional = TRUE, overal_uncertainty = TRUE) %>% add_confint() %>% 
+      add_column(intercept = coef(.x)[1]) %>% 
+      mutate(across(c(est, lower_ci, upper_ci), ~.x + intercept))
+  },
+  .id = "habitat")
+
+raw_df <- bind_rows("1-ha" = model.frame(g_1ha), "CF" = model.frame(g_cf), .id = "habitat")
+
+g <- 
+  ggplot(size_curves, aes(x = log_size_prev, y = est)) +
+  geom_pointdensity(data = raw_df, aes(y = log_size), method="kde2d") +
+  scale_color_viridis_c() +
+  geom_line() +
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.6) +
+  facet_wrap(~habitat) +
+  theme_bw() +
+  labs(y = "log(size_t+1)", x = "log(size_t)", color = "point density")
+```
+
+Figure S3.
+The curves and confidence intervals plotted here are the same as those in Figure 3a–c but separated by habitat (1-ha fragments in left panel, continuous forest plots in right panel) and with raw data superimposed.
+For size (a), raw data are plotted as points with color representing point density to assist in visualizing overplotted data.
+For survival (b) and flowering (c), raw data (survived or flowered = 1, died or didn't flower = 0) are shown using violin plots with the median marked by a vertical line.
+
+```{r fig.width=7, fig.height=7}
+library(patchwork)
+(g + theme(axis.title.x = element_blank()))/
+(s + theme(axis.title.x = element_blank()))/
+(f) + plot_annotation(tag_levels = "a")
+```
+
+\pagebreak
+
 # References

### 2021-09-01 doc/supplemental.Rmd
added figure to supplemental

@@
 library(performance)
+library(gratia)
+library(ggpointdensity)
 source(here::here("R", "utils.R"))
@@
 withr::with_dir(here::here(), {
-  targets::tar_load(c(
+  tar_load(c(
     s_cf,
@@
     f_cf,
-    f_1ha,
+    f_1ha
   ))
@@
 ```{r}
-library(ggpointdensity)
 size_curves <-
@@
 
+```{r fig.width=7, fig.height=7}
+(g + theme(axis.title.x = element_blank()))/
+(s + theme(axis.title.x = element_blank()))/
+(f) + plot_annotation(tag_levels = "a")
+```
+
 Figure S3.
@@
 For survival (b) and flowering (c), raw data (survived or flowered = 1, died or didn't flower = 0) are shown using violin plots with the median marked by a vertical line.
-
-```{r fig.width=7, fig.height=7}
-library(patchwork)
-(g + theme(axis.title.x = element_blank()))/
-(s + theme(axis.title.x = element_blank()))/
-(f) + plot_annotation(tag_levels = "a")
-```
 

### 2021-09-01 _targets.R

@@ -31,23 +31,23 @@ tar_plan(
   xa_raw  = read_csv(xa_file),
   xa_spei = calc_spei_xa(xa_raw, scale = 3),
   xa_lag  = lag_spei(xa_spei, maxlag),
-  
+
   # Prep demographic data
   tar_target(demog_file, here("data", "Ha_survey_with_Zombies.csv"), format = "file"),
   demog_raw  = read_fix_demog(demog_file),
   ## filter_dupes() removes some duplicate HA id numbers.  Will eventually be fixed in raw data
-  demog_done = demog_raw %>% filter_dupes() %>% wrangle_demog(), 
+  demog_done = demog_raw %>% filter_dupes() %>% wrangle_demog(),
   model_data        = join_filter_demog_spei(demog_done, xa_lag),
   model_data_cf     = dplyr::filter(model_data, habitat == "CF"),
   model_data_1ha    = dplyr::filter(model_data, habitat == "1-ha"),
   model_data_cf_sub = subset_cf(model_data),
-  
+
   # Data validation
   tar_render(validate_data, "doc/validate_data.Rmd", deployment = "main"),
-  
+
   # Select number of knots for GAMs
   # tar_render(knot_choice, "doc/knot_choice.Rmd", deployment = "main"),
-  
+
   # Fit demographic models
   tar_target(s_cf,  fit_surv(model_data_cf, k = c(10,15,15))),
   tar_target(s_1ha, fit_surv(model_data_1ha, k = c(10,10,10))),
@@ -55,7 +55,7 @@ tar_plan(
   tar_target(g_1ha, fit_growth(model_data_1ha, k = c(10,5,15))),
   tar_target(f_cf, fit_flwr(model_data_cf, k = c(10,15,15))),
   tar_target(f_1ha, fit_flwr(model_data_1ha, k = c(10,10,18))),
-  
+
   # Validate and summarize results
   ### Check for edf differences due to sample size
   tar_target(g_cf_sub, fit_growth(model_data_cf_sub, k = c(25,5,15))),
@@ -67,15 +67,15 @@ tar_plan(
   normals = normals_data(),
   normals_plot = plot_normals(normals),
   plot_dates = get_plot_daterange(model_data, maxlag),
-  
+
   eda_spei = plot_eda_spei(xa_lag, plot_dates),
   eda_surv_ts = plot_eda_surv_ts(model_data, plot_dates),
[…]

### 2021-09-01 conflicts.R

@@ -6,4 +6,3 @@ conflict_prefer("lag", "dplyr")
 conflict_prefer("slice", "dplyr")
 conflict_prefer("summarize", "dplyr")
 conflict_prefer("Lag", "tsModel")
-conflict_prefer("compose", "flextable")

### 2021-09-01 packages.R

@@ -25,3 +25,4 @@ library(Hmisc)
 library(qqplotr)
 library(pander)
 library(scales)
+library(ggpointdensity)

### 2021-09-01 README.Rmd

@@
 
-**NOTE**: The full pipeline will take a *very long* time to run.
-I suggest switching off individual-level random effects (more details in comments in `_targets.R`) if you want to run the project on your own computer.
-Without individual-level random effects the pipeline should run relatively quickly (< 1hr, I think).
-
 **Using HPC**:
-Also, you may need to install packages on the remote HPC if using `tar_make_clustermq()` with the SSH option.
+You may need to install packages on the remote HPC if using `tar_make_clustermq()` with the SSH option.
 Unfortunately, I'm not sure if there is an easy way to do this using `renv`.

### 2021-09-01 README.Rmd
add Zenodo DOI

@@
 
-[![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml)
+[![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml) [![DOI](https://zenodo.org/badge/289078614.svg)](https://zenodo.org/badge/latestdoi/289078614)
 

### 2021-09-01 doc/paper.Rmd
add raw data plot supplement. closes #79
add Zenodo DOI

@@
 
-Data used in this study are available at the Dryad Digital Repository; R code for analyses and visualizations are archived at Zenodo [links included upon acceptance].
+Data used in this study are available at the Dryad Digital Repository [links included upon acceptance]; R code for analyses and visualizations are archived with Zenodo at <https://doi.org/10.5281/zenodo.5364591>.
 
@@
 The bands depicting the 95% confidence interval include uncertainty in the intercept and uncertainty due to smoothness selection; the smooths for 1-ha fragments and continuous forest are fit in separate models.
-(d) Plant size distribution by habitat type (solid line = 1-ha fragments, dashed line = Continuous Forest)."
+(d) Plant size distribution by habitat type (solid line = 1-ha fragments, dashed line = Continuous Forest).
+The curves in a–c are shown with raw data superimposed in Figure S3."
 ```

### 2021-09-02 doc/paper.Rmd
proofread

@@
 Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (*Heliconia acuminata*, Heliconiaceae).
-We used distributed lag non-linear models to understand the delayed effects of standardized precipitation evapotranspiration index (SPEI) on survival, growth, and flowering.
+We used distributed lag non-linear models to understand the delayed effects of climate (measured as standardized precipitation evapotranspiration index, SPEI) on survival, growth, and flowering.
 We detected delayed effects of climate up to 36 months.
-Drought two dry seasons prior to the February census decreased survival and increased flowering probability while drought in the wet season a year prior to the census decreased flowering probability and increased growth.
+Extremes in SPEI in the previous year reduced survival, drought in the wet season 8--11 months prior to the February census increased growth, and drought two dry seasons prior increased flowering probability.
 The effects of extremes in precipitation on survival and growth were more pronounced in forest fragments compared to continuous forest.
@@
   group_by(habitat) %>% 
-  summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants))
+  summarize(mean_n_plants = mean(n_plants), sd_n_plants = sd(n_plants), median_n_plants = median(n_plants), min = min(n_plants), max = max(n_plants))
 plants_per_plot
@@
 Our dataset comprised `r format(summary_stats$n_plants[2], big.mark = ",")` plants in continuous forest and `r format(summary_stats$n_plants[1], big.mark = ",")` plants in forest fragments.
-Plots in CF had on average 2.7-fold more plants than plots in 1-ha fragments (CF = 681 ± 493 SD; 1-ha = 253 ± 30 SD).
+Plots in CF had on average more than twice as many plants than plots in 1-ha fragments (CF median = 788, range = [201, 1549]; 1-ha median = 339, range = [297, 400]).
 
@@
 In this study, we used the product of shoot number and plant height as our measure of plant size.
-Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
-After the initial census year, between 80% and 97% of marked plants were found.
+Preliminary analysis showed that the product of shoot number and height was a better predictor of total leaf area (which in turn is assumed to be a strong predictor of aboveground biomass) than either shoot number or height alone (Table S2<!--# update crossref manually -->).
+After the initial census year, between 80% and 97% of marked plants were found at each survey.
 Of plants that had missing values for some years, but were found again in a subsequent year, 95% had 2 or fewer years of missing values.
@@
 SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H.
-acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
+acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010; @mckeeRelationshipDroughtFrequency1993].
 Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
-SPEI calculations were made using the `SPEI` package [@begueriaSPEICalculationStandardised2017].
+SPEI calculations were made using the `SPEI` package in R [@begueriaSPEICalculationStandardised2017].
 The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought sometimes differed (Figure S2; Table S1 <!--# update crossrefs manually -->).
@@
 To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs, @gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
-<!--# DLNMs solve the problems created by including all possible climate windows in a single model---namely collinearity of predictors and the arbitrary choice of window length -->They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
-<!--# The shape of the function can be penalized so that it is only as complex as the data supports, thus reducing effective degrees of freedom --> Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
+DLNMs capture how potentially delayed effects of predictor variables (e.g.
+SPEI) affect an outcome (e.g.
+growth) well beyond the event period.
+DLNMs avoid the problems of including all possible climate windows in a single model---namely collinearity of predictors and the arbitrary choice of window length.
+They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
+This models a non-linear relationship between predictor and response (e.g.
+between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+The shape of the function can be penalized so that it is only as complex as the data supports, thus reducing effective degrees of freedom.
+We used a crossbasis function of SPEI and lag time with possible lags from 0--36 months.
 We chose 36 months as a maximum lag because prior transplant experiments with *H.
@@
 We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
-A random effect of plant ID was not included in survival models since each plant only dies once.
 Individual-level random effects were not included due to computational limitations in fitting random intercepts for all of the nearly 5,000 plants in continuous forest plots.
@@
 
-The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
+The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` R package [@woodGeneralizedAdditiveModels2017].
 The crossbasis smooth was coded as:
@@
 `L` is an $n \times m$ matrix with columns describing the lag structure, i.e. the integers 0 through 36.
-The spline basis, `"cb"` is provided by the `dlnm` package, and we specified the marginal bases as cubic regression splines with the `xt` argument.
-For the full code used to fit the models, see <!--# insert citation to Zenodo archive -->.
+The spline basis, `"cb"` is provided by the `dlnm` package [@gasparriniDistributedLagLinear2011], and we specified the marginal bases as cubic regression splines with the `xt` argument.
+For the full code used to fit the models, see the Data Availability Statement.
 
 We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
-This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
+This was modeled with a scaled t family error distribution because residuals were leptokurtic when a Gaussian family was used.
 Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
@@
 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
-Plants were significantly larger in continuous forest (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time.
+Plants were significantly larger in continuous forest on average (Table \@ref(tab:intercept-table)), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time.
 
@@
 Decreasing SPEI at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats.
-In continuous forest, but not 1 ha fragments, SPEI at lags of 22--25 months had a significant effect on plant size in the opposite direction, with wet conditions at that lag time resulting in larger predicted sizes.
-At longer lags of 26--36 months positive effects of drought on plant size are predicted for both continuous forest and 1 ha fragments by our models, although the area that reaches statistical significance is larger for continuous forest.
+In continuous forest, but not 1-ha fragments, SPEI at lags of 22--25 months had a significant effect on plant size in the opposite direction, with wet conditions at that lag time resulting in larger predicted sizes.
+At longer lags of 26--36 months positive effects of drought on plant size are predicted for both continuous forest and 1-ha fragments by our models, although the area of SPEI--lag space that reaches statistical significance is larger for continuous forest.
 
@@
 In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1) and droughts two dry seasons prior (lags 15--20) increased the probability of flowering.
-The shape of the crossbasis smooth for 1 ha fragments suggested that moderate drought (-1.5 \<SPEI\< -1) in the previous 0--18 months and at lags of 28--36 months significantly increased flowering probability slightly (the highest probability estimated is 0.004; Figure \@ref(fig:flwr-spei)b).
-The effects of drought on flowering probability appeared stronger in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
+The shape of the crossbasis smooth for 1-ha fragments suggested that moderate drought (-1.5 \<SPEI\< -1) in the previous 0--18 months and at lags of 28--36 months significantly increased flowering probability slightly (the highest probability estimated is 0.004; Figure \@ref(fig:flwr-spei)b).
+The effects of drought on flowering probability appeared stronger in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)c).
 We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were significantly more likely to be larger and flower again.
@@
 While intact forest and its canopy buffer populations from climatic extremes, populations in fragments---especially near edges with high contrast matrix---likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
-We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying the observed reduced plant growth and survival in forest fragments.
+We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying the observed reduced plant survival and size in forest fragments.
 
@@
 These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation [@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
-There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].
+There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pauCloudsTemperatureDrive2013].
 
@@
 To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021; @ogleQuantifyingEcologicalMemory2015].
-Our extension of this approach allows for modeling of non-linear responses to the environment in addition to an unbiased way of identifying these delayed effects without overfitting.
 Our results are consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.
+Additionally, our method allowed us to capture significant non-linear responses to SPEI at different lag times, which appear to differ in shape between habitats.
 
@@
 
-Interestingly, we found significant detrimental effects of unusually wet conditions at some lag times for all vital rates except survival in continuous forest and flowering probability in 1 ha fragments.
+Interestingly, we found significant detrimental effects of unusually wet conditions at some lag times for all vital rates except survival in continuous forest and flowering probability in 1-ha fragments.
 Although the literature focuses primarily on the effects of droughts in the tropics, some research shows that unusually wet conditions can also be detrimental to tropical plants [@estebanOtherSideDroughts2021].
@@
 Drought conditions could be favorable for *H.
-accuminata* flowering due to increased temperatures or decreased cloudiness associated with droughts[@pauCloudsTemperatureDrive2013], and the effects could be delayed due to the development time of inflorescences.
-A better understanding of the developmental effects of climate on *Heliconia* is needed to determine if this is the best explanation for lagged effects of drought on flowering.
+accuminata* flowering due to increased temperatures or decreased cloudiness associated with droughts [@pauCloudsTemperatureDrive2013], and the effects could be delayed due to the development time of inflorescences.
 
@@
 This is consistent with the results of prior experiments in which *H.
-acuminata* plants transplanted into hotter, drier fragments lost and then recovered far more leaf area than control plants[@brunaExperimentalAssessmentHeliconia2002].
+acuminata* plants transplanted into hotter, drier fragments lost and then recovered far more leaf area than control plants [@brunaExperimentalAssessmentHeliconia2002].
 
@@
 Alternatively the drought could kill plants immediately.
-In both cases, if mortality is recorded in the Spring, both would appear as a delayed effect of the drought, but only the first case is a truly delayed physiological response to drought, the apparent delayed response in the second case is due to the timing of the response in relation to the census date.
-In our case, this potential explanation of "observational delays" applies only to plant size and survival, as the flowering season coincided with the yearly census.
+If mortality is recorded in Spring, then mortality would appear as a delayed effect of the drought in both hypothetical cases, but only the first case is a truly delayed physiological response to drought.
+The apparent delayed response in the second case is due to the timing of the response in relation to the census date.
+In our analysis, this potential explanation of "observational delays" applies only to plant size and survival, as the flowering season coincided with the yearly census.
 It also only applies to lags of up to 12 months since mortality, size, and flowering are recorded yearly.
@@
 Over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation [@broadbentForestFragmentationEdge2008].
-Climate models predict a future of extremes for these forests---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015].
+Climate models predict a future of extremes for these forests---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffyProjectionsFutureMeteorological2015].
 Our results support the hypothesis that populations in Amazonian forest fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014].
@@
 Recent research, including our own presented here, shows that lagged effects of climate drivers on demography may be the norm [@eversLaggedDormantSeason2021].
-As such, we suggest that all demographers investigate the possibility of lagged environmental effects, ideally using methods that do not require an *a priori* choice of lag times such as DLNMs, automated algorithms for choosing critical weather windows[@vandepolIdentifyingBestClimatic2016], or Bayesian methods that incorporate "ecological memory" [@ogleQuantifyingEcologicalMemory2015].
+As such, we suggest that all demographers investigate the possibility of lagged environmental effects, ideally using methods that do not require an *a priori* choice of lag times such as DLNMs, automated algorithms for choosing critical weather windows [@vandepolIdentifyingBestClimatic2016], or Bayesian methods that incorporate "ecological memory" [@ogleQuantifyingEcologicalMemory2015].
 
@@
 @tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data.
-@compagnoni2021 showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather.
+Work in progress from @compagnoni2021 showed that spatial replication may alleviate this requirement, and in fact in our study we were able to detect significant lagged effects with only 10 years of data and multiple plots experiencing slightly different weather.
 Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
@@
 Error bars represent the standard deviation.
-(b) The proportion of plants surviving from one transition year to the next varied from 0.98 (CF in 1998-1999) to 0.91 (CF in 2003-2004).
+(b) The proportion of plants surviving from one transition year to the next varied from 0.98 (CF in 1998-1999) to 0.93 (CF in 2003-2004).
 (c) The proportion of *H.

### 2021-09-02 review/cover letter.Rmd
coverletter for resubmission

---
title: ""
author: ""
output: word_document
---

Dear handling editor,

We identified three main suggestions from reviewer and editor comments:

1) A need for a more in-depth explanation of distributed lag non-linear models, both for a technical audience (for replicability) and for a general ecology audience (for interpretabillity). We have expanded and revised the statistical modeling subsection to improve on this area.

2) A need for more discussion of the caveats and limitations of our modeling.  For example, since we were not able to test for a statistical interaction between habitat and lagged SPEI, we have revised the results to be more clear about which conclusions are strongly supported by statistical results.

3) Finally, reviewer 3's comments made us re-evaluate our approach to limiting the number of "knots" (or maximum basis complexity) in our models.  This (along with an updated version of the data with some duplicate plants corrected) lead to some cascading changes in our methods, results, and discussion, although the general conclusion that there are lagged effects of SPEI that differ among vital rates and habitats remains unchanged.  Increasing $k$, the maximum basis complexity, resulted in changes to the shape of the crossbasis smooths, and have ensured that increasing $k$ further yet does not result in further changes to the shape of the smooths (see model validation supplement). The change in crossbasis shape changed the statistical significance of some lag times for some models. This increase in $k$ increased the computational time for models in a way that scales multiplicatively with random effects terms such that it was no longer feasible to include individual-level random effects.  We ensured that removing these individual-level random effects did not bias the results using some data subsets (now mentioned in the updated methods).

The suggestions from all reviewers really improved the manuscript, and we are very thankful for such thorough and thoughtful reviewers. We have addressed all suggestions and comments in detail in our response to reviewer comments.

Sincerely,

Eric Scott

### 2021-09-02 review/Response to reviews.Rmd
add raw data plot supplement. closes #79
proofread
add line refs

@@
 ---
-
-Cascading effects of increasing `k`:
-
--   Changed shape of crossbasis functions (edf changes, now not always more complex in fragments)
-
--   Changes in which lags are significant
-
--   Individual-level random effects impossible to fit due to computational limitations.
-    Using a sub-sample, checked with and without invididual-level random effects.
-    Changes r2, but very minimal changes to shape of smooths, edfs, and significance of smooths.
-    Dropped from flowering model.
-
-For coverletter:
-
-we identified \_\_\_ main suggestions
-
-1.  more in depth explanation of model (both technical and for broad audience)
-2.
-something about changing declaritive statements about habitat:SPEI,Lag interaction.
-Added caveats about limitations of DLNM approach.
-Expanded on this in future directions to draw attention to the importance of including interactions for demography.
-3.  R\_ suggested increasing k, important suggestion. While it didn't alter the conclusions, it did require considerable work and further evaluation of results. new appendix, review rewrite
 
@@
 
-We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages.
-<!--# add line reference before submiting! --> We are not entirely sure if this is the type of description the reviewer was looking for.
+We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages (L229--239).
+We are not entirely sure if this is the type of description the reviewer was looking for.
 If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
@@
 We have replaced many of the parenthetical p-values reported in the results section with references to this table.
+Additionally, we have added supplemental Figure S3 which shows the curves as separate panels for continuous forest and fragments with raw data superimposed.
 
@@
 
-We have added additional detail to the conclusions recommending that all demographers consider lagged effects when possible in light of recent research (including this study).
-<!--# insert line ref -->
+We have added additional detail to the conclusions recommending that all demographers consider lagged effects when possible in light of recent research (including this study) (L536--541).
 
@@
 
-While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
+While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now L465--469) and delayed effects on growth (L 445--450, now L475--480), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
 We have added a paragraph to the discussion regarding effects of wet conditions on tropical plants.
@@
 We believe a thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper.
+Additionally, the paper originally describing SPI recommends 3-month scale for droughts defined as depletion of soil moisture (McKee et al. 1993).
 
@@
 We added a paragraph to the methods explaining our strategy for visualizing the GAM results.
-
-<!--# see also response to comments where we decide to backtransform univariate smooths -->
 

### 2021-09-08 review/Response to reviews.Rmd
added line number references

@@
 We edited a sentence in the introduction that cites Esteban et al. (2021) to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
-*"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction...".* Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon.
+*"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction..."* (L67--69).
+Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon (L64).
 
@@
 
-At the reviewer's suggestion, we have added a short paragraph to the introduction explaining why the pressures of habitat fragmentation and climate change might interact (i.e. be more than additive).
+At the reviewer's suggestion, we have added a short paragraph (starting on L74) to the introduction explaining why the pressures of habitat fragmentation and climate change might interact (i.e.
+be more than additive).
 
@@
 The main limitations of this method are of data (see also reviewer 2's comment \@ref(R2-15)).
-We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment.
+We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment (L542--550).
 
@@
 
-We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest."
+We have replaced this sentence.
+See our response to \@ref(R2-5).
 
@@
 acuminata*, but we have revised this paragraph (see also our response to \@ref(R2-17)).
-It now includes the following possible explanation: *"Drought conditions could be favorable for H.
-accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts (Pau et al.
-2013),* *and the effects could be delayed due to the development time of inflorescences."*
+It now includes the following possible explanation: *"Drought conditions could be favorable for* H.
+accuminata *flowering due to increased temperatures or decreased cloudiness associated with droughts (Pau et al.
+2013),* *and the effects could be delayed due to the development time of inflorescences."*(L470-473).
 
@@
 We have edited the abstract.
-Now the sentence to which the reviewer refers reads: *"We used distributed lag non-linear models to understand the delayed effects of standardized precipitation evapotranspiration index (SPEI) on survival, growth, and flowering."*
+Now the sentence to which the reviewer refers reads: *"We used distributed lag non-linear models to understand the delayed effects of climate (measured as standardized precipitation evapotranspiration index, SPEI) on survival, growth, and flowering."* (L33).
 
@@
 small or large individuals) contribute most to population declines in fragments.
-We have re-ordered this sentence to emphasize the hypothesis that fragmentation alters survival, growth, or reproduction with the caveat that few studies have determined which of these vital rate changes are driving extinctions.
+We have re-ordered this sentence (now L50) to emphasize the hypothesis that fragmentation alters survival, growth, or reproduction with the caveat that few studies have determined which of these vital rate changes are driving extinctions.
 
@@
 The reviewer is correct that the height of a plant growing more horizontally (for example if it was knocked over by a branchfall) could be shorter than the length from the base to the end of the tallest leaf.
-We edited the description to read: "...the height of each plant (i.e. distance from the ground to the tallest leaf tip)"
+We edited the description to read: *"...the height of each plant (i.e. distance from the ground to the tallest leaf tip)"* (L152)
 
@@
 
-We have added a sentence to the results that refers to this figure.
+We have added a sentence to the results that refers to this figure (L284).
 
@@
 We thank the reviewer for pointing out this mistake.
-We have edited this line to read "Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats." See also our response to \@ref(R1-figs).
+We have edited this line to read "*Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats.*" (L303).
+See also our response to \@ref(R1-figs).
 
@@
 This sentence meant to describe the difference between CF and fragments, not the absolute size in fragments.
-It has been re-worded to clarify.
+It has been re-worded to clarify (L348).
 
@@
 whether the shape of the crossbasis smooth was *significantly* different between habitats), we did test for a main effect of habitat by looking for overlap in 84% confidence intervals of model intercepts (see also response to \@ref(R2-6)).
-However, to reduce confusion here, we have changed "significantly" to "substantially" so as not to imply statistical significance for all of "magnitude, direction and lag time".
+However, to reduce confusion here, we have removed "significantly" so as not to imply statistical significance for all of "magnitude, directions and lag times" (L. 377).
 See also our response to \@ref(R2-7).
@@
 As reviewer 3 pointed out (\@ref(R3-11)), our extension to previous approaches is the use of a 2-dimensional spline to allow for fitting non-linear responses to the environment.
-We have edited this sentence accordingly.
+We have edited this sentence accordingly (L436).
 
@@
 accuminata* survival, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
-We have therefore revised the example in this paragraph to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
+We have therefore revised the example in this paragraph (starting on L462) to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
 See also our response to \@ref(R1-8).
@@
 
-We agree with the reviewer and have replaced this sentence with a more in-depth explanation of Bruna et al. 2002.
+We agree with the reviewer and have replaced this sentence with a more in-depth explanation of Bruna et al. 2002 (L480).
 
@@
 Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
-We have revised this paragraph to emphasize the above points.
+We have revised this paragraph (starting on L498) to emphasize the above points.
 
@@
 While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now L465--469) and delayed effects on growth (L 445--450, now L475--480), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
-We have added a paragraph to the discussion regarding effects of wet conditions on tropical plants.
+We have added a paragraph (L542) to the discussion regarding effects of wet conditions on tropical plants.
 
@@
 This sentence meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
-We've edited this sentence to read: "The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g.
-monthly) as climate drivers."
+We've edited this sentence to read: "*The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g.
+monthly) as climate drivers.*" (L533).
 
@@
 Of plants that had missing values for some years, but were found again in a subsequent year, 95% had 2 or fewer years of missing values.
-Therefore, plants that were not found for three consecutive surveys, and no subsequent survey, were considered to have died in the transition year after their last observation."*
+Therefore, plants that were not found for three consecutive surveys, and no subsequent survey, were considered to have died in the transition year after their last observation."* (L159--164).
 
@@
 We have changed $y_i$ to $y_{ij}$ and added an explanation of the indexing as suggested.
-We have also corrected the error in the random effect equation so that only one variance is estimated.
+We have also corrected the error in the random effect equation so that only one variance is included.
 $d_{it}$ is intended to indicate that each plant observation is associated with a measure of SPEI for the past 36 months.
@@
 We think that this is an intuitive way to display the results.
-We added a paragraph to the methods explaining our strategy for visualizing the GAM results.
+We added a paragraph to the methods (L260) explaining our strategy for visualizing the GAM results.
 
@@
 This is correct---our expansion was not the fitting of lagged effects with a spline, but the use of a 2 dimensional spline to *also* fit non-linear reaction norms.
-We have edited this sentence accordingly.
+We have edited this sentence accordingly (L436--438).
 
@@
 
-We We have revised this paragraph to clarify.
+We We have revised this paragraph to clarify (L498).
 See our response to \@ref(R2-20).

### 2021-09-08 doc/supplemental.Rmd
added reference to model validation appendix

@@
 ---
-title: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
+title: "Appendix B: Supplemental figures and tables"
+subtitle: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
 author: 

### 2021-09-08 doc/validate_models.Rmd
added figure to supplemental
added reference to model validation appendix

@@
 ---
-title: "Model validation plots"
-author: "Eric R. Scott"
+title: "Appendix A: Model Validation"
+subtitle: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
+author: 
+  - Eric R. Scott
+  - María Uriarte
+  - Emilio M. Bruna
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
+  pdf_document:
+    toc: yes
   html_document:
@@
     toc_float: yes
+linkcolor: blue
 ---
@@
 ```{r eval=TRUE}
+# Function sourced from R/utils.R
 check_res_edf(s_1ha)
 ```
+
 Adequate k for crossbasis.
@@
 
-The crossbasis is much smoother
+The crossbasis is smoother, but shows a similar pattern.
 

### 2021-09-08 doc/paper.Rmd
added reference to model validation appendix
emilio suggestions

@@
 To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs, @gasparriniPenalizedFrameworkDistributed2017].
-DLNMs capture how potentially delayed effects of predictor variables (e.g.
-SPEI) affect an outcome (e.g.
-growth) well beyond the event period.
+DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
 DLNMs avoid the problems of including all possible climate windows in a single model---namely collinearity of predictors and the arbitrary choice of window length.
 They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
-This models a non-linear relationship between predictor and response (e.g.
-between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
+This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
 The shape of the function can be penalized so that it is only as complex as the data supports, thus reducing effective degrees of freedom.
@@
 Penalized cubic regression splines were used for both marginal bases of the crossbasis function.
-We increased *k*, the maximum basis dimension, as necessary to be sure it was adequate.
+We increased *k*, the maximum basis dimension, as necessary to be sure it was adequate (Appendix A).
 Because of penalization, the maximum basis dimension is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
@@
 
-*Flowering:* The overall proportion of plants that flowered was very low.
-The proportion of plants flowering was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]*100`% vs. `r summary_stats$flwr_mean[1]*100`%, respectively).
+*Flowering:* The overall proportion of plants that flowered was very low, but the proportion of plants flowering was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]*100`% vs.
+`r summary_stats$flwr_mean[1]*100`%, respectively).
 The intercepts for the continuous forest and 1 ha fragment models were significantly different, indicating a main effect of habitat on flowering probability (Table \@ref(tab:intercept-table)).
@@
 The most striking differences between habitat types coincided with relatively extreme weather conditions.
-In 2003, during the onset of a severe drought, the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1 ha fragments.
-In 2006, following a wet spring and winter (SPEI \> 1), the trend was reversed when only 14% of reproductive-sized plants flowered in continuous forest and 26% flowered in 1 ha fragments (Figure \@ref(fig:eda)c).
-It is important to keep in mind that 1 ha fragment plots have a lower density of plants on average and fewer reproductive-sized plants compared to continuous forest plots, so total numbers of flowering plants was always lower in fragments.
+In 2003, during the onset of a severe drought, the proportion of potentially reproductive plants (i.e.
+plants above the size-threshold for reproduction) that actually flowered was 28% in continuous forest vs.
+17% in 1-ha fragments.
+In 2006, following a wet spring and winter (SPEI \> 1), the trend was reversed: only 14% of these plants flowered in continuous forest vs. 26% in 1-ha fragments (Figure \@ref(fig:eda)c).
+Note that plots in 1-ha fragments generally have fewer and smaller plants than those in continuous forest, so total numbers of flowering plants was always lower in fragments.
 
@@
 This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
-However, under extreme conditions, survival probability is higher in continuous forest compared to fragments by up to 0.042.
+However, under extreme conditions, survival probability is up to 0.042 higher in continuous forest than fragments.
 
@@
 Interestingly, we found significant detrimental effects of unusually wet conditions at some lag times for all vital rates except survival in continuous forest and flowering probability in 1-ha fragments.
-Although the literature focuses primarily on the effects of droughts in the tropics, some research shows that unusually wet conditions can also be detrimental to tropical plants [@estebanOtherSideDroughts2021].
-Wet conditions could be associated with increased cloud cover which could reduce photosynthetic activity and decrease growth, or with wind storms which could increase tree- and branch-fall---a source of mortality for understory plants.
-Waterlogged soils can also decrease growth and increase mortality [@parentOverviewPlantResponses2008].
+While most studies investigating how precipitation extremes influence tropical plants have focused on droughts [@lewis2010AmazonDrought2011; @phillips2009; @williamsonAmazonianTreeMortality2000], more recent work has shown that unusually wet conditions can also be detrimental [@estebanOtherSideDroughts2021].
+For instance, the increased cloud cover associated with elevated precipitation could lead to reduced photosynthetic activity and growth; plants in the resulting saturated soils could also have lower growth and elevated mortality [@parentOverviewPlantResponses2008].
+The wind storms accompanying extreme precipitation events [@espirito-santoStormIntensityOldgrowth2010; @negron-juarezVulnerabilityAmazonForests2018] could increase the likelihood of tree-, branch-, and litter-fall, all of which are sources of mortality for understory plants [@scariotSeedlingMortalityLitterfall2000; @ssali2019].
 Finally, cool temperatures associated with unusually wet conditions may decrease flowering [@pauCloudsTemperatureDrive2013].

### 2021-09-08 doc/validate_models.Rmd
update validation appendix

@@
 ```{r eval=TRUE}
-# Function sourced from R/utils.R
+check_res_edf <- function(model) {
+  res <- residuals(model)
+  mgcv::gam(res ~ te(spei_history, L, k = c(20, 35), bs = "cs"),
+      gamma = 1.4,
+      data = model.frame(model)) %>%
+    gratia::edf() %>% 
+    dplyr::mutate(edf = round(edf, 1))
+}
+
 check_res_edf(s_1ha)

### 2021-09-08 doc/paper.Rmd
Edit to abstract to fit word limit

@@
 It is often hypothesized that altered environmental conditions in fragments drive declines in reproduction, recruitment, or survivorship.
-The Amazon basin, in addition to experiencing continuing fragmentation, is warming and experiencing changes in precipitation leading to altered frequency and intensity of droughts and unusually wet periods.
+The Amazon basin, in addition to experiencing continuing fragmentation, is experiencing climate change related increases in the frequency and intensity of droughts and unusually wet periods.
 Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear.
@@
 Extremes in SPEI in the previous year reduced survival, drought in the wet season 8--11 months prior to the February census increased growth, and drought two dry seasons prior increased flowering probability.
-The effects of extremes in precipitation on survival and growth were more pronounced in forest fragments compared to continuous forest.
+Effects of extremes in precipitation on survival and growth were more pronounced in forest fragments compared to continuous forest.
 The complex delayed effects of climate and habitat fragmentation in our study point to the importance of long-term demography experiments in understanding the effects of anthropogenic change on plant populations.

### 2021-09-08 _targets.R
rename supplemental

@@ -61,7 +61,7 @@ tar_plan(
   tar_target(g_cf_sub, fit_growth(model_data_cf_sub, k = c(25,5,15))),
   tar_target(f_cf_sub, fit_flwr(model_data_cf_sub, k = c(10,15,15))),
   tar_target(s_cf_sub, fit_surv(model_data_cf_sub, k = c(10,15,15))),
-  tar_render(validate_models, "doc/validate_models.Rmd", deployment = "main"),
+  tar_render(validate_models, "doc/appendix_A_model_validation.Rmd", deployment = "main"),
 
   # Descriptive / Exploratory Data Analysis Figures
   normals = normals_data(),
@@ -141,6 +141,6 @@ tar_plan(
   tar_target(la_file, here("data", "HA-la-stems-ht.xlsx"), format = "file"),
   la_data = read_tidy_la(la_file),
 
-  tar_render(supplemental, "doc/supplemental.Rmd", deployment = "main")
+  tar_render(supplemental, "doc/appendix_B_supplemental_figs.Rmd", deployment = "main")
 
 )

### 2021-09-08 review/Response to reviews.Rmd
reorganize response to reviewers

@@
 ---
+
+Dear Editor,
+
+Thank you for the opportunity to submit a revised version of our manuscript "Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments" to Global Change Biology The feedback we received from the referees was clear, fair, and very helpful, for which we would like to extend our thanks.
+
+Overall the referees were pleased with the submission, but they also had several excellent suggestions for improving the analyses, interpretation, and clarity of our presentation.
+We have carefully gone through the manuscript to incorporate their suggestions.
+We identified three main themes in the reviews:
+
+1) A need for a more in-depth explanation of distributed lag non-linear models, both for a technical audience (for replicability) and for a general ecology audience (for interpretabillity).
+We have expanded and revised the statistical modeling subsection to improve on this area.
+
+2) A need for more discussion of the caveats and limitations of our modeling.
+For example, since we were not able to test for a statistical interaction between habitat and lagged SPEI, we have revised the results to be more clear about which conclusions are strongly supported by statistical results.
+
+3) Finally, reviewer 3's comments made us re-evaluate our approach to limiting the number of "knots" (or maximum basis complexity) in our models.
+This (along with an updated version of the data with some duplicate plants corrected) lead to some cascading changes in our methods, results, and discussion, although the general conclusion that there are lagged effects of SPEI that differ among vital rates and habitats remains unchanged.
+Increasing $k$, the maximum basis complexity, resulted in changes to the shape of the crossbasis smooths, and have ensured that increasing \$k\$ further yet does not result in further changes to the shape of the smooths (see model validation supplement).
+The change in crossbasis shape changed the statistical significance of some lag times for some models.
+This increase in $k$ increased the computational time for models in a way that scales multiplicatively with random effects terms such that it was no longer feasible to include individual-level random effects.
+We ensured that removing these individual-level random effects did not bias the results using some data subsets (now mentioned in the updated methods).
+
+The suggestions from all reviewers really improved the manuscript, and we are very thankful for such thorough and thoughtful reviewers.
+We have addressed all suggestions and comments in detail in our response to reviewer comments below.
+
+Sincerely,
+
+Eric Scott
 

### 2021-09-16 doc/paper.Rmd
change graphics device to ragg

@@
 	fig.path = "figures/",
-	dev = "png",
+	dev = "ragg_png",
 	dpi = 300,
@@
 library(pander)
+library(ragg)
 ```
@@
 
-```{r eda_hires, fig.height=7, fig.width=7, dev='tiff', include=FALSE, dpi=600}
+```{r eda_hires, include=FALSE}
+agg_tiff(here::here("doc", "figures", "eda_hires.tiff"),
+        width = 7, height = 7, units = "in", res = 300)
 tar_read(eda_plot) & theme(text = element_text(size = 10))
+invisible(dev.off())
 ```
@@
 
-```{r surv-curve_hires, fig.height=3.5, fig.width=7, dev="tiff", dpi=600, include=FALSE}
-tar_read(surv_curve) + theme(text = element_text(size = 10))
-```
-
-```{r surv-curve_vect, fig.height=3.5, fig.width=7, dev="pdf", include=FALSE}
+```{r surv-curve_vect, fig.height=3.5, fig.width=7, dev="postscript", include=FALSE}
 tar_read(surv_curve) + theme(text = element_text(size = 10))
@@
 
-```{r size-plot_hires, fig.width=7, fig.height=7, dev='tiff', dpi=600, include=FALSE}
+```{r size-plot_hires, include=FALSE}
+agg_tiff(here::here("doc", "figures", "size-plot_hires.tiff"),
+        width = 7, height = 7, units = "in", res = 300)
 tar_read(size_plot) & theme(text = element_text(size = 10))
+invisible(dev.off())
 ```
@@
 
-```{r surv-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
+```{r surv-spei_hires, include=FALSE}
+agg_tiff(here::here("doc", "figures", "surv-spei_hires.tiff"),
+        width = 7, height = 7.5, units = "in", res = 300)
 tar_read(s_spei_plot) & theme(text = element_text(size = 10))
+invisible(dev.off())
 ```
@@
 
-```{r growth-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
+```{r growth-spei_hires, include=FALSE}
+agg_tiff(here::here("doc", "figures", "growth-spei_hires.tiff"),
+        width = 7, height = 7.5, units = "in", res = 300)
 tar_read(g_spei_plot) & theme(text = element_text(size = 10))
+invisible(dev.off())
 ```
@@
 
-```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='tiff', dpi=600, include=FALSE}
+```{r flwr-spei_hires, fig.height=7.5, fig.width=7, dev='agg_tiff', fig.ext='tiff', dpi=300, include=FALSE}
+agg_tiff(here::here("doc", "figures", "flwr-spei_hires.tiff"),
+        width = 7, height = 7.5, units = "in", res = 300)
 tar_read(f_spei_plot) & theme(text = element_text(size = 10))
+invisible(dev.off())
 ```

### 2021-09-20 doc/paper.Rmd
export fig for graphical abstract
update data availability statement

@@
 National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
-This is publication No. \_\_\_ in the BDFFP Technical Series.
+This is publication No. 824 in the BDFFP Technical Series.
 The authors delcare no conflicts of interest.
@@
 
-Data used in this study are available at the Dryad Digital Repository [links included upon acceptance]; R code for analyses and visualizations are archived with Zenodo at <https://doi.org/10.5281/zenodo.5364591>.
+All data and code for analyses and visualizations are archived with Zenodo at <https://doi.org/10.5281/zenodo.5364591>.
 
@@
 
+<!-- For graphical abstract -->
+
+```{r}
+agg_tiff(here::here("doc", "figures", "surv-spei_graphical-abstract.tiff"),
+        width = 7, height = 8, units = "in", res = 300)
+tar_read(s_spei_plot) & theme(text = element_text(size = 14))
+invisible(dev.off())
+```
+
 ```{r}

### 2021-10-28 README.Rmd
update readme with publication

--- 2021-09-01 README.Rmd
@@
 
-[![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml)
+[![Run Tests](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml/badge.svg)](https://github.com/BrunaLab/HeliconiaDemography/actions/workflows/run-testthat.yaml) [![Zenodo DOI](https://zenodo.org/badge/289078614.svg)](https://zenodo.org/badge/latestdoi/289078614) [![Paper DOI](https://img.shields.io/badge/DOI-10.1111%2Fgcb.15900-blue)](https://doi.org/10.1111/gcb.15900)
 
@@
 
-This repository contains the data and code for a manuscript written for submission to Global Change Biology tentatively titled: "Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments"
+This repository contains the data and code for:
+
+Scott ER, Uriarte M, Bruna EM (2021) Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments. Global Change Biology. https://doi.org/10.1111/gcb.15900
+
+## How to cite
+
+If you use the data or code in this repository, please cite the [Zenodo archive](https://zenodo.org/badge/latestdoi/289078614) as well as the Global Change Biology [paper](https://doi.org/10.1111/gcb.15900).
 
@@
 
-**Using HPC**:
-You may need to install packages on the remote HPC if using `tar_make_clustermq()` with the SSH option.
-Unfortunately, I'm not sure if there is an easy way to do this using `renv`.
-You should, however, make sure the same versions of the packages are installed locally and on the HPC, especially `gratia` which I'm using a development version of that has important bugs fixed, but not new bugs introduced.
-
 ## Additional Dependencies
@@
 You might need to install it for the main text to knit correctly to Word.
-
-<!--
-### Licenses
--->
-<!-- **Text and figures :** [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/) -->
-
-<!-- <!-- **Code :** See the [DESCRIPTION](DESCRIPTION) file -->
-
-<!-- **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse -->
